### 2020-06-25 analysis/paper/26-SAM_asim.Rmd
burn it all

---
title: "SAM simétrico y asimétrico"
author: "Elio Campitelli"
output: 
   # pdf_document:
   #    latex_engine: xelatex
   powerpoint_presentation:
    # ioslides_presentation:
        fig_height: 7.6
        fig_width: 12.8
        reference_doc: "../templates/template.pptx" 
        slide_level: 2
urlcolor: blue
header_includes:
   - \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
# Notification
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(metR)
library(data.table)
library(ggplot2)
library(magrittr)
library(ggh4x)
library(ggperiodic)
library(asymsam)

# source(here("scripts", "helperfun.R"))
# source(here("scripts", "data_locations.R"))
theme_set(theme_asymsam())
ZeroBreaks <- AnchorBreaks(0, NULL, 0)

lev.lab <- AddSuffix(" hPa")
# knitr_set_timer()
# knitr_set_cache(cache.extra = 44)
```

```{r, cache = FALSE}
hgt <- ReadNetCDF(ERA5(), 
                  subset = list(time = c("1979-01-01", "2018-12-31"),
                                latitude = c(-90, 10)),
                  vars = c(hgt = "z")) %>% 
   normalise_coords() %>% 
   .[, hgt := hgt/9.8] %>% 
   .[, dataset := "ERA5"] %>%  
   .[, hgt_a := hgt - mean(hgt), by = .(lon, lat, lev, month(time), dataset)] %>% 
   .[, hgt_m := mean(hgt_a), by = .(lat, lev, time)] %>% 
   .[, hgt_z := hgt_a - hgt_m]
```

# SAM tradicional

(Primera componente principal de las anomalías temporales de altura geopotencial mensual en 700hPA pesado por la raiz del coseno de la latitud)

---

```{r}
SAM <- hgt[lat <= -20] %>% 
   .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
   .[, .(eof = list(eof_flip(EOF(hgt ~ time | lon + lat, n = 1, data = .SD)))), 
     by = .(dataset, lev)]
```

```{r, fig.cap = "Patrón espacial del SAM. ERA5 en sombrado, NCEP en contornos."}
SAM[lev == 700, eof[[1]]$right, by = dataset] %>% 
   periodic(lon = c(0, 360)) %>% 
   ggplot(aes(lon, lat)) +
   geom_contour_fill(aes(z = hgt), breaks = ZeroBreaks) +
   geom_qmap(~.x[lat < -20]) +
   scale_x_longitude() +
   scale_y_latitude() +
   scale_fill_divergent() +
   coord_polar()
```

```{r, fig.cap = "Proyección de la anomalía de altura geopotencial en el índice SAM de 700hPa -ERA5. (Comparar con Baldwin y THompson 2009 figura 8, columna izquierda)"}
levs <- c(700, 300, 30, 3)

SAM[lev == 700, eof[[1]]$left, by = .(dataset)] %>% 
   .[dataset == "ERA5"] %>% 
   setnames("hgt", "SAM") %>% 
   .[, SAM := SAM/sd(SAM), by = .(dataset, PC)] %>% 
   .[hgt[lev %in% levs & dataset == "ERA5" & lat <= -20], on = .NATURAL] %>% 
   .[, FitLm(hgt_a, SAM), by = .(lon, lat, lev)] %>% 
   rm_intercept() -> proj

proj %>% 
   periodic(lon = c(0, 360)) %>% 
   ggplot(aes(lon, lat)) +
   geom_contour_fill(aes(z = estimate), global.breaks = FALSE, breaks = ZeroBreaks) +
   geom_qmap(~.[lat < -20]) +
   geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
                 global.breaks = FALSE, breaks = ZeroBreaks) +
   # geom_text_contour(aes(z = estimate), stroke = 0.1, rotate = FALSE, 
   #                   global.breaks = FALSE, breaks = ZeroBreaks) +
   scale_x_longitude() +
   scale_y_latitude(limits = c(-90, -20)) +
   scale_fill_divergent() +
   coord_polar() +
   facet_wrap(lev~., ncol = 2, labeller = labeller(lev = lev.lab)) 
```

# Método BT09

Método:

1. Calcular anomalías tamporales de geopotencial del lado polar de 20°.

2. Calcular el promedio zonal de las anomalías temporale de geopotencial en cada nivel.

3. Hacer EOF de este promedio zonal **de forma independiente para cada nivel** (y pesando por la raiz del coseno de la latitud).

Datos: ERA5 y NCEP para el período 1979-2018

---

```{r}
bt09 <- hgt[lat <= -20, .(hgt = mean(hgt_a)), by = .(lat, lev, time, dataset)] %>% 
   .[, hgt := -hgt*sqrt(cos(lat*pi/180))] %>% 
   .[, .(eof = list(EOF(hgt ~ time | lat, n = 1:2, data = .SD))), by = .(lev, dataset)]
```

```{r}
extremes <- bt09[, eof[[1]]$sdev, by = .(dataset, lev)] %>% 
   .[PC == "PC1"] %>% 
   .[, rbind(.SD[which.min(r2)], 
             .SD[lev > 100][which.max(r2)],
             .SD[lev <= 100][which.max(r2)],
             .SD[lev == 700]), by = .(dataset)] 
```

```{r, fig.cap = "Varianza explicada por el EOF1 y EOF2 para cada nivel."}
bt09[, eof[[1]]$sdev, by = .(dataset, lev)] %>% 
   ggplot(aes(lev, r2)) +
   geom_line(aes(color = PC)) +
   ggrepel::geom_text_repel(data = extremes, aes(label = AddSuffix(" hPA")(lev))) +
   scale_x_level() +
   coord_flip() +
   facet_grid(~dataset)
```

```{r, fig.cap = "Patŕon espacial de altura geopotencial media zonal asociado a cada nivel. ERA5 en sombrado, NCEP en contornos."}
bt09 %>% 
   .[, eof[[1]]$right, by = .(dataset, lev)] %>% 
   copy() %>% 
   # .[dataset == "ERA5"] %>%
   .[PC == "PC2" & dataset == "NCEP", hgt := hgt*sign(hgt[lat == -45]), by = .(dataset, lev)] %>%
   .[PC == "PC2" & dataset == "ERA5", hgt := hgt*sign(hgt[lat == -30]), by = .(dataset, lev)] %>%
   # .[PC == "PC1"] %>% 
   ggplot(aes(lat, lev)) +
   geom_contour_fill(aes(z = hgt)) +
   geom_hline(data = extremes, aes(yintercept = lev)) +
   # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = hgt)) +
   scale_y_level() + 
   scale_x_latitude(ticks = 15) +
   scale_fill_divergent() +
   facet_grid(dataset~PC)
```

```{r, fig.cap = "Igual que la figura anterior pero con los patrones espaciales multiplicados por el valor de sus respectivos valores principales."}
bt09 %>% 
   .[, denormalise(eof[[1]], "right"), by = .(dataset, lev)] %>% 
   copy() %>% 
   # .[dataset == "ERA5"] %>%
   .[PC == "PC2" & dataset == "NCEP", hgt := hgt*sign(hgt[lat == -45]), by = .(dataset, lev)] %>%
   .[PC == "PC2" & dataset == "ERA5", hgt := hgt*sign(hgt[lat == -30]), by = .(dataset, lev)] %>%
   # .[PC == "PC1"] %>% 
   ggplot(aes(lat, lev)) +
   geom_contour_fill(aes(z = hgt)) +
   geom_hline(data = extremes, aes(yintercept = lev)) +
   # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = hgt)) +
   scale_y_level() + 
   scale_x_latitude(ticks = 15) +
   scale_fill_divergent() +
   facet_grid(dataset~PC)
```

```{r, fig.cap = "EOF1 para los niveles marcados en la figura anterior."}
bt09 %>% 
   .[extremes[, .(lev, dataset)], on = .NATURAL] %>% 
   .[, eof[[1]]$right, by = .(dataset, lev)] %>%
   .[PC == "PC1"] %>% 
   ggplot(aes(lat, hgt)) +
   geom_line(aes(color = factor(lev))) +
   scale_x_latitude(ticks = 15) +
   facet_grid(~dataset) 
```

```{r, fig.cap = "Correlación cruzada entre las series temporales de SAM simétrico en cada nivel."}
bt09 %>% 
   .[dataset == "ERA5"] %>% 
   .[, eof[[1]]$left, by = .(lev, dataset)] %>%
   .[PC == "PC1"] %>% 
   .[, widyr::pairwise_cor(.SD, lev, time, hgt), by = .(dataset, PC)] %>% 
   ggplot(aes(item1, item2)) +
   geom_contour_fill(data = ~.x[dataset == "ERA5"], aes(z = correlation), na.fill = TRUE) +
   scale_fill_divergent() +
   scale_x_level() +
   scale_y_level() +
   coord_equal()
```

```{r, fig.cap = "Covarianza de la anomalía de altura geopotencial y el índice SAM simétrico en cada nivel -ERA5. (Comparar con Baldwin y THompson 2009 figura 8, columna derecha)"}
bt09_2d <- bt09 %>% 
   .[, eof[[1]]$left, by = .(lev, dataset)] %>% 
   # .[PC == "PC1"] %>% 
   setnames("hgt", "EOF") %>%
   .[hgt[lat <= -20 & lev %in% unique(c(levs, extremes$lev))], on = .NATURAL, allow.cartesian = TRUE] %>% 
   .[, FitLm(hgt, EOF = EOF/sd(EOF)), by = .(lon, lat, lev, dataset, PC)] %>% 
   rm_intercept()

bt09_2d %>% 
   .[lev %in% levs] %>% 
   .[dataset == "ERA5"] %>% 
   periodic(lon = c(0, 360)) %>% 
   ggplot(aes(lon, lat)) +
   geom_qmap(~.x[lat <= -20]) +
   geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
   scale_x_longitude() +
   scale_y_latitude(limits = c(NA, -20)) +
   coord_polar() +
   facet_grid(PC~lev,  labeller = labeller(lev =  lev.lab))

```

```{r}
last_plot()
```

```{r bt09, fig.cap = "Igual a la figura anterior pero para los niveles extremos"}
bt09_2d %>% 
   .[extremes, on = .NATURAL] %>% 
   .[dataset == "ERA5"] %>% 
   periodic(lon = c(0, 360)) %>% 
   ggplot(aes(lon, lat)) +
   geom_qmap(~.x[lat <= -20]) +
   geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
   scale_x_longitude() +
   scale_y_latitude(limits = c(NA, -20)) +
   coord_polar() +
   facet_grid(PC~lev,  labeller = labeller(lev =  lev.lab)) 
```

# Separando simétrico y asimétrico

El método de BT09 define el modo anular usando el geopotencial medio, pero no es simple ver cómo usarlo para separar la parte simétrica de la asimétrica. Lo que voy a hacer es otra cosa.

1. Computar el patrón SAM de forma "tradicional" usando las anomalías temporales de geopotencial en 700hPa.

2. Calcular la parte simétrica y la parte asimétrica de ese patrón.

3. Para cada campo de anomalía de geopotencial, calcular la regresión múltiple lineal con la parte simétrica y la asimétrica del patrón. Es decir, si SAM_sym y SAM_asym son los patrones simétricos y asimétricos del SAM computados en el paso 2 y HGT es el campo de anomalías de geopotencial en 700hPa de una determinada fecha, hacer la regresión HGT ~ a\*SAM_sym + b\*SAM_asym. (La regresión pesada por el coseno de la latitud).

4. La serie temporal de a y b son los índices de SAM simétrico y asimétrico.

```{r, fig.cap = "Parte simétrica y parte asimétrica del patrón de SAM en el hemisferio sur."}
sam_sep <- SAM[, eof[[1]]$right, by = .(dataset, lev)] %>% 
   setnames("hgt", "full") %>% 
   .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev, dataset)]

lab_sam <-  c(full = "Completo", 
              asym = "Asimétrico",
              sym  = "Simétrico")

sam_sep %>% 
   .[dataset == "ERA5"] %>% 
   .[lev %in% levs] %>%
   rm_singleton() %>%
   melt(id.vars = c("lon", "lat", "lev")) %>% 
   .[variable == "asym", value := value + rnorm(.N, 0, sd = 1e-6)] %>%
   .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)] %>% 
   periodic(lon = c(0, 360)) %>% 
   ggplot(aes(lon, lat)) +
   # geom_contour2(aes(z = value, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
   geom_contour_fill(aes(z = value), global.breaks = FALSE) +
   geom_qmap() +
   scale_x_longitude() +
   scale_y_latitude(limits = c(NA, -20)) +
   scale_fill_divergent() +
   coord_polar() +
   facet_grid(variable~lev, labeller = labeller(variable = lab_sam, lev = lev.lab)) 
```

```{r, fig.cap = "Igual que la figura anterior pero para el hemisferio norte"}
# sam_sep %>% 
#    .[dataset == "ERA5"] %>% 
#    .[as.numeric(hemisphere) == 1] %>%
#    .[lev %in% levs] %>% 
#    rm_singleton() %>%
#    melt(id.vars = c("lon", "lat", "lev")) %>% 
#    .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)] %>% 
#    periodic(lon = c(0, 360)) %>% 
#    ggplot(aes(lon, lat)) +
#    geom_contour_fill(aes(z = value), global.breaks = FALSE) +
#    # geom_contour2(aes(z = value, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
#    geom_qmap(lat >= 20) +
#    scale_x_longitude(trans = scales::reverse_trans()) +
#    scale_y_latitude(trans = scales::reverse_trans()) +
#    scale_fill_divergent() +
#    coord_polar() +
#    facet_grid(variable~lev, labeller = labeller(variable = lab_sam, lev = lev.lab)) 
```

```{r, fig.cap = "Relación entre el EOF1 calculado acá y el índice AAO provisto por NCEP"}
aao <- readRDS(data_path("derived"))

SAM[lev == 700, eof[[1]]$left, by = .(dataset)] %>%
   .[dataset == "ERA5"] %>%
   .[aao, on = "time"] %>%
   ggplot(aes(hgt, aao)) +
   geom_point() +
   labs(x = "EOF1", y = "AAO")
```

```{r}
indexes <- sam_sep %>% 
   hgt[., on = .NATURAL]

full <- indexes[, FitLm(hgt_a, full, weights = cos(lat*pi/180), r2 = TRUE),
                by = .(dataset, lev, time)] %>% 
   rm_intercept()

pcor <- indexes[, partial_cor(hgt_a, sym, asym, weights = cos(lat*pi/180)), 
                by = .(dataset, lev, time)]

sep <- indexes[, FitLm(hgt_a, sym, asym, weights = cos(lat*pi/180), r2 = TRUE), 
               by = .(dataset, lev, time)] %>% 
   rm_intercept()

indexes <- rbind(full, sep) %>% 
   .[, estimate_norm := estimate/sd(estimate[term == "full"]), by = .(lev, dataset)] %>% 
   .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
   pcor[., on = .NATURAL]

```

```{r, fig.cap = "Correlación entre el BT09 y las partes del SAM."}
bt09[, eof[[1]]$left, by = .(lev, dataset)] %>%
   .[PC == "PC1"] %>% 
   .[indexes, on = .NATURAL, allow.cartesian = TRUE] %>% 
   .[, cor(hgt, estimate), by = .(lev, dataset, term)] %>% 
   ggplot(aes(lev, V1)) +
   geom_line(aes(color = term)) +
   scale_x_level() +
   # scale_y_continuous(NULL, limits = c(0, 1)) +
   scale_color_brewer("Parte", palette = "Set1", labels = lab_sam) +
   coord_flip(ylim = c(0, 1))
```

```{r, fig.cap="Series temporales del SAM simétrico y el SAM asimétrico para cada dataset (normalizados usando el desvio estándar de la serie completa.)"}
indexes %>% 
   .[lev == 700 & term != "full"] %>% 
   ggplot(aes(time, estimate_norm)) +
   geom_line(aes(color = term)) +
   # geom_smooth(method = "lm", aes(color = term)) +
   scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
   scale_y_continuous(NULL) +
   scale_x_datetime(NULL)
```

```{r}
indexes %>% 
   .[lev == 700 & term != "sym"] %>% 
   ggplot(aes(time, r.squared)) +
   geom_line(aes(color = term)) +
   # geom_smooth(method = "lm", aes(color = term)) +
   scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
   scale_y_continuous(NULL) +
   scale_x_datetime(NULL) 
```

```{r, fig.cap = "Correlaciones cruzadas de los índices entre niveles para el hemisferio sur."}
indexes %>% 
   .[dataset == "ERA5"] %>%
   .[, widyr::pairwise_cor(.SD, lev, time, estimate), by = .(dataset, term)] %>% 
   ggplot(aes(item1, item2)) +
   geom_contour_fill(aes(z = correlation), na.fill = 1) +
   scale_fill_divergent() +
   scale_x_level() +
   scale_y_level() +
   coord_equal() +
   facet_grid(dataset ~ term, labeller = labeller(term = lab_sam))
```

```{r, fig.cap = "Promedio mensual del valor absoluto de cada componente SAM"}
indexes %>% 
   .[, .(estimate = mean(abs(estimate_norm))), by = .(dataset, lev, term, month(time))] %>% 
   periodic(month = 1:13) %>% 
   ggplot(aes(month, lev)) +
   geom_contour_fill(aes(z = estimate)) +
   scale_y_level() +
   scale_x_continuous(breaks = 1:13, expand = c(0, 0), 
                      labels = c(month.abb, month.abb[1])) +
   scale_fill_viridis_c() +
   facet_grid(term ~ ., labeller = labeller(term = lab_sam)) 
```

```{r, fig.cap = "Promedio mensual de abs(sym) / (abs(sym) + abs(asym))."}
indexes_wide <- dcast(indexes, dataset + lev + time ~ term, value.var = "estimate_norm")

indexes_wide %>% 
   copy() %>% 
   .[, mean(abs(sym)/(abs(sym) + abs(asym))), by = .(dataset, lev, month(time))] %>% 
   periodic(month = 1:13) %>% 
   ggplot(aes(month, lev)) +
   geom_contour_fill(aes(z = V1)) +
   scale_y_level() +
   scale_x_continuous(breaks = 1:13, expand = c(0, 0), labels = c(month.abb, month.abb[1])) +
   scale_fill_viridis_c()  
```

```{r, fig.cap = "Regresión entre el SAM completo y el simétrico o el asimétrico"}
vertical_sym <- indexes %>% 
   # .[lev %in% levs] %>%
   .[dataset == "ERA5"] %>%
   # .[, estimate := estimate/sd(estimate), by = .(dataset, lev, hemisphere, term)] %>% 
   dcast(dataset + time  + lev ~ term, value.var = "estimate") %>% 
   .[, FitLm(full, asym, sym, se = TRUE), by = .(dataset, lev)] %>% 
   rm_intercept() %>% 
   .[term == "sym"] 
# vertical_sym[, .SD[which.max(estimate)], by = .(hemisphere, dataset)]
# %>%
extremes_sym <- vertical_sym %>% 
   .[, rbind(.SD[which.min(estimate)], 
             .SD[lev > 200][which.max(estimate)],
             .SD[lev > 200][which.min(estimate)],
             .SD[lev <= 200][which.max(estimate)],
             .SD[lev == 700]), by = .(dataset)] %>% 
   unique()

vertical_sym %>% 
   ggplot(aes(lev, estimate)) +
   geom_line() +
   geom_text(data = extremes_sym, aes(label = AddSuffix(" hPa")(lev))) +
   geom_hline(yintercept = 0.5, linetype = 2) +
   scale_x_level() +
   scale_y_continuous("Simetría", limits  = c(0, 1)) +
   coord_flip()
```

```{r, fig.cap = "Relación entre índice de SAM completo y los índices separados."}
indexes %>% 
   .[extremes_sym[, .(dataset, lev)], on = .NATURAL] %>%
   # .[lev %in% extremes_sym$lev] %>%
   .[dataset == "ERA5"] %>% 
   .[, estimate := estimate/sd(estimate), by = .(dataset, lev, term)] %>% 
   dcast(dataset + time  + lev ~ term, value.var = "estimate") %>% 
   tidyr::pivot_longer(c("asym", "sym")) %>% 
   ggplot(aes(full, value, color = name)) +
   geom_point() +
   geom_smooth(method = "lm") +
   scale_x_continuous("SAM completo") +
   scale_y_continuous("Índices separados") +
   scale_color_brewer(palette = "Set1") +
   facet_nested(lev ~ ., scales = "free")
```

```{r, fig.cap = "Coeficientes anteriores vs. varianza de cada patrón / varianza del patrón completo."}
sam_sep %>% 
   .[, .(sym = var(sym*cos(lat*pi/180))/var(full*cos(lat*pi/180))),
     by = .(dataset, lev)] %>% 
   # .[lev %in% levs] %>% 
   melt(id.vars = c("dataset", "lev"), variable.name = "term", value.name = "estimate2")  %>% 
   vertical_sym[., on = .NATURAL] %>% 
   ggplot(aes(estimate, estimate2)) +
   geom_point() +
   geom_abline() + 
   scale_x_continuous("Simetría.") +
   scale_y_continuous("var(patrón simétrico) / var(patrón total)") +
   coord_equal()
```

```{r, fig.cap = "Covarianza de las anomalías temporales de geopotencial con el índice correspondiente a cada nivel para el hemisferio sur."}

indexes_2d <- hgt[extremes, on = .NATURAL] %>% 
   indexes_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
   .[, rbind(as.data.table(FitLm(hgt_a, sym, asym)),
             as.data.table(FitLm(hgt_a, full))),
     by = .(lon, lat, dataset, lev)] %>% 
   rm_intercept() %>% 
   periodic(lon = c(0, 360)) %>% 
   .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)]

indexes_2d %>% 
   .[lat <= -20] %>% 
   ggplot(aes(lon, lat)) +
   geom_contour_fill(aes(z = estimate), global.breaks = FALSE) +
   geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), size = 0.3,
                 global.breaks = FALSE) +
   geom_qmap(~.x[lat <= -20]) +
   # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
   scale_x_longitude() +
   scale_y_latitude(limits = c(NA, -20)) +
   scale_fill_divergent() +
   coord_polar() +
   facet_grid(term ~ lev, labeller = labeller(term = lab_sam, lev = lev.lab))
```

```{r, fig.cap = "Regresión con anomalía mensual de OLR de los índices para distintos niveles en el hemisferio SUR."}
index_wide <- indexes %>% 
   dcast(dataset + time + lev ~ term, value.var = "estimate_norm")

# OLR() %>% 
#    ReadNetCDF(subset = list(lat = -90:10)) %>% 
#    na.omit() %>% 
#    .[, olr_a := Anomaly(olr), by = .(lon, lat, month(time))] %>% 
#    .[index_wide[lev %in% c(50, 700) & dataset == "ERA5"], 
#      on = "time", allow.cartesian = TRUE] %>% 
#    .[, FitLm(olr_a, asym, sym, se = TRUE), by = .(lon, lat, lev, dataset)] %>%
#    rm_intercept() %>% 
#    .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, lev, dataset)] %>% 
#    .[dataset == "ERA5"] %>% 
#    ggplot(aes(lon, lat)) +
#    geom_contour_fill(aes(z = estimate)) +
#    geom_qmap() +
#    geom_contour2(aes(z = p.val), breaks = 0.05) +
#    stat_subset(aes(subset = p.val <= 0.05 & is.cross(lon, lat)), size = 0.5, alpha = 0.5) +
#    scale_x_longitude() +
#    scale_y_latitude() +
#    scale_fill_divergent() +
#    coord_quickmap(ylim = c(-90, 10)) +
#    facet_grid(term~lev)
```

```{r}
pcor %>% 
   .[lev == 700 & term != "full"] %>% 
   ggplot(aes(time, partial_correlation^2)) +
   geom_line(aes(color = term)) +
   # geom_smooth(method = "lm", aes(color = term)) +
   scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
   scale_y_continuous(NULL) +
   scale_x_datetime(NULL) +
   facet_grid(. ~ dataset)
```

```{r}
pcor %>% 
   .[, mean(partial_correlation^2), by = .(dataset, lev, term, month(time))] %>% 
   ggplot(aes(lev, month)) +
   geom_contour_fill(aes(z = V1)) +
   scale_x_level() +
   coord_flip() +
   facet_grid(term ~ .)
```

```{r}
indexes %>%
   copy() %>% 
   .[term != "full"] %>% 
   .[, denom := sum(partial_correlation^2), by = .(dataset, lev, time)] %>% 
   .[, mean(partial_correlation^2/denom), by = .(dataset, lev, term)] %>%
   ggplot(aes(lev, V1)) +
   geom_line(aes(color = term)) +
   scale_x_level() +
   scale_y_continuous(limits = c(0, 1)) +
   scale_color_brewer(palette = "Set1", labels = lab_sam) +
   coord_flip()
```

```{r}
indexes %>% 
   .[term != "full"] %>%
   .[, mean(partial_correlation^2), by = .(dataset, lev, month(time), term)] %>% 
   ggplot(aes(month, lev)) +
   geom_contour_fill(aes(z = V1)) +
   scale_y_level() +
   scale_x_continuous(expand = c(0, 0), 
                      breaks = 1:12, 
                      labels = month.abb) +
   scale_fill_viridis_c() +
   facet_grid(term ~ .)
```

```{r}
indexes %>% 
   .[term == "sym"] %>% 
   .[, mean(partial_correlation^2/r.squared), by = .(dataset, lev, month(time), term)] %>% 
   ggplot(aes(month, lev)) +
   geom_contour_fill(aes(z = V1)) +
   scale_x_continuous(expand = c(0, 0), 
                      breaks = 1:12, 
                      labels = month.abb) +
   scale_y_level() +
   scale_fill_divergent(midpoint = 0.5) 
```

```{r}
dcast(pcor, dataset + lev + time ~ term, value.var = "partial_correlation") %>% 
   .[, cor(asym^2, sym^2), by = .(dataset, lev)] %>% 
   ggplot(aes(lev, V1)) +
   geom_line() +
   scale_x_level() +
   coord_flip()
```

### 2020-06-25 analysis/paper/28-SAM-cEOF.Rmd
burn it all

---
title: "SAM y cEOF"
author: "Elio Campitelli"
output: 
   # pdf_document:
   #    latex_engine: xelatex
   powerpoint_presentation:
    # ioslides_presentation:
        fig_height: 7.6
        fig_width: 12.8
        reference_doc: template.pptx
        slide_level: 2
urlcolor: blue
header_includes:
   - \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
# Notification

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(metR)
library(data.table)
library(ggplot2)
library(magrittr)
library(patchwork)
library(here)
library(ggh4x)
library(glue)
library(ggperiodic)
library(gt)
here <- here::here

source(here("scripts", "helperfun.R"))
source(here("scripts", "data_locations.R"))

hemisphere <- function(lat) {
  factor(ifelse(lat < 0, "sh", "nh"), levels = c("nh", "sh"),
         labels = c("Hemisferio norte", 
                    "Hemisferio sur"), 
         ordered = TRUE)
}

mean_season <- function(data) {
  data %>% 
    .[is.full_season(time)] %>%  
    .[, lapply(.SD, mean), by = .(time = seasonally(time))]   
}

table_elio <- function(data) {
  gt::tab_options(data,
                  table_body.hlines.width = 0, 
                  table_body.vlines.width = 0,
                  row_group.font.weight = "lighter", 
                  column_labels.font.weight = "bold")

}

lab_sam <-  c(full = "Completo", 
              asym = "Asimétrico",
              sym  = "Simétrico")

ZeroBreaks <- AnchorBreaks(0, NULL, 0)
knitr_set_timer()
knitr_set_cache(cache.extra = 40)
```

```{r}
hgt <- ReadNetCDF(ERA5(), vars = auto_scale(hgt = "z"), 
                  subset = list(lev = list(50, 200))) %>% 
  .[, hgt := hgt/9.8] %>% 
  .[, hgt_a := Anomaly(hgt), by = .(lon, lat, lev, month(time))]
```

```{r}
sam <-  hgt[lat %between% c(-80, -20), eof_asym(hgt_a, lon, lat, time, n = 1), by = lev]
```

```{r, fig.cap = "Regresión de altura geopotencial e índices de SAM simétrico y asimétrico."}
regr <- hgt[dcast(sam, time + lev + PC ~ term, value.var = "estimate_norm"), 
            on = .NATURAL, allow.cartesian = TRUE] %>% 
  .[lat <= -20] %>% 
  .[, FitLm(hgt_a, sym, asym), by = .(lon, lat, lev)] %>% 
  rm_intercept()

regr %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
  geom_map2(lat <= -20) + 
  scale_x_longitude() +
  scale_y_latitude() +
  scale_fill_divergent() +
  coord_polar() +
  facet_grid(lev~term, labeller = labeller(term = lab_sam, 
                                           lev = AddSuffix(" hPa")))
```

```{r, fig.cap = "Regresión de la altura geopotencial con el índice SAM asimétrico en sombreado y el simétrico en contornos."}
sam_regr <- sam %>% 
  .[term != "full"] %>% 
  dcast(time + lev ~ term, value.var = "estimate_norm") %>% 
  .[hgt[lat <= -20], on = .NATURAL] %>% 
  .[is.full_season(time)] %>%  
  .[, lapply(.SD, mean), by = .(lon, lat, lev, time = seasonally(time))] %>% 
  .[, FitLm(hgt_a, sym, asym, se = TRUE), by = .(lon, lat, lev, season(time))]

sam_regr %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = estimate), data = ~.x[term == "asym"], 
                    breaks = ZeroBreaks, global.breaks = FALSE) +
  geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
                data = ~.x[term == "sym"], breaks = ZeroBreaks, global.breaks = FALSE) +
  geom_map2(lat <= -20) +
  scale_x_longitude() +
  scale_y_latitude() +
  scale_fill_divergent(guide = "none") +
  coord_polar() +
  facet_grid(lev ~ season, labeller = labeller(lev = AddSuffix(" hPa")))
```

```{r fig.cap = "Igual que la figura anterior, pero computando el SAM para cada estación por separado."}
# hgt[lat <= -20] %>% 
#   .[is.full_season(time)] %>%  
#   .[, lapply(.SD, mean), by = .(lon, lat, lev, time = seasonally(time))] 
# 
# sam_season <-  hgt[lat %between% c(-80, -20)] %>% 
#   .[, season := season(time)] %>% 
#   .[, eof_asym(hgt_a, lon, lat, time, n = 1), by = .(lev, season)]
# 
# hgt_season <- hgt[lat <= -20] %>% 
#   .[is.full_season(time)] %>%  
#   .[, lapply(.SD, mean), by = .(lon, lat, lev, time = seasonally(time))] 
# 
# sam_season %>% 
#   .[term != "full"] %>% 
#   dcast(time + lev ~ term, value.var = "estimate_norm") %>% 
#   .[hgt_season, on = .NATURAL] %>% 
#   .[, FitLm(hgt_a, sym, asym, se = TRUE), by = .(lon, lat, lev, season(time))] %>% 
#   periodic(lon = c(0, 360)) %>% 
#   ggplot(aes(lon, lat)) +
#   geom_contour_fill(aes(z = estimate), data = ~.x[term == "asym"], 
#                     breaks = ZeroBreaks, global.breaks = FALSE) +
#   geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
#                 data = ~.x[term == "sym"], breaks = ZeroBreaks, global.breaks = FALSE) +
#   geom_map2(lat <= -20) +
#   scale_x_longitude() +
#   scale_y_latitude() +
#   scale_fill_divergent(guide = "none") +
#   coord_polar() +
#   facet_grid(lev ~ season, labeller = labeller(lev = AddSuffix(" hPa")))
```

Todo lo que sigue es usando el SAM coomputado para todo el año.

---

Para los siguientes gráficos, una nota sobre la metodología:

Para el cálculo de SAM simétrico y asimétrico hago una regresión múltiple. De ahí obtengo un R2 que es la varianza del campo de geopotencial explicada por el modelo hgt ~ sam_simetrico + sam_asimetrico. Ese R2 lo llamo "R2 total". Luego, también calculo los coeficientes de correlación parcial entre cada parte y el campo total. El cuadrado de esa variable lo llamo "R2 parcial".

La razón entre el r2 parcial de la parte asimétrica y el r2 total puede servir como una estimación del nivel de simétría de cada evento SAM.

La suma de R2 parciales NO es igual al R2 total.

Todo esto es un Work In Progress. Creo que es mejor pensar en un ANOVA.

```{r, fig.cap = "Relación entre R2 total y la suma de R2 parciales. Notar que la suma de parciales puede ser mayor a 1."}
sam %>% 
  copy() %>% 
  .[term != "full"] %>% 
  .[, .(r_parcial = sum(partial.r.squared),
        r_total = r.squared[1]), by = .(lev, time)] %>% 
  ggplot(aes(r_parcial, r_total)) +
  geom_point() +
  scale_x_continuous("Suma de r2 parciales") +
  scale_y_continuous("R2 total")
```

```{r, fig.cap = "Coeficiente de determinación parcial medio para cada parte del sam."}
sam %>% 
  copy() %>% 
  .[term != "full"] %>% 
  .[, mean(partial.r.squared), by = .(lev, term, month(time))] %>% 
  ggplot(aes(month, V1)) +
  geom_line(aes(color = term)) +
  scale_color_brewer(palette = "Set1", labels = lab_sam) +
  scale_x_continuous(breaks = 1:12, labels = month.abb_sp) +
  facet_grid(lev ~., scales = "free", labeller = labeller(lev = AddSuffix(" hPa")))
```

```{r, fig.cap = "R2 parcial / r2 total medio para cada parte"}
sam %>% 
  copy() %>% 
  .[term != "full"] %>% 
  .[, mean(partial.r.squared/r.squared), by = .(lev, term, month(time))] %>% 
  ggplot(aes(month, V1)) +
  geom_hline(yintercept = 0.5) +
  geom_line(aes(color = term)) +
  scale_color_brewer(palette = "Set1", labels = lab_sam) +
  scale_x_continuous(breaks = 1:12, labels = month.abb_sp) +
  facet_grid(lev ~., scales = "free", labeller = labeller(lev = AddSuffix(" hPa")))
```

# Relación con el EOF complejo

```{r}
ceof <- hgt[, compute_ceof(hgt, lon, lat, lev, time, temporal = FALSE, lats.eof = c(-80, -20))]
```

```{r, fig.cap = "Parte real de las componentes principales complejas conjuntas en sombreado y regresión de la paret asimétrica del primer EOF en contornos."}
ceof$right %>% 
  copy() %>% 
  .[, hgt := Re(hgt)] %>% 
  .[, hgt := hgt + rnorm(.N, sd = 1e-7)] %>% 
  .[regr[term != "sym"], on = .NATURAL, allow.cartesian = TRUE] %>% 
  na.omit() %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = hgt), breaks = ZeroBreaks, global.breaks = FALSE) +
  geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
  geom_map2(lat <= -20) +
  scale_x_longitude() +
  scale_y_latitude() +
  scale_fill_divergent() +
  facet_grid(lev ~ PC) +
  coord_polar()

```

```{r, fig.cap = "Parte imaginaria de las componentes principales complejas conjuntas en sombreado y regresión de la paret asimétrica del primer EOF en contornos."}
ceof$right %>% 
  copy() %>% 
  .[, hgt := Im(hgt)] %>% 
  .[, hgt := hgt + rnorm(.N, sd = 1e-7)] %>% 
  .[regr[term != "sym"], on = .NATURAL, allow.cartesian = TRUE] %>% 
  na.omit() %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = hgt), breaks = ZeroBreaks, global.breaks = FALSE) +
  geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
  geom_map2(lat <= -20) +
  scale_x_longitude() +
  scale_y_latitude() +
  scale_fill_divergent() +
  facet_grid(lev ~ PC) +
  coord_polar()
```

```{r, fig.cap = "r2 entre eof complejo y sam asimétrico para cada PC y parte real e imaginaria."}
sam_asym <- sam[term == "asym"] %>% 
  rm_singleton()

lm_regr <- ceof$left %>% 
  sep_ReIm(hgt, FALSE) %>% 
  .[sam_asym, on = .NATURAL] %>% 
  .[, FitLm(estimate_norm, Real = R, Imaginary = I, se = TRUE, r2 = TRUE), by = .(PC, lev)] %>% 
  rm_intercept()

ceof$left %>% 
  sep_ReIm(hgt) %>% 
  .[copy(sam)[, PC := NULL], on = .NATURAL, allow.cartesian = TRUE] %>% 
  .[term == "asym"] %>% 
  .[, item := interaction(PC, part)] %>% 
  .[, cor(estimate, hgt)^2, by = .(item, lev)] %>%  
  .[, c("PC", "part") := tstrsplit(as.character(item), split = "\\.")] %>% 
  # lm_regr[., on = c("lev", term = "part", "PC")] %>% 
  dcast(lev + PC ~ part, value.var = "V1") %>% 
  .[lm_regr[, .(full = r.squared[1]), by = .(PC, lev)], on = .NATURAL] %>% 
  gt::gt(rowname_col = "PC", groupname_col = c("lev"))  %>% 
  gt::fmt_number(vars(Imaginary, Real, full), decimals = 2) %>% 
  gt::cols_label(full = "Im + Real") %>% 
  gt::tab_footnote("R2 de la regresión múltiple", locations = gt::cells_column_labels("full")) %>% 
  table_elio() %>% 
  gt::gtsave("tabla1.png", path = "/home/elio.campitelli/onda3/fig/28-SAM-cEOF/") %>%
  knitr::include_graphics()
```

```{r, fig.cap = "Coeficientes de regresión múltiple para el modelo asym ~ im + real."}
lm_regr %>% 
  copy() %>% 
  .[, c("estimate", "std.error") := .(round(estimate, 1), 2*signif(std.error, 2))] %>%
  dcast(PC + lev ~ term, value.var = c("estimate", "std.error")) %>% 
  gt::gt(rowname_col = "PC", groupname_col = c("lev")) %>% 
  gt::cols_merge_uncert(col_val = "estimate_Imaginary", 
                        col_uncert = "std.error_Imaginary") %>% 
  gt::cols_merge_uncert(col_val = "estimate_Real", 
                        col_uncert = "std.error_Real") %>% 
  gt::cols_label(estimate_Imaginary = "Imaginary", 
                 estimate_Real = "Real") %>% 
  table_elio() %>% 
  gt::gtsave("tabla2.png", path = "/home/elio.campitelli/onda3/fig/28-SAM-cEOF/") %>%
  knitr::include_graphics()
```

### Conclusión

* La parte asimétrica del SAM se proyecta sobre los EOF complejos (no es sorprendente).

* En la tropósfera, se proyecta sobre el PC2 
- principalmente en la parte real   
(OJO: este resultado no es del todo robusto y depende de la definción del dominio usado para el eof complejo. Si se usa todo el hemisferio al sur de los 20°S en vez de cortar en 80°S la repartición es más pareja)

* En la estratósfera se proyecta sobre el PC1 de froma pareja. 
- La combinación lineal es pareja, pero este resultado también es ligeramente sensible la definición de domino; aunque no tanto como la conclusión anterior.

### 2020-06-25 analysis/paper/paper.Rmd
burn it all

---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

```{r}
library(asymsam)
library(ggplot2)
```

```{r}
ggplot() +
  geom_qmap()
```

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

<!-- Here's some example analysis code: -->

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot

```{r demo-inline-code}
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2020-06-25 R/ReIm.R

@@ -1,0 +1,30 @@
+#' Splits complex numbers into real and imaginary vectors
+#'
+#' @param complex vector of complex values
+#' @param data a data.table
+#' @param column unquoted name of the column to split into real and imaginary part
+#' @param longer logical indicating wether to return a "longer" version of the data.frame
+#'
+#' @export
+ReIm <- function(complex) {
+  list(R = Re(complex), I = Im(complex))
+}
+
+#' @export
+#' @rdname ReIm
+sep_ReIm <- function(data, column, longer = TRUE) {
+  R <- part <- I <- NULL
+  names <- c("R", "I")
+  expr <- quote(data.table::copy(data)[, (names) := ReIm(column)])
+  expr  <-  do.call(substitute, list(expr,
+                                     list(col = substitute(column))))
+  data <- eval(expr)
+
+  if (isTRUE(longer)) {
+    data[, deparse(substitute(column)) := NULL]
+    data <- data.table::setDT(tidyr::pivot_longer(data, R:I, names_to = "part", values_to = deparse(substitute(column))))
+    data[, part := factor(part, levels = c("R", "I"), labels = c("Real", "Imaginary"))]
+  }
+
+  return(data[])
+}

### 2020-06-25 R/asymsam-package.R

@@ -1,0 +1,10 @@
+#' @keywords internal
+"_PACKAGE"
+
+# The following block is used by usethis to automatically manage
+# roxygen namespace tags. Modify with care!
+## usethis namespace: start
+#' @importFrom magrittr %>%
+## usethis namespace: end
+NULL
+

### 2020-06-25 R/data_locations.R

@@ -1,0 +1,65 @@
+#' @export
+print.nc_file <- function(x, ...) {
+   print(metR::GlanceNetCDF(x))
+}
+
+#' @param type character indicating the type of data. The resulting base will be type_data
+#' @param ... characters to be contactenadted with [path.file()]
+#' @export
+#' @rdname data_locations
+data_path <- function(type = c("raw", "derived"), ...) {
+   here::here("analysis", "data", paste0(type[1], "_data"), ...)
+}
+
+#' @export
+#' @rdname data_locations
+AAO <- function() {
+   file <- data_path("derived", "aao.Rds")
+   checkmate::assert_access(file, access = "r")
+   return(file)
+}
+
+#' Data locations
+#'
+#' Functions that return the location of different datasets. They check that
+#' relevant file exist and then return it with a class that has a neat print method.
+#'
+#'
+#' @export
+#' @rdname data_locations
+SST <- function() {
+   file <- here::here("DATA", "reanalysis", "ERSST_V5", "sst.ersst.mnmean.nc")
+   checkmate::assert_access(file, access = "r")
+   class(file) <- c("nc_file", class(file))
+   return(file)
+}
+
+
+#' @export
+#' @rdname data_locations
+ERA5 <- function() {
+   file <- data_path("raw", "era5.z.mon.mean.nc")
+   checkmate::assert_access(file, access = "r")
+   class(file) <- c("nc_file", class(file))
+   return(file)
+}
+
+
+
+#' @export
+#' @rdname data_locations
+OLR <- function() {
+   file <- here::here("DATA", "reanalysis", "NOAA", "olr.mon.mean.nc")
+   checkmate::assert_access(file, access = "r")
+   class(file) <- c("nc_file", class(file))
+   return(file)
+}
+
+#' @export
+#' @rdname data_locations
+CMAP <- function() {
+   file <- here::here("DATA", "reanalysis", "CMAP", "precip.mon.mean.nc")
+   checkmate::assert_access(file, access = "r")
+   class(file) <- c("nc_file", class(file))
+   return(file)
+}

### 2020-06-25 R/eof_asym.R

@@ -1,0 +1,58 @@
+#' Computes symmetric and asymmetric Principal Components
+#'
+#' Computes EOF of a field (weighted by sqrt(cos(lat))), separates it between
+#' its zonally symmetric and zonally asymmetric parts and finally, computes
+#' time series of those patterns.
+#'
+#' @param value value of the variable
+#' @param lon,lat,time latitude, longitude and time
+#' @param n numeric vector of components to compute.
+#'
+#' @export
+eof_asym <- function(value, lon, lat, time, n = 1) {
+  sym <- asym <- full <- PC <- estimate_norm <- term <- estimate <- r.squared <-  partial.r.squared <- adj.r.squared <- NULL
+  data <- data.table::data.table(value, lon, lat, time, key = c("lon", "lat", "time"))
+
+  lab_sam <-  c(full = "Full",
+                asym = "Asymmetric",
+                sym  = "Symmetric")
+
+  eof <- data %>%
+    .[, full := value*sqrt(cos(lat*pi/180))] %>%
+    metR::EOF(full ~ time | lon + lat, n = n, data = .)
+  eof$right[, c("sym", "asym") := list(mean(full), metR::Anomaly(full)), by = .(lat, PC)]
+
+  data[, full := NULL]
+  indexes <- eof$right %>%
+    data[., on = .NATURAL, allow.cartesian = TRUE]
+
+  pcor <- indexes[, partial_cor(value, sym, asym, weights = cos(lat*pi/180)),
+                  keyby = .(time, PC)]
+
+  indexes[, rbind(data.table::as.data.table(lm_lite(value, full, weights = cos(lat*pi/180), r2 = TRUE)),
+                  data.table::as.data.table(metR::FitLm(value, sym, asym, weights = cos(lat*pi/180), r2 = TRUE))),
+          keyby = .(time, PC)] %>%
+    .[term != "(Intercept)"] %>%
+    .[, estimate_norm := estimate/stats::sd(estimate[term == "full"]), by = PC] %>%
+    .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>%
+    .[term != "full", partial.r.squared := pcor$partial_correlation^2] %>%
+    .[term == "full", partial.r.squared := r.squared] %>%
+    .[, adj.r.squared := NULL] %>%
+    .[]
+}
+
+#' Flip signs of an EOF
+#'
+#'
[…]

### 2020-06-25 R/eof_methods.R

@@ -1,0 +1,129 @@
+## Methos for EOF
+#
+# cut.eof <- function(eof, n) {
+#    var <- attr(eof, "suffix")
+#    value.var <- attr(eof, "value.var")
+#    return(structure(lapply(as.list(eof), function(x) {
+#       x[as.numeric(get(var)) %in% n]
+#    }),
+#    class = c("eof", "list"),
+#    suffix = var,
+#    value.var = value.var))
+# }
+
+# screeplot.eof <- function(eof, n = "all") {
+#    var <- attr(eof, "suffix")
+#    r2 <- "r2"
+#    if (n[1] == "all") n <- as.numeric(unique(eof$sdev[[var]]))
+#    ggplot(eof$sdev[as.numeric(get(var)) %in% seq_along(n)], aes_(as.name(var), as.name(r2))) +
+#       geom_point()
+# }
+#
+# autoplot.eof <- function(eof, n = "all") {
+#    screeplot(eof, n)
+# }
+#
+# predict.eof <- function(eof, n = NULL) {
+#    ` %>% ` <- magrittr::`%>%`
+#    if (!inherits(eof, "eof")) {
+#       stop("eof must be an EOF object")
+#    }
+#
+#    if(!is.null(n)) eof <- cut(eof, n)
+#
+#    value.var <- attr(eof, "value.var")
+#    pc <- attr(eof, "suffix")
+#
+#    right.vars <- colnames(eof$right)[!(colnames(eof$right) %in% c(pc, value.var))]
+#    right.formula <- as.formula(paste0(pc, " ~ ", paste0(right.vars, collapse = "+")))
+#
+#    right <- eof$right %>%
+#       .[eof$sdev, on = pc] %>%
+#       .[, (value.var) := get(value.var)*sd] %>%
+#       metR:::.tidy2matrix(right.formula, value.var)
+#
+#    left.vars <- colnames(eof$left)[!(colnames(eof$left) %in% c(pc, value.var))]
+#    left.formula <- as.formula(paste0(pc, " ~ ", paste0(left.vars, collapse = "+")))
+#    left <- metR:::.tidy2matrix(eof$left, left.formula, value.var)
+#
+#    dt <- cbind(.extend.dt(left$coldims, each = nrow(right$coldims)),
+#                .extend.dt(right$coldims, n = nrow(left$coldims)),
+#                c(t(right$matrix)%*%left$matrix))
+#    colnames(dt)[length(colnames(dt))] <- value.var
[…]

### 2020-06-25 R/helperfun.R

@@ -1,0 +1,1333 @@
+
+#
+#
+# ## Archivo con funciones
+#
+# #Librerías
+# library(ggplot2)
+# library(ggforce)
+# library(stringi)
+# # library(ggthemes)
+# library(magrittr)
+# library(data.table)
+# library(lubridate)
+# library(akima)
+# library(compiler)
+# # library(RColorBrewer)
+# enableJIT(0)
+#
+# # source(here::here("scripts/eof_methods.R"))
+#
+#
+# # Repetir vector en negativo
+# fill_neg <- function(vector) {
+#    c(-vector[order(-vector)], vector)
+# }
+#
+#
+#
+# # Nombres de los meses en español
+# month.abb_sp <- c("Ene", "Feb", "Mar", "Abr", "May", "Jun",
+#                   "Jul", "Ago", "Sep", "Oct", "Nov", "Dic")
+# names(month.abb_sp) <- as.character(1:12)
+#
+# names(month.abb) <- as.character(1:12)
+#
+#
+# # # Para interpolación en data table
+# # Interpolate.DT <- function(z, x, y, yo = unique(y), xo = unique(x), ...){
+# #    na <- is.na(z)
+# #    int <- akima::interp(x = x[!na], y = y[!na], z = z[!na], yo = yo, xo = xo, ...)
+# #    names <- c(deparse(substitute(x)),
+# #               deparse(substitute(y)),
+# #               deparse(substitute(z)))    # muy feo, sí
+# #    r <- with(int, {
+# #       grid <- expand.grid(x, y)
+# #       r <- list(grid[,1], grid[, 2], c(z))
+# #       names(r) <- names
+# #       return(r)
+# #    })
+# # }
+#
+# # Función que hace autocorrelograma y su test según Anderson o large lag.
+# acf.sig <- function(x, lag.max=0.3*length(x), alpha = 0.05,
+#                     method=c("anderson","large.lag", "salas"), sided="one") {
+#    autocor <- acf(x, lag.max=lag.max, plot = F)$acf
+#    N <- length(x)
+#    e <- -1/(N-1)
+#    if (method[1]=="anderson"){
+#       var <- (N-2)/(N-1)^2
+#    } else if (method[1]=="large.lag"){
+#       var <- vector()
+#       for (i in 1:length(autocor)){
+#          v <- ifelse(i==1,1/N, 1/N * (1+2*sum(autocor[1:i-1]^2)))
+#          var<- c(var, v)
+#       }
[…]

### 2020-06-25 R/hemisphere.R

@@ -1,0 +1,11 @@
+#' Convert latitude into hemispheres
+#'
+#' @param lat latitude
+#'
+#' @export
+hemisphere <- function(lat) {
+  factor(ifelse(lat < 0, "sh", "nh"), levels = c("nh", "sh"),
+         labels = c("Northern Hemisphere",
+                    "Southern Hemisphere"),
+         ordered = TRUE)
+}

### 2020-06-25 R/lm_lite.R

@@ -1,0 +1,77 @@
+#' Fast linear regression
+#'
+#' Computes regression coefficient quickly for the case
+#' of simple lienar regression using the simple
+#' covariance/variance formula.
+#'
+#' @param y dependent variable
+#' @param x independent variable
+#' @param weights vector of weights
+#' @param r2 logical indicating wether to compute r.quared
+#'
+#' @export
+lm_lite <- function(y, x, weights = rep(1, length(y)), r2 = FALSE) {
+  cov <- stats::cov.wt(cbind(y, x), wt = weights, cor = r2)
+  estimate <- cov$cov[1, 2]/var.wt(x, w = weights)
+  out <- list(term = deparse(substitute(x)),
+              estimate = estimate)
+
+  if (r2) {
+    out$r.squared <-  cov$cor[1, 2]^2
+    out$adj.r.squared <- cov$cor[1, 2]^2   # Not real!!
+  }
+  return(out)
+}
+
+
+#' Weighted vaiance
+#' @keywords internal
+var.wt <- function(x, w, na.rm = FALSE) {
+  if (na.rm) {
+    w <- w[i <- !is.na(x)]
+    x <- x[i]
+  }
+  sum.w <- sum(w)
+  (sum(w*x^2) * sum.w - sum(w*x)^2) / (sum.w^2 - sum(w^2))
+}
+
+
+#' Compute p.values
+#'
+#' Computes p.values from the result of [metR::FitLm()] (or any assumed student-t-distributed
+#' statistic) with optional adjustments.
+#'
+#' @param estimate estimate of the regression (or other fit)
+#' @param std.error standard error
+#' @param df degrees of freedom
+#' @param adjustment method for adjustment, see [stats::p.adjust()].
+#'
+#' @export
+Pvaluate <- function(estimate, std.error, df, adjustment = "none") {
+  stats::p.adjust(2*stats::pt(abs(estimate)/std.error, df, lower.tail = FALSE), method = adjustment)
+}
+
+
+#' Remove intercept
+#'
+#' Removes intercept term from the result of [metR::FitLm()]. More to the point,
+#' it removes rows from a data.frame in which a column named "term" equals "(Intercept)".
+#'
+#' @param data dataframe
+#'
+#' @export
+rm_intercept <- function(data) {
+  data[data$term != "(Intercept)"]
+}
+
+#' Remove singleton dimensions
+#'
+#' Removes any column from a dataframe that holds only
[…]

### 2020-06-25 R/maps.R

@@ -1,0 +1,50 @@
+#' Quick maps
+#'
+#'
+#' @param subset a function that will be applied to the map data.
+#' @param color,size,fill parameters passed to [ggplot2::geom_polygon()]
+#' @param wrap vector of length 2 interpreted as the longitude range of a global map (see [maps::map()])
+#' @param keep proportions of point to retain (passed to [rmapshaper::ms_simplify()])
+#' @param ... other arguments passed to [ggplot2::geom_polygon()]
+#'
+#' @export
+geom_qmap <- function(subset = identity, color = "black", size = 0.2,
+                      fill = NA, wrap = c(0, 360),
+                      keep = 0.015, ...) {
+  lon <- lat <- group <- NULL
+  data <- ggplot2::fortify(map_simple(wrap = wrap, keep  = keep)) %>%
+    data.table::as.data.table() %>%
+    .[, c("long", "lat", "group")] %>%
+    data.table::setnames("long", "lon")
+  subset <- purrr::as_mapper(subset)
+  data <- subset(data)
+
+  ggplot2::geom_polygon(data = data,
+                        ggplot2::aes(lon, lat, group = group),
+                        color = color,
+                        size = size,
+                        fill = fill,
+                        ...)
+}
+
+map_simple_ <- function(wrap = c(0, 360), keep = 0.015) {
+  map <- maps::map("world", fill = TRUE,
+                   col = "transparent", plot = FALSE, wrap = wrap)
+  IDs <- vapply(strsplit(map$names, ":"), function(x) x[1],
+                "")
+  proj <- sp::CRS("+proj=longlat +datum=WGS84")
+  map <- maptools::map2SpatialPolygons(map, IDs = IDs,
+                                       proj4string = proj)
+
+  if (keep != 1) {
+    map <- rmapshaper::ms_simplify(map, keep = keep)
+  }
+  map
+}
+
+
+
+
+#' @export
+#' @rdname geom_qmap
+map_simple <- memoise::memoise(map_simple_)

### 2020-06-25 R/normalise_coords.R

@@ -1,0 +1,31 @@
+#' Normalises spatio-temporal dimensions
+#'
+#' @param data a data.frame
+#' @param rules a list of "rules". Each element is a possible column name,
+#' which will be renamed using the name of the element of the list.
+#' @param extra an extra list of rules to add to the default one.
+#'
+#' @export
+normalise_coords <- function(data,
+                             rules =  list(lev = c("level"),
+                                           lat = c("latitude"),
+                                           lon = c("longitude", "long"),
+                                           time = c("date")),
+                             extra = list()) {
+
+  checkmate::assert_list(rules, types = "character", names = "named", any.missing = FALSE)
+  checkmate::assert_list(extra, types = "character", names = "named", any.missing = FALSE)
+
+  rules <- c(rules, extra)
+
+  for (f in seq_along(rules)) {
+    old <- colnames(data)[colnames(data) %in% rules[[f]]]
+
+    if (length(old) != 0) {
+      data.table::setnames(data,
+                           old,
+                           names(rules)[[f]], skip_absent = TRUE)
+    }
+  }
+  return(invisible(data))
+}

### 2020-06-25 R/partial_cor.R

@@ -1,0 +1,20 @@
+#' Partial correlation between x1 and y and x2 and y.
+#'
+#' @param y variable to which partial correlation will be computed
+#' @param x1,x2 variables that will be correlated with y
+#' @param weights vector of weights
+#'
+#' @export
+partial_cor <- function(y, x1, x2, weights = rep(1, length(y))) {
+  rho_x1x2 <- stats::cov.wt(cbind(x1, x2), wt = weights, cor = TRUE)$cor[1, 2]
+  rho_x1y <- stats::cov.wt(cbind(x1, y), wt = weights, cor = TRUE)$cor[1, 2]
+  rho_x2y <- stats::cov.wt(cbind(x2, y), wt = weights, cor = TRUE)$cor[1, 2]
+
+  estimate1 <- (rho_x1y - rho_x2y*rho_x1x2 ) / (sqrt(1 - rho_x2y^2) * sqrt(1 - rho_x1x2^2))
+  estimate2 <- (rho_x2y - rho_x1y*rho_x1x2 ) / (sqrt(1 - rho_x1y^2) * sqrt(1 - rho_x1x2^2))
+
+  list(term = c(deparse(substitute(x1)),
+                deparse(substitute(x2))),
+       partial_correlation = c(estimate1, estimate2))
+}
+

### 2020-06-25 R/theme.R

@@ -1,0 +1,25 @@
+#' Main theme used in the paper
+#'
+#' Based on [ggplot2::theme_minimal()] but with minimal (wink) modifications.
+#'
+#'
+#' @param base_size base font size, given in pts
+#'
+#' @export
+theme_asymsam <- function(base_size = 16) {
+  ggplot2::theme_minimal(base_size = 16) +
+    ggplot2::theme(
+      strip.background = ggplot2::element_rect(fill = NA, color = "gray30"),
+      # text = element_text(family = font_rc),
+      legend.position = "bottom", legend.box = "vertical",
+      panel.spacing.y = grid::unit(5, "mm"),
+      panel.spacing.x = grid::unit(5, "mm"),
+      legend.spacing = grid::unit(2, "mm"),
+      panel.border = ggplot2::element_rect(colour = "black", fill = NA),
+      plot.margin = grid::unit(rep(3, 4), "mm"),
+      # legend.title = element_blank(),
+      legend.box.spacing = grid::unit(3, "mm"),
+      legend.margin = ggplot2::margin(t = -5),
+      panel.grid = ggplot2::element_line(color = "gray10", size = 0.4, linetype = 3),
+      panel.ontop = TRUE)
+}

### 2020-06-25 analysis/scripts/01 - get-data.R

@@ -1,0 +1,88 @@
+## Downloads data from the oficial channels and
+##  - AAO from NOAA's Climate Prediction Centre (see ?rsoi::download_aao)
+##  - GISTEMP from NOAA's Physical Sciences Laboratory (https://psl.noaa.gov/data/gridded/data.gisstemp.html)
+##  - CMAP from NOAA's Physical Sciences Laboratory (https://psl.noaa.gov/data/gridded/data.cmap.html)
+##  - ERA5 Geopotential height from Climate Data Store (https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means?tab=overview)
+##
+
+library(magrittr)
+
+force <- FALSE
+
+files <- list(aao = "aao.csv",
+              gistemp = "air.2x2.1200.mon.anom.land.nc",
+              cmap = "precip.mon.mean.nc",
+              era5 =  "era5.z.mon.mean.nc")
+
+# Download AAO ------------------------------------------------------------
+file <- files$aao
+aao <- rsoi::download_aao(use_cache = !force, file = asymsam::data_path("raw", file)) %>%
+  data.table::as.data.table() %>%
+  .[, .(time = lubridate::as_datetime(Date), aao = AAO)]
+
+saveRDS(aao, asymsam::data_path("derived", "aao.Rds"))
+
+# Download GISTEMP --------------------------------------------------------
+url <- "ftp://ftp.cdc.noaa.gov/Datasets/gistemp/landonly/1200km/air.2x2.1200.mon.anom.land.nc"
+file <- files$gistemp
+
+if (!force && file.exists(asymsam::data_path("raw", file))) {
+  message("File already exists.")
+} else {
+  download.file(url, destfile = asymsam::data_path("raw", file))
+}
+
+# Download CMAP -----------------------------------------------------------
+url <- "ftp://ftp.cdc.noaa.gov/Datasets/cmap/std/precip.mon.mean.nc"
+file <- files$cmap
+
+if (!force && file.exists(asymsam::data_path("raw", file))) {
+  message("File already exists.")
+} else {
+  download.file(url, destfile = asymsam::data_path("raw", file))
+}
+
+# Download ERA5 -----------------------------------------------------------
+user <- Sys.getenv("CDSUSER", NA)
+key <- Sys.getenv("CDSKEY", NA)
+file <- files$era5
+
+if (!force && file.exists(asymsam::data_path("raw", file))) {
[…]

### 2020-06-25 analysis/paper/paper.Rmd
Adds some figures

@@
 ---
-title: "Title Goes Here"
-author:
-  - Myfirstname Mylastname:
-      email: fl@oneplace.edu
-      institute: [UofO]
-      correspondence: true
-  - Yourfirstname Yourlastname:
-      email: fl@another.edu
-      institute: [UofA]
-      correspondence: false
-institute:
-  - UofO: University of One Place
-  - UofA: University of Another Place
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+title: "SAM simétrico y asimétrico"
+author: "Elio Campitelli"
 output: 
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
-abstract: |
-  Text of abstract
-keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
+   # pdf_document:
+   #    latex_engine: xelatex
+   powerpoint_presentation:
+    # ioslides_presentation:
+        fig_height: 7.6
+        fig_width: 12.8
+        reference_doc: "../templates/template.pptx" 
+        slide_level: 2
+urlcolor: blue
+header_includes:
+   - \usepackage[utf8]{inputenc}
 ---
@@
 
-```{r setup, echo = FALSE}
+```{r setup, include=FALSE}
 knitr::opts_chunk$set(
@@
 )
+
+library(metR)
+library(data.table)
+library(ggplot2)
+library(magrittr)
+library(ggh4x)
+library(ggperiodic)
+library(asymsam)
+
+# source(here("scripts", "helperfun.R"))
+# source(here("scripts", "data_locations.R"))
+theme_set(theme_asymsam())
+ZeroBreaks <- AnchorBreaks(0, NULL, 0)
+
+lev.lab <- AddSuffix(" hPa")
+
+guide_colorstrip_bottom <- function(width = 25, height = 0.5, ...) {
+  guide_colorstrip(title.position = "top", title.hjust = 0.5,
+                   barheight = height,
+                   barwidth = width, ...)
+}
+
+main_levs <- c(700, 300, 30)
+# knitr_set_timer()
+# knitr_set_cache(cache.extra = 44)
+```
+
+```{r read-hgt, cache = FALSE}
+hgt <- ReadNetCDF(ERA5(), 
+                  subset = list(time = c("1979-01-01", "2018-12-31"),
+                                latitude = c(-90, 10)),
+                  vars = c(hgt = "z")) %>% 
+  normalise_coords() %>% 
+  .[, hgt := hgt/9.8] %>% 
+  .[, dataset := "ERA5"] %>%  
+  .[, hgt_a := hgt - mean(hgt), by = .(lon, lat, lev, month(time), dataset)] %>% 
+  .[, hgt_m := mean(hgt_a), by = .(lat, lev, time)] %>% 
+  .[, hgt_z := hgt_a - hgt_m]
+```
+
+```{r compute-SAM}
+SAM <- hgt[lat <= -20] %>% 
+  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
+  .[, .(eof = list(eof_flip(EOF(hgt ~ time | lon + lat, n = 1, data = .SD)))), 
+    by = .(dataset, lev)]
 ```
@@
 
-```{r}
-library(asymsam)
-library(ggplot2)
+# Methods
+
+```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height.
+Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right).
+Arbitrary units."}
+sam_sep <- SAM[, eof[[1]]$right, by = .(dataset, lev)] %>% 
+  setnames("hgt", "full") %>% 
+  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev, dataset)]
+
+lab_sam <-  c(full = "Full", 
+              asym = "Asymmetric",
+              sym  = "Symmetric")
+
+sam_sep %>% 
+  .[dataset == "ERA5"] %>% 
+  .[lev == 700] %>%
+  rm_singleton() %>% 
+  melt(id.vars = c("lon", "lat")) %>%
+
+  # .[variable == "asym", value := value + rnorm(.N, 0, sd = 1e-6)] %>%
+  .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)] %>% 
+  periodic(lon = c(0, 360)) %>% 
+  ggplot(aes(lon, lat)) +
+  # geom_contour2(aes(z = value, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
+  geom_contour_fill(aes(z = value), global.breaks = FALSE) +
+  geom_contour_tanaka(aes(z = value), global.breaks = FALSE) +
+  geom_qmap() +
+  scale_x_longitude() +
+  scale_y_latitude(limits = c(NA, -20)) +
+  scale_fill_divergent(guide = "none") +
+  coord_polar() +
+  facet_grid(~variable, labeller = labeller(variable = lab_sam, lev = lev.lab)) 
+```
+
+# Results
+
+```{r compute-asymsam}
+indexes <-  hgt[lat <= -20, eof_asym(hgt_a, lon, lat, time, n = 1), by = lev]
+indexes_wide <- dcast(indexes, + lev + time ~ term, value.var = "estimate_norm")
+```
+
+```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM."}
+indexes %>% 
+  .[lev == 700 & term != "full"] %>% 
+  ggplot(aes(time, estimate_norm)) +
+  geom_line(aes(color = term)) +
+  # geom_smooth(method = "lm", aes(color = term)) +
+  scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
+  scale_y_continuous(NULL) +
+  scale_x_datetime(NULL)
 ```
@@
 ```{r}
-ggplot() +
-  geom_qmap()
+indexes_wide[, cor(asym, sym), by = lev] %>% 
+  ggplot(aes(lev, V1)) +
+  geom_line() +
+  scale_x_level() + 
+  scale_y_continuous(limits = c(0, 1)) +
+  coord_flip()
 ```
 
-Here is a citation [@Marwick2017]
+```{r}
 
-# Background
+indexes_2d <- hgt[lev %in% main_levs] %>% 
+  indexes_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
+  .[, rbind(as.data.table(FitLm(hgt_a, sym, asym)),
+            as.data.table(FitLm(hgt_a, full))),
+    by = .(lon, lat, lev)] %>% 
+  rm_intercept() %>% 
+  periodic(lon = c(0, 360)) %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)]
 
-# Methods
-
-# Results
-
-<!-- Here's some example analysis code: -->
-
-```{r get-data, eval = FALSE}
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
-```
-
-```{r demo-plot, fig.cap="A plot of random numbers"}
-plot(rnorm(10))
-```
-
-Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot
-
-```{r demo-inline-code}
-x <- round(pi, 2)
+indexes_2d %>% 
+  .[lat <= -20] %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
+  # geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
+                # size = 0.3,
+                # global.breaks = FALSE) +
+  geom_contour_tanaka(aes(z = estimate),
+                      breaks = ZeroBreaks,
+                      size = 0.3,
+                      global.breaks = FALSE) +
+  geom_qmap(~.x[lat <= -20]) +
+  # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
+  scale_x_longitude() +
+  scale_y_latitude(limits = c(NA, -20)) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
+                       limits = c(-80, 80),
+                       oob = scales::squish,
+                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
+  scale_linetype(guide = "none") +
+  coord_polar() +
+  facet_grid(term ~ lev, labeller = labeller(term = lab_sam, lev = lev.lab))
 ```

### 2020-06-25 R/theme.R

@@ -9,17 +9,17 @@
 theme_asymsam <- function(base_size = 16) {
   ggplot2::theme_minimal(base_size = 16) +
     ggplot2::theme(
-      strip.background = ggplot2::element_rect(fill = NA, color = "gray30"),
+      # strip.background = ggplot2::element_rect(fill = NA, color = "gray30"),
       # text = element_text(family = font_rc),
       legend.position = "bottom", legend.box = "vertical",
-      panel.spacing.y = grid::unit(5, "mm"),
-      panel.spacing.x = grid::unit(5, "mm"),
-      legend.spacing = grid::unit(2, "mm"),
-      panel.border = ggplot2::element_rect(colour = "black", fill = NA),
-      plot.margin = grid::unit(rep(3, 4), "mm"),
-      # legend.title = element_blank(),
-      legend.box.spacing = grid::unit(3, "mm"),
-      legend.margin = ggplot2::margin(t = -5),
+      # panel.spacing.y = grid::unit(5, "mm"),
+      # panel.spacing.x = grid::unit(5, "mm"),
+      # legend.spacing = grid::unit(2, "mm"),
+      # panel.border = ggplot2::element_rect(colour = "black", fill = NA),
+      # plot.margin = grid::unit(rep(3, 4), "mm"),
+      # # legend.title = element_blank(),
+      # legend.box.spacing = grid::unit(3, "mm"),
+      # legend.margin = ggplot2::margin(t = -5),
       panel.grid = ggplot2::element_line(color = "gray10", size = 0.4, linetype = 3),
       panel.ontop = TRUE)
 }

### 2020-06-29 analysis/paper/paper.Rmd
draft figs

@@
 <!-- With the following code you can access and display values from the yml header above. -->
-
+<!-- 
 Keywords: `r rmarkdown::metadata$keywords`
@@
 Highlights: `r rmarkdown::metadata$highlights`
+-->
 
@@
   collapse = TRUE,
+  cache = TRUE,
   warning = FALSE,
@@
 indexes %>% 
-  .[lev == 700 & term != "full"] %>% 
+  .[lev %in% c(30, 700) & term != "full"] %>% 
   ggplot(aes(time, estimate_norm)) +
@@
   scale_y_continuous(NULL) +
-  scale_x_datetime(NULL)
+  scale_x_datetime(NULL) +
+  facet_wrap(lev~., labeller = labeller(lev = lev.lab), ncol = 1)
 ```
 
-```{r}
+```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level."}
 indexes_wide[, cor(asym, sym), by = lev] %>% 
@@
   scale_x_level() + 
-  scale_y_continuous(limits = c(0, 1)) +
+  scale_y_continuous("Correlation", limits = c(0, 1)) + 
   coord_flip()
@@
 
-```{r}
+```{r cross-correlation, fig.cap = "Cross correlation between levels of the Full, Asymmetric and Symmetric SAM."}
+indexes %>% 
+  copy() %>% 
+  .[, widyr::pairwise_cor(.SD, lev, time, estimate), by = .(term)] %>% 
+  ggplot(aes(item1, item2)) +
+  geom_contour_fill(aes(z = correlation), na.fill = 1) +
+  scale_fill_divergent("Cross-correlation") +
+  scale_x_level() +
+  scale_y_level() +
+  coord_equal() +
+  facet_grid(. ~ term, labeller = labeller(term = lab_sam))
+```
 
+```{r compute-regr-pattern}
 indexes_2d <- hgt[lev %in% main_levs] %>% 
@@
   .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] 
+```
 
+```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM.
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef."}
 indexes_2d %>% 
@@
 
-Here is an example of inline code `r x` in the middle of a sentence.
+```{r compute-regr-pattern-season}
+indexes_2d <- hgt[lev %in% main_levs] %>% 
+  indexes_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
+  .[, rbind(as.data.table(FitLm(hgt_a, sym, asym)),
+            as.data.table(FitLm(hgt_a, full))),
+    by = .(lon, lat, lev, season(time))] %>% 
+  rm_intercept() %>% 
+  periodic(lon = c(0, 360)) %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)]
+
+```
+
+```{r 2d-regr-700, fig.cap = "Seasonal regression patterns of geopotential height at 700 hPa with the Full, Asymmetric and Symmetric SAM.
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef."}
+indexes_2d %>% 
+  .[lat <= -20] %>% 
+  .[lev == 700] %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
+  # geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
+  # size = 0.3,
+  # global.breaks = FALSE) +
+  geom_contour_tanaka(aes(z = estimate),
+                      breaks = ZeroBreaks,
+                      size = 0.3,
+                      global.breaks = FALSE) +
+  geom_qmap(~.x[lat <= -20]) +
+  # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
+  scale_x_longitude() +
+  scale_y_latitude(limits = c(NA, -20)) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
+                       limits = c(-80, 80),
+                       oob = scales::squish,
+                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
+  scale_linetype(guide = "none") +
+  coord_polar() +
+  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab))
+```
+
+```{r 2d-regr-30, fig.cap = "Seasonal regression patterns of geopotential height at 30 hPa with the Full, Asymmetric and Symmetric SAM.
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef."}
+indexes_2d %>% 
+  .[lat <= -20] %>% 
+  .[lev == 30] %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
+  # geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
+  # size = 0.3,
+  # global.breaks = FALSE) +
+  geom_contour_tanaka(aes(z = estimate),
+                      breaks = ZeroBreaks,
+                      size = 0.3,
+                      global.breaks = FALSE) +
+  geom_qmap(~.x[lat <= -20]) +
+  # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
+  scale_x_longitude() +
+  scale_y_latitude(limits = c(NA, -20)) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
+                       limits = c(-80, 80),
+                       oob = scales::squish,
+                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
+  scale_linetype(guide = "none") +
+  coord_polar() +
+  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab))
+```
+
+```{r wave-amplitude-700, fig.cap = "Planteray wave squared amplitudes for the regression patterns at 700 hPa."}
+indexes_2d %>% 
+  .[ lev == 700 & lat < -20] %>%
+  .[, FitWave(estimate, 1:3), by = .(lat, lev, season, term)] %>% 
+  ggplot(aes(lat, amplitude^2)) +
+  geom_line(aes(color = factor(k))) +
+  scale_color_brewer("Wave-number", palette = "Set1") +
+  scale_y_continuous("Squared amplitude") +
+  scale_x_latitude(ticks = 15) +
+  coord_flip() +
+  facet_grid(term~season, labeller = labeller(term = lab_sam)) 
+```
+
+```{r wave-amplitude-30, fig.cap = "Planteray wave squared amplitudes for the regression patterns at 30 hPa."}
+indexes_2d %>% 
+  .[ lev == 30 & lat < -20] %>%
+  .[, FitWave(estimate, 1:3), by = .(lat, lev, season, term)] %>% 
+  ggplot(aes(lat, amplitude^2)) +
+  geom_line(aes(color = factor(k))) +
+  scale_color_brewer("Wave-number", palette = "Set1") +
+  scale_y_continuous("Squared amplitude") +
+  scale_x_latitude(ticks = 15) +
+  coord_flip() +
+  facet_grid(term~season, labeller = labeller(term = lab_sam)) 
+```
+
+```{r compute-pp-regr} 
+pp_regr <- CMAP() %>% 
+  ReadNetCDF() %>% 
+  normalise_coords() %>% 
+  .[, precip := Anomaly(precip), by = .(lon, lat)] %>% 
+  .[indexes_wide[lev == 700], on = .NATURAL] %>% 
+  .[, FitLm(precip, asym, sym, se = TRUE, r2 = TRUE), by = .(lon, lat)] %>% 
+  rm_intercept() %>% 
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)]
+```
+
+```{r regr-pp, fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
+pp_regr %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate), na.fill = TRUE, breaks = AnchorBreaks(0, 0.1, 0)) +
+  geom_contour2(aes(z = p.value), breaks = 0.01) +
+  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
+  stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
+  geom_qmap() +
+  scale_x_longitude() +
+  scale_y_latitude() +
+  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
+  coord_quickmap() +
+  facet_wrap(~term, labeller = labeller(term = lab_sam))
+```
+
+```{r}
+air_regr <- GISTEMP() %>% 
+  ReadNetCDF(subset = list(time = range(indexes$time))) %>% 
+  normalise_coords() %>% 
+  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
+  .[indexes_wide[lev == 700], on = .NATURAL] %>% 
+  .[, FitLm(air, asym, sym), by = .(lon, lat)] %>% 
+  rm_intercept()
+```
+
+```{r,  fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
+air_regr %>% 
+  ggplot(aes(lon, lat)) +
+  # geom_raster(aes(fill = estimate)) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.1, 0)) +
+  # geom_contour2(aes(z = p.value), breaks = 0.01) +
+  stat_subset(aes(subset = is.na(estimate)), geom = "raster", fill = "gray50") +
+  # stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
+  geom_qmap() +
+  scale_x_longitude() +
+  scale_y_latitude() +
+  scale_fill_divergent("Surface temperature", limits = c(-0.5, 0.5), 
+                       oob = scales::squish,
+                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
+  coord_quickmap() +
+  facet_wrap(~term, labeller = labeller(term = lab_sam))
+```
 

### 2020-06-29 R/data_locations.R

@@ -4,7 +4,7 @@ print.nc_file <- function(x, ...) {
 }
 
 #' @param type character indicating the type of data. The resulting base will be type_data
-#' @param ... characters to be contactenadted with [path.file()]
+#' @param ... characters to be contactenadted with [file.path()]
 #' @export
 #' @rdname data_locations
 data_path <- function(type = c("raw", "derived"), ...) {
@@ -58,7 +58,17 @@ OLR <- function() {
 #' @export
 #' @rdname data_locations
 CMAP <- function() {
-   file <- here::here("DATA", "reanalysis", "CMAP", "precip.mon.mean.nc")
+   file <- data_path("raw", "precip.mon.mean.nc")
+   checkmate::assert_access(file, access = "r")
+   class(file) <- c("nc_file", class(file))
+   return(file)
+}
+
+
+#' @export
+#' @rdname data_locations
+GISTEMP <- function() {
+   file <- data_path("raw", "air.2x2.1200.mon.anom.land.nc")
    checkmate::assert_access(file, access = "r")
    class(file) <- c("nc_file", class(file))
    return(file)

### 2020-06-29 analysis/scripts/01 - get-data.R
Creates data dirs if needed

@@ -14,6 +14,9 @@ files <- list(aao = "aao.csv",
               cmap = "precip.mon.mean.nc",
               era5 =  "era5.z.mon.mean.nc")
 
+dir.create(asymsam::data_path("raw"), recursive = TRUE, showWarnings = FALSE)
+dir.create(asymsam::data_path("derived"), recursive = TRUE, showWarnings = FALSE)
+
 # Download AAO ------------------------------------------------------------
 file <- files$aao
 aao <- rsoi::download_aao(use_cache = !force, file = asymsam::data_path("raw", file)) %>%

### 2020-06-29 analysis/paper/paper.Rmd
Moaaar figures

@@
   scale_x_datetime(NULL) +
-  facet_wrap(lev~., labeller = labeller(lev = lev.lab), ncol = 1)
+  facet_grid(lev~., labeller = labeller(lev = lev.lab))
+```
+
+```{r asymsam-timeseries-seasonally, fig.cap = "Seasonal time series for the asymmetric SAM and symmetric SAM."}
+indexes %>% 
+  .[lev %in% c(30, 700) & term != "full"] %>% 
+  .[, .(estimate_norm = mean(estimate_norm)), by = .(term, lev, time = seasonally(time))] %>% 
+  .[, season := season(time)] %>% 
+  ggplot(aes(time, estimate_norm)) +
+  geom_line(aes(color = term)) +
+  # geom_smooth(method = "lm", aes(color = term)) +
+  scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
+  scale_y_continuous(NULL) +
+  scale_x_datetime(NULL) +
+  facet_grid(season~lev, labeller = labeller(lev = lev.lab))
 ```
@@
 ```{r}
-air_regr <- GISTEMP() %>% 
-  ReadNetCDF(subset = list(time = range(indexes$time))) %>% 
-  normalise_coords() %>% 
+time_subset <- list(time = range(indexes$time))
+air <- NOAATemp() %>% 
+  ReadNetCDF(vars = "air", subset = time_subset) %>% 
+  normalise_coords()
+
+air_regr <- air %>%
+  copy() %>% 
+  .[, nas := mean(is.na(air)), by = .(lon, lat)] %>% 
+  .[nas >= 0.15, air := NA] %>% 
   # .[, air := Anomaly(air), by = .(lon, lat)] %>%
   .[indexes_wide[lev == 700], on = .NATURAL] %>% 
-  .[, FitLm(air, asym, sym), by = .(lon, lat)] %>% 
+  .[, rbind(as.data.table(FitLm(air, asym, sym, se = TRUE)),
+            as.data.table(FitLm(air, full, se = TRUE))),
+    by = .(lon, lat)] %>% 
   rm_intercept()
@@
 air_regr %>%
+  copy() %>% 
+  .[lat <= -60, estimate := NA] %>%
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)] %>% 
   ggplot(aes(lon, lat)) +
-  # geom_raster(aes(fill = estimate)) +
-  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.1, 0)) +
-  # geom_contour2(aes(z = p.value), breaks = 0.01) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0,
+                    breaks = AnchorBreaks(0, 0.05, 0)) +
+  geom_contour2(aes(z = p.value), breaks = 0.05) +
+  stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat)), size = 0.4, alpha = 0.8) +
   stat_subset(aes(subset = is.na(estimate)), geom = "raster", fill = "gray50") +
-  # stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
   geom_qmap() +
@@
   scale_y_latitude() +
-  scale_fill_divergent("Surface temperature", limits = c(-0.5, 0.5), 
-                       oob = scales::squish,
-                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
+  scale_fill_divergent("Surface temperature", 
+                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.05, 0)) +
   coord_quickmap() +
-  facet_wrap(~term, labeller = labeller(term = lab_sam))
+  facet_grid(.~term, labeller = labeller(term = lab_sam))
+```
+
+```{r, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
+air_regr_season <- air %>% 
+  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
+  .[indexes_wide[lev == 700], on = .NATURAL] %>% 
+  rm_singleton() %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
+  .[, nas := mean(is.na(air)), by = .(lon, lat, season(time))] %>% 
+  .[nas >= 0.15, air := NA] %>% 
+  .[, rbind(as.data.table(FitLm(air, asym, sym, se = TRUE)),
+            as.data.table(FitLm(air, full, se = TRUE))),
+    by = .(lon, lat, season(time))] %>% 
+  rm_intercept()
+
+air_regr_season %>% 
+  copy() %>% 
+  .[lat <= -61, estimate := NA] %>%
+  .[lat >= 65, estimate := NA] %>%
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0,
+                    breaks = AnchorBreaks(0, 0.15, 0)) +
+  # geom_raster(aes(fill = estimate)) +
+  geom_contour2(aes(z = p.value), breaks = 0.05) +
+  stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat)), size = 0.4, alpha = 0.8) +
+  stat_subset(aes(subset = is.na(estimate)), geom = "raster", fill = "gray50") +
+  geom_qmap() +
+  scale_x_longitude() +
+  scale_y_latitude() +
+  scale_fill_divergent("Surface temperature",  limits = c(-0.8, .8), oob = scales::squish,
+                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.15, 0)) +
+  coord_quickmap() +
+  facet_grid(season~term, labeller = labeller(term = lab_sam))
 ```

### 2020-06-29 R/data_locations.R

@@ -67,6 +67,25 @@ CMAP <- function() {
 
 #' @export
 #' @rdname data_locations
+NOAATemp <- function() {
+   file <- data_path("raw", "air.mon.anom.nc")
+   checkmate::assert_access(file, access = "r")
+   class(file) <- c("nc_file", class(file))
+   return(file)
+}
+
+#' @export
+#' @rdname data_locations
+HadCRUT <- function() {
+   file <- data_path("raw", "HadCRUT.4.6.0.0.median.nc")
+   checkmate::assert_access(file, access = "r")
+   class(file) <- c("nc_file", class(file))
+   return(file)
+}
+
+
+#' @export
+#' @rdname data_locations
 GISTEMP <- function() {
    file <- data_path("raw", "air.2x2.1200.mon.anom.land.nc")
    checkmate::assert_access(file, access = "r")

### 2020-06-29 R/eof_asym.R

@@ -29,7 +29,7 @@ eof_asym <- function(value, lon, lat, time, n = 1) {
   pcor <- indexes[, partial_cor(value, sym, asym, weights = cos(lat*pi/180)),
                   keyby = .(time, PC)]
 
-  indexes[, rbind(data.table::as.data.table(lm_lite(value, full, weights = cos(lat*pi/180), r2 = TRUE)),
+  indexes[, rbind(data.table::as.data.table(metR::FitLm(value, full, weights = cos(lat*pi/180), r2 = TRUE)),
                   data.table::as.data.table(metR::FitLm(value, sym, asym, weights = cos(lat*pi/180), r2 = TRUE))),
           keyby = .(time, PC)] %>%
     .[term != "(Intercept)"] %>%

### 2020-06-29 analysis/scripts/01 - get-data.R

@@ -11,7 +11,9 @@ force <- FALSE
 
 files <- list(aao = "aao.csv",
               gistemp = "air.2x2.1200.mon.anom.land.nc",
+              NOAAGlobalTemp = "air.mon.anom.nc",
               cmap = "precip.mon.mean.nc",
+              hadcrut = "HadCRUT.4.6.0.0.median.nc",
               era5 =  "era5.z.mon.mean.nc")
 
 # Download AAO ------------------------------------------------------------
@@ -31,6 +33,27 @@ if (!force && file.exists(asymsam::data_path("raw", file))) {
 } else {
   download.file(url, destfile = asymsam::data_path("raw", file))
 }
+
+# Download NOAAGlobalTemp --------------------------------------------------------
+url <- "ftp://ftp.cdc.noaa.gov/Datasets/noaaglobaltemp/air.mon.anom.nc"
+file <- files$NOAAGlobalTemp
+
+if (!force && file.exists(asymsam::data_path("raw", file))) {
+  message("File already exists.")
+} else {
+  download.file(url, destfile = asymsam::data_path("raw", file))
+}
+
+# Download HadCRUT4 --------------------------------------------------------
+url <- "https://crudata.uea.ac.uk/cru/data/temperature/HadCRUT.4.6.0.0.median.nc"
+file <- files$hadcrut
+
+if (!force && file.exists(asymsam::data_path("raw", file))) {
+  message("File already exists.")
+} else {
+  download.file(url, destfile = asymsam::data_path("raw", file))
+}
+
 
 # Download CMAP -----------------------------------------------------------
 url <- "ftp://ftp.cdc.noaa.gov/Datasets/cmap/std/precip.mon.mean.nc"

### 2020-06-29 analysis/paper/paper.Rmd
Removes unnecessary library() call

@@
 library(magrittr)
-library(ggh4x)
 library(ggperiodic)
@@
 
-# source(here("scripts", "helperfun.R"))
-# source(here("scripts", "data_locations.R"))
 theme_set(theme_asymsam())
@@
 main_levs <- c(700, 300, 30)
-# knitr_set_timer()
-# knitr_set_cache(cache.extra = 44)
 ```

### 2020-06-30 analysis/paper/paper.Rmd
Efficiency improvements

@@
 
-```{r read-hgt, cache = FALSE}
+```{r read-hgt}
 hgt <- ReadNetCDF(ERA5(), 
                   subset = list(time = c("1979-01-01", "2018-12-31"),
-                                latitude = c(-90, 10)),
+                                latitude = c(-90, 10),
+                                level = as.list(main_levs)),
                   vars = c(hgt = "z")) %>% 
@@
   .[, hgt := hgt/9.8] %>% 
-  .[, dataset := "ERA5"] %>%  
-  .[, hgt_a := hgt - mean(hgt), by = .(lon, lat, lev, month(time), dataset)] %>% 
-  .[, hgt_m := mean(hgt_a), by = .(lat, lev, time)] %>% 
-  .[, hgt_z := hgt_a - hgt_m]
+  .[, hgt_a := hgt - mean(hgt), by = .(lon, lat, lev, month(time))]
 ```
@@
   .[, .(eof = list(eof_flip(EOF(hgt ~ time | lon + lat, n = 1, data = .SD)))), 
-    by = .(dataset, lev)]
+    by = .(lev)]
 ```
@@
 Arbitrary units."}
-sam_sep <- SAM[, eof[[1]]$right, by = .(dataset, lev)] %>% 
+sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
   setnames("hgt", "full") %>% 
-  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev, dataset)]
+  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)]
 
@@
 sam_sep %>% 
-  .[dataset == "ERA5"] %>% 
   .[lev == 700] %>%
@@
 ```{r compute-asymsam}
-indexes <-  hgt[lat <= -20, eof_asym(hgt_a, lon, lat, time, n = 1), by = lev]
+indexes_file <- data_path("derived", "indexes.Rds")
+if (file.exists(indexes_file)) {
+  indexes <- readRDS(indexes_file)
+} else {
+  stop("Run 02-compute-eofs.R")
+}
+
 indexes_wide <- dcast(indexes, + lev + time ~ term, value.var = "estimate_norm")

### 2020-06-30 analysis/scripts/02-compute-eofs.R

@@ -1,0 +1,21 @@
+library(metR)
+library(magrittr)
+library(asymsam)
+
+indexes_file <- data_path("derived", "indexes.Rds")
+
+if (!file.exists(indexes_file)) {
+  hgt <- ReadNetCDF(ERA5(),
+                    subset = list(time = c("1979-01-01", "2018-12-31"),
+                                  latitude = c(-90, 10)),
+                    vars = c(hgt = "z")) %>%
+    normalise_coords() %>%
+    .[, hgt := hgt/9.8] %>%
+    .[, hgt_a := hgt - mean(hgt), by = .(lon, lat, lev, month(time))]
+
+  indexes <- hgt[lat <= -20, eof_asym(hgt_a, lon, lat, time, n = 1), by = lev]
+
+  saveRDS(indexes, file = data_path("derived", "indexes.Rds"))
+
+}
+

### 2020-06-30 analysis/paper/paper.Rmd
seasonal pp regression

@@
 ```{r compute-pp-regr} 
-pp_regr <- CMAP() %>% 
+pp <- CMAP() %>% 
   ReadNetCDF() %>% 
   normalise_coords() %>% 
-  .[, precip := Anomaly(precip), by = .(lon, lat)] %>% 
+  .[, precip := Anomaly(precip), by = .(lon, lat)]
+
+pp_regr <- pp %>% 
   .[indexes_wide[lev == 700], on = .NATURAL] %>% 
@@
   scale_y_latitude() +
-  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
+  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
+                       breaks = AnchorBreaks(0, 0.1, 0)) +
   coord_quickmap() +
@@
 
-```{r}
+```{r regr-pp-season, fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
+pp_regr_season <- pp %>% 
+  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
+  .[indexes_wide[lev == 700], on = .NATURAL] %>% 
+  rm_singleton() %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
+  .[, nas := mean(is.na(precip)), by = .(lon, lat, season(time))] %>% 
+  .[nas >= 0.15, precip := NA] %>% 
+  .[, rbind(as.data.table(FitLm(precip, asym, sym, se = TRUE)),
+            as.data.table(FitLm(precip, full, se = TRUE))),
+    by = .(lon, lat, season(time))] %>% 
+  rm_intercept() %>% 
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)]
+
+pp_regr_season %>% 
+  ggplot(aes(lon, lat)) +
+  # geom_raster(aes(fill = estimate)) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.5, 0)) +
+  geom_contour2(aes(z = p.value), breaks = 0.01) +
+  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
+  stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
+  geom_qmap() +
+  scale_x_longitude() +
+  scale_y_latitude() +
+  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
+                       breaks = AnchorBreaks(0, 0.5, 0)) +
+  coord_quickmap() +
+  facet_grid(season~term, labeller = labeller(term = lab_sam))
+```
+
+```{r compute-air-regr}
 time_subset <- list(time = range(indexes$time))
@@
 
-```{r, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
+```{r regr-air, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
 air_regr %>%
@@
 
-```{r, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
+```{r regr-air-season, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
 air_regr_season <- air %>% 
@@
 
-```{r}
+```{r git-info, cache = FALSE}
 # what commit is this file at? 

### 2020-07-07 analysis/paper/paper.Rmd
paper figures

--- 2020-06-30 analysis/paper/paper.Rmd
@@
 ---
-title: "SAM simétrico y asimétrico"
-author: "Elio Campitelli"
+twocol: TRUE  #  Use FALSE for submission. 
+title: Title here
+author1: 
+  name: Elio Campitelli
+  email: elio.campitelli@cima.fcen.uba.ar
+author2: 
+  name: Leandro Díaz
+affiliation: "CIMA UBA blablabla"
+extra_authors: 
+  - name: Carolina Vera
+abstract: |
+  Enter the text of your abstract here.
+This is a sample American Meteorological Society (AMS) \LaTeX\ template.
+This document provides authors with instructions on the use of the AMS \LaTeX\ template.
+Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document.
+The template.tex file should be modified by authors for their own manuscript.
+statement: |
+  This is significant becasue... 
+bibliography: references.bib
 output: 
-   # pdf_document:
-   #    latex_engine: xelatex
-   powerpoint_presentation:
-    # ioslides_presentation:
-        fig_height: 7.6
-        fig_width: 12.8
-        reference_doc: "../templates/template.pptx" 
-        slide_level: 2
-urlcolor: blue
-header_includes:
-   - \usepackage[utf8]{inputenc}
+  rticles::ams_article:
+    latex_engine: "xelatex"
 ---
 
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(fig.path = "",     # AMS required
+                      out.extra = "",    # To force figure labels
+                      echo = FALSE)      # Don't show R code
 
-<!-- With the following code you can access and display values from the yml header above. -->
-<!-- 
-Keywords: `r rmarkdown::metadata$keywords`
+# This helps to use appendix figures.
+appendcaption <- function (x, options) {
+  `%n%` <- knitr:::`%n%`
+  rw = options$resize.width
+  rh = options$resize.height
+  resize1 = resize2 = ""
+  if (!is.null(rw) || !is.null(rh)) {
+    resize1 = sprintf("\\resizebox{%s}{%s}{", rw %n% "!", 
+                      rh %n% "!")
+    resize2 = "} "
+  }
+  tikz = knitr:::is_tikz_dev(options)
+  a = options$fig.align
+  fig.cur = options$fig.cur %n% 1L
+  fig.num = options$fig.num %n% 1L
+  animate = options$fig.show == "animate"
+  fig.ncol = options$fig.ncol %n% fig.num
+  if (is.null(fig.sep <- options$fig.sep)) {
+    fig.sep = character(fig.num)
+    if (fig.ncol < fig.num) 
+      fig.sep[seq(fig.ncol, fig.num - 1L, fig.ncol)] = "\\newline"
+  }
+  sep.cur = NULL
+  if (!tikz && animate && fig.cur < fig.num) 
+    return("")
+  usesub = length(subcap <- options$fig.subcap) && fig.num > 
+    1
+  ai = options$fig.show != "hold"
+  plot1 = ai || fig.cur <= 1L
+  plot2 = ai || fig.cur == fig.num
+  align1 = if (plot1) 
+    switch(a, left = "\n\n", center = "\n\n{\\centering ", 
+           right = "\n\n\\hfill{}", "\n")
+  align2 = if (plot2) 
+    switch(a, left = "\\hfill{}\n\n", center = "\n\n}\n\n", 
+           right = "\n\n", "")
+  cap = options$fig.cap
+  appendix = !is.null(options$appendix)
+  scap = options$fig.scap
+  fig1 = fig2 = ""
+  mcap = fig.num > 1L && options$fig.show == "asis" && !length(subcap)
+  sub1 = sub2 = ""
+  if (length(cap) && !is.na(cap)) {
+    lab = paste0(options$fig.lp, options$label)
+    if (plot1) {
+      pos = options$fig.pos
+      if (pos != "" && !grepl("^[[{]", pos)) 
+        pos = sprintf("[%s]", pos)
+      fig1 = sprintf("\\begin{%s}%s", options$fig.env, 
+                     pos)
+    }
+    if (usesub) {
+      sub1 = sprintf("\\subfloat[%s%s]{", subcap, knitr:::create_label(lab, 
+                                                               fig.cur, latex = TRUE))
+      sub2 = "}"
+      sep.cur = fig.sep[fig.cur]
+      if (is.na(sep.cur)) 
+        sep.cur = NULL
+    }
+    if (plot2) {
+      if (is.null(scap) && !grepl("[{].*?[:.;].*?[}]", 
+                                  cap)) {
+        scap = strsplit(cap, "[:.;]( |\\\\|$)")[[1L]][1L]
+      }
+      scap = if (is.null(scap) || is.na(scap)) 
+        ""
+      else sprintf("[%s]", scap)
+      if (cap == "") {
+        cap = ""
+      } else if (appendix == TRUE) {
+        cap = sprintf("\\appendcaption{%s}{%s}%s\n", options$label, cap, knitr:::create_label(lab, 
+                                                                  if (mcap) 
+                                                                    fig.cur, latex = TRUE))
+      } else {
+        cap = sprintf("\\caption%s{%s}%s\n", scap, cap, knitr:::create_label(lab, 
+                                                                  if (mcap) 
+                                                                    fig.cur, latex = TRUE))
+      }
+      fig2 = sprintf("%s\\end{%s}\n", cap, options$fig.env)
+    }
+  }
+  else if (knitr:::pandoc_to(c("latex", "beamer"))) {
+    align.env = switch(a, left = "flushleft", center = "center", 
+                       right = "flushright")
+    align1 = if (plot1) 
+      if (a == "default") 
+        "\n"
+    else sprintf("\n\n\\begin{%s}", align.env)
+    align2 = if (plot2) 
+      if (a == "default") 
+        ""
+    else sprintf("\\end{%s}\n\n", align.env)
+  }
+  ow = options$out.width
+  if (animate && identical(ow, "\\maxwidth")) 
+    ow = NULL
+  if (is.numeric(ow)) 
+    ow = paste0(ow, "px")
+  size = paste(c(sprintf("width=%s", ow), sprintf("height=%s", 
+                                                  options$out.height), options$out.extra), collapse = ",")
+  paste0(fig1, align1, sub1, resize1, if (tikz) {
+    sprintf("\\input{%s}", x)
+  }
+  else if (animate) {
+    aniopts = options$aniopts
+    aniopts = if (is.na(aniopts)) 
+      NULL
+    else gsub(";", ",", aniopts)
+    size = paste(c(size, sprintf("%s", aniopts)), collapse = ",")
+    if (nzchar(size)) 
+      size = sprintf("[%s]", size)
+    sprintf("\\animategraphics%s{%s}{%s}{%s}{%s}", size, 
+            1/options$interval, sub(sprintf("%d$", fig.num), 
+                                    "", xfun::sans_ext(x)), 1L, fig.num)
+  }
+  else {
+    if (nzchar(size)) 
+      size = sprintf("[%s]", size)
+    res = sprintf("\\includegraphics%s{%s} ", size, if (getOption("knitr.include_graphics.ext", 
+                                                                  FALSE)) 
+      x
+      else xfun::sans_ext(x))
+    lnk = options$fig.link
+    if (is.null(lnk) || is.na(lnk)) 
+      res
+    else sprintf("\\href{%s}{%s}", lnk, res)
+  }, resize2, sub2, sep.cur, align2, fig2)
+}
 
-Highlights: `r rmarkdown::metadata$highlights`
--->
+knitr::knit_hooks$set(plot = appendcaption)
+```
 
-<!-- The following code chunk defines some general settings how code chunks should behave. -->
-
-```{r setup, include=FALSE}
+```{r, warning=FALSE, message = FALSE}
 knitr::opts_chunk$set(
@@
   comment = "#>",
-  fig.path = "../figures/",
+  fig.path = "",
+  cache.extra = 2,
   dpi = 300
@@
 
-theme_set(theme_asymsam())
+theme_set(theme_asymsam(base_size = 9))
 ZeroBreaks <- AnchorBreaks(0, NULL, 0)
-
+axis_labs_smol <- theme(axis.text = element_text(size = 7))
 lev.lab <- AddSuffix(" hPa")
@@
 
+knitr::opts_hooks$set(invisible = function(options) {
+  options$include = FALSE
+  options
+})
+
 main_levs <- c(700, 300, 30)
 ```
+
+<!-- The actual document text starts here: -->
+
+# Introduction
+
+blabalblba introdocction. SAM... yada yada circulation.. 
+`r paste(stringi::stri_rand_lipsum(3), collapse = "\n\n")`
 
@@
 
-<!-- The actual document text starts here: -->
+# Methods
 
-# Introduction
+### Data
 
-# Methods
+We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach].
+
+Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0 degree latitude x 5.0 degree longitude global grid [@vose2012; @smith2008].
+
+We used monthly precipitation data from CPC Merged Analysis of Precipitation [@xie1997] 2.5 degree latitude x 2.5 degree longitude.
+
+### Definition of indexes
+
+`r paste(stringi::stri_rand_lipsum(4, FALSE), collapse = "\n\n")`
 
@@
 Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right).
-Arbitrary units."}
+Arbitrary units.", fig.width=6, fig.height=3, fig.env="figure*"}
 sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
@@
   melt(id.vars = c("lon", "lat")) %>% 
-
-  # .[variable == "asym", value := value + rnorm(.N, 0, sd = 1e-6)] %>%
+  .[variable == "sym", value := value + rnorm(.N, 0, sd = 1e-6)] %>%
   .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)] %>% 
@@
   ggplot(aes(lon, lat)) +
-  # geom_contour2(aes(z = value, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
-  geom_contour_fill(aes(z = value), global.breaks = FALSE) +
-  geom_contour_tanaka(aes(z = value), global.breaks = FALSE) +
+  # geom_contour(aes(z = value, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
+  geom_contour_fill(aes(z = value), global.breaks = FALSE, breaks = ZeroBreaks) +
+  geom_contour_tanaka(aes(z = value), global.breaks = FALSE, breaks = ZeroBreaks) +
   geom_qmap() +
   scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20)) +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
   scale_fill_divergent(guide = "none") +
   coord_polar() +
-  facet_grid(~variable, labeller = labeller(variable = lab_sam, lev = lev.lab)) 
+  facet_grid(~variable, labeller = labeller(variable = lab_sam, lev = lev.lab)) +
+  axis_labs_smol
 ```
@@
 
-```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM."}
+`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
+
+```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM.", fig.width=6, fig.height=3, fig.env="figure*"}
 indexes %>% 
   .[lev %in% c(30, 700) & term != "full"] %>% 
+  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
   ggplot(aes(time, estimate_norm)) +
   geom_line(aes(color = term)) +
-  # geom_smooth(method = "lm", aes(color = term)) +
   scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
@@
   scale_x_datetime(NULL) +
-  facet_grid(lev~., labeller = labeller(lev = lev.lab))
+  facet_grid(lev~., labeller = labeller(lev = lev.lab), scales = "free")
 ```
 
-```{r asymsam-timeseries-seasonally, fig.cap = "Seasonal time series for the asymmetric SAM and symmetric SAM."}
-indexes %>% 
-  .[lev %in% c(30, 700) & term != "full"] %>% 
-  .[, .(estimate_norm = mean(estimate_norm)), by = .(term, lev, time = seasonally(time))] %>% 
-  .[, season := season(time)] %>% 
-  ggplot(aes(time, estimate_norm)) +
-  geom_line(aes(color = term)) +
-  # geom_smooth(method = "lm", aes(color = term)) +
-  scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
-  scale_y_continuous(NULL) +
-  scale_x_datetime(NULL) +
-  facet_grid(season~lev, labeller = labeller(lev = lev.lab))
-```
+`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
 
-```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level."}
+`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
+
+`r paste(stringi::stri_rand_lipsum(1, FALSE), collapse = "\n\n")`
+
+```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level.", fig.width=3, fig.height=3.5}
 indexes_wide[, cor(asym, sym), by = lev] %>% 
@@
 
-```{r cross-correlation, fig.cap = "Cross correlation between levels of the Full, Asymmetric and Symmetric SAM."}
+`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
+
+```{r cross-correlation, fig.cap = "Cross correlation between levels of the Full, Asymmetric and Symmetric SAM.", fig.width = 6, fig.height=4, fig.env="figure*"}
+tri <- matrix(c(1000, 1000, 1000, 1, 1, 1, 1000, 1000), ncol = 2, byrow = TRUE) %>% 
+  as.data.table() %>% 
+  setnames(., colnames(.), c("item1", "item2"))
+
+logbreaks <- function(range) {
+  ticks <- ggplot2:::calc_logticks(minpow = log10(min(range)), maxpow = log10(max(range)))
+  ticks$value[ticks$end == 0.2]
+}
+
 indexes %>% 
   copy() %>% 
-  .[, widyr::pairwise_cor(.SD, lev, time, estimate), by = .(term)] %>% 
-  ggplot(aes(item1, item2)) +
-  geom_contour_fill(aes(z = correlation), na.fill = 1) +
-  scale_fill_divergent("Cross-correlation") +
-  scale_x_level() +
-  scale_y_level() +
+  .[, widyr::pairwise_cor(.SD, lev, time, estimate, diag = FALSE), by = .(term)] %>% 
+  ggplot(aes(item2, item1)) +
+  geom_contour_fill(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
+  geom_polygon(data = tri, fill = "white") +
+  scale_fill_divergent("Cross-correlation", 
+                       breaks = MakeBreaks(binwidth = 0.1),
+                       guide = guide_colorstrip_bottom()) +
+  scale_x_level(minor_breaks = logbreaks, position = "top") +
+  scale_y_level(minor_breaks = logbreaks) +
   coord_equal() +
-  facet_grid(. ~ term, labeller = labeller(term = lab_sam))
+  facet_grid(. ~ term, labeller = labeller(term = lab_sam)) +
+  theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside")
 ```
 
+`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
+
 ```{r compute-regr-pattern}
-indexes_2d <- hgt[lev %in% main_levs] %>% 
+indexes_2d_yearly <- hgt[lev %in% main_levs] %>% 
   indexes_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
@@
 ```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef."}
-indexes_2d %>% 
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.env = "figure*", fig.width=6, fig.height=7}
+indexes_2d_yearly %>% 
   .[lat <= -20] %>% 
@@
   scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20)) +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
-                       limits = c(-80, 80),
+                       limits = c(-85, 85),
                        oob = scales::squish,
@@
   coord_polar() +
-  facet_grid(term ~ lev, labeller = labeller(term = lab_sam, lev = lev.lab))
+  facet_grid(term ~ lev, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  axis_labs_smol 
 ```
+`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
 
@@
   .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] 
-
 ```
@@
 ```{r 2d-regr-700, fig.cap = "Seasonal regression patterns of geopotential height at 700 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef."}
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
 indexes_2d %>% 
@@
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
-  # geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
-  # size = 0.3,
-  # global.breaks = FALSE) +
+  geom_contour_fill(aes(z = estimate),  breaks = AnchorBreaks(0, binwidth = 5, 0),
+                    global.breaks = FALSE) +
   geom_contour_tanaka(aes(z = estimate),
-                      breaks = ZeroBreaks,
+                      breaks = AnchorBreaks(0, binwidth = 5, 0),
                       size = 0.3,
@@
   geom_qmap(~.x[lat <= -20]) +
-  # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
   scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20)) +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
-                       limits = c(-80, 80),
+                       limits = c(-55, 55),
                        oob = scales::squish,
-                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
+                       breaks = AnchorBreaks(0, binwidth = 5, 0)) +
   scale_linetype(guide = "none") +
   coord_polar() +
-  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab))
+  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  axis_labs_smol
 ```
 
+`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
+
+```{r wave-amplitude-700, fig.cap = "Planteray wave amplitude for the regression patterns at 700 hPa.", fig.width=6, fig.height=4}
+indexes_2d %>% 
+  .[ lev == 700 & lat < -20] %>%
+  .[, FitWave(estimate, 0:3), by = .(lat, lev, season, term)] %>% 
+  ggplot(aes(lat, amplitude)) +
+  geom_line(aes(color = factor(k))) +
+  scale_color_brewer("Wave-number", palette = "Set1") +
+  scale_y_continuous("Amplitude") +
+  scale_x_latitude(ticks = 15) +
+  coord_flip() +
+  facet_grid(term~season, labeller = labeller(term = lab_sam), scales = "free_x")
+```
+
+`r paste(stringi::stri_rand_lipsum(1, FALSE), collapse = "\n\n")`
+
 ```{r 2d-regr-30, fig.cap = "Seasonal regression patterns of geopotential height at 30 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef."}
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
 indexes_2d %>% 
@@
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
-  # geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
-  # size = 0.3,
-  # global.breaks = FALSE) +
+  geom_contour_fill(aes(z = estimate),  breaks = AnchorBreaks(0, binwidth = 20, 0),
+                    global.breaks = FALSE) +
   geom_contour_tanaka(aes(z = estimate),
-                      breaks = ZeroBreaks,
+                      breaks = AnchorBreaks(0, binwidth = 20, 0),
                       size = 0.3,
@@
   geom_qmap(~.x[lat <= -20]) +
-  # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
   scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20)) +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
-                       limits = c(-80, 80),
+                       limits = c(-105, 105),
                        oob = scales::squish,
-                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
+                       breaks = AnchorBreaks(0, binwidth = 20, 0)) +
   scale_linetype(guide = "none") +
   coord_polar() +
-  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab))
+  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  axis_labs_smol
 ```
 
-```{r wave-amplitude-700, fig.cap = "Planteray wave squared amplitudes for the regression patterns at 700 hPa."}
+`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
+
+```{r wave-amplitude-30, fig.cap = "Planteray wave amplitude for the regression patterns at 30 hPa.", fig.width=6, fig.height=4}
 indexes_2d %>% 
-  .[ lev == 700 & lat < -20] %>%
-  .[, FitWave(estimate, 1:3), by = .(lat, lev, season, term)] %>% 
-  ggplot(aes(lat, amplitude^2)) +
+  .[ lev == 30 & lat < -20] %>%
+  .[, FitWave(estimate, 0:3), by = .(lat, lev, season, term)] %>% 
+  ggplot(aes(lat, amplitude)) +
   geom_line(aes(color = factor(k))) +
@@
   coord_flip() +
-  facet_grid(term~season, labeller = labeller(term = lab_sam)) 
+  facet_grid(term~season, labeller = labeller(term = lab_sam), scales = "free_x") 
+```
+`r paste(stringi::stri_rand_lipsum(1, FALSE), collapse = "\n\n")`
+
+```{r compute-pp-regr} 
+pp <- CMAP() %>% 
+  ReadNetCDF(subset = list(lat = c(-90, -20))) %>% 
+  normalise_coords() %>% 
+  .[, precip := Anomaly(precip), by = .(lon, lat)]
+
+pp_regr_season <- pp %>% 
+  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
+  .[indexes_wide[lev == 700], on = .NATURAL] %>% 
+  rm_singleton() %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
+  .[, nas := mean(is.na(precip)), by = .(lon, lat, season(time))] %>% 
+  .[nas >= 0.15, precip := NA] %>% 
+  .[, rbind(as.data.table(FitLm(precip, asym, sym, se = TRUE)),
+            as.data.table(FitLm(precip, full, se = TRUE))),
+    by = .(lon, lat, season(time))] %>% 
+  rm_intercept() %>% 
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)]
+
 ```
 
-```{r wave-amplitude-30, fig.cap = "Planteray wave squared amplitudes for the regression patterns at 30 hPa."}
-indexes_2d %>% 
-  .[ lev == 30 & lat < -20] %>%
-  .[, FitWave(estimate, 1:3), by = .(lat, lev, season, term)] %>% 
-  ggplot(aes(lat, amplitude^2)) +
-  geom_line(aes(color = factor(k))) +
-  scale_color_brewer("Wave-number", palette = "Set1") +
-  scale_y_continuous("Squared amplitude") +
-  scale_x_latitude(ticks = 15) +
-  coord_flip() +
-  facet_grid(term~season, labeller = labeller(term = lab_sam)) 
-```
-
-```{r compute-pp-regr} 
-pp_regr <- CMAP() %>% 
-  ReadNetCDF() %>% 
-  normalise_coords() %>% 
-  .[, precip := Anomaly(precip), by = .(lon, lat)] %>% 
-  .[indexes_wide[lev == 700], on = .NATURAL] %>% 
-  .[, FitLm(precip, asym, sym, se = TRUE, r2 = TRUE), by = .(lon, lat)] %>% 
-  rm_intercept() %>% 
-  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)]
-```
-
-```{r regr-pp, fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
-pp_regr %>% 
+```{r pp-regr-global, fig.cap = "caption", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
+pp_regr_season %>%
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), na.fill = TRUE, breaks = AnchorBreaks(0, 0.1, 0)) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.25, 0)) +
   geom_contour2(aes(z = p.value), breaks = 0.01) +
@@
   stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
-  geom_qmap() +
+  geom_qmap(~.x[lat <= -20]) +
+  scale_x_longitude() +
+  scale_y_latitude(labels = NULL) +
+  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
+                       breaks = AnchorBreaks(0, 0.25, 0)) +
+  coord_polar() +
+  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  axis_labs_smol
+```
+
+```{r regr-pp-season, fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", invisible = TRUE}
+
+is.sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
+  data[, lon %between% lons & lat %between% lats]
+}
+
+filter_sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
+  data[is.sa(data, lons = lons, lats = lats)]
+}
+
+pp_regr_season %>% 
+  filter_sa() %>% 
+  .[, lon := ConvertLongitude(lon)] %>% 
+  ggplot(aes(lon, lat)) +
+  # geom_raster(aes(fill = estimate)) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.25, 0)) +
+  geom_contour2(aes(z = p.value), breaks = 0.01) +
+  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
+  stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
+  geom_qmap(~filter_sa(.x)[, lon := ConvertLongitude(lon)], keep = 0.02) +
   scale_x_longitude() +
   scale_y_latitude() +
-  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
-  coord_quickmap() +
-  facet_wrap(~term, labeller = labeller(term = lab_sam))
+  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
+                       breaks = AnchorBreaks(0, 0.25, 0)) +
+  # coord_polar() +
+  ggalt::coord_proj("+proj=moll +lon_0=-60") +
+  facet_grid(season~term, labeller = labeller(term = lab_sam)) +
+  axis_labs_smol
 ```
 
-```{r}
+```{r compute-air-regr}
 time_subset <- list(time = range(indexes$time))
 air <- NOAATemp() %>% 
-  ReadNetCDF(vars = "air", subset = time_subset) %>% 
+  ReadNetCDF(vars = "air", subset = c(time_subset, list(lat = c(-90, 10)))) %>% 
   normalise_coords()
@@
 air_regr <- air %>% 
-  copy() %>% 
-  .[, nas := mean(is.na(air)), by = .(lon, lat)] %>% 
-  .[nas >= 0.15, air := NA] %>% 
-  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
-  .[indexes_wide[lev == 700], on = .NATURAL] %>% 
-  .[, rbind(as.data.table(FitLm(air, asym, sym, se = TRUE)),
-            as.data.table(FitLm(air, full, se = TRUE))),
-    by = .(lon, lat)] %>% 
-  rm_intercept()
-```
-
-```{r, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
-air_regr %>%
-  copy() %>% 
-  .[lat <= -60, estimate := NA] %>%
-  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)] %>% 
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), na.fill = 0,
-                    breaks = AnchorBreaks(0, 0.05, 0)) +
-  geom_contour2(aes(z = p.value), breaks = 0.05) +
-  stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat)), size = 0.4, alpha = 0.8) +
-  stat_subset(aes(subset = is.na(estimate)), geom = "raster", fill = "gray50") +
-  geom_qmap() +
-  scale_x_longitude() +
-  scale_y_latitude() +
-  scale_fill_divergent("Surface temperature", 
-                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.05, 0)) +
-  coord_quickmap() +
-  facet_grid(.~term, labeller = labeller(term = lab_sam))
-```
-
-```{r, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
-air_regr_season <- air %>% 
   # .[, air := Anomaly(air), by = .(lon, lat)] %>%
@@
   rm_intercept() 
+```
 
-air_regr_season %>% 
+```{r regr-air-season, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", fig.width = 10, fig.height = 7, fig.env = "sidewaysfigure"}
+air_regr %>% 
   copy() %>% 
@@
   stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat)), size = 0.4, alpha = 0.8) +
-  stat_subset(aes(subset = is.na(estimate)), geom = "raster", fill = "gray50") +
-  geom_qmap() +
+  # stat_subset(aes(subset = is.na(estimate)), geom = "raster", fill = "gray50") +
+  geom_qmap(~.x[lat <= 10]) +
   scale_x_longitude() +
-  scale_y_latitude() +
+  scale_y_latitude(labels = NULL) +
   scale_fill_divergent("Surface temperature",  limits = c(-0.8, .8), oob = scales::squish,
                        guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.15, 0)) +
-  coord_quickmap() +
-  facet_grid(season~term, labeller = labeller(term = lab_sam))
+  coord_polar() +
+  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  axis_labs_smol
 ```
 
-# Discussion
+\acknowledgments
+CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/
 
-# Conclusion
+NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/
 
-# Acknowledgements
+\appendix
 
-<!-- The following line inserts a page break  -->
+\appendixtitle{Extra figures}
 
-\newpage
+```{r lag-correlation, fig.cap = "Lag-correlation between Symmetric and Asymmetric SAM at each level.", fig.width = 4, fig.height=4, appendix = TRUE}
+ccf_dt <- function(..., plot = FALSE) {
+  stat <- ccf(..., plot = plot)
+  list(lag = stat$lag[, 1, 1], 
+       acf = stat$acf[, 1, 1])
+}
 
-# References
-
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-
-<div id="refs"></div>
-
-\newpage
-
-### Colophon
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
+indexes_wide %>% 
+  .[, ccf_dt(asym, sym, lag.max = 5), by = .(lev)] %>% 
+  ggplot(aes(lag, lev)) +
+  geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
+  geom_vline(xintercept = 0) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.05)) +
+  scale_x_continuous("\u2190 Asym leads Sym\t \t\t Sym leads Asym \U2192", breaks = seq(-5, 5, 1), 
+                     expand = c(0, 0)) +
+  scale_y_level()
 ```
 
-The current Git commit details are:
+```{r}
+my_spectrum <- function(data, spans = NULL, R = 1000, ..., probs = 0.95) {
+  mtm <- spec.pgram(data, spans = spans, ..., plot = FALSE)
 
-```{r}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+  out <- as.data.table(mtm[c("freq", "spec")])
+
+  out[, boot_null := null_spec(data, spans = spans, R = R, ..., probs = probs)]
+
+  return(out[])
+}
+
+null_spec <- memoise::memoise(function(data, spans, R = 1000, ..., probs = 0.95) {
+
+  b <- boot::boot(data, function(d, i) spec.pgram(d[i], spans = spans, 
+                                                  ...,
+                                                  plot = FALSE)$spec, 
+                  R = R)
+
+  apply(b$t, 2, quantile, probs = probs)
+})
+
+indexes %>% 
+  copy() %>% 
+  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
+  .[, my_spectrum(estimate_norm, 5), by = .(term, lev)] -> spec
 ```
+
+```{r spectrum, fig.cap = "Foutier spectrum", appendix = TRUE}
+spec %>%
+  .[lev %in% c(30, 700)] %>% 
+  ggplot(aes(1/freq/12, spec)) +
+  # geom_ribbon(aes(group = interaction(term, rleid(sign(spec - boot_null))), 
+  #                 fill = sign(spec - boot_null),
+  #                 ymin = spec, ymax = boot_null)) +
+    geom_line(aes(color = term)) +
+  geom_line(aes(color = term, y = boot_null)) +
+  annotation_logticks(sides = "b") +
+  scale_color_brewer(palette = "Set1") +
+  scale_fill_divergent() +
+  scale_x_log10() +
+  facet_grid(lev ~ .,labeller = labeller(lev = lev.lab))
+```
+
+```{r trends, fig.cap = "Trends for each index at each level. Shading indicates the 95\\% confidence interval.", fig.width = 3, fig.height=3.5, appendix = TRUE}
+indexes %>%
+  copy() %>% 
+  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
+  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
+  rm_intercept() %>% 
+  .[, term := NULL] %>% 
+  .[, estimate := estimate*3600*24*365*10] %>% 
+  .[, std.error := std.error*3600*24*365*10] %>% 
+  setnames("type", "term") %>% 
+  ggplot(aes(lev, estimate)) +
+  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96, fill = term), alpha = 0.3) +
+  geom_line(aes(color = term)) + 
+  geom_hline(yintercept = 0) +
+  scale_x_level() +
+  coord_flip() 
+```

### 2020-07-07 analysis/paper/paper.tex
paper figures

%% Version 5.0, 2 January 2020
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TemplateV5.tex --  LaTeX-based template for submissions to the
% American Meteorological Society
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PREAMBLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Start with one of the following:
% DOUBLE-SPACED VERSION FOR SUBMISSION TO THE AMS
\documentclass[twocol]{ametsocV5}

% TWO-COLUMN JOURNAL PAGE LAYOUT---FOR AUTHOR USE ONLY
% \documentclass[twocol]{ametsocV5}

% Enter packages here. If too many math alphabets are used,
% remove unnecessary packages or define hmmax and bmmax as necessary.

\newcommand{\hmmax}{0}
\newcommand{\bmmax}{0}
\usepackage{amsmath,amsfonts,amssymb,bm}
\usepackage{mathptmx}%{times}
\usepackage{newtxtext}
\usepackage{newtxmath}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% To be entered by author:

%% May use \\ to break lines in title:

\title{Title here}

\authors{Elio Campitelli
\correspondingauthor{Elio Campitelli,elio.campitelli@cima.fcen.uba.ar}
and Leandro Díaz
}

\affiliation{CIMA UBA blablabla}

\extraauthor{Carolina Vera
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%
% Enter your abstract here
% Abstracts should not exceed 250 words in length!
%

\abstract{Enter the text of your abstract here. This is a sample American
Meteorological Society (AMS) \LaTeX~template. This document provides
authors with instructions on the use of the AMS \LaTeX~template. Authors
should refer to the file amspaper.tex to review the actual \LaTeX~code
used to create this document. The template.tex file should be modified
by authors for their own manuscript.}

\begin{document}

%% Necessary!
\maketitle

\bibliographystyle{ametsoc2014}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% SIGNIFICANCE STATEMENT/CAPSULE SUMMARY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% If you are including an optional significance statement for a journal article or a required capsule summary for BAMS
% (see www.ametsoc.org/ams/index.cfm/publications/authors/journal-and-bams-authors/formatting-and-manuscript-components for details),
% please apply the necessary command as shown below:
%
\statement
This is significant becasue\ldots{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MAIN BODY OF PAPER
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

\section{Introduction}

blabalblba introdocction. SAM\ldots{} yada yada circulation.. Lorem
ipsum dolor sit amet, hac curae turpis sollicitudin vulputate quisque
sapien nullam curae integer. Torquent porttitor et et vulputate blandit
velit sed sed at velit eget arcu. Senectus, taciti justo eu sem purus
vestibulum pellentesque quis ac. Ac facilisi, amet consectetur mauris
fringilla sit. Est fusce, augue orci dictumst lacinia et elementum
dapibus orci eu, luctus ut luctus. Phasellus magna ut felis vitae metus
non non. In maecenas taciti. Hendrerit bibendum placerat. Molestie est
enim ante amet natoque consequat hendrerit, pharetra ridiculus.
Ultricies mauris donec leo eu, parturient praesent. Ridiculus sociis
ligula, ultricies augue, sagittis, turpis et. A netus at sagittis enim
nisl aliquet sed.

Justo id ut dolor faucibus ac justo orci volutpat vivamus nisl. Eu
sollicitudin, semper sed vitae et praesent vitae diam venenatis. Felis
vitae, velit risus id donec et. Neque nec id. Tellus nisl, vivamus sed
faucibus, non, per scelerisque ac arcu eget mus himenaeos imperdiet.
Purus eu quis felis mauris, ultricies et luctus integer orci. Lobortis
eget nec cras egestas feugiat. At mattis eget mus aliquam pellentesque
mauris elit sollicitudin tempor. Est mauris sed convallis at purus.

Purus scelerisque condimentum blandit interdum pharetra commodo. Neque
sed tristique natoque condimentum quis mollis et tortor egestas augue
sem est neque. Tortor dictum, pretium sociis auctor odio. Quam velit,
sapien, sed, tempor quis ut, ac auctor orci eleifend. Venenatis ut in
euismod nec nunc velit sed imperdiet. Lectus bibendum enim morbi dui sem
ut pellentesque venenatis diam. Ut, tincidunt senectus vel sed conubia
dapibus feugiat purus. Dui donec eu lacus semper tortor eget. Magna
proin amet justo. A lacus est ornare auctor sed vel et. Duis, amet
parturient nisl tempor praesent libero elementum erat ac et.

\section{Methods}

\subsubsection{Data}

We used monthly geopotential height at 2.5 longitude by 2.5 latitude
resolution from ERA5 \citep{hersbach}.

Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0
degree latitude x 5.0 degree longitude global grid
\citep{vose2012, smith2008}.

We used monthly precipitation data from CPC Merged Analysis of
Precipitation \citep{xie1997} 2.5 degree latitude x 2.5 degree
longitude.

\subsubsection{Definition of indexes}

Sagittis, malesuada, lacus integer vehicula. Diam curae, tempus feugiat
vivamus nascetur augue eu, vitae non arcu. Vivamus aptent viverra
blandit gravida consectetur eleifend habitasse? Aliquam ante. Venenatis
mauris parturient per, per et ut cum. Dapibus nulla tempor. Tempor
libero cubilia enim vestibulum. Et, magna elementum in conubia ut. Dis
non eleifend consectetur a et. Nulla libero nunc varius porta, eros eros
pretium platea malesuada velit. Dolor, suspendisse, suspendisse magna et
quis magnis ultricies! Luctus vestibulum arcu, neque sapien.

Nisl quis sed, fringilla laoreet felis. Felis ac vel curabitur himenaeos
pretium. Et enim sed vitae vitae nec ac dapibus aliquet. Massa libero
non egestas, tempus tempor commodo amet, suspendisse, orci. Nostra, orci
penatibus diam aliquet neque sed. Eu fusce, rutrum at ut in, curae. Ex
cum massa efficitur augue et nulla consequat ornare in. Elementum velit,
non aliquet ut. Ultricies ut arcu diam eros scelerisque lobortis semper.
In semper ligula varius natoque purus in commodo. Dis tellus elit quam
egestas maximus. Curae aliquam dolor platea, penatibus sed. Dui habitant
aenean habitasse nulla dui amet.

Vehicula sed ligula nunc fringilla. Ornare morbi, eu, non ac, et ipsum
phasellus. Felis non tincidunt litora. Hendrerit elit, nisi nam sem
auctor eu ultrices amet. Per accumsan pharetra sapien lacinia molestie,
suspendisse. In vehicula elementum vestibulum vestibulum ut taciti, erat
consequat sed. Non non et augue volutpat non ante congue quam. Risus mi
nec laoreet viverra non. Ligula platea porta nisi faucibus, vestibulum
lacus. Purus nec imperdiet sed sodales sed sed in sed mauris. Quam
semper suscipit nunc felis sed cras amet ultrices maecenas. Nullam
luctus magnis amet non id at, tempus, ullamcorper. Id interdum tortor
sed viverra imperdiet porttitor et phasellus. Ac facilisis mus.

Nec fusce, ipsum a porttitor, magnis in vestibulum. Eu cras mi,
inceptos. Nec class ad fames litora nisl pellentesque vestibulum. Nullam
fames amet diam aliquam eget. Donec, non mollis aptent urna condimentum
tellus, torquent. Sit nunc lectus interdum. Maximus eu mattis libero
pellentesque vel suspendisse ullamcorper. Auctor dolor non sociosqu et
porttitor pulvinar, turpis gravida in auctor ipsum cursus. Nulla
venenatis fringilla sollicitudin, accumsan libero nulla aliquet volutpat
mauris lectus at auctor! Parturient nisl, metus etiam eu aptent purus,
ac vestibulum placerat sed, a quam tellus. Vulputate placerat euismod.
Commodo urna, neque penatibus? Pharetra elementum curae, sed.

\begin{figure*}
\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height]{Spatial patterns of the first EOF of 700 hPa geopotential height. Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right). Arbitrary units.}\label{fig:method}
\end{figure*}

\section{Results}

Dignissim nibh sed eu, sed dapibus, tempor. Libero ligula accumsan enim
fringilla dapibus hendrerit integer quisque sapien. Nam convallis eu,
magnis. Mauris eu at bibendum, elit mus. Vehicula posuere amet viverra
volutpat nec fringilla orci netus. Malesuada imperdiet leo iaculis at
varius massa a sit leo ad odio cum. Justo orci quis in ante convallis.
Sed sit justo cras in, mus ac, elementum? Rhoncus fermentum, at risus ut
tincidunt nec, sit per, lacus. Eros inceptos posuere facilisis, sed,
magna suspendisse non sed sapien porttitor. Sit, augue facilisis sed. Et
senectus odio, ipsum fringilla et, iaculis duis. Erat ac integer
consectetur cubilia. Condimentum neque suscipit vel aliquet egestas
pellentesque.

Fringilla non dis nisl erat ut ligula. Eros egestas ut ut congue
lobortis sem? Aliquet nunc diam tortor. Semper conubia consequat nec ex
nibh laoreet. Sed malesuada luctus a viverra netus congue a diam ut
lorem. Vestibulum, nec varius nisl non ornare morbi. Sed tellus,
ultricies non ac. Ridiculus, cubilia dui sit ac massa suspendisse
maximus. Penatibus, taciti vel sed. Ut pellentesque vestibulum sed
consequat ultricies! Nisl mi tristique vulputate vitae tristique
consequat. At accumsan class est mattis. Interdum sociosqu lobortis
dapibus nullam fermentum himenaeos.

\begin{figure*}
\includegraphics{asymsam-timeseries-1} \caption[Time series for the asymmetric SAM and symmetric SAM]{Time series for the asymmetric SAM and symmetric SAM.}\label{fig:asymsam-timeseries}
\end{figure*}

Euismod non magnis, ligula sed eu velit enim. Laoreet nulla vel vivamus
et ultrices accumsan tempor fames at sed. Eleifend sociis tristique nibh
nunc tempor. Nostra sed iaculis, ut vehicula eu nulla sed venenatis.
Pharetra risus, orci erat convallis sit sociis sed tortor. Amet, mauris
eros magnis praesent aptent auctor scelerisque. Faucibus in tempus in
lobortis semper, tempus, in. Amet volutpat class tristique sapien sociis
habitant posuere et auctor. Nulla litora eu hac neque efficitur risus
donec. Nam dui fermentum mollis libero sapien, taciti et vulputate, in.
Eu vestibulum felis suscipit gravida aenean ligula. Cras eu nunc
conubia, sed nibh inceptos id accumsan. Lacus, eu ante aliquet.

Pulvinar magna phasellus amet per vel lacus ante lectus. Integer
vestibulum hac sodales nibh ac donec augue maximus. Ullamcorper vivamus
finibus enim quis. Fringilla nibh sed pellentesque adipiscing facilisis
sed. Pretium interdum nibh amet nunc quam velit imperdiet. In sed in
lobortis et conubia. Praesent sed pretium auctor. Sit eleifend laoreet
sed erat a ultrices cursus hendrerit nisl. Tellus habitasse aliquet
molestie pretium faucibus. Fames eu convallis penatibus morbi tincidunt
vestibulum, ligula leo. Pharetra facilisis ac mi. Dapibus congue eu eros
accumsan dictum suscipit in diam donec in interdum id. Dictumst ac id
amet iaculis ut aptent.

Netus quam. Euismod senectus donec pharetra felis eget. Penatibus eu in
varius morbi interdum erat. Sed felis finibus mauris interdum. Ut,
euismod, a lacus justo nulla porta faucibus ac nec, odio dignissim orci.
Et, porta, blandit tincidunt. Ullamcorper in egestas scelerisque luctus
accumsan mauris nam dis diam. Sollicitudin quis vestibulum cum fames
maximus odio aliquam sed. Aliquam ornare ut malesuada ut eu eros.

Leo maecenas aptent aenean diam, arcu. Pulvinar sociis vulputate eu non
amet maecenas posuere. Commodo leo phasellus. Eu ut suscipit nibh in sed
conubia fermentum sit tristique. Urna morbi consequat dui sed lorem
ridiculus eros, odio non neque? Ante dolor eu id hac. In risus dolor
nunc urna arcu. Consectetur sociis platea curabitur nulla, platea in eu.

Ut pharetra amet nam augue imperdiet. Eros eget orci cubilia rutrum
fusce eu in in. Magnis tincidunt orci. In tellus nisl suspendisse ex
ante mauris nam. A ac sed leo neque taciti tincidunt diam laoreet.
Semper odio. Rhoncus habitant sed sed mollis ultricies vehicula ipsum
urna ac nunc nulla nunc? Magna non mauris, montes ullamcorper. Dolor sed
maximus dui ac porttitor at cubilia conubia gravida ut.

\begin{figure}
\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric and Asymmetric SAM at each level]{Correlation between the Symmetric and Asymmetric SAM at each level.}\label{fig:cor-lev}
\end{figure}

Blandit rutrum, nisi vitae, a velit sodales eu. Vestibulum non, dolor
in, sit. Libero nullam sociosqu per tellus non nunc quam fusce nisl. Sed
metus etiam faucibus amet eget velit aliquam. Eget eget nibh commodo.
Consectetur tristique netus ac arcu. Ipsum ullamcorper nec euismod velit
quis lacinia mus. Enim lectus est in sed eu efficitur himenaeos auctor
nec. Justo libero montes pellentesque augue. Tempus laoreet pellentesque
sociosqu tincidunt varius ac eleifend, non vitae. Mattis, eleifend
sociis, nascetur eget sed aenean, nec.

Egestas finibus gravida luctus per facilisi. Ornare quam, erat amet
maximus nisi posuere potenti consequat ante eu arcu ac? Fermentum sociis
class justo accumsan sapien penatibus consequat, ut ex at dis, etiam.
Sem parturient facilisis montes feugiat. Et urna velit praesent vehicula
quisque magna in. Tellus, at sem eros fusce ultrices pretium inceptos
dui, tristique. Mi feugiat volutpat ut a condimentum ultricies, purus
pellentesque in. Morbi ut mauris conubia tellus mi, quis. Nulla
fermentum ad ut dui sem ultrices ut, ante. Per enim vestibulum sit
congue natoque egestas nec erat vehicula aenean bibendum. Non ultricies
ultrices nibh accumsan laoreet eros nunc nisi pulvinar sit sed.
Pellentesque turpis adipiscing efficitur aptent tincidunt facilisi mi,
quisque. Mi, pulvinar at non, phasellus.

\begin{figure*}
\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the Full, Asymmetric and Symmetric SAM]{Cross correlation between levels of the Full, Asymmetric and Symmetric SAM.}\label{fig:cross-correlation}
\end{figure*}

Consequat porttitor eget diam fusce phasellus a orci natoque in. Fusce
orci convallis vel. Eu aliquet ligula risus nunc lacinia sem ut. At
morbi ipsum nam senectus, odio. Amet et consequat erat. Tincidunt, ad
risus sed! Interdum eget enim nascetur sed eleifend nisl cum vivamus.
Placerat natoque eu risus ut nec, tellus tortor, donec. Quam litora
pulvinar ac sem, tortor. Sodales ligula praesent interdum erat enim
lectus donec dolor ligula donec. Et ipsum dis. Laoreet, id auctor
efficitur elementum vitae commodo posuere lorem sem lacus et, platea
sapien. Ligula primis inceptos accumsan, torquent. Elementum laoreet
vitae nulla aliquam in magna condimentum ex blandit proin.

Tincidunt tristique ante nulla sit a semper urna auctor. Ridiculus, nibh
sagittis. Ullamcorper bibendum in conubia consectetur montes, viverra.
Vivamus non diam. Vel efficitur interdum sed suscipit magna sed massa
magnis nunc. Taciti inceptos ipsum massa nunc. Ac, tellus curabitur amet
congue hac metus rhoncus. Posuere tortor rhoncus, nec ultrices enim sit
habitasse mattis eget platea. Integer dui at sapien tincidunt phasellus
enim at euismod lobortis.

\begin{figure*}
\includegraphics{2d-regr-1} \caption[Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM]{Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.}\label{fig:2d-regr}
\end{figure*}

Nisi porttitor et nibh consectetur porttitor a dapibus blandit faucibus.
Non dis quis donec, laoreet. Suspendisse primis, aenean justo ex nullam
ex cursus. Nec pulvinar dictumst sed. Conubia, nulla sit, et mi
malesuada at ex sit dui est. Mauris quisque lectus facilisis parturient
dictumst ultrices. Non a non, in, vestibulum, scelerisque ligula mauris,
dictum? Urna eget id litora in. Platea ac turpis ultrices enim sagittis
et purus? Sagittis sollicitudin est purus mauris ultricies nisi aliquet
ante, nam laoreet. Pellentesque nascetur adipiscing mus platea congue eu
ac velit. In ut tellus elementum neque ridiculus facilisis.

Leo nec suscipit sit duis. Taciti per tempus sed vel aliquet. Amet
feugiat. In ut ut dui per nisl ridiculus. Et urna venenatis scelerisque
ut nec feugiat ad. Elementum vestibulum aliquet ullamcorper magna aenean
lacus habitasse, laoreet mauris. Consectetur ornare molestie ultrices.
Diam magnis purus et lacinia curae. Faucibus nulla a. Sociis parturient
facilisis maximus nisi dis a. Tempus id venenatis vitae, cubilia
ultricies, eget nulla. Quis amet per velit. Ac mattis sed, placerat
donec ac, maximus tempor et quisque. Nulla vestibulum sed nulla eget
eget vestibulum dictumst gravida ac viverra habitant, placerat.

\begin{sidewaysfigure}
\includegraphics{2d-regr-700-1} \caption[Seasonal regression patterns of geopotential height at 700 hPa with the Full, Asymmetric and Symmetric SAM]{Seasonal regression patterns of geopotential height at 700 hPa with the Full, Asymmetric and Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.}\label{fig:2d-regr-700}
\end{sidewaysfigure}

Ligula nunc eget lacus fermentum, nisl ut donec venenatis. Primis sed.
Turpis vel tempor condimentum ac magnis nam, sed. Dolor blandit ornare
sed in ut nisl. Dignissim posuere aptent. Consequat ac neque at
himenaeos nulla. Faucibus nascetur, elit tellus. Fringilla erat sed
sodales class, lobortis. Amet sed suscipit. In et purus ante felis
tortor ex potenti. Sed sagittis, suspendisse aliquet sapien accumsan,
luctus suscipit. Turpis varius ex maecenas, nascetur arcu ultrices quis,
enim. Sed ullamcorper non, suspendisse praesent porttitor sit dui
aliquet. Fermentum dis sit nunc tristique, ac nec, praesent hendrerit.

Fermentum at odio eu ultrices amet, vel ridiculus integer. Mollis, sed
habitasse ut, mollis at id donec porttitor suspendisse? Sagittis magna
sed mauris mi euismod varius sapien pretium pharetra. Ex ac sed sapien,
orci. Sed sed feugiat eget justo a euismod. Augue magna nullam justo id
molestie egestas placerat. Ornare maximus in sed ridiculus, praesent.
Sed, nec blandit. Nascetur pellentesque ligula tellus congue.

\begin{figure}
\includegraphics{wave-amplitude-700-1} \caption[Planteray wave amplitude for the regression patterns at 700 hPa]{Planteray wave amplitude for the regression patterns at 700 hPa.}\label{fig:wave-amplitude-700}
\end{figure}

Velit ex et tortor dapibus pharetra convallis. A sed ridiculus.
Efficitur ut ad, ligula nisl, mi in cursus tellus quis dictum malesuada
rutrum elit venenatis. Ac et. Cum lorem fames ut. Ultricies sed a nunc
mi ligula. At ut maecenas luctus sociosqu laoreet. Duis netus quam
varius mi convallis nunc a. Non lacus sit nullam odio porttitor
vestibulum. Cum donec ligula per sed, elit. Quis nibh libero cubilia
viverra feugiat justo dapibus sit. Dictumst montes ut sit! Sed nunc
auctor nullam porttitor hendrerit ornare egestas. Vivamus ac sit, hac
cum ut, a proin, quis, non, pulvinar. Purus quam suscipit amet sed nec
feugiat dictumst elit vestibulum habitasse mauris.

\begin{sidewaysfigure}
\includegraphics{2d-regr-30-1} \caption[Seasonal regression patterns of geopotential height at 30 hPa with the Full, Asymmetric and Symmetric SAM]{Seasonal regression patterns of geopotential height at 30 hPa with the Full, Asymmetric and Symmetric SAM. The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.}\label{fig:2d-regr-30}
\end{sidewaysfigure}

Mus, bibendum viverra non ante velit auctor in, interdum turpis. Aliquet
egestas, quam porta a eu sed ac donec sed elementum sit. Ligula donec
nec facilisis efficitur finibus, egestas porta sed, nisi nunc, elit
pretium. Orci vestibulum dui vitae amet nulla, sagittis tincidunt sem
sit, amet. Mi risus penatibus, magna, risus turpis, ut posuere praesent
a fames pulvinar lacus. Sit maecenas leo vestibulum urna massa luctus,
vitae, justo eros libero ut! Condimentum a blandit tellus pellentesque
etiam ut mauris. Sollicitudin erat et amet, tempor donec aenean arcu
maximus leo etiam. Felis class lectus vulputate vulputate pellentesque
praesent. Non tristique quis molestie nec maecenas vel eget, felis dui.
Accumsan aenean vitae urna arcu nam mauris nisl. In inceptos magnis
mollis massa habitasse interdum etiam pretium facilisis. Velit erat
convallis etiam felis luctus sem amet posuere.

Etiam, enim, magna. Ut erat, commodo sem euismod vel quam dui. Efficitur
ridiculus id a, mattis facilisis cursus vitae, felis. In et et, est
maximus luctus sed habitant aptent. Eget libero justo interdum aenean
finibus mauris pulvinar et duis pulvinar. Nisl et in cum sagittis
luctus. Libero justo ut risus cras sapien. Felis porta, lacus ipsum,
condimentum eros integer. Nunc per lectus lorem per quisque. Tempus
magna sodales vestibulum faucibus.

\begin{figure}
\includegraphics{wave-amplitude-30-1} \caption[Planteray wave amplitude for the regression patterns at 30 hPa]{Planteray wave amplitude for the regression patterns at 30 hPa.}\label{fig:wave-amplitude-30}
\end{figure}

Himenaeos a per at ut, erat felis congue. Varius pharetra eu maecenas
non sed lectus amet egestas in. Aenean nullam nascetur mollis sit. In,
donec consequat per ridiculus platea risus risus neque. Facilisi risus
suspendisse imperdiet a at taciti lobortis. Fringilla egestas,
sollicitudin natoque. Nam diam eget sed, eleifend ut faucibus lacinia
maximus. Vestibulum aliquet curabitur sapien efficitur morbi elementum
nam donec turpis. Egestas egestas nunc auctor nunc cursus nec semper est
vel, felis.

\begin{sidewaysfigure}
\includegraphics{pp-regr-global-1} \caption[caption]{caption}\label{fig:pp-regr-global}
\end{sidewaysfigure}

\begin{sidewaysfigure}
\includegraphics{regr-air-season-1} \caption[Regression pattern of surface temperature with Asymmetric and Symmetric SAM]{Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:regr-air-season}
\end{sidewaysfigure}

\acknowledgments

CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder,
Colorado, USA, from their Web site at https://psl.noaa.gov/

NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the
NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at
https://psl.noaa.gov/

\appendix

\appendixtitle{Extra figures}

\begin{figure}
\includegraphics{lag-correlation-1} \appendcaption{lag-correlation}{Lag-correlation between Symmetric and Asymmetric SAM at each level.}\label{fig:lag-correlation}
\end{figure}

\begin{figure}
\includegraphics{spectrum-1} \appendcaption{spectrum}{Foutier spectrum}\label{fig:spectrum}
\end{figure}

\begin{figure}
\includegraphics{trends-1} \appendcaption{trends}{Trends for each index at each level. Shading indicates the 95\% confidence interval.}\label{fig:trends}
\end{figure}

\end{document}

### 2020-07-07 analysis/scripts/26-SAM_asim.Rmd
paper figures

--- 2020-06-25 analysis/paper/26-SAM_asim.Rmd
@@
 lev.lab <- AddSuffix(" hPa")
+
+guide_colorstrip_bottom <- function(width = 25, height = 0.5, ...) {
+   guide_colorstrip(title.position = "top", title.hjust = 0.5,
+                    barheight = height,
+                    barwidth = width, ...)
+}
+
 # knitr_set_timer()
@@
 
-```{r, fig.cap = "Patrón espacial del SAM. ERA5 en sombrado, NCEP en contornos."}
+```{r, fig.cap = "Spatial pattern of the first EOF of 700hPa geopotential height. Arbitrary units."}
 SAM[lev == 700, eof[[1]]$right, by = dataset] %>% 
@@
    scale_y_latitude() +
-   scale_fill_divergent() +
-   coord_polar()
+   scale_fill_divergent(guide = "none") +
+   coord_polar() +
+   theme(panel.border = element_blank())
 ```
@@
 ```{r, fig.cap = "Proyección de la anomalía de altura geopotencial en el índice SAM de 700hPa -ERA5. (Comparar con Baldwin y THompson 2009 figura 8, columna izquierda)"}
-levs <- c(700, 300, 30, 3)
+main_levs <- c(700, 300, 30, 3)
 
@@
    .[, SAM := SAM/sd(SAM), by = .(dataset, PC)] %>% 
-   .[hgt[lev %in% levs & dataset == "ERA5" & lat <= -20], on = .NATURAL] %>% 
+   .[hgt[lev %in% main_levs & dataset == "ERA5" & lat <= -20], on = .NATURAL] %>% 
    .[, FitLm(hgt_a, SAM), by = .(lon, lat, lev)] %>% 
@@
    geom_qmap(~.[lat < -20]) +
-   geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
+   geom_contour_tanaka(aes(z = estimate), 
                        global.breaks = FALSE, breaks = ZeroBreaks) +
@@
    scale_y_latitude(limits = c(-90, -20)) +
-   scale_fill_divergent() +
+   scale_fill_divergent(guide = "none") +
+   scale_linetype(guide = "none") +
    coord_polar() +
@@
    setnames("hgt", "EOF") %>%
-   .[hgt[lat <= -20 & lev %in% unique(c(levs, extremes$lev))], on = .NATURAL, allow.cartesian = TRUE] %>% 
+   .[hgt[lat <= -20 & lev %in% unique(c(main_levs, extremes$lev))], on = .NATURAL, allow.cartesian = TRUE] %>% 
    .[, FitLm(hgt, EOF = EOF/sd(EOF)), by = .(lon, lat, lev, dataset, PC)] %>% 
@@
 bt09_2d %>% 
-   .[lev %in% levs] %>% 
+   .[lev %in% main_levs] %>% 
    .[dataset == "ERA5"] %>% 
@@
 
-lab_sam <-  c(full = "Completo", 
-              asym = "Asimétrico",
-              sym  = "Simétrico")
+lab_sam <-  c(full = "Full", 
+              asym = "Asymmetric",
+              sym  = "Symmetric")
 
@@
    .[dataset == "ERA5"] %>% 
-   .[lev %in% levs] %>%
+   .[lev %in% main_levs] %>%
    rm_singleton() %>%
@@
 #    .[as.numeric(hemisphere) == 1] %>%
-#    .[lev %in% levs] %>% 
+#    .[lev %in% main_levs] %>% 
 #    rm_singleton() %>%
@@
 ```{r, fig.cap = "Relación entre el EOF1 calculado acá y el índice AAO provisto por NCEP"}
-aao <- readRDS(data_path("derived"))
+aao <- readRDS(data_path("derived", "aao.Rds"))
 
@@
 indexes %>% 
-   .[dataset == "ERA5"] %>%
-   .[, widyr::pairwise_cor(.SD, lev, time, estimate), by = .(dataset, term)] %>% 
+   # .[dataset == "ERA5"] %>%
+   .[, widyr::pairwise_cor(.SD, lev, time, estimate), by = .(term)] %>% 
    ggplot(aes(item1, item2)) +
@@
    coord_equal() +
-   facet_grid(dataset ~ term, labeller = labeller(term = lab_sam))
+   facet_grid(. ~ term, labeller = labeller(term = lab_sam))
 ```
@@
 vertical_sym <- indexes %>% 
-   # .[lev %in% levs] %>%
-   .[dataset == "ERA5"] %>%
+   # .[lev %in% main_levs] %>%
+   # .[dataset == "ERA5"] %>%
    # .[, estimate := estimate/sd(estimate), by = .(dataset, lev, hemisphere, term)] %>% 
-   dcast(dataset + time  + lev ~ term, value.var = "estimate") %>% 
-   .[, FitLm(full, asym, sym, se = TRUE), by = .(dataset, lev)] %>% 
+   dcast(time  + lev ~ term, value.var = "estimate") %>% 
+   .[, FitLm(full, asym, sym, se = TRUE), by = .(lev)] %>% 
    rm_intercept() %>% 
@@
      by = .(dataset, lev)] %>% 
-   # .[lev %in% levs] %>% 
+   # .[lev %in% main_levs] %>% 
    melt(id.vars = c("dataset", "lev"), variable.name = "term", value.name = "estimate2")  %>% 

### 2020-07-07 analysis/scripts/paper_figs.Rmd
paper figures

--- 2020-06-30 analysis/paper/paper.Rmd
@@
 ZeroBreaks <- AnchorBreaks(0, NULL, 0)
-
+axis_labs_smol <- theme(axis.text = element_text(size = 7))
 lev.lab <- AddSuffix(" hPa")
@@
   .[lev %in% c(30, 700) & term != "full"] %>% 
+  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
   ggplot(aes(time, estimate_norm)) +
   geom_line(aes(color = term)) +
-  # geom_smooth(method = "lm", aes(color = term)) +
   scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
@@
   scale_x_datetime(NULL) +
-  facet_grid(lev~., labeller = labeller(lev = lev.lab))
+  facet_grid(lev~., labeller = labeller(lev = lev.lab), scales = "free")
 ```
 
-```{r asymsam-timeseries-seasonally, fig.cap = "Seasonal time series for the asymmetric SAM and symmetric SAM."}
+```{r lag-correlation, fig.cap = "Lag-correlation between Symmetric and Asymmetric SAM at each level."}
+ccf_dt <- function(..., plot = FALSE) {
+  stat <- ccf(..., plot = plot)
+  list(lag = stat$lag[, 1, 1], 
+       acf = stat$acf[, 1, 1])
+}
+
+indexes_wide %>% 
+  .[, ccf_dt(asym, sym, lag.max = 5), by = .(lev)] %>% 
+  ggplot(aes(lag, lev)) +
+  geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
+  geom_vline(xintercept = 0) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.05)) +
+  scale_x_continuous("\u2190 Asym leads Sym\t \t\t Sym leads Asym \U2192", breaks = seq(-5, 5, 1), 
+                     expand = c(0, 0)) +
+  scale_y_level()
+```
+
+```{r}
+my_spectrum <- function(data, spans = NULL, R = 1000, ..., probs = 0.95) {
+  mtm <- spec.pgram(data, spans = spans, ..., plot = FALSE)
+
+  out <- as.data.table(mtm[c("freq", "spec")])
+
+  out[, boot_null := null_spec(data, spans = spans, R = R, ..., probs = probs)]
+
+  return(out[])
+}
+
+null_spec <- memoise::memoise(function(data, spans, R = 1000, ..., probs = 0.95) {
+
+  b <- boot::boot(data, function(d, i) spec.pgram(d[i], spans = spans, 
+                                                  ...,
+                                                  plot = FALSE)$spec, 
+                  R = R)
+
+  apply(b$t, 2, quantile, probs = probs)
+})
+
 indexes %>% 
-  .[lev %in% c(30, 700) & term != "full"] %>% 
-  .[, .(estimate_norm = mean(estimate_norm)), by = .(term, lev, time = seasonally(time))] %>% 
-  .[, season := season(time)] %>% 
-  ggplot(aes(time, estimate_norm)) +
+  copy() %>% 
+  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
+  .[, my_spectrum(estimate_norm, 5), by = .(term, lev)] -> spec
+```
+
+```{r}
+spec %>%
+  .[lev %in% c(30, 700)] %>% 
+  ggplot(aes(1/freq/12, spec)) +
+  # geom_ribbon(aes(group = interaction(term, rleid(sign(spec - boot_null))), 
+  #                 fill = sign(spec - boot_null),
+  #                 ymin = spec, ymax = boot_null)) +
     geom_line(aes(color = term)) +
-  # geom_smooth(method = "lm", aes(color = term)) +
-  scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
-  scale_y_continuous(NULL) +
-  scale_x_datetime(NULL) +
-  facet_grid(season~lev, labeller = labeller(lev = lev.lab))
+  geom_line(aes(color = term, y = boot_null)) +
+  annotation_logticks(sides = "b") +
+  scale_color_brewer(palette = "Set1") +
+  scale_fill_divergent() +
+  scale_x_log10() +
+  facet_grid(lev ~ .,labeller = labeller(lev = lev.lab))
+```
+
+```{r trends, fig.cap = "Trends (in standard deviations per decade) for each index at each level. Shading indicates the 95% confidence interval."}
+indexes %>%
+  copy() %>% 
+  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
+  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
+  rm_intercept() %>% 
+  .[, term := NULL] %>% 
+  .[, estimate := estimate*3600*24*365*10] %>% 
+  .[, std.error := std.error*3600*24*365*10] %>% 
+  setnames("type", "term") %>% 
+  ggplot(aes(lev, estimate)) +
+  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96, fill = term), alpha = 0.3) +
+  geom_line(aes(color = term)) + 
+  geom_hline(yintercept = 0) +
+  scale_x_level() +
+  coord_flip() 
 ```
@@
 ```{r cross-correlation, fig.cap = "Cross correlation between levels of the Full, Asymmetric and Symmetric SAM."}
+
+tri <- matrix(c(1000, 1000, 1000, 1, 1, 1, 1000, 1000), ncol = 2, byrow = TRUE) %>% 
+  as.data.table() %>% 
+  setnames(., colnames(.), c("item1", "item2"))
+
+logbreaks <- function(range) {
+  ticks <- ggplot2:::calc_logticks(minpow = log10(min(range)), maxpow = log10(max(range)))
+  ticks$value[ticks$end == 0.2]
+}
+
 indexes %>% 
   copy() %>% 
-  .[, widyr::pairwise_cor(.SD, lev, time, estimate), by = .(term)] %>% 
-  ggplot(aes(item1, item2)) +
-  geom_contour_fill(aes(z = correlation), na.fill = 1) +
-  scale_fill_divergent("Cross-correlation") +
-  scale_x_level() +
-  scale_y_level() +
+  .[, widyr::pairwise_cor(.SD, lev, time, estimate, diag = FALSE), by = .(term)] %>% 
+  ggplot(aes(item2, item1)) +
+  geom_contour_fill(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
+  geom_polygon(data = tri, fill = "white") +
+  scale_fill_divergent("Cross-correlation", 
+                       breaks = MakeBreaks(binwidth = 0.1),
+                       guide = guide_colorstrip_bottom()) +
+  scale_x_level(minor_breaks = logbreaks, position = "top") +
+  scale_y_level(minor_breaks = logbreaks) +
   coord_equal() +
-  facet_grid(. ~ term, labeller = labeller(term = lab_sam))
+  facet_grid(. ~ term, labeller = labeller(term = lab_sam)) +
+  theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside")
 ```
@@
 ```{r compute-regr-pattern}
-indexes_2d <- hgt[lev %in% main_levs] %>% 
+indexes_2d_yearly <- hgt[lev %in% main_levs] %>% 
   indexes_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
@@
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef."}
-indexes_2d %>% 
+indexes_2d_yearly %>% 
   .[lat <= -20] %>% 
@@
   coord_polar() +
-  facet_grid(term ~ lev, labeller = labeller(term = lab_sam, lev = lev.lab))
+  facet_grid(term ~ lev, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  axis_labs_smol 
 ```
@@
   geom_qmap(~.x[lat <= -20]) +
-  # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
   scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20)) +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
-                       limits = c(-80, 80),
+                       limits = c(-55, 55),
                        oob = scales::squish,
-                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
+                       breaks = AnchorBreaks(0, binwidth = 5, 0)) +
   scale_linetype(guide = "none") +
   coord_polar() +
-  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab))
+  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  axis_labs_smol
+```
+
+```{r wave-amplitude-700, fig.cap = "Planteray wave amplitude for the regression patterns at 700 hPa."}
+indexes_2d %>% 
+  .[ lev == 700 & lat < -20] %>%
+  .[, FitWave(estimate, 0:3), by = .(lat, lev, season, term)] %>% 
+  ggplot(aes(lat, amplitude)) +
+  geom_line(aes(color = factor(k))) +
+  scale_color_brewer("Wave-number", palette = "Set1") +
+  scale_y_continuous("Amplitude") +
+  scale_x_latitude(ticks = 15) +
+  coord_flip() +
+  facet_grid(term~season, labeller = labeller(term = lab_sam), scales = "free_x")
 ```
@@
   geom_qmap(~.x[lat <= -20]) +
-  # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
   scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20)) +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
-                       limits = c(-80, 80),
+                       limits = c(-105, 105),
                        oob = scales::squish,
-                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
+                       breaks = AnchorBreaks(0, binwidth = 20, 0)) +
   scale_linetype(guide = "none") +
   coord_polar() +
-  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab))
+  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  axis_labs_smol
 ```
 
-```{r wave-amplitude-700, fig.cap = "Planteray wave squared amplitudes for the regression patterns at 700 hPa."}
+```{r wave-amplitude-30, fig.cap = "Planteray wave amplitude for the regression patterns at 30 hPa."}
 indexes_2d %>% 
-  .[ lev == 700 & lat < -20] %>%
-  .[, FitWave(estimate, 1:3), by = .(lat, lev, season, term)] %>% 
-  ggplot(aes(lat, amplitude^2)) +
+  .[ lev == 30 & lat < -20] %>%
+  .[, FitWave(estimate, 0:3), by = .(lat, lev, season, term)] %>% 
+  ggplot(aes(lat, amplitude)) +
   geom_line(aes(color = factor(k))) +
@@
   coord_flip() +
-  facet_grid(term~season, labeller = labeller(term = lab_sam)) 
-```
-
-```{r wave-amplitude-30, fig.cap = "Planteray wave squared amplitudes for the regression patterns at 30 hPa."}
-indexes_2d %>% 
-  .[ lev == 30 & lat < -20] %>%
-  .[, FitWave(estimate, 1:3), by = .(lat, lev, season, term)] %>% 
-  ggplot(aes(lat, amplitude^2)) +
-  geom_line(aes(color = factor(k))) +
-  scale_color_brewer("Wave-number", palette = "Set1") +
-  scale_y_continuous("Squared amplitude") +
-  scale_x_latitude(ticks = 15) +
-  coord_flip() +
-  facet_grid(term~season, labeller = labeller(term = lab_sam)) 
+  facet_grid(term~season, labeller = labeller(term = lab_sam), scales = "free_x") 
 ```
@@
 ```{r compute-pp-regr} 
-pp_regr <- CMAP() %>% 
-  ReadNetCDF() %>% 
+pp <- CMAP() %>% 
+  ReadNetCDF(subset = list(lat = c(-90, 20))) %>% 
   normalise_coords() %>% 
-  .[, precip := Anomaly(precip), by = .(lon, lat)] %>% 
+  .[, precip := Anomaly(precip), by = .(lon, lat)]
+
+pp_regr_season <- pp %>% 
+  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
   .[indexes_wide[lev == 700], on = .NATURAL] %>% 
-  .[, FitLm(precip, asym, sym, se = TRUE, r2 = TRUE), by = .(lon, lat)] %>% 
+  rm_singleton() %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
+  .[, nas := mean(is.na(precip)), by = .(lon, lat, season(time))] %>% 
+  .[nas >= 0.15, precip := NA] %>% 
+  .[, rbind(as.data.table(FitLm(precip, asym, sym, se = TRUE)),
+            as.data.table(FitLm(precip, full, se = TRUE))),
+    by = .(lon, lat, season(time))] %>% 
   rm_intercept() %>% 
-  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term)]
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)]
+
 ```
 
-```{r regr-pp, fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
-pp_regr %>% 
+```{r pp-regr-global}
+pp_regr_season %>%
+  # filter_sa() %>% 
+  # .[, lon := ConvertLongitude(lon)] %>% 
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), na.fill = TRUE, breaks = AnchorBreaks(0, 0.1, 0)) +
+  # geom_raster(aes(fill = estimate)) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.25, 0)) +
   geom_contour2(aes(z = p.value), breaks = 0.01) +
@@
   stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
-  geom_qmap() +
+  geom_qmap(~.x[lat <= -20]) +
+  scale_x_longitude() +
+  scale_y_latitude(labels = NULL) +
+  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
+                       breaks = AnchorBreaks(0, 0.25, 0)) +
+  coord_polar() +
+  facet_grid(season~term, labeller = labeller(term = lab_sam)) +
+  axis_labs_smol
+```
+
+```{r regr-pp-season, fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
+
+is.sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
+  data[, lon %between% lons & lat %between% lats]
+}
+
+filter_sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
+  data[is.sa(data, lons = lons, lats = lats)]
+}
+
+pp_regr_season %>% 
+  filter_sa() %>% 
+  .[, lon := ConvertLongitude(lon)] %>% 
+  ggplot(aes(lon, lat)) +
+  # geom_raster(aes(fill = estimate)) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.25, 0)) +
+  geom_contour2(aes(z = p.value), breaks = 0.01) +
+  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
+  stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
+  geom_qmap(~filter_sa(.x)[, lon := ConvertLongitude(lon)], keep = 0.02) +
   scale_x_longitude() +
   scale_y_latitude() +
-  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
-  coord_quickmap() +
-  facet_wrap(~term, labeller = labeller(term = lab_sam))
+  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
+                       breaks = AnchorBreaks(0, 0.25, 0)) +
+  # coord_polar() +
+  ggalt::coord_proj("+proj=moll +lon_0=-60") +
+  facet_grid(season~term, labeller = labeller(term = lab_sam)) +
+  axis_labs_smol
 ```
 
-```{r}
+```{r compute-air-regr}
 time_subset <- list(time = range(indexes$time))
 air <- NOAATemp() %>% 
-  ReadNetCDF(vars = "air", subset = time_subset) %>% 
+  ReadNetCDF(vars = "air", subset = c(time_subset, list(lat = c(-90, 10)))) %>% 
   normalise_coords()
@@
   rm_intercept() 
+```
 
-air_regr_season %>% 
+```{r regr-air-season, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas."}
+air_regr %>% 
   copy() %>% 
@@
   stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat)), size = 0.4, alpha = 0.8) +
-  stat_subset(aes(subset = is.na(estimate)), geom = "raster", fill = "gray50") +
-  geom_qmap() +
+  # stat_subset(aes(subset = is.na(estimate)), geom = "raster", fill = "gray50") +
+  geom_qmap(~.x[lat <= 10]) +
   scale_x_longitude() +
-  scale_y_latitude() +
+  scale_y_latitude(labels = NULL) +
   scale_fill_divergent("Surface temperature",  limits = c(-0.8, .8), oob = scales::squish,
                        guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.15, 0)) +
-  coord_quickmap() +
-  facet_grid(season~term, labeller = labeller(term = lab_sam))
+  coord_polar() +
+  facet_grid(season~term, labeller = labeller(term = lab_sam)) +
+  axis_labs_smol
 ```

### 2020-07-07 R/theme.R

@@ -7,7 +7,7 @@
 #'
 #' @export
 theme_asymsam <- function(base_size = 16) {
-  ggplot2::theme_minimal(base_size = 16) +
+  ggplot2::theme_minimal(base_size = base_size) +
     ggplot2::theme(
       # strip.background = ggplot2::element_rect(fill = NA, color = "gray30"),
       # text = element_text(family = font_rc),

### 2020-07-07 analysis/paper/paper.Rmd
progress

@@
 ---
-twocol: TRUE  #  Use FALSE for submission. 
+twocol: FALSE  #  Use FALSE for submission. 
 title: Title here
@@
 statement: |
-  This is significant becasue... 
-bibliography: references.bib
+  This is significant becasue I wrote it.
+bibliography: AsymSAM.bib
 output: 
@@
     latex_engine: "xelatex"
+header-includes: 
+ - \usepackage{gensymb}
 ---
@@
   comment = "#>",
-  fig.path = "",
-  cache.extra = 2,
+  cache.path = "../cache/",
+  cache.extra = 3,
   dpi = 300
@@
 library(asymsam)
+library(latex2exp)
 
-theme_set(theme_asymsam(base_size = 9))
+theme_set(theme_asymsam(base_size = 10))
 ZeroBreaks <- AnchorBreaks(0, NULL, 0)
-axis_labs_smol <- theme(axis.text = element_text(size = 7))
+axis_labs_smol <- theme(axis.text = element_text(size = 6))
 lev.lab <- AddSuffix(" hPa")
@@
 
-blabalblba introdocction. SAM... yada yada circulation.. 
-`r paste(stringi::stri_rand_lipsum(3), collapse = "\n\n")`
+yada yada SAM yada yada circulation.. yada yada so important. yada yada many impacts.
 
@@
     by = .(lev)]
+
+dims <- hgt[lat <= -20] %>% 
+  .[, .(x = uniqueN(lon), y = uniqueN(lat), t = uniqueN(time))]
 ```
@@
 
-We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach].
+We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach] for the period 1979 to 2018 (inclusive).
 
 Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0 degree latitude x 5.0 degree longitude global grid [@vose2012; @smith2008].
+The same analysis was carried out using CRUTEM4 [@osborn2014] (not shown).
 
@@
 
-`r paste(stringi::stri_rand_lipsum(4, FALSE), collapse = "\n\n")`
+We defined the Southern Annular Mode (SAM) as the leading EOF of the monthly anomalies of geopotential field at 700 hPa south of 20\degree S (citation?).
+The EOF was performed by computing the Singular Value Decomposition of the data matrix consisting in `r dims$t` rows and  `r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points of latitude).
+The values where weighted by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999].
+This same method was used at the rest of the levels considered in this paper.
+
+To separate between the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern.
+The results are shown in Figure\ \ref{fig:method} for 700hPa.
+The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
+We then compute the "Full", "Asymmetric" and "Symmetric" indexes, by regressing each geopotential field on these patterns (weighting by the cosine of latitude).
+
+The three indexes are normalised by dividing them by the standard deviation of the "Full" index at each level.
+This means that comparing the magnitude between indexes is meaningful, but it also means that not every index will have unit standard deviation.
 
@@
 
+lab_eof <- c(full = TeX("$\\mathrm{EOF_1}(\\lambda, \\phi)$"),
+             asym = TeX("$\\mathrm{EOF_1}^*(\\lambda, \\phi)$"), 
+             sym  = TeX("$\\[\\mathrm{EOF_1}\\](\\lambda, \\phi)$"))
+
 sam_sep %>% 
@@
   .[variable == "sym", value := value + rnorm(.N, 0, sd = 1e-6)] %>%
-  .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)] %>% 
+  .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)] -> sam_plot
+
+levels(sam_plot$variable) <- lab_eof
+
+sam_plot %>% 
   periodic(lon = c(0, 360)) %>%
@@
   coord_polar() +
-  facet_grid(~variable, labeller = labeller(variable = lab_sam, lev = lev.lab)) +
+  facet_grid(~variable, labeller = labeller(variable = label_parsed, lev = lev.lab)) +
   axis_labs_smol
 ```
+
+### Significance
+
+We adjusted p-values for False Detection Rate following @wilks2016.
 
@@
 
-`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
-
 ```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM.", fig.width=6, fig.height=3, fig.env="figure*"}
@@
 
-`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
-
-`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
-
-`r paste(stringi::stri_rand_lipsum(1, FALSE), collapse = "\n\n")`
-
 ```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level.", fig.width=3, fig.height=3.5}
@@
 ```
-
-`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
 
@@
 
-`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
-
 ```{r wave-amplitude-700, fig.cap = "Planteray wave amplitude for the regression patterns at 700 hPa.", fig.width=6, fig.height=4}
@@
 ```
-
-`r paste(stringi::stri_rand_lipsum(1, FALSE), collapse = "\n\n")`
 
@@
 
-`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
-
 ```{r wave-amplitude-30, fig.cap = "Planteray wave amplitude for the regression patterns at 30 hPa.", fig.width=6, fig.height=4}
@@
 ```
-`r paste(stringi::stri_rand_lipsum(1, FALSE), collapse = "\n\n")`
 
@@
 
-```{r pp-regr-global, fig.cap = "caption", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
+```{r pp-regr-global,  fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
 pp_regr_season %>%
@@
 
+\bibliography{AsymSAM}
+
 \appendix
@@
 
-```{r lag-correlation, fig.cap = "Lag-correlation between Symmetric and Asymmetric SAM at each level.", fig.width = 4, fig.height=4, appendix = TRUE}
+```{r A1, fig.cap = "Lag-correlation between Symmetric and Asymmetric SAM at each level.", fig.width = 4, fig.height=4, appendix = TRUE}
 ccf_dt <- function(..., plot = FALSE) {
@@
 
-```{r spectrum, fig.cap = "Foutier spectrum", appendix = TRUE}
+```{r A2, fig.cap = "Foutier spectrum", appendix = TRUE}
 spec %>%
@@
 
-```{r trends, fig.cap = "Trends for each index at each level. Shading indicates the 95\\% confidence interval.", fig.width = 3, fig.height=3.5, appendix = TRUE}
+```{r A3, fig.cap = "Trends for each index at each level. Shading indicates the 95\\% confidence interval.", fig.width = 3, fig.height=3.5, appendix = TRUE}
 indexes %>%

### 2020-07-07 analysis/paper/paper.tex
progress

@@
 % DOUBLE-SPACED VERSION FOR SUBMISSION TO THE AMS
-\documentclass[twocol]{ametsocV5}
+\documentclass[]{ametsocV5}
 
@@
 \usepackage{newtxmath}
+
+\usepackage{gensymb}
 
@@
 \statement
-This is significant becasue\ldots{}
+This is significant becasue I wrote it.
 
@@
 
-blabalblba introdocction. SAM\ldots{} yada yada circulation.. Lorem
-ipsum dolor sit amet, hac curae turpis sollicitudin vulputate quisque
-sapien nullam curae integer. Torquent porttitor et et vulputate blandit
-velit sed sed at velit eget arcu. Senectus, taciti justo eu sem purus
-vestibulum pellentesque quis ac. Ac facilisi, amet consectetur mauris
-fringilla sit. Est fusce, augue orci dictumst lacinia et elementum
-dapibus orci eu, luctus ut luctus. Phasellus magna ut felis vitae metus
-non non. In maecenas taciti. Hendrerit bibendum placerat. Molestie est
-enim ante amet natoque consequat hendrerit, pharetra ridiculus.
-Ultricies mauris donec leo eu, parturient praesent. Ridiculus sociis
-ligula, ultricies augue, sagittis, turpis et. A netus at sagittis enim
-nisl aliquet sed.
-
-Justo id ut dolor faucibus ac justo orci volutpat vivamus nisl. Eu
-sollicitudin, semper sed vitae et praesent vitae diam venenatis. Felis
-vitae, velit risus id donec et. Neque nec id. Tellus nisl, vivamus sed
-faucibus, non, per scelerisque ac arcu eget mus himenaeos imperdiet.
-Purus eu quis felis mauris, ultricies et luctus integer orci. Lobortis
-eget nec cras egestas feugiat. At mattis eget mus aliquam pellentesque
-mauris elit sollicitudin tempor. Est mauris sed convallis at purus.
-
-Purus scelerisque condimentum blandit interdum pharetra commodo. Neque
-sed tristique natoque condimentum quis mollis et tortor egestas augue
-sem est neque. Tortor dictum, pretium sociis auctor odio. Quam velit,
-sapien, sed, tempor quis ut, ac auctor orci eleifend. Venenatis ut in
-euismod nec nunc velit sed imperdiet. Lectus bibendum enim morbi dui sem
-ut pellentesque venenatis diam. Ut, tincidunt senectus vel sed conubia
-dapibus feugiat purus. Dui donec eu lacus semper tortor eget. Magna
-proin amet justo. A lacus est ornare auctor sed vel et. Duis, amet
-parturient nisl tempor praesent libero elementum erat ac et.
+yada yada SAM yada yada circulation.. yada yada so important. yada yada
+many impacts.
 
@@
 We used monthly geopotential height at 2.5 longitude by 2.5 latitude
-resolution from ERA5 \citep{hersbach}.
+resolution from ERA5 \citep{hersbach} for the period 1979 to 2018
+(inclusive).
 
@@
 degree latitude x 5.0 degree longitude global grid
-\citep{vose2012, smith2008}.
+\citep{vose2012, smith2008}. The same analysis was carried out using
+CRUTEM4 \citep{osborn2014} (not shown).
 
@@
 
-Sagittis, malesuada, lacus integer vehicula. Diam curae, tempus feugiat
-vivamus nascetur augue eu, vitae non arcu. Vivamus aptent viverra
-blandit gravida consectetur eleifend habitasse? Aliquam ante. Venenatis
-mauris parturient per, per et ut cum. Dapibus nulla tempor. Tempor
-libero cubilia enim vestibulum. Et, magna elementum in conubia ut. Dis
-non eleifend consectetur a et. Nulla libero nunc varius porta, eros eros
-pretium platea malesuada velit. Dolor, suspendisse, suspendisse magna et
-quis magnis ultricies! Luctus vestibulum arcu, neque sapien.
+We defined the Southern Annular Mode (SAM) as the leading EOF of the
+monthly anomalies of geopotential field at 700 hPa south of 20\degree S
+(citation?). The EOF was performed by computing the Singular Value
+Decomposition of the data matrix consisting in 481 rows and 4176 columns
+(144 points of longitude and 29 points of latitude). The values where
+weighted by the square root of the cosine of latitude to account for the
+non-equal area of each gridpoint \citep{chung1999}. This same method was
+used at the rest of the levels considered in this paper.
 
-Nisl quis sed, fringilla laoreet felis. Felis ac vel curabitur himenaeos
-pretium. Et enim sed vitae vitae nec ac dapibus aliquet. Massa libero
-non egestas, tempus tempor commodo amet, suspendisse, orci. Nostra, orci
-penatibus diam aliquet neque sed. Eu fusce, rutrum at ut in, curae. Ex
-cum massa efficitur augue et nulla consequat ornare in. Elementum velit,
-non aliquet ut. Ultricies ut arcu diam eros scelerisque lobortis semper.
-In semper ligula varius natoque purus in commodo. Dis tellus elit quam
-egestas maximus. Curae aliquam dolor platea, penatibus sed. Dui habitant
-aenean habitasse nulla dui amet.
+To separate between the zonally symmetric and asymmetric components of
+the SAM, we computed the zonal mean and anomalies of the full SAM
+spatial pattern. The results are shown in Figure~\ref{fig:method} for
+700hPa. The full spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is
+the sum of the zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\))
+and symmetric (\([\mathrm{EOF_1}](\lambda, \phi)\)) components. We then
+compute the ``Full'', ``Asymmetric'' and ``Symmetric'' indexes, by
+regressing each geopotential field on these patterns (weighting by the
+cosine of latitude).
 
-Vehicula sed ligula nunc fringilla. Ornare morbi, eu, non ac, et ipsum
-phasellus. Felis non tincidunt litora. Hendrerit elit, nisi nam sem
-auctor eu ultrices amet. Per accumsan pharetra sapien lacinia molestie,
-suspendisse. In vehicula elementum vestibulum vestibulum ut taciti, erat
-consequat sed. Non non et augue volutpat non ante congue quam. Risus mi
-nec laoreet viverra non. Ligula platea porta nisi faucibus, vestibulum
-lacus. Purus nec imperdiet sed sodales sed sed in sed mauris. Quam
-semper suscipit nunc felis sed cras amet ultrices maecenas. Nullam
-luctus magnis amet non id at, tempus, ullamcorper. Id interdum tortor
-sed viverra imperdiet porttitor et phasellus. Ac facilisis mus.
-
-Nec fusce, ipsum a porttitor, magnis in vestibulum. Eu cras mi,
-inceptos. Nec class ad fames litora nisl pellentesque vestibulum. Nullam
-fames amet diam aliquam eget. Donec, non mollis aptent urna condimentum
-tellus, torquent. Sit nunc lectus interdum. Maximus eu mattis libero
-pellentesque vel suspendisse ullamcorper. Auctor dolor non sociosqu et
-porttitor pulvinar, turpis gravida in auctor ipsum cursus. Nulla
-venenatis fringilla sollicitudin, accumsan libero nulla aliquet volutpat
-mauris lectus at auctor! Parturient nisl, metus etiam eu aptent purus,
-ac vestibulum placerat sed, a quam tellus. Vulputate placerat euismod.
-Commodo urna, neque penatibus? Pharetra elementum curae, sed.
+The three indexes are normalised by dividing them by the standard
+deviation of the ``Full'' index at each level. This means that comparing
+the magnitude between indexes is meaningful, but it also means that not
+every index will have unit standard deviation.
 
@@
 
+\subsubsection{Significance}
+
+We adjusted p-values for False Detection Rate following
+\citet{wilks2016}.
+
 \section{Results}
-
-Dignissim nibh sed eu, sed dapibus, tempor. Libero ligula accumsan enim
-fringilla dapibus hendrerit integer quisque sapien. Nam convallis eu,
-magnis. Mauris eu at bibendum, elit mus. Vehicula posuere amet viverra
-volutpat nec fringilla orci netus. Malesuada imperdiet leo iaculis at
-varius massa a sit leo ad odio cum. Justo orci quis in ante convallis.
-Sed sit justo cras in, mus ac, elementum? Rhoncus fermentum, at risus ut
-tincidunt nec, sit per, lacus. Eros inceptos posuere facilisis, sed,
-magna suspendisse non sed sapien porttitor. Sit, augue facilisis sed. Et
-senectus odio, ipsum fringilla et, iaculis duis. Erat ac integer
-consectetur cubilia. Condimentum neque suscipit vel aliquet egestas
-pellentesque.
-
-Fringilla non dis nisl erat ut ligula. Eros egestas ut ut congue
-lobortis sem? Aliquet nunc diam tortor. Semper conubia consequat nec ex
-nibh laoreet. Sed malesuada luctus a viverra netus congue a diam ut
-lorem. Vestibulum, nec varius nisl non ornare morbi. Sed tellus,
-ultricies non ac. Ridiculus, cubilia dui sit ac massa suspendisse
-maximus. Penatibus, taciti vel sed. Ut pellentesque vestibulum sed
-consequat ultricies! Nisl mi tristique vulputate vitae tristique
-consequat. At accumsan class est mattis. Interdum sociosqu lobortis
-dapibus nullam fermentum himenaeos.
 
@@
 
-Euismod non magnis, ligula sed eu velit enim. Laoreet nulla vel vivamus
-et ultrices accumsan tempor fames at sed. Eleifend sociis tristique nibh
-nunc tempor. Nostra sed iaculis, ut vehicula eu nulla sed venenatis.
-Pharetra risus, orci erat convallis sit sociis sed tortor. Amet, mauris
-eros magnis praesent aptent auctor scelerisque. Faucibus in tempus in
-lobortis semper, tempus, in. Amet volutpat class tristique sapien sociis
-habitant posuere et auctor. Nulla litora eu hac neque efficitur risus
-donec. Nam dui fermentum mollis libero sapien, taciti et vulputate, in.
-Eu vestibulum felis suscipit gravida aenean ligula. Cras eu nunc
-conubia, sed nibh inceptos id accumsan. Lacus, eu ante aliquet.
-
-Pulvinar magna phasellus amet per vel lacus ante lectus. Integer
-vestibulum hac sodales nibh ac donec augue maximus. Ullamcorper vivamus
-finibus enim quis. Fringilla nibh sed pellentesque adipiscing facilisis
-sed. Pretium interdum nibh amet nunc quam velit imperdiet. In sed in
-lobortis et conubia. Praesent sed pretium auctor. Sit eleifend laoreet
-sed erat a ultrices cursus hendrerit nisl. Tellus habitasse aliquet
-molestie pretium faucibus. Fames eu convallis penatibus morbi tincidunt
-vestibulum, ligula leo. Pharetra facilisis ac mi. Dapibus congue eu eros
-accumsan dictum suscipit in diam donec in interdum id. Dictumst ac id
-amet iaculis ut aptent.
-
-Netus quam. Euismod senectus donec pharetra felis eget. Penatibus eu in
-varius morbi interdum erat. Sed felis finibus mauris interdum. Ut,
-euismod, a lacus justo nulla porta faucibus ac nec, odio dignissim orci.
-Et, porta, blandit tincidunt. Ullamcorper in egestas scelerisque luctus
-accumsan mauris nam dis diam. Sollicitudin quis vestibulum cum fames
-maximus odio aliquam sed. Aliquam ornare ut malesuada ut eu eros.
-
-Leo maecenas aptent aenean diam, arcu. Pulvinar sociis vulputate eu non
-amet maecenas posuere. Commodo leo phasellus. Eu ut suscipit nibh in sed
-conubia fermentum sit tristique. Urna morbi consequat dui sed lorem
-ridiculus eros, odio non neque? Ante dolor eu id hac. In risus dolor
-nunc urna arcu. Consectetur sociis platea curabitur nulla, platea in eu.
-
-Ut pharetra amet nam augue imperdiet. Eros eget orci cubilia rutrum
-fusce eu in in. Magnis tincidunt orci. In tellus nisl suspendisse ex
-ante mauris nam. A ac sed leo neque taciti tincidunt diam laoreet.
-Semper odio. Rhoncus habitant sed sed mollis ultricies vehicula ipsum
-urna ac nunc nulla nunc? Magna non mauris, montes ullamcorper. Dolor sed
-maximus dui ac porttitor at cubilia conubia gravida ut.
-
 \begin{figure}
@@
 
-Blandit rutrum, nisi vitae, a velit sodales eu. Vestibulum non, dolor
-in, sit. Libero nullam sociosqu per tellus non nunc quam fusce nisl. Sed
-metus etiam faucibus amet eget velit aliquam. Eget eget nibh commodo.
-Consectetur tristique netus ac arcu. Ipsum ullamcorper nec euismod velit
-quis lacinia mus. Enim lectus est in sed eu efficitur himenaeos auctor
-nec. Justo libero montes pellentesque augue. Tempus laoreet pellentesque
-sociosqu tincidunt varius ac eleifend, non vitae. Mattis, eleifend
-sociis, nascetur eget sed aenean, nec.
-
-Egestas finibus gravida luctus per facilisi. Ornare quam, erat amet
-maximus nisi posuere potenti consequat ante eu arcu ac? Fermentum sociis
-class justo accumsan sapien penatibus consequat, ut ex at dis, etiam.
-Sem parturient facilisis montes feugiat. Et urna velit praesent vehicula
-quisque magna in. Tellus, at sem eros fusce ultrices pretium inceptos
-dui, tristique. Mi feugiat volutpat ut a condimentum ultricies, purus
-pellentesque in. Morbi ut mauris conubia tellus mi, quis. Nulla
-fermentum ad ut dui sem ultrices ut, ante. Per enim vestibulum sit
-congue natoque egestas nec erat vehicula aenean bibendum. Non ultricies
-ultrices nibh accumsan laoreet eros nunc nisi pulvinar sit sed.
-Pellentesque turpis adipiscing efficitur aptent tincidunt facilisi mi,
-quisque. Mi, pulvinar at non, phasellus.
-
 \begin{figure*}
@@
 \end{figure*}
-
-Consequat porttitor eget diam fusce phasellus a orci natoque in. Fusce
-orci convallis vel. Eu aliquet ligula risus nunc lacinia sem ut. At
-morbi ipsum nam senectus, odio. Amet et consequat erat. Tincidunt, ad
-risus sed! Interdum eget enim nascetur sed eleifend nisl cum vivamus.
-Placerat natoque eu risus ut nec, tellus tortor, donec. Quam litora
-pulvinar ac sem, tortor. Sodales ligula praesent interdum erat enim
-lectus donec dolor ligula donec. Et ipsum dis. Laoreet, id auctor
-efficitur elementum vitae commodo posuere lorem sem lacus et, platea
-sapien. Ligula primis inceptos accumsan, torquent. Elementum laoreet
-vitae nulla aliquam in magna condimentum ex blandit proin.
-
-Tincidunt tristique ante nulla sit a semper urna auctor. Ridiculus, nibh
-sagittis. Ullamcorper bibendum in conubia consectetur montes, viverra.
-Vivamus non diam. Vel efficitur interdum sed suscipit magna sed massa
-magnis nunc. Taciti inceptos ipsum massa nunc. Ac, tellus curabitur amet
-congue hac metus rhoncus. Posuere tortor rhoncus, nec ultrices enim sit
-habitasse mattis eget platea. Integer dui at sapien tincidunt phasellus
-enim at euismod lobortis.
 
@@
 
-Nisi porttitor et nibh consectetur porttitor a dapibus blandit faucibus.
-Non dis quis donec, laoreet. Suspendisse primis, aenean justo ex nullam
-ex cursus. Nec pulvinar dictumst sed. Conubia, nulla sit, et mi
-malesuada at ex sit dui est. Mauris quisque lectus facilisis parturient
-dictumst ultrices. Non a non, in, vestibulum, scelerisque ligula mauris,
-dictum? Urna eget id litora in. Platea ac turpis ultrices enim sagittis
-et purus? Sagittis sollicitudin est purus mauris ultricies nisi aliquet
-ante, nam laoreet. Pellentesque nascetur adipiscing mus platea congue eu
-ac velit. In ut tellus elementum neque ridiculus facilisis.
+Sem tortor mus non tristique augue. Donec vel feugiat sit aliquet lorem
+dui lobortis dolor risus. Proin, ac eu, mauris. Dolor id habitasse,
+curabitur platea ante. Rhoncus volutpat himenaeos vestibulum ut bibendum
+felis non. Ligula placerat ac luctus ad. Ante, lorem volutpat eu dapibus
+pharetra arcu. Sollicitudin eu tortor amet. Maecenas nulla euismod ac
+suspendisse metus. A ac, hendrerit, sit sollicitudin ac primis
+ullamcorper tempor id, mauris. Sem diam, sem curabitur. Risus, nullam
+vel diam hendrerit massa nec non. Donec amet amet ultrices ac taciti
+aliquam. Enim arcu et sed sit.
 
-Leo nec suscipit sit duis. Taciti per tempus sed vel aliquet. Amet
-feugiat. In ut ut dui per nisl ridiculus. Et urna venenatis scelerisque
-ut nec feugiat ad. Elementum vestibulum aliquet ullamcorper magna aenean
-lacus habitasse, laoreet mauris. Consectetur ornare molestie ultrices.
-Diam magnis purus et lacinia curae. Faucibus nulla a. Sociis parturient
-facilisis maximus nisi dis a. Tempus id venenatis vitae, cubilia
-ultricies, eget nulla. Quis amet per velit. Ac mattis sed, placerat
-donec ac, maximus tempor et quisque. Nulla vestibulum sed nulla eget
-eget vestibulum dictumst gravida ac viverra habitant, placerat.
+Ligula sed sociosqu imperdiet magna curabitur, et, elit lobortis ut,
+sed. Placerat id eros quis vivamus sed, in felis. Egestas phasellus at
+sagittis sed senectus malesuada. Vestibulum at pretium elementum id,
+facilisi. Porttitor luctus vel tempus diam nec per, nibh nisi. Viverra
+turpis nisl sed nec varius, et ridiculus in amet. Amet sapien nec
+convallis arcu dis et mauris dolor. Vestibulum, in, in tincidunt tellus
+risus class quisque, augue feugiat.
 
@@
 
-Ligula nunc eget lacus fermentum, nisl ut donec venenatis. Primis sed.
-Turpis vel tempor condimentum ac magnis nam, sed. Dolor blandit ornare
-sed in ut nisl. Dignissim posuere aptent. Consequat ac neque at
-himenaeos nulla. Faucibus nascetur, elit tellus. Fringilla erat sed
-sodales class, lobortis. Amet sed suscipit. In et purus ante felis
-tortor ex potenti. Sed sagittis, suspendisse aliquet sapien accumsan,
-luctus suscipit. Turpis varius ex maecenas, nascetur arcu ultrices quis,
-enim. Sed ullamcorper non, suspendisse praesent porttitor sit dui
-aliquet. Fermentum dis sit nunc tristique, ac nec, praesent hendrerit.
-
-Fermentum at odio eu ultrices amet, vel ridiculus integer. Mollis, sed
-habitasse ut, mollis at id donec porttitor suspendisse? Sagittis magna
-sed mauris mi euismod varius sapien pretium pharetra. Ex ac sed sapien,
-orci. Sed sed feugiat eget justo a euismod. Augue magna nullam justo id
-molestie egestas placerat. Ornare maximus in sed ridiculus, praesent.
-Sed, nec blandit. Nascetur pellentesque ligula tellus congue.
-
 \begin{figure}
@@
 \end{figure}
-
-Velit ex et tortor dapibus pharetra convallis. A sed ridiculus.
-Efficitur ut ad, ligula nisl, mi in cursus tellus quis dictum malesuada
-rutrum elit venenatis. Ac et. Cum lorem fames ut. Ultricies sed a nunc
-mi ligula. At ut maecenas luctus sociosqu laoreet. Duis netus quam
-varius mi convallis nunc a. Non lacus sit nullam odio porttitor
-vestibulum. Cum donec ligula per sed, elit. Quis nibh libero cubilia
-viverra feugiat justo dapibus sit. Dictumst montes ut sit! Sed nunc
-auctor nullam porttitor hendrerit ornare egestas. Vivamus ac sit, hac
-cum ut, a proin, quis, non, pulvinar. Purus quam suscipit amet sed nec
-feugiat dictumst elit vestibulum habitasse mauris.
 
@@
 
-Mus, bibendum viverra non ante velit auctor in, interdum turpis. Aliquet
-egestas, quam porta a eu sed ac donec sed elementum sit. Ligula donec
-nec facilisis efficitur finibus, egestas porta sed, nisi nunc, elit
-pretium. Orci vestibulum dui vitae amet nulla, sagittis tincidunt sem
-sit, amet. Mi risus penatibus, magna, risus turpis, ut posuere praesent
-a fames pulvinar lacus. Sit maecenas leo vestibulum urna massa luctus,
-vitae, justo eros libero ut! Condimentum a blandit tellus pellentesque
-etiam ut mauris. Sollicitudin erat et amet, tempor donec aenean arcu
-maximus leo etiam. Felis class lectus vulputate vulputate pellentesque
-praesent. Non tristique quis molestie nec maecenas vel eget, felis dui.
-Accumsan aenean vitae urna arcu nam mauris nisl. In inceptos magnis
-mollis massa habitasse interdum etiam pretium facilisis. Velit erat
-convallis etiam felis luctus sem amet posuere.
-
-Etiam, enim, magna. Ut erat, commodo sem euismod vel quam dui. Efficitur
-ridiculus id a, mattis facilisis cursus vitae, felis. In et et, est
-maximus luctus sed habitant aptent. Eget libero justo interdum aenean
-finibus mauris pulvinar et duis pulvinar. Nisl et in cum sagittis
-luctus. Libero justo ut risus cras sapien. Felis porta, lacus ipsum,
-condimentum eros integer. Nunc per lectus lorem per quisque. Tempus
-magna sodales vestibulum faucibus.
-
 \begin{figure}
@@
 
-Himenaeos a per at ut, erat felis congue. Varius pharetra eu maecenas
-non sed lectus amet egestas in. Aenean nullam nascetur mollis sit. In,
-donec consequat per ridiculus platea risus risus neque. Facilisi risus
-suspendisse imperdiet a at taciti lobortis. Fringilla egestas,
-sollicitudin natoque. Nam diam eget sed, eleifend ut faucibus lacinia
-maximus. Vestibulum aliquet curabitur sapien efficitur morbi elementum
-nam donec turpis. Egestas egestas nunc auctor nunc cursus nec semper est
-vel, felis.
-
 \begin{sidewaysfigure}
-\includegraphics{pp-regr-global-1} \caption[caption]{caption}\label{fig:pp-regr-global}
+\includegraphics{pp-regr-global-1} \caption[Regression pattern of precipiation with Asymmetric and Symmetric SAM]{Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:pp-regr-global}
 \end{sidewaysfigure}
@@
 \begin{figure}
-\includegraphics{lag-correlation-1} \appendcaption{lag-correlation}{Lag-correlation between Symmetric and Asymmetric SAM at each level.}\label{fig:lag-correlation}
+\includegraphics{A1-1} \appendcaption{A1}{Lag-correlation between Symmetric and Asymmetric SAM at each level.}\label{fig:A1}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{spectrum-1} \appendcaption{spectrum}{Foutier spectrum}\label{fig:spectrum}
+\includegraphics{A2-1} \appendcaption{A2}{Foutier spectrum}\label{fig:A2}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{trends-1} \appendcaption{trends}{Trends for each index at each level. Shading indicates the 95\% confidence interval.}\label{fig:trends}
+\includegraphics{A3-1} \appendcaption{A3}{Trends for each index at each level. Shading indicates the 95\% confidence interval.}\label{fig:A3}
 \end{figure}

### 2020-07-07 R/eof_asym.R

@@ -26,18 +26,13 @@ eof_asym <- function(value, lon, lat, time, n = 1) {
   indexes <- eof$right %>%
     data[., on = .NATURAL, allow.cartesian = TRUE]
 
-  pcor <- indexes[, partial_cor(value, sym, asym, weights = cos(lat*pi/180)),
-                  keyby = .(time, PC)]
-
   indexes[, rbind(data.table::as.data.table(metR::FitLm(value, full, weights = cos(lat*pi/180), r2 = TRUE)),
-                  data.table::as.data.table(metR::FitLm(value, sym, asym, weights = cos(lat*pi/180), r2 = TRUE))),
+                  data.table::as.data.table(metR::FitLm(value,  sym, weights = cos(lat*pi/180), r2 = TRUE)),
+                  data.table::as.data.table(metR::FitLm(value, asym, weights = cos(lat*pi/180), r2 = TRUE))),
           keyby = .(time, PC)] %>%
     .[term != "(Intercept)"] %>%
     .[, estimate_norm := estimate/stats::sd(estimate[term == "full"]), by = PC] %>%
     .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>%
-    .[term != "full", partial.r.squared := pcor$partial_correlation^2] %>%
-    .[term == "full", partial.r.squared := r.squared] %>%
-    .[, adj.r.squared := NULL] %>%
     .[]
 }
 

### 2020-07-07 analysis/scripts/02-compute-eofs.R

@@ -1,6 +1,7 @@
 library(metR)
 library(magrittr)
 library(asymsam)
+library(data.table)
 
 indexes_file <- data_path("derived", "indexes.Rds")
 

### 2020-07-08 analysis/paper/paper.Rmd
spectrum

@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(fig.path = "",     # AMS required
-                      out.extra = "",    # To force figure labels
-                      echo = FALSE)      # Don't show R code
-
-# This helps to use appendix figures.
-appendcaption <- function (x, options) {
-  `%n%` <- knitr:::`%n%`
-  rw = options$resize.width
-  rh = options$resize.height
-  resize1 = resize2 = ""
-  if (!is.null(rw) || !is.null(rh)) {
-    resize1 = sprintf("\\resizebox{%s}{%s}{", rw %n% "!", 
-                      rh %n% "!")
-    resize2 = "} "
-  }
-  tikz = knitr:::is_tikz_dev(options)
-  a = options$fig.align
-  fig.cur = options$fig.cur %n% 1L
-  fig.num = options$fig.num %n% 1L
-  animate = options$fig.show == "animate"
-  fig.ncol = options$fig.ncol %n% fig.num
-  if (is.null(fig.sep <- options$fig.sep)) {
-    fig.sep = character(fig.num)
-    if (fig.ncol < fig.num) 
-      fig.sep[seq(fig.ncol, fig.num - 1L, fig.ncol)] = "\\newline"
-  }
-  sep.cur = NULL
-  if (!tikz && animate && fig.cur < fig.num) 
-    return("")
-  usesub = length(subcap <- options$fig.subcap) && fig.num > 
-    1
-  ai = options$fig.show != "hold"
-  plot1 = ai || fig.cur <= 1L
-  plot2 = ai || fig.cur == fig.num
-  align1 = if (plot1) 
-    switch(a, left = "\n\n", center = "\n\n{\\centering ", 
-           right = "\n\n\\hfill{}", "\n")
-  align2 = if (plot2) 
-    switch(a, left = "\\hfill{}\n\n", center = "\n\n}\n\n", 
-           right = "\n\n", "")
-  cap = options$fig.cap
-  appendix = !is.null(options$appendix)
-  scap = options$fig.scap
-  fig1 = fig2 = ""
-  mcap = fig.num > 1L && options$fig.show == "asis" && !length(subcap)
-  sub1 = sub2 = ""
-  if (length(cap) && !is.na(cap)) {
-    lab = paste0(options$fig.lp, options$label)
-    if (plot1) {
-      pos = options$fig.pos
-      if (pos != "" && !grepl("^[[{]", pos)) 
-        pos = sprintf("[%s]", pos)
-      fig1 = sprintf("\\begin{%s}%s", options$fig.env, 
-                     pos)
-    }
-    if (usesub) {
-      sub1 = sprintf("\\subfloat[%s%s]{", subcap, knitr:::create_label(lab, 
-                                                               fig.cur, latex = TRUE))
-      sub2 = "}"
-      sep.cur = fig.sep[fig.cur]
-      if (is.na(sep.cur)) 
-        sep.cur = NULL
-    }
-    if (plot2) {
-      if (is.null(scap) && !grepl("[{].*?[:.;].*?[}]", 
-                                  cap)) {
-        scap = strsplit(cap, "[:.;]( |\\\\|$)")[[1L]][1L]
-      }
-      scap = if (is.null(scap) || is.na(scap)) 
-        ""
-      else sprintf("[%s]", scap)
-      if (cap == "") {
-        cap = ""
-      } else if (appendix == TRUE) {
-        cap = sprintf("\\appendcaption{%s}{%s}%s\n", options$label, cap, knitr:::create_label(lab, 
-                                                                  if (mcap) 
-                                                                    fig.cur, latex = TRUE))
-      } else {
-        cap = sprintf("\\caption%s{%s}%s\n", scap, cap, knitr:::create_label(lab, 
-                                                                  if (mcap) 
-                                                                    fig.cur, latex = TRUE))
-      }
-      fig2 = sprintf("%s\\end{%s}\n", cap, options$fig.env)
-    }
-  }
-  else if (knitr:::pandoc_to(c("latex", "beamer"))) {
-    align.env = switch(a, left = "flushleft", center = "center", 
-                       right = "flushright")
-    align1 = if (plot1) 
-      if (a == "default") 
-        "\n"
-    else sprintf("\n\n\\begin{%s}", align.env)
-    align2 = if (plot2) 
-      if (a == "default") 
-        ""
-    else sprintf("\\end{%s}\n\n", align.env)
-  }
-  ow = options$out.width
-  if (animate && identical(ow, "\\maxwidth")) 
-    ow = NULL
-  if (is.numeric(ow)) 
-    ow = paste0(ow, "px")
-  size = paste(c(sprintf("width=%s", ow), sprintf("height=%s", 
-                                                  options$out.height), options$out.extra), collapse = ",")
-  paste0(fig1, align1, sub1, resize1, if (tikz) {
-    sprintf("\\input{%s}", x)
-  }
-  else if (animate) {
-    aniopts = options$aniopts
-    aniopts = if (is.na(aniopts)) 
-      NULL
-    else gsub(";", ",", aniopts)
-    size = paste(c(size, sprintf("%s", aniopts)), collapse = ",")
-    if (nzchar(size)) 
-      size = sprintf("[%s]", size)
-    sprintf("\\animategraphics%s{%s}{%s}{%s}{%s}", size, 
-            1/options$interval, sub(sprintf("%d$", fig.num), 
-                                    "", xfun::sans_ext(x)), 1L, fig.num)
-  }
-  else {
-    if (nzchar(size)) 
-      size = sprintf("[%s]", size)
-    res = sprintf("\\includegraphics%s{%s} ", size, if (getOption("knitr.include_graphics.ext", 
-                                                                  FALSE)) 
-      x
-      else xfun::sans_ext(x))
-    lnk = options$fig.link
-    if (is.null(lnk) || is.na(lnk)) 
-      res
-    else sprintf("\\href{%s}{%s}", lnk, res)
-  }, resize2, sub2, sep.cur, align2, fig2)
-}
-
-knitr::knit_hooks$set(plot = appendcaption)
-```
-
-```{r, warning=FALSE, message = FALSE}
+```{r setup, echo = FALSE,warning=FALSE, message = FALSE}
 knitr::opts_chunk$set(
+	echo = FALSE,
+	fig.path = "",
+	message = FALSE,
+	warning = FALSE,
+	cache = TRUE,
+	cache.extra = 3,
+	cache.path = "../cache/",
 	collapse = TRUE,
-  cache = TRUE,
-  warning = FALSE,
-  message = FALSE,
-  echo = FALSE,
 	comment = "#>",
-  cache.path = "../cache/",
-  cache.extra = 3,
-  dpi = 300
+	dpi = 300,
+	out.extra = ""
 )
-
 library(metR)
@@
 
-```{r}
-my_spectrum <- function(data, spans = NULL, R = 1000, ..., probs = 0.95) {
-  mtm <- spec.pgram(data, spans = spans, ..., plot = FALSE)
-
-  out <- as.data.table(mtm[c("freq", "spec")])
-
-  out[, boot_null := null_spec(data, spans = spans, R = R, ..., probs = probs)]
-
-  return(out[])
-}
-
-null_spec <- memoise::memoise(function(data, spans, R = 1000, ..., probs = 0.95) {
-
-  b <- boot::boot(data, function(d, i) spec.pgram(d[i], spans = spans, 
-                                                  ...,
-                                                  plot = FALSE)$spec, 
-                  R = R)
-
-  apply(b$t, 2, quantile, probs = probs)
-})
-
+```{r compute-spectrums}
 indexes %>% 
-  copy() %>% 
+  .[lev %in% c(30, 700)] %>% 
   # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
-  .[, my_spectrum(estimate_norm, 5), by = .(term, lev)] -> spec
+  .[, fftspectrum(estimate_norm, 5, B = 5000), by = .(term, lev)] -> spec
 ```
 
-```{r A2, fig.cap = "Foutier spectrum", appendix = TRUE}
+```{r A2, fig.cap = "Fourier spectrum of each timeseries. The shading indicates de 95\\% qunatile derived fitting an AR process and bootstrapping 5000 simulated samples.", appendix = TRUE}
 spec %>%
-  .[lev %in% c(30, 700)] %>% 
   ggplot(aes(1/freq/12, spec)) +
-  # geom_ribbon(aes(group = interaction(term, rleid(sign(spec - boot_null))), 
-  #                 fill = sign(spec - boot_null),
-  #                 ymin = spec, ymax = boot_null)) +
+  geom_area(aes(fill = term, y = ar_spectrum_upp), position = "dodge", alpha = 0.2) +
+  # geom_line(aes(color = term, y = ar_spectrum), alpha = 0.5) +
   geom_line(aes(color = term)) +
-  geom_line(aes(color = term, y = boot_null)) +
   annotation_logticks(sides = "b") +
-  scale_color_brewer(palette = "Set1") +
-  scale_fill_divergent() +
-  scale_x_log10() +
-  facet_grid(lev ~ .,labeller = labeller(lev = lev.lab))
+  scale_color_brewer(NULL, palette = "Set1", aesthetics = c("color", "fill")) +
+  scale_x_log10("Period (years)") +
+  facet_grid(term ~ lev,labeller = labeller(lev = lev.lab, term = lab_sam))
 ```
 
-```{r A3, fig.cap = "Trends for each index at each level. Shading indicates the 95\\% confidence interval.", fig.width = 3, fig.height=3.5, appendix = TRUE}
+```{r A3, fig.cap = "Autocorrelation functions of each timeseries", appendix = TRUE}
+acf_dt <- function(...) {
+  a <- acf(..., plot = FALSE)
+  list(acf = as.vector(a$acf),
+       lag = as.vector(a$lag))
+}
+indexes %>% 
+  .[lev %in% c(30, 700)] %>% 
+  .[, acf_dt(estimate_norm), by = .(lev, term)] %>% 
+  ggplot(aes(lag, acf)) +
+  geom_col(aes(fill = term)) +
+  scale_fill_brewer(palette = "Set1") +
+  facet_grid(term ~ lev,labeller = labeller(lev = lev.lab, term = lab_sam))
+```
+
+```{r A4, fig.cap = "Trends for each index at each level. Shading indicates the 95\\% confidence interval.", fig.width = 3, fig.height=5, appendix = TRUE}
 indexes %>%
@@
   geom_line(aes(color = term)) +
+  scale_fill_brewer(palette = "Set1", aesthetics = c("color", "fill")) +
   geom_hline(yintercept = 0) +
   scale_x_level() +
-  coord_flip() 
+  coord_flip() +
+  facet_grid(term ~ .,labeller = labeller(term = lab_sam))
 ```

### 2020-07-08 analysis/paper/paper.tex
spectrum

@@
 \begin{figure}
-\includegraphics{A2-1} \appendcaption{A2}{Foutier spectrum}\label{fig:A2}
+\includegraphics{A2-1} \appendcaption{A2}{Fourier spectrum of each timeseries.
+The shading indicates de 95\% qunatile derived fitting an AR process and bootstrapping 5000 simulated samples.}\label{fig:A2}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A3-1} \appendcaption{A3}{Trends for each index at each level. Shading indicates the 95\% confidence interval.}\label{fig:A3}
+\includegraphics{A3-1} \appendcaption{A3}{Autocorrelation functions of each timeseries}\label{fig:A3}
+\end{figure}
+
+\begin{figure}
+\includegraphics{A4-1} \appendcaption{A4}{Trends for each index at each level. Shading indicates the 95\% confidence interval.}\label{fig:A4}
 \end{figure}

### 2020-07-08 R/spectrum.R

@@ -1,0 +1,58 @@
+#' FFT spectrum
+#'
+#' A light wrapper around [stats::spec.pgram()] and [stats::ar()]. Returns the spectrum of
+#' the signal and the spectrum of the fitted autoregressiv model.
+#'
+#' @param x numeric vector
+#' @param B number of bootstrap samples.
+#' @param spans vector of odd integers giving the widths of modified Daniell
+#' smoothers to be used to smooth the periodogram
+#' @param ... other arguments passed to [stats::spec.pgram()]
+#'
+#' @export
+fftspectrum <- function(x, spans = NULL, B = 10000, ..., probs = 0.95) {
+  mtm <- spec.pgram(x, spans = spans, ..., plot = FALSE)
+
+  out <- as.data.table(mtm[c("freq", "spec")])
+
+  ar <- ar(ts(x))
+  # rho <- a$ar
+  # var <- a$var.pred
+  out[, ar_spectrum := arspectrum(mtm$freq, ar$ar, ar$var.pred)]
+  out[, ar_spectrum_upp := null_ar_spectrum(B = B, length(x), ar, spans = spans, ..., probs = probs)]
+
+  return(out[])
+}
+
+arspectrum <- function(freq, rho, var) {
+  k <- seq_along(rho)
+  e <- vapply(freq, function(f) sum(rho * exp(-2*1i*pi*f*k)), complex(1))
+  var / (Mod(1 - e))^2
+}
+
+
+
+null_ar_spectrum_ <- function(B = 100, n, ar, spans = NULL, ..., probs = 0.95) {
+  y <- as.vector(arima.sim(model = list(ar = ar$ar), n = n))*sqrt(ar$var.pred)
+  nfreq <- length(spec.pgram(y, spans = spans, ..., plot = FALSE)$spec)
+  boots <- vapply(seq_len(B), function(b) {
+    y <- as.vector(arima.sim(model = list(ar = ar$ar), n = n))*sqrt(ar$var.pred)
+    spec.pgram(y, spans = spans, plot = FALSE)$spec
+
+  }, numeric(nfreq))
+
+  apply(boots, 1, quantile, probs = probs)
+}
+
+null_ar_spectrum <- memoise::memoise(null_ar_spectrum_)
+
+null_spec <- memoise::memoise(function(x, spans, B = 1000, ..., probs = 0.95) {
+
+  b <- boot::boot(x, function(d, i) spec.pgram(d[i], spans = spans,
+                                                  ...,
[…]

### 2020-07-16 analysis/paper/paper.Rmd
some results

@@
 
-main_levs <- c(700, 300, 30)
+main_levs <- c(700, 300, 30, 50)
 ```
@@
 
-### Definition of indexes
+### Definition of indices
 
@@
 The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
-We then compute the "Full", "Asymmetric" and "Symmetric" indexes, by regressing each geopotential field on these patterns (weighting by the cosine of latitude).
+We then compute the "Full", "Asymmetric" and "Symmetric" indices, by regressing each geopotential field on these patterns (weighting by the cosine of latitude).
 
-The three indexes are normalised by dividing them by the standard deviation of the "Full" index at each level.
-This means that comparing the magnitude between indexes is meaningful, but it also means that not every index will have unit standard deviation.
+The three indices are normalised by dividing them by the standard deviation of the "Full" index at each level.
+This means that comparing the magnitude between indices is meaningful, but it also means that not every index will have unit standard deviation.
 
@@
   setnames("hgt", "full") %>% 
-  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)]
+  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)] %>% 
+  rm_singleton()
 
@@
 
-sam_sep %>% 
-  .[lev == 700] %>%
-  rm_singleton() %>% 
-  melt(id.vars = c("lon", "lat")) %>% 
+sam_plot <- sam_sep %>% 
+  .[lev %in% c(700)] %>%
+  melt(id.vars = c("lon", "lat", "lev")) %>% 
   .[variable == "sym", value := value + rnorm(.N, 0, sd = 1e-6)] %>%
-  .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)] -> sam_plot
+  .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)]
 
@@
   coord_polar() +
-  facet_grid(~variable, labeller = labeller(variable = label_parsed, lev = lev.lab)) +
-  axis_labs_smol
+  facet_grid(lev~variable, labeller = labeller(variable = label_parsed, lev = lev.lab)) +
+  axis_labs_smol +
+  facet_tags()
 ```
@@
 ```{r compute-asymsam}
-indexes_file <- data_path("derived", "indexes.Rds")
-if (file.exists(indexes_file)) {
-  indexes <- readRDS(indexes_file)
+indices_file <- data_path("derived", "indices.Rds")
+if (file.exists(indices_file)) {
+  indices <- readRDS(indices_file)
 } else {
@@
 
-indexes_wide <- dcast(indexes, + lev + time ~ term, value.var = "estimate_norm")
+indices_wide <- dcast(indices, + lev + time ~ term, value.var = "estimate_norm")
 ```
@@
 ```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM.", fig.width=6, fig.height=3, fig.env="figure*"}
-indexes %>% 
+indices %>% 
   .[lev %in% c(30, 700) & term != "full"] %>% 
@@
   scale_y_continuous(NULL) +
-  scale_x_datetime(NULL) +
-  facet_grid(lev~., labeller = labeller(lev = lev.lab), scales = "free")
+  scale_x_datetime(NULL, expand = c(0, 0)) +
+  facet_grid(lev~., labeller = labeller(lev = lev.lab), scales = "free") +
+  facet_tags(stroke = 0.2)
 ```
 
+Figure~\ref{fig:asymsam-timeseries} shows the resulting Asymmetric and Symmetric time series corresponding to 700 and 30hPa. blablababla
+
+* stratosphere clearly nor normally distributed. a lot of values near 0 and some relatively high outliers. Especially true int he case of the asymmetric index. High frequency variablity.
+
+* In both levels, there's correlation between the series (expected),
+
+```{r compute-cors}
+cors <- indices_wide[, cor(asym, sym), by = lev] 
+```
+
+Correlations between the Asymmetric and Symmetric series are rather constant throught the troposphere, fluctiating between `r signif(cors[lev >= 100, min(V1)], 2)` and `r signif(cors[lev >= 100, max(V1)], 2)` (Figure~\ref{fig:cor-lev}).
+Futhermore, the cross-correlation of each series across levels --shown in Figure~\ref{fig:cross-correlation}-- are high in the trosposphere (greater than 0.9) for both indices.
+This suggests that both the Asymmetric and the Symemtric component of the tropospheric SAM are highly vertically coherent, both in their individual evolution and their temporal relationship.
+This is to be expected since the SAM is mostly equivalent barotropic (citaaaa).
+
+In the stratosphere the situation is different.
+As can be seen in Figure~\ref{fig:cor-lev}, the relationship between the Asymmetric and Symmetric indices varies with height above 100 hPa.
+It starts to decreese right over the tropopause, reaches a minium of `r signif(min(cors$V1), 2)` at `r cors[which.min(V1), lev]` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis.
+The cross-correlation across levels in the stratosphere is generally weaker than in the troposphere (Figure~\ref{fig:cross-correlation}).
+Furthermore, above 100 hPa, the cross-correlation decreases more rapdily with height for the Symmetric SAM than for the Asymmetric SAM as evidenced by the wider dark red areas near the diagonal in Figure~\ref{fig:cross-correlation}b) vs.
+Figure~\ref{fig:cross-correlation}c).
+Moreover, the stratospheric Symmetric SAM seems to be slightly more connected to the trosposphere than the Asymmetric SAM; this can be seen by the lower correlation values in the top right quadrant of Figure~\ref{fig:cross-correlation}b) in comparison with Figure~\ref{fig:cross-correlation}c).
+
 ```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level.", fig.width=3, fig.height=3.5}
-indexes_wide[, cor(asym, sym), by = lev] %>% 
+cors %>% 
   ggplot(aes(lev, V1)) +
@@
 
-indexes %>% 
+indices %>% 
   copy() %>% 
@@
   geom_contour_fill(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
+  geom_text_contour(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1), 
+                    color = "white", rotate = FALSE, stroke.color = "black", stroke = 0.2, 
+                    skip = 1, size = 2) +
   geom_polygon(data = tri, fill = "white") +
@@
   facet_grid(. ~ term, labeller = labeller(term = lab_sam)) +
-  theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside")
+  theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside") +
+  facet_tags(stroke = 0.2)
 ```
 
+Figure~\ref{fig:cross-correlation}a) show the cross-correlation across levels for the Full SAM index. .....
+
 ```{r compute-regr-pattern}
-indexes_2d_yearly <- hgt[lev %in% main_levs] %>% 
-  indexes_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
-  .[, rbind(as.data.table(FitLm(hgt_a, sym, asym)),
-            as.data.table(FitLm(hgt_a, full))),
+indices_2d_yearly <- hgt[lev %in% c(main_levs)] %>% 
+  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
+  .[, rbind(as.data.table(FitLm(hgt_a, sym, asym, se = TRUE)),
+            as.data.table(FitLm(hgt_a, full, se = TRUE))),
     by = .(lon, lat, lev)] %>% 
   rm_intercept() %>% 
+  .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, lev)] %>% 
   periodic(lon = c(0, 360)) %>% 
@@
 
+To undertand the spatial patterns associated with both indeces we regressed monthly geopotential anomalies into both indeces using multiple regression (Figure A6 illustrates the difference between computing two simple regressions and one multiple regression).
+
 ```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM.
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.env = "figure*", fig.width=6, fig.height=7}
-indexes_2d_yearly %>% 
+indices_2d_yearly %>% 
   .[lat <= -20] %>% 
@@
   geom_qmap(~.x[lat <= -20]) +
-  # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
+  # geom_contour2(aes(z = p.val), breaks = 0.01) +
+  # stat_subset(aes(subset = p.val <= 0.01 & is.cross(lon, lat, 2)), geom = "point", size = 0.5, alpha = 0.3) +
   scale_x_longitude() +
@@
   facet_grid(term ~ lev, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  facet_tags() +
   axis_labs_smol 
 ```
-`r paste(stringi::stri_rand_lipsum(2, FALSE), collapse = "\n\n")`
+
+Figure~\ref{fig:2d-regr} shows the spatial year-long regression for selected levels.
+In the troposphere the Full annular mode is clearly "contaminated" with well known zonal asymemtries (panels g and j) which are succesfully sepparated by our methodology (panels h, i, k and l).
+In the stratosphere, the spatial pattern associated with the Full SAM is much more clearly dominated by a zonally symmetric, monopolar structure (panels a and d) that is, however, not perfectly centered in the south pole.
+The monopoloe obtained by multiple regression with the Asymmetric and Symmetric SAM (panels c and f in Figure~\ref{fig:2d-regr}) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 pattern (panels b and e).
+
+```{r vertical-regression, fig.cap = "Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South.
+Values are standarised by the standard deviation at each level.
+(this caption needs some love)", fig.width = 4, fig.height=3}
+ReadNetCDF(ERA5(), 
+           subset = list(time = c("1979-01-01", "2018-12-31"),
+                         latitude = c(-65, -40)),
+           vars = c(hgt = "z")) %>% 
+  normalise_coords() %>% 
+  .[, .(hgt = mean(hgt)), by = .(lon, lev, time)] %>% 
+  .[, hgt := hgt/9.8] %>% 
+  .[, hgt_a := hgt - mean(hgt), by = .(lon, lev, month(time))] %>% 
+  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
+  .[, FitLm(hgt_a, sym, asym, se = TRUE),
+    by = .(lon, lev)] %>% 
+  rm_intercept() %>% 
+  .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, lev)] %>% 
+  periodic(lon = c(0, 360)) %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] -> vertical_regr
+
+vertical_regr %>% 
+  copy() %>% 
+  .[term == "asym"] %>% 
+  .[, estimate := estimate/sd(estimate), by = .(lev, term)] %>%
+  ggplot(aes(lon, lev)) +
+  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, 0.5, 0)) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(15), 
+                       oob = scales::squish,
+                       breaks = MakeBreaks(binwidth = 0.5, exclude = 0)) +
+  scale_y_level() +
+  scale_x_longitude() 
+```
 
 ```{r compute-regr-pattern-season}
-indexes_2d <- hgt[lev %in% main_levs] %>% 
-  indexes_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
+indices_2d <- hgt[lev %in% main_levs] %>% 
+  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
   .[, rbind(as.data.table(FitLm(hgt_a, sym, asym)),
@@
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
-indexes_2d %>% 
+indices_2d %>% 
   .[lat <= -20] %>% 
@@
   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  facet_tags() +
   axis_labs_smol
@@
 ```{r wave-amplitude-700, fig.cap = "Planteray wave amplitude for the regression patterns at 700 hPa.", fig.width=6, fig.height=4}
-indexes_2d %>% 
+indices_2d %>% 
   .[ lev == 700 & lat < -20] %>%
@@
   coord_flip() +
-  facet_grid(term~season, labeller = labeller(term = lab_sam), scales = "free_x")
+  facet_grid(term~season, labeller = labeller(term = lab_sam), scales = "free_x") +
+  facet_tags("tr")
 ```
@@
 ```{r wave-amplitude-30, fig.cap = "Planteray wave amplitude for the regression patterns at 30 hPa.", fig.width=6, fig.height=4}
-indexes_2d %>% 
+indices_2d %>% 
   .[ lev == 30 & lat < -20] %>%
@@
   coord_flip() +
-  facet_grid(term~season, labeller = labeller(term = lab_sam), scales = "free_x") 
+  facet_grid(term~season, labeller = labeller(term = lab_sam), scales = "free_x") +
+  facet_tags("tr")
 ```
@@
   # .[, air := Anomaly(air), by = .(lon, lat)] %>%
-  .[indexes_wide[lev == 700], on = .NATURAL] %>% 
+  .[indices_wide[lev == 700], on = .NATURAL] %>% 
   rm_singleton() %>% 
@@
   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  axis_labs_smol
-```
-
-```{r regr-pp-season, fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", invisible = TRUE}
-
-is.sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
-  data[, lon %between% lons & lat %between% lats]
-}
-
-filter_sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
-  data[is.sa(data, lons = lons, lats = lats)]
-}
-
-pp_regr_season %>% 
-  filter_sa() %>% 
-  .[, lon := ConvertLongitude(lon)] %>% 
-  ggplot(aes(lon, lat)) +
-  # geom_raster(aes(fill = estimate)) +
-  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.25, 0)) +
-  geom_contour2(aes(z = p.value), breaks = 0.01) +
-  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
-  stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
-  geom_qmap(~filter_sa(.x)[, lon := ConvertLongitude(lon)], keep = 0.02) +
-  scale_x_longitude() +
-  scale_y_latitude() +
-  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
-                       breaks = AnchorBreaks(0, 0.25, 0)) +
-  # coord_polar() +
-  ggalt::coord_proj("+proj=moll +lon_0=-60") +
-  facet_grid(season~term, labeller = labeller(term = lab_sam)) +
+  facet_tags() +
   axis_labs_smol
@@
 ```{r compute-air-regr}
-time_subset <- list(time = range(indexes$time))
+time_subset <- list(time = range(indices$time))
 air <- NOAATemp() %>% 
@@
 
-indexes_wide %>% 
+indices_wide %>% 
   .[, ccf_dt(asym, sym, lag.max = 5), by = .(lev)] %>% 
@@
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.05)) +
-  scale_x_continuous("\u2190 Asym leads Sym\t \t\t Sym leads Asym \U2192", breaks = seq(-5, 5, 1), 
+  scale_x_continuous("Asym leads Sym\t \t\t Sym leads Asym", breaks = seq(-5, 5, 1), 
                      expand = c(0, 0)) +
@@
 
-```{r compute-spectrums}
-indexes %>% 
-  .[lev %in% c(30, 700)] %>% 
-  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
-  .[, fftspectrum(estimate_norm, 5, B = 5000), by = .(term, lev)] -> spec
+```{r ccf-levels, fig.cap = "Cross-correlation functions for each index and two differnet base levels."}
+ccf_dt <- function(...) {
+  a <- ccf(..., plot = FALSE)
+  list(acf = as.vector(a$acf),
+       lag = as.vector(a$lag))
+}
+
+index2 <- copy(indices)
+index2[, estimate_700 := estimate[lev == 700], by = .(term, time)]
+index2[, estimate_10 := estimate[lev == 10], by = .(term, time)]
+
+rbind("700" = index2[, ccf_dt(x = estimate, y = estimate_700, lag.max = 5), by = .(lev, term)],
+      "10" = index2[, ccf_dt(x = estimate, y = estimate_10, lag.max = 5), by = .(lev, term)], 
+      idcol = "base_level") %>% 
+  ggplot(aes(lag, lev)) +
+  geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.1, 0)) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
+                       breaks = AnchorBreaks(0, 0.1, 0)) +
+  scale_y_level() +
+  scale_x_continuous(breaks = seq(-10, 10)) +
+  facet_grid(term~base_level, labeller = labeller(base_level = function(x) paste0("Base level: ", x, " hPa"), 
+                                                  term = lab_sam)) +
+  facet_tags() 
 ```
 
-```{r A2, fig.cap = "Fourier spectrum of each timeseries. The shading indicates de 95\\% qunatile derived fitting an AR process and bootstrapping 5000 simulated samples.", appendix = TRUE}
+```{r compute-spectrums}
+indices %>% 
+  .[lev %in% c(50, 700)] %>% 
+  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
+  .[, fftspectrum(estimate, 5, B = 5000, probs = c(0.025, 0.975)), by = .(term, lev)] -> spec
+```
+
+```{r A2, fig.cap = "Fourier spectrum of each timeseries.
+The shading indicates de 95\\% area derived by fitting an AR process to each series and bootstrapping 5000 simulated samples.", appendix = TRUE}
 spec %>%
+  .[term != "full"] %>% 
   ggplot(aes(1/freq/12, spec)) +
-  geom_area(aes(fill = term, y = ar_spectrum_upp), position = "dodge", alpha = 0.2) +
-  # geom_line(aes(color = term, y = ar_spectrum), alpha = 0.5) +
+  geom_ribbon(aes(fill = term, ymin = `2%`, ymax = `98%`), position = "dodge", alpha = 0.1) +
+  geom_line(aes(color = term, y = ar_spectrum), alpha = 0.15) +
   geom_line(aes(color = term)) +
@@
   scale_color_brewer(NULL, palette = "Set1", aesthetics = c("color", "fill")) +
-  scale_x_log10("Period (years)") +
-  facet_grid(term ~ lev,labeller = labeller(lev = lev.lab, term = lab_sam))
+  scale_x_log10("Period (years)", sec.axis = sec_axis(~.*12)) +
+  facet_grid(lev ~ term,labeller = labeller(lev = lev.lab, term = lab_sam), scales = "free") +
+  facet_tags() 
 ```
@@
 }
-indexes %>% 
+indices %>% 
   .[lev %in% c(30, 700)] %>% 
@@
   scale_fill_brewer(palette = "Set1") +
-  facet_grid(term ~ lev,labeller = labeller(lev = lev.lab, term = lab_sam))
+  facet_grid(term ~ lev,labeller = labeller(lev = lev.lab, term = lab_sam)) +
+  facet_tags("tr") 
 ```
@@
 ```{r A4, fig.cap = "Trends for each index at each level. Shading indicates the 95\\% confidence interval.", fig.width = 3, fig.height=5, appendix = TRUE}
-indexes %>%
+indices %>%
   copy() %>% 
@@
   coord_flip() +
-  facet_grid(term ~ .,labeller = labeller(term = lab_sam))
+  facet_grid(term ~ .,labeller = labeller(term = lab_sam)) +
+  facet_tags() 
 ```
+
+```{r A5, fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", appendix = TRUE}
+
+is.sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
+  data[, lon %between% lons & lat %between% lats]
+}
+
+filter_sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
+  data[is.sa(data, lons = lons, lats = lats)]
+}
+
+pp_regr_season %>% 
+  filter_sa() %>% 
+  .[, lon := ConvertLongitude(lon)] %>% 
+  ggplot(aes(lon, lat)) +
+  # geom_raster(aes(fill = estimate)) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.25, 0)) +
+  geom_contour2(aes(z = p.value), breaks = 0.01) +
+  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
+  stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
+  geom_qmap(~filter_sa(.x)[, lon := ConvertLongitude(lon)], keep = 0.02) +
+  scale_x_longitude() +
+  scale_y_latitude() +
+  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
+                       breaks = AnchorBreaks(0, 0.25, 0)) +
+  # coord_polar() +
+  ggalt::coord_proj("+proj=moll +lon_0=-60") +
+  facet_grid(season~term, labeller = labeller(term = lab_sam)) +
+  facet_tags() +
+  axis_labs_smol 
+```
+
+```{r A6, appendix = TRUE, fig.cap = "Regressions maps resulting from performing one multiple regression (a. and b.) and from performing two simple regressions (c. and d.)"}
+regr_dif <- hgt[lev %in% 700] %>% 
+  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
+  .[,  rbind(idcol = "regression", multiple = as.data.table(FitLm(hgt_a, sym, asym)),
+             single = rbind(as.data.table(FitLm(hgt_a, sym)),
+                            as.data.table(FitLm(hgt_a, asym)))),
+    by = .(lon, lat)]
+
+regr_dif %>% 
+  rm_intercept() %>% 
+  periodic(lon = c(0, 360)) %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
+  .[lat <= -20] %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
+  # geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
+  # size = 0.3,
+  # global.breaks = FALSE) +
+  geom_contour_tanaka(aes(z = estimate),
+                      breaks = ZeroBreaks,
+                      size = 0.3,
+                      global.breaks = FALSE) +
+  geom_qmap(~.x[lat <= -20]) +
+  # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
+  scale_x_longitude() +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
+                       limits = c(-85, 85),
+                       oob = scales::squish,
+                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
+  scale_linetype(guide = "none") +
+  coord_polar() +
+  facet_grid(term ~ regression, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  facet_tags() +
+  axis_labs_smol 
+```

### 2020-07-16 analysis/paper/paper.tex
some results

@@
 
-\subsubsection{Definition of indexes}
+\subsubsection{Definition of indices}
 
@@
 and symmetric (\([\mathrm{EOF_1}](\lambda, \phi)\)) components. We then
-compute the ``Full'', ``Asymmetric'' and ``Symmetric'' indexes, by
+compute the ``Full'', ``Asymmetric'' and ``Symmetric'' indices, by
 regressing each geopotential field on these patterns (weighting by the
@@
 
-The three indexes are normalised by dividing them by the standard
+The three indices are normalised by dividing them by the standard
 deviation of the ``Full'' index at each level. This means that comparing
-the magnitude between indexes is meaningful, but it also means that not
+the magnitude between indices is meaningful, but it also means that not
 every index will have unit standard deviation.
@@
 
+Figure\textasciitilde{}\ref{fig:asymsam-timeseries} shows the resulting
+Asymmetric and Symmetric time series corresponding to 700 and 30hPa.
+blablababla
+
+\begin{itemize}
+\item
+  stratosphere clearly nor normally distributed. a lot of values near 0
+  and some relatively high outliers. Especially true int he case of the
+  asymmetric index. High frequency variablity.
+\item
+  In both levels, there's correlation between the series (expected),
+\end{itemize}
+
+Correlations between the Asymmetric and Symmetric series are rather
+constant throught the troposphere, fluctiating between 0.39 and 0.45
+(Figure\textasciitilde{}\ref{fig:cor-lev}). Futhermore, the
+cross-correlation of each series across levels --shown in
+Figure\textasciitilde{}\ref{fig:cross-correlation}-- are high in the
+trosposphere (greater than 0.9) for both indices. This suggests that
+both the Asymmetric and the Symemtric component of the tropospheric SAM
+are highly vertically coherent, both in their individual evolution and
+their temporal relationship. This is to be expected since the SAM is
+mostly equivalent barotropic (citaaaa).
+
+In the stratosphere the situation is different. As can be seen in
+Figure\textasciitilde{}\ref{fig:cor-lev}, the relationship between the
+Asymmetric and Symmetric indices varies with height above 100 hPa. It
+starts to decreese right over the tropopause, reaches a minium of 0.21
+at 20 hPa and then it increases again monotonically with height up to
+the uppermost level of the reanalysis. The cross-correlation across
+levels in the stratosphere is generally weaker than in the troposphere
+(Figure\textasciitilde{}\ref{fig:cross-correlation}). Furthermore, above
+100 hPa, the cross-correlation decreases more rapdily with height for
+the Symmetric SAM than for the Asymmetric SAM as evidenced by the wider
+dark red areas near the diagonal in
+Figure\textasciitilde{}\ref{fig:cross-correlation}b)
+vs.~Figure\textasciitilde{}\ref{fig:cross-correlation}c). Moreover, the
+stratospheric Symmetric SAM seems to be slightly more connected to the
+trosposphere than the Asymmetric SAM; this can be seen by the lower
+correlation values in the top right quadrant of
+Figure\textasciitilde{}\ref{fig:cross-correlation}b) in comparison with
+Figure\textasciitilde{}\ref{fig:cross-correlation}c).
+
 \begin{figure}
@@
 
+Figure\textasciitilde{}\ref{fig:cross-correlation}a) show the
+cross-correlation across levels for the Full SAM index. \ldots{}..
+
+To undertand the spatial patterns associated with both indeces we
+regressed monthly geopotential anomalies into both indeces using
+multiple regression (Figure A6 illustrates the difference between
+computing two simple regressions and one multiple regression).
+
 \begin{figure*}
@@
 
-Sem tortor mus non tristique augue. Donec vel feugiat sit aliquet lorem
-dui lobortis dolor risus. Proin, ac eu, mauris. Dolor id habitasse,
-curabitur platea ante. Rhoncus volutpat himenaeos vestibulum ut bibendum
-felis non. Ligula placerat ac luctus ad. Ante, lorem volutpat eu dapibus
-pharetra arcu. Sollicitudin eu tortor amet. Maecenas nulla euismod ac
-suspendisse metus. A ac, hendrerit, sit sollicitudin ac primis
-ullamcorper tempor id, mauris. Sem diam, sem curabitur. Risus, nullam
-vel diam hendrerit massa nec non. Donec amet amet ultrices ac taciti
-aliquam. Enim arcu et sed sit.
+Figure\textasciitilde{}\ref{fig:2d-regr} shows the spatial year-long
+regression for selected levels. In the troposphere the Full annular mode
+is clearly ``contaminated'' with well known zonal asymemtries (panels g
+and j) which are succesfully sepparated by our methodology (panels h, i,
+k and l). In the stratosphere, the spatial pattern associated with the
+Full SAM is much more clearly dominated by a zonally symmetric,
+monopolar structure (panels a and d) that is, however, not perfectly
+centered in the south pole. The monopoloe obtained by multiple
+regression with the Asymmetric and Symmetric SAM (panels c and f in
+Figure\textasciitilde{}\ref{fig:2d-regr}) is much more symmetric and the
+shift from total symmetry is captured by the regression pattern of the
+Asymmetric SAM as a wave-1 pattern (panels b and e).
 
-Ligula sed sociosqu imperdiet magna curabitur, et, elit lobortis ut,
-sed. Placerat id eros quis vivamus sed, in felis. Egestas phasellus at
-sagittis sed senectus malesuada. Vestibulum at pretium elementum id,
-facilisi. Porttitor luctus vel tempus diam nec per, nibh nisi. Viverra
-turpis nisl sed nec varius, et ridiculus in amet. Amet sapien nec
-convallis arcu dis et mauris dolor. Vestibulum, in, in tincidunt tellus
-risus class quisque, augue feugiat.
+\begin{figure}
+\includegraphics{vertical-regression-1} \caption[Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South]{Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South.
+Values are standarised by the standard deviation at each level.
+(this caption needs some love)}\label{fig:vertical-regression}
+\end{figure}
 
@@
 \begin{figure}
+\includegraphics{ccf-levels-1} \caption[Cross-correlation functions for each index and two differnet base levels]{Cross-correlation functions for each index and two differnet base levels.}\label{fig:ccf-levels}
+\end{figure}
+
+\begin{figure}
 \includegraphics{A2-1} \appendcaption{A2}{Fourier spectrum of each timeseries.
-The shading indicates de 95\% qunatile derived fitting an AR process and bootstrapping 5000 simulated samples.}\label{fig:A2}
+The shading indicates de 95\% area derived by fitting an AR process to each series and bootstrapping 5000 simulated samples.}\label{fig:A2}
 \end{figure}
@@
 
+\begin{figure}
+\includegraphics{A5-1} \appendcaption{A5}{Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:A5}
+\end{figure}
+
+\begin{figure}
+\includegraphics{A6-1} \appendcaption{A6}{Regressions maps resulting from performing one multiple regression (a. and b.) and from performing two simple regressions (c. and d.)}\label{fig:A6}
+\end{figure}
+
 \end{document}

### 2020-07-16 R/ggplot-tags.R

@@ -1,0 +1,87 @@
+#' Adds tags to facets
+#'
+#' @export
+facet_tags <- function(position = "tl", tag_pool = letters, open = "", close = ")", stroke = 0.15, stroke.color = "white",
+                 color = "black") {
+  if (!is.list(position)) {
+    position <- switch(position,
+                       tl = list(x = 0.05, y = 0.95, hjust = 0, vjust = 1),
+                       tr = list(x = 0.95, y = 0.95, hjust = 1, vjust = 1),
+                       bl = list(x = 0.05, y = 0.05, hjust = 0, vjust = 0),
+                       br = list(x = 0.95, y = 0.05, hjust = 1, vjust = 0),
+                       position)
+  }
+
+
+  structure(list(tag_pool = tag_pool, open = open, close = close, position = position,
+                 color = color,
+                 stroke = stroke, stroke.color = stroke.color),
+            class = "ggtags")
+}
+
+#' @export
+ggplot_add.ggtags <- function(object, plot, object_name) {
+  class(plot) <- c("ggtagged", class(plot))
+  attr(plot, "tag_options") <- object
+  plot
+}
+
+#' @export
+print.ggtagged <- function(x, newpage = is.null(vp), vp = NULL, ...) {
+  ggplot2::set_last_plot(x)
+  if (newpage) grid::grid.newpage()
+
+  # Record dependency on 'ggplot2' on the display list
+  # (AFTER grid.newpage())
+  grDevices::recordGraphics(
+    requireNamespace("ggplot2", quietly = TRUE),
+    list(),
+    getNamespace("ggplot2")
+  )
+
+  tag_options <- attr(x, "tag_options")
+  gb <- ggplot2::ggplot_build(x)
+  lay <- gb$layout$layout
+
+  x = tag_options$position$x
+  y = tag_options$position$y
+
+  hjust = tag_options$position$hjust
+  vjust = tag_options$position$vjust
+
+  gt <- ggplot2::ggplot_gtable(gb)
+  panels <- which(grepl("panel", gt$layout$name))
+  for (p in seq_along(panels)) {
+
+    if (tag_options$stroke == 0) {
+      tagGrob <- grid::textGrob(paste0(tag_options$open, tag_options$tag_pool[p], tag_options$close),
+                                            x = x, y = y, gp = grid::gpar(col = tag_options$color),
+                                            hjust = hjust, vjust = vjust)
+    } else {
[…]

### 2020-07-16 R/nc-management.R

@@ -1,0 +1,48 @@
+#' Tools to manage netcdf4 files
+#'
+#' Wrappers around nco commands .
+#'
+#' @param in_files character vector of files
+#' @param out_file file to save output
+#' @param overwrite wether to overwrite exiting files
+#' @param verbose print out commands
+#'
+#' @export
+#' @rdname nc-files
+nc_append <- function(in_files, out_file, verbose = TRUE) {
+  if (file.exists(out_file)) file.remove(out_file)
+
+  in_files <- paste0("\"", in_files, "\"")
+  out_file <- paste0("\"", out_file, "\"")
+
+  for (file in in_files) {
+    command <- paste("ncks -A", file, out_file)
+    if (verbose) {
+      message("Processing ", file, " with command:\n", command)
+    }
+    system(command)
+  }
+  return(out_file)
+
+}
+
+#' @export
+#' @rdname nc-files
+nc_concatenate <- function(in_files, out_file, verbose = TRUE, overwrite = FALSE) {
+
+  if (file.exists(out_file) & overwrite == FALSE) {
+    if (verbose) message("Non existent file: ", out_file)
+    return(out_file)
+  }
+  if (file.exists(out_file)) file.remove(out_file)
+
+  in_files <- paste0("\"", in_files, "\"")
+  out_file <- paste0("\"", out_file, "\"")
+
+
+  command <- paste("ncrcat", paste(in_files, collapse = " "), out_file)
+  if (verbose) {
+    message("Concatenating files with command:\n", command)
+  }
+  system(command)
+}

### 2020-07-16 R/spectrum.R

@@ -19,7 +19,7 @@ fftspectrum <- function(x, spans = NULL, B = 10000, ..., probs = 0.95) {
   # rho <- a$ar
   # var <- a$var.pred
   out[, ar_spectrum := arspectrum(mtm$freq, ar$ar, ar$var.pred)]
-  out[, ar_spectrum_upp := null_ar_spectrum(B = B, length(x), ar, spans = spans, ..., probs = probs)]
+  out[, c(scales::percent(probs)) := null_ar_spectrum(B = B, length(x), ar, spans = spans, ..., probs = probs)]
 
   return(out[])
 }
@@ -40,8 +40,7 @@ null_ar_spectrum_ <- function(B = 100, n, ar, spans = NULL, ..., probs = 0.95) {
     spec.pgram(y, spans = spans, plot = FALSE)$spec
 
   }, numeric(nfreq))
-
-  apply(boots, 1, quantile, probs = probs)
+  data.table::as.data.table(t(apply(boots, 1, quantile, probs = probs)))
 }
 
 null_ar_spectrum <- memoise::memoise(null_ar_spectrum_)

### 2020-07-16 analysis/scripts/01-get-data.R

@@ -108,7 +108,15 @@ if (!force && file.exists(asymsam::data_path("raw", file))) {
 }
 
 
-# if (zip) {
-#   utils::zip(zipfile = asymsam::data_path("raw", "data_raw.zip"),
-#              files = data_path("raw", unlist(files)))
-# }
+for (y in seq(1979, 2018)) {
+  url <- paste0("ftp://ftp.cdc.noaa.gov/Datasets/ncep.reanalysis.dailyavgs/pressure/hgt.", y, ".nc")
+  dir.create(asymsam::data_path("raw", "ncep"), )
+  file <- asymsam::data_path("raw", "ncep", paste0("hgt.daily.", y, ".nc"))
+  if (!file.exists(file)) {
+    download.file(url, destfile = file)
+  }
+}
+
+ncepfiles <- list.files(asymsam::data_path("raw", "ncep"), full.names = TRUE)
+
+asymsam::nc_concatenate(ncepfiles, asymsam::data_path("raw", "ncep.daily.nc"), overwrite = TRUE)

### 2020-07-16 analysis/scripts/02-compute-eofs.R

@@ -3,9 +3,9 @@ library(magrittr)
 library(asymsam)
 library(data.table)
 
-indexes_file <- data_path("derived", "indexes.Rds")
+indices_file <- data_path("derived", "indices.Rds")
 
-if (!file.exists(indexes_file)) {
+if (!file.exists(indices_file)) {
   hgt <- ReadNetCDF(ERA5(),
                     subset = list(time = c("1979-01-01", "2018-12-31"),
                                   latitude = c(-90, 10)),
@@ -14,9 +14,9 @@ if (!file.exists(indexes_file)) {
     .[, hgt := hgt/9.8] %>%
     .[, hgt_a := hgt - mean(hgt), by = .(lon, lat, lev, month(time))]
 
-  indexes <- hgt[lat <= -20, eof_asym(hgt_a, lon, lat, time, n = 1), by = lev]
+  indices <- hgt[lat <= -20, eof_asym(hgt_a, lon, lat, time, n = 1), by = lev]
 
-  saveRDS(indexes, file = data_path("derived", "indexes.Rds"))
+  saveRDS(indices, file = data_path("derived", "indices.Rds"))
 
 }
 

### 2020-07-16 analysis/scripts/03-compute-daily-eofs.R

@@ -1,0 +1,69 @@
+ccf_dt <- function(...) {
+  a <- ccf(..., plot = FALSE)
+  list(acf = as.vector(a$acf),
+       lag = as.vector(a$lag))
+}
+
+SAM <- ReadNetCDF(ERA5(),
+                  subset = list(time = c("1979-01-01", "2018-12-31"),
+                                latitude = c(-90, -20)),
+                  vars = c(hgt = "z")) %>%
+  normalise_coords() %>%
+  .[, hgt := hgt/9.8] %>%
+  .[, hgt_a := hgt - mean(hgt), by = .(lon, lat, lev, month(time))] %>%
+  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>%
+  .[, .(eof = list(eof_flip(EOF(hgt ~ time | lon + lat, n = 1, data = .SD)))),
+    by = .(lev)]
+
+sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>%
+  setnames("hgt", "full") %>%
+  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)]
+
+
+metadata <- GlanceNetCDF(asymsam::data_path("raw", "ncep.daily.nc"))
+
+
+
+make_projection <- function(lev) {
+  this_lev <- lev
+  day_hgt <- ReadNetCDF(asymsam::data_path("raw", "ncep.daily.nc"),
+                        vars = c(hgt = "hgt"),
+                        subset = list(lat = c(-90, -20),
+                                      level = this_lev,
+                                      time = c("1979-01-01", "2018-12-31"))) %>%
+    asymsam::normalise_coords() %>%
+    .[, month := month(time[1]), by = time]
+
+  day_hgt <- sam_sep[lev == this_lev][day_hgt, on = c("lon", "lat", "lev")]
+
+  day_hgt[, hgt_m := mean(hgt), by = .(lon, lat, month, lev)]
+  day_hgt[, hgt_a := hgt - hgt_m]
+
+  day_hgt[, coslat := cos(lat*pi/180)]
+  asym <- day_hgt[, .(estimate = sum(hgt_a*asym*coslat)/sum(coslat)), by = .(time, lev)]
+  sym <- day_hgt[, .(estimate = sum(hgt_a*sym*coslat)/sum(coslat)), by = .(time, lev)]
+  rbind(asym = asym, sym = sym, idcol = "term")
+}
+
[…]

### 2020-07-16 analysis/paper/paper.Rmd
more results

@@
   cache = TRUE,
-  cache.extra = 3,
+  cache.extra = 42,
   cache.path = "../cache/",
@@
   axis_labs_smol +
-  facet_tags()
+  tag_facets()
 ```
@@
 # Results
+
+## Temporal evolution
 
@@
   facet_grid(lev~., labeller = labeller(lev = lev.lab), scales = "free") +
-  facet_tags(stroke = 0.2)
+  tag_facets(stroke = 0.2)
 ```
 
-Figure~\ref{fig:asymsam-timeseries} shows the resulting Asymmetric and Symmetric time series corresponding to 700 and 30hPa. blablababla
+Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and Symmetric time series corresponding to 700 and 30hPa. blablababla
 
@@
 
-Correlations between the Asymmetric and Symmetric series are rather constant throught the troposphere, fluctiating between `r signif(cors[lev >= 100, min(V1)], 2)` and `r signif(cors[lev >= 100, max(V1)], 2)` (Figure~\ref{fig:cor-lev}).
-Futhermore, the cross-correlation of each series across levels --shown in Figure~\ref{fig:cross-correlation}-- are high in the trosposphere (greater than 0.9) for both indices.
+Correlations between the Asymmetric and Symmetric series are rather constant throught the troposphere, fluctiating between `r signif(cors[lev >= 100, min(V1)], 2)` and `r signif(cors[lev >= 100, max(V1)], 2)` (Figure \ref{fig:cor-lev}).
+Futhermore, the cross-correlation of each series across levels --shown in Figure \ref{fig:cross-correlation}-- are high in the trosposphere (greater than 0.9) for both indices.
 This suggests that both the Asymmetric and the Symemtric component of the tropospheric SAM are highly vertically coherent, both in their individual evolution and their temporal relationship.
@@
 In the stratosphere the situation is different.
-As can be seen in Figure~\ref{fig:cor-lev}, the relationship between the Asymmetric and Symmetric indices varies with height above 100 hPa.
+As can be seen in Figure \ref{fig:cor-lev}, the relationship between the Asymmetric and Symmetric indices varies with height above 100 hPa.
 It starts to decreese right over the tropopause, reaches a minium of `r signif(min(cors$V1), 2)` at `r cors[which.min(V1), lev]` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis.
-The cross-correlation across levels in the stratosphere is generally weaker than in the troposphere (Figure~\ref{fig:cross-correlation}).
-Furthermore, above 100 hPa, the cross-correlation decreases more rapdily with height for the Symmetric SAM than for the Asymmetric SAM as evidenced by the wider dark red areas near the diagonal in Figure~\ref{fig:cross-correlation}b) vs.
-Figure~\ref{fig:cross-correlation}c).
-Moreover, the stratospheric Symmetric SAM seems to be slightly more connected to the trosposphere than the Asymmetric SAM; this can be seen by the lower correlation values in the top right quadrant of Figure~\ref{fig:cross-correlation}b) in comparison with Figure~\ref{fig:cross-correlation}c).
+The cross-correlation across levels in the stratosphere is generally weaker than in the troposphere (Figure \ref{fig:cross-correlation}).
+Furthermore, above 100 hPa, the cross-correlation decreases more rapdily with height for the Symmetric SAM than for the Asymmetric SAM as evidenced by the wider dark red areas near the diagonal in Figure \ref{fig:cross-correlation}b) vs.
+Figure \ref{fig:cross-correlation}c).
+Moreover, the stratospheric Symmetric SAM seems to be slightly more connected to the trosposphere than the Asymmetric SAM; this can be seen by the lower correlation values in the top right quadrant of Figure \ref{fig:cross-correlation}b) in comparison with Figure \ref{fig:cross-correlation}c).
 
-```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level.", fig.width=3, fig.height=3.5}
+```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level.", fig.width=3, fig.height=3}
 cors %>% 
@@
   theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside") +
-  facet_tags(stroke = 0.2)
+  tag_facets("br", stroke = 0.2)
 ```
 
-Figure~\ref{fig:cross-correlation}a) show the cross-correlation across levels for the Full SAM index. .....
+Figure \ref{fig:cross-correlation}a) show the cross-correlation across levels for the Full SAM index. .....
 
@@
 ```
+
+## Spatial patterns
 
@@
   # stat_subset(aes(subset = p.val <= 0.01 & is.cross(lon, lat, 2)), geom = "point", size = 0.5, alpha = 0.3) +
+  # geom_hline(yintercept = c(-65, -40)) +
   scale_x_longitude() +
@@
   facet_grid(term ~ lev, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  facet_tags() +
+  tag_facets() +
   axis_labs_smol 
@@
 
-Figure~\ref{fig:2d-regr} shows the spatial year-long regression for selected levels.
+Figure \ref{fig:2d-regr} shows the spatial year-long regression for selected levels.
 In the troposphere the Full annular mode is clearly "contaminated" with well known zonal asymemtries (panels g and j) which are succesfully sepparated by our methodology (panels h, i, k and l).
 In the stratosphere, the spatial pattern associated with the Full SAM is much more clearly dominated by a zonally symmetric, monopolar structure (panels a and d) that is, however, not perfectly centered in the south pole.
-The monopoloe obtained by multiple regression with the Asymmetric and Symmetric SAM (panels c and f in Figure~\ref{fig:2d-regr}) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 pattern (panels b and e).
+The monopoloe obtained by multiple regression with the Asymmetric and Symmetric SAM (panels c and f in Figure \ref{fig:2d-regr}) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 pattern (panels b and e).
+
+```{r wave-amplitude, fig.cap = "Planteray wave amplitude for the regression patterns at 50 and 700 hPa.", fig.width=3, fig.height=4}
+indices_2d_yearly %>% 
+  .[lat < -20] %>%
+  .[lev %in% c(50, 700)] %>% 
+  .[, FitWave(estimate, 0:3), by = .(lat, lev, term)] %>% 
+  ggplot(aes(lat, amplitude)) +
+  geom_line(aes(color = factor(k))) +
+  scale_color_brewer("Wave-number", palette = "Set1") +
+  scale_y_continuous("Amplitude") +
+  scale_x_latitude(ticks = 15) +
+  coord_flip() +
+  facet_grid(term~lev, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free_x") +
+  tag_facets("tr")
+```
+
+The amplitude of each zonal wave number at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the zonal mean.
+Comparing between rows, this Figure quantifies the relatively clean separation between the zonally symmetric and zonally asymmetric structures, as its evident how the mixture of waves of the Full field (first row) is very similar to the sum of the waves of the Asymmetric and Symmetric field (second and third row, respectively).
+The second row of Figure \ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b), while in the stratosphere it is composed of zonal waves 3 to 1 in decreasing level of importance.
 
 ```{r vertical-regression, fig.cap = "Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South.
-Values are standarised by the standard deviation at each level.
 (this caption needs some love)", fig.width = 4, fig.height=3}
@@
   normalise_coords() %>% 
-  .[, .(hgt = mean(hgt)), by = .(lon, lev, time)] %>% 
+  .[, .(hgt = weighted.mean(hgt, w = cos(lat*pi/180))), by = .(lon, lev, time)] %>% 
   .[, hgt := hgt/9.8] %>% 
@@
   .[term == "asym"] %>% 
-  .[, estimate := estimate/sd(estimate), by = .(lev, term)] %>%
+  # .[, estimate := estimate/sd(estimate), by = .(lev, term)] %>%
   ggplot(aes(lon, lev)) +
-  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, 0.5, 0)) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(15), 
+  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, 10, 0)) +
+  geom_contour_tanaka(aes(z = estimate),
+                      breaks = AnchorBreaks(0, 10, 0),
+                      size = 0.3,
+                      global.breaks = FALSE) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(18), 
                        oob = scales::squish,
-                       breaks = MakeBreaks(binwidth = 0.5, exclude = 0)) +
+                       breaks = MakeBreaks(binwidth = 10, exclude = 0)) +
   scale_y_level() +
-  scale_x_longitude() 
+  # geom_hline(yintercept = 300) +
+  scale_x_longitude(ticks = 60) 
 ```
+
+From Figure \ref{fig:2d-regr} it appears that the vertical structure of the Asymmetric SAM is equivalent barotropic in the troposphere but baroclinic in the stratosphere.
+Anomalies are centerd in the same locations in the troposphere (panels h and k), but show westerly displacement in the stratosphere (panesl b and e).
+This is expanded in Figure \ref{fig:vertical-regression}, which shows a vertical crossection of the regression coefficient corresponding to the middle row of Figure \ref{fig:2d-regr}, area-weighted averaged between 65 and 40 degrees South.
+Below 100 hPa, anomalies are completely vertical, while above they show an important westerly tilt with height.
 
@@
   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  facet_tags() +
+  tag_facets() +
   axis_labs_smol
-```
-
-```{r wave-amplitude-700, fig.cap = "Planteray wave amplitude for the regression patterns at 700 hPa.", fig.width=6, fig.height=4}
-indices_2d %>% 
-  .[ lev == 700 & lat < -20] %>%
-  .[, FitWave(estimate, 0:3), by = .(lat, lev, season, term)] %>% 
-  ggplot(aes(lat, amplitude)) +
-  geom_line(aes(color = factor(k))) +
-  scale_color_brewer("Wave-number", palette = "Set1") +
-  scale_y_continuous("Amplitude") +
-  scale_x_latitude(ticks = 15) +
-  coord_flip() +
-  facet_grid(term~season, labeller = labeller(term = lab_sam), scales = "free_x") +
-  facet_tags("tr")
 ```
@@
   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  facet_tags() +
+  tag_facets() +
   axis_labs_smol
@@
 
-```{r wave-amplitude-30, fig.cap = "Planteray wave amplitude for the regression patterns at 30 hPa.", fig.width=6, fig.height=4}
-indices_2d %>% 
-  .[ lev == 30 & lat < -20] %>%
-  .[, FitWave(estimate, 0:3), by = .(lat, lev, season, term)] %>% 
-  ggplot(aes(lat, amplitude)) +
-  geom_line(aes(color = factor(k))) +
-  scale_color_brewer("Wave-number", palette = "Set1") +
-  scale_y_continuous("Squared amplitude") +
-  scale_x_latitude(ticks = 15) +
-  coord_flip() +
-  facet_grid(term~season, labeller = labeller(term = lab_sam), scales = "free_x") +
-  facet_tags("tr")
-```
+### Impacts
 
@@
 
-```
-
-```{r pp-regr-global,  fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
-pp_regr_season %>%
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.25, 0)) +
-  geom_contour2(aes(z = p.value), breaks = 0.01) +
-  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
-  stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
-  geom_qmap(~.x[lat <= -20]) +
-  scale_x_longitude() +
-  scale_y_latitude(labels = NULL) +
-  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
-                       breaks = AnchorBreaks(0, 0.25, 0)) +
-  coord_polar() +
-  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  facet_tags() +
-  axis_labs_smol
 ```
@@
   .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
   ggplot(aes(lon, lat)) +
@@
   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  facet_tags() +
+  tag_facets() +
+  axis_labs_smol
+```
+
+```{r pp-regr-global,  fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
+pp_regr_season %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.25, 0)) +
+  geom_contour2(aes(z = p.value), breaks = 0.01) +
+  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
+  stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
+  geom_qmap(~.x[lat <= -20]) +
+  scale_x_longitude() +
+  scale_y_latitude(labels = NULL) +
+  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
+                       breaks = AnchorBreaks(0, 0.25, 0)) +
+  coord_polar() +
+  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  tag_facets() +
   axis_labs_smol
@@
 \bibliography{AsymSAM}
+
+\newpage
 
@@
                                                   term = lab_sam)) +
-  facet_tags() 
+  tag_facets() 
 ```
@@
   facet_grid(lev ~ term,labeller = labeller(lev = lev.lab, term = lab_sam), scales = "free") +
-  facet_tags() 
+  tag_facets() 
 ```
@@
   facet_grid(term ~ lev,labeller = labeller(lev = lev.lab, term = lab_sam)) +
-  facet_tags("tr") 
+  tag_facets("tr") 
 ```
@@
   facet_grid(term ~ .,labeller = labeller(term = lab_sam)) +
-  facet_tags() 
+  tag_facets() 
 ```
@@
   facet_grid(season~term, labeller = labeller(term = lab_sam)) +
-  facet_tags() +
+  tag_facets() +
   axis_labs_smol 
@@
   facet_grid(term ~ regression, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  facet_tags() +
+  tag_facets() +
   axis_labs_smol 
 ```
+
+```{r A7, fig.cap = " "}
+regr_dif %>% 
+    rm_intercept() %>% 
+  .[lat < -20] %>%
+  # .[lev %in% c(50, 700)] %>% 
+  .[, FitWave(estimate, 0:3), by = .(lat, regression, term)] %>% 
+  ggplot(aes(lat, amplitude)) +
+  geom_line(aes(color = factor(k))) +
+  scale_color_brewer("Wave-number", palette = "Set1") +
+  scale_y_continuous("Amplitude") +
+  scale_x_latitude(ticks = 15) +
+  coord_flip() +
+  facet_grid(term~regression, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free_x") +
+  tag_facets("tr")
+```

### 2020-07-16 analysis/paper/paper.tex
more results

@@
 
+\subsection{Temporal evolution}
+
 \begin{figure*}
@@
 
-Figure\textasciitilde{}\ref{fig:asymsam-timeseries} shows the resulting
-Asymmetric and Symmetric time series corresponding to 700 and 30hPa.
-blablababla
+Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and
+Symmetric time series corresponding to 700 and 30hPa. blablababla
 
@@
 constant throught the troposphere, fluctiating between 0.39 and 0.45
-(Figure\textasciitilde{}\ref{fig:cor-lev}). Futhermore, the
-cross-correlation of each series across levels --shown in
-Figure\textasciitilde{}\ref{fig:cross-correlation}-- are high in the
-trosposphere (greater than 0.9) for both indices. This suggests that
-both the Asymmetric and the Symemtric component of the tropospheric SAM
-are highly vertically coherent, both in their individual evolution and
-their temporal relationship. This is to be expected since the SAM is
-mostly equivalent barotropic (citaaaa).
+(Figure \ref{fig:cor-lev}). Futhermore, the cross-correlation of each
+series across levels --shown in Figure \ref{fig:cross-correlation}-- are
+high in the trosposphere (greater than 0.9) for both indices. This
+suggests that both the Asymmetric and the Symemtric component of the
+tropospheric SAM are highly vertically coherent, both in their
+individual evolution and their temporal relationship. This is to be
+expected since the SAM is mostly equivalent barotropic (citaaaa).
 
-In the stratosphere the situation is different. As can be seen in
-Figure\textasciitilde{}\ref{fig:cor-lev}, the relationship between the
-Asymmetric and Symmetric indices varies with height above 100 hPa. It
-starts to decreese right over the tropopause, reaches a minium of 0.21
-at 20 hPa and then it increases again monotonically with height up to
-the uppermost level of the reanalysis. The cross-correlation across
-levels in the stratosphere is generally weaker than in the troposphere
-(Figure\textasciitilde{}\ref{fig:cross-correlation}). Furthermore, above
-100 hPa, the cross-correlation decreases more rapdily with height for
-the Symmetric SAM than for the Asymmetric SAM as evidenced by the wider
-dark red areas near the diagonal in
-Figure\textasciitilde{}\ref{fig:cross-correlation}b)
-vs.~Figure\textasciitilde{}\ref{fig:cross-correlation}c). Moreover, the
-stratospheric Symmetric SAM seems to be slightly more connected to the
-trosposphere than the Asymmetric SAM; this can be seen by the lower
-correlation values in the top right quadrant of
-Figure\textasciitilde{}\ref{fig:cross-correlation}b) in comparison with
-Figure\textasciitilde{}\ref{fig:cross-correlation}c).
+In the stratosphere the situation is different. As can be seen in Figure
+\ref{fig:cor-lev}, the relationship between the Asymmetric and Symmetric
+indices varies with height above 100 hPa. It starts to decreese right
+over the tropopause, reaches a minium of 0.21 at 20 hPa and then it
+increases again monotonically with height up to the uppermost level of
+the reanalysis. The cross-correlation across levels in the stratosphere
+is generally weaker than in the troposphere (Figure
+\ref{fig:cross-correlation}). Furthermore, above 100 hPa, the
+cross-correlation decreases more rapdily with height for the Symmetric
+SAM than for the Asymmetric SAM as evidenced by the wider dark red areas
+near the diagonal in Figure \ref{fig:cross-correlation}b) vs.~Figure
+\ref{fig:cross-correlation}c). Moreover, the stratospheric Symmetric SAM
+seems to be slightly more connected to the trosposphere than the
+Asymmetric SAM; this can be seen by the lower correlation values in the
+top right quadrant of Figure \ref{fig:cross-correlation}b) in comparison
+with Figure \ref{fig:cross-correlation}c).
 
@@
 
-Figure\textasciitilde{}\ref{fig:cross-correlation}a) show the
-cross-correlation across levels for the Full SAM index. \ldots{}..
+Figure \ref{fig:cross-correlation}a) show the cross-correlation across
+levels for the Full SAM index. \ldots{}..
+
+\subsection{Spatial patterns}
 
@@
 
-Figure\textasciitilde{}\ref{fig:2d-regr} shows the spatial year-long
-regression for selected levels. In the troposphere the Full annular mode
-is clearly ``contaminated'' with well known zonal asymemtries (panels g
-and j) which are succesfully sepparated by our methodology (panels h, i,
-k and l). In the stratosphere, the spatial pattern associated with the
-Full SAM is much more clearly dominated by a zonally symmetric,
-monopolar structure (panels a and d) that is, however, not perfectly
-centered in the south pole. The monopoloe obtained by multiple
-regression with the Asymmetric and Symmetric SAM (panels c and f in
-Figure\textasciitilde{}\ref{fig:2d-regr}) is much more symmetric and the
-shift from total symmetry is captured by the regression pattern of the
-Asymmetric SAM as a wave-1 pattern (panels b and e).
+Figure \ref{fig:2d-regr} shows the spatial year-long regression for
+selected levels. In the troposphere the Full annular mode is clearly
+``contaminated'' with well known zonal asymemtries (panels g and j)
+which are succesfully sepparated by our methodology (panels h, i, k and
+l). In the stratosphere, the spatial pattern associated with the Full
+SAM is much more clearly dominated by a zonally symmetric, monopolar
+structure (panels a and d) that is, however, not perfectly centered in
+the south pole. The monopoloe obtained by multiple regression with the
+Asymmetric and Symmetric SAM (panels c and f in Figure
+\ref{fig:2d-regr}) is much more symmetric and the shift from total
+symmetry is captured by the regression pattern of the Asymmetric SAM as
+a wave-1 pattern (panels b and e).
+
+\begin{figure}
+\includegraphics{wave-amplitude-1} \caption[Planteray wave amplitude for the regression patterns at 50 and 700 hPa]{Planteray wave amplitude for the regression patterns at 50 and 700 hPa.}\label{fig:wave-amplitude}
+\end{figure}
+
+The amplitude of each zonal wave number at each latitude at 50 hPa and
+700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number
+zero represents the zonal mean. Comparing between rows, this Figure
+quantifies the relatively clean separation between the zonally symmetric
+and zonally asymmetric structures, as its evident how the mixture of
+waves of the Full field (first row) is very similar to the sum of the
+waves of the Asymmetric and Symmetric field (second and third row,
+respectively). The second row of Figure \ref{fig:wave-amplitude} shows
+that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the
+stratosphere (panel b), while in the stratosphere it is composed of
+zonal waves 3 to 1 in decreasing level of importance.
 
@@
 \includegraphics{vertical-regression-1} \caption[Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South]{Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South.
-Values are standarised by the standard deviation at each level.
 (this caption needs some love)}\label{fig:vertical-regression}
 \end{figure}
+
+From Figure \ref{fig:2d-regr} it appears that the vertical structure of
+the Asymmetric SAM is equivalent barotropic in the troposphere but
+baroclinic in the stratosphere. Anomalies are centerd in the same
+locations in the troposphere (panels h and k), but show westerly
+displacement in the stratosphere (panesl b and e). This is expanded in
+Figure \ref{fig:vertical-regression}, which shows a vertical crossection
+of the regression coefficient corresponding to the middle row of Figure
+\ref{fig:2d-regr}, area-weighted averaged between 65 and 40 degrees
+South. Below 100 hPa, anomalies are completely vertical, while above
+they show an important westerly tilt with height.
 
@@
 
-\begin{figure}
-\includegraphics{wave-amplitude-700-1} \caption[Planteray wave amplitude for the regression patterns at 700 hPa]{Planteray wave amplitude for the regression patterns at 700 hPa.}\label{fig:wave-amplitude-700}
-\end{figure}
-
 \begin{sidewaysfigure}
@@
 
-\begin{figure}
-\includegraphics{wave-amplitude-30-1} \caption[Planteray wave amplitude for the regression patterns at 30 hPa]{Planteray wave amplitude for the regression patterns at 30 hPa.}\label{fig:wave-amplitude-30}
-\end{figure}
+\subsubsection{Impacts}
+
+\begin{sidewaysfigure}
+\includegraphics{regr-air-season-1} \caption[Regression pattern of surface temperature with Asymmetric and Symmetric SAM]{Regression pattern of surface temperature with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:regr-air-season}
+\end{sidewaysfigure}
 
@@
 P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:pp-regr-global}
-\end{sidewaysfigure}
-
-\begin{sidewaysfigure}
-\includegraphics{regr-air-season-1} \caption[Regression pattern of surface temperature with Asymmetric and Symmetric SAM]{Regression pattern of surface temperature with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:regr-air-season}
 \end{sidewaysfigure}
@@
 
+\begin{figure}
+\includegraphics{A7-1} \caption[ ]{ }\label{fig:A7}
+\end{figure}
+
 \end{document}

### 2020-07-16 R/ggplot-tags.R

@@ -1,7 +1,8 @@
 #' Adds tags to facets
 #'
 #' @export
-facet_tags <- function(position = "tl", tag_pool = letters, open = "", close = ")", stroke = 0.15, stroke.color = "white",
+tag_facets <- function(position = "tl", tag_levels = "a", tag_pool = NULL, tag_prefix = "",
+                       tag_suffix = ")", stroke = 0.15, stroke.color = "white",
                  color = "black") {
   if (!is.list(position)) {
     position <- switch(position,
@@ -13,7 +14,8 @@ facet_tags <- function(position = "tl", tag_pool = letters, open = "", close = "
   }
 
 
-  structure(list(tag_pool = tag_pool, open = open, close = close, position = position,
+  structure(list(tag_pool = tag_pool, tag_levels = tag_levels, open = tag_prefix,
+                 close = tag_suffix, position = position,
                  color = color,
                  stroke = stroke, stroke.color = stroke.color),
             class = "ggtags")
@@ -40,25 +42,43 @@ print.ggtagged <- function(x, newpage = is.null(vp), vp = NULL, ...) {
   )
 
   tag_options <- attr(x, "tag_options")
+
   gb <- ggplot2::ggplot_build(x)
   lay <- gb$layout$layout
 
-  x = tag_options$position$x
-  y = tag_options$position$y
+  x <-  tag_options$position$x
+  y <-  tag_options$position$y
 
-  hjust = tag_options$position$hjust
-  vjust = tag_options$position$vjust
+  hjust <- tag_options$position$hjust
+  vjust <- tag_options$position$vjust
 
   gt <- ggplot2::ggplot_gtable(gb)
   panels <- which(grepl("panel", gt$layout$name))
+
+  i_panels <- seq_along(panels)
+
+  if (is.null(tag_options$tag_pool)) {
+   tag_pool <-  switch(tag_options$tag_levels,
+                       a = letters[i_panels],
+                       A = LETTERS[i_panels],
+                       "1" = i_panels,
+                       "i" = tolower(utils::as.roman(i_panels)),
+                       "I" = utils::as.roman(i_panels),
+                       stop("tag_levels is not valis")
+                       )
+
+  } else {
+    tag_pool <- tag_options$tag_pool
+  }
+
   for (p in seq_along(panels)) {
 
     if (tag_options$stroke == 0) {
-      tagGrob <- grid::textGrob(paste0(tag_options$open, tag_options$tag_pool[p], tag_options$close),
[…]

### 2020-07-21 analysis/AGU/agu-abstract.Rmd
more results and agu

---
title: "Teasing out the asymmetric component of the Southern Annular Mode"
output: pdf_document
---

The conception of the Southern Annular Mode as a mostly zonally symmetric pattern is reflected not only in its name, but also on the various methods used to characterise it. Gong and Wang (1999) and many others define the SAM index based on zonal mean pressure anomalies. Baldwind & Thompson also put forward an index based on performin EOF on zonal mean geopotential height. However, composites and regression patterns based on these indices consistently result in zonally assymetric patterns partially embebed in the zonally symmetric structure. Citar a @fogt2012 de alguna manera? Here, we propose a straightforward method of creating two sister indices of Symemtric and Asymmetric SAM and show that they can be used to separate between the zonally symmetric and asymmetric structure. We show that the observed trends towards more positive SAM [@folgt2020] is completely explained by the Symmetric component, but that the relationship between SAM and surface variables such as temperature and precipitation is associated more strongly with the Asymmetric component, particularly over the Southern continents.

```{r eval=FALSE, include=FALSE}
nchar(gsub(" ", "", "The conception of the Southern Annular Mode as a mostly zonally symmetric pattern is reflected not only in its name, but also on the various methods used to characterise it. Gong and Wang (1999) and many others define the SAM index based on zonal mean pressure anomalies. Baldwind & Thompson also put forward an index based on performin EOF on zonal mean geopotential height. However, composites and regression patterns based on these indices consistently result in zonally assymetric patterns partially embebed in the zonally symmetric structure. Citar a @fogt2012 de alguna manera? Here, we propose a straightforward method of creating two sister indices of Symemtric and Asymmetric SAM and show that they can be used to separate between the zonally symmetric and asymmetric structure. We show that the observed trends towards more positive SAM [@folgt2020] is completely explained by the Symmetric component, but that the relationship between SAM and surface variables such as temperature and precipitation is associated more strongly with the Asymmetric component, particularly over the Southern continents."))
```

### 2020-07-21 analysis/paper/paper.Rmd
more results and agu

@@
   cache = TRUE,
-  cache.extra = 42,
+  cache.extra = 41,
   cache.path = "../cache/",
@@
 library(latex2exp)
+library(tagger)
 
@@
   indices <- readRDS(indices_file)
+  indices[, `:=`(estimate = -estimate, estimate_norm = -estimate_norm)]
 } else {
@@
   facet_grid(lev~., labeller = labeller(lev = lev.lab), scales = "free") +
-  tag_facets(stroke = 0.2)
+  tag_facets()
 ```
 
-Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and Symmetric time series corresponding to 700 and 30hPa. blablababla
+Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and Symmetric time series corresponding to 700 and 30hPa. blablababla #FIXME
 
@@
   theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside") +
-  tag_facets("br", stroke = 0.2)
+  tag_facets(position = "br")
 ```
 
-Figure \ref{fig:cross-correlation}a) show the cross-correlation across levels for the Full SAM index. .....
+Figure \ref{fig:cross-correlation}a) show the cross-correlation across levels for the Full SAM index. #FIXME
+
+<!-- ```{r} -->
+<!-- indices_wide %>%  -->
+<!--   .[, FitLm(as.numeric(time), asym, sym, se = TRUE), by = .(lev)] %>%  -->
+<!--   rm_intercept() %>%  -->
+<!--   # .[, term := NULL] %>%  -->
+<!--   .[, estimate := estimate/(3600*24*365*10)] %>% -->
+<!--   .[, std.error := std.error/(3600*24*365*10)] %>% -->
+<!--   # setnames("type", "term") %>%  -->
+<!--   ggplot(aes(lev, estimate)) + -->
+<!--   geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) + -->
+<!--   geom_line() + -->
+<!--   scale_fill_brewer(palette = "Set1", aesthetics = c("color", "fill")) + -->
+<!--   geom_hline(yintercept = 0) + -->
+<!--   scale_x_level() + -->
+<!--   scale_y_continuous("Standard deviations per decade") + -->
+<!--   coord_flip() + -->
+<!--   facet_grid(~ term,labeller = labeller(term = lab_sam)) + -->
+<!--   tag_facets()  -->
+<!-- ``` -->
+
+```{r trends, fig.cap = "Trends for each index at each level. Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=3}
+indices %>%
+  copy() %>% 
+  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
+  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
+  rm_intercept() %>% 
+  .[, term := NULL] %>% 
+  .[, estimate := estimate*3600*24*365*10] %>% 
+  .[, std.error := std.error*3600*24*365*10] %>% 
+  setnames("type", "term") %>% 
+  ggplot(aes(lev, estimate)) +
+  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) +
+  geom_line() +
+  scale_fill_brewer(palette = "Set1", aesthetics = c("color", "fill")) +
+  geom_hline(yintercept = 0) +
+  scale_x_level() +
+  scale_y_continuous("Standard deviations per decade") +
+  coord_flip() +
+  facet_grid(~ term,labeller = labeller(term = lab_sam)) +
+  tag_facets() 
+```
+
+Figure \ref{fig:trends} shows normalised decadal trends for each index for the whole period along with a 95% interval in shading.
+There is a statistically significant increase in positive SAM in the tropostphere (panel a), which has been already documented in other studies (e.g.
+@fogt2020).
+Panels b and c of Figure \ref{fig:trends} show that this increase is evident only in the Symmetric component.
+This distinction should prove useful when attributing trends in other variables such as temperature and precipitation to trends in the SAM.
+
+## Spatial patterns
 
@@
 
-## Spatial patterns
-
-To undertand the spatial patterns associated with both indeces we regressed monthly geopotential anomalies into both indeces using multiple regression (Figure A6 illustrates the difference between computing two simple regressions and one multiple regression).
+To undertand the spatial patterns associated with both indeces, we regressed monthly geopotential anomalies into both indeces using multiple regression (Figure A6 illustrates the difference between computing two simple regressions and one multiple regression).
 
@@
   facet_grid(term~lev, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free_x") +
-  tag_facets("tr")
+  tag_facets(position = "tr")
 ```
@@
 
-```{r compute-regr-pattern-season}
-indices_2d <- hgt[lev %in% main_levs] %>% 
-  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
-  .[, rbind(as.data.table(FitLm(hgt_a, sym, asym)),
-            as.data.table(FitLm(hgt_a, full))),
-    by = .(lon, lat, lev, season(time))] %>% 
-  rm_intercept() %>% 
-  periodic(lon = c(0, 360)) %>% 
-  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] 
+<!-- ```{r compute-regr-pattern-season} -->
+<!-- indices_2d <- hgt[lev %in% main_levs] %>%  -->
+<!--   indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>%  -->
+<!--   .[, rbind(as.data.table(FitLm(hgt_a, sym, asym)), -->
+<!--             as.data.table(FitLm(hgt_a, full))), -->
+<!--     by = .(lon, lat, lev, season(time))] %>%  -->
+<!--   rm_intercept() %>%  -->
+<!--   periodic(lon = c(0, 360)) %>%  -->
+<!--   .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)]  -->
+<!-- ``` -->
+
+<!-- ```{r 2d-regr-700, fig.cap = "Seasonal regression patterns of geopotential height at 700 hPa with the Full, Asymmetric and Symmetric SAM.
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"} -->
+<!-- indices_2d %>%  -->
+<!--   .[lat <= -20] %>%  -->
+<!--   .[lev == 700] %>%  -->
+<!--   ggplot(aes(lon, lat)) + -->
+<!--   geom_contour_fill(aes(z = estimate),  breaks = AnchorBreaks(0, binwidth = 5, 0), -->
+<!--                     global.breaks = FALSE) + -->
+<!--   geom_contour_tanaka(aes(z = estimate), -->
+<!--                       breaks = AnchorBreaks(0, binwidth = 5, 0), -->
+<!--                       size = 0.3, -->
+<!--                       global.breaks = FALSE) + -->
+<!--   geom_qmap(~.x[lat <= -20]) + -->
+<!--   scale_x_longitude() + -->
+<!--   scale_y_latitude(limits = c(NA, -20), labels = NULL) + -->
+<!--   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), -->
+<!--                        limits = c(-55, 55), -->
+<!--                        oob = scales::squish, -->
+<!--                        breaks = AnchorBreaks(0, binwidth = 5, 0)) + -->
+<!--   scale_linetype(guide = "none") + -->
+<!--   coord_polar() + -->
+<!--   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) + -->
+<!--   tag_facets() + -->
+<!--   axis_labs_smol -->
+<!-- ``` -->
+
+<!-- ```{r 2d-regr-30, fig.cap = "Seasonal regression patterns of geopotential height at 30 hPa with the Full, Asymmetric and Symmetric SAM.
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"} -->
+<!-- indices_2d %>%  -->
+<!--   .[lat <= -20] %>%  -->
+<!--   .[lev == 30] %>%  -->
+<!--   ggplot(aes(lon, lat)) + -->
+<!--   geom_contour_fill(aes(z = estimate),  breaks = AnchorBreaks(0, binwidth = 20, 0), -->
+<!--                     global.breaks = FALSE) + -->
+<!--   geom_contour_tanaka(aes(z = estimate), -->
+<!--                       breaks = AnchorBreaks(0, binwidth = 20, 0), -->
+<!--                       size = 0.3, -->
+<!--                       global.breaks = FALSE) + -->
+<!--   geom_qmap(~.x[lat <= -20]) + -->
+<!--   scale_x_longitude() + -->
+<!--   scale_y_latitude(limits = c(NA, -20), labels = NULL) + -->
+<!--   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), -->
+<!--                        limits = c(-105, 105), -->
+<!--                        oob = scales::squish, -->
+<!--                        breaks = AnchorBreaks(0, binwidth = 20, 0)) + -->
+<!--   scale_linetype(guide = "none") + -->
+<!--   coord_polar() + -->
+<!--   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) + -->
+<!--   tag_facets() + -->
+<!--   axis_labs_smol -->
+<!-- ``` -->
+
+## Impacts
+
+### Temperature
+
+```{r compute-air-regr}
+time_subset <- list(time = range(indices$time))
+air <- NOAATemp() %>% 
+  ReadNetCDF(vars = "air", subset = c(time_subset, list(lat = c(-90, 10)))) %>% 
+  normalise_coords()  %>%
+  # Interpolate at lon = 0. Otherwise, can't close contours. 
+  qwrap(lon = c(0, 360) ~ c(-2.5, 360)) %>% 
+  .[ , Interpolate(air ~ lon + lat, x.out = c(0, unique(lon)), y.out = unique(lat)), by = time] %>% 
+  .[lon != -2.5]
+
+air_regr <- air %>% 
+  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
+  .[indices_wide[lev == 700], on = .NATURAL] %>% 
+  rm_singleton() %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
+  .[, nas := mean(is.na(air)), by = .(lon, lat, season(time))] %>% 
+  .[nas >= 0.15, air := NA] %>% 
+  .[, rbind(as.data.table(FitLm(air, asym, sym, se = TRUE)),
+            as.data.table(FitLm(air, full, se = TRUE))),
+    by = .(lon, lat, season(time))] %>% 
+  rm_intercept() 
 ```
 
-```{r 2d-regr-700, fig.cap = "Seasonal regression patterns of geopotential height at 700 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
-indices_2d %>% 
-  .[lat <= -20] %>% 
-  .[lev == 700] %>% 
+```{r regr-air-season, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
+Gray areas have more than 15\\% of missing data.", fig.width = 10, fig.height = 7, fig.env = "sidewaysfigure"}
+air_regr %>% 
+  copy() %>% 
+  # .[lat <= -61, estimate := NA] %>%
+  .[lat >= 65, estimate := NA] %>%
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
+  periodic(lon = c(0, 360)) %>% 
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate),  breaks = AnchorBreaks(0, binwidth = 5, 0),
-                    global.breaks = FALSE) +
-  geom_contour_tanaka(aes(z = estimate),
-                      breaks = AnchorBreaks(0, binwidth = 5, 0),
-                      size = 0.3,
-                      global.breaks = FALSE) +
-  geom_qmap(~.x[lat <= -20]) +
+  geom_contour_fill(aes(z = estimate), na.fill = 0,
+                    breaks = AnchorBreaks(0, 0.15, 0)) +
+  # geom_raster(aes(fill = estimate)) +
+  geom_contour2(aes(z = p.value), breaks = 0.05) +
+  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 0.5, fill = "gray", color = "gray30") +
+  geom_qmap(~.x[lat <= 0]) +
   scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
-                       limits = c(-55, 55),
-                       oob = scales::squish,
-                       breaks = AnchorBreaks(0, binwidth = 5, 0)) +
-  scale_linetype(guide = "none") +
+  scale_y_latitude(labels = NULL, limits = c(NA, 0)) +
+  scale_fill_divergent("Surface temperature",  limits = c(-0.8, .8), oob = scales::squish,
+                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.15, 0)) +
   coord_polar() +
   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets() +
+  tag_facets("cr", tag_suffix = "") +
   axis_labs_smol
@@
 
-```{r 2d-regr-30, fig.cap = "Seasonal regression patterns of geopotential height at 30 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
-indices_2d %>% 
-  .[lat <= -20] %>% 
-  .[lev == 30] %>% 
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate),  breaks = AnchorBreaks(0, binwidth = 20, 0),
-                    global.breaks = FALSE) +
-  geom_contour_tanaka(aes(z = estimate),
-                      breaks = AnchorBreaks(0, binwidth = 20, 0),
-                      size = 0.3,
-                      global.breaks = FALSE) +
-  geom_qmap(~.x[lat <= -20]) +
-  scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
-                       limits = c(-105, 105),
-                       oob = scales::squish,
-                       breaks = AnchorBreaks(0, binwidth = 20, 0)) +
-  scale_linetype(guide = "none") +
-  coord_polar() +
-  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets() +
-  axis_labs_smol
-```
+Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface temperature for each trimester.
+It is evident that the Asymmetric and Symmetric SAM indices are associated with overall distinct temperature patterns which can be obscured when using the Full SAM index.
+The Symmetric SAM signal is weaker than the Asymmetric SAM, as evidenced by the relatively smaller and les sstatistically significant regression coefficients in row 3 of Figure \ref{fig:regr-air-season} compared with row 2.
 
-### Impacts
+In DJF (column a), the strong negative signal in the tropical Pacific in panel a.1 is mostly associated with the Asymmetric component (panel a.2), as is it largely absent in the Symmetric component (panel a.3).
+Furthermore, the Asymmetric SAM is also asociated with low temperature anomalies in the Indian ocean, but this signal is obscured by the Symmetric variability and thus lost in the Full SAM.
+Over the continents, the Asymmetric SAM is assoiated with negative temperature anomalies which, again, mostly disappear in the Full SAM regression.
+
+The patterns seen in MAM and JJA (columns b and c) are not robustly significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following @wilks2016.
+Nevertheless, it is interesting to note that in both trimesters, the sign of the regression is consistently flipped between the Asymmetric and Symmetric regressions.
+In South America, for example, the Asymmetric SAM is associated with positive temperature anomalies in MAM and negative temperature anomales in JJA, while the oposite is the case for the Symemtric SAM.
+
+Finally, in SON (column d), there is no significant temperature signal associated with the Symmetric SAM (panel d.3), while the Asymmetric SAM shows a relatively robust signal in the equatorial Pacitic, Australia, and even Southeast South America.
+This strong signals are reduced in intensity in panel a.3.
+
+### Precipitation
 
@@
     by = .(lon, lat, season(time))] %>% 
-  rm_intercept() %>% 
-  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)]
+  rm_intercept()
 
-```
-
-```{r compute-air-regr}
-time_subset <- list(time = range(indices$time))
-air <- NOAATemp() %>% 
-  ReadNetCDF(vars = "air", subset = c(time_subset, list(lat = c(-90, 10)))) %>% 
-  normalise_coords()
-
-air_regr <- air %>% 
-  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
-  .[indices_wide[lev == 700], on = .NATURAL] %>% 
-  rm_singleton() %>% 
-  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
-  .[, nas := mean(is.na(air)), by = .(lon, lat, season(time))] %>% 
-  .[nas >= 0.15, air := NA] %>% 
-  .[, rbind(as.data.table(FitLm(air, asym, sym, se = TRUE)),
-            as.data.table(FitLm(air, full, se = TRUE))),
-    by = .(lon, lat, season(time))] %>% 
-  rm_intercept() 
-```
-
-```{r regr-air-season, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", fig.width = 10, fig.height = 7, fig.env = "sidewaysfigure"}
-air_regr %>% 
-  copy() %>% 
-  .[lat <= -61, estimate := NA] %>%
-  .[lat >= 65, estimate := NA] %>%
-  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
-  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), na.fill = 0,
-                    breaks = AnchorBreaks(0, 0.15, 0)) +
-  # geom_raster(aes(fill = estimate)) +
-  geom_contour2(aes(z = p.value), breaks = 0.05) +
-  stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat)), size = 0.4, alpha = 0.8) +
-  # stat_subset(aes(subset = is.na(estimate)), geom = "raster", fill = "gray50") +
-  geom_qmap(~.x[lat <= 10]) +
-  scale_x_longitude() +
-  scale_y_latitude(labels = NULL) +
-  scale_fill_divergent("Surface temperature",  limits = c(-0.8, .8), oob = scales::squish,
-                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.15, 0)) +
-  coord_polar() +
-  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets() +
-  axis_labs_smol
 ```
@@
 pp_regr_season %>%
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)] %>% 
   ggplot(aes(lon, lat)) +
@@
   facet_grid(term ~ lev,labeller = labeller(lev = lev.lab, term = lab_sam)) +
-  tag_facets("tr") 
-```
-
-```{r A4, fig.cap = "Trends for each index at each level. Shading indicates the 95\\% confidence interval.", fig.width = 3, fig.height=5, appendix = TRUE}
-indices %>%
-  copy() %>% 
-  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
-  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
-  rm_intercept() %>% 
-  .[, term := NULL] %>% 
-  .[, estimate := estimate*3600*24*365*10] %>% 
-  .[, std.error := std.error*3600*24*365*10] %>% 
-  setnames("type", "term") %>% 
-  ggplot(aes(lev, estimate)) +
-  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96, fill = term), alpha = 0.3) +
-  geom_line(aes(color = term)) +
-  scale_fill_brewer(palette = "Set1", aesthetics = c("color", "fill")) +
-  geom_hline(yintercept = 0) +
-  scale_x_level() +
-  coord_flip() +
-  facet_grid(term ~ .,labeller = labeller(term = lab_sam)) +
-  tag_facets() 
+  tag_facets(position = "tr") 
 ```
@@
 
-```{r A7, fig.cap = " "}
+```{r A7, appendix = TRUE, fig.cap = " "}
 regr_dif %>% 
@@
   facet_grid(term~regression, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free_x") +
-  tag_facets("tr")
+  tag_facets(position = "tr")
 ```
+
+```{r A8, appendix = TRUE, fig.cap = "Decadal trends of SAM indices for each season. "}
+indices %>% 
+  copy() %>% 
+  .[, season := season(time)] %>% 
+  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
+  .[, FitLm(estimate_norm, time, se = TRUE), by = .(type = term, lev, season)] %>% 
+  rm_intercept() %>% 
+  .[, term := NULL] %>% 
+  .[, estimate := estimate*3600*24*365*10] %>% 
+  .[, std.error := std.error*3600*24*365*10] %>% 
+  setnames("type", "term") %>% 
+  ggplot(aes(lev, estimate)) +
+  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) +
+  geom_line() +
+  scale_fill_brewer(palette = "Set1", aesthetics = c("color", "fill")) +
+  geom_hline(yintercept = 0) +
+  scale_x_level() +
+  scale_y_continuous("Standard deviations per decade") +
+  coord_flip() +
+  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
+  tag_facets() 
+```

### 2020-07-21 analysis/paper/paper.tex
more results and agu

@@
 Symmetric time series corresponding to 700 and 30hPa. blablababla
+\#FIXME
 
@@
 Figure \ref{fig:cross-correlation}a) show the cross-correlation across
-levels for the Full SAM index. \ldots{}..
+levels for the Full SAM index. \#FIXME
+
+\begin{figure}
+\includegraphics{trends-1} \caption[Trends for each index at each level]{Trends for each index at each level. Shading indicates the 95\% confidence interval.}\label{fig:trends}
+\end{figure}
+
+Figure \ref{fig:trends} shows normalised decadal trends for each index
+for the whole period along with a 95\% interval in shading. There is a
+statistically significant increase in positive SAM in the tropostphere
+(panel a), which has been already documented in other studies (e.g.
+\citet{fogt2020}). Panels b and c of Figure \ref{fig:trends} show that
+this increase is evident only in the Symmetric component. This
+distinction should prove useful when attributing trends in other
+variables such as temperature and precipitation to trends in the SAM.
 
@@
 
-To undertand the spatial patterns associated with both indeces we
+To undertand the spatial patterns associated with both indeces, we
 regressed monthly geopotential anomalies into both indeces using
@@
 
-\begin{sidewaysfigure}
-\includegraphics{2d-regr-700-1} \caption[Seasonal regression patterns of geopotential height at 700 hPa with the Full, Asymmetric and Symmetric SAM]{Seasonal regression patterns of geopotential height at 700 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.}\label{fig:2d-regr-700}
-\end{sidewaysfigure}
+\subsection{Impacts}
 
-\begin{sidewaysfigure}
-\includegraphics{2d-regr-30-1} \caption[Seasonal regression patterns of geopotential height at 30 hPa with the Full, Asymmetric and Symmetric SAM]{Seasonal regression patterns of geopotential height at 30 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.}\label{fig:2d-regr-30}
-\end{sidewaysfigure}
-
-\subsubsection{Impacts}
+\subsubsection{Temperature}
 
@@
 \includegraphics{regr-air-season-1} \caption[Regression pattern of surface temperature with Asymmetric and Symmetric SAM]{Regression pattern of surface temperature with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:regr-air-season}
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
+Gray areas have more than 15\% of missing data.}\label{fig:regr-air-season}
 \end{sidewaysfigure}
+
+Figure \ref{fig:regr-air-season} shows regression coefficients of each
+index at 700 hPa with surface temperature for each trimester. It is
+evident that the Asymmetric and Symmetric SAM indices are associated
+with overall distinct temperature patterns which can be obscured when
+using the Full SAM index. The Symmetric SAM signal is weaker than the
+Asymmetric SAM, as evidenced by the relatively smaller and les
+sstatistically significant regression coefficients in row 3 of Figure
+\ref{fig:regr-air-season} compared with row 2.
+
+In DJF (column a), the strong negative signal in the tropical Pacific in
+panel a.1 is mostly associated with the Asymmetric component (panel
+a.2), as is it largely absent in the Symmetric component (panel a.3).
+Furthermore, the Asymmetric SAM is also asociated with low temperature
+anomalies in the Indian ocean, but this signal is obscured by the
+Symmetric variability and thus lost in the Full SAM. Over the
+continents, the Asymmetric SAM is assoiated with negative temperature
+anomalies which, again, mostly disappear in the Full SAM regression.
+
+The patterns seen in MAM and JJA (columns b and c) are not robustly
+significant in the sense that there are no areas with p-values below
+0.05 when controlling for FDR following \citet{wilks2016}. Nevertheless,
+it is interesting to note that in both trimesters, the sign of the
+regression is consistently flipped between the Asymmetric and Symmetric
+regressions. In South America, for example, the Asymmetric SAM is
+associated with positive temperature anomalies in MAM and negative
+temperature anomales in JJA, while the oposite is the case for the
+Symemtric SAM.
+
+Finally, in SON (column d), there is no significant temperature signal
+associated with the Symmetric SAM (panel d.3), while the Asymmetric SAM
+shows a relatively robust signal in the equatorial Pacitic, Australia,
+and even Southeast South America. This strong signals are reduced in
+intensity in panel a.3.
+
+\subsubsection{Precipitation}
 
@@
 \begin{figure}
-\includegraphics{A4-1} \appendcaption{A4}{Trends for each index at each level. Shading indicates the 95\% confidence interval.}\label{fig:A4}
-\end{figure}
-
-\begin{figure}
 \includegraphics{A5-1} \appendcaption{A5}{Regression pattern of precipiation with Asymmetric and Symmetric SAM.

### 2020-07-21 R/ggplot-tags.R

@@ -1,107 +1,0 @@
-#' Adds tags to facets
-#'
-#' @export
-tag_facets <- function(position = "tl", tag_levels = "a", tag_pool = NULL, tag_prefix = "",
-                       tag_suffix = ")", stroke = 0.15, stroke.color = "white",
-                 color = "black") {
-  if (!is.list(position)) {
-    position <- switch(position,
-                       tl = list(x = 0.05, y = 0.95, hjust = 0, vjust = 1),
-                       tr = list(x = 0.95, y = 0.95, hjust = 1, vjust = 1),
-                       bl = list(x = 0.05, y = 0.05, hjust = 0, vjust = 0),
-                       br = list(x = 0.95, y = 0.05, hjust = 1, vjust = 0),
-                       position)
-  }
-
-
-  structure(list(tag_pool = tag_pool, tag_levels = tag_levels, open = tag_prefix,
-                 close = tag_suffix, position = position,
-                 color = color,
-                 stroke = stroke, stroke.color = stroke.color),
-            class = "ggtags")
-}
-
-#' @export
-ggplot_add.ggtags <- function(object, plot, object_name) {
-  class(plot) <- c("ggtagged", class(plot))
-  attr(plot, "tag_options") <- object
-  plot
-}
-
-#' @export
-print.ggtagged <- function(x, newpage = is.null(vp), vp = NULL, ...) {
-  ggplot2::set_last_plot(x)
-  if (newpage) grid::grid.newpage()
-
-  # Record dependency on 'ggplot2' on the display list
-  # (AFTER grid.newpage())
-  grDevices::recordGraphics(
-    requireNamespace("ggplot2", quietly = TRUE),
-    list(),
-    getNamespace("ggplot2")
-  )
-
-  tag_options <- attr(x, "tag_options")
-
-  gb <- ggplot2::ggplot_build(x)
-  lay <- gb$layout$layout
-
-  x <-  tag_options$position$x
-  y <-  tag_options$position$y
-
-  hjust <- tag_options$position$hjust
-  vjust <- tag_options$position$vjust
-
-  gt <- ggplot2::ggplot_gtable(gb)
-  panels <- which(grepl("panel", gt$layout$name))
-
-  i_panels <- seq_along(panels)
-
-  if (is.null(tag_options$tag_pool)) {
-   tag_pool <-  switch(tag_options$tag_levels,
-                       a = letters[i_panels],
-                       A = LETTERS[i_panels],
-                       "1" = i_panels,
[…]

### 2020-07-21 R/theme.R

@@ -12,6 +12,8 @@ theme_asymsam <- function(base_size = 16) {
       # strip.background = ggplot2::element_rect(fill = NA, color = "gray30"),
       # text = element_text(family = font_rc),
       legend.position = "bottom", legend.box = "vertical",
+      tagger.panel.tag.background = ggplot2::element_rect(color = NA),
+      strip.text = element_text(size = 12),
       # panel.spacing.y = grid::unit(5, "mm"),
       # panel.spacing.x = grid::unit(5, "mm"),
       # legend.spacing = grid::unit(2, "mm"),

### 2020-07-21 analysis/AGU/agu-abstract.R

@@ -1,0 +1,8 @@
+---
+title: "Teasing out the asymmetric component of the Southern Annular Mode"
+output: pdf_document
+---
+
+The conception of the Southern Annular Mode as a mostly zonally symmetric pattern is reflected not only in its name, but also on the various methods used to characterise it. Gong and Wang (1999) and many others define the SAM index based on zonal mean pressure anomalies. Baldwind & Thompson also put forward an index based on performin EOF on zonal mean geopotential height. However, composites, as well as regression patterns, based on these indinces un
+[@fogt2012 ]
+. Here, we propose a straightforward method of 

### 2020-07-28 analysis/paper/paper.Rmd
render figuras

@@
   cache = TRUE,
-  cache.extra = 41,
+  cache.extra = 42,
   cache.path = "../cache/",
@@
 This distinction should prove useful when attributing trends in other variables such as temperature and precipitation to trends in the SAM.
+
+```{r}
+constants <- indices_wide %>% 
+  .[, FitLm(full, asym, sym), by = .(lev)] %>% 
+  rm_intercept() %>% 
+  setnames("estimate", "constant")
+
+constants %>% 
+  .[term == "asym"] %>% 
+  ggplot(aes(lev, constant)) +
+  geom_line() +
+  scale_x_level() +
+  coord_flip()
+
+asymmetry <- constants[indices, on = .NATURAL] %>% 
+  .[term == "full", constant := 1] %>%
+  # .[, estimate_norm := estimate_norm*constant] %>%
+  dcast(lev + time ~ term, value.var = "estimate_norm") %>% 
+  .[, asymmetry := atan(abs(asym)/abs(sym))/(pi/2)]
+
+asymmetry %>% 
+  .[, median(asymmetry), by = .(lev, month(time))] %>% 
+  ggplot(aes(month, lev)) +
+  geom_contour_fill(aes(z = V1)) +
+  scale_fill_divergent(midpoint = 0.5) +
+  scale_y_level()
+
+asymmetry %>% 
+  .[lev %in% main_levs] %>% 
+  ggplot(aes(time, asymmetry))  +
+  geom_line() +
+  facet_grid(lev~.)
+
+asymmetry %>% 
+  # .[lev %in% main_levs] %>% 
+  .[, FitLm(asymmetry, time, se = TRUE), by = lev] %>% 
+  rm_intercept() %>% 
+  # .[, term := NULL] %>% 
+  .[, estimate := estimate*3600*24*365*10] %>% 
+  .[, std.error := std.error*3600*24*365*10] %>% 
+  # setnames("type", "term") %>% 
+  ggplot(aes(lev, estimate)) +
+  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) +
+  geom_line() +
+  scale_fill_brewer(palette = "Set1", aesthetics = c("color", "fill")) +
+  geom_hline(yintercept = 0) +
+  scale_x_level() +
+  scale_y_continuous("Standard deviations per decade") +
+  coord_flip() +
+  tag_facets()
+
+```
 

### 2020-07-31 analysis/paper/paper.Rmd
Adds dependencies

@@
   # coord_polar() +
-  ggalt::coord_proj("+proj=moll +lon_0=-60") +
+  # ggalt::coord_proj("+proj=moll +lon_0=-60") +
   facet_grid(season~term, labeller = labeller(term = lab_sam)) +

### 2020-08-06 analysis/scripts/01-get-data.R
Corrects data

@@ -107,16 +107,16 @@ if (!force && file.exists(asymsam::data_path("raw", file))) {
                               path = asymsam::data_path())
 }
 
-
-for (y in seq(1979, 2018)) {
-  url <- paste0("ftp://ftp.cdc.noaa.gov/Datasets/ncep.reanalysis.dailyavgs/pressure/hgt.", y, ".nc")
-  dir.create(asymsam::data_path("raw", "ncep"), )
-  file <- asymsam::data_path("raw", "ncep", paste0("hgt.daily.", y, ".nc"))
-  if (!file.exists(file)) {
-    download.file(url, destfile = file)
-  }
-}
-
-ncepfiles <- list.files(asymsam::data_path("raw", "ncep"), full.names = TRUE)
-
-asymsam::nc_concatenate(ncepfiles, asymsam::data_path("raw", "ncep.daily.nc"), overwrite = TRUE)
+#
+# for (y in seq(1979, 2018)) {
+#   url <- paste0("ftp://ftp.cdc.noaa.gov/Datasets/ncep.reanalysis.dailyavgs/pressure/hgt.", y, ".nc")
+#   dir.create(asymsam::data_path("raw", "ncep"), )
+#   file <- asymsam::data_path("raw", "ncep", paste0("hgt.daily.", y, ".nc"))
+#   if (!file.exists(file)) {
+#     download.file(url, destfile = file)
+#   }
+# }
+#
+# ncepfiles <- list.files(asymsam::data_path("raw", "ncep"), full.names = TRUE)
+#
+# asymsam::nc_concatenate(ncepfiles, asymsam::data_path("raw", "ncep.daily.nc"), overwrite = TRUE)

### 2020-08-06 analysis/scripts/02-compute-eofs.R

@@ -5,7 +5,10 @@ library(data.table)
 
 indices_file <- data_path("derived", "indices.Rds")
 
+message("Hellow, you're running the script to compute symmetric and asymmetric SAM.\nI'll be your messenger for the day.")
+
 if (!file.exists(indices_file)) {
+  message("Reading data")
   hgt <- ReadNetCDF(ERA5(),
                     subset = list(time = c("1979-01-01", "2018-12-31"),
                                   latitude = c(-90, 10)),
@@ -14,9 +17,12 @@ if (!file.exists(indices_file)) {
     .[, hgt := hgt/9.8] %>%
     .[, hgt_a := hgt - mean(hgt), by = .(lon, lat, lev, month(time))]
 
+  message("Computing eofs... (this might take a while, you can go make some coffee).")
   indices <- hgt[lat <= -20, eof_asym(hgt_a, lon, lat, time, n = 1), by = lev]
-
+  message("Done! Saving file.")
   saveRDS(indices, file = data_path("derived", "indices.Rds"))
-
+  message("Everything's ready. Have a nice day!")
+} else {
+  message("The file ", indices_file, "already exists, so you don't need to do anything. Horray!")
 }
 

### 2020-08-06 analysis/paper/paper.Rmd
resultados

--- 2020-07-21 analysis/paper/paper.Rmd
@@
 
-```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM.", fig.width=6, fig.height=3, fig.env="figure*"}
+```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM and density estimates.", fig.width=6, fig.height=3, fig.env="figure*"}
 indices %>% 
@@
   geom_line(aes(color = term)) +
+  ggridges::geom_vridgeline(stat="ydensity", scale = -4e8, fill = NA,
+                            aes(color = term, width = ..density.., x = min(time - 1e7))) +
   scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
   scale_y_continuous(NULL) +
-  scale_x_datetime(NULL, expand = c(0, 0)) +
+  # scale_x_datetime(NULL, expand = c(0, 0)) +
   facet_grid(lev~., labeller = labeller(lev = lev.lab), scales = "free") +
@@
 
-Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and Symmetric time series corresponding to 700 and 30hPa. blablababla #FIXME
+Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and Symmetric time series corresponding to 700 and 30hPa.  #FIXME
 
-* stratosphere clearly nor normally distributed. a lot of values near 0 and some relatively high outliers. Especially true int he case of the asymmetric index. High frequency variablity.
+At first glance the series can be distinguished by their distributions.
+Whereas the tropospheric incides are aproximately normally distributed, the stratospheric indices are more long-tailed; that is, extreme values (both negative and positive) abound.
+The Asymmetric series have both more variability in the higher frequencies than the Symmetric series.
 
-* In both levels, there's correlation between the series (expected),
+The stratospheric Symmetric SAM varies strongly with  a two-year period, which can be seen using spectral methods (Figure A3) or in the autocorrelation structure (Figure A4).
+There is a local peak at 2 years in the periodigram of the tropospheric Symmetric SAM also, although it's not statistically significant.
+In the troposphere the most significant peak of variability is found in the Asymmetric index at around `r 12*.3` months.
 
@@
 
-Correlations between the Asymmetric and Symmetric series are rather constant throught the troposphere, fluctiating between `r signif(cors[lev >= 100, min(V1)], 2)` and `r signif(cors[lev >= 100, max(V1)], 2)` (Figure \ref{fig:cor-lev}).
+Correlations between the Asymmetric and Symmetric series are constant throught the troposphere, fluctiating between `r signif(cors[lev >= 100, min(V1)], 2)` and `r signif(cors[lev >= 100, max(V1)], 2)` (Figure \ref{fig:cor-lev}).
 Futhermore, the cross-correlation of each series across levels --shown in Figure \ref{fig:cross-correlation}-- are high in the trosposphere (greater than 0.9) for both indices.
@@
 
-Figure \ref{fig:cross-correlation}a) show the cross-correlation across levels for the Full SAM index. #FIXME
+Figure \ref{fig:cross-correlation}a) show the cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
+The high values below 100 hPa in panel reflects the vertical (zero-lag) coherency of the tropospheric SAM.
+Above 100 hPa correlation between levels falls of rapidly, indicating less coherent (zero-lag) variability.
+At the same time, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
+By examining panels b and c we see that the Asymemtric and Symmetric SAM  share the smae high level of coherency in the tropospherem but that the Asymmetric SAM is relatively more coherent in the stratosphere.
+The tropospheric-stratospheric connection is only present in the Symmetric SAM.
 
@@
 
+??
+
 ```{r compute-pp-regr} 
@@
 
-```{r ccf-levels, fig.cap = "Cross-correlation functions for each index and two differnet base levels."}
+```{r A2 ccf-levels, fig.cap = "Cross-correlation functions for each index and two differnet base levels."}
 ccf_dt <- function(...) {
@@
   # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
-  .[, fftspectrum(estimate, 5, B = 5000, probs = c(0.025, 0.975)), by = .(term, lev)] -> spec
+  .[, fftspectrum(estimate, c(3, 5), B = 5000, probs = c(0.025, 0.975)), by = .(term, lev)] -> spec
 ```
 
-```{r A2, fig.cap = "Fourier spectrum of each timeseries.
+```{r A3, fig.cap = "Fourier spectrum of each timeseries.
 The shading indicates de 95\\% area derived by fitting an AR process to each series and bootstrapping 5000 simulated samples.", appendix = TRUE}
 spec %>%
-  .[term != "full"] %>% 
+  # .[term != "full"] %>% 
   ggplot(aes(1/freq/12, spec)) +
@@
 
-```{r A3, fig.cap = "Autocorrelation functions of each timeseries", appendix = TRUE}
+```{r A4, fig.cap = "Autocorrelation functions of each timeseries", appendix = TRUE}
 acf_dt <- function(...) {
@@
   .[, lon := ConvertLongitude(lon)] %>% 
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)] %>% 
   ggplot(aes(lon, lat)) +
@@
 
-```{r A6, appendix = TRUE, fig.cap = "Regressions maps resulting from performing one multiple regression (a. and b.) and from performing two simple regressions (c. and d.)"}
+```{r A6, appendix = TRUE, fig.cap = "Regressions maps resulting from performing one multiple regression (column a) and from performing two simple regressions (column b)"}
 regr_dif <- hgt[lev %in% 700] %>% 
@@
   facet_grid(term ~ regression, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets() +
+  tag_facets("cr") +
   axis_labs_smol 
@@
 
-```{r A7, appendix = TRUE, fig.cap = " "}
+```{r A7, appendix = TRUE, fig.cap = "Zonal waves derives from the regression maps from performing one multiple regression (column a) and from performing two simple regressions (column b)"}
 regr_dif %>% 
@@
   facet_grid(term~regression, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free_x") +
-  tag_facets(position = "tr")
+  tag_facets("cr", position = "tr")
 ```

### 2020-08-06 analysis/paper/paper.tex
render figuras
resultados

@@
 \begin{figure*}
-\includegraphics{asymsam-timeseries-1} \caption[Time series for the asymmetric SAM and symmetric SAM]{Time series for the asymmetric SAM and symmetric SAM.}\label{fig:asymsam-timeseries}
+\includegraphics{asymsam-timeseries-1} \caption[Time series for the asymmetric SAM and symmetric SAM and density estimates]{Time series for the asymmetric SAM and symmetric SAM and density estimates.}\label{fig:asymsam-timeseries}
 \end{figure*}
@@
 Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and
-Symmetric time series corresponding to 700 and 30hPa. blablababla
-\#FIXME
+Symmetric time series corresponding to 700 and 30hPa. \#FIXME
 
-\begin{itemize}
-\item
-  stratosphere clearly nor normally distributed. a lot of values near 0
-  and some relatively high outliers. Especially true int he case of the
-  asymmetric index. High frequency variablity.
-\item
-  In both levels, there's correlation between the series (expected),
-\end{itemize}
+At first glance the series can be distinguished by their distributions.
+Whereas the tropospheric incides are aproximately normally distributed,
+the stratospheric indices are more long-tailed; that is, extreme values
+(both negative and positive) abound. The Asymmetric series have both
+more variability in the higher frequencies than the Symmetric series.
 
-Correlations between the Asymmetric and Symmetric series are rather
-constant throught the troposphere, fluctiating between 0.39 and 0.45
-(Figure \ref{fig:cor-lev}). Futhermore, the cross-correlation of each
-series across levels --shown in Figure \ref{fig:cross-correlation}-- are
-high in the trosposphere (greater than 0.9) for both indices. This
-suggests that both the Asymmetric and the Symemtric component of the
-tropospheric SAM are highly vertically coherent, both in their
-individual evolution and their temporal relationship. This is to be
-expected since the SAM is mostly equivalent barotropic (citaaaa).
+The stratospheric Symmetric SAM varies strongly with a two-year period,
+which can be seen using spectral methods (Figure A3) or in the
+autocorrelation structure (Figure A4). There is a local peak at 2 years
+in the periodigram of the tropospheric Symmetric SAM also, although it's
+not statistically significant. In the troposphere the most significant
+peak of variability is found in the Asymmetric index at around 3.6
+months.
+
+Correlations between the Asymmetric and Symmetric series are constant
+throught the troposphere, fluctiating between 0.39 and 0.45 (Figure
+\ref{fig:cor-lev}). Futhermore, the cross-correlation of each series
+across levels --shown in Figure \ref{fig:cross-correlation}-- are high
+in the trosposphere (greater than 0.9) for both indices. This suggests
+that both the Asymmetric and the Symemtric component of the tropospheric
+SAM are highly vertically coherent, both in their individual evolution
+and their temporal relationship. This is to be expected since the SAM is
+mostly equivalent barotropic (citaaaa).
 
@@
 Figure \ref{fig:cross-correlation}a) show the cross-correlation across
-levels for the Full SAM index. \#FIXME
+levels for the Full, Symmetric and Asymmetric SAM indices. The high
+values below 100 hPa in panel reflects the vertical (zero-lag) coherency
+of the tropospheric SAM. Above 100 hPa correlation between levels falls
+of rapidly, indicating less coherent (zero-lag) variability. At the same
+time, there is a non negligible correlation between the troposphere and
+the lower-to-middle stratosphere. By examining panels b and c we see
+that the Asymemtric and Symmetric SAM share the smae high level of
+coherency in the tropospherem but that the Asymmetric SAM is relatively
+more coherent in the stratosphere. The tropospheric-stratospheric
+connection is only present in the Symmetric SAM.
 
@@
 
+??
+
 \begin{sidewaysfigure}
@@
 \begin{figure}
-\includegraphics{ccf-levels-1} \caption[Cross-correlation functions for each index and two differnet base levels]{Cross-correlation functions for each index and two differnet base levels.}\label{fig:ccf-levels}
+\includegraphics{A2 ccf-levels-1} \caption[Cross-correlation functions for each index and two differnet base levels]{Cross-correlation functions for each index and two differnet base levels.}\label{fig:A2 ccf-levels}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A2-1} \appendcaption{A2}{Fourier spectrum of each timeseries.
-The shading indicates de 95\% area derived by fitting an AR process to each series and bootstrapping 5000 simulated samples.}\label{fig:A2}
+\includegraphics{A3-1} \appendcaption{A3}{Fourier spectrum of each timeseries.
+The shading indicates de 95\% area derived by fitting an AR process to each series and bootstrapping 5000 simulated samples.}\label{fig:A3}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A3-1} \appendcaption{A3}{Autocorrelation functions of each timeseries}\label{fig:A3}
+\includegraphics{A4-1} \appendcaption{A4}{Autocorrelation functions of each timeseries}\label{fig:A4}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A6-1} \appendcaption{A6}{Regressions maps resulting from performing one multiple regression (a. and b.) and from performing two simple regressions (c. and d.)}\label{fig:A6}
+\includegraphics{A6-1} \appendcaption{A6}{Regressions maps resulting from performing one multiple regression (column a) and from performing two simple regressions (column b)}\label{fig:A6}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A7-1} \caption[ ]{ }\label{fig:A7}
+\includegraphics{A7-1} \appendcaption{A7}{Zonal waves derives from the regression maps from performing one multiple regression (column a) and from performing two simple regressions (column b)}\label{fig:A7}
+\end{figure}
+
+\begin{figure}
+\includegraphics{A8-1} \appendcaption{A8}{Decadal trends of SAM indices for each season. }\label{fig:A8}
 \end{figure}

### 2020-09-01 analysis/paper/paper.Rmd

@@
 ---
-twocol: FALSE  #  Use FALSE for submission. 
+twocol: TRUE  #  Use FALSE for submission. 
 title: Title here
@@
  - \usepackage{gensymb}
+ - \usepackage{subfig}
 ---
@@
 main_levs <- c(700, 300, 30, 50)
+
+combine_lists <- function(...) {
+  X <- list(...)
+  names <- names(X[[1]])
+  names(names) <- names
+  lapply(names, function(n) {
+    unlist(lapply(X, function(x) x[[n]]))
+  })
+}
 ```
@@
 
-We used monthly precipitation data from CPC Merged Analysis of Precipitation [@xie1997] 2.5 degree latitude x 2.5 degree longitude.
+We used monthly precipitation data from CPC Merged Analysis of Precipitation [@xie1997] 2.5 degree latitude x 2.5 degree longitude. CPCC: [schneider2015]  #FIXME
 
@@
 Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right).
-Arbitrary units.", fig.width=6, fig.height=3, fig.env="figure*"}
+Arbitrary units.", fig.width=6, fig.height=2.5, fig.env="figure*"}
 sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
@@
   coord_polar() +
-  facet_grid(lev~variable, labeller = labeller(variable = label_parsed, lev = lev.lab)) +
+  facet_grid(.~variable, labeller = labeller(variable = label_parsed, lev = lev.lab)) +
   axis_labs_smol +
@@
 
+```{r}
+# indices %>% 
+#   copy() %>% 
+#   .[lev %in% c(50, 700)] %>% 
+#   .[, cumestimate := cumsum(estimate_norm), by = .(lev, term, season(time))] %>% 
+#   
+#   
+#   ggplot(aes(time, cumestimate)) +
+#   geom_line(aes(color = term)) +
+#   # ggridges::geom_vridgeline(stat="ydensity", scale = -4e8, fill = NA,
+#   #                           aes(color = term, width = ..density.., x = min(time - 1e7))) +
+#   scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam) +
+#   scale_y_continuous(NULL) +
+#   # scale_x_datetime(NULL, expand = c(0, 0)) +
+#   facet_grid(lev~season(time), labeller = labeller(lev = lev.lab)) +
+#   tag_facets()
+
+```
+
 ```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM and density estimates.", fig.width=6, fig.height=3, fig.env="figure*"}
 indices %>% 
-  .[lev %in% c(30, 700) & term != "full"] %>% 
+  .[lev %in% c(50, 700) & term != "full"] %>% 
   # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
@@
                             aes(color = term, width = ..density.., x = min(time - 1e7))) +
-  scale_color_brewer(NULL, palette = "Set1", labels = lab_sam) +
+  # geom_smooth(aes(color = term), method = "loess", n = 481, span = 10*12/481, method.args = list(degree = 1)) +
+  # geom_point(data = ~.x[lev == 50][time == as.Date("1987-11-01") & term == "sym" | 
+  #                         time == as.Date("1987-08-01") & term == "asym"]) +
+  ggforce::geom_mark_circle(data = ~.x[lev == 50][time == as.Date("1987-11-01") & term == "sym" | 
+                                                    time == as.Date("1987-08-01") & term == "asym"],
+                            expand = unit(2, "mm")) +
+  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam) +
   scale_y_continuous(NULL) +
   # scale_x_datetime(NULL, expand = c(0, 0)) +
-  facet_grid(lev~., labeller = labeller(lev = lev.lab), scales = "free") +
+  facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
   tag_facets()
@@
 
-Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and Symmetric time series corresponding to 700 and 30hPa.  #FIXME
+Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and Symmetric time series corresponding to 700 and 50hPa.  #FIXME
 
@@
 
-The stratospheric Symmetric SAM varies strongly with  a two-year period, which can be seen using spectral methods (Figure A3) or in the autocorrelation structure (Figure A4).
+The stratospheric Symmetric SAM varies strongly witha two-year period, which can be seen using spectral methods (Figure A3) or in the autocorrelation structure (Figure A4).
 There is a local peak at 2 years in the periodigram of the tropospheric Symmetric SAM also, although it's not statistically significant.
@@
 ```{r compute-cors}
-cors <- indices_wide[, cor(asym, sym), by = lev] 
+ccf_dt <- function(..., plot = FALSE) {
+  stat <- ccf(..., plot = plot)
+  list(lag = stat$lag[, 1, 1], 
+       acf = stat$acf[, 1, 1])
+}
+
+cors <- indices_wide %>% 
+  .[, ccf_dt(asym, sym, lag.max = 5), by = .(lev)]
 ```
 
-Correlations between the Asymmetric and Symmetric series are constant throught the troposphere, fluctiating between `r signif(cors[lev >= 100, min(V1)], 2)` and `r signif(cors[lev >= 100, max(V1)], 2)` (Figure \ref{fig:cor-lev}).
-Futhermore, the cross-correlation of each series across levels --shown in Figure \ref{fig:cross-correlation}-- are high in the trosposphere (greater than 0.9) for both indices.
-This suggests that both the Asymmetric and the Symemtric component of the tropospheric SAM are highly vertically coherent, both in their individual evolution and their temporal relationship.
-This is to be expected since the SAM is mostly equivalent barotropic (citaaaa).
+From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric and Symmetric time series appear to be correlated.
+Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987 marked with a circle).
+We show these correlations, across all the levels of the reanalysis and for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}.
 
-In the stratosphere the situation is different.
-As can be seen in Figure \ref{fig:cor-lev}, the relationship between the Asymmetric and Symmetric indices varies with height above 100 hPa.
-It starts to decreese right over the tropopause, reaches a minium of `r signif(min(cors$V1), 2)` at `r cors[which.min(V1), lev]` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis.
-The cross-correlation across levels in the stratosphere is generally weaker than in the troposphere (Figure \ref{fig:cross-correlation}).
-Furthermore, above 100 hPa, the cross-correlation decreases more rapdily with height for the Symmetric SAM than for the Asymmetric SAM as evidenced by the wider dark red areas near the diagonal in Figure \ref{fig:cross-correlation}b) vs.
-Figure \ref{fig:cross-correlation}c).
-Moreover, the stratospheric Symmetric SAM seems to be slightly more connected to the trosposphere than the Asymmetric SAM; this can be seen by the lower correlation values in the top right quadrant of Figure \ref{fig:cross-correlation}b) in comparison with Figure \ref{fig:cross-correlation}c).
-
-```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level.", fig.width=3, fig.height=3}
+```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level for lag zero and lag -1 (Asymmetric leads Symmetric).", fig.width=3, fig.height=3}
 cors %>% 
-  ggplot(aes(lev, V1)) +
-  geom_line() +
+  .[lag %in% c(0, -1)] %>% 
+  ggplot(aes(lev, acf)) +
+  geom_line(aes(color = factor(lag, levels = c(0, -1), ordered = TRUE),
+                linetype = factor(lag, levels = c(0, -1), ordered = TRUE))) +
   scale_x_level() +
   scale_y_continuous("Correlation", limits = c(0, 1)) + 
+  scale_linetype_manual(NULL, values = c(1, 5), labels = c("0" = "Zero-lag", 
+                                                         "-1" = "Asymmetric leads\nSymmetric (1 month)")) +
+  scale_color_brewer(NULL, palette = "Dark2", labels = c("0" = "Zero-lag", 
+                                                         "-1" = "Asymmetric leads\nSymmetric (1 month)")) +
   coord_flip()
@@
 
-```{r cross-correlation, fig.cap = "Cross correlation between levels of the Full, Asymmetric and Symmetric SAM.", fig.width = 6, fig.height=4, fig.env="figure*"}
+Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctiating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
+One-month-lag correlations are similarly constant but significantly reduced, hovering around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`.
+In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis.
+At the same time, one-month-lag correlations increase with height.
+
+```{r cross-correlation, fig.cap = "Cross correlation between levels of the Full, Asymmetric and Symmetric SAM.", fig.width = 6, fig.height=3, fig.env="figure*"}
 tri <- matrix(c(1000, 1000, 1000, 1, 1, 1, 1000, 1000), ncol = 2, byrow = TRUE) %>% 
@@
 
-indices %>% 
+crosscor <- indices %>% 
   copy() %>% 
-  .[, widyr::pairwise_cor(.SD, lev, time, estimate, diag = FALSE), by = .(term)] %>% 
+  .[, widyr::pairwise_cor(.SD, lev, time, estimate, diag = FALSE), by = .(term)] 
+crosscor %>% 
   ggplot(aes(item2, item1)) +
   geom_contour_fill(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
+  geom_contour_tanaka(aes(z = correlation), size = 0.1, na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
   geom_text_contour(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1), 
@@
 
-Figure \ref{fig:cross-correlation}a) show the cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
-The high values below 100 hPa in panel reflects the vertical (zero-lag) coherency of the tropospheric SAM.
-Above 100 hPa correlation between levels falls of rapidly, indicating less coherent (zero-lag) variability.
-At the same time, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
-By examining panels b and c we see that the Asymemtric and Symmetric SAM  share the smae high level of coherency in the tropospherem but that the Asymmetric SAM is relatively more coherent in the stratosphere.
-The tropospheric-stratospheric connection is only present in the Symmetric SAM.
+Figure \ref{fig:cross-correlation}a) shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
+For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposfere.
+Above 100 hPa correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
+Still there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
+Examining panels b and c, we see that the Asymemtric and Symmetric SAM  share the same high level of coherency in the troposphere but they differ in their stratospheric behaviour.
+As evidenced by the wider dark red areas near the diagonal in Figure \ref{fig:cross-correlation}b) vs.
+Figure \ref{fig:cross-correlation}c), stratospheric coherency is stronger for the Asymmetric SAM than the Symemtric SAM.
+The stratospheric Symmetric SAM seems to connect more strongly to the trosposphere than the Asymmetric SAM; this can be seen by the lower correlation values in the top right left of Figure \ref{fig:cross-correlation}b) in comparison with Figure \ref{fig:cross-correlation}c).
 
@@
 <!--   geom_line() + -->
-<!--   scale_fill_brewer(palette = "Set1", aesthetics = c("color", "fill")) + -->
+<!--   scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) + -->
 <!--   geom_hline(yintercept = 0) + -->
@@
 
-```{r trends, fig.cap = "Trends for each index at each level. Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=3}
-indices %>%
+```{r trends, fig.cap = "Decadal normalised trends for each index at each level for annual (row a) and seasonal values (rows b-e) for the period 1979-2018.
+Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
+annual_trend <- indices %>%
   copy() %>% 
@@
   .[, term := NULL] %>% 
+  .[, season := "Annual"]
+
+indices %>% 
+  copy() %>% 
+  # .[time > as.Date("1990-01-01")] %>%
+  .[, season := season(time)] %>% 
+  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
+  .[, FitLm(estimate_norm, time, se = TRUE), by = .(type = term, lev, season)] %>% 
+  rm_intercept() %>% 
+  .[, term := NULL] %>% 
+  rbind(annual_trend) %>% 
   .[, estimate := estimate*3600*24*365*10] %>% 
@@
   setnames("type", "term") %>% 
+  .[, season := forcats::fct_relevel(season, "Annual")] %>% 
   ggplot(aes(lev, estimate)) +
@@
   geom_line() +
-  scale_fill_brewer(palette = "Set1", aesthetics = c("color", "fill")) +
+  scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
   geom_hline(yintercept = 0) +
@@
   coord_flip() +
-  facet_grid(~ term,labeller = labeller(term = lab_sam)) +
-  tag_facets() 
+  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
+  tag_facets("rc") 
 ```
 
-Figure \ref{fig:trends} shows normalised decadal trends for each index for the whole period along with a 95% interval in shading.
-There is a statistically significant increase in positive SAM in the tropostphere (panel a), which has been already documented in other studies (e.g.
-@fogt2020).
-Panels b and c of Figure \ref{fig:trends} show that this increase is evident only in the Symmetric component.
-This distinction should prove useful when attributing trends in other variables such as temperature and precipitation to trends in the SAM.
+Figure \ref{fig:trends} shows normalised decadal trends for each index for the whole period 1979-2918 along with the 95% confidence interval in shading for the whole year (row a) and separed by trimesters (rows b through e).
+As documented by #FIXME (e.g.
+@fogt2020), there is a statistically significant increase towards more positive SAM (panel a.1), which is XX only in Summer and Autumn (panels b.1 and c.1).
+We observe these increases mainly in the troposphere, reaching their maximum at at 100 hPa in Summer.
+By separating the SAM signal in its Asymmetric and Symmetric parts, we can not only see that these trends are almost entirely due to the Symmetric component (columns 2 vs.
+columns 3), but in some cases the trends become more clear.
+In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the signal; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel b.1 and b.3).
+In Autumn, using the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full index.
+
+```{r regression-90}
+local <- indices %>%
+  .[time >= as.Date("1990-01-01")] %>% 
+  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
+  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term, season(time))] %>% 
+  rm_intercept() %>% 
+  .[, term := NULL] %>% 
+  .[, estimate := estimate*3600*24*365*10] %>% 
+  .[, std.error := std.error*3600*24*365*10]
+
+trends_text <- local[, paste0(signif(estimate, 2), " \\pm ", signif(std.error*1.96, 2)), 
+                     by = .(term = type, season, lev)]
+```
+
+We stress that these are only linear trends during the whole period and the absence of a statistically significant signal should not be taken as evidence of no sistematic change.
+In particular, going back to Figure \ref{fig:asymsam-timeseries}, we can see an evident change in the stratospheric Asymemtric component (red line in panel a) between the 90's, when we see a dominance of extreme negative values, and the 00's, when we see the inverse.
+This change is restricted to the Winter months: the linear trend for JJA starting in 1990 for the Asymmetric component at 50hPa is $`r trends_text[term == "asym" & lev == 50 & season == "JJA"]$V1`$.
+
+These results point to very different behaviours between the stratospheric and tropospheric SAM. #FIXME
+
+This suggests that both the Asymmetric and the Symemtric component of the tropospheric SAM are highly vertically coherent, both in their individual evolution and their temporal relationship.
+This is to be expected since the SAM is mostly equivalent barotropic (citaaaa).
 
@@
   indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
-  .[, rbind(as.data.table(FitLm(hgt_a, sym, asym, se = TRUE)),
-            as.data.table(FitLm(hgt_a, full, se = TRUE))),
+  .[, combine_lists(FitLm(hgt_a, sym, asym, se = TRUE),
+                    FitLm(hgt_a, full, se = TRUE)),
     by = .(lon, lat, lev)] %>% 
@@
 
-To undertand the spatial patterns associated with both indeces, we regressed monthly geopotential anomalies into both indeces using multiple regression (Figure A6 illustrates the difference between computing two simple regressions and one multiple regression).
+To undertand the spatial patterns associated with both indeces, we regressed monthly geopotential anomalies into both indeces using multiple regression
 
 ```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.env = "figure*", fig.width=6, fig.height=7}
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.env = "figure*", fig.width=7, fig.height=5}
 indices_2d_yearly %>% 
   .[lat <= -20] %>% 
+  .[lev %in% c(50, 700)] %>% 
   ggplot(aes(lon, lat)) +
   geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
-  # geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
-  # size = 0.3,
-  # global.breaks = FALSE) +
   geom_contour_tanaka(aes(z = estimate),
@@
   geom_qmap(~.x[lat <= -20]) +
-  # geom_contour2(aes(z = p.val), breaks = 0.01) +
-  # stat_subset(aes(subset = p.val <= 0.01 & is.cross(lon, lat, 2)), geom = "point", size = 0.5, alpha = 0.3) +
-  # geom_hline(yintercept = c(-65, -40)) +
   scale_x_longitude() +
@@
   coord_polar() +
-  facet_grid(term ~ lev, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets() +
+  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  tag_facets("rc") +
   axis_labs_smol 
@@
 
-Figure \ref{fig:2d-regr} shows the spatial year-long regression for selected levels.
-In the troposphere the Full annular mode is clearly "contaminated" with well known zonal asymemtries (panels g and j) which are succesfully sepparated by our methodology (panels h, i, k and l).
-In the stratosphere, the spatial pattern associated with the Full SAM is much more clearly dominated by a zonally symmetric, monopolar structure (panels a and d) that is, however, not perfectly centered in the south pole.
-The monopoloe obtained by multiple regression with the Asymmetric and Symmetric SAM (panels c and f in Figure \ref{fig:2d-regr}) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 pattern (panels b and e).
+Figure \ref{fig:2d-regr} shows the spatial year-long regression #FIXME.
+Column a are regressinos using the Full SAM, while columns b and c are regression coefficients computed in a multiple regression of geopotential height on the Asymmetric and Symmetric indices at the same time.
+Thus, they are to be interpreted as the patterns associated with each index, controling for the (linear) effect of the other.
+(Figure A6 #FIXME illustrates the difference between computing two simple regressions and one multiple regression.)
+
+In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the south pole.
+The monopole obtained by multiple regression with the Asymmetric and Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel b.1).
+
+In the troposphere, panel a.2 show the well known zonally symmetrical annular mode *contaminated* with zonal asymmetries in the form of a wave-3.
+The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
+The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel c.2).
+#FIXME (agregar algo más?)
 
@@
   geom_line(aes(color = factor(k))) +
-  scale_color_brewer("Wave-number", palette = "Set1") +
+  scale_color_brewer("Wave-number", palette = "Dark2") +
   scale_y_continuous("Amplitude") +
-  scale_x_latitude(ticks = 15) +
+  scale_x_latitude(ticks = 20) +
   coord_flip() +
@@
 
-The amplitude of each zonal wave number at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the zonal mean.
+The amplitude of each zonal wave number at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
 Comparing between rows, this Figure quantifies the relatively clean separation between the zonally symmetric and zonally asymmetric structures, as its evident how the mixture of waves of the Full field (first row) is very similar to the sum of the waves of the Asymmetric and Symmetric field (second and third row, respectively).
-The second row of Figure \ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b), while in the stratosphere it is composed of zonal waves 3 to 1 in decreasing level of importance.
+The second row of Figure \ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance.
 
-```{r vertical-regression, fig.cap = "Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South.
-(this caption needs some love)", fig.width = 4, fig.height=3}
-ReadNetCDF(ERA5(), 
+```{r compute-vertical-regression}
+# vertical_regr <- ReadNetCDF(ERA5(), 
+#            subset = list(time = c("1979-01-01", "2018-12-31"),
+#                          latitude = c(-65, -40)),
+#            vars = c(hgt = "z")) %>% 
+#   normalise_coords() %>% 
+#   .[, .(hgt = weighted.mean(hgt, w = cos(lat*pi/180))), by = .(lon, lev, time)] %>% 
+#   .[, hgt := hgt/9.8] %>% 
+#   .[, hgt_a := hgt - mean(hgt), by = .(lon, lev, month(time))] %>% 
+#   indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
+#   .[, FitLm(hgt_a, sym, asym, se = TRUE),
+#     by = .(lon, lev)] %>% 
+#   rm_intercept() %>% 
+#   .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, lev)] %>% 
+#   periodic(lon = c(0, 360)) %>% 
+#   .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)]
+
+indices_wide %>% 
+  copy() %>% 
+  setnames("lev", "index_level") -> indices_for_regr
+
+vertical_regr <- ReadNetCDF(ERA5(), 
                             subset = list(time = c("1979-01-01", "2018-12-31"),
@@
   .[, hgt_a := hgt - mean(hgt), by = .(lon, lev, month(time))] %>% 
-  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
-  .[, FitLm(hgt_a, sym, asym, se = TRUE),
-    by = .(lon, lev)] %>% 
+  indices_for_regr[., on = .NATURAL, allow.cartesian = TRUE] %>% 
+  .[, combine_lists(FitLm(hgt_a, full, se = TRUE),
+                    FitLm(hgt_a, sym, asym, se = TRUE)),
+    by = .(lon, lev, index_level)] %>% 
   rm_intercept() %>% 
-  .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, lev)] %>% 
+  .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, index_level, lev)] %>% 
   periodic(lon = c(0, 360)) %>% 
-  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] -> vertical_regr
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] 
+```
 
+```{r vertical-regression, fig.cap = "Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South.
+(\\#FIXME this caption needs some love)", fig.width = 3, fig.height=3}
 vertical_regr %>% 
-  copy() %>% 
+  .[index_level %in% c(50, 700)] %>% 
   .[term == "asym"] %>%
-  # .[, estimate := estimate/sd(estimate), by = .(lev, term)] %>%
   ggplot(aes(lon, lev)) +
@@
                       global.breaks = FALSE) +
+  # geom_contour2(aes(z = p.val), breaks = 0.01) +
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(18), 
@@
   scale_y_level() +
-  # geom_hline(yintercept = 300) +
-  scale_x_longitude(ticks = 60) 
+  scale_x_longitude(ticks = 60)  +
+  facet_grid(index_level~., labeller = labeller(index_level = lev.lab)) +
+  tag_facets(position = "bl") 
 ```
 
-From Figure \ref{fig:2d-regr} it appears that the vertical structure of the Asymmetric SAM is equivalent barotropic in the troposphere but baroclinic in the stratosphere.
-Anomalies are centerd in the same locations in the troposphere (panels h and k), but show westerly displacement in the stratosphere (panesl b and e).
-This is expanded in Figure \ref{fig:vertical-regression}, which shows a vertical crossection of the regression coefficient corresponding to the middle row of Figure \ref{fig:2d-regr}, area-weighted averaged between 65 and 40 degrees South.
-Below 100 hPa, anomalies are completely vertical, while above they show an important westerly tilt with height.
+To analyse the vertical structure of the geopotential anomalies asociated with the asymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65ªS and 40ªS for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
+
+The geopotential anomalies associated with the stratospheric SAM (panel a) are clearly constrained to the stratosphere, which underscores the disconnect between the stratospheric and tropospheric symetric SAM.
+The vertical structure this signal tilts about 60ª to the West between 100 hPa and 1 hPa, suggesting baroclinic processes and polarward transport of heat (#FIXME is this ok?).
+Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
+
+The tropospheric asymmetric SAM has significant signals that extend upwards the uppermost levels of the reanalysis.
+In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
+#FIXME moaaaaar.
+
+Interestingly, the structures shown in Figure \ref{fig:vertical-regression} are surprisignly robust to the choice of index level.
+For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in panel a.
+Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b.
+The pattern cross-correlation between levels of each segment of the atmosphere is greater than 0.9 (Figure A#fixme).
+The patterns mainly change in amplitude.
+The tropospheric pattern is maximised by the 300 hPa Asymmetric SAM index and the stratospheric pattern increased monotonically with height.
 
@@
 ```{r compute-pp-regr} 
-pp <- CMAP() %>% 
-  ReadNetCDF(subset = list(lat = c(-90, -20))) %>% 
+# pp <- CMAP() %>% 
+#   ReadNetCDF(subset = list(lat = c(-90, -20))) %>% 
+#   normalise_coords() %>% 
+#   .[, precip := Anomaly(precip), by = .(lon, lat)]
+
+continents <- list(africa  = list(lon = c(0, 60),
+                                  lat = c(-40, 0)),
+                   oceania = list(lon = c(100, 180),
+                                  lat = c(-50, -5)),
+                   america = list(lon = c(265, 325),
+                                  lat = c(-60, -10)))
+
+add_continent <- function(dt) {
+  dt[, continent := fcase(lon %between% continents$africa$lon & 
+                            lat %between% continents$africa$lat, "africa",
+                          lon %between% continents$oceania$lon & 
+                            lat %between% continents$oceania$lat, "oceania",
+                          lon %between% continents$america$lon &
+                            lat %between% continents$america$lat, "america")] %>% 
+    .[, continent := factor(continent, levels = c("africa", "oceania", "america"), ordered = TRUE)] %>% 
+    .[]
+}
+
+pp_regr_season <- CPCC() %>% 
+  ReadNetCDF(subset = list(lon = lapply(continents, function(x) x$lon),
+                           lat = lapply(continents, function(x) x$lat))) %>% 
+  na.omit() %>% 
   normalise_coords() %>%
-  .[, precip := Anomaly(precip), by = .(lon, lat)]
-
-pp_regr_season <- pp %>% 
+  .[, days := lubridate::days_in_month(time[1]), by = time] %>% 
+  .[, precip := precip/days] %>% 
   # .[, air := Anomaly(air), by = .(lon, lat)] %>%
@@
   rm_singleton() %>% 
-  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
+  add_continent() %>% 
+  .[!is.na(continent)] %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, continent, time = seasonally(time))] %>%
+  .[, `:=`(precip_a = Anomaly(precip),
+           precip_a_p = log(precip) - log(mean(precip, na.rm = TRUE))),
+    by = .(lon, lat, season(time))] %>% 
   .[, nas := mean(is.na(precip)), by = .(lon,  lat, season(time))] %>% 
   .[nas >= 0.15, precip := NA] %>% 
-  .[, rbind(as.data.table(FitLm(precip, asym, sym, se = TRUE)),
-            as.data.table(FitLm(precip, full, se = TRUE))),
+  na.omit() %>% 
+  .[is.finite(precip_a_p)] %>% 
+  .[, combine_lists(FitLm(precip_a, asym, sym, se = TRUE),
+                    FitLm(precip_a, full, se = TRUE)),
     by = .(lon, lat, season(time))] %>% 
@@
 
+filter_sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
+  data[is.sa(data, lons = lons, lats = lats)]
+}
+
 ```
 
-```{r pp-regr-global,  fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"}
+```{r def-plot_pp}
+plot_pp <- function(this_continent) {
   pp_regr_season %>%
-  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)] %>% 
+    add_continent() %>% 
+    # .[lat <= -10] %>% 
+    .[continent == this_continent] %>%
+    .[, p.value := Pvaluate(estimate, std.error, df, "none"), by = .(season, continent, term)] %>%
+    .[, p.value2 := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, continent, term)] %>%
     ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.25, 0)) +
-  geom_contour2(aes(z = p.value), breaks = 0.01) +
-  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
-  stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
-  geom_qmap(~.x[lat <= -20]) +
-  scale_x_longitude() +
-  scale_y_latitude(labels = NULL) +
-  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
-                       breaks = AnchorBreaks(0, 0.25, 0)) +
-  coord_polar() +
+    geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.5, 0)) +
+    # geom_raster(aes(fill = estimate/30)) +
+    geom_contour2(aes(z = p.value2), breaks = 0.01) +
+    # geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
+    # stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat, 1)), size = 0.4, alpha = 0.3) +
+    geom_qmap(~.x %>% add_continent() %>% .[continent == this_continent], keep = 0.015, weighting = 0.9) +
+    scale_x_longitude(ticks = 15) +
+    scale_y_latitude(ticks = 15, labels = NULL) +
+    scale_fill_divergent("Precipitation", 
+                         limits = c(-3.4, 3.4),
+                         oob = scales::squish,
+                         guide = guide_colorstrip_bottom(), 
+                         breaks = AnchorBreaks(0, 0.5, 0)) +
+    coord_quickmap(xlim = continents[[this_continent]]$lon, 
+                   ylim = continents[[this_continent]]$lat) +
     facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets() +
+    tag_facets(position = "br") +
     axis_labs_smol
+}
 ```
 
+```{r pp-regr-global-africa,  fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
+Africa", fig.width = 6, fig.height = 4.5, fig.env = "figure*"}
+plot_pp("africa")
+
+```
+
+```{r pp-regr-global-oceania,  fig.cap = "Same but for oceania", fig.width = 6, fig.height = 4.5, fig.env = "figure*"}
+plot_pp("oceania") +
+  tag_facets(position = "bl")
+
+```
+
+```{r pp-regr-global-america,  fig.cap = "Same but for america", fig.width = 6, fig.height = 6, fig.env = "figure*"}
+plot_pp("america")
+
+```
+
+## Conclusiones
+
+This suggests that both the Asymmetric and the Symemtric component of the tropospheric SAM are highly vertically coherent, both in their individual evolution and their temporal relationship.
+This is to be expected since the SAM is mostly equivalent barotropic (citaaaa).
+
 \acknowledgments
-CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/
+CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ #FIXME
 
@@
   annotation_logticks(sides = "b") +
-  scale_color_brewer(NULL, palette = "Set1", aesthetics = c("color", "fill")) +
+  scale_color_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill")) +
   scale_x_log10("Period (years)", sec.axis = sec_axis(~.*12)) +
@@
   geom_col(aes(fill = term)) +
-  scale_fill_brewer(palette = "Set1") +
+  scale_fill_brewer(palette = "Dark2") +
   facet_grid(term ~ lev,labeller = labeller(lev = lev.lab, term = lab_sam)) +
   tag_facets(position = "tr") 
-```
-
-```{r A5, fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM. P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", appendix = TRUE}
-
-is.sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
-  data[, lon %between% lons & lat %between% lats]
-}
-
-filter_sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
-  data[is.sa(data, lons = lons, lats = lats)]
-}
-
-pp_regr_season %>% 
-  filter_sa() %>% 
-  .[, lon := ConvertLongitude(lon)] %>% 
-  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)] %>% 
-  ggplot(aes(lon, lat)) +
-  # geom_raster(aes(fill = estimate)) +
-  geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.25, 0)) +
-  geom_contour2(aes(z = p.value), breaks = 0.01) +
-  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
-  stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat)), size = 0.4, alpha = 0.3) +
-  geom_qmap(~filter_sa(.x)[, lon := ConvertLongitude(lon)], keep = 0.02) +
-  scale_x_longitude() +
-  scale_y_latitude() +
-  scale_fill_divergent("Precipitation", guide = guide_colorstrip_bottom(), 
-                       breaks = AnchorBreaks(0, 0.25, 0)) +
-  # coord_polar() +
-  # ggalt::coord_proj("+proj=moll +lon_0=-60") +
-  facet_grid(season~term, labeller = labeller(term = lab_sam)) +
-  tag_facets() +
-  axis_labs_smol 
 ```
@@
   geom_line(aes(color = factor(k))) +
-  scale_color_brewer("Wave-number", palette = "Set1") +
+  scale_color_brewer("Wave-number", palette = "Dark2") +
   scale_y_continuous("Amplitude") +
@@
 
-```{r A8, appendix = TRUE, fig.cap = "Decadal trends of SAM indices for each season. "}
-indices %>% 
+```{r A8, appendix = TRUE, fig.cap = "Pattern cross-correlation \\#FIXME!"}
+vertical_regr %>% 
+  .[term == "asym"] %>% 
   copy() %>% 
-  .[, season := season(time)] %>% 
-  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
-  .[, FitLm(estimate_norm, time, se = TRUE), by = .(type = term, lev, season)] %>% 
-  rm_intercept() %>% 
-  .[, term := NULL] %>% 
-  .[, estimate := estimate*3600*24*365*10] %>% 
-  .[, std.error := std.error*3600*24*365*10] %>% 
-  setnames("type", "term") %>% 
-  ggplot(aes(lev, estimate)) +
-  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) +
-  geom_line() +
-  scale_fill_brewer(palette = "Set1", aesthetics = c("color", "fill")) +
-  geom_hline(yintercept = 0) +
-  scale_x_level() +
-  scale_y_continuous("Standard deviations per decade") +
-  coord_flip() +
-  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
-  tag_facets() 
+  .[, id := interaction(lev, lon)] %>% 
+  widyr::pairwise_cor(index_level, id, estimate) %>% 
+  ggplot(aes(item2, item1)) +
+  geom_contour_fill(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
+  geom_contour_tanaka(aes(z = correlation), size = 0.1, na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
+  geom_text_contour(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1), 
+                    color = "white", rotate = FALSE, stroke.color = "black", stroke = 0.2, 
+                    skip = 1, size = 4) +
+  geom_polygon(data = tri, fill = "white") +
+  scale_fill_divergent("Cross-correlation", 
+                       breaks = MakeBreaks(binwidth = 0.1),
+                       guide = guide_colorstrip_bottom()) +
+  scale_x_level(minor_breaks = logbreaks, position = "top") +
+  scale_y_level(minor_breaks = logbreaks) +
+  coord_equal() +
+  theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside") 
 ```

### 2020-09-01 analysis/paper/paper.tex

@@
 % DOUBLE-SPACED VERSION FOR SUBMISSION TO THE AMS
-\documentclass[]{ametsocV5}
+\documentclass[twocol]{ametsocV5}
 
@@
 \usepackage{gensymb}
+\usepackage{subfig}
 
@@
 Precipitation \citep{xie1997} 2.5 degree latitude x 2.5 degree
-longitude.
+longitude. CPCC: {[}schneider2015{]} \#FIXME
 
@@
 Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and
-Symmetric time series corresponding to 700 and 30hPa. \#FIXME
+Symmetric time series corresponding to 700 and 50hPa. \#FIXME
 
@@
 
-The stratospheric Symmetric SAM varies strongly with a two-year period,
+The stratospheric Symmetric SAM varies strongly witha two-year period,
 which can be seen using spectral methods (Figure A3) or in the
@@
 
-Correlations between the Asymmetric and Symmetric series are constant
-throught the troposphere, fluctiating between 0.39 and 0.45 (Figure
-\ref{fig:cor-lev}). Futhermore, the cross-correlation of each series
-across levels --shown in Figure \ref{fig:cross-correlation}-- are high
-in the trosposphere (greater than 0.9) for both indices. This suggests
-that both the Asymmetric and the Symemtric component of the tropospheric
-SAM are highly vertically coherent, both in their individual evolution
-and their temporal relationship. This is to be expected since the SAM is
-mostly equivalent barotropic (citaaaa).
-
-In the stratosphere the situation is different. As can be seen in Figure
-\ref{fig:cor-lev}, the relationship between the Asymmetric and Symmetric
-indices varies with height above 100 hPa. It starts to decreese right
-over the tropopause, reaches a minium of 0.21 at 20 hPa and then it
-increases again monotonically with height up to the uppermost level of
-the reanalysis. The cross-correlation across levels in the stratosphere
-is generally weaker than in the troposphere (Figure
-\ref{fig:cross-correlation}). Furthermore, above 100 hPa, the
-cross-correlation decreases more rapdily with height for the Symmetric
-SAM than for the Asymmetric SAM as evidenced by the wider dark red areas
-near the diagonal in Figure \ref{fig:cross-correlation}b) vs.~Figure
-\ref{fig:cross-correlation}c). Moreover, the stratospheric Symmetric SAM
-seems to be slightly more connected to the trosposphere than the
-Asymmetric SAM; this can be seen by the lower correlation values in the
-top right quadrant of Figure \ref{fig:cross-correlation}b) in comparison
-with Figure \ref{fig:cross-correlation}c).
+From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric
+and Symmetric time series appear to be correlated. Moreover, looking at
+the extremes in the stratosphere, the Symmetric serie appears to lag the
+Asymmetric series (see, for example, the positive events on late 1987
+marked with a circle). We show these correlations, across all the levels
+of the reanalysis and for zero and -1 lag (Asymmetric index leading the
+Symmetric index), in Figure \ref{fig:cor-lev}.
 
 \begin{figure}
-\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric and Asymmetric SAM at each level]{Correlation between the Symmetric and Asymmetric SAM at each level.}\label{fig:cor-lev}
+\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric and Asymmetric SAM at each level for lag zero and lag -1 (Asymmetric leads Symmetric)]{Correlation between the Symmetric and Asymmetric SAM at each level for lag zero and lag -1 (Asymmetric leads Symmetric).}\label{fig:cor-lev}
 \end{figure}
+
+Zero-lag correlations between the Asymmetric and Symmetric series are
+relatively constant throught the troposphere, fluctiating between 0.39
+and 0.45. One-month-lag correlations are similarly constant but
+significantly reduced, hovering around 0.17. In the stratosphere,
+zero-lag correlations drop to a minimum of 0.21 at 20 hPa and then it
+increases again monotonically with height up to the uppermost level of
+the reanalysis. At the same time, one-month-lag correlations increase
+with height.
 
@@
 
-Figure \ref{fig:cross-correlation}a) show the cross-correlation across
-levels for the Full, Symmetric and Asymmetric SAM indices. The high
-values below 100 hPa in panel reflects the vertical (zero-lag) coherency
-of the tropospheric SAM. Above 100 hPa correlation between levels falls
-of rapidly, indicating less coherent (zero-lag) variability. At the same
-time, there is a non negligible correlation between the troposphere and
-the lower-to-middle stratosphere. By examining panels b and c we see
-that the Asymemtric and Symmetric SAM share the smae high level of
-coherency in the tropospherem but that the Asymmetric SAM is relatively
-more coherent in the stratosphere. The tropospheric-stratospheric
-connection is only present in the Symmetric SAM.
+Figure \ref{fig:cross-correlation}a) shows (zero-lag) cross-correlation
+across levels for the Full, Symmetric and Asymmetric SAM indices. For
+the Full SAM (panel a), high values below 100 hPa reflect the vertical
+(zero-lag) coherency throughout the troposfere. Above 100 hPa
+correlation between levels falls off more rapidly, indicating less
+coherent (zero-lag) variability. Still there is a non negligible
+correlation between the troposphere and the lower-to-middle
+stratosphere. Examining panels b and c, we see that the Asymemtric and
+Symmetric SAM share the same high level of coherency in the troposphere
+but they differ in their stratospheric behaviour. As evidenced by the
+wider dark red areas near the diagonal in Figure
+\ref{fig:cross-correlation}b) vs.~Figure \ref{fig:cross-correlation}c),
+stratospheric coherency is stronger for the Asymmetric SAM than the
+Symemtric SAM. The stratospheric Symmetric SAM seems to connect more
+strongly to the trosposphere than the Asymmetric SAM; this can be seen
+by the lower correlation values in the top right left of Figure
+\ref{fig:cross-correlation}b) in comparison with Figure
+\ref{fig:cross-correlation}c).
 
-\begin{figure}
-\includegraphics{trends-1} \caption[Trends for each index at each level]{Trends for each index at each level. Shading indicates the 95\% confidence interval.}\label{fig:trends}
-\end{figure}
+\begin{figure*}
+\includegraphics{trends-1} \caption[Decadal normalised trends for each index at each level for annual (row a) and seasonal values (rows b-e) for the period 1979-2018]{Decadal normalised trends for each index at each level for annual (row a) and seasonal values (rows b-e) for the period 1979-2018.
+Shading indicates the 95\% confidence interval.}\label{fig:trends}
+\end{figure*}
 
 Figure \ref{fig:trends} shows normalised decadal trends for each index
-for the whole period along with a 95\% interval in shading. There is a
-statistically significant increase in positive SAM in the tropostphere
-(panel a), which has been already documented in other studies (e.g.
-\citet{fogt2020}). Panels b and c of Figure \ref{fig:trends} show that
-this increase is evident only in the Symmetric component. This
-distinction should prove useful when attributing trends in other
-variables such as temperature and precipitation to trends in the SAM.
+for the whole period 1979-2918 along with the 95\% confidence interval
+in shading for the whole year (row a) and separed by trimesters (rows b
+through e). As documented by \#FIXME (e.g. \citet{fogt2020}), there is a
+statistically significant increase towards more positive SAM (panel
+a.1), which is XX only in Summer and Autumn (panels b.1 and c.1). We
+observe these increases mainly in the troposphere, reaching their
+maximum at at 100 hPa in Summer. By separating the SAM signal in its
+Asymmetric and Symmetric parts, we can not only see that these trends
+are almost entirely due to the Symmetric component (columns 2
+vs.~columns 3), but in some cases the trends become more clear. In
+Summer, the Asymmetric SAM has a statistically non significant negative
+trend in the middle troposphere that obscures the signal; as a result,
+trends computed using only the Symmetric component are more clear
+(compare the shading region in panel b.1 and b.3). In Autumn, using the
+Symmetric SAM reveals a statistically significant positive trend in the
+stratosphere that is not significant using the Full index.
+
+We stress that these are only linear trends during the whole period and
+the absence of a statistically significant signal should not be taken as
+evidence of no sistematic change. In particular, going back to Figure
+\ref{fig:asymsam-timeseries}, we can see an evident change in the
+stratospheric Asymemtric component (red line in panel a) between the
+90's, when we see a dominance of extreme negative values, and the 00's,
+when we see the inverse. This change is restricted to the Winter months:
+the linear trend for JJA starting in 1990 for the Asymmetric component
+at 50hPa is \(0.37 \pm 0.22\).
+
+These results point to very different behaviours between the
+stratospheric and tropospheric SAM. \#FIXME
+
+This suggests that both the Asymmetric and the Symemtric component of
+the tropospheric SAM are highly vertically coherent, both in their
+individual evolution and their temporal relationship. This is to be
+expected since the SAM is mostly equivalent barotropic (citaaaa).
 
@@
 regressed monthly geopotential anomalies into both indeces using
-multiple regression (Figure A6 illustrates the difference between
-computing two simple regressions and one multiple regression).
+multiple regression
 
@@
 
-Figure \ref{fig:2d-regr} shows the spatial year-long regression for
-selected levels. In the troposphere the Full annular mode is clearly
-``contaminated'' with well known zonal asymemtries (panels g and j)
-which are succesfully sepparated by our methodology (panels h, i, k and
-l). In the stratosphere, the spatial pattern associated with the Full
-SAM is much more clearly dominated by a zonally symmetric, monopolar
-structure (panels a and d) that is, however, not perfectly centered in
-the south pole. The monopoloe obtained by multiple regression with the
-Asymmetric and Symmetric SAM (panels c and f in Figure
-\ref{fig:2d-regr}) is much more symmetric and the shift from total
-symmetry is captured by the regression pattern of the Asymmetric SAM as
-a wave-1 pattern (panels b and e).
+Figure \ref{fig:2d-regr} shows the spatial year-long regression \#FIXME.
+Column a are regressinos using the Full SAM, while columns b and c are
+regression coefficients computed in a multiple regression of
+geopotential height on the Asymmetric and Symmetric indices at the same
+time. Thus, they are to be interpreted as the patterns associated with
+each index, controling for the (linear) effect of the other. (Figure A6
+\#FIXME illustrates the difference between computing two simple
+regressions and one multiple regression.)
+
+In the stratosphere, the spatial pattern associated with the Full SAM is
+more clearly dominated by a zonally symmetric, monopolar structure
+(panel a.1) which is, however, not perfectly centered in the south pole.
+The monopole obtained by multiple regression with the Asymmetric and
+Symmetric SAM (panel c.1) is much more symmetric and the shift from
+total symmetry is captured by the regression pattern of the Asymmetric
+SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the
+Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel
+b.1).
+
+In the troposphere, panel a.2 show the well known zonally symmetrical
+annular mode \emph{contaminated} with zonal asymmetries in the form of a
+wave-3. The regression using the Asymmetric and Symmetric SAM indices
+successfully disentangle both structures. The Asymmetric component gives
+rise to a cleaner zonal wave (panel b.2) and the Symemtric component is
+assosiated with an trully annular mode, almost devoid of zonal
+asymmetries (panel c.2). \#FIXME (agregar algo más?)
 
@@
 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number
-zero represents the zonal mean. Comparing between rows, this Figure
-quantifies the relatively clean separation between the zonally symmetric
-and zonally asymmetric structures, as its evident how the mixture of
-waves of the Full field (first row) is very similar to the sum of the
-waves of the Asymmetric and Symmetric field (second and third row,
-respectively). The second row of Figure \ref{fig:wave-amplitude} shows
-that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the
-stratosphere (panel b), while in the stratosphere it is composed of
-zonal waves 3 to 1 in decreasing level of importance.
+zero represents the amplitude of the zonal mean. Comparing between rows,
+this Figure quantifies the relatively clean separation between the
+zonally symmetric and zonally asymmetric structures, as its evident how
+the mixture of waves of the Full field (first row) is very similar to
+the sum of the waves of the Asymmetric and Symmetric field (second and
+third row, respectively). The second row of Figure
+\ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly
+dominated by wave 1 in the stratosphere (panel b), while in the
+troposphere it is composed of zonal waves 3 to 1 in decreasing level of
+importance.
 
@@
 \includegraphics{vertical-regression-1} \caption[Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South]{Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South.
-(this caption needs some love)}\label{fig:vertical-regression}
+(\#FIXME this caption needs some love)}\label{fig:vertical-regression}
 \end{figure}
 
-From Figure \ref{fig:2d-regr} it appears that the vertical structure of
-the Asymmetric SAM is equivalent barotropic in the troposphere but
-baroclinic in the stratosphere. Anomalies are centerd in the same
-locations in the troposphere (panels h and k), but show westerly
-displacement in the stratosphere (panesl b and e). This is expanded in
-Figure \ref{fig:vertical-regression}, which shows a vertical crossection
-of the regression coefficient corresponding to the middle row of Figure
-\ref{fig:2d-regr}, area-weighted averaged between 65 and 40 degrees
-South. Below 100 hPa, anomalies are completely vertical, while above
-they show an important westerly tilt with height.
+To analyse the vertical structure of the geopotential anomalies
+asociated with the asymetric SAM index, we show a vertical cross section
+of regressions of mean geopotential height between 65ªS and 40ªS for the
+50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM
+index (panel b) (Figure \ref{fig:vertical-regression}).
+
+The geopotential anomalies associated with the stratospheric SAM (panel
+a) are clearly constrained to the stratosphere, which underscores the
+disconnect between the stratospheric and tropospheric symetric SAM. The
+vertical structure this signal tilts about 60ª to the West between 100
+hPa and 1 hPa, suggesting baroclinic processes and polarward transport
+of heat (\#FIXME is this ok?). Interestingly, the signal in the
+stratosphere maximises near 10 hPa despite using the 50 hPa index for
+the regression.
+
+The tropospheric asymmetric SAM has significant signals that extend
+upwards the uppermost levels of the reanalysis. In the troposphere, the
+wave-3 structure is equivalent barotropic with maximum amplitude at
+roughly 250 hPa. \#FIXME moaaaaar.
+
+Interestingly, the structures shown in Figure
+\ref{fig:vertical-regression} are surprisignly robust to the choice of
+index level. For any stratospheric (above 100 hPa) index, the resulting
+anomalies are very similar to the wave-1 structure with maximum near 10
+hPa in panel a. Conversely, for any tropospheric (below 100 hPa) index,
+the result is very similar to panel b. The pattern cross-correlation
+between levels of each segment of the atmosphere is greater than 0.9
+(Figure A\#fixme). The patterns mainly change in amplitude. The
+tropospheric pattern is maximised by the 300 hPa Asymmetric SAM index
+and the stratospheric pattern increased monotonically with height.
 
@@
 
-\begin{sidewaysfigure}
-\includegraphics{pp-regr-global-1} \caption[Regression pattern of precipiation with Asymmetric and Symmetric SAM]{Regression pattern of precipiation with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:pp-regr-global}
-\end{sidewaysfigure}
+\begin{figure*}
+\includegraphics{pp-regr-global-africa-1} \caption[Regression pattern of precipiation with Asymmetric and Symmetric SAM]{Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
+Africa}\label{fig:pp-regr-global-africa}
+\end{figure*}
+
+\begin{figure*}
+\includegraphics{pp-regr-global-oceania-1} \caption[Same but for oceania]{Same but for oceania}\label{fig:pp-regr-global-oceania}
+\end{figure*}
+
+\begin{figure*}
+\includegraphics{pp-regr-global-america-1} \caption[Same but for america]{Same but for america}\label{fig:pp-regr-global-america}
+\end{figure*}
+
+\subsection{Conclusiones}
+
+This suggests that both the Asymmetric and the Symemtric component of
+the tropospheric SAM are highly vertically coherent, both in their
+individual evolution and their temporal relationship. This is to be
+expected since the SAM is mostly equivalent barotropic (citaaaa).
 
@@
 CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder,
-Colorado, USA, from their Web site at https://psl.noaa.gov/
+Colorado, USA, from their Web site at https://psl.noaa.gov/ \#FIXME
 
@@
 \begin{figure}
-\includegraphics{A5-1} \appendcaption{A5}{Regression pattern of precipiation with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:A5}
-\end{figure}
-
-\begin{figure}
 \includegraphics{A6-1} \appendcaption{A6}{Regressions maps resulting from performing one multiple regression (column a) and from performing two simple regressions (column b)}\label{fig:A6}
@@
 \begin{figure}
-\includegraphics{A8-1} \appendcaption{A8}{Decadal trends of SAM indices for each season. }\label{fig:A8}
+\includegraphics{A8-1} \appendcaption{A8}{Pattern cross-correlation \#FIXME!}\label{fig:A8}
 \end{figure}

### 2020-09-01 analysis/paper/paper.Rmd
casi draft

@@
 lev.lab <- AddSuffix(" hPa")
+no_grid <- theme(panel.grid = element_blank())
 
@@
 combine_lists <- function(...) {
+  # browser(expr = length(x) != length(y))
   X <- list(...)
@@
 }
+
+lab_sam <-  c(full = "Full", 
+              asym = "Asymmetric",
+              sym  = "Symmetric")
 ```
@@
 SAM <- hgt[lat <= -20] %>% 
+  .[lev == 700] %>% 
   .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
-  .[, .(eof = list(eof_flip(EOF(hgt ~ time | lon + lat, n = 1, data = .SD)))), 
+  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
     by = .(lev)]
@@
 
-We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach] for the period 1979 to 2018 (inclusive).
+We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach] for the period 1979 to 2018.
 
@@
   setnames("hgt", "full") %>% 
-  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)] %>% 
-  rm_singleton()
-
-lab_sam <-  c(full = "Full", 
-              asym = "Asymmetric",
-              sym  = "Symmetric")
+  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev)]
 
@@
   .[lev %in% c(700)] %>%
-  melt(id.vars = c("lon", "lat", "lev")) %>% 
-  .[variable == "sym", value := value + rnorm(.N, 0, sd = 1e-6)] %>%
+  rm_singleton() %>% 
+  melt(id.vars = c("lon", "lat")) %>% 
+  .[, value := value + rnorm(.N, sd = 1e-6)] %>% 
   .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)]
@@
   scale_y_latitude(limits = c(NA, -20), labels = NULL) +
+  scale_fill_divergent(guide = "none", high = "#731C1F", low = "#323071") +
   scale_fill_divergent(guide = "none") +
@@
   axis_labs_smol +
+  no_grid +
   tag_facets()
@@
 
-```{r}
-# indices %>% 
-#   copy() %>% 
-#   .[lev %in% c(50, 700)] %>% 
-#   .[, cumestimate := cumsum(estimate_norm), by = .(lev, term, season(time))] %>% 
-#   
-#   
-#   ggplot(aes(time, cumestimate)) +
-#   geom_line(aes(color = term)) +
-#   # ggridges::geom_vridgeline(stat="ydensity", scale = -4e8, fill = NA,
-#   #                           aes(color = term, width = ..density.., x = min(time - 1e7))) +
-#   scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam) +
-#   scale_y_continuous(NULL) +
-#   # scale_x_datetime(NULL, expand = c(0, 0)) +
-#   facet_grid(lev~season(time), labeller = labeller(lev = lev.lab)) +
-#   tag_facets()
-
+```{r test-aao}
+# test that my indices hace the correct sign. The Full index needs to be **positively** correlated
+# with the AAO index.
+aao <- rsoi::download_aao(TRUE, data_path("raw", "aao.csv")) %>% 
+  as.data.table() %>% 
+  .[, .(time = lubridate::as_datetime(Date), aao = AAO)]
+
+cor_aao <- indices[term == "full" & lev == 700] %>% 
+  .[aao, on = .NATURAL] %>% 
+  na.omit() %>% 
+  .[, cor(estimate, aao)]
+
+if (sign(cor_aao) != 1) {
+  stop("AAO and full index are not positively correlated!")
+}
 ```
@@
   copy() %>% 
-  .[, widyr::pairwise_cor(.SD, lev, time, estimate, diag = FALSE), by = .(term)] 
+  .[, widyr::pairwise_cor(.SD, lev, time, estimate, diag = TRUE), by = .(term)] 
 crosscor %>% 
   ggplot(aes(item2, item1)) +
-  geom_contour_fill(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
+  geom_contour_fill(aes(z = correlation), breaks = MakeBreaks(binwidth = 0.1)) +
   geom_contour_tanaka(aes(z = correlation), size = 0.1, na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
-  geom_text_contour(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1), 
+  geom_text_contour(aes(z = correlation), breaks = MakeBreaks(binwidth = 0.1), 
                     color = "white", rotate = FALSE, stroke.color = "black", stroke = 0.2, 
@@
   periodic(lon = c(0, 360)) %>% 
-  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
+  .[, season := "Annual"]
 ```
@@
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
+  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0), global.breaks = FALSE) +
   geom_contour_tanaka(aes(z = estimate),
-                      breaks = ZeroBreaks,
+                      breaks = AnchorBreaks(0, exclude =  0),
                       size = 0.3,
@@
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
-                       limits = c(-85, 85),
+                       limits = c(-45, 45),
                        oob = scales::squish,
-                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
+                       breaks = AnchorBreaks(0, 10, 0)) +
   scale_linetype(guide = "none") +
   coord_polar() +
-  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  facet_grid(lev ~ term) + # labeller = labeller(term = lab_sam, lev = lev.lab)) +
   tag_facets("rc") +
-  axis_labs_smol 
+  axis_labs_smol  +
+  no_grid
 ```
@@
 Figure \ref{fig:2d-regr} shows the spatial year-long regression #FIXME.
-Column a are regressinos using the Full SAM, while columns b and c are regression coefficients computed in a multiple regression of geopotential height on the Asymmetric and Symmetric indices at the same time.
+Column a are regressinos using the Full SAM, while columns 2 and 3 are regression coefficients computed in a multiple regression of geopotential height on the Asymmetric and Symmetric indices at the same time.
 Thus, they are to be interpreted as the patterns associated with each index, controling for the (linear) effect of the other.
@@
 In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the south pole.
-The monopole obtained by multiple regression with the Asymmetric and Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel b.1).
-
-In the troposphere, panel a.2 show the well known zonally symmetrical annular mode *contaminated* with zonal asymmetries in the form of a wave-3.
+The monopole obtained by multiple regression with the Asymmetric and Symmetric SAM (panel a.3) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel a.2).
+
+In the troposphere, panel b.1 shows the well known zonally symmetrical annular mode *contaminated* with zonal asymmetries in the form of a wave-3.
 The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
-The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel c.2).
+The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel b.3).
+note that the wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern and asociated with @raphael2004's ZW3 index (see Figure 1 from that paper).
 #FIXME (agregar algo más?)
 
-```{r wave-amplitude, fig.cap = "Planteray wave amplitude for the regression patterns at 50 and 700 hPa.", fig.width=3, fig.height=4}
+```{r calculate-raphael}
+# 50°E, 166°E and 76°W. 
+raphael <- ReadNetCDF(ERA5(), 
+           subset = list(time = c("1979-01-01", "2018-12-31"),
+                         latitude = -49,
+                         level = 500),
+           vars = c(hgt = "z")) %>% 
+  normalise_coords() %>% 
+  .[, hgt := hgt/9.8] %>% 
+  .[lon %~% ConvertLongitude(c(-50, -166, 76))] %>% 
+  .[, hgt_s := frollmean(hgt, 3, NA, align = "center"), by = .(lon, lat)] %>% 
+  na.omit() %>% 
+  .[, hgt_s := scale(hgt_s), by = .(lon, lat, month(time))] %>% 
+  .[, .(raphael = mean(hgt_s)), by = time]
+```
+
+```{r wave-amplitude, fig.cap = "Planteray wave amplitude for the regression patterns at 50 and 700 hPa. Note the varying x axis.", fig.width=3, fig.height=3}
 indices_2d_yearly %>% 
@@
   coord_flip() +
-  facet_grid(term~lev, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free_x") +
-  tag_facets(position = "tr")
+  facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free") +
+  tag_facets("cr", position = "tr")  +
+  no_grid +
+  theme(strip.text = element_text(size = 7))
 ```
@@
 The amplitude of each zonal wave number at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
-Comparing between rows, this Figure quantifies the relatively clean separation between the zonally symmetric and zonally asymmetric structures, as its evident how the mixture of waves of the Full field (first row) is very similar to the sum of the waves of the Asymmetric and Symmetric field (second and third row, respectively).
-The second row of Figure \ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance.
+Comparing between rows, this Figure quantifies the relatively clean separation between the zonally symmetric and zonally asymmetric structures, as its evident how the mixture of waves of the Full field (column a) is very similar to the sum of the waves of the Asymmetric and Symmetric field (columns b and c, respectively).
+Column b of Figure \ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b.1), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel b.2).
 
@@
 
-<!-- ```{r compute-regr-pattern-season} -->
-<!-- indices_2d <- hgt[lev %in% main_levs] %>%  -->
-<!--   indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>%  -->
-<!--   .[, rbind(as.data.table(FitLm(hgt_a, sym, asym)), -->
-<!--             as.data.table(FitLm(hgt_a, full))), -->
-<!--     by = .(lon, lat, lev, season(time))] %>%  -->
-<!--   rm_intercept() %>%  -->
-<!--   periodic(lon = c(0, 360)) %>%  -->
-<!--   .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)]  -->
-<!-- ``` -->
-
-<!-- ```{r 2d-regr-700, fig.cap = "Seasonal regression patterns of geopotential height at 700 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"} -->
-<!-- indices_2d %>%  -->
-<!--   .[lat <= -20] %>%  -->
-<!--   .[lev == 700] %>%  -->
-<!--   ggplot(aes(lon, lat)) + -->
-<!--   geom_contour_fill(aes(z = estimate),  breaks = AnchorBreaks(0, binwidth = 5, 0), -->
-<!--                     global.breaks = FALSE) + -->
-<!--   geom_contour_tanaka(aes(z = estimate), -->
-<!--                       breaks = AnchorBreaks(0, binwidth = 5, 0), -->
-<!--                       size = 0.3, -->
-<!--                       global.breaks = FALSE) + -->
-<!--   geom_qmap(~.x[lat <= -20]) + -->
-<!--   scale_x_longitude() + -->
-<!--   scale_y_latitude(limits = c(NA, -20), labels = NULL) + -->
-<!--   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), -->
-<!--                        limits = c(-55, 55), -->
-<!--                        oob = scales::squish, -->
-<!--                        breaks = AnchorBreaks(0, binwidth = 5, 0)) + -->
-<!--   scale_linetype(guide = "none") + -->
-<!--   coord_polar() + -->
-<!--   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) + -->
-<!--   tag_facets() + -->
-<!--   axis_labs_smol -->
-<!-- ``` -->
-
-<!-- ```{r 2d-regr-30, fig.cap = "Seasonal regression patterns of geopotential height at 30 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.width =10, fig.height=7, fig.env = "sidewaysfigure"} -->
-<!-- indices_2d %>%  -->
-<!--   .[lat <= -20] %>%  -->
-<!--   .[lev == 30] %>%  -->
-<!--   ggplot(aes(lon, lat)) + -->
-<!--   geom_contour_fill(aes(z = estimate),  breaks = AnchorBreaks(0, binwidth = 20, 0), -->
-<!--                     global.breaks = FALSE) + -->
-<!--   geom_contour_tanaka(aes(z = estimate), -->
-<!--                       breaks = AnchorBreaks(0, binwidth = 20, 0), -->
-<!--                       size = 0.3, -->
-<!--                       global.breaks = FALSE) + -->
-<!--   geom_qmap(~.x[lat <= -20]) + -->
-<!--   scale_x_longitude() + -->
-<!--   scale_y_latitude(limits = c(NA, -20), labels = NULL) + -->
-<!--   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), -->
-<!--                        limits = c(-105, 105), -->
-<!--                        oob = scales::squish, -->
-<!--                        breaks = AnchorBreaks(0, binwidth = 20, 0)) + -->
-<!--   scale_linetype(guide = "none") + -->
-<!--   coord_polar() + -->
-<!--   facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) + -->
-<!--   tag_facets() + -->
-<!--   axis_labs_smol -->
-<!-- ``` -->
+```{r enso-cor}
+mei_sam_cor <- rsoi::download_mei(TRUE, data_path("derived", "mei.csv")) %>% 
+  as.data.table() %>% 
+  .[, .(time = lubridate::as_datetime(Date), mei = MEI)] %>% 
+  .[indices, on = "time"] %>% 
+  .[, with(cor.test(estimate, mei), 
+           list(cor = estimate, 
+                low = conf.int[1], 
+                upp = conf.int[2])), by = .(lev, term)] 
+# enso_sam_cor %>% 
+#   ggplot(aes(lev, cor)) +
+#   geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha = 0.2) +
+#   geom_line(aes(color = term)) +
+#   scale_x_level() +
+#   coord_flip()
+
+cor_texts <- mei_sam_cor[lev == 700, 
+                          paste0(signif(cor, 2)), # " (CI: ", signif(low, 2), "–", signif(upp, 2), ")"), 
+                          by = term] %>% 
+  dcast(1~term)
+
+mei_sam_partial <- rsoi::download_mei(TRUE, data_path("derived", "mei.csv")) %>% 
+  as.data.table() %>% 
+  .[, .(time = lubridate::as_datetime(Date), mei = MEI)] %>% 
+  .[indices_wide, on = "time"] %>% 
+  .[lev == 700] %>% 
+  .[, partial_cor(mei, asym, sym)] %>% 
+  .[, partial_correlation := signif(partial_correlation, 2)] %>% 
+  dcast(.~term)
+```
+
+The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the teleconnection pattern associated with the ENSO.
+Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
+The correlation between the full SAM and the ENSO as measured by the Multivariate Enso Index [@wolter2011] is `r cor_texts$full`.
+This relationship is captured entirely the Asymmetric SAM, as this index has a partial correlation of `r mei_sam_partial$asym` with the MEI, whereas the Symmetric SAM has null partial correlation with the MEI.
 
@@
 
+# mei <- rsoi::download_mei() %>% 
+#   as.data.table() %>% 
+#   .[, .(time = lubridate::as_datetime(Date), mei = MEI)]
+
 air_regr <- air %>% 
@@
   .[indices_wide[lev == 700], on = .NATURAL] %>% 
+  # mei[., on = .NATURAL] %>% 
   rm_singleton() %>% 
@@
 P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
-Gray areas have more than 15\\% of missing data.", fig.width = 10, fig.height = 7, fig.env = "sidewaysfigure"}
+Gray areas have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*"}
 air_regr %>% 
+  # .[season == "DJF"] %>% 
   copy() %>% 
@@
   .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
-  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
+  # .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>%
   periodic(lon = c(0, 360)) %>% 
@@
                     breaks = AnchorBreaks(0, 0.15, 0)) +
+  geom_contour_tanaka(aes(z = estimate), na.fill = 0,
+                      breaks = AnchorBreaks(0, 0.15, 0)) +
   # geom_raster(aes(fill = estimate)) +
-  geom_contour2(aes(z = p.value), breaks = 0.05) +
-  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 0.5, fill = "gray", color = "gray30") +
+  geom_contour_fill(aes(z = p.value), breaks = c(0, 0.05), color = "black", fill = "gray50", alpha = 0.2) +
+  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), fill = "gray", color = "gray30") +
   geom_qmap(~.x[lat <= 0]) +
@@
   coord_polar() +
-  facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  facet_grid(season ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
   tag_facets("cr", tag_suffix = "") +
+  no_grid +
   axis_labs_smol
@@
 
-??
+Regression of the SAM indicies with seasonal mean precipitation are shown Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australia and New Zealand, and South America respectively.
+(We didn't detect any significant signal in South Africa.)
+
+```{r}
+indices_2d_seasonal <- hgt[lev %in% 700] %>% 
+  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>% 
+  .[, combine_lists(FitLm(hgt_a, sym, asym, se = TRUE),
+                    FitLm(hgt_a, full, se = TRUE)),
+    by = .(lon, lat, season(time))] %>% 
+  rm_intercept() %>% 
+  .[, p.val := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
+  periodic(lon = c(0, 360)) %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)]
+
+indices_2d <- rbind(indices_2d_yearly[lev == 700] %>% .[, lev := NULL], indices_2d_seasonal) %>% 
+  .[, season := forcats::fct_relevel(season, "Annual")] 
+```
 
@@
 
-pp_regr_season <- CPCC() %>% 
-  ReadNetCDF(subset = list(lon = lapply(continents, function(x) x$lon),
-                           lat = lapply(continents, function(x) x$lat))) %>% 
+pp <- CPCC() %>% 
+  ReadNetCDF(subset = list(lat = c(-90, 0),
+                           time = range(indices$time))) %>% 
   na.omit() %>% 
@@
   .[, days := lubridate::days_in_month(time[1]), by = time] %>% 
-  .[, precip := precip/days] %>% 
-  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
+  .[, precip := precip/days]
+
+# mei <- rsoi::download_mei() %>% 
+#   as.data.table() %>% 
+#   .[, .(time = lubridate::as_datetime(Date), mei = MEI)]
+
+pp_regr_season <- pp %>% 
   .[indices_wide[lev == 700], on = .NATURAL] %>% 
-  rm_singleton() %>% 
-  add_continent() %>% 
-  .[!is.na(continent)] %>% 
-  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, continent, time = seasonally(time))] %>%
-  .[, `:=`(precip_a = Anomaly(precip),
-           precip_a_p = log(precip) - log(mean(precip, na.rm = TRUE))),
-    by = .(lon, lat, season(time))] %>% 
+  na.omit() %>% 
+  .[, precip_a := Anomaly(precip), by = .(lon, lat, month(time))] %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
+  .[, precip_a := Anomaly(precip), by = .(lon, lat, season(time))] %>%
   .[, nas := mean(is.na(precip)), by = .(lon,  lat, season(time))] %>% 
   .[nas >= 0.15, precip := NA] %>% 
-  na.omit() %>% 
-  .[is.finite(precip_a_p)] %>% 
+  # na.omit() %>% 
+  .[is.finite(precip_a)] %>%
   .[, combine_lists(FitLm(precip_a, asym, sym, se = TRUE),
@@
     by = .(lon, lat, season(time))] %>% 
-  rm_intercept()
-
-filter_sa <- function(data, lons = c(265, 325), lats = c(-80, -10)) {
-  data[is.sa(data, lons = lons, lats = lats)]
-}
-
+  rm_intercept() %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)]
+
+pp_regr_annual <- pp %>% 
+  .[indices_wide[lev == 700], on = .NATURAL] %>% 
+  na.omit() %>% 
+  .[, precip_a := Anomaly(precip), by = .(lon, lat, month(time))] %>% 
+  .[, nas := mean(is.na(precip)), by = .(lon,  lat)] %>%
+  .[nas >= 0.15, precip := NA] %>% 
+  .[is.finite(precip_a)] %>%
+  .[, combine_lists(FitLm(precip_a, asym, sym, se = TRUE),
+                    FitLm(precip_a, full, se = TRUE)),
+    by = .(lon, lat)] %>% 
+  rm_intercept() %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
+  .[, season := "Annual"]
+pp_regr <- rbind(pp_regr_season, pp_regr_annual) %>% 
+  .[, season := forcats::fct_relevel(season, "Annual")] 
 ```
@@
 ```{r def-plot_pp}
-plot_pp <- function(this_continent) {
-  pp_regr_season %>%
+plot_pp <- function(this_continent, data = pp_regr) {
+  data %>%
+    .[term != "mei"] %>% 
     add_continent() %>% 
@@
     .[continent == this_continent] %>%
-    .[, p.value := Pvaluate(estimate, std.error, df, "none"), by = .(season, continent, term)] %>%
-    .[, p.value2 := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, continent, term)] %>%
+    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, continent, term)] %>%
     ggplot(aes(lon, lat)) +
-    geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.5, 0)) +
+    geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.1, 0)) +
+    stat_subset(aes(subset = is.na(estimate)), geom = "tile", color = "gray50") +
     # geom_raster(aes(fill = estimate/30)) +
-    geom_contour2(aes(z = p.value2), breaks = 0.01) +
+    geom_contour2(aes(z = p.value), breaks = 0.05) +
     # geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = 1) +
-    # stat_subset(aes(subset = p.value <= 0.01 & is.cross(lon, lat, 1)), size = 0.4, alpha = 0.3) +
+    # stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat, 2)), size = 0.4, alpha = 0.8) +
+    geom_contour2(data = indices_2d %>% copy() %>% add_continent() %>% .[continent == this_continent],
+                  color = "gray50", size = 0.2, global.breaks = FALSE, 
+                aes(z = estimate, linetype = factor(-sign(..level..)))) +
     geom_qmap(~.x %>% add_continent() %>% .[continent == this_continent], keep = 0.015, weighting = 0.9) +
-    scale_x_longitude(ticks = 15) +
-    scale_y_latitude(ticks = 15, labels = NULL) +
+    scale_x_longitude(breaks = NULL) +
+    scale_y_latitude(breaks = NULL) +
     scale_fill_divergent("Precipitation", 
-                         limits = c(-3.4, 3.4),
+                         limits = c(-.8, .8),
+                         oob = scales::squish,
+                         # label = scales::percent_format(),
+                         guide = guide_colorstrip_bottom(), 
+                         breaks = AnchorBreaks(0, 0.1, 0)) +
+    scale_linetype(guide = "none") +
+    coord_quickmap(xlim = continents[[this_continent]]$lon, 
+                   ylim = continents[[this_continent]]$lat) +
+    facet_grid(season ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+    tag_facets("cr", position = "bl")  +
+    no_grid +
+    axis_labs_smol
+}
+
+```
+
+```{r pp-regr-africa,  fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
+Africa", fig.width = 6, fig.height = 4.5, fig.env = "figure*"}
+# plot_pp("africa") 
+```
+
+```{r pp-regr-oceania, dependson=plot_pp,  fig.cap = "Same but for oceania", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
+plot_pp("oceania") 
+```
+
+In Australia (Figure \ref{fig:pp-regr-oceania}), the annual-level regression shows that the Full SAM is associated with a statistically significant increase in precipitation in the Southeastern region (panel a.1), which reproduces the results from @gillett2006.
+The separation between Asymmetric and Symmetric SAM suggest that this increase is explained by the Symmetric SAM only in the East coast (panel c.1), which is consistent with the increased easterly flow clearly seen in relation with this index.
+The Asymmetric SAM appears related to increased precipitation in the West coast of Southeastern Australia (panel b.2), explained by the anomalous *westerly* circulation transporting moist air to the continent.
+
+The seasonal-level regressions show statistically significant anomalies only in SON, with a pattern similar to the annual-level regression (panel a.5).
+Panels b.5 and c.5 don't show a clear separation between the Asymmetric and Symmetric SAM.
+If anything, the positive and more significant regression coefficients in panel b.5 vs pane c.5 would suggest more influence of the Asymmetric than the Symmetric SAM, going against the interpretation gathered from the annual-level regressions.
+This Spring signal is broadly consistent with @hendon2007, but whereas @hendon2007 also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coeffcients have the consistent sign).
+
+```{r pp-regr-america, dependson = plot_pp, fig.cap = "Same but for america", fig.width = 6, fig.height = 7.6, fig.env = "figure*"}
+plot_pp("america")
+
+```
+
+In South America (Figure \ref{fig:pp-regr-america}) , the annual-level regression shows that the SAM is associated with statistically significant precipitation decreases both in Southeastern South America and Southern Chile (panel a.1).
+Both signals were observed by @silvestri2009 and the former also by @gillett2006.
+Panels b.1 and c.1 show a remarkable clean separation between the Asymmeric SAM --associated with the signal located in Southestern South America-- and the Symmetric SAM --associated with the signal in Chile.
+This separation is consistent with the mechanisms responsible for these effects.
+In Southeastern South America, anomalous meridional winds lead to less precipitation by inhibiting moisture convergence from the South American Low Level Jet [@silvestri2009].
+In Southern Chile, the reduced westerly flow reduce moisture transport from the Pacific Ocean (cita??).
+The is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 and that is explained by the Asymmmetric SAM.
+
+The negative relationship between precipiation and SAM index in Southeastern South America and Southern Chile is evident in all seasons except Winter (although not always stastitically significant) and it appears to maximise in Autumn.
+
+### Sea ice
+
+```{r read-ice}
+ice <- readRDS(data_path("derived", "ice.Rds"))
+ice_proj <- readRDS(data_path("derived", "ice_proj.Rds"))
+
+new_proj <- "+proj=stere +lat_0=-90 +lat_ts=-70 +lon_0=-180 +k=1 +x_0=0 +y_0=0 +a=6378273 +b=6356889.449 +units=m +no_defs"
+
+project_ice <- function(data) {
+  browser()
+  data[, c("x", "y") := proj4::project(list(x, y), proj = ice_proj, inverse = TRUE)] %>% 
+    .[, x := ConvertLongitude(x, group = group)] %>% 
+    .[]
+
+}
+```
+
+```{r compte-ice-regression}
+ice_regr <- ice %>% 
+  .[indices_wide[lev == 700], on = .NATURAL] %>% 
+  # .[season(time) %in% c("JJA", "SON")] %>%
+  .[concentration <= 0.15, concentration := 0] %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(x, y, time = seasonally(time))] %>% 
+  # .[season(time) == "SON"] %>% 
+  .[, combine_lists(FitLm(concentration, full, se = TRUE),
+                    FitLm(concentration, asym, sym, se = TRUE)),
+    by = .(x, y, season(time))] %>%
+  rm_intercept() 
+```
+
+```{r regr-ice, fig.cap = "Seasonal regression of SAM indices with sea ice concentration. \\#FIXME", fig.env="figure*",  fig.width = 6, fig.height = 8}
+ice_regr %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>%
+  .[, pval := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)] %>%
+  .[, c("lon", "lat") :=  proj4::project(list(x, y), proj = ice_proj, inverse = TRUE)] %>% 
+  # .[season == "SON"] %>%
+  .[, lon := ConvertLongitude(lon)] %>% 
+  na.omit() %>% 
+  .[estimate != 0] %>%
+
+  ggplot(aes(lon, lat)) +
+  scattermore::geom_scattermore(aes(color = estimate)) +
+  stat_subset(aes(subset = pval < 0.01 & is.cross(x, y, 3)),
+              geom = "point", size = 0.01, alpha = .6) +
+  geom_contour2(data = indices_2d_seasonal[lat <= -50], color = "gray50", size = 0.3,
+                aes(z = estimate, linetype = factor(-sign(..level..)))) +
+  geom_qmap(~.x[lat <= -60]) +
+  scale_color_divergent("Sea Ice",
+                        limits = c(-.1, .1), 
                         oob = scales::squish, 
                         guide = guide_colorstrip_bottom(),
-                         breaks = AnchorBreaks(0, 0.5, 0)) +
-    coord_quickmap(xlim = continents[[this_continent]]$lon, 
-                   ylim = continents[[this_continent]]$lat) +
-    facet_grid(term ~ season, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-    tag_facets(position = "br") +
+                        breaks = AnchorBreaks(0, 0.025, 0)) +
+  scale_x_longitude() +
+  scale_y_latitude(limit = c(-90, NA), labels = NULL) +
+  scale_linetype(guide = "none") +
+  coord_polar() +
+  facet_grid(season ~ term, labeller = labeller(term = lab_sam)) +
+  tag_facets("cr") +
+  no_grid +
   axis_labs_smol
-}
-```
-
-```{r pp-regr-global-africa,  fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
-Africa", fig.width = 6, fig.height = 4.5, fig.env = "figure*"}
-plot_pp("africa")
-
-```
-
-```{r pp-regr-global-oceania,  fig.cap = "Same but for oceania", fig.width = 6, fig.height = 4.5, fig.env = "figure*"}
-plot_pp("oceania") +
-  tag_facets(position = "bl")
-
-```
-
-```{r pp-regr-global-america,  fig.cap = "Same but for america", fig.width = 6, fig.height = 6, fig.env = "figure*"}
-plot_pp("america")
-
-```
-
-## Conclusiones
+```
+
+Regressions between the Full SAM index and Antarctic Sea Ice Concentrations (Figure \ref{fig:regr-ice}) show a great deal of variability across seasons.
+The only statistically significant signal is in Spring, when we observe negative concentraion anomalies in the Northen Weddell Sea (panel a.4) explained by the Asymmetric SAM (panel b.4).
+Both in Winter and in Spring the Asymmetric SAM is associated with bigger Sea Ice Concentration anomalies in West Antarctica than East Antarctica, with generally decreased concentration East of the Antarctic Peninsula and increased concentration to the West, as expected from the anomalous circulation correlated with this index.
+The Symmetric SAM signal appears more evenly distributed accross the whole ice sheet.
+
+## Conclusions
+
+@silvestri2009 showed that impacts linked to the SAM changed rather dramatically before and after 1980, and that this change can be explained by changes in the regional-scale structure of the SAM-related circulation.
+In particular, the decreased in precipitation in South America (Figure \ref{fig:pp-regr-america} panel a.1) was absent or even switched sign.
+
+@silvestri2009 also thowed
+
+Therefore, is very likely that if we were to repeat this analysis using pre-satellite data, the resulting Asymmetric SAM would look very differnt.
 
@@
 This is to be expected since the SAM is mostly equivalent barotropic (citaaaa).
+
+@hosking2013 observed a pattern similar to the one near the Amundsen Sea in panel a.4 associated with the strengthening of the ASL.
+The positive anomalies of SIC near the Amundsen sea being inconsistent with the anomalies of meridional wind associated with the ASL.
+Row 4 in Figure \ref{fig:regr-ice} might shed some light to resolve this mistery.
+The SAM is evidently related to the ASL through it's Asymmetric component, and since the ASymmetric SAM is correlated with the Symmetric SAM, the ALS will be rleated to the symemtric SAM.
+Consiering the similarities between the the SIC anomalies observed by @hosking2013 (in its Figure 7) and Figure \ref{fig:regr-ice} panel a.4, it is very likely that what was observed in that paper was a residual effect of the SAM.
+#FIXME!!
 

### 2020-09-01 analysis/paper/paper.tex
casi draft

@@
 We used monthly geopotential height at 2.5 longitude by 2.5 latitude
-resolution from ERA5 \citep{hersbach} for the period 1979 to 2018
-(inclusive).
+resolution from ERA5 \citep{hersbach} for the period 1979 to 2018.
 
@@
 Figure \ref{fig:2d-regr} shows the spatial year-long regression \#FIXME.
-Column a are regressinos using the Full SAM, while columns b and c are
+Column a are regressinos using the Full SAM, while columns 2 and 3 are
 regression coefficients computed in a multiple regression of
@@
 The monopole obtained by multiple regression with the Asymmetric and
-Symmetric SAM (panel c.1) is much more symmetric and the shift from
+Symmetric SAM (panel a.3) is much more symmetric and the shift from
 total symmetry is captured by the regression pattern of the Asymmetric
@@
 Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel
-b.1).
+a.2).
 
-In the troposphere, panel a.2 show the well known zonally symmetrical
+In the troposphere, panel b.1 shows the well known zonally symmetrical
 annular mode \emph{contaminated} with zonal asymmetries in the form of a
@@
 assosiated with an trully annular mode, almost devoid of zonal
-asymmetries (panel c.2). \#FIXME (agregar algo más?)
+asymmetries (panel b.3). note that the wave-3 pattern observed in panel
+b.2 is rotated by half a wavelength from the average position of the
+mean wave-3 pattern and asociated with \citet{raphael2004}'s ZW3 index
+(see Figure 1 from that paper). \#FIXME (agregar algo más?)
 
 \begin{figure}
-\includegraphics{wave-amplitude-1} \caption[Planteray wave amplitude for the regression patterns at 50 and 700 hPa]{Planteray wave amplitude for the regression patterns at 50 and 700 hPa.}\label{fig:wave-amplitude}
+\includegraphics{wave-amplitude-1} \caption[Planteray wave amplitude for the regression patterns at 50 and 700 hPa]{Planteray wave amplitude for the regression patterns at 50 and 700 hPa.
+Note the varying x axis.}\label{fig:wave-amplitude}
 \end{figure}
@@
 zonally symmetric and zonally asymmetric structures, as its evident how
-the mixture of waves of the Full field (first row) is very similar to
-the sum of the waves of the Asymmetric and Symmetric field (second and
-third row, respectively). The second row of Figure
-\ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly
-dominated by wave 1 in the stratosphere (panel b), while in the
-troposphere it is composed of zonal waves 3 to 1 in decreasing level of
-importance.
+the mixture of waves of the Full field (column a) is very similar to the
+sum of the waves of the Asymmetric and Symmetric field (columns b and c,
+respectively). Column b of Figure \ref{fig:wave-amplitude} shows that
+the Asymmetric SAM is overwhelmingly dominated by wave 1 in the
+stratosphere (panel b.1), while in the troposphere it is composed of
+zonal waves 3 to 1 in decreasing level of importance (panel b.2).
 
@@
 
+The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very
+similar to the teleconnection pattern associated with the ENSO. Indeed,
+\citet{fogt2011} showed that there is a significant relationship between
+the SAM and the ENSO. The correlation between the full SAM and the ENSO
+as measured by the Multivariate Enso Index \citep{wolter2011} is -0.19.
+This relationship is captured entirely the Asymmetric SAM, as this index
+has a partial correlation of -0.27 with the MEI, whereas the Symmetric
+SAM has null partial correlation with the MEI.
+
 \subsection{Impacts}
@@
 
-\begin{sidewaysfigure}
+\begin{figure*}
 \includegraphics{regr-air-season-1} \caption[Regression pattern of surface temperature with Asymmetric and Symmetric SAM]{Regression pattern of surface temperature with Asymmetric and Symmetric SAM.
@@
 Gray areas have more than 15\% of missing data.}\label{fig:regr-air-season}
-\end{sidewaysfigure}
+\end{figure*}
 
@@
 
-??
+Regression of the SAM indicies with seasonal mean precipitation are
+shown Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america}
+for Australia and New Zealand, and South America respectively. (We
+didn't detect any significant signal in South Africa.)
 
 \begin{figure*}
-\includegraphics{pp-regr-global-africa-1} \caption[Regression pattern of precipiation with Asymmetric and Symmetric SAM]{Regression pattern of precipiation with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
-Africa}\label{fig:pp-regr-global-africa}
+\includegraphics{pp-regr-oceania-1} \caption[Same but for oceania]{Same but for oceania}\label{fig:pp-regr-oceania}
 \end{figure*}
 
+In Australia (Figure \ref{fig:pp-regr-oceania}), the annual-level
+regression shows that the Full SAM is associated with a statistically
+significant increase in precipitation in the Southeastern region (panel
+a.1), which reproduces the results from \citet{gillett2006}. The
+separation between Asymmetric and Symmetric SAM suggest that this
+increase is explained by the Symmetric SAM only in the East coast (panel
+c.1), which is consistent with the increased easterly flow clearly seen
+in relation with this index. The Asymmetric SAM appears related to
+increased precipitation in the West coast of Southeastern Australia
+(panel b.2), explained by the anomalous \emph{westerly} circulation
+transporting moist air to the continent.
+
+The seasonal-level regressions show statistically significant anomalies
+only in SON, with a pattern similar to the annual-level regression
+(panel a.5). Panels b.5 and c.5 don't show a clear separation between
+the Asymmetric and Symmetric SAM. If anything, the positive and more
+significant regression coefficients in panel b.5 vs pane c.5 would
+suggest more influence of the Asymmetric than the Symmetric SAM, going
+against the interpretation gathered from the annual-level regressions.
+This Spring signal is broadly consistent with \citet{hendon2007}, but
+whereas \citet{hendon2007} also detected a strong signal in Summer,
+panel a.2 shows no statistically significant association (although the
+coeffcients have the consistent sign).
+
 \begin{figure*}
-\includegraphics{pp-regr-global-oceania-1} \caption[Same but for oceania]{Same but for oceania}\label{fig:pp-regr-global-oceania}
+\includegraphics{pp-regr-america-1} \caption[Same but for america]{Same but for america}\label{fig:pp-regr-america}
 \end{figure*}
 
+In South America (Figure \ref{fig:pp-regr-america}) , the annual-level
+regression shows that the SAM is associated with statistically
+significant precipitation decreases both in Southeastern South America
+and Southern Chile (panel a.1). Both signals were observed by
+\citet{silvestri2009} and the former also by \citet{gillett2006}. Panels
+b.1 and c.1 show a remarkable clean separation between the Asymmeric SAM
+--associated with the signal located in Southestern South America-- and
+the Symmetric SAM --associated with the signal in Chile. This separation
+is consistent with the mechanisms responsible for these effects. In
+Southeastern South America, anomalous meridional winds lead to less
+precipitation by inhibiting moisture convergence from the South American
+Low Level Jet \citep{silvestri2009}. In Southern Chile, the reduced
+westerly flow reduce moisture transport from the Pacific Ocean (cita??).
+The is a small area of increased precipitation with SAM near central
+Argentina which is also present in the station-based analysis by
+\citet{gillett2006} and that is explained by the Asymmmetric SAM.
+
+The negative relationship between precipiation and SAM index in
+Southeastern South America and Southern Chile is evident in all seasons
+except Winter (although not always stastitically significant) and it
+appears to maximise in Autumn.
+
+\subsubsection{Sea ice}
+
 \begin{figure*}
-\includegraphics{pp-regr-global-america-1} \caption[Same but for america]{Same but for america}\label{fig:pp-regr-global-america}
+\includegraphics{regr-ice-1} \caption[Seasonal regression of SAM indices with sea ice concentration]{Seasonal regression of SAM indices with sea ice concentration. \#FIXME}\label{fig:regr-ice}
 \end{figure*}
 
-\subsection{Conclusiones}
+Regressions between the Full SAM index and Antarctic Sea Ice
+Concentrations (Figure \ref{fig:regr-ice}) show a great deal of
+variability across seasons. The only statistically significant signal is
+in Spring, when we observe negative concentraion anomalies in the
+Northen Weddell Sea (panel a.4) explained by the Asymmetric SAM (panel
+b.4). Both in Winter and in Spring the Asymmetric SAM is associated with
+bigger Sea Ice Concentration anomalies in West Antarctica than East
+Antarctica, with generally decreased concentration East of the Antarctic
+Peninsula and increased concentration to the West, as expected from the
+anomalous circulation correlated with this index. The Symmetric SAM
+signal appears more evenly distributed accross the whole ice sheet.
+
+\subsection{Conclusions}
+
+\citet{silvestri2009} showed that impacts linked to the SAM changed
+rather dramatically before and after 1980, and that this change can be
+explained by changes in the regional-scale structure of the SAM-related
+circulation. In particular, the decreased in precipitation in South
+America (Figure \ref{fig:pp-regr-america} panel a.1) was absent or even
+switched sign.
+
+\citet{silvestri2009} also thowed
+
+Therefore, is very likely that if we were to repeat this analysis using
+pre-satellite data, the resulting Asymmetric SAM would look very
+differnt.
 
@@
 expected since the SAM is mostly equivalent barotropic (citaaaa).
+
+\citet{hosking2013} observed a pattern similar to the one near the
+Amundsen Sea in panel a.4 associated with the strengthening of the ASL.
+The positive anomalies of SIC near the Amundsen sea being inconsistent
+with the anomalies of meridional wind associated with the ASL. Row 4 in
+Figure \ref{fig:regr-ice} might shed some light to resolve this mistery.
+The SAM is evidently related to the ASL through it's Asymmetric
+component, and since the ASymmetric SAM is correlated with the Symmetric
+SAM, the ALS will be rleated to the symemtric SAM. Consiering the
+similarities between the the SIC anomalies observed by
+\citet{hosking2013} (in its Figure 7) and Figure \ref{fig:regr-ice}
+panel a.4, it is very likely that what was observed in that paper was a
+residual effect of the SAM. \#FIXME!!
 

### 2020-09-01 R/data_locations.R

@@ -66,6 +66,15 @@ CMAP <- function() {
 
 #' @export
 #' @rdname data_locations
+GPCP <- function() {
+   file <- data_path("raw", "gpcp.precip.mon.mean.nc")
+   checkmate::assert_access(file, access = "r")
+   class(file) <- c("nc_file", class(file))
+   return(file)
+}
+
+#' @export
+#' @rdname data_locations
 CPCC <- function() {
    file <- data_path("raw", "precip.mon.total.v2018.nc")
    checkmate::assert_access(file, access = "r")

### 2020-09-01 R/helperfun.R

@@ -16,7 +16,6 @@
 # # library(RColorBrewer)
 # enableJIT(0)
 #
-# # source(here::here("scripts/eof_methods.R"))
 #
 #
 # # Repetir vector en negativo

### 2020-09-01 R/lm_lite.R

@@ -35,6 +35,14 @@ var.wt <- function(x, w, na.rm = FALSE) {
   (sum(w*x^2) * sum.w - sum(w*x)^2) / (sum.w^2 - sum(w^2))
 }
 
+#' @export
+Detrend <- function(y, x) {
+   nas <- is.na(y)
+   m <- mean(y, na.rm = TRUE)
+   if (!hasArg(x)) x <- seq_along(y)
+   y[!nas] <- .lm.fit(cbind(1, x[!nas]), y[!nas])$residuals
+   return(y + m)
+}
 
 #' Compute p.values
 #'

### 2020-09-01 analysis/scripts/01-get-data.R

@@ -15,6 +15,7 @@ files <- list(aao = "aao.csv",
               cmap = "precip.mon.mean.nc",
               hadcrut = "HadCRUT.4.6.0.0.median.nc",
               cpcc = "precip.mon.total.v2018.nc",
+              gpcp = "gpcp.precip.mon.mean.nc",
               era5 =  "era5.z.mon.mean.nc")
 
 dir.create(asymsam::data_path("raw"), recursive = TRUE, showWarnings = FALSE)
@@ -79,6 +80,15 @@ if (!force && file.exists(asymsam::data_path("raw", file))) {
   download.file(url, destfile = asymsam::data_path("raw", file))
 }
 
+# Download GPCP -----------------------------------------------------------
+url <- "ftp://ftp.cdc.noaa.gov/Datasets/gpcp/precip.mon.mean.nc"
+file <- files$gpcp
+
+if (!force && file.exists(asymsam::data_path("raw", file))) {
+  message("File already exists.")
+} else {
+  download.file(url, destfile = asymsam::data_path("raw", file))
+}
 
 
 # Download ERA5 -----------------------------------------------------------
@@ -133,3 +143,72 @@ if (!force && file.exists(asymsam::data_path("raw", file))) {
 # ncepfiles <- list.files(asymsam::data_path("raw", "ncep"), full.names = TRUE)
 #
 # asymsam::nc_concatenate(ncepfiles, asymsam::data_path("raw", "ncep.daily.nc"), overwrite = TRUE)
+
+# Antarctic sea ice
+# from https://nsidc.org/data/G02135
+months <- 1:12
+years <- 1979:2019
+dates <- data.table::CJ(years, months)
+# There are no data from 3 December 1987 to 13 January 1988 due to satellite problems. :sob:
+dates <- dates[!(paste0(years, months) %in% c(198712, 19881))]
+
+make_url <- function(year, month) {
+  month_zero <- formatC(month, width = 2, flag = "0")
+  month_Abb <- month.abb[month]
+
+  file_name <- paste0("S_", paste0(year, month_zero), "_concentration_v3.0.tif")
+  url <- paste0("ftp://sidads.colorado.edu/DATASETS/NOAA/G02135/south/monthly/geotiff/",
+                month_zero, "_", month_Abb, "/", file_name)
+
+  list(file_name,
+       url)
+}
+
+dates[, c("file_name", "url") := make_url(years, months)]
+p <- txtProgressBar(max = 100, style = 3)
[…]

### 2020-09-01 analysis/paper/paper.Rmd
mas draft

@@
 
-theme_set(theme_asymsam(base_size = 10))
+theme_set(theme_asymsam(base_size = 9))
 ZeroBreaks <- AnchorBreaks(0, NULL, 0)
@@
 
+@fogt2012 studied the characteristics of the asymmetric structure of the SAM.
+It computed the zonally anomalous component of mean sea level pressure (MSLP) composites for positive and negative SAM events and created two indices by projecting MSLP fields ontno them.
+However, the use of composites leads to some issues that affects the interpretability of the results.
+First, they can be dependent on the choice of threshold used to define positive and negative events.
+Secondly, by discarting data that don't meet the threshold, they don't use all the information avaiable.
+Due to the realtive short timeframe used, this leads to some composites being composed of as little as 4 years.
+Third, the resulting composites corresponding to each polarity and season are derived from the average of different amount of fields and from different years.
+This last issue is particularly important in light of the changing structure of the SAM before and after 1980 [@silvestri2009].
+In @fogt2012, the DJF SAM+ composite uses only 7 years, 5 of which are later than 1988, whereas all of the 8 years used for their DJF SAM- composite are from earlier than 1988.
+
 ```{r read-hgt}
@@
 # Methods
-
-### Data
-
-We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach] for the period 1979 to 2018.
-
-Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0 degree latitude x 5.0 degree longitude global grid [@vose2012; @smith2008].
-The same analysis was carried out using CRUTEM4 [@osborn2014] (not shown).
-
-We used monthly precipitation data from CPC Merged Analysis of Precipitation [@xie1997] 2.5 degree latitude x 2.5 degree longitude. CPCC: [schneider2015]  #FIXME
 
@@
 ```
+
+### Data
+
+We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach2020.] for the period 1979 to 2018.
+
+Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0 degree latitude x 5.0 degree longitude global grid [@vose2012; @smith2008].
+The same analysis was carried out using CRUTEM4 [@osborn2014] (not shown).
+
+We used monthly precipitation data from CPC Merged Analysis of Precipitation [@xie1997] 2.5 degree latitude x 2.5 degree longitude. CPCC: [schneider2015]  #FIXME
 
@@
 
-These results point to very different behaviours between the stratospheric and tropospheric SAM. #FIXME
-
-This suggests that both the Asymmetric and the Symemtric component of the tropospheric SAM are highly vertically coherent, both in their individual evolution and their temporal relationship.
-This is to be expected since the SAM is mostly equivalent barotropic (citaaaa).
+Figure \ref{fig:r-squared-trend} shows decadal trends for the explained variance of each index.
+There is no evidencie of a significant trend in the stratosphere.
+In the troposphere, there is a positive trend for the Asymmetric SAM and no significant trend for the Symmetric SAM.
+This suggest that the SAM has become more asymmetric in the period from 1979 to 2018.
+The change is slight, though; of the order of 1% icreased explained variance per decade.
+
+```{r r-squared-trend, fig.height=5, fig.width=3, fig.cap = "Decadal trends for explained variance of each index at each level for the period 1979-2018.
+Shading indicates the 95\\% confidence interval."}
+indices %>% 
+  .[term != "full"] %>% 
+  .[, FitLm(r.squared, time, se = TRUE), by = .(lev, type = term)] %>% 
+  rm_intercept() %>%
+  .[, estimate := estimate*3600*24*365*10] %>% 
+  .[, std.error := std.error*3600*24*365*10] %>% 
+  ggplot(aes(lev, estimate)) +
+  geom_hline(yintercept = 0) +
+  geom_ribbon(aes(fill = type, ymin = estimate - 1.96*std.error, ymax = estimate + 1.96*std.error), alpha = 0.2) +
+  geom_line(aes(color = type)) +
+  scale_x_level() +
+  scale_y_continuous("Trend in R^2") +
+  scale_fill_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill"), labels = lab_sam) +
+  coord_flip() 
+```
 
@@
 Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b.
-The pattern cross-correlation between levels of each segment of the atmosphere is greater than 0.9 (Figure A#fixme).
+The pattern cross-correlation between levels of each segment of the atmosphere is greater than 0.9 (Figure A8).
 The patterns mainly change in amplitude.
@@
     by = .(lon, lat, season(time))] %>% 
-  rm_intercept() 
+  rm_intercept() %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] 
 ```
@@
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), na.fill = 0,
+  geom_contour_fill(aes(z = estimate), 
                     breaks = AnchorBreaks(0, 0.15, 0)) +
-  geom_contour_tanaka(aes(z = estimate), na.fill = 0,
+  geom_contour_tanaka(aes(z = estimate),
                       breaks = AnchorBreaks(0, 0.15, 0)) +
@@
   geom_contour_fill(aes(z = p.value), breaks = c(0, 0.05), color = "black", fill = "gray50", alpha = 0.2) +
-  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), fill = "gray", color = "gray30") +
+  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), fill = "gray") +
+  geom_contour2(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), color = "gray50") +
   geom_qmap(~.x[lat <= 0]) +
@@
   scale_y_latitude(labels = NULL, limits = c(NA, 0)) +
-  scale_fill_divergent("Surface temperature",  limits = c(-0.8, .8), oob = scales::squish,
+  scale_fill_divergent("Surface temperature",  
+                       limits = c(-0.8, .8), 
+                       oob = scales::squish,
                        guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.15, 0)) +
@@
 
-```{r}
+```{r calc-indices2d-seasonal}
 indices_2d_seasonal <- hgt[lev %in% 700] %>% 
@@
                 aes(z = estimate, linetype = factor(-sign(..level..)))) +
-    geom_qmap(~.x %>% add_continent() %>% .[continent == this_continent], keep = 0.015, weighting = 0.9) +
+    geom_qmap(~.x[lat <= 10], keep = 0.015, weighting = 0.9) +
     scale_x_longitude(breaks = NULL) +
@@
 
-```{r pp-regr-oceania, dependson=plot_pp,  fig.cap = "Same but for oceania", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
+```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Same but for oceania", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
 plot_pp("oceania") 
@@
 
-```{r pp-regr-america, dependson = plot_pp, fig.cap = "Same but for america", fig.width = 6, fig.height = 7.6, fig.env = "figure*"}
+```{r pp-regr-america, dependson = "def-plot_pp", fig.cap = "Same but for america", fig.width = 6, fig.height = 7.6, fig.env = "figure*"}
 plot_pp("america")
@@
 
-In South America (Figure \ref{fig:pp-regr-america}) , the annual-level regression shows that the SAM is associated with statistically significant precipitation decreases both in Southeastern South America and Southern Chile (panel a.1).
-Both signals were observed by @silvestri2009 and the former also by @gillett2006.
-Panels b.1 and c.1 show a remarkable clean separation between the Asymmeric SAM --associated with the signal located in Southestern South America-- and the Symmetric SAM --associated with the signal in Chile.
+In South America (Figure \ref{fig:pp-regr-america}), the annual-level regression shows that the SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1).
+
+Panels b.1 and c.1 show a remarkably clean separation between the Asymmeric SAM --associated with the Southeastern South American and Southern Brazilian signals-- and the Symmetric SAM --associated with the signal in Southern Chile.
 This separation is consistent with the mechanisms responsible for these effects.
+In Southern Chile, the reduced westerly flow reduce moisture transport from the Pacific Ocean (cita??
+#FIXME).
+
 In Southeastern South America, anomalous meridional winds lead to less precipitation by inhibiting moisture convergence from the South American Low Level Jet [@silvestri2009].
-In Southern Chile, the reduced westerly flow reduce moisture transport from the Pacific Ocean (cita??).
-The is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 and that is explained by the Asymmmetric SAM.
-
-The negative relationship between precipiation and SAM index in Southeastern South America and Southern Chile is evident in all seasons except Winter (although not always stastitically significant) and it appears to maximise in Autumn.
+The increased precipiation in the South Atlantic Convergence Zone, on the other hand, appears to be related to modulation of the anomalous SAM circulation in SACZ events which lead to more frequent and intense SACZ events during positive SAM [@rosso2018].
+
+There is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 and that is explained by the Asymmmetric SAM.
+
+Except during Winter, the seasonal-level regression all show these similar patterns although not as cleanly and only in some cases statistically significant.
 
@@
   ggplot(aes(lon, lat)) +
-  scattermore::geom_scattermore(aes(color = estimate)) +
+  scattermore::geom_scattermore(aes(color = estimate), pixels = c(256, 256)) +
   stat_subset(aes(subset = pval < 0.01 & is.cross(x, y, 3)),
@@
 
-@silvestri2009 showed that impacts linked to the SAM changed rather dramatically before and after 1980, and that this change can be explained by changes in the regional-scale structure of the SAM-related circulation.
-In particular, the decreased in precipitation in South America (Figure \ref{fig:pp-regr-america} panel a.1) was absent or even switched sign.
-
-@silvestri2009 also thowed
-
-Therefore, is very likely that if we were to repeat this analysis using pre-satellite data, the resulting Asymmetric SAM would look very differnt.
-
-This suggests that both the Asymmetric and the Symemtric component of the tropospheric SAM are highly vertically coherent, both in their individual evolution and their temporal relationship.
-This is to be expected since the SAM is mostly equivalent barotropic (citaaaa).
-
-@hosking2013 observed a pattern similar to the one near the Amundsen Sea in panel a.4 associated with the strengthening of the ASL.
+We presented a method to systematically separate the zonally asymmetric and zonally symmetric components of the SAM. 
+By itself, the spatial structure and temporal evolution of the SAM shows a strong separation between the stratosphere SAMs.
+The first EOF in the stratosphere is monopolar in nature while the first EOF in the troposphere is a proper annular mode.
+Their repespective departures from the zonal mean are also different: a zonal wave 1 dominates the stratospheric SAM, waves 3 and 2 dominate the tropospheric SAM.
+Furthermore, there is little temporal correlation between the tropospheric and stratospheric time series.
+
+The zonal asymmetric component of the SAM at each level is even more decupled between the troposphere and the stratosphere.
+Their temporal evolution shows essentialy zero correlation (Figure \ref{fig:cross-correlation}) and the signal associated with the stratospheric Asymmetric SAM is completely restricted to the stratosphere (Figure \ref{fig:vertical-regression} panel a).
+Geopotential height anomalies associated with the tropospheric Asymmetric SAM, on the other hand, do extend to the stratosphere, but those anomalies do not project strongly into the stratospheric Asymmetric SAM.
+
+We show that the observed positive trends towards positive SAM is restricted to the tropospheric SAM and is explained by the Symmetric component (Figure \ref{fig:trends]).
+However, the degree of asymmetry appears to have increased slightly in the last 40 years, as the Asymmetric SAM explains an increasingly proportion of the variance (Figure \ref{fig:r-squared-trend}).
+
+In terms of impacts, we see that ..Temperature... #FIXME
+
+The patterns of SAM-associated precipitation anomalies are similarly well separated.
+In South America, we show that negative anomalies observed in Chile related to the SAM are well explained by the Symmetric compomnet, while the precipitation dipole in Southern South America and the South Atlantic Convergence Zone is explained by the Asymmetric component.
+
+<!-- @hosking2013 observed a pattern similar to the one near the Amundsen Sea in panel a.4 associated with the strengthening of the ASL.
 The positive anomalies of SIC near the Amundsen sea being inconsistent with the anomalies of meridional wind associated with the ASL.
@@
 #FIXME!!
+-->
+
+### Limitations
+
+Our method assumes linearity in the asymmetric component of the SAM.
+That is, assumes that zonal symmetries associated with positive SAM are oposite and equal to the ones associated with negatie SAM.
+@fogt2012's composites suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
+Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for for SAM+ and SAM- show pattern linear correlations greater than -0.7 for all seasons and are visually very linear (Figure A9).
+Therefore, we belive that our method is at the very least a reasonable approximation of the fenomenon.
+
+We also asumed that the structure of the SAM zonal anomalies is stable in all seasons.
+Again, this is not unreasonable, as geopotential zonal anomalies computed by projecting the first EOF *of each season* are very similar to each other (Figure A10).
+
+@silvestri2009 showed that impacts linked to the SAM changed rather dramatically before and after 1980.
+In particular, the negative relationship with precipitation in South America (consistent with Figure \ref{fig:pp-regr-america} panel a.1) was absent in some areas and switched sign in other in the earlier period.
+The correlation between ENSO and SAM is similarly non-stationary, also disapearing before 1973.
+
+Seeing as both the ENSO-SAM relationship and most of the precipitation imacts in South America are captured by the Asymmetric SAM, the results presented here are most likely period-dependent.
+Therefore, is very likely that if we were to repeat this analysis using pre-satellite data, the resulting Asymmetric SAM would look very different.
 
@@
 
-```{r A2 ccf-levels, fig.cap = "Cross-correlation functions for each index and two differnet base levels."}
+```{r A2 ccf-levels, fig.cap = "Cross-correlation functions for each index and two differnet base levels.", fig.height=5, fig.width=6}
 ccf_dt <- function(...) {
@@
 ```{r A3, fig.cap = "Fourier spectrum of each timeseries.
-The shading indicates de 95\\% area derived by fitting an AR process to each series and bootstrapping 5000 simulated samples.", appendix = TRUE}
+The shading indicates de 95\\% area derived by fitting an AR process to each series and bootstrapping 5000 simulated samples.", appendix = TRUE, fig.height=4, fig.width=7}
 spec %>%
@@
   ggplot(aes(1/freq/12, spec)) +
-  geom_ribbon(aes(fill = term, ymin = `2%`, ymax = `98%`), position = "dodge", alpha = 0.1) +
-  geom_line(aes(color = term, y = ar_spectrum), alpha = 0.15) +
-  geom_line(aes(color = term)) +
+  geom_ribbon(aes(ymin = `2%`, ymax = `98%`), position = "dodge", alpha = 0.1) +
+  geom_line(aes(y = ar_spectrum), alpha = 0.15) +
+  geom_line() +
   annotation_logticks(sides = "b") +
@@
 
-```{r A4, fig.cap = "Autocorrelation functions of each timeseries", appendix = TRUE}
+```{r A4, fig.cap = "Autocorrelation functions of each timeseries", appendix = TRUE, fig.width=4, fig.height=4}
 acf_dt <- function(...) {
@@
 
-```{r A6, appendix = TRUE, fig.cap = "Regressions maps resulting from performing one multiple regression (column a) and from performing two simple regressions (column b)"}
-regr_dif <- hgt[lev %in% 700] %>% 
-  indices_wide[., on = .NATURAL, allow.cartesian = TRUE] %>% 
-  .[,  rbind(idcol = "regression", multiple = as.data.table(FitLm(hgt_a, sym, asym)),
-             single = rbind(as.data.table(FitLm(hgt_a, sym)),
-                            as.data.table(FitLm(hgt_a, asym)))),
-    by = .(lon, lat)]
-
-regr_dif %>% 
-  rm_intercept() %>% 
-  periodic(lon = c(0, 360)) %>% 
-  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>% 
-  .[lat <= -20] %>% 
-  ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), breaks = ZeroBreaks, global.breaks = FALSE) +
-  # geom_contour2(aes(z = estimate, linetype = factor(-sign(..level..))), 
-  # size = 0.3,
-  # global.breaks = FALSE) +
-  geom_contour_tanaka(aes(z = estimate),
-                      breaks = ZeroBreaks,
-                      size = 0.3,
-                      global.breaks = FALSE) +
-  geom_qmap(~.x[lat <= -20]) +
-  # geom_contour2(data = ~.x[dataset == "NCEP"], aes(z = V1)) +
-  scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
-                       limits = c(-85, 85),
-                       oob = scales::squish,
-                       breaks = MakeBreaks(binwidth = 20, exclude = 0)) +
-  scale_linetype(guide = "none") +
-  coord_polar() +
-  facet_grid(term ~ regression, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets("cr") +
-  axis_labs_smol 
-```
-
-```{r A7, appendix = TRUE, fig.cap = "Zonal waves derives from the regression maps from performing one multiple regression (column a) and from performing two simple regressions (column b)"}
-regr_dif %>% 
-  rm_intercept() %>% 
-  .[lat < -20] %>%
-  # .[lev %in% c(50, 700)] %>% 
-  .[, FitWave(estimate, 0:3), by = .(lat, regression, term)] %>% 
-  ggplot(aes(lat, amplitude)) +
-  geom_line(aes(color = factor(k))) +
-  scale_color_brewer("Wave-number", palette = "Dark2") +
-  scale_y_continuous("Amplitude") +
-  scale_x_latitude(ticks = 15) +
-  coord_flip() +
-  facet_grid(term~regression, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free_x") +
-  tag_facets("cr", position = "tr")
-```
-
-```{r A8, appendix = TRUE, fig.cap = "Pattern cross-correlation \\#FIXME!"}
+```{r A8, appendix = TRUE, fig.cap = "Pattern cross-correlation \\#FIXME!", fig.height=4, fig.width=4}
 vertical_regr %>% 
@@
 ```
+
+```{r A9, fig.cap = "700 hPa Geopotetnial height zonal anomalies of composites of positive and negative SAM months selected using 1 standard deviation as threshhold.
+Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.", apendix = TRUE, fig.height = 5, fig.width = 7}
+composites <- hgt[lev == 700] %>%
+  indices_wide[., on = .NATURAL] %>% 
+  .[, sign := ifelse(abs(full) > sd(full), sign(full), 0), by = season(time)] %>% 
+  .[, .(hgt = mean(hgt_a), n = .N), by = .(lon, lat, sign, season(time))] %>% 
+  .[, z := Anomaly(hgt), by = .(lat, sign, season)]
+
+composites_cors <- composites %>% 
+  dcast(season + lat + lon ~ sign, value.var = "z") %>% 
+  .[, .(cor = cor(`-1`, `1`)), by = season]
+
+tags <- composites[order(sign, season)][sign != 0, paste0("n: ", unique(n)), by = .(sign, season)]
+
+composites %>% 
+  .[sign != 0] %>% 
+  composites_cors[., on = "season"] %>% 
+  .[order(season)] %>% 
+  .[, season_text := paste0(season, "\ncor: ", signif(cor, 2))] %>% 
+  .[, season_text := factor(season_text, levels = unique(season_text), ordered = TRUE)] %>% 
+  periodic(lon = c(0, 360)) %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = z), breaks = AnchorBreaks(0,exclude = 0), global.breaks = FALSE) +
+  geom_contour_tanaka(aes(z = z), breaks = AnchorBreaks(0,exclude = 0), global.breaks = FALSE) +
+
+  geom_qmap(~.x[lat <= 0]) +
+  scale_x_longitude() +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
+                       breaks = AnchorBreaks(0, binwidth = 15, exclude = 0)) +
+  coord_polar() +
+  axis_labs_smol +
+  no_grid +
+  facet_grid(sign~season_text, labeller = labeller(sign = c("-1" = "SAM-", "1" = "SAM+"))) +
+  tag_facets(tag_pool = tags$V1, tag_suffix = "")
+
+```
+
+```{r A10, fig.cap = "Zonal of projection of 700 hPa onto the first EOF of each season.",  apendix = TRUE, fig.width=4, fig.height=4}
+set.seed(42)
+SAM_season <- hgt[lat <= -20] %>% 
+  .[lev == 700] %>% 
+  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
+  .[, season := season(time)] %>% 
+  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
+    by = .(season)]
+
+SAM_season[, eof[[1]]$left, by = season] %>% 
+  copy() %>% 
+  setnames("hgt", "EOF") %>% 
+  .[, EOF := EOF/sd(EOF), by = .(season)] %>% 
+  hgt[., on = "time"] %>% 
+  .[, .(proj = sum(EOF*hgt_a)), by = .(lon, lat, season)] %>% 
+  .[season == "DJF", proj := -proj] %>% 
+  .[, proj_z := Anomaly(proj), by = .(lat, season)] %>% 
+   periodic(lon = c(0, 360)) %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = proj_z), breaks = AnchorBreaks(0,exclude = 0), global.breaks = FALSE) +
+  geom_contour_tanaka(aes(z = proj_z), breaks = AnchorBreaks(0,exclude = 0), global.breaks = FALSE) +
+  geom_qmap(~.x[lat <= 0]) +
+  scale_x_longitude() +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
+  scale_fill_divergent(NULL, guide = "none") +
+  coord_polar() +
+  axis_labs_smol +
+  no_grid +
+  facet_wrap(.~season)
+```

### 2020-09-01 analysis/paper/paper.tex
mas draft

@@
 
+\citet{fogt2012} studied the characteristics of the asymmetric structure
+of the SAM. It computed the zonally anomalous component of mean sea
+level pressure (MSLP) composites for positive and negative SAM events
+and created two indices by projecting MSLP fields ontno them. However,
+the use of composites leads to some issues that affects the
+interpretability of the results. First, they can be dependent on the
+choice of threshold used to define positive and negative events.
+Secondly, by discarting data that don't meet the threshold, they don't
+use all the information avaiable. Due to the realtive short timeframe
+used, this leads to some composites being composed of as little as 4
+years. Third, the resulting composites corresponding to each polarity
+and season are derived from the average of different amount of fields
+and from different years. This last issue is particularly important in
+light of the changing structure of the SAM before and after 1980
+\citep{silvestri2009}. In \citet{fogt2012}, the DJF SAM+ composite uses
+only 7 years, 5 of which are later than 1988, whereas all of the 8 years
+used for their DJF SAM- composite are from earlier than 1988.
+
 \section{Methods}
-
-\subsubsection{Data}
-
-We used monthly geopotential height at 2.5 longitude by 2.5 latitude
-resolution from ERA5 \citep{hersbach} for the period 1979 to 2018.
-
-Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0
-degree latitude x 5.0 degree longitude global grid
-\citep{vose2012, smith2008}. The same analysis was carried out using
-CRUTEM4 \citep{osborn2014} (not shown).
-
-We used monthly precipitation data from CPC Merged Analysis of
-Precipitation \citep{xie1997} 2.5 degree latitude x 2.5 degree
-longitude. CPCC: {[}schneider2015{]} \#FIXME
 
@@
 \end{figure*}
+
+\subsubsection{Data}
+
+We used monthly geopotential height at 2.5 longitude by 2.5 latitude
+resolution from ERA5 \citep{hersbach2020} for the period 1979 to 2018.
+
+Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0
+degree latitude x 5.0 degree longitude global grid
+\citep{vose2012, smith2008}. The same analysis was carried out using
+CRUTEM4 \citep{osborn2014} (not shown).
+
+We used monthly precipitation data from CPC Merged Analysis of
+Precipitation \citep{xie1997} 2.5 degree latitude x 2.5 degree
+longitude. CPCC: {[}schneider2015{]} \#FIXME
 
@@
 
-These results point to very different behaviours between the
-stratospheric and tropospheric SAM. \#FIXME
+Figure \ref{fig:r-squared-trend} shows decadal trends for the explained
+variance of each index. There is no evidencie of a significant trend in
+the stratosphere. In the troposphere, there is a positive trend for the
+Asymmetric SAM and no significant trend for the Symmetric SAM. This
+suggest that the SAM has become more asymmetric in the period from 1979
+to 2018. The change is slight, though; of the order of 1\% icreased
+explained variance per decade.
 
-This suggests that both the Asymmetric and the Symemtric component of
-the tropospheric SAM are highly vertically coherent, both in their
-individual evolution and their temporal relationship. This is to be
-expected since the SAM is mostly equivalent barotropic (citaaaa).
+\begin{figure}
+\includegraphics{r-squared-trend-1} \caption[Decadal trends for explained variance of each index at each level for the period 1979-2018]{Decadal trends for explained variance of each index at each level for the period 1979-2018.
+Shading indicates the 95\% confidence interval.}\label{fig:r-squared-trend}
+\end{figure}
 
@@
 between levels of each segment of the atmosphere is greater than 0.9
-(Figure A\#fixme). The patterns mainly change in amplitude. The
-tropospheric pattern is maximised by the 300 hPa Asymmetric SAM index
-and the stratospheric pattern increased monotonically with height.
+(Figure A8). The patterns mainly change in amplitude. The tropospheric
+pattern is maximised by the 300 hPa Asymmetric SAM index and the
+stratospheric pattern increased monotonically with height.
 
@@
 
-In South America (Figure \ref{fig:pp-regr-america}) , the annual-level
+In South America (Figure \ref{fig:pp-regr-america}), the annual-level
 regression shows that the SAM is associated with statistically
-significant precipitation decreases both in Southeastern South America
-and Southern Chile (panel a.1). Both signals were observed by
-\citet{silvestri2009} and the former also by \citet{gillett2006}. Panels
-b.1 and c.1 show a remarkable clean separation between the Asymmeric SAM
---associated with the signal located in Southestern South America-- and
-the Symmetric SAM --associated with the signal in Chile. This separation
-is consistent with the mechanisms responsible for these effects. In
-Southeastern South America, anomalous meridional winds lead to less
+significant precipitation decrease in Southeastern South America (SESA)
+and Southern Chile and non-significant increase in South Brazil, near
+the South Atlantic Convergence Zone (SACZ) (panel a.1).
+
+Panels b.1 and c.1 show a remarkably clean separation between the
+Asymmeric SAM --associated with the Southeastern South American and
+Southern Brazilian signals-- and the Symmetric SAM --associated with the
+signal in Southern Chile. This separation is consistent with the
+mechanisms responsible for these effects. In Southern Chile, the reduced
+westerly flow reduce moisture transport from the Pacific Ocean (cita??
+\#FIXME).
+
+In Southeastern South America, anomalous meridional winds lead to less
 precipitation by inhibiting moisture convergence from the South American
-Low Level Jet \citep{silvestri2009}. In Southern Chile, the reduced
-westerly flow reduce moisture transport from the Pacific Ocean (cita??).
-The is a small area of increased precipitation with SAM near central
+Low Level Jet \citep{silvestri2009}. The increased precipiation in the
+South Atlantic Convergence Zone, on the other hand, appears to be
+related to modulation of the anomalous SAM circulation in SACZ events
+which lead to more frequent and intense SACZ events during positive SAM
+\citep{rosso2018}.
+
+There is a small area of increased precipitation with SAM near central
 Argentina which is also present in the station-based analysis by
@@
 
-The negative relationship between precipiation and SAM index in
-Southeastern South America and Southern Chile is evident in all seasons
-except Winter (although not always stastitically significant) and it
-appears to maximise in Autumn.
+Except during Winter, the seasonal-level regression all show these
+similar patterns although not as cleanly and only in some cases
+statistically significant.
 
@@
 
+We presented a method to systematically separate the zonally asymmetric
+and zonally symmetric components of the SAM. By itself, the spatial
+structure and temporal evolution of the SAM shows a strong separation
+between the stratosphere SAMs. The first EOF in the stratosphere is
+monopolar in nature while the first EOF in the troposphere is a proper
+annular mode. Their repespective departures from the zonal mean are also
+different: a zonal wave 1 dominates the stratospheric SAM, waves 3 and 2
+dominate the tropospheric SAM. Furthermore, there is little temporal
+correlation between the tropospheric and stratospheric time series.
+
+The zonal asymmetric component of the SAM at each level is even more
+decupled between the troposphere and the stratosphere. Their temporal
+evolution shows essentialy zero correlation (Figure
+\ref{fig:cross-correlation}) and the signal associated with the
+stratospheric Asymmetric SAM is completely restricted to the
+stratosphere (Figure \ref{fig:vertical-regression} panel a).
+Geopotential height anomalies associated with the tropospheric
+Asymmetric SAM, on the other hand, do extend to the stratosphere, but
+those anomalies do not project strongly into the stratospheric
+Asymmetric SAM.
+
+We show that the observed positive trends towards positive SAM is
+restricted to the tropospheric SAM and is explained by the Symmetric
+component (Figure \textbackslash{}ref\{fig:trends{]}). However, the
+degree of asymmetry appears to have increased slightly in the last 40
+years, as the Asymmetric SAM explains an increasingly proportion of the
+variance (Figure \ref{fig:r-squared-trend}).
+
+In terms of impacts, we see that ..Temperature\ldots{} \#FIXME
+
+The patterns of SAM-associated precipitation anomalies are similarly
+well separated. In South America, we show that negative anomalies
+observed in Chile related to the SAM are well explained by the Symmetric
+compomnet, while the precipitation dipole in Southern South America and
+the South Atlantic Convergence Zone is explained by the Asymmetric
+component.
+
+\subsubsection{Limitations}
+
+Our method assumes linearity in the asymmetric component of the SAM.
+That is, assumes that zonal symmetries associated with positive SAM are
+oposite and equal to the ones associated with negatie SAM.
+\citet{fogt2012}'s composites suggest that this might not be entirely
+valid, although we argue that much of that apparent non-linearity is due
+to the heterogenous nature of the selected years for constructing the
+composites. Using our data (from 1979 to 2018), seasonal composites of
+zonal anomalies of 700 hPa geopotential height for for SAM+ and SAM-
+show pattern linear correlations greater than -0.7 for all seasons and
+are visually very linear (Figure A9). Therefore, we belive that our
+method is at the very least a reasonable approximation of the fenomenon.
+
+We also asumed that the structure of the SAM zonal anomalies is stable
+in all seasons. Again, this is not unreasonable, as geopotential zonal
+anomalies computed by projecting the first EOF \emph{of each season} are
+very similar to each other (Figure A10).
+
 \citet{silvestri2009} showed that impacts linked to the SAM changed
-rather dramatically before and after 1980, and that this change can be
-explained by changes in the regional-scale structure of the SAM-related
-circulation. In particular, the decreased in precipitation in South
-America (Figure \ref{fig:pp-regr-america} panel a.1) was absent or even
-switched sign.
+rather dramatically before and after 1980. In particular, the negative
+relationship with precipitation in South America (consistent with Figure
+\ref{fig:pp-regr-america} panel a.1) was absent in some areas and
+switched sign in other in the earlier period. The correlation between
+ENSO and SAM is similarly non-stationary, also disapearing before 1973.
 
-\citet{silvestri2009} also thowed
-
-Therefore, is very likely that if we were to repeat this analysis using
-pre-satellite data, the resulting Asymmetric SAM would look very
-differnt.
-
-This suggests that both the Asymmetric and the Symemtric component of
-the tropospheric SAM are highly vertically coherent, both in their
-individual evolution and their temporal relationship. This is to be
-expected since the SAM is mostly equivalent barotropic (citaaaa).
-
-\citet{hosking2013} observed a pattern similar to the one near the
-Amundsen Sea in panel a.4 associated with the strengthening of the ASL.
-The positive anomalies of SIC near the Amundsen sea being inconsistent
-with the anomalies of meridional wind associated with the ASL. Row 4 in
-Figure \ref{fig:regr-ice} might shed some light to resolve this mistery.
-The SAM is evidently related to the ASL through it's Asymmetric
-component, and since the ASymmetric SAM is correlated with the Symmetric
-SAM, the ALS will be rleated to the symemtric SAM. Consiering the
-similarities between the the SIC anomalies observed by
-\citet{hosking2013} (in its Figure 7) and Figure \ref{fig:regr-ice}
-panel a.4, it is very likely that what was observed in that paper was a
-residual effect of the SAM. \#FIXME!!
+Seeing as both the ENSO-SAM relationship and most of the precipitation
+imacts in South America are captured by the Asymmetric SAM, the results
+presented here are most likely period-dependent. Therefore, is very
+likely that if we were to repeat this analysis using pre-satellite data,
+the resulting Asymmetric SAM would look very different.
 
@@
 \begin{figure}
-\includegraphics{A6-1} \appendcaption{A6}{Regressions maps resulting from performing one multiple regression (column a) and from performing two simple regressions (column b)}\label{fig:A6}
+\includegraphics{A8-1} \appendcaption{A8}{Pattern cross-correlation \#FIXME!}\label{fig:A8}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A7-1} \appendcaption{A7}{Zonal waves derives from the regression maps from performing one multiple regression (column a) and from performing two simple regressions (column b)}\label{fig:A7}
+\includegraphics{A9-1} \caption[700 hPa Geopotetnial height zonal anomalies of composites of positive and negative SAM months selected using 1 standard deviation as threshhold]{700 hPa Geopotetnial height zonal anomalies of composites of positive and negative SAM months selected using 1 standard deviation as threshhold.
+Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A9}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A8-1} \appendcaption{A8}{Pattern cross-correlation \#FIXME!}\label{fig:A8}
+\includegraphics{A10-1} \caption[Zonal of projection of 700 hPa onto the first EOF of each season]{Zonal of projection of 700 hPa onto the first EOF of each season.}\label{fig:A10}
 \end{figure}

### 2020-09-04 analysis/paper/paper.Rmd
borrador

@@
  - \usepackage{subfig}
+ - \usepackage[inline]{showlabels}
 ---
@@
 
-### Significance
+### Regressions
+
+Explain multiple regression interpretation #FIXME.
 
@@
 
-Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and Symmetric time series corresponding to 700 and 50hPa.  #FIXME
-
+Figure \ref{fig:asymsam-timeseries} shows the Asymmetric and Symmetric time series corresponding to 700 and 50hPa.
 At first glance the series can be distinguished by their distributions.
@@
                     color = "white", rotate = FALSE, stroke.color = "black", stroke = 0.2, 
-                    skip = 1, size = 2) +
+                    skip = 1, size = 2, 
+                    data = ~.x[item2 > item1]) +
   geom_polygon(data = tri, fill = "white") +
@@
 Figure \ref{fig:trends} shows normalised decadal trends for each index for the whole period 1979-2918 along with the 95% confidence interval in shading for the whole year (row a) and separed by trimesters (rows b through e).
-As documented by #FIXME (e.g.
-@fogt2020), there is a statistically significant increase towards more positive SAM (panel a.1), which is XX only in Summer and Autumn (panels b.1 and c.1).
+As previously documented (e.g.
+@fogt2020), there is a statistically significant increase towards more positive SAM (panel a.1), which is evident only in Summer and Autumn (panels b.1 and c.1).
 We observe these increases mainly in the troposphere, reaching their maximum at at 100 hPa in Summer.
@@
 
-```{r r-squared-trend, fig.height=5, fig.width=3, fig.cap = "Decadal trends for explained variance of each index at each level for the period 1979-2018.
+```{r r-squared-trend, fig.height=3, fig.width=3, fig.cap = "Decadal trends for explained variance of each index at each level for the period 1979-2018.
 Shading indicates the 95\\% confidence interval."}
@@
   geom_hline(yintercept = 0) +
-  geom_ribbon(aes(fill = type, ymin = estimate - 1.96*std.error, ymax = estimate + 1.96*std.error), alpha = 0.2) +
+  geom_ribbon(aes(fill = type, color = type, 
+                  ymin = estimate - 1.96*std.error, ymax = estimate + 1.96*std.error), 
+              size = 0.2, alpha = 0.2) +
   geom_line(aes(color = type)) +
@@
 
-To undertand the spatial patterns associated with both indeces, we regressed monthly geopotential anomalies into both indeces using multiple regression
-
 ```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.", fig.env = "figure*", fig.width=7, fig.height=5}
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.
+Points marked on panel b.2 are the location of the reference points used by \\cite{raphael04} for its Zonal Wave 3 index.", fig.env = "figure*", fig.width=7, fig.height=5}
 indices_2d_yearly %>% 
@@
   geom_qmap(~.x[lat <= -20]) +
+  geom_point(data = data.frame(lon = ConvertLongitude(c(50, 166, -76)),
+                               lat = -49, term = "asym", lev = 700), 
+             shape = 21, color = "black", fill = "#de3e80") +
   scale_x_longitude() +
@@
   coord_polar() +
-  facet_grid(lev ~ term) + # labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets("rc") +
+  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  tag_facets("cr") +
   axis_labs_smol  +
@@
 
-Figure \ref{fig:2d-regr} shows the spatial year-long regression #FIXME.
-Column a are regressinos using the Full SAM, while columns 2 and 3 are regression coefficients computed in a multiple regression of geopotential height on the Asymmetric and Symmetric indices at the same time.
-Thus, they are to be interpreted as the patterns associated with each index, controling for the (linear) effect of the other.
-(Figure A6 #FIXME illustrates the difference between computing two simple regressions and one multiple regression.)
+To show if, and to what extent, the Asymmetric and Symmetric SAM inidices indeed capture the asymmetric and symmetric compoment of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index.
+Figure \ref{fig:2d-regr} shows these regressions.
+Regression coefficients in column a are computed using the Full SAM.
+Regression coefficients in columns b and c are computed using multiple regression using the Asymmetric and Symmetric indices at the same time.
+Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other index.
 
 In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the south pole.
-The monopole obtained by multiple regression with the Asymmetric and Symmetric SAM (panel a.3) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel a.2).
-
-In the troposphere, panel b.1 shows the well known zonally symmetrical annular mode *contaminated* with zonal asymmetries in the form of a wave-3.
+The monopole obtained by multiple regression with the Asymmetric and Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel b.1).
+
+In the troposphere, panel a.2 shows the well known zonally symmetrical annular mode *contaminated* with zonal asymmetries in the form of a wave-3.
 The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
-The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel b.3).
-note that the wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern and asociated with @raphael2004's ZW3 index (see Figure 1 from that paper).
-#FIXME (agregar algo más?)
-
-```{r calculate-raphael}
+The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel c.2).
+The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern asociated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure.
+Thus, the tropospheric Asymmetric SAM index responds to zonal displacements in the position of the wave-3 pattern.
+
+```{r compute-raphael}
 # 50°E, 166°E and 76°W. 
@@
   .[, hgt := hgt/9.8] %>% 
-  .[lon %~% ConvertLongitude(c(-50, -166, 76))] %>% 
+  .[, hgt := Anomaly(hgt), by = .(lat, time)] %>% 
+  .[lon %~% ConvertLongitude(c(50, 166, -76))] %>% 
   .[, hgt_s := frollmean(hgt, 3, NA, align = "center"), by = .(lon, lat)] %>% 
@@
   # geom_contour2(aes(z = p.val), breaks = 0.01) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(18), 
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(13), 
                        oob = scales::squish,
@@
 
-The geopotential anomalies associated with the stratospheric SAM (panel a) are clearly constrained to the stratosphere, which underscores the disconnect between the stratospheric and tropospheric symetric SAM.
+The geopotential anomalies associated with the stratospheric SAM (panel a) are clearly constrained to the stratosphere, which underscores the disconnect between the stratospheric and tropospheric Asymmetric SAM.
 The vertical structure this signal tilts about 60ª to the West between 100 hPa and 1 hPa, suggesting baroclinic processes and polarward transport of heat (#FIXME is this ok?).
@@
 
-The tropospheric asymmetric SAM has significant signals that extend upwards the uppermost levels of the reanalysis.
+The tropospheric Asymmetric SAM (panel b) has significant signals that extend upwards to the uppermost levels of the reanalysis.
 In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
-#FIXME moaaaaar.
-
-Interestingly, the structures shown in Figure \ref{fig:vertical-regression} are surprisignly robust to the choice of index level.
+The anomalies are much more intense in the Western hemisphere, where they extent into the stratosphere.
+In the Eastern hemisphere the wave-3 signal is weaker and confined to the troosphere while negative anomalies dominate in the stratosphere.
+So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
+
+The structures shown in panels a and b in  Figure \ref{fig:vertical-regression} are surprisignly robust to the choice of index level.
 For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in panel a.
 Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b.
-The pattern cross-correlation between levels of each segment of the atmosphere is greater than 0.9 (Figure A8).
+Pattern cross-correlation is greater than 0.9 within stratospheric index levels and tropospheric index levels and less than .4 accross stratospheric and tropospheric levels (Figure A8), i.e.
+regressions obtained using Asymmetric SAM indices in the troposphere are very similar to each other and different to the ones obtaines using indices in the stratosphere and vice versa.
 The patterns mainly change in amplitude.
@@
 The correlation between the full SAM and the ENSO as measured by the Multivariate Enso Index [@wolter2011] is `r cor_texts$full`.
-This relationship is captured entirely the Asymmetric SAM, as this index has a partial correlation of `r mei_sam_partial$asym` with the MEI, whereas the Symmetric SAM has null partial correlation with the MEI.
+This relationship is captured entirely the Asymmetric SAM, as this index has a partial correlation of `r mei_sam_partial$asym` with the MEI, whereas the Symmetric SAM's partial correlation with the MEI is essentially null (`r mei_sam_partial$sym`).
 
@@
 
-In DJF (column a), the strong negative signal in the tropical Pacific in panel a.1 is mostly associated with the Asymmetric component (panel a.2), as is it largely absent in the Symmetric component (panel a.3).
+In DJF (row 1), the strong negative signal in the tropical Pacific in panel a.1 is mostly associated with the Asymmetric component (panel b.1), as is it largely absent in the Symmetric component (panel c.1).
 Furthermore, the Asymmetric SAM is also asociated with low temperature anomalies in the Indian ocean, but this signal is obscured by the Symmetric variability and thus lost in the Full SAM.
 Over the continents, the Asymmetric SAM is assoiated with negative temperature anomalies which, again, mostly disappear in the Full SAM regression.
-
-The patterns seen in MAM and JJA (columns b and c) are not robustly significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following @wilks2016.
+Temperature regression with the Symmetric SAM (panel c.1) shows a ring of negative anomalies around Antarctica surrounded by positive anomalies.
+This pattern of anomalies is consistent with thermal wind balance and the intesification and pole-ward migration of the westerlies commonly linked to the SAM.
+
+The patterns seen in MAM and JJA (rows 2 and 3) are not robustly significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following @wilks2016.
 Nevertheless, it is interesting to note that in both trimesters, the sign of the regression is consistently flipped between the Asymmetric and Symmetric regressions.
@@
 
-Finally, in SON (column d), there is no significant temperature signal associated with the Symmetric SAM (panel d.3), while the Asymmetric SAM shows a relatively robust signal in the equatorial Pacitic, Australia, and even Southeast South America.
-This strong signals are reduced in intensity in panel a.3.
+Finally, in SON (row 4) there is no significant temperature signal associated with the Symmetric SAM (panel c.4), while the Asymmetric SAM shows a relatively robust signal in the equatorial Pacitic, Australia, and even Southeast South America.
 
@@
 
-```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Same but for oceania", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
+```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
 plot_pp("oceania")
+
 ```
@@
 plot_pp("america")
-
 ```
@@
 
-We presented a method to systematically separate the zonally asymmetric and zonally symmetric components of the SAM. 
-By itself, the spatial structure and temporal evolution of the SAM shows a strong separation between the stratosphere SAMs.
+We presented a method to systematically separate the zonally asymmetric and zonally symmetric components of the SAM into two indices.
+As expected, they are significantly correlated with each other, but they also have their own unique dynamics such as differnet vertical coherence patterns, variability and long term trends.
+
+The spatial structure and temporal evolution of the SAM show a strong separation between the stratosphere SAMs.
 The first EOF in the stratosphere is monopolar in nature while the first EOF in the troposphere is a proper annular mode.
-Their repespective departures from the zonal mean are also different: a zonal wave 1 dominates the stratospheric SAM, waves 3 and 2 dominate the tropospheric SAM.
+Their repespective departures from the zonal mean are also different: a zonal wave 1 dominates the stratospheric SAM; waves 3 and 2 dominate the tropospheric SAM.
 Furthermore, there is little temporal correlation between the tropospheric and stratospheric time series.
@@
 
-We show that the observed positive trends towards positive SAM is restricted to the tropospheric SAM and is explained by the Symmetric component (Figure \ref{fig:trends]).
+We show that the observed positive trend towards positive SAM is restricted to the tropospheric SAM and is explained by the Symmetric component (Figure \ref{fig:trends]).
 However, the degree of asymmetry appears to have increased slightly in the last 40 years, as the Asymmetric SAM explains an increasingly proportion of the variance (Figure \ref{fig:r-squared-trend}).
 
-In terms of impacts, we see that ..Temperature... #FIXME
+Temperature and precipitation anomalies associated with the SAM respond to various processes in different locations.
+The Asymmetric and Symmetric indices do a reasonable job separating some of them.
+The strong sumertime temperature anomalies in the equatorial Pacific are explained by the Asymmetric component, most likely due to its relationship with ENSO.
+The Symmetric SAM, on the other hand, captures the change in meridional temperature gradient, which is linked to the SAM-zonal wind relationship by thermal wind balance.
 
 The patterns of SAM-associated precipitation anomalies are similarly well separated.
-In South America, we show that negative anomalies observed in Chile related to the SAM are well explained by the Symmetric compomnet, while the precipitation dipole in Southern South America and the South Atlantic Convergence Zone is explained by the Asymmetric component.
+In South America, we show that negative anomalies observed in Chile related to the SAM are well explained by the Symmetric component, while the precipitation dipole in Southern South America and the South Atlantic Convergence Zone is explained by the Asymmetric component.
 
@@
 
-### Limitations
+### Limitations {-}
 
@@
   .[, id := interaction(lev, lon)] %>% 
-  widyr::pairwise_cor(index_level, id, estimate) %>% 
+  widyr::pairwise_cor(index_level, id, estimate, diag = TRUE) %>% 
+  as.data.table() %>% 
   ggplot(aes(item2, item1)) +
@@
                     color = "white", rotate = FALSE, stroke.color = "black", stroke = 0.2, 
-                    skip = 1, size = 4) +
+                    skip = 1, size = 4,
+                    data = ~.x[item2 > item1]) +
   geom_polygon(data = tri, fill = "white") +
-  scale_fill_divergent("Cross-correlation", 
+  scale_fill_divergent("Pattern correlation", 
                        breaks = MakeBreaks(binwidth = 0.1),

### 2020-09-04 analysis/paper/paper.tex
borrador

@@
 \usepackage{subfig}
+\usepackage[inline]{showlabels}
 
@@
 
-\subsubsection{Significance}
+\subsubsection{Regressions}
+
+Explain multiple regression interpretation \#FIXME.
 
@@
 
-Figure \ref{fig:asymsam-timeseries} shows the resulting Asymmetric and
-Symmetric time series corresponding to 700 and 50hPa. \#FIXME
-
-At first glance the series can be distinguished by their distributions.
-Whereas the tropospheric incides are aproximately normally distributed,
-the stratospheric indices are more long-tailed; that is, extreme values
-(both negative and positive) abound. The Asymmetric series have both
-more variability in the higher frequencies than the Symmetric series.
+Figure \ref{fig:asymsam-timeseries} shows the Asymmetric and Symmetric
+time series corresponding to 700 and 50hPa. At first glance the series
+can be distinguished by their distributions. Whereas the tropospheric
+incides are aproximately normally distributed, the stratospheric indices
+are more long-tailed; that is, extreme values (both negative and
+positive) abound. The Asymmetric series have both more variability in
+the higher frequencies than the Symmetric series.
 
@@
 in shading for the whole year (row a) and separed by trimesters (rows b
-through e). As documented by \#FIXME (e.g. \citet{fogt2020}), there is a
+through e). As previously documented (e.g. \citet{fogt2020}), there is a
 statistically significant increase towards more positive SAM (panel
-a.1), which is XX only in Summer and Autumn (panels b.1 and c.1). We
-observe these increases mainly in the troposphere, reaching their
+a.1), which is evident only in Summer and Autumn (panels b.1 and c.1).
+We observe these increases mainly in the troposphere, reaching their
 maximum at at 100 hPa in Summer. By separating the SAM signal in its
@@
 
-To undertand the spatial patterns associated with both indeces, we
-regressed monthly geopotential anomalies into both indeces using
-multiple regression
-
 \begin{figure*}
 \includegraphics{2d-regr-1} \caption[Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM]{Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.}\label{fig:2d-regr}
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.
+Points marked on panel b.2 are the location of the reference points used by \cite{raphael04} for its Zonal Wave 3 index.}\label{fig:2d-regr}
 \end{figure*}
 
-Figure \ref{fig:2d-regr} shows the spatial year-long regression \#FIXME.
-Column a are regressinos using the Full SAM, while columns 2 and 3 are
-regression coefficients computed in a multiple regression of
-geopotential height on the Asymmetric and Symmetric indices at the same
-time. Thus, they are to be interpreted as the patterns associated with
-each index, controling for the (linear) effect of the other. (Figure A6
-\#FIXME illustrates the difference between computing two simple
-regressions and one multiple regression.)
+To show if, and to what extent, the Asymmetric and Symmetric SAM
+inidices indeed capture the asymmetric and symmetric compoment of the
+SAM respectively, we computed the spatial regression of geopotential
+height anomalies on these indices and the Full SAM index. Figure
+\ref{fig:2d-regr} shows these regressions. Regression coefficients in
+column a are computed using the Full SAM. Regression coefficients in
+columns b and c are computed using multiple regression using the
+Asymmetric and Symmetric indices at the same time. Thus, they are to be
+interpreted as the patterns associated with each index, removing the
+variability (linearly) explained by the other index.
 
@@
 The monopole obtained by multiple regression with the Asymmetric and
-Symmetric SAM (panel a.3) is much more symmetric and the shift from
+Symmetric SAM (panel c.1) is much more symmetric and the shift from
 total symmetry is captured by the regression pattern of the Asymmetric
@@
 Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel
-a.2).
+b.1).
 
-In the troposphere, panel b.1 shows the well known zonally symmetrical
+In the troposphere, panel a.2 shows the well known zonally symmetrical
 annular mode \emph{contaminated} with zonal asymmetries in the form of a
@@
 assosiated with an trully annular mode, almost devoid of zonal
-asymmetries (panel b.3). note that the wave-3 pattern observed in panel
-b.2 is rotated by half a wavelength from the average position of the
-mean wave-3 pattern and asociated with \citet{raphael2004}'s ZW3 index
-(see Figure 1 from that paper). \#FIXME (agregar algo más?)
+asymmetries (panel c.2). The wave-3 pattern observed in panel b.2 is
+rotated by half a wavelength from the average position of the mean
+wave-3 pattern asociated with \citet{raphael2004}'s ZW3 index, whose
+reference locations are marked with points in the figure. Thus, the
+tropospheric Asymmetric SAM index responds to zonal displacements in the
+position of the wave-3 pattern.
 
@@
 a) are clearly constrained to the stratosphere, which underscores the
-disconnect between the stratospheric and tropospheric symetric SAM. The
-vertical structure this signal tilts about 60ª to the West between 100
-hPa and 1 hPa, suggesting baroclinic processes and polarward transport
-of heat (\#FIXME is this ok?). Interestingly, the signal in the
-stratosphere maximises near 10 hPa despite using the 50 hPa index for
-the regression.
+disconnect between the stratospheric and tropospheric Asymmetric SAM.
+The vertical structure this signal tilts about 60ª to the West between
+100 hPa and 1 hPa, suggesting baroclinic processes and polarward
+transport of heat (\#FIXME is this ok?). Interestingly, the signal in
+the stratosphere maximises near 10 hPa despite using the 50 hPa index
+for the regression.
 
-The tropospheric asymmetric SAM has significant signals that extend
-upwards the uppermost levels of the reanalysis. In the troposphere, the
-wave-3 structure is equivalent barotropic with maximum amplitude at
-roughly 250 hPa. \#FIXME moaaaaar.
+The tropospheric Asymmetric SAM (panel b) has significant signals that
+extend upwards to the uppermost levels of the reanalysis. In the
+troposphere, the wave-3 structure is equivalent barotropic with maximum
+amplitude at roughly 250 hPa. The anomalies are much more intense in the
+Western hemisphere, where they extent into the stratosphere. In the
+Eastern hemisphere the wave-3 signal is weaker and confined to the
+troosphere while negative anomalies dominate in the stratosphere. So,
+while the tropospheric Asymmetric SAM index is associated with
+stratospheric geopotential anomalies, these do not project strongly onto
+the stratospheric Asymmetric SAM.
 
-Interestingly, the structures shown in Figure
+The structures shown in panels a and b in Figure
 \ref{fig:vertical-regression} are surprisignly robust to the choice of
@@
 hPa in panel a. Conversely, for any tropospheric (below 100 hPa) index,
-the result is very similar to panel b. The pattern cross-correlation
-between levels of each segment of the atmosphere is greater than 0.9
-(Figure A8). The patterns mainly change in amplitude. The tropospheric
-pattern is maximised by the 300 hPa Asymmetric SAM index and the
-stratospheric pattern increased monotonically with height.
+the result is very similar to panel b. Pattern cross-correlation is
+greater than 0.9 within stratospheric index levels and tropospheric
+index levels and less than .4 accross stratospheric and tropospheric
+levels (Figure A8), i.e.~regressions obtained using Asymmetric SAM
+indices in the troposphere are very similar to each other and different
+to the ones obtaines using indices in the stratosphere and vice versa.
+The patterns mainly change in amplitude. The tropospheric pattern is
+maximised by the 300 hPa Asymmetric SAM index and the stratospheric
+pattern increased monotonically with height.
 
@@
 has a partial correlation of -0.27 with the MEI, whereas the Symmetric
-SAM has null partial correlation with the MEI.
+SAM's partial correlation with the MEI is essentially null
+(\ensuremath{-7.9\times 10^{-4}}).
 
@@
 
-In DJF (column a), the strong negative signal in the tropical Pacific in
+In DJF (row 1), the strong negative signal in the tropical Pacific in
 panel a.1 is mostly associated with the Asymmetric component (panel
-a.2), as is it largely absent in the Symmetric component (panel a.3).
+b.1), as is it largely absent in the Symmetric component (panel c.1).
 Furthermore, the Asymmetric SAM is also asociated with low temperature
@@
 anomalies which, again, mostly disappear in the Full SAM regression.
+Temperature regression with the Symmetric SAM (panel c.1) shows a ring
+of negative anomalies around Antarctica surrounded by positive
+anomalies. This pattern of anomalies is consistent with thermal wind
+balance and the intesification and pole-ward migration of the westerlies
+commonly linked to the SAM.
 
-The patterns seen in MAM and JJA (columns b and c) are not robustly
+The patterns seen in MAM and JJA (rows 2 and 3) are not robustly
 significant in the sense that there are no areas with p-values below
@@
 
-Finally, in SON (column d), there is no significant temperature signal
-associated with the Symmetric SAM (panel d.3), while the Asymmetric SAM
+Finally, in SON (row 4) there is no significant temperature signal
+associated with the Symmetric SAM (panel c.4), while the Asymmetric SAM
 shows a relatively robust signal in the equatorial Pacitic, Australia,
-and even Southeast South America. This strong signals are reduced in
-intensity in panel a.3.
+and even Southeast South America.
 
@@
 \begin{figure*}
-\includegraphics{pp-regr-oceania-1} \caption[Same but for oceania]{Same but for oceania}\label{fig:pp-regr-oceania}
+\includegraphics{pp-regr-oceania-1} \caption[Regression pattern of precipiation with Asymmetric and Symmetric SAM]{Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:pp-regr-oceania}
 \end{figure*}
@@
 We presented a method to systematically separate the zonally asymmetric
-and zonally symmetric components of the SAM. By itself, the spatial
-structure and temporal evolution of the SAM shows a strong separation
-between the stratosphere SAMs. The first EOF in the stratosphere is
-monopolar in nature while the first EOF in the troposphere is a proper
-annular mode. Their repespective departures from the zonal mean are also
-different: a zonal wave 1 dominates the stratospheric SAM, waves 3 and 2
-dominate the tropospheric SAM. Furthermore, there is little temporal
-correlation between the tropospheric and stratospheric time series.
+and zonally symmetric components of the SAM into two indices. As
+expected, they are significantly correlated with each other, but they
+also have their own unique dynamics such as differnet vertical coherence
+patterns, variability and long term trends.
+
+The spatial structure and temporal evolution of the SAM show a strong
+separation between the stratosphere SAMs. The first EOF in the
+stratosphere is monopolar in nature while the first EOF in the
+troposphere is a proper annular mode. Their repespective departures from
+the zonal mean are also different: a zonal wave 1 dominates the
+stratospheric SAM; waves 3 and 2 dominate the tropospheric SAM.
+Furthermore, there is little temporal correlation between the
+tropospheric and stratospheric time series.
 
@@
 
-We show that the observed positive trends towards positive SAM is
+We show that the observed positive trend towards positive SAM is
 restricted to the tropospheric SAM and is explained by the Symmetric
@@
 
-In terms of impacts, we see that ..Temperature\ldots{} \#FIXME
+Temperature and precipitation anomalies associated with the SAM respond
+to various processes in different locations. The Asymmetric and
+Symmetric indices do a reasonable job separating some of them. The
+strong sumertime temperature anomalies in the equatorial Pacific are
+explained by the Asymmetric component, most likely due to its
+relationship with ENSO. The Symmetric SAM, on the other hand, captures
+the change in meridional temperature gradient, which is linked to the
+SAM-zonal wind relationship by thermal wind balance.
 
@@
 observed in Chile related to the SAM are well explained by the Symmetric
-compomnet, while the precipitation dipole in Southern South America and
+component, while the precipitation dipole in Southern South America and
 the South Atlantic Convergence Zone is explained by the Asymmetric
@@
 
-\subsubsection{Limitations}
+\subsubsection*{Limitations}
+\addcontentsline{toc}{subsubsection}{Limitations}
 

### 2020-09-11 analysis/paper/paper.Rmd
fraft

@@
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.
-Points marked on panel b.2 are the location of the reference points used by \\cite{raphael04} for its Zonal Wave 3 index.", fig.env = "figure*", fig.width=7, fig.height=5}
+Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for its Zonal Wave 3 index.", fig.env = "figure*", fig.width=7, fig.height=5}
 indices_2d_yearly %>% 
@@
                                lat = -49, term = "asym", lev = 700), 
-             shape = 21, color = "black", fill = "#de3e80") +
+             shape = 21, size = 2.5, color = "black", fill = "#de3e80") +
   scale_x_longitude() +

### 2020-09-11 analysis/paper/paper.Rmd
Correcciones de Leandro

--- 2020-09-04 analysis/paper/paper.Rmd
@@
 
-To separate between the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern.
-The results are shown in Figure\ \ref{fig:method} for 700hPa.
+To separate between the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure\ \ref{fig:method} for 700hPa.
 The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
-We then compute the "Full", "Asymmetric" and "Symmetric" indices, by regressing each geopotential field on these patterns (weighting by the cosine of latitude).
-
+We then compute the "Full", "Asymmetric" and "Symmetric" indices, by regressing each geopotential field on the respective patterns (weighting by the cosine of latitude).
 The three indices are normalised by dividing them by the standard deviation of the "Full" index at each level.
-This means that comparing the magnitude between indices is meaningful, but it also means that not every index will have unit standard deviation.
+As a result, the magnitude between indices is comparable.
+However, only "Full" index will have unit standard deviation per definition.
 
@@
 We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach2020.] for the period 1979 to 2018.
-
 Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0 degree latitude x 5.0 degree longitude global grid [@vose2012; @smith2008].
 The same analysis was carried out using CRUTEM4 [@osborn2014] (not shown).
-
-We used monthly precipitation data from CPC Merged Analysis of Precipitation [@xie1997] 2.5 degree latitude x 2.5 degree longitude. CPCC: [schneider2015]  #FIXME
+We used monthly precipitation data from CPC Merged Analysis of Precipitation [@xie1997] 2.5 degree latitude x 2.5 degree longitude.
+CPCC: [schneider2015]  #FIXME
 
@@
 
-Figure \ref{fig:asymsam-timeseries} shows the Asymmetric and Symmetric time series corresponding to 700 and 50hPa.
+The temporal evolution of the Assymmetric and Symmetric SAM was firstly asssesed.
+Figure \ref{fig:asymsam-timeseries} shows the corresponding time series for 700 and 50hPa and their corresponding density estimates.
+Month-to-month variability is evident for both indices, with noisy variations in the low frequency.
 At first glance the series can be distinguished by their distributions.
-Whereas the tropospheric incides are aproximately normally distributed, the stratospheric indices are more long-tailed; that is, extreme values (both negative and positive) abound.
+Whereas the tropospheric indices are aproximately normally distributed, the stratospheric indices are more long-tailed; that is, extreme values (both negative and positive) abound.
 The Asymmetric series have both more variability in the higher frequencies than the Symmetric series.
@@
 Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987 marked with a circle).
-We show these correlations, across all the levels of the reanalysis and for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}.
+We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}.
+Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctiating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
+One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`.
+In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis.
+At the same time, one-month-lag correlations increase with height.
+As a consequence, statospheric Assymmetric index tend to precede corresponding Symmetric index.
 
@@
 ```
-
-Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctiating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
-One-month-lag correlations are similarly constant but significantly reduced, hovering around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`.
-In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis.
-At the same time, one-month-lag correlations increase with height.
 
@@
 Above 100 hPa correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
-Still there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
+Therefore, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
 Examining panels b and c, we see that the Asymemtric and Symmetric SAM  share the same high level of coherency in the troposphere but they differ in their stratospheric behaviour.
-As evidenced by the wider dark red areas near the diagonal in Figure \ref{fig:cross-correlation}b) vs.
-Figure \ref{fig:cross-correlation}c), stratospheric coherency is stronger for the Asymmetric SAM than the Symemtric SAM.
-The stratospheric Symmetric SAM seems to connect more strongly to the trosposphere than the Asymmetric SAM; this can be seen by the lower correlation values in the top right left of Figure \ref{fig:cross-correlation}b) in comparison with Figure \ref{fig:cross-correlation}c).
+Stratospheric coherency is stronger for the Asymmetric SAM than the Symemtric SAM.
+The stratospheric Symmetric SAM seems to connect more strongly to the trosposphere than the Asymmetric SAM.
 
@@
 
-Figure \ref{fig:trends} shows normalised decadal trends for each index for the whole period 1979-2918 along with the 95% confidence interval in shading for the whole year (row a) and separed by trimesters (rows b through e).
+The trends for each of the indices (Full, Symmetric, Assymetric) were evaluate for the whole period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separed by trimesters.
 As previously documented (e.g.
@@
 
-In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the south pole.
+In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the South Pole.
 The monopole obtained by multiple regression with the Asymmetric and Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel b.1).
@@
 The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern asociated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure.
-Thus, the tropospheric Asymmetric SAM index responds to zonal displacements in the position of the wave-3 pattern.
+Thus, the tropospheric Asymmetric SAM index responds to zonal displacements in the position of the climatological wave-3 pattern, previously described in the literature (e.g.
+....).
 
@@
 
-The amplitude of each zonal wave number at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
-Comparing between rows, this Figure quantifies the relatively clean separation between the zonally symmetric and zonally asymmetric structures, as its evident how the mixture of waves of the Full field (column a) is very similar to the sum of the waves of the Asymmetric and Symmetric field (columns b and c, respectively).
+The amplitude of first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
+This Figure quantifies the relatively clean separation between the zonally symmetric and zonally asymmetric structures, as its evident how the mixture of waves of the Full field (column a) is very similar to the sum of the waves of the Asymmetric and Symmetric field (columns b and c, respectively).
 Column b of Figure \ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b.1), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel b.2).
@@
 To analyse the vertical structure of the geopotential anomalies asociated with the asymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65ªS and 40ªS for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
-
-The geopotential anomalies associated with the stratospheric SAM (panel a) are clearly constrained to the stratosphere, which underscores the disconnect between the stratospheric and tropospheric Asymmetric SAM.
-The vertical structure this signal tilts about 60ª to the West between 100 hPa and 1 hPa, suggesting baroclinic processes and polarward transport of heat (#FIXME is this ok?).
+The geopotential anomalies associated with the stratospheric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
+The vertical structure of this signal tilts about 60ª to the West between 100 hPa and 1 hPa, suggesting baroclinic processes and poleward transport of heat (#FIXME is this ok?).
 Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
@@
 Furthermore, the Asymmetric SAM is also asociated with low temperature anomalies in the Indian ocean, but this signal is obscured by the Symmetric variability and thus lost in the Full SAM.
-Over the continents, the Asymmetric SAM is assoiated with negative temperature anomalies which, again, mostly disappear in the Full SAM regression.
+Over the continents, the Asymmetric SAM is associated with negative temperature anomalies that disappear in the Full SAM regression.
 Temperature regression with the Symmetric SAM (panel c.1) shows a ring of negative anomalies around Antarctica surrounded by positive anomalies.
-This pattern of anomalies is consistent with thermal wind balance and the intesification and pole-ward migration of the westerlies commonly linked to the SAM.
+This pattern of anomalies is consistent with thermal wind balance and the intesification and poleward migration of the westerlies commonly linked to the SAM.
 
@@
 
-Regression of the SAM indicies with seasonal mean precipitation are shown Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australia and New Zealand, and South America respectively.
-(We didn't detect any significant signal in South Africa.)
+Regression of the SAM indicies with seasonal mean precipitation are shown Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
+South Africa is not shown because any significant signal in South Africa was detected.
 
@@
 
-In Australia (Figure \ref{fig:pp-regr-oceania}), the annual-level regression shows that the Full SAM is associated with a statistically significant increase in precipitation in the Southeastern region (panel a.1), which reproduces the results from @gillett2006.
+In Australia, the annual regression shows that the Full SAM is associated with a statistically significant increase in precipitation in the Southeastern region (Figure \ref{fig:pp-regr-oceania} panel a.1), which reproduces the results from @gillett2006.
 The separation between Asymmetric and Symmetric SAM suggest that this increase is explained by the Symmetric SAM only in the East coast (panel c.1), which is consistent with the increased easterly flow clearly seen in relation with this index.
@@
 
-We presented a method to systematically separate the zonally asymmetric and zonally symmetric components of the SAM into two indices.
-As expected, they are significantly correlated with each other, but they also have their own unique dynamics such as differnet vertical coherence patterns, variability and long term trends.
+In this study, we presented a novel method to separate the zonally asymmetric and zonally symmetric contributions of the SAM into two indices.
+As expected, these are significantly correlated between them, but each of them has its own unique dynamics such as differnet vertical coherence patterns, variability and long term trends.
 
 The spatial structure and temporal evolution of the SAM show a strong separation between the stratosphere SAMs.
-The first EOF in the stratosphere is monopolar in nature while the first EOF in the troposphere is a proper annular mode.
+The first EOF in the stratosphere has a monopolar structure while the first EOF in the troposphere is a proper annular mode.
 Their repespective departures from the zonal mean are also different: a zonal wave 1 dominates the stratospheric SAM; waves 3 and 2 dominate the tropospheric SAM.
@@
 @fogt2012's composites suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
-Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for for SAM+ and SAM- show pattern linear correlations greater than -0.7 for all seasons and are visually very linear (Figure A9).
+Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for for SAM+ and SAM- show spatial pattern correlations greater than -0.7 for all seasons and are visually very linear (Figure A9).
 Therefore, we belive that our method is at the very least a reasonable approximation of the fenomenon.

### 2020-09-23 analysis/paper/paper.Rmd
llegamos a la conclusion!

@@
 )
+
+knitr::opts_hooks$set(fig.cap = function(options) {
+  options$cache <- FALSE
+  options
+})
+
 library(metR)
@@
 library(tagger)
+library(patchwork)
+library(knitr)
 
@@
 
+### Data
+
+To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
+We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.
+
+We describe the relationship between the SAM indices and temperature and precipitation.
+We use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree lagitude by 5\degree longitude.
+For precipitation, we use monthly, 0.5\degree latitude by 0.5\degree longitude data from the Global Precipitation Climatology Centre [@schneider2015].
+
 ### Definition of indices
 
-We defined the Southern Annular Mode (SAM) as the leading EOF of the monthly anomalies of geopotential field at 700 hPa south of 20\degree S (citation?).
-The EOF was performed by computing the Singular Value Decomposition of the data matrix consisting in `r dims$t` rows and  `r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points of latitude).
-The values where weighted by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999].
-This same method was used at the rest of the levels considered in this paper.
+Traditionally the Souther Annular Mode (SAM) is defined as de leading Empiral Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels [@ho2012].
+Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the the leading EOF of the monthly anomalies of geopotential field south of 20\degree S at each level.
+We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in `r dims$t` rows and  `r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points of latitude).
+We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999].
 
@@
 
-### Data
-
-We used monthly geopotential height at 2.5 longitude by 2.5 latitude resolution from ERA5 [@hersbach2020.] for the period 1979 to 2018.
-Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0 degree latitude x 5.0 degree longitude global grid [@vose2012; @smith2008].
-The same analysis was carried out using CRUTEM4 [@osborn2014] (not shown).
-We used monthly precipitation data from CPC Merged Analysis of Precipitation [@xie1997] 2.5 degree latitude x 2.5 degree longitude.
-CPCC: [schneider2015]  #FIXME
+Our method assumes linearity in the asymmetric component of the SAM.
+That is, we assume that zonal symmetries associated with positive SAM are oposite and equal to the ones associated with negatie SAM.
+@fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
+Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (smaller than negative 1 standard deviation) show relatively high pattern correlations all seasons and are visually very linear (Figure A9).
+Therefore, we belive that our method is at the very least a reasonable approximation of the phenomenon.
+
+By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons.
+Geopotential zonal anomalies computed by projecting the first EOF *of each season* are very similar to each other (Figure A10) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (between MAM and SON).
+Based on this, we believe that our initial assumption is not unreasonable.
+
+Finally, we assume that the zonally asymmetric pattern is stationary in time.
+@silvestri2009 suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979-2018) so it's unlikely that we could observe significant changes.
+Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change (Figure A11).
 
@@
 
-Explain multiple regression interpretation #FIXME.
-
-We adjusted p-values for False Detection Rate following @wilks2016.
+We perform linear regression to quantify the association between the SAM indices and other variables.
+Since the Asymmetric and Symmetric SAM indices are significantly correlated with each other, to capture the variability explained uniquely by each index we use one multiple linear regression instead of two simple linear regressions.
+To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Asymmetric SAM ($SAM_a$) and Symmetric SAM ($SAM_s$) we fit the ecuation
+
+$$
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) SAM_a + \beta(\lambda, \phi) SAM_s + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
+$$
+
+where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear coefficients, $X_0$ and $\epsilon$ are the constant and error terms.
+From this equarion, $\alpha$ represents the (linear) association of $X$ with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; in other words, it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and viceversa for $\beta$.
+
+At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions.
+In such case, using naive p-values to test for significance leads to misleading results [@walker1914;@katz1991].
+While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjunting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue.
+Therefore, p-values showed in regression fields are all adjusted following @benjamini1995.
+
+When performing a separate regression for each trimester (DJF, MAM, JJA, SON) we first average the relevant variables to obtain a single value for each year and each trimester.
 
@@
 ```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM and density estimates.", fig.width=6, fig.height=3, fig.env="figure*"}
+
+indices %>% 
+  .[lev %in% c(50, 700) & term != "full"] %>% 
+  ggplot(aes(estimate_norm)) +
+  geom_density(aes(color = term)) +
+  facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
+  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam, guide = "none") +
+  scale_y_continuous(NULL, trans = scales::reverse_trans(), labels = NULL, breaks = NULL) +
+  scale_x_continuous(NULL, labels = NULL, breaks = NULL) +
+  coord_flip()  +
+  theme(strip.text = element_blank()) +
+
   indices %>% 
@@
   geom_line(aes(color = term)) +
-  ggridges::geom_vridgeline(stat="ydensity", scale = -4e8, fill = NA,
-                            aes(color = term, width = ..density.., x = min(time - 1e7))) +
+  # ggridges::geom_vridgeline(stat="ydensity", scale = -4e8, fill = NA,
+  #                           aes(color = term, width = ..density.., x = min(time - 1e7))) +
   # geom_smooth(aes(color = term), method = "loess", n = 481, span = 10*12/481, method.args = list(degree = 1)) +
@@
   #                         time == as.Date("1987-08-01") & term == "asym"]) +
-  ggforce::geom_mark_circle(data = ~.x[lev == 50][time == as.Date("1987-11-01") & term == "sym" | 
-                                                    time == as.Date("1987-08-01") & term == "asym"],
-                            expand = unit(2, "mm")) +
   scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam) +
@@
   facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
-  tag_facets()
+  tag_facets() +
+  plot_layout(widths = c(1, 4))
 ```
@@
 The temporal evolution of the Assymmetric and Symmetric SAM was firstly asssesed.
-Figure \ref{fig:asymsam-timeseries} shows the corresponding time series for 700 and 50hPa and their corresponding density estimates.
+Figure \ref{fig:asymsam-timeseries} shows the corresponding time series for 700 hPa and 50 hPa and their corresponding density estimates.
+We selected these two levels as representative of the tropospheric and stratospheric variability respectively.
+As will be shown later, both indices are highly coherent within each atmospheric layer, therefore is reasonable to take one level as representative of each layer.
+
 Month-to-month variability is evident for both indices, with noisy variations in the low frequency.
 At first glance the series can be distinguished by their distributions.
-Whereas the tropospheric indices are aproximately normally distributed, the stratospheric indices are more long-tailed; that is, extreme values (both negative and positive) abound.
+Compared to the stratospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound.
 The Asymmetric series have both more variability in the higher frequencies than the Symmetric series.
 
-The stratospheric Symmetric SAM varies strongly witha two-year period, which can be seen using spectral methods (Figure A3) or in the autocorrelation structure (Figure A4).
+The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A3).
+This might suggests a link between stratospheric SAM variability and QBO.
 There is a local peak at 2 years in the periodigram of the tropospheric Symmetric SAM also, although it's not statistically significant.
@@
 From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric and Symmetric time series appear to be correlated.
-Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987 marked with a circle).
+Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987).
 We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}.
@@
 One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`.
-In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis.
+In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care).
 At the same time, one-month-lag correlations increase with height.
-As a consequence, statospheric Assymmetric index tend to precede corresponding Symmetric index.
+As a consequence, statospheric Asymmetric index tend to precede corresponding Symmetric index.
 
@@
 
-Figure \ref{fig:cross-correlation}a) shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
+Figure \ref{fig:cross-correlation}a shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
 For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposfere.
@@
   copy() %>% 
-  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
+  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
   .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
@@
   .[, season := season(time)] %>% 
-  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
-  .[, FitLm(estimate_norm, time, se = TRUE), by = .(type = term, lev, season)] %>% 
+  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
+  .[, FitLm(estimate_norm, time = time, se = TRUE), by = .(type = term, lev, season)] %>% 
   rm_intercept() %>% 
@@
   ggplot(aes(lev, estimate)) +
-  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) +
+  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), 
+              fill = "#95a3ab", color = "black", 
+              alpha = .6, size = 0.1) +
   geom_line() +
@@
   facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
-  tag_facets("rc") 
+  tag_facets("cr") 
 ```
@@
 The trends for each of the indices (Full, Symmetric, Assymetric) were evaluate for the whole period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separed by trimesters.
-As previously documented (e.g.
-@fogt2020), there is a statistically significant increase towards more positive SAM (panel a.1), which is evident only in Summer and Autumn (panels b.1 and c.1).
-We observe these increases mainly in the troposphere, reaching their maximum at at 100 hPa in Summer.
-By separating the SAM signal in its Asymmetric and Symmetric parts, we can not only see that these trends are almost entirely due to the Symmetric component (columns 2 vs.
-columns 3), but in some cases the trends become more clear.
-In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the signal; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel b.1 and b.3).
+The Full SAM index presents a statistically significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throught the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
+The seasonal trends (rest of column a) indicate that positive trends are present in  Autumn and particularly in in Summer, where the 100 hPa maximum is much more defined.
+Positive trends have been documented by previous studies [e.g.
+@fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.
+
+By separating the SAM signal in its Asymmetric and Symmetric parts, we can not only see that these trends are almost entirely due to the Symmetric component (colum b vs.
+column c), but in some cases the trends become more clear.
+In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full index; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
 In Autumn, using the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full index.
@@
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.
-Points marked on panel b.2 are the location of the reference points used by \\cite{raphael04} for its Zonal Wave 3 index.", fig.env = "figure*", fig.width=7, fig.height=5}
+Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index.", fig.env = "figure*", fig.width=7, fig.height=5}
 indices_2d_yearly %>% 
@@
   .[lev %in% c(50, 700)] %>%
+  .[, env := metR::WaveEnvelope(estimate), by = .(lat, term, lev)] %>%
   ggplot(aes(lon, lat)) +
@@
                       global.breaks = FALSE) +
+  # geom_contour2(aes(z = env)) +
   geom_qmap(~.x[lat <= -20]) +
@@
 
-In the troposphere, panel a.2 shows the well known zonally symmetrical annular mode *contaminated* with zonal asymmetries in the form of a wave-3.
+In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3.
 The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
@@
 The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern asociated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure.
-Thus, the tropospheric Asymmetric SAM index responds to zonal displacements in the position of the climatological wave-3 pattern, previously described in the literature (e.g.
-....).
+Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
 
@@
 The amplitude of first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
-This Figure quantifies the relatively clean separation between the zonally symmetric and zonally asymmetric structures, as its evident how the mixture of waves of the Full field (column a) is very similar to the sum of the waves of the Asymmetric and Symmetric field (columns b and c, respectively).
-Column b of Figure \ref{fig:wave-amplitude} shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b.1), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel b.2).
+Column `r get_col(term == "asym")` shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel `r get_col(term== "asym" & lev == 50)`), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel `r get_col(term== "asym" & lev == 700)`).
+Looking at panel `r get_tag(term == "asym" & lev == 700)` from Figure \ref{fig:2d-regr}, it becomes apparent that zonal wavess 1 and 2 modulate the amplitude of zonal wave 3, which --as mentioned before-- is larger in the Western Hemisphere than in the Easten Hemisphere.
 
@@
 
-To analyse the vertical structure of the geopotential anomalies asociated with the asymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65ªS and 40ªS for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
-The geopotential anomalies associated with the stratospheric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
-The vertical structure of this signal tilts about 60ª to the West between 100 hPa and 1 hPa, suggesting baroclinic processes and poleward transport of heat (#FIXME is this ok?).
+To analyse the vertical structure of the geopotential anomalies asociated with the asymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
+The geopotential anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
+The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes.
 Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
@@
 Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b.
-Pattern cross-correlation is greater than 0.9 within stratospheric index levels and tropospheric index levels and less than .4 accross stratospheric and tropospheric levels (Figure A8), i.e.
-regressions obtained using Asymmetric SAM indices in the troposphere are very similar to each other and different to the ones obtaines using indices in the stratosphere and vice versa.
 The patterns mainly change in amplitude.
-The tropospheric pattern is maximised by the 300 hPa Asymmetric SAM index and the stratospheric pattern increased monotonically with height.
 
 ```{r enso-cor}
-mei_sam_cor <- rsoi::download_mei(TRUE, data_path("derived", "mei.csv")) %>% 
+oni <- rsoi::download_oni(TRUE, data_path("derived", "oni.csv")) %>% 
   as.data.table() %>% 
-  .[, .(time = lubridate::as_datetime(Date), mei = MEI)] %>% 
-  .[indices, on = "time"] %>% 
-  .[, with(cor.test(estimate, mei), 
+  .[, .(time = lubridate::as_datetime(Date), value = ONI)]
+
+mei <- rsoi::download_mei(TRUE, data_path("derived", "mei.csv")) %>% 
+  as.data.table() %>% 
+  .[, .(time = lubridate::as_datetime(Date), value = MEI)]
+
+soi <- rsoi::download_soi(TRUE, data_path("derived", "soi.csv")) %>% 
+  as.data.table() %>% 
+  .[, .(time = lubridate::as_datetime(Date), value = -SOI)]
+
+enso <- rbindlist(list(oni = oni, soi = soi, mei = mei), idcol = "index")
+
+enso_sam_cor <- enso %>% 
+  .[indices, on = "time", allow.cartesian = TRUE] %>% 
+  .[, with(cor.test(estimate, value), 
            list(cor = estimate, 
                 low = conf.int[1], 
-                upp = conf.int[2])), by = .(lev, term)] 
-# enso_sam_cor %>% 
-#   ggplot(aes(lev, cor)) +
-#   geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha = 0.2) +
+                upp = conf.int[2])), by = .(lev, term, index)]
+
+# ggplot(enso_sam_cor, aes(lev, cor, group = interaction(term, index))) +
+#   geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha  = .3) +
 #   geom_line(aes(color = term)) +
@@
 
-cor_texts <- mei_sam_cor[lev == 700, 
+cor_texts <- enso_sam_cor[lev == 700, 
                          paste0(signif(cor, 2)), # " (CI: ", signif(low, 2), "–", signif(upp, 2), ")"), 
-                          by = term] %>% 
-  dcast(1~term)
-
-mei_sam_partial <- rsoi::download_mei(TRUE, data_path("derived", "mei.csv")) %>% 
-  as.data.table() %>% 
-  .[, .(time = lubridate::as_datetime(Date), mei = MEI)] %>% 
-  .[indices_wide, on = "time"] %>% 
-  .[lev == 700] %>% 
-  .[, partial_cor(mei, asym, sym)] %>% 
+                         by = .(term, index)] %>% 
+  dcast(index~term)
+
+enso_sam_partial <- enso %>% 
+  .[indices_wide[lev == 700], on = "time"] %>% 
+  .[, partial_cor(value, asym, sym), by = index] %>% 
   .[, partial_correlation := signif(partial_correlation, 2)] %>% 
-  dcast(.~term)
-```
-
-The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the teleconnection pattern associated with the ENSO.
+  dcast(index~term)
+```
+
+```{r}
+#  enso %>% 
+#   .[indices, on = "time", allow.cartesian = TRUE] %>% 
+#   .[, with(cor.test(estimate, value), 
+#            list(cor = estimate, 
+#                 low = conf.int[1], 
+#                 upp = conf.int[2])), by = .(lev, term, index, season(time))] %>% 
+#   .[index == "oni"] %>% 
+# ggplot(aes(lev, cor)) +
+#   geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha  = .3) +
+#   geom_line(aes(color = term)) +
+#   scale_x_level() +
+#   coord_flip() +
+#   facet_wrap(~season)
+# 
+# enso %>% 
+#   .[indices_wide, on = "time", allow.cartesian = TRUE] %>% 
+#   .[, partial_cor(value, asym, sym), by = .(index, lev, season(time))] %>% 
+#   .[index == "oni"] %>%
+# ggplot(aes(lev, partial_correlation)) +
+#   # geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha  = .3) +
+#   geom_line(aes(color = term, group = interaction(term, index))) +
+#   scale_x_level() +
+#   coord_flip() +
+#   facet_wrap(~season)
+```
+
+The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the full SAM and the ENSO as measured by the Multivariate Enso Index [@wolter2011] is `r cor_texts$full`.
-This relationship is captured entirely the Asymmetric SAM, as this index has a partial correlation of `r mei_sam_partial$asym` with the MEI, whereas the Symmetric SAM's partial correlation with the MEI is essentially null (`r mei_sam_partial$sym`).
+The correlation between the full SAM and the ENSO as measured by the Oceanic Niño Index [@bamston1997] (ONI) is `r cor_texts[index == "oni"]$full`.
+This relationship is captured entirely the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni"]$asym` with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni"]$sym`).
+We performed the same analysis using the Multivariate Enso Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987] to conclude that these results do not depend on the ENSO index used.
 
 ## Impacts
-
-### Temperature
 
@@
 
-# mei <- rsoi::download_mei() %>% 
-#   as.data.table() %>% 
-#   .[, .(time = lubridate::as_datetime(Date), mei = MEI)]
+# air <- ReadNetCDF("~/DATOS/reanalysis/ERA5/mon/era5sl.mon.mean.nc", 
+#            vars = c(air = "t2m"), subset = c(time_subset, list(lat = c(-90, 10)))) %>% 
+#   normalise_coords()
 
@@
   # .[lat <= -61, estimate := NA] %>%
-  .[lat >= 65, estimate := NA] %>%
   .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
@@
   geom_contour_fill(aes(z = estimate), 
-                    breaks = AnchorBreaks(0, 0.15, 0)) +
-  geom_contour_tanaka(aes(z = estimate),
-                      breaks = AnchorBreaks(0, 0.15, 0)) +
+                    breaks = AnchorBreaks(0, 0.1, 0)) +
+  # geom_contour_tanaka(aes(z = estimate),
+  #                     breaks = AnchorBreaks(0, 0.1, 0)) +
   # geom_raster(aes(fill = estimate)) +
-  geom_contour_fill(aes(z = p.value), breaks = c(0, 0.05), color = "black", fill = "gray50", alpha = 0.2) +
+  # geom_contour_fill(aes(z = p.value), breaks = c(0, 0.05), fill = "gray50", alpha = 0.2) +
+  geom_contour2(aes(z = p.value), breaks = c(0, 0.05), color = "black", size = 0.3) +
   geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), fill = "gray") +
@@
   scale_fill_divergent("Surface temperature",  
-                       limits = c(-0.8, .8), 
+                       limits = c(-0.6, .6), 
                        oob = scales::squish,
-                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.15, 0)) +
+                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
   coord_polar() +
@@
 
+The SAM has been shown to be associated with important surface variables such as temperature and precipitacion [e.g.
+@gillett2006, and see @fogt2020 for a review].
+Naturally, most studies on the surface impacts of the SAM are based on an index identical or analogous to what we call Full SAM index (@fogt2012 being the only exception that we are aware of).
+We regress surface temperature and precipitation onto each of the three SAM indices to see if there are diferent surface impact associated with the asymmetric and symmetric SAM circulation.
+
 Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface temperature for each trimester.
-It is evident that the Asymmetric and Symmetric SAM indices are associated with overall distinct temperature patterns which can be obscured when using the Full SAM index.
-The Symmetric SAM signal is weaker than the Asymmetric SAM, as evidenced by the relatively smaller and les sstatistically significant regression coefficients in row 3 of Figure \ref{fig:regr-air-season} compared with row 2.
-
-In DJF (row 1), the strong negative signal in the tropical Pacific in panel a.1 is mostly associated with the Asymmetric component (panel b.1), as is it largely absent in the Symmetric component (panel c.1).
-Furthermore, the Asymmetric SAM is also asociated with low temperature anomalies in the Indian ocean, but this signal is obscured by the Symmetric variability and thus lost in the Full SAM.
-Over the continents, the Asymmetric SAM is associated with negative temperature anomalies that disappear in the Full SAM regression.
-Temperature regression with the Symmetric SAM (panel c.1) shows a ring of negative anomalies around Antarctica surrounded by positive anomalies.
-This pattern of anomalies is consistent with thermal wind balance and the intesification and poleward migration of the westerlies commonly linked to the SAM.
-
-The patterns seen in MAM and JJA (rows 2 and 3) are not robustly significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following @wilks2016.
-Nevertheless, it is interesting to note that in both trimesters, the sign of the regression is consistently flipped between the Asymmetric and Symmetric regressions.
-In South America, for example, the Asymmetric SAM is associated with positive temperature anomalies in MAM and negative temperature anomales in JJA, while the oposite is the case for the Symemtric SAM.
-
-Finally, in SON (row 4) there is no significant temperature signal associated with the Symmetric SAM (panel c.4), while the Asymmetric SAM shows a relatively robust signal in the equatorial Pacitic, Australia, and even Southeast South America.
-
-### Precipitation
-
-Regression of the SAM indicies with seasonal mean precipitation are shown Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
-South Africa is not shown because any significant signal in South Africa was detected.
+In Summer positive values of the Full SAM index (panel `r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
+The ring is not zonally symmetric, as there are three clear local maximums around 30\degree W, 15\degree E and 50\degree E and a local minimum (with negative sign) around 120\degree W.
+In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
+Panels `r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of the Asymmetric and Symmetric SAM, respectively.
+Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the Asymetric SAM regression map, while temperature patterns linked to positive Symmetric SAM show a more zonally consistent ring and less relation to the tropics.
+Noticeable, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to positive values of Asymmetric SAM.
+This signal is not present in the regression pattern with the Full SAM.
+Spring (row `r get_row(season == "SON")`) features very similar patterns but of generally smaller in magnitude and statistical significance.
+
+In Autumn and Winter (rows `r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))`) the postive ring is only present through its local maximums in the regression with the Full SAM.
+There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
+These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following @wilks2016.
+However, repeating this analysis with 2-meter temperature from ERA5 resulted in similar patterns that were statistically significant.
+Moreover, similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
+
+The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons --although with varying intensity and small-scale details-- is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM.
+It's then not surprising to see it more clearly in association with the Symemtric SAM (at least in Summer and Spring).
+
+These results suggests that Asymmetric and Symmetric SAM indices are associated with overall distinct temperature patterns which may not be apparent when using the Full SAM index.
+
+Figure \ref{fig:regr-air-season} column b can be partially compared with Figure 11 from @fogt2012.
+Although they used station data from 1958 to 2001, a lot of the characteristics are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.
+
+Regression of the SAM indicies with seasonal mean precipitation and 700 hPa geopotential height are shown Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
+South Africa is not shown because no significant signal was detected there.
 
@@
 
-pp <- CPCC() %>% 
+pp <- GPCC() %>% 
   ReadNetCDF(subset = list(lat = c(-90, 0),
@@
 plot_pp <- function(this_continent, data = pp_regr) {
+  gh <- indices_2d %>% 
+    copy() %>%
+    .[, c("u", "v") := GeostrophicWind(estimate, lon, lat, cyclical = TRUE), by = .(term, season)] %>%
+    add_continent() %>% 
+    .[continent == this_continent]
+
   data %>%
@@
     # stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat, 2)), size = 0.4, alpha = 0.8) +
-    geom_contour2(data = indices_2d %>% copy() %>% add_continent() %>% .[continent == this_continent],
+    geom_contour2(data = gh,
                   color = "gray50", size = 0.2, global.breaks = FALSE, 
@@
 plot_pp("oceania") 
-
-```
-
-In Australia, the annual regression shows that the Full SAM is associated with a statistically significant increase in precipitation in the Southeastern region (Figure \ref{fig:pp-regr-oceania} panel a.1), which reproduces the results from @gillett2006.
-The separation between Asymmetric and Symmetric SAM suggest that this increase is explained by the Symmetric SAM only in the East coast (panel c.1), which is consistent with the increased easterly flow clearly seen in relation with this index.
-The Asymmetric SAM appears related to increased precipitation in the West coast of Southeastern Australia (panel b.2), explained by the anomalous *westerly* circulation transporting moist air to the continent.
-
-The seasonal-level regressions show statistically significant anomalies only in SON, with a pattern similar to the annual-level regression (panel a.5).
-Panels b.5 and c.5 don't show a clear separation between the Asymmetric and Symmetric SAM.
-If anything, the positive and more significant regression coefficients in panel b.5 vs pane c.5 would suggest more influence of the Asymmetric than the Symmetric SAM, going against the interpretation gathered from the annual-level regressions.
+```
+
+In Australia, the annual regression shows that the Full SAM index is positively associated with precipitation in the Southeastern region (Figure \ref{fig:pp-regr-oceania} panel a.1), which reproduces the results from @gillett2006.
+The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (panel c.1).
+Geopotential anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by @hendon2007.
+The Asymmetric SAM appears related to increased precipitation in the West coast of Southeastern Australia (panel b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
+
 This Spring signal is broadly consistent with @hendon2007, but whereas @hendon2007 also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coeffcients have the consistent sign).
+
+The seasonal-level regressions show statistically significant anomalies only in Spring, when positive Full SAM is associated with positive precipitation anomalies in Eastern Australia (panel a.5).
+In this trimester the Symmetric SAM seems to be associated with precipitation in a relatively reduced area of the East Coast (panel c.5) while the positive precipitation anomalies related with positive Asymmetric SAM affect all Estern Australia (panel b.5).
+
+In Summer, positive Full SAM index is associated with with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (panel a.2).
+The Eastern part being dominated by the relationship with the Symmetric SAM and the Western, by the Asymemtric SAM.
+In Autumn, the regression with Full SAM shows positive values in the North, similar to Summer, and a broad area of positive values in the North-East to South-West direction.
+This structure seems to be associated with the Symmetric SAM, while the Northern positive values are associated with the Asymmetric SAM.
+In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the Symmetric SAM.
+None of these regression coefficients are statistically signifincat at the 95% level
 
@@
 In South America (Figure \ref{fig:pp-regr-america}), the annual-level regression shows that the SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1).
-
 Panels b.1 and c.1 show a remarkably clean separation between the Asymmeric SAM --associated with the Southeastern South American and Southern Brazilian signals-- and the Symmetric SAM --associated with the signal in Southern Chile.
-This separation is consistent with the mechanisms responsible for these effects.
-In Southern Chile, the reduced westerly flow reduce moisture transport from the Pacific Ocean (cita??
-#FIXME).
-
-In Southeastern South America, anomalous meridional winds lead to less precipitation by inhibiting moisture convergence from the South American Low Level Jet [@silvestri2009].
-The increased precipiation in the South Atlantic Convergence Zone, on the other hand, appears to be related to modulation of the anomalous SAM circulation in SACZ events which lead to more frequent and intense SACZ events during positive SAM [@rosso2018].
+
+Except Winter, seasonal-level regressions mirror this same pattern.
+Even if not statistically significant,  they all show negative values in Southeastern South America and Southern Chile along with positive values in Southern Brazil in relation with the Full SAM.
+The separation of these features between the Asymmetric SAM and Symmetric SAM regression maps is also rather consistent.
+
+The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous Easterly flow over Southern Chile.
+This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region.
+On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
+This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region [@silvestri2009, @zamboni2010].
+This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events [@rosso2018].
 
@@
 
-Except during Winter, the seasonal-level regression all show these similar patterns although not as cleanly and only in some cases statistically significant.
-
-### Sea ice
-
-```{r read-ice}
-ice <- readRDS(data_path("derived", "ice.Rds"))
-ice_proj <- readRDS(data_path("derived", "ice_proj.Rds"))
-
-new_proj <- "+proj=stere +lat_0=-90 +lat_ts=-70 +lon_0=-180 +k=1 +x_0=0 +y_0=0 +a=6378273 +b=6356889.449 +units=m +no_defs"
-
-project_ice <- function(data) {
-  browser()
-  data[, c("x", "y") := proj4::project(list(x, y), proj = ice_proj, inverse = TRUE)] %>% 
-    .[, x := ConvertLongitude(x, group = group)] %>% 
-    .[]
-
-}
-```
-
-```{r compte-ice-regression}
-ice_regr <- ice %>% 
-  .[indices_wide[lev == 700], on = .NATURAL] %>% 
-  # .[season(time) %in% c("JJA", "SON")] %>%
-  .[concentration <= 0.15, concentration := 0] %>% 
-  .[, lapply(.SD, mean, na.rm = TRUE), by = .(x, y, time = seasonally(time))] %>% 
-  # .[season(time) == "SON"] %>% 
-  .[, combine_lists(FitLm(concentration, full, se = TRUE),
-                    FitLm(concentration, asym, sym, se = TRUE)),
-    by = .(x, y, season(time))] %>%
-  rm_intercept() 
-```
-
-```{r regr-ice, fig.cap = "Seasonal regression of SAM indices with sea ice concentration. \\#FIXME", fig.env="figure*",  fig.width = 6, fig.height = 8}
-ice_regr %>% 
-  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>%
-  .[, pval := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)] %>%
-  .[, c("lon", "lat") :=  proj4::project(list(x, y), proj = ice_proj, inverse = TRUE)] %>% 
-  # .[season == "SON"] %>%
-  .[, lon := ConvertLongitude(lon)] %>% 
-  na.omit() %>% 
-  .[estimate != 0] %>%
-
-  ggplot(aes(lon, lat)) +
-  scattermore::geom_scattermore(aes(color = estimate), pixels = c(256, 256)) +
-  stat_subset(aes(subset = pval < 0.01 & is.cross(x, y, 3)),
-              geom = "point", size = 0.01, alpha = .6) +
-  geom_contour2(data = indices_2d_seasonal[lat <= -50], color = "gray50", size = 0.3,
-                aes(z = estimate, linetype = factor(-sign(..level..)))) +
-  geom_qmap(~.x[lat <= -60]) +
-  scale_color_divergent("Sea Ice",
-                        limits = c(-.1, .1), 
-                        oob = scales::squish, 
-                        guide = guide_colorstrip_bottom(),
-                        breaks = AnchorBreaks(0, 0.025, 0)) +
-  scale_x_longitude() +
-  scale_y_latitude(limit = c(-90, NA), labels = NULL) +
-  scale_linetype(guide = "none") +
-  coord_polar() +
-  facet_grid(season ~ term, labeller = labeller(term = lab_sam)) +
-  tag_facets("cr") +
-  no_grid +
-  axis_labs_smol
-```
-
-Regressions between the Full SAM index and Antarctic Sea Ice Concentrations (Figure \ref{fig:regr-ice}) show a great deal of variability across seasons.
-The only statistically significant signal is in Spring, when we observe negative concentraion anomalies in the Northen Weddell Sea (panel a.4) explained by the Asymmetric SAM (panel b.4).
-Both in Winter and in Spring the Asymmetric SAM is associated with bigger Sea Ice Concentration anomalies in West Antarctica than East Antarctica, with generally decreased concentration East of the Antarctic Peninsula and increased concentration to the West, as expected from the anomalous circulation correlated with this index.
-The Symmetric SAM signal appears more evenly distributed accross the whole ice sheet.
-
 ## Conclusions
 
-In this study, we presented a novel method to separate the zonally asymmetric and zonally symmetric contributions of the SAM into two indices.
-As expected, these are significantly correlated between them, but each of them has its own unique dynamics such as differnet vertical coherence patterns, variability and long term trends.
-
-The spatial structure and temporal evolution of the SAM show a strong separation between the stratosphere SAMs.
-The first EOF in the stratosphere has a monopolar structure while the first EOF in the troposphere is a proper annular mode.
-Their repespective departures from the zonal mean are also different: a zonal wave 1 dominates the stratospheric SAM; waves 3 and 2 dominate the tropospheric SAM.
-Furthermore, there is little temporal correlation between the tropospheric and stratospheric time series.
-
-The zonal asymmetric component of the SAM at each level is even more decupled between the troposphere and the stratosphere.
-Their temporal evolution shows essentialy zero correlation (Figure \ref{fig:cross-correlation}) and the signal associated with the stratospheric Asymmetric SAM is completely restricted to the stratosphere (Figure \ref{fig:vertical-regression} panel a).
-Geopotential height anomalies associated with the tropospheric Asymmetric SAM, on the other hand, do extend to the stratosphere, but those anomalies do not project strongly into the stratospheric Asymmetric SAM.
-
-We show that the observed positive trend towards positive SAM is restricted to the tropospheric SAM and is explained by the Symmetric component (Figure \ref{fig:trends]).
-However, the degree of asymmetry appears to have increased slightly in the last 40 years, as the Asymmetric SAM explains an increasingly proportion of the variance (Figure \ref{fig:r-squared-trend}).
-
-Temperature and precipitation anomalies associated with the SAM respond to various processes in different locations.
-The Asymmetric and Symmetric indices do a reasonable job separating some of them.
-The strong sumertime temperature anomalies in the equatorial Pacific are explained by the Asymmetric component, most likely due to its relationship with ENSO.
-The Symmetric SAM, on the other hand, captures the change in meridional temperature gradient, which is linked to the SAM-zonal wind relationship by thermal wind balance.
-
-The patterns of SAM-associated precipitation anomalies are similarly well separated.
-In South America, we show that negative anomalies observed in Chile related to the SAM are well explained by the Symmetric component, while the precipitation dipole in Southern South America and the South Atlantic Convergence Zone is explained by the Asymmetric component.
-
-<!-- @hosking2013 observed a pattern similar to the one near the Amundsen Sea in panel a.4 associated with the strengthening of the ASL.
-The positive anomalies of SIC near the Amundsen sea being inconsistent with the anomalies of meridional wind associated with the ASL.
-Row 4 in Figure \ref{fig:regr-ice} might shed some light to resolve this mistery.
-The SAM is evidently related to the ASL through it's Asymmetric component, and since the ASymmetric SAM is correlated with the Symmetric SAM, the ALS will be rleated to the symemtric SAM.
-Consiering the similarities between the the SIC anomalies observed by @hosking2013 (in its Figure 7) and Figure \ref{fig:regr-ice} panel a.4, it is very likely that what was observed in that paper was a residual effect of the SAM.
-#FIXME!!
--->
-
-### Limitations {-}
-
-Our method assumes linearity in the asymmetric component of the SAM.
-That is, assumes that zonal symmetries associated with positive SAM are oposite and equal to the ones associated with negatie SAM.
-@fogt2012's composites suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
-Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for for SAM+ and SAM- show spatial pattern correlations greater than -0.7 for all seasons and are visually very linear (Figure A9).
-Therefore, we belive that our method is at the very least a reasonable approximation of the fenomenon.
-
-We also asumed that the structure of the SAM zonal anomalies is stable in all seasons.
-Again, this is not unreasonable, as geopotential zonal anomalies computed by projecting the first EOF *of each season* are very similar to each other (Figure A10).
-
-@silvestri2009 showed that impacts linked to the SAM changed rather dramatically before and after 1980.
-In particular, the negative relationship with precipitation in South America (consistent with Figure \ref{fig:pp-regr-america} panel a.1) was absent in some areas and switched sign in other in the earlier period.
+In this study we tried to systematically characterise the variability of the zonally symmetric and zonally asymmetric structure of the SAM.
+By projecting monthly geopotential fields at each level with the corresponding asymmetric and symmetric pattern, we created two indices representing the zonally asymmetric and zonally symmetric contributions of the SAM respectively.
+
+As expected, the Asymmetric SAM index correlates strongly with the Symmetric SAM index.
+In the troposphere, this correlation is maximum at zero lag, while in the stratosphere is maximised with the Asymmetric SAM leading the Symmetric SAM by one month.
+Since most indices of the SAM are calculated using suftace or near-surface conditions, this result would suggest that they might not be sensitive to the most dramatic changes in SAM variability.
+
+The two-year periodcty we found in the stratospheric Symmetric SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
+There is evidence of influence between the QBO and the Northern Annular Mode [e.g.
+@holton1980, @watson2014, @zhang2020], so it's not unlikely that the SAM would be similarly connected.
+However establishing this link would require further research.
+
+As documented by previous studies, such as @fogt2020 (and references therein), we observe a positive trend towards positive SAM in Summer and Autumn.
+We show that these trends are maximised at the 100 hPa level and are explained by the zonally symmetric component.
+We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not apparent in the Full SAM index.
+In contrast to @fogt2012 we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
+
+In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM is much closer to being trully annular than the patterns associated with the Full SAM index.
+The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific region and whose phae is rotated a quarter wavelength from the mean zonal wave 3 described by @raphael2004's index.
+This pattern extends in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimum for capturing this variability.
+
+This wave-3 pattern is similar to the Pacific-South American Pattern, which is a teleconnection pattern linked to ENSO variability.
+We found that the significant correlation that exists between the Full SAM index and the Oceanic Niño Index is captured entirely by the Asymmetric SAM index.
+This suggests that ENSO is linked to SAM exclusively through the variability in the latter's Asymmetric compoment.
+
+Temperature anomalies associated with the Full SAM broadly show a pattern of negative anomalies at polar latitudes surrounded by positive anomalies, but with many deviations from symmetry.
+The Asymemtric SAM index explains a big portion of these deviations.
+In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship delineated above.
+These are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [@cazes-boezio2003, @fogt2011, @cai2020a].
+
+In Australia the Full SAM is associated with positive precipitation anomalies in South East and this is explained by the Symemtric SAM.
+However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
+
+In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
+This features are cleanly separated between the Asymmetric and Symmetric components.
+The Symmetric SAM explains the negative anomalies in Southern Chile and the Asymmetric SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
+Individual seasons mostly follow this pattern.
+
+@silvestri2009 suggests that precipitation impacts linked to the SAM changed rather dramatically before and after 1980.
+In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in other in the earlier period.
 The correlation between ENSO and SAM is similarly non-stationary, also disapearing before 1973.
@@
 \acknowledgments
-CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ #FIXME
-
 NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/
@@
   tag_facets() 
-```
-
-```{r A4, fig.cap = "Autocorrelation functions of each timeseries", appendix = TRUE, fig.width=4, fig.height=4}
-acf_dt <- function(...) {
-  a <- acf(..., plot = FALSE)
-  list(acf = as.vector(a$acf),
-       lag = as.vector(a$lag))
-}
-indices %>% 
-  .[lev %in% c(30, 700)] %>% 
-  .[, acf_dt(estimate_norm), by = .(lev, term)] %>% 
-  ggplot(aes(lag, acf)) +
-  geom_col(aes(fill = term)) +
-  scale_fill_brewer(palette = "Dark2") +
-  facet_grid(term ~ lev,labeller = labeller(lev = lev.lab, term = lab_sam)) +
-  tag_facets(position = "tr") 
 ```
@@
   indices_wide[., on = .NATURAL] %>% 
-  .[, sign := ifelse(abs(full) > sd(full), sign(full), 0), by = season(time)] %>% 
+  .[, sign := ifelse(abs(full) > sd(full), sign(full), 0), by = month(time)] %>% 
   .[, .(hgt = mean(hgt_a), n = .N), by = .(lon, lat, sign, season(time))] %>% 
@@
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
+                       limits = c(-74, 74),
                        breaks = AnchorBreaks(0, binwidth = 15, exclude = 0)) +
@@
     by = .(season)]
+
+SAM_season[, eof[[1]]$left, by = season] %>% 
+  widyr::pairwise_cor(season, time, hgt)
+
+cors <- SAM_season[, eof[[1]]$left, by = season] %>% 
+  copy() %>% 
+  setnames("hgt", "EOF") %>% 
+  .[, EOF := EOF/sd(EOF), by = .(season)] %>% 
+  hgt[., on = "time"] %>% 
+  .[, .(proj = sum(EOF*hgt_a)), by = .(lon, lat, season)] %>% 
+  .[season == "DJF", proj := -proj] %>% 
+  .[, proj_z := Anomaly(proj), by = .(lat, season)] %>% 
+  .[ , f := interaction(lon, lat)] %>% 
+  widyr::pairwise_cor(season, f, proj_z, upper = FALSE)
 
@@
 ```
+
+```{r A11, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height computed for the periods 1979 to 1998 and 1999 to 2018.
+Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right).
+Pattern correlation between both fields is 0.77.
+Arbitrary units.", fig.width=6, fig.height=5, apendix = TRUE}
+
+prepost_SAM <- hgt[lat <= -20] %>% 
+  .[lev == 700] %>%
+  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
+  .[, year := year(time)] %>% 
+  .[, period := factor(ifelse(year(time) <= 1998, "pre", "post"),
+                       levels = c("pre", "post"), ordered = TRUE,
+                       labels = c("1979-1998", "1999-2018"))] %>%
+  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
+    by = .(lev, period)]
+
+sam_sep <- prepost_SAM[, eof[[1]]$right, by = .(lev, period)] %>% 
+  setnames("hgt", "full") %>% 
+  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev, period)]
+
+sam_plot <- sam_sep %>% 
+  .[lev %in% c(700)] %>%
+  rm_singleton() %>% 
+  melt(id.vars = c("lon", "lat", "period")) %>% 
+  .[variable == "asym"] %>% 
+  .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)] %>% 
+  .[, value := value/sd(value), by = .(period)]
+
+cor <- sam_plot %>% 
+  dcast(lon + lat ~ period, value.var = "value") %>% 
+  .[, cor(`1979-1998`, `1999-2018`)]
+
+sam_plot %>% 
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+  # geom_contour(aes(z = value, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
+  geom_contour_fill(aes(z = value), breaks = ZeroBreaks) +
+  geom_contour_tanaka(aes(z = value), breaks = ZeroBreaks) +
+  geom_qmap() +
+  scale_x_longitude() +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
+  scale_fill_divergent(guide = "none") +
+  coord_polar() +
+  facet_wrap(period~., labeller = labeller(variable = label_parsed)) +
+  axis_labs_smol +
+  no_grid 
+```

### 2020-09-23 analysis/paper/paper.tex
fraft
llegamos a la conclusion!

@@
 
+\subsubsection{Data}
+
+To describe the Southern Annular Mode and its variability we used
+monthly geopotential height at 2.5\degree longitude by
+2.5\degree latitude of horizontal resolution and 37 vertical isobaric
+levels from ERA5 \citep{hersbach2020} for the period 1979 to 2018. We
+restrict our analysis to the post-satellite era to avoid any confounding
+factors arising from the introduction of satellite observations.
+
+We describe the relationship between the SAM indices and temperature and
+precipitation. We use temperature data from NOAA's Merged Land Ocean
+Global Surface Temperature Analysis V4.0.1 \citep{vose2012, smith2008},
+which blends land and ocean temperature analysis into a monthly global
+grid 5\degree lagitude by 5\degree longitude. For precipitation, we use
+monthly, 0.5\degree latitude by 0.5\degree longitude data from the
+Global Precipitation Climatology Centre \citep{schneider2015}.
+
 \subsubsection{Definition of indices}
 
-We defined the Southern Annular Mode (SAM) as the leading EOF of the
-monthly anomalies of geopotential field at 700 hPa south of 20\degree S
-(citation?). The EOF was performed by computing the Singular Value
-Decomposition of the data matrix consisting in 481 rows and 4176 columns
-(144 points of longitude and 29 points of latitude). The values where
-weighted by the square root of the cosine of latitude to account for the
-non-equal area of each gridpoint \citep{chung1999}. This same method was
-used at the rest of the levels considered in this paper.
+Traditionally the Souther Annular Mode (SAM) is defined as de leading
+Empiral Orthogonal Mode (EOF) of sea level pressure or geopotential
+height at lower levels \citep{ho2012}. Following \citet{baldwin2001}, we
+extend that definition vertically and use the term SAM to refer to the
+the leading EOF of the monthly anomalies of geopotential field south of
+20\degree S at each level. We performed EOFs by computing the Singular
+Value Decomposition of the data matrix consisting in 481 rows and 4176
+columns (144 points of longitude and 29 points of latitude). We weighted
+the values by the square root of the cosine of latitude to account for
+the non-equal area of each gridpoint \citep{chung1999}.
 
@@
 the SAM, we computed the zonal mean and anomalies of the full SAM
-spatial pattern. The results are shown in Figure~\ref{fig:method} for
-700hPa. The full spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is
-the sum of the zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\))
-and symmetric (\([\mathrm{EOF_1}](\lambda, \phi)\)) components. We then
+spatial pattern, as shown in Figure~\ref{fig:method} for 700hPa. The
+full spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is the sum of
+the zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\)) and
+symmetric (\([\mathrm{EOF_1}](\lambda, \phi)\)) components. We then
 compute the ``Full'', ``Asymmetric'' and ``Symmetric'' indices, by
-regressing each geopotential field on these patterns (weighting by the
-cosine of latitude).
-
-The three indices are normalised by dividing them by the standard
-deviation of the ``Full'' index at each level. This means that comparing
-the magnitude between indices is meaningful, but it also means that not
-every index will have unit standard deviation.
+regressing each geopotential field on the respective patterns (weighting
+by the cosine of latitude). The three indices are normalised by dividing
+them by the standard deviation of the ``Full'' index at each level. As a
+result, the magnitude between indices is comparable. However, only
+``Full'' index will have unit standard deviation per definition.
 
@@
 
-\subsubsection{Data}
+Our method assumes linearity in the asymmetric component of the SAM.
+That is, we assume that zonal symmetries associated with positive SAM
+are oposite and equal to the ones associated with negatie SAM.
+\citet{fogt2012}'s composites (their Figure 4) suggest that this might
+not be entirely valid, although we argue that much of that apparent
+non-linearity is due to the heterogenous nature of the selected years
+for constructing the composites. Using our data (from 1979 to 2018),
+seasonal composites of zonal anomalies of 700 hPa geopotential height
+for SAM+ (Full SAM index greater than 1 standard deviation) and SAM-
+(smaller than negative 1 standard deviation) show relatively high
+pattern correlations all seasons and are visually very linear (Figure
+A9). Therefore, we belive that our method is at the very least a
+reasonable approximation of the phenomenon.
 
-We used monthly geopotential height at 2.5 longitude by 2.5 latitude
-resolution from ERA5 \citep{hersbach2020} for the period 1979 to 2018.
+By computing a single EOF pattern using data for all months we are
+assuming that the zonal anomalies of the SAM are the same in all
+seasons. Geopotential zonal anomalies computed by projecting the first
+EOF \emph{of each season} are very similar to each other (Figure A10)
+and show pattern correlations between 0.65 (DJF with JJA) and 0.9
+(between MAM and SON). Based on this, we believe that our initial
+assumption is not unreasonable.
 
-Monthly temperature NOAA Global Surface Temperature (NOAAGlobalTemp) 5.0
-degree latitude x 5.0 degree longitude global grid
-\citep{vose2012, smith2008}. The same analysis was carried out using
-CRUTEM4 \citep{osborn2014} (not shown).
-
-We used monthly precipitation data from CPC Merged Analysis of
-Precipitation \citep{xie1997} 2.5 degree latitude x 2.5 degree
-longitude. CPCC: {[}schneider2015{]} \#FIXME
+Finally, we assume that the zonally asymmetric pattern is stationary in
+time. \citet{silvestri2009} suggest that this might not be the case
+between 1958 and 2004 but the period we analyse is much shorter
+(1979-2018) so it's unlikely that we could observe significant changes.
+Moreover, zonal asymmetry of the spatial patterns for the two halves of
+the period (1979 to 1998 and 1999 to 2018) show no systematic change
+(Figure A11).
 
@@
 
-Explain multiple regression interpretation \#FIXME.
+We perform linear regression to quantify the association between the SAM
+indices and other variables. Since the Asymmetric and Symmetric SAM
+indices are significantly correlated with each other, to capture the
+variability explained uniquely by each index we use one multiple linear
+regression instead of two simple linear regressions. To obtain the
+linear coefficients of a variable \(X\) (geopotential, temperature,
+precipitation, etc\ldots{}) with the Asymmetric SAM (\(SAM_a\)) and
+Symmetric SAM (\(SAM_s\)) we fit the ecuation
 
-We adjusted p-values for False Detection Rate following
-\citet{wilks2016}.
+\[
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) SAM_a + \beta(\lambda, \phi) SAM_s + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
+\]
+
+where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is
+the time, \(\alpha\) and \(\beta\) are the linear coefficients, \(X_0\)
+and \(\epsilon\) are the constant and error terms. From this equarion,
+\(\alpha\) represents the (linear) association of \(X\) with the
+variability of the Asymmetric SAM that is not explained by the
+variability of the Symmetric SAM; in other words, it is proportional to
+the partial correlation of \(X\) and the Asymmetric SAM, controlling for
+the effect of the Symmetric SAM and viceversa for \(\beta\).
+
+At 2.5\degree by 2.5\degree resolution, a single regression field is
+composed of thousands of regressions. In such case, using naive p-values
+to test for significance leads to misleading results
+\citep{walker1914, katz1991}. While there are multiple proposed
+solutions in the literature, \citet{wilks2016} suggests that adjunting
+p-values by controlling for the False Discovery Rate
+\citep{benjamini1995} is a simple and effective method to ameliorate
+this issue. Therefore, p-values showed in regression fields are all
+adjusted following \citet{benjamini1995}.
+
+When performing a separate regression for each trimester (DJF, MAM, JJA,
+SON) we first average the relevant variables to obtain a single value
+for each year and each trimester.
 
@@
 
-Figure \ref{fig:asymsam-timeseries} shows the Asymmetric and Symmetric
-time series corresponding to 700 and 50hPa. At first glance the series
-can be distinguished by their distributions. Whereas the tropospheric
-incides are aproximately normally distributed, the stratospheric indices
-are more long-tailed; that is, extreme values (both negative and
-positive) abound. The Asymmetric series have both more variability in
-the higher frequencies than the Symmetric series.
+The temporal evolution of the Assymmetric and Symmetric SAM was firstly
+asssesed. Figure \ref{fig:asymsam-timeseries} shows the corresponding
+time series for 700 hPa and 50 hPa and their corresponding density
+estimates. We selected these two levels as representative of the
+tropospheric and stratospheric variability respectively. As will be
+shown later, both indices are highly coherent within each atmospheric
+layer, therefore is reasonable to take one level as representative of
+each layer.
 
-The stratospheric Symmetric SAM varies strongly witha two-year period,
-which can be seen using spectral methods (Figure A3) or in the
-autocorrelation structure (Figure A4). There is a local peak at 2 years
-in the periodigram of the tropospheric Symmetric SAM also, although it's
-not statistically significant. In the troposphere the most significant
-peak of variability is found in the Asymmetric index at around 3.6
-months.
+Month-to-month variability is evident for both indices, with noisy
+variations in the low frequency. At first glance the series can be
+distinguished by their distributions. Compared to the stratospheric
+indices, the stratospheric indices are much more long-tailed; that is,
+extreme values (both negative and positive) abound. The Asymmetric
+series have both more variability in the higher frequencies than the
+Symmetric series.
+
+The stratospheric Symmetric SAM varies strongly with a two-year period,
+which can be seen by spectral analysis (Figure A3). This might suggests
+a link between stratospheric SAM variability and QBO. There is a local
+peak at 2 years in the periodigram of the tropospheric Symmetric SAM
+also, although it's not statistically significant. In the troposphere
+the most significant peak of variability is found in the Asymmetric
+index at around 3.6 months.
 
@@
 the extremes in the stratosphere, the Symmetric serie appears to lag the
-Asymmetric series (see, for example, the positive events on late 1987
-marked with a circle). We show these correlations, across all the levels
-of the reanalysis and for zero and -1 lag (Asymmetric index leading the
-Symmetric index), in Figure \ref{fig:cor-lev}.
+Asymmetric series (see, for example, the positive events on late 1987).
+We show these correlations, across all the levels of the reanalysis for
+zero and -1 lag (Asymmetric index leading the Symmetric index), in
+Figure \ref{fig:cor-lev}. Zero-lag correlations between the Asymmetric
+and Symmetric series are relatively constant throught the troposphere,
+fluctiating between 0.39 and 0.45. One-month-lag correlations are
+similarly constant but significantly reduced to around 0.17. In the
+stratosphere, zero-lag correlations drop to a minimum of 0.21 at 20 hPa
+and then it increases again monotonically with height up to the
+uppermost level of the reanalysis (although results near the top of the
+models are to be interpreted with care). At the same time, one-month-lag
+correlations increase with height. As a consequence, statospheric
+Asymmetric index tend to precede corresponding Symmetric index.
 
@@
 
-Zero-lag correlations between the Asymmetric and Symmetric series are
-relatively constant throught the troposphere, fluctiating between 0.39
-and 0.45. One-month-lag correlations are similarly constant but
-significantly reduced, hovering around 0.17. In the stratosphere,
-zero-lag correlations drop to a minimum of 0.21 at 20 hPa and then it
-increases again monotonically with height up to the uppermost level of
-the reanalysis. At the same time, one-month-lag correlations increase
-with height.
-
 \begin{figure*}
@@
 
-Figure \ref{fig:cross-correlation}a) shows (zero-lag) cross-correlation
+Figure \ref{fig:cross-correlation}a shows (zero-lag) cross-correlation
 across levels for the Full, Symmetric and Asymmetric SAM indices. For
@@
 correlation between levels falls off more rapidly, indicating less
-coherent (zero-lag) variability. Still there is a non negligible
+coherent (zero-lag) variability. Therefore, there is a non negligible
 correlation between the troposphere and the lower-to-middle
@@
 Symmetric SAM share the same high level of coherency in the troposphere
-but they differ in their stratospheric behaviour. As evidenced by the
-wider dark red areas near the diagonal in Figure
-\ref{fig:cross-correlation}b) vs.~Figure \ref{fig:cross-correlation}c),
-stratospheric coherency is stronger for the Asymmetric SAM than the
-Symemtric SAM. The stratospheric Symmetric SAM seems to connect more
-strongly to the trosposphere than the Asymmetric SAM; this can be seen
-by the lower correlation values in the top right left of Figure
-\ref{fig:cross-correlation}b) in comparison with Figure
-\ref{fig:cross-correlation}c).
+but they differ in their stratospheric behaviour. Stratospheric
+coherency is stronger for the Asymmetric SAM than the Symemtric SAM. The
+stratospheric Symmetric SAM seems to connect more strongly to the
+trosposphere than the Asymmetric SAM.
 
@@
 
-Figure \ref{fig:trends} shows normalised decadal trends for each index
-for the whole period 1979-2918 along with the 95\% confidence interval
-in shading for the whole year (row a) and separed by trimesters (rows b
-through e). As previously documented (e.g. \citet{fogt2020}), there is a
-statistically significant increase towards more positive SAM (panel
-a.1), which is evident only in Summer and Autumn (panels b.1 and c.1).
-We observe these increases mainly in the troposphere, reaching their
-maximum at at 100 hPa in Summer. By separating the SAM signal in its
-Asymmetric and Symmetric parts, we can not only see that these trends
-are almost entirely due to the Symmetric component (columns 2
-vs.~columns 3), but in some cases the trends become more clear. In
-Summer, the Asymmetric SAM has a statistically non significant negative
-trend in the middle troposphere that obscures the signal; as a result,
-trends computed using only the Symmetric component are more clear
-(compare the shading region in panel b.1 and b.3). In Autumn, using the
-Symmetric SAM reveals a statistically significant positive trend in the
-stratosphere that is not significant using the Full index.
+The trends for each of the indices (Full, Symmetric, Assymetric) were
+evaluate for the whole period 1979-2018 at each level (Figure
+\ref{fig:trends}) for the whole year and separed by trimesters. The Full
+SAM index presents a statistically significant trend (panel a.1) that
+extends throught the troposphere up to about 50 hPa and reaches its
+maximum value at 100 hPa. The seasonal trends (rest of column a)
+indicate that positive trends are present in Autumn and particularly in
+in Summer, where the 100 hPa maximum is much more defined. Positive
+trends have been documented by previous studies \citep[e.g.][ and
+references therein]{fogt2020} using indices of the SAM based on surface
+or near-surface circulation.
+
+By separating the SAM signal in its Asymmetric and Symmetric parts, we
+can not only see that these trends are almost entirely due to the
+Symmetric component (colum b vs.~column c), but in some cases the trends
+become more clear. In Summer, the Asymmetric SAM has a statistically non
+significant negative trend in the middle troposphere that obscures the
+trend in the Full index; as a result, trends computed using only the
+Symmetric component are more clear (compare the shading region in panel
+a.2 and c.2). In Autumn, using the Symmetric SAM reveals a statistically
+significant positive trend in the stratosphere that is not significant
+using the Full index.
 
@@
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.
-Points marked on panel b.2 are the location of the reference points used by \cite{raphael04} for its Zonal Wave 3 index.}\label{fig:2d-regr}
+Points marked on panel b.2 are the location of the reference points used by \cite{raphael2004} for their Zonal Wave 3 index.}\label{fig:2d-regr}
 \end{figure*}
@@
 more clearly dominated by a zonally symmetric, monopolar structure
-(panel a.1) which is, however, not perfectly centered in the south pole.
+(panel a.1) which is, however, not perfectly centered in the South Pole.
 The monopole obtained by multiple regression with the Asymmetric and
@@
 
-In the troposphere, panel a.2 shows the well known zonally symmetrical
-annular mode \emph{contaminated} with zonal asymmetries in the form of a
+In the troposphere, panel a.2 shows the well known combination of
+zonally symmetrical annular mode with zonal asymmetries in the form of a
 wave-3. The regression using the Asymmetric and Symmetric SAM indices
@@
 reference locations are marked with points in the figure. Thus, the
-tropospheric Asymmetric SAM index responds to zonal displacements in the
-position of the wave-3 pattern.
+tropospheric Asymmetric SAM index represents a zonal displacement in the
+position of the climatological wave-3 pattern.
 
@@
 
-The amplitude of each zonal wave number at each latitude at 50 hPa and
+The amplitude of first zonal wave numbers at each latitude at 50 hPa and
 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number
-zero represents the amplitude of the zonal mean. Comparing between rows,
-this Figure quantifies the relatively clean separation between the
-zonally symmetric and zonally asymmetric structures, as its evident how
-the mixture of waves of the Full field (column a) is very similar to the
-sum of the waves of the Asymmetric and Symmetric field (columns b and c,
-respectively). Column b of Figure \ref{fig:wave-amplitude} shows that
-the Asymmetric SAM is overwhelmingly dominated by wave 1 in the
-stratosphere (panel b.1), while in the troposphere it is composed of
-zonal waves 3 to 1 in decreasing level of importance (panel b.2).
+zero represents the amplitude of the zonal mean. Column b shows that the
+Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere
+(panel b), while in the troposphere it is composed of zonal waves 3 to 1
+in decreasing level of importance (panel b). Looking at panel b.2 from
+Figure \ref{fig:2d-regr}, it becomes apparent that zonal wavess 1 and 2
+modulate the amplitude of zonal wave 3, which --as mentioned before-- is
+larger in the Western Hemisphere than in the Easten Hemisphere.
 
@@
 asociated with the asymetric SAM index, we show a vertical cross section
-of regressions of mean geopotential height between 65ªS and 40ªS for the
-50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM
-index (panel b) (Figure \ref{fig:vertical-regression}).
-
-The geopotential anomalies associated with the stratospheric SAM (panel
-a) are clearly constrained to the stratosphere, which underscores the
-disconnect between the stratospheric and tropospheric Asymmetric SAM.
-The vertical structure this signal tilts about 60ª to the West between
-100 hPa and 1 hPa, suggesting baroclinic processes and polarward
-transport of heat (\#FIXME is this ok?). Interestingly, the signal in
-the stratosphere maximises near 10 hPa despite using the 50 hPa index
-for the regression.
+of regressions of mean geopotential height between 65\degree S and
+40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the
+700 hPa Asymmetric SAM index (panel b) (Figure
+\ref{fig:vertical-regression}). The geopotential anomalies associated
+with the stratospheric Asymmetric SAM (panel a) are clearly constrained
+to the stratosphere, which underscores the uncoupling between the
+stratospheric and tropospheric Asymmetric SAM. The vertical structure of
+this signal tilts about 60\degree to the West between 100 hPa and 1 hPa,
+suggesting baroclinic processes. Interestingly, the signal in the
+stratosphere maximises near 10 hPa despite using the 50 hPa index for
+the regression.
 
@@
 hPa in panel a. Conversely, for any tropospheric (below 100 hPa) index,
-the result is very similar to panel b. Pattern cross-correlation is
-greater than 0.9 within stratospheric index levels and tropospheric
-index levels and less than .4 accross stratospheric and tropospheric
-levels (Figure A8), i.e.~regressions obtained using Asymmetric SAM
-indices in the troposphere are very similar to each other and different
-to the ones obtaines using indices in the stratosphere and vice versa.
-The patterns mainly change in amplitude. The tropospheric pattern is
-maximised by the 300 hPa Asymmetric SAM index and the stratospheric
-pattern increased monotonically with height.
+the result is very similar to panel b. The patterns mainly change in
+amplitude.
 
 The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very
-similar to the teleconnection pattern associated with the ENSO. Indeed,
-\citet{fogt2011} showed that there is a significant relationship between
-the SAM and the ENSO. The correlation between the full SAM and the ENSO
-as measured by the Multivariate Enso Index \citep{wolter2011} is -0.19.
-This relationship is captured entirely the Asymmetric SAM, as this index
-has a partial correlation of -0.27 with the MEI, whereas the Symmetric
-SAM's partial correlation with the MEI is essentially null
-(\ensuremath{-7.9\times 10^{-4}}).
+similar to the Pacific-South American Pattern \citep{mo1987, kidson1988}
+which is a teleconnection pattern associated with the ENSO
+\citep{karoly1989}. Indeed, \citet{fogt2011} showed that there is a
+significant relationship between the SAM and the ENSO. The correlation
+between the full SAM and the ENSO as measured by the Oceanic Niño Index
+\citep{bamston1997} (ONI) is -0.16. This relationship is captured
+entirely the Asymmetric SAM, as this index has a partial correlation of
+-0.26 with the ONI controlling for the effect of the Symmetric SAM,
+whereas the Symmetric SAM's partial correlation with the ONI is
+essentially null (0.019). We performed the same analysis using the
+Multivariate Enso Index \citep{wolter2011} and the Southern Oscillation
+Index \citep{ropelewski1987} to conclude that these results do not
+depend on the ENSO index used.
 
 \subsection{Impacts}
-
-\subsubsection{Temperature}
 
@@
 
+The SAM has been shown to be associated with important surface variables
+such as temperature and precipitacion \citep[e.g.][and see
+\citet{fogt2020} for a review]{gillett2006}. Naturally, most studies on
+the surface impacts of the SAM are based on an index identical or
+analogous to what we call Full SAM index (\citet{fogt2012} being the
+only exception that we are aware of). We regress surface temperature and
+precipitation onto each of the three SAM indices to see if there are
+diferent surface impact associated with the asymmetric and symmetric SAM
+circulation.
+
 Figure \ref{fig:regr-air-season} shows regression coefficients of each
-index at 700 hPa with surface temperature for each trimester. It is
-evident that the Asymmetric and Symmetric SAM indices are associated
-with overall distinct temperature patterns which can be obscured when
-using the Full SAM index. The Symmetric SAM signal is weaker than the
-Asymmetric SAM, as evidenced by the relatively smaller and les
-sstatistically significant regression coefficients in row 3 of Figure
-\ref{fig:regr-air-season} compared with row 2.
+index at 700 hPa with surface temperature for each trimester. In Summer
+positive values of the Full SAM index (panel a.1) are associated with
+negative temperature anomalies near Antarctica which are surrounded by a
+ring of positive anomalies. The ring is not zonally symmetric, as there
+are three clear local maximums around 30\degree W, 15\degree E and
+50\degree E and a local minimum (with negative sign) around
+120\degree W. In the tropics, there are negative anomalies in the
+equatorial Pacific, consistent with the negative correlation between SAM
+and ENSO. Panels b.1 and c.1 show temperature anomalies associated with
+positive values of the Asymmetric and Symmetric SAM, respectively. Both
+the local maximums in the ring and the anomalies in the Pacific regions
+are present mostly on the Asymetric SAM regression map, while
+temperature patterns linked to positive Symmetric SAM show a more
+zonally consistent ring and less relation to the tropics. Noticeable,
+temperature anomalies in the Indian ocean, South Africa and Australia
+are strongly related to positive values of Asymmetric SAM. This signal
+is not present in the regression pattern with the Full SAM. Spring (row
+4) features very similar patterns but of generally smaller in magnitude
+and statistical significance.
 
-In DJF (row 1), the strong negative signal in the tropical Pacific in
-panel a.1 is mostly associated with the Asymmetric component (panel
-b.1), as is it largely absent in the Symmetric component (panel c.1).
-Furthermore, the Asymmetric SAM is also asociated with low temperature
-anomalies in the Indian ocean, but this signal is obscured by the
-Symmetric variability and thus lost in the Full SAM. Over the
-continents, the Asymmetric SAM is assoiated with negative temperature
-anomalies which, again, mostly disappear in the Full SAM regression.
-Temperature regression with the Symmetric SAM (panel c.1) shows a ring
-of negative anomalies around Antarctica surrounded by positive
-anomalies. This pattern of anomalies is consistent with thermal wind
-balance and the intesification and pole-ward migration of the westerlies
-commonly linked to the SAM.
+In Autumn and Winter (rows 2 and 3) the postive ring is only present
+through its local maximums in the regression with the Full SAM. There
+are also negative anomalies in Southern Australia, and positive
+anomalies over New Zealand and Southern South America. These patterns
+are not significant in the sense that there are no areas with p-values
+below 0.05 when controlling for FDR following \citet{wilks2016}.
+However, repeating this analysis with 2-meter temperature from ERA5
+resulted in similar patterns that were statistically significant.
+Moreover, similar features were observed in station measurements by
+\citet{jones2019}, although using data from 1957 to 2016.
 
-The patterns seen in MAM and JJA (rows 2 and 3) are not robustly
-significant in the sense that there are no areas with p-values below
-0.05 when controlling for FDR following \citet{wilks2016}. Nevertheless,
-it is interesting to note that in both trimesters, the sign of the
-regression is consistently flipped between the Asymmetric and Symmetric
-regressions. In South America, for example, the Asymmetric SAM is
-associated with positive temperature anomalies in MAM and negative
-temperature anomales in JJA, while the oposite is the case for the
-Symemtric SAM.
+The pattern of negative anomalies in the pole surrounded by positive
+anomalies roughly seen in all seasons --although with varying intensity
+and small-scale details-- is consistent with the intensification and
+poleward migration of the westerlies commonly linked to the SAM. It's
+then not surprising to see it more clearly in association with the
+Symemtric SAM (at least in Summer and Spring).
 
-Finally, in SON (row 4) there is no significant temperature signal
-associated with the Symmetric SAM (panel c.4), while the Asymmetric SAM
-shows a relatively robust signal in the equatorial Pacitic, Australia,
-and even Southeast South America.
+These results suggests that Asymmetric and Symmetric SAM indices are
+associated with overall distinct temperature patterns which may not be
+apparent when using the Full SAM index.
 
-\subsubsection{Precipitation}
+Figure \ref{fig:regr-air-season} column b can be partially compared with
+Figure 11 from \citet{fogt2012}. Although they used station data from
+1958 to 2001, a lot of the characteristics are reproduced here, such as
+the strong signal in New Zealand and Australia in Summer and Spring.
 
-Regression of the SAM indicies with seasonal mean precipitation are
-shown Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america}
-for Australia and New Zealand, and South America respectively. (We
-didn't detect any significant signal in South Africa.)
+Regression of the SAM indicies with seasonal mean precipitation and 700
+hPa geopotential height are shown Figures \ref{fig:pp-regr-oceania} and
+\ref{fig:pp-regr-america} for Australasia and South America
+respectively. South Africa is not shown because no significant signal
+was detected there.
 
@@
 
-In Australia (Figure \ref{fig:pp-regr-oceania}), the annual-level
-regression shows that the Full SAM is associated with a statistically
-significant increase in precipitation in the Southeastern region (panel
-a.1), which reproduces the results from \citet{gillett2006}. The
-separation between Asymmetric and Symmetric SAM suggest that this
-increase is explained by the Symmetric SAM only in the East coast (panel
-c.1), which is consistent with the increased easterly flow clearly seen
-in relation with this index. The Asymmetric SAM appears related to
-increased precipitation in the West coast of Southeastern Australia
-(panel b.2), explained by the anomalous \emph{westerly} circulation
-transporting moist air to the continent.
+In Australia, the annual regression shows that the Full SAM index is
+positively associated with precipitation in the Southeastern region
+(Figure \ref{fig:pp-regr-oceania} panel a.1), which reproduces the
+results from \citet{gillett2006}. The separation between Asymmetric and
+Symmetric SAM suggest that this positive anomaly is explained by the
+Symmetric SAM only in the East coast (panel c.1). Geopotential anomalies
+associated with this index (black contours) are indicative of easterly
+flow from the Tasman Sea, which could explain the positive anomalies in
+precipitation as found by \citet{hendon2007}. The Asymmetric SAM appears
+related to increased precipitation in the West coast of Southeastern
+Australia (panel b.2), which could similarly be explained by the
+anomalous westerly circulation transporting moist air to the continent
+from the Indian Ocean.
 
-The seasonal-level regressions show statistically significant anomalies
-only in SON, with a pattern similar to the annual-level regression
-(panel a.5). Panels b.5 and c.5 don't show a clear separation between
-the Asymmetric and Symmetric SAM. If anything, the positive and more
-significant regression coefficients in panel b.5 vs pane c.5 would
-suggest more influence of the Asymmetric than the Symmetric SAM, going
-against the interpretation gathered from the annual-level regressions.
 This Spring signal is broadly consistent with \citet{hendon2007}, but
@@
 coeffcients have the consistent sign).
+
+The seasonal-level regressions show statistically significant anomalies
+only in Spring, when positive Full SAM is associated with positive
+precipitation anomalies in Eastern Australia (panel a.5). In this
+trimester the Symmetric SAM seems to be associated with precipitation in
+a relatively reduced area of the East Coast (panel c.5) while the
+positive precipitation anomalies related with positive Asymmetric SAM
+affect all Estern Australia (panel b.5).
+
+In Summer, positive Full SAM index is associated with with positive
+precipitation anomalies in Western and Eastern Australia, particularly
+in the North East (panel a.2). The Eastern part being dominated by the
+relationship with the Symmetric SAM and the Western, by the Asymemtric
+SAM. In Autumn, the regression with Full SAM shows positive values in
+the North, similar to Summer, and a broad area of positive values in the
+North-East to South-West direction. This structure seems to be
+associated with the Symmetric SAM, while the Northern positive values
+are associated with the Asymmetric SAM. In Winter we see the same NE to
+SW aligned anomaly (although with much reduced amplitude) that is also
+present only in relation with the Symmetric SAM. None of these
+regression coefficients are statistically signifincat at the 95\% level
 
@@
 and Southern Chile and non-significant increase in South Brazil, near
-the South Atlantic Convergence Zone (SACZ) (panel a.1).
+the South Atlantic Convergence Zone (SACZ) (panel a.1). Panels b.1 and
+c.1 show a remarkably clean separation between the Asymmeric SAM
+--associated with the Southeastern South American and Southern Brazilian
+signals-- and the Symmetric SAM --associated with the signal in Southern
+Chile.
 
-Panels b.1 and c.1 show a remarkably clean separation between the
-Asymmeric SAM --associated with the Southeastern South American and
-Southern Brazilian signals-- and the Symmetric SAM --associated with the
-signal in Southern Chile. This separation is consistent with the
-mechanisms responsible for these effects. In Southern Chile, the reduced
-westerly flow reduce moisture transport from the Pacific Ocean (cita??
-\#FIXME).
+Except Winter, seasonal-level regressions mirror this same pattern. Even
+if not statistically significant, they all show negative values in
+Southeastern South America and Southern Chile along with positive values
+in Southern Brazil in relation with the Full SAM. The separation of
+these features between the Asymmetric SAM and Symmetric SAM regression
+maps is also rather consistent.
 
-In Southeastern South America, anomalous meridional winds lead to less
-precipitation by inhibiting moisture convergence from the South American
-Low Level Jet \citep{silvestri2009}. The increased precipiation in the
-South Atlantic Convergence Zone, on the other hand, appears to be
-related to modulation of the anomalous SAM circulation in SACZ events
-which lead to more frequent and intense SACZ events during positive SAM
+The anomalous circulation at 700 hPa associated with the Symmetric SAM
+(panel c.1) indicate anomalous Easterly flow over Southern Chile. This
+leads to reduced influx of moist air from the Pacific Ocean which, is
+the main source of precipitable water in that region. On the other hand,
+the anomalous circulation associated with positive values of Asymmetric
+SAM (panel b.1) in the Atlantic is anticyclonic in the South and
+cyclonic in the North. This creates anomalous South-Easterly flow over
+Southeastern South America, which inhibits the flow of the Low Level Jet
+to the region \citep[\citet{zamboni2010}]{silvestri2009}. This same
+pattern was found to be associated with increased precipitation in
+Southern Brazil during South Atlantic Convergence Zone events
 \citep{rosso2018}.
@@
 
-Except during Winter, the seasonal-level regression all show these
-similar patterns although not as cleanly and only in some cases
-statistically significant.
-
-\subsubsection{Sea ice}
-
-\begin{figure*}
-\includegraphics{regr-ice-1} \caption[Seasonal regression of SAM indices with sea ice concentration]{Seasonal regression of SAM indices with sea ice concentration. \#FIXME}\label{fig:regr-ice}
-\end{figure*}
-
-Regressions between the Full SAM index and Antarctic Sea Ice
-Concentrations (Figure \ref{fig:regr-ice}) show a great deal of
-variability across seasons. The only statistically significant signal is
-in Spring, when we observe negative concentraion anomalies in the
-Northen Weddell Sea (panel a.4) explained by the Asymmetric SAM (panel
-b.4). Both in Winter and in Spring the Asymmetric SAM is associated with
-bigger Sea Ice Concentration anomalies in West Antarctica than East
-Antarctica, with generally decreased concentration East of the Antarctic
-Peninsula and increased concentration to the West, as expected from the
-anomalous circulation correlated with this index. The Symmetric SAM
-signal appears more evenly distributed accross the whole ice sheet.
-
 \subsection{Conclusions}
 
-We presented a method to systematically separate the zonally asymmetric
-and zonally symmetric components of the SAM into two indices. As
-expected, they are significantly correlated with each other, but they
-also have their own unique dynamics such as differnet vertical coherence
-patterns, variability and long term trends.
+In this study we tried to systematically characterise the variability of
+the zonally symmetric and zonally asymmetric structure of the SAM. By
+projecting monthly geopotential fields at each level with the
+corresponding asymmetric and symmetric pattern, we created two indices
+representing the zonally asymmetric and zonally symmetric contributions
+of the SAM respectively.
 
-The spatial structure and temporal evolution of the SAM show a strong
-separation between the stratosphere SAMs. The first EOF in the
-stratosphere is monopolar in nature while the first EOF in the
-troposphere is a proper annular mode. Their repespective departures from
-the zonal mean are also different: a zonal wave 1 dominates the
-stratospheric SAM; waves 3 and 2 dominate the tropospheric SAM.
-Furthermore, there is little temporal correlation between the
-tropospheric and stratospheric time series.
+As expected, the Asymmetric SAM index correlates strongly with the
+Symmetric SAM index. In the troposphere, this correlation is maximum at
+zero lag, while in the stratosphere is maximised with the Asymmetric SAM
+leading the Symmetric SAM by one month. Since most indices of the SAM
+are calculated using suftace or near-surface conditions, this result
+would suggest that they might not be sensitive to the most dramatic
+changes in SAM variability.
 
-The zonal asymmetric component of the SAM at each level is even more
-decupled between the troposphere and the stratosphere. Their temporal
-evolution shows essentialy zero correlation (Figure
-\ref{fig:cross-correlation}) and the signal associated with the
-stratospheric Asymmetric SAM is completely restricted to the
-stratosphere (Figure \ref{fig:vertical-regression} panel a).
-Geopotential height anomalies associated with the tropospheric
-Asymmetric SAM, on the other hand, do extend to the stratosphere, but
-those anomalies do not project strongly into the stratospheric
-Asymmetric SAM.
+The two-year periodcty we found in the stratospheric Symmetric SAM might
+point to a link between the SAM and the Quasi Biennial Oscillation.
+There is evidence of influence between the QBO and the Northern Annular
+Mode \citep[e.g.][\citet{watson2014}, \citet{zhang2020}]{holton1980}, so
+it's not unlikely that the SAM would be similarly connected. However
+establishing this link would require further research.
 
-We show that the observed positive trend towards positive SAM is
-restricted to the tropospheric SAM and is explained by the Symmetric
-component (Figure \textbackslash{}ref\{fig:trends{]}). However, the
-degree of asymmetry appears to have increased slightly in the last 40
-years, as the Asymmetric SAM explains an increasingly proportion of the
-variance (Figure \ref{fig:r-squared-trend}).
+As documented by previous studies, such as \citet{fogt2020} (and
+references therein), we observe a positive trend towards positive SAM in
+Summer and Autumn. We show that these trends are maximised at the 100
+hPa level and are explained by the zonally symmetric component. We also
+find a statistically significant positive trend in the Symmetric
+component of the SAM in the stratosphere that is not apparent in the
+Full SAM index. In contrast to \citet{fogt2012} we find some evidence of
+the SAM becoming more zonally asymmetric, as there is a slight positive
+trend in the variance explained by the as the Asymmetric SAM explains an
+increasingly proportion of the total variance.
 
-Temperature and precipitation anomalies associated with the SAM respond
-to various processes in different locations. The Asymmetric and
-Symmetric indices do a reasonable job separating some of them. The
-strong sumertime temperature anomalies in the equatorial Pacific are
-explained by the Asymmetric component, most likely due to its
-relationship with ENSO. The Symmetric SAM, on the other hand, captures
-the change in meridional temperature gradient, which is linked to the
-SAM-zonal wind relationship by thermal wind balance.
+In the troposphere, the spatial patterns of geopotential associated with
+the Symmetric SAM is much closer to being trully annular than the
+patterns associated with the Full SAM index. The Asymmetric SAM, on the
+other hand, describes a wave-3 pattern with maximum amplitude in the
+Pacific region and whose phae is rotated a quarter wavelength from the
+mean zonal wave 3 described by \citet{raphael2004}'s index. This pattern
+extends in the troposphere but its maximum is located at 250 hPa, which
+also could suggest that surface-based indices are not optimum for
+capturing this variability.
 
-The patterns of SAM-associated precipitation anomalies are similarly
-well separated. In South America, we show that negative anomalies
-observed in Chile related to the SAM are well explained by the Symmetric
-component, while the precipitation dipole in Southern South America and
-the South Atlantic Convergence Zone is explained by the Asymmetric
-component.
+This wave-3 pattern is similar to the Pacific-South American Pattern,
+which is a teleconnection pattern linked to ENSO variability. We found
+that the significant correlation that exists between the Full SAM index
+and the Oceanic Niño Index is captured entirely by the Asymmetric SAM
+index. This suggests that ENSO is linked to SAM exclusively through the
+variability in the latter's Asymmetric compoment.
 
-\subsubsection*{Limitations}
-\addcontentsline{toc}{subsubsection}{Limitations}
+Temperature anomalies associated with the Full SAM broadly show a
+pattern of negative anomalies at polar latitudes surrounded by positive
+anomalies, but with many deviations from symmetry. The Asymemtric SAM
+index explains a big portion of these deviations. In particular, the
+positive phase of the Asymmetric SAM is associated with colder
+temperatures over Southern Brazil, South Africa and Southern Australia,
+as well as the negative anomalies in the equatorial Pacific consistent
+with the ENSO-SAM relationship delineated above. These are particularly
+clear in the DJF and SON trimesters, which include the months in which
+the ENSO teleconnection is more active \citep[\citet{fogt2011},
+\citet{cai2020a}]{cazes-boezio2003}.
 
-Our method assumes linearity in the asymmetric component of the SAM.
-That is, assumes that zonal symmetries associated with positive SAM are
-oposite and equal to the ones associated with negatie SAM.
-\citet{fogt2012}'s composites suggest that this might not be entirely
-valid, although we argue that much of that apparent non-linearity is due
-to the heterogenous nature of the selected years for constructing the
-composites. Using our data (from 1979 to 2018), seasonal composites of
-zonal anomalies of 700 hPa geopotential height for for SAM+ and SAM-
-show pattern linear correlations greater than -0.7 for all seasons and
-are visually very linear (Figure A9). Therefore, we belive that our
-method is at the very least a reasonable approximation of the fenomenon.
+In Australia the Full SAM is associated with positive precipitation
+anomalies in South East and this is explained by the Symemtric SAM.
+However, the Asymmetric SAM is associated with a small area of positive
+precipitation anomalies in the Eastern Coast of West Australia, maybe
+due to advection of moist air from the Indian Ocean.
 
-We also asumed that the structure of the SAM zonal anomalies is stable
-in all seasons. Again, this is not unreasonable, as geopotential zonal
-anomalies computed by projecting the first EOF \emph{of each season} are
-very similar to each other (Figure A10).
+In South America, precipitation anomalies associated with the Full SAM
+are negative both in Southern Chile and Southeastern South America, and
+positive in Southern Brazil. This features are cleanly separated between
+the Asymmetric and Symmetric components. The Symmetric SAM explains the
+negative anomalies in Southern Chile and the Asymmetric SAM, the
+negative-positive dipole between Southeastern South America and Southern
+Brazil. Individual seasons mostly follow this pattern.
 
-\citet{silvestri2009} showed that impacts linked to the SAM changed
-rather dramatically before and after 1980. In particular, the negative
-relationship with precipitation in South America (consistent with Figure
-\ref{fig:pp-regr-america} panel a.1) was absent in some areas and
-switched sign in other in the earlier period. The correlation between
-ENSO and SAM is similarly non-stationary, also disapearing before 1973.
+\citet{silvestri2009} suggests that precipitation impacts linked to the
+SAM changed rather dramatically before and after 1980. In particular,
+the negative relationship with precipitation in South America was absent
+in some areas and switched sign in other in the earlier period. The
+correlation between ENSO and SAM is similarly non-stationary, also
+disapearing before 1973.
 
@@
 \acknowledgments
-
-CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder,
-Colorado, USA, from their Web site at https://psl.noaa.gov/ \#FIXME
 
@@
 \begin{figure}
-\includegraphics{A4-1} \appendcaption{A4}{Autocorrelation functions of each timeseries}\label{fig:A4}
-\end{figure}
-
-\begin{figure}
 \includegraphics{A8-1} \appendcaption{A8}{Pattern cross-correlation \#FIXME!}\label{fig:A8}
@@
 
+\begin{verbatim}
+#> # A tibble: 12 x 3
+#>    item1 item2 correlation
+#>    <fct> <fct>       <dbl>
+#>  1 MAM   DJF     -1.16e-33
+#>  2 JJA   DJF      3.07e-34
+#>  3 SON   DJF      2.55e-33
+#>  4 DJF   MAM     -1.16e-33
+#>  5 JJA   MAM      1.84e-33
+#>  6 SON   MAM      1.53e-32
+#>  7 DJF   JJA      3.07e-34
+#>  8 MAM   JJA      1.84e-33
+#>  9 SON   JJA     -4.03e-33
+#> 10 DJF   SON      2.55e-33
+#> 11 MAM   SON      1.53e-32
+#> 12 JJA   SON     -4.03e-33
+\end{verbatim}
+
 \begin{figure}
@@
 
+\begin{figure}
+\includegraphics{A11-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height computed for the periods 1979 to 1998 and 1999 to 2018]{Spatial patterns of the first EOF of 700 hPa geopotential height computed for the periods 1979 to 1998 and 1999 to 2018.
+Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right).
+Pattern correlation between both fields is 0.77.
+Arbitrary units.}\label{fig:A11}
+\end{figure}
+
 \end{document}

### 2020-09-23 R/data_locations.R

@@ -75,7 +75,7 @@ GPCP <- function() {
 
 #' @export
 #' @rdname data_locations
-CPCC <- function() {
+GPCC <- function() {
    file <- data_path("raw", "precip.mon.total.v2018.nc")
    checkmate::assert_access(file, access = "r")
    class(file) <- c("nc_file", class(file))

### 2020-09-23 R/maps.R

@@ -8,11 +8,14 @@
 #' @param ... other arguments passed to [ggplot2::geom_polygon()]
 #'
 #' @export
-geom_qmap <- function(subset = identity, color = "black", size = 0.2,
+geom_qmap <- function(subset = identity, color = "gray50", size = 0.2,
                       fill = NA, wrap = c(0, 360), weighting = 0.7,
                       keep = 0.015, ...) {
   lon <- lat <- group <- NULL
-  data <- ggplot2::fortify(map_simple(wrap = wrap, keep  = keep, weighting = weighting)) %>%
+  data <- map_simple(wrap = wrap, keep  = keep, weighting = weighting)
+
+  data <- data %>%
+    fortify() %>%
     data.table::as.data.table() %>%
     .[, c("long", "lat", "group")] %>%
     data.table::setnames("long", "lon")
@@ -27,7 +30,7 @@ geom_qmap <- function(subset = identity, color = "black", size = 0.2,
                         ...)
 }
 
-map_simple_ <- function(wrap = c(0, 360), keep = 0.015, weighting = 0.7) {
+map_simple <- function(wrap = c(0, 360), keep = 0.015, weighting = 0.7) {
   map <- maps::map("world", fill = TRUE,
                    col = "transparent", plot = FALSE, wrap = wrap)
   IDs <- vapply(strsplit(map$names, ":"), function(x) x[1],
@@ -42,9 +45,12 @@ map_simple_ <- function(wrap = c(0, 360), keep = 0.015, weighting = 0.7) {
   map
 }
 
+fortify <- ggplot2::fortify
+
+# zzz.R
+.onLoad <- function(pkgname, libname) {
+  map_simple <<- memoise::memoise(map_simple)
+  fortify <<- memoise::memoise(fortify)
+}
 
 
-
-#' @export
-#' @rdname geom_qmap
-map_simple <- memoise::memoise(map_simple_)

### 2020-09-23 analysis/scripts/01-get-data.R

@@ -14,7 +14,7 @@ files <- list(aao = "aao.csv",
               NOAAGlobalTemp = "air.mon.anom.nc",
               cmap = "precip.mon.mean.nc",
               hadcrut = "HadCRUT.4.6.0.0.median.nc",
-              cpcc = "precip.mon.total.v2018.nc",
+              gpcc = "precip.mon.total.v2018.nc",
               gpcp = "gpcp.precip.mon.mean.nc",
               era5 =  "era5.z.mon.mean.nc")
 
@@ -28,6 +28,20 @@ aao <- rsoi::download_aao(use_cache = !force, file = asymsam::data_path("raw", f
   .[, .(time = lubridate::as_datetime(Date), aao = AAO)]
 
 saveRDS(aao, asymsam::data_path("derived", "aao.Rds"))
+
+# Download QBO
+qbourl <- "https://psl.noaa.gov/data/correlation/qbo.data"
+
+qbo <- data.table::fread(qbourl, na.strings = "-999.00", skip = 1, header = FALSE,
+             nrows = 73, col.names = c("year", as.character(1:12))) %>%
+  data.table::melt(id.var = "year", value.name = "qbo") %>%
+  na.omit() %>%
+  .[, time := lubridate::make_datetime(year, variable)] %>%
+  .[, year := NULL] %>%
+  .[, variable := NULL]
+
+saveRDS(qbo, asymsam::data_path("derived", "qbo.Rds"))
+
 
 # Download GISTEMP --------------------------------------------------------
 url <- "ftp://ftp.cdc.noaa.gov/Datasets/gistemp/landonly/1200km/air.2x2.1200.mon.anom.land.nc"
@@ -72,7 +86,7 @@ if (!force && file.exists(asymsam::data_path("raw", file))) {
 
 # Download CMAP -----------------------------------------------------------
 url <- "ftp://ftp.cdc.noaa.gov/Datasets/gpcc/full_v2018/precip.mon.total.v2018.nc"
-file <- files$cpcc
+file <- files$gpcc
 
 if (!force && file.exists(asymsam::data_path("raw", file))) {
   message("File already exists.")
@@ -144,71 +158,71 @@ if (!force && file.exists(asymsam::data_path("raw", file))) {
 #
 # asymsam::nc_concatenate(ncepfiles, asymsam::data_path("raw", "ncep.daily.nc"), overwrite = TRUE)
 
-# Antarctic sea ice
-# from https://nsidc.org/data/G02135
-months <- 1:12
-years <- 1979:2019
[…]

### 2020-09-23 analysis/paper/paper.Rmd
casi casi

@@
 ---
-twocol: TRUE  #  Use FALSE for submission. 
-title: Title here
+twocol: FALSE  #  Use FALSE for submission. 
+title: New method to describe the zonal symmetries and asymmetries of the Southern Annular Mode
 author1: 
@@
   name: Leandro Díaz
-affiliation: "CIMA UBA blablabla"
+affiliation: |
+  Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+  CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+  CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
 extra_authors: 
@@
 
-yada yada SAM yada yada circulation.. yada yada so important. yada yada many impacts.
-
-@fogt2012 studied the characteristics of the asymmetric structure of the SAM.
-It computed the zonally anomalous component of mean sea level pressure (MSLP) composites for positive and negative SAM events and created two indices by projecting MSLP fields ontno them.
-However, the use of composites leads to some issues that affects the interpretability of the results.
-First, they can be dependent on the choice of threshold used to define positive and negative events.
-Secondly, by discarting data that don't meet the threshold, they don't use all the information avaiable.
-Due to the realtive short timeframe used, this leads to some composites being composed of as little as 4 years.
-Third, the resulting composites corresponding to each polarity and season are derived from the average of different amount of fields and from different years.
-This last issue is particularly important in light of the changing structure of the SAM before and after 1980 [@silvestri2009].
-In @fogt2012, the DJF SAM+ composite uses only 7 years, 5 of which are later than 1988, whereas all of the 8 years used for their DJF SAM- composite are from earlier than 1988.
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a; fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [@fogt2020].
+Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
+
+However, computed as the leading Empirical Orthogonal Function (EOF) of Sea Level Pressure or low-level geopotential height, the SAM spatial structure contains noticeable deviations from this zonally symmetric description, particularly in the Pacific Ocean region.
+These zonal asymmetries are not widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009].
+At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availabilty of dense observations in the Southern Hemisphere [@jones2009].
+
+We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of @fogt2012.
+However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988).
+This is particularly important due to the inhomogeneities in renalysis products prior to the sattelite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].
+
+Our objective is, then, to systematically characterise the zonally asymmetric component of the SAM variability by constructing two indices which aim to capture exclusively the variability of the symmetric and asymmetric component each.
+We then analyse their temporal variability, trends and vertical coherence.
+We study the spatial patterns described by the variability exclusive to each index.
+Finally, we investigate theri relationship with temperature and recipitation anomalies.
 
@@
 The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
-We then compute the "Full", "Asymmetric" and "Symmetric" indices, by regressing each geopotential field on the respective patterns (weighting by the cosine of latitude).
+We then compute the "Full SAM", "Asymmetric SAM" and "Symmetric SAM" indices as the regression coefficients of the regression ofeach monthly geopotential field on the respective patterns (weighting by the cosine of latitude).
 The three indices are normalised by dividing them by the standard deviation of the "Full" index at each level.
@@
 However, only "Full" index will have unit standard deviation per definition.
+From the regression, we also use the explained variance of each pattern as a indication of the degere of symmetry or asymmetry of each monthly field.
 
 ```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height.
-Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right).
+(a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
 Arbitrary units.", fig.width=6, fig.height=2.5, fig.env="figure*"}
@@
 
-```{r asymsam-timeseries, fig.cap = "Time series for the asymmetric SAM and symmetric SAM and density estimates.", fig.width=6, fig.height=3, fig.env="figure*"}
-
+```{r asymsam-timeseries, fig.cap = "Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
+To the left, probability density estimate of each index.
+Series are standarised by the standard deviation of the Full SAM at each level.", fig.width=6, fig.height=3, fig.env="figure*"}
 indices %>% 
@@
   scale_y_continuous(NULL) +
-  # scale_x_datetime(NULL, expand = c(0, 0)) +
+  scale_x_datetime(NULL, expand = c(0, 0)) +
   facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
@@
 At the same time, one-month-lag correlations increase with height.
-As a consequence, statospheric Asymmetric index tend to precede corresponding Symmetric index.
-
-```{r cor-lev, fig.cap = "Correlation between the Symmetric and Asymmetric SAM at each level for lag zero and lag -1 (Asymmetric leads Symmetric).", fig.width=3, fig.height=3}
+As a consequence, statospheric Symmetric index tend to precede corresponding Asymmetric index.
+
+```{r cor-lev, fig.cap = "Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Symmetric leads Asymmetric).", fig.width=3, fig.height=3}
 cors %>% 
@@
   scale_linetype_manual(NULL, values = c(1, 5), labels = c("0" = "Zero-lag", 
-                                                           "-1" = "Asymmetric leads\nSymmetric (1 month)")) +
+                                                           "-1" = "Symmetric leads\nAsymmetric by 1 month")) +
   scale_color_brewer(NULL, palette = "Dark2", labels = c("0" = "Zero-lag", 
-                                                         "-1" = "Asymmetric leads\nSymmetric (1 month)")) +
+                                                         "-1" = "Symmetric leads\nAsymmetric by 1 month")) +
   coord_flip()
@@
 
-```{r cross-correlation, fig.cap = "Cross correlation between levels of the Full, Asymmetric and Symmetric SAM.", fig.width = 6, fig.height=3, fig.env="figure*"}
+```{r cross-correlation, fig.cap = "Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM.", fig.width = 6, fig.height=3, fig.env="figure*"}
 tri <- matrix(c(1000, 1000, 1000, 1, 1, 1, 1000, 1000), ncol = 2, byrow = TRUE) %>% 
@@
 
-```{r trends, fig.cap = "Decadal normalised trends for each index at each level for annual (row a) and seasonal values (rows b-e) for the period 1979-2018.
+```{r trends, fig.cap = "Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
 Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
@@
   copy() %>% 
-  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
+  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
   .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
@@
   .[, season := season(time)] %>% 
-  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
+  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
   .[, FitLm(estimate_norm, time = time, se = TRUE), by = .(type = term, lev, season)] %>% 
@@
   .[, std.error := std.error*3600*24*365*10] %>% 
+  .[, pval := Pvaluate(estimate, std.error, df, "fdr")] %>% 
   setnames("type", "term") %>% 
@@
   ggplot(aes(lev, estimate)) +
+  geom_hline(yintercept = 0) +
   geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), 
@@
   geom_line() +
+  # stat_subset(aes(subset = pval <= 0.05), shape = 4, size = 2) +
   scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
-  geom_hline(yintercept = 0) +
   scale_x_level() +
@@
 
-```{r r-squared-trend, fig.height=3, fig.width=3, fig.cap = "Decadal trends for explained variance of each index at each level for the period 1979-2018.
+```{r r-squared-trend, fig.height=3, fig.width=3, fig.cap = "Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979-2018.
 Shading indicates the 95\\% confidence interval."}
@@
   scale_x_level() +
-  scale_y_continuous("Trend in R^2") +
+  scale_y_continuous("Change in explained variance per decade", label = scales::percent_format()) +
   scale_fill_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill"), labels = lab_sam) +
@@
 
-```{r 2d-regr, fig.cap = "Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM.
+```{r 2d-regr, fig.cap = "Regression patterns of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM.
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.
-Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index.", fig.env = "figure*", fig.width=7, fig.height=5}
+Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index.
+", fig.env = "figure*", fig.width=7, fig.height=5}
 indices_2d_yearly %>% 
@@
   scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL,
+                   minor_breaks = NULL) +
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
@@
   axis_labs_smol  +
-  no_grid
+  no_grid +
+  theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
 ```
@@
 
-```{r wave-amplitude, fig.cap = "Planteray wave amplitude for the regression patterns at 50 and 700 hPa. Note the varying x axis.", fig.width=3, fig.height=3}
+```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.
+Note the differnet x axis.", fig.width=3, fig.height=3}
 indices_2d_yearly %>% 
@@
   indices_for_regr[., on = .NATURAL, allow.cartesian = TRUE] %>% 
-  .[, combine_lists(FitLm(hgt_a, full, se = TRUE),
-                    FitLm(hgt_a, sym, asym, se = TRUE)),
+  .[, FitLm(hgt_a, asym, sym, se = TRUE),
     by = .(lon, lev, index_level)] %>% 
@@
 
-```{r vertical-regression, fig.cap = "Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South.
-(\\#FIXME this caption needs some love)", fig.width = 3, fig.height=3}
+```{r vertical-regression, fig.cap = "Regression between monthly geopotential anomalies (meters) averaged betweeen 65\\degree and 40\\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
+(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa.", fig.width = 3, fig.height=3}
 vertical_regr %>% 
@@
 
-```{r}
-#  enso %>% 
-#   .[indices, on = "time", allow.cartesian = TRUE] %>% 
-#   .[, with(cor.test(estimate, value), 
-#            list(cor = estimate, 
-#                 low = conf.int[1], 
-#                 upp = conf.int[2])), by = .(lev, term, index, season(time))] %>% 
-#   .[index == "oni"] %>% 
-# ggplot(aes(lev, cor)) +
-#   geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha  = .3) +
-#   geom_line(aes(color = term)) +
-#   scale_x_level() +
-#   coord_flip() +
-#   facet_wrap(~season)
-# 
-# enso %>% 
-#   .[indices_wide, on = "time", allow.cartesian = TRUE] %>% 
-#   .[, partial_cor(value, asym, sym), by = .(index, lev, season(time))] %>% 
-#   .[index == "oni"] %>%
-# ggplot(aes(lev, partial_correlation)) +
-#   # geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha  = .3) +
-#   geom_line(aes(color = term, group = interaction(term, index))) +
-#   scale_x_level() +
-#   coord_flip() +
-#   facet_wrap(~season)
-```
-
 The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
@@
 The correlation between the full SAM and the ENSO as measured by the Oceanic Niño Index [@bamston1997] (ONI) is `r cor_texts[index == "oni"]$full`.
-This relationship is captured entirely the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni"]$asym` with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni"]$sym`).
+Consistent with @fan2007, we show that this relationship is captured entirely the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni"]$asym` with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni"]$sym`).
 We performed the same analysis using the Multivariate Enso Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987] to conclude that these results do not depend on the ENSO index used.
@@
 
-```{r regr-air-season, fig.cap = "Regression pattern of surface temperature with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
-Gray areas have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*"}
+```{r regr-air-season, fig.cap = "Regression pattern of season mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM.
+Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
+Gray areas in Antarctica are areas with have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*"}
 air_regr %>% 
@@
   scale_x_longitude() +
-  scale_y_latitude(labels = NULL, limits = c(NA, 0)) +
+  scale_y_latitude(labels = NULL, limits = c(NA, 0), 
+                   minor_breaks = NULL) +
   scale_fill_divergent("Surface temperature",  
@@
   no_grid +
-  axis_labs_smol
+  axis_labs_smol +
+  theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
 ```
@@
 
-```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
+```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression pattern of (row 1) annual and (rows 2 to 5) season mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM.
+thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
 plot_pp("oceania") 
@@
 
-```{r pp-regr-america, dependson = "def-plot_pp", fig.cap = "Same but for america", fig.width = 6, fig.height = 7.6, fig.env = "figure*"}
+```{r pp-regr-america, dependson = "def-plot_pp", fig.cap = "Same as Figure \\ref{fig:pp-regr-oceania} but for South America.", fig.width = 6, fig.height = 7.6, fig.env = "figure*"}
 plot_pp("america")
@@
 In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship delineated above.
-These are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [@cazes-boezio2003, @fogt2011, @cai2020a].
+These are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [@cazes-boezio2003; @fogt2011; @cai2020a].
 
@@
 
-```{r A1, fig.cap = "Lag-correlation between Symmetric and Asymmetric SAM at each level.", fig.width = 4, fig.height=4, appendix = TRUE}
+```{r A1, fig.cap = "Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
+Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.", fig.width = 4, fig.height=4, appendix = TRUE}
 ccf_dt <- function(..., plot = FALSE) {
@@
   geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
+  geom_contour_tanaka(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
   geom_vline(xintercept = 0) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.05)) +
-  scale_x_continuous("Asym leads Sym\t \t\t Sym leads Asym", breaks = seq(-5, 5, 1), 
+  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(15),
+                       breaks = AnchorBreaks(0, 0.05),
+                       limits = c(NA, .59)) +
+  scale_x_continuous("Lag (months)\nSymmetric leads Asymmetric\t \tAsymmetric leads Symmetric", breaks = seq(-5, 5, 1), 
                      expand = c(0, 0)) +
   scale_y_level()
-```
-
-```{r A2 ccf-levels, fig.cap = "Cross-correlation functions for each index and two differnet base levels.", fig.height=5, fig.width=6}
-ccf_dt <- function(...) {
-  a <- ccf(..., plot = FALSE)
-  list(acf = as.vector(a$acf),
-       lag = as.vector(a$lag))
-}
-
-index2 <- copy(indices)
-index2[, estimate_700 := estimate[lev == 700], by = .(term, time)]
-index2[, estimate_10 := estimate[lev == 10], by = .(term, time)]
-
-rbind("700" = index2[, ccf_dt(x = estimate, y = estimate_700, lag.max = 5), by = .(lev, term)],
-      "10" = index2[, ccf_dt(x = estimate, y = estimate_10, lag.max = 5), by = .(lev, term)], 
-      idcol = "base_level") %>% 
-  ggplot(aes(lag, lev)) +
-  geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.1, 0)) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
-                       breaks = AnchorBreaks(0, 0.1, 0)) +
-  scale_y_level() +
-  scale_x_continuous(breaks = seq(-10, 10)) +
-  facet_grid(term~base_level, labeller = labeller(base_level = function(x) paste0("Base level: ", x, " hPa"), 
-                                                  term = lab_sam)) +
-  tag_facets() 
 ```
@@
   # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
-  .[, fftspectrum(estimate, c(3, 5), B = 5000, probs = c(0.025, 0.975)), by = .(term, lev)] -> spec
-```
-
-```{r A3, fig.cap = "Fourier spectrum of each timeseries.
-The shading indicates de 95\\% area derived by fitting an AR process to each series and bootstrapping 5000 simulated samples.", appendix = TRUE, fig.height=4, fig.width=7}
+  .[, fftspectrum(estimate, c(3, 5), B = 5000, probs = c(0, 0.95)), by = .(term, lev)] -> spec
+```
+
+```{r A3, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with withs 3 and 5.
+The shading indicates de 95\\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\\% of the simulated sampels had an amplitude equal or lower).
+The light line indicates the theoretical expected amplitude from the autorregressive model.", appendix = TRUE, fig.height=4, fig.width=7}
 spec %>%
@@
   ggplot(aes(1/freq/12, spec)) +
-  geom_ribbon(aes(ymin = `2%`, ymax = `98%`), position = "dodge", alpha = 0.1) +
+  geom_ribbon(aes(ymin = 0, ymax = `95%`), position = "dodge", alpha = 0.1) +
   geom_line(aes(y = ar_spectrum), alpha = 0.15) +
@@
   scale_color_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill")) +
-  scale_x_log10("Period (years)", sec.axis = sec_axis(~.*12)) +
+  scale_x_log10("Period (years)", sec.axis = sec_axis(~.*12, name = "Period (months)")) +
+  scale_y_continuous("Spectrum") +
   facet_grid(lev ~ term,labeller = labeller(lev = lev.lab, term = lab_sam), scales = "free") +
@@
 
-```{r A8, appendix = TRUE, fig.cap = "Pattern cross-correlation \\#FIXME!", fig.height=4, fig.width=4}
-vertical_regr %>% 
-  .[term == "asym"] %>% 
-  copy() %>% 
-  .[, id := interaction(lev, lon)] %>% 
-  widyr::pairwise_cor(index_level, id, estimate, diag = TRUE) %>% 
-  as.data.table() %>% 
-  ggplot(aes(item2, item1)) +
-  geom_contour_fill(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
-  geom_contour_tanaka(aes(z = correlation), size = 0.1, na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
-  geom_text_contour(aes(z = correlation), na.fill = .99, breaks = MakeBreaks(binwidth = 0.1), 
-                    color = "white", rotate = FALSE, stroke.color = "black", stroke = 0.2, 
-                    skip = 1, size = 4,
-                    data = ~.x[item2 > item1]) +
-  geom_polygon(data = tri, fill = "white") +
-  scale_fill_divergent("Pattern correlation", 
-                       breaks = MakeBreaks(binwidth = 0.1),
-                       guide = guide_colorstrip_bottom()) +
-  scale_x_level(minor_breaks = logbreaks, position = "top") +
-  scale_y_level(minor_breaks = logbreaks) +
-  coord_equal() +
-  theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside") 
-```
-
-```{r A9, fig.cap = "700 hPa Geopotetnial height zonal anomalies of composites of positive and negative SAM months selected using 1 standard deviation as threshhold.
+```{r A9, fig.cap = "700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\\pm1$ standard deviation as threshhold.
 Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.", apendix = TRUE, fig.height = 5, fig.width = 7}
@@
 
-```{r A10, fig.cap = "Zonal of projection of 700 hPa onto the first EOF of each season.",  apendix = TRUE, fig.width=4, fig.height=4}
+```{r A10, fig.cap = "Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently.",  apendix = TRUE, fig.width=7, fig.height=2.5}
 set.seed(42)
@@
 
-SAM_season[, eof[[1]]$left, by = season] %>% 
-  widyr::pairwise_cor(season, time, hgt)
-
-cors <- SAM_season[, eof[[1]]$left, by = season] %>% 
+patterns <- SAM_season[, eof[[1]]$left, by = season] %>% 
   copy() %>% 
@@
   hgt[., on = "time"] %>% 
-  .[, .(proj = sum(EOF*hgt_a)), by = .(lon, lat, season)] %>% 
-  .[season == "DJF", proj := -proj] %>% 
-  .[, proj_z := Anomaly(proj), by = .(lat, season)] %>% 
+  .[, FitLm(hgt_a, EOF), by = .(lon, lat, season)] %>%
+  rm_intercept() %>% 
+  .[season == "DJF", estimate := -estimate] %>% 
+  .[, estimate_a := Anomaly(estimate), by = .(lat, season)]
+
+cors <- patterns %>% 
   .[ , f := interaction(lon, lat)] %>% 
-  widyr::pairwise_cor(season, f, proj_z, upper = FALSE)
-
-SAM_season[, eof[[1]]$left, by = season] %>% 
-  copy() %>% 
-  setnames("hgt", "EOF") %>% 
-  .[, EOF := EOF/sd(EOF), by = .(season)] %>% 
-  hgt[., on = "time"] %>% 
-  .[, .(proj = sum(EOF*hgt_a)), by = .(lon, lat, season)] %>% 
-  .[season == "DJF", proj := -proj] %>% 
-  .[, proj_z := Anomaly(proj), by = .(lat, season)] %>% 
+  widyr::pairwise_cor(season, f, estimate_a, upper = FALSE)
+
+patterns %>% 
   periodic(lon = c(0, 360)) %>% 
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = proj_z), breaks = AnchorBreaks(0,exclude = 0), global.breaks = FALSE) +
-  geom_contour_tanaka(aes(z = proj_z), breaks = AnchorBreaks(0,exclude = 0), global.breaks = FALSE) +
+  geom_contour_fill(aes(z = estimate_a), breaks = AnchorBreaks(0, exclude = 0), global.breaks = FALSE) +
+  geom_contour_tanaka(aes(z = estimate_a), breaks = AnchorBreaks(0, exclude = 0), global.breaks = FALSE) +
   geom_qmap(~.x[lat <= 0]) +
@@
   scale_y_latitude(limits = c(NA, -20), labels = NULL) +
-  scale_fill_divergent(NULL, guide = "none") +
+  scale_fill_divergent(NULL, 
+                       guide = guide_colorstrip_bottom(),
+                       breaks = AnchorBreaks(0, 10, exclude = 0)) +
   coord_polar() +
@@
   no_grid +
-  facet_wrap(.~season)
-```
-
-```{r A11, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height computed for the periods 1979 to 1998 and 1999 to 2018.
-Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right).
-Pattern correlation between both fields is 0.77.
-Arbitrary units.", fig.width=6, fig.height=5, apendix = TRUE}
-
-prepost_SAM <- hgt[lat <= -20] %>% 
+  facet_grid(.~season)
+```
+
+```{r A11, fig.cap = "Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 to 1998 and 1999 to 2018.
+Pattern correlation between both fields is 0.76.", fig.width=6, fig.height=5, apendix = TRUE}
+SAM_prepost <- hgt[lat <= -20] %>% 
   .[lev == 700] %>%
@@
   .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
-    by = .(lev, period)]
-
-sam_sep <- prepost_SAM[, eof[[1]]$right, by = .(lev, period)] %>% 
+    by = .(period)]
+
+sam_sep <- SAM_prepost[, eof[[1]]$right, by = .(period)] %>% 
   setnames("hgt", "full") %>% 
-  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, lev, period)]
-
-sam_plot <- sam_sep %>% 
-  .[lev %in% c(700)] %>%
+  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, period)]
+
+patterns <- SAM_prepost[, eof[[1]]$left, by = .(period)] %>% 
+  copy() %>% 
+  setnames("hgt", "EOF") %>% 
+  .[, EOF := EOF/sd(EOF), by = .(period)] %>% 
   rm_singleton() %>% 
-  melt(id.vars = c("lon", "lat", "period")) %>% 
-  .[variable == "asym"] %>% 
-  .[, variable := factor(variable, levels = names(lab_sam), ordered = TRUE)] %>% 
-  .[, value := value/sd(value), by = .(period)]
-
-cor <- sam_plot %>% 
-  dcast(lon + lat ~ period, value.var = "value") %>% 
+  hgt[., on = .NATURAL] %>% 
+  .[lev == 700] %>% 
+  .[, FitLm(hgt_a, EOF), by = .(lon, lat, period)] %>%
+  rm_intercept() %>% 
+  # .[season == "DJF", estimate := -estimate] %>% 
+  .[, estimate_a := Anomaly(estimate), by = .(lat, period)]
+
+cor <- patterns %>% 
+  dcast(lon + lat ~ period, value.var = "estimate_a") %>% 
   .[, cor(`1979-1998`, `1999-2018`)]
 
-sam_plot %>% 
+patterns %>% 
   periodic(lon = c(0, 360)) %>%
   ggplot(aes(lon, lat)) +
-  # geom_contour(aes(z = value, linetype = factor(-sign(..level..))), global.breaks = FALSE) +
-  geom_contour_fill(aes(z = value), breaks = ZeroBreaks) +
-  geom_contour_tanaka(aes(z = value), breaks = ZeroBreaks) +
+  geom_contour_fill(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
+  geom_contour_tanaka(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
   geom_qmap() +
@@
   scale_y_latitude(limits = c(NA, -20), labels = NULL) +
-  scale_fill_divergent(guide = "none") +
+  scale_fill_divergent(NULL, 
+                       guide = guide_colorstrip_bottom(),
+                       breaks = AnchorBreaks(0, 5, exclude = 0)) +
   coord_polar() +

### 2020-09-23 analysis/paper/paper.tex
casi casi

@@
 
-\title{Title here}
+\title{New method to describe the zonal symmetries and asymmetries of the
+Southern Annular Mode}
 
@@
 
-\affiliation{CIMA UBA blablabla}
+\affiliation{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales,
+Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires,
+Argentina CONICET -- Universidad de Buenos Aires, Centro de
+Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el
+Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires,
+Argentina}
 
@@
 
-yada yada SAM yada yada circulation.. yada yada so important. yada yada
-many impacts.
+The Southern Annular Mode (SAM) is the main mode of variability in the
+Southern Hemisphere extratropical circulation \citep{rogers1982} in
+daily, monthly, and decadal timescales {[}\citet{baldwin2001a};
+fogt2006{]} and exerts an important influence in weather conditions such
+as temperature and precipitation anomalies and sea ice concentration
+\citep{fogt2020}. Its positive phase is traditionally described as
+anomalously low pressures over Antarctica surrounded by a ring of
+anomalous high pressures in middle-to-high latitudes.
 
-\citet{fogt2012} studied the characteristics of the asymmetric structure
-of the SAM. It computed the zonally anomalous component of mean sea
-level pressure (MSLP) composites for positive and negative SAM events
-and created two indices by projecting MSLP fields ontno them. However,
-the use of composites leads to some issues that affects the
-interpretability of the results. First, they can be dependent on the
-choice of threshold used to define positive and negative events.
-Secondly, by discarting data that don't meet the threshold, they don't
-use all the information avaiable. Due to the realtive short timeframe
-used, this leads to some composites being composed of as little as 4
-years. Third, the resulting composites corresponding to each polarity
-and season are derived from the average of different amount of fields
-and from different years. This last issue is particularly important in
-light of the changing structure of the SAM before and after 1980
-\citep{silvestri2009}. In \citet{fogt2012}, the DJF SAM+ composite uses
-only 7 years, 5 of which are later than 1988, whereas all of the 8 years
-used for their DJF SAM- composite are from earlier than 1988.
+However, computed as the leading Empirical Orthogonal Function (EOF) of
+Sea Level Pressure or low-level geopotential height, the SAM spatial
+structure contains noticeable deviations from this zonally symmetric
+description, particularly in the Pacific Ocean region. These zonal
+asymmetries are not widely studied, but previous work suggest that they
+strongly modulate the regional impacts of the SAM, going as far as
+reversing its relationship between precipitation in South America
+\citep{silvestri2009}. At the very least, the fact that the SAM is not
+entirely zonally symmetric hinders our ability to reconstruct its
+historical variability prior to the availabilty of dense observations in
+the Southern Hemisphere \citep{jones2009}.
+
+We are not aware of any previous work which quantifies the temporal
+variability of the asymmetric component of the SAM with the exception of
+\citet{fogt2012}. However, their methods based on composites of positive
+and negative SAM events leads to some issues, such as spatial patterns
+derived from as little as 4 cases and from imbalanced periods (for
+example, 5 of the 7 cases in their DJF SAM+ composite are from later
+than 1988, whereas all of the 8 years in their DJF SAM- composite are
+from earlier than 1988). This is particularly important due to the
+inhomogeneities in renalysis products prior to the sattelite era and the
+possible change in the asymmetric structure of the SAM
+\citep{silvestri2009}.
+
+Our objective is, then, to systematically characterise the zonally
+asymmetric component of the SAM variability by constructing two indices
+which aim to capture exclusively the variability of the symmetric and
+asymmetric component each. We then analyse their temporal variability,
+trends and vertical coherence. We study the spatial patterns described
+by the variability exclusive to each index. Finally, we investigate
+theri relationship with temperature and recipitation anomalies.
 
@@
 symmetric (\([\mathrm{EOF_1}](\lambda, \phi)\)) components. We then
-compute the ``Full'', ``Asymmetric'' and ``Symmetric'' indices, by
-regressing each geopotential field on the respective patterns (weighting
-by the cosine of latitude). The three indices are normalised by dividing
-them by the standard deviation of the ``Full'' index at each level. As a
-result, the magnitude between indices is comparable. However, only
-``Full'' index will have unit standard deviation per definition.
+compute the ``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM''
+indices as the regression coefficients of the regression ofeach monthly
+geopotential field on the respective patterns (weighting by the cosine
+of latitude). The three indices are normalised by dividing them by the
+standard deviation of the ``Full'' index at each level. As a result, the
+magnitude between indices is comparable. However, only ``Full'' index
+will have unit standard deviation per definition. From the regression,
+we also use the explained variance of each pattern as a indication of
+the degere of symmetry or asymmetry of each monthly field.
 
@@
 \includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height]{Spatial patterns of the first EOF of 700 hPa geopotential height.
-Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right).
+(a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
 Arbitrary units.}\label{fig:method}
@@
 \begin{figure*}
-\includegraphics{asymsam-timeseries-1} \caption[Time series for the asymmetric SAM and symmetric SAM and density estimates]{Time series for the asymmetric SAM and symmetric SAM and density estimates.}\label{fig:asymsam-timeseries}
+\includegraphics{asymsam-timeseries-1} \caption[Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa]{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
+To the left, probability density estimate of each index.
+Series are standarised by the standard deviation of the Full SAM at each level.}\label{fig:asymsam-timeseries}
 \end{figure*}
@@
 correlations increase with height. As a consequence, statospheric
-Asymmetric index tend to precede corresponding Symmetric index.
+Symmetric index tend to precede corresponding Asymmetric index.
 
 \begin{figure}
-\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric and Asymmetric SAM at each level for lag zero and lag -1 (Asymmetric leads Symmetric)]{Correlation between the Symmetric and Asymmetric SAM at each level for lag zero and lag -1 (Asymmetric leads Symmetric).}\label{fig:cor-lev}
+\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Symmetric leads Asymmetric)]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Symmetric leads Asymmetric).}\label{fig:cor-lev}
 \end{figure}
@@
 \begin{figure*}
-\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the Full, Asymmetric and Symmetric SAM]{Cross correlation between levels of the Full, Asymmetric and Symmetric SAM.}\label{fig:cross-correlation}
+\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM.}\label{fig:cross-correlation}
 \end{figure*}
@@
 \begin{figure*}
-\includegraphics{trends-1} \caption[Decadal normalised trends for each index at each level for annual (row a) and seasonal values (rows b-e) for the period 1979-2018]{Decadal normalised trends for each index at each level for annual (row a) and seasonal values (rows b-e) for the period 1979-2018.
+\includegraphics{trends-1} \caption[Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
 Shading indicates the 95\% confidence interval.}\label{fig:trends}
@@
 \begin{figure}
-\includegraphics{r-squared-trend-1} \caption[Decadal trends for explained variance of each index at each level for the period 1979-2018]{Decadal trends for explained variance of each index at each level for the period 1979-2018.
+\includegraphics{r-squared-trend-1} \caption[Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979-2018]{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979-2018.
 Shading indicates the 95\% confidence interval.}\label{fig:r-squared-trend}
@@
 \begin{figure*}
-\includegraphics{2d-regr-1} \caption[Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM]{Regression patterns of geopotential height at 30, 300 and 700 hPa with the Full, Asymmetric and Symmetric SAM.
+\includegraphics{2d-regr-1} \caption[Regression patterns of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM]{Regression patterns of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM.
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.
-Points marked on panel b.2 are the location of the reference points used by \cite{raphael2004} for their Zonal Wave 3 index.}\label{fig:2d-regr}
+Points marked on panel b.2 are the location of the reference points used by \cite{raphael2004} for their Zonal Wave 3 index.
+}\label{fig:2d-regr}
 \end{figure*}
@@
 \begin{figure}
-\includegraphics{wave-amplitude-1} \caption[Planteray wave amplitude for the regression patterns at 50 and 700 hPa]{Planteray wave amplitude for the regression patterns at 50 and 700 hPa.
-Note the varying x axis.}\label{fig:wave-amplitude}
+\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.
+Note the differnet x axis.}\label{fig:wave-amplitude}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{vertical-regression-1} \caption[Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South]{Asymmetric coefficient of the multiple regression of mean monthly geopotential height anomalies between 65 and 40 South.
-(\#FIXME this caption needs some love)}\label{fig:vertical-regression}
+\includegraphics{vertical-regression-1} \caption[Regression between monthly geopotential anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM)]{Regression between monthly geopotential anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
+(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa.}\label{fig:vertical-regression}
 \end{figure}
@@
 between the full SAM and the ENSO as measured by the Oceanic Niño Index
-\citep{bamston1997} (ONI) is -0.16. This relationship is captured
-entirely the Asymmetric SAM, as this index has a partial correlation of
--0.26 with the ONI controlling for the effect of the Symmetric SAM,
-whereas the Symmetric SAM's partial correlation with the ONI is
-essentially null (0.019). We performed the same analysis using the
-Multivariate Enso Index \citep{wolter2011} and the Southern Oscillation
-Index \citep{ropelewski1987} to conclude that these results do not
-depend on the ENSO index used.
+\citep{bamston1997} (ONI) is -0.16. Consistent with \citet{fan2007}, we
+show that this relationship is captured entirely the Asymmetric SAM, as
+this index has a partial correlation of -0.26 with the ONI controlling
+for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial
+correlation with the ONI is essentially null (0.019). We performed the
+same analysis using the Multivariate Enso Index \citep{wolter2011} and
+the Southern Oscillation Index \citep{ropelewski1987} to conclude that
+these results do not depend on the ENSO index used.
 
@@
 \begin{figure*}
-\includegraphics{regr-air-season-1} \caption[Regression pattern of surface temperature with Asymmetric and Symmetric SAM]{Regression pattern of surface temperature with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
-Gray areas have more than 15\% of missing data.}\label{fig:regr-air-season}
+\includegraphics{regr-air-season-1} \caption[Regression pattern of season mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM]{Regression pattern of season mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM.
+Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
+Gray areas in Antarctica are areas with have more than 15\% of missing data.}\label{fig:regr-air-season}
 \end{figure*}
@@
 \begin{figure*}
-\includegraphics{pp-regr-oceania-1} \caption[Regression pattern of precipiation with Asymmetric and Symmetric SAM]{Regression pattern of precipiation with Asymmetric and Symmetric SAM.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.}\label{fig:pp-regr-oceania}
+\includegraphics{pp-regr-oceania-1} \caption[Regression pattern of (row 1) annual and (rows 2 to 5) season mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM]{Regression pattern of (row 1) annual and (rows 2 to 5) season mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM.
+thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
 \end{figure*}
@@
 \begin{figure*}
-\includegraphics{pp-regr-america-1} \caption[Same but for america]{Same but for america}\label{fig:pp-regr-america}
+\includegraphics{pp-regr-america-1} \caption{Same as Figure \ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
 \end{figure*}
@@
 clear in the DJF and SON trimesters, which include the months in which
-the ENSO teleconnection is more active \citep[\citet{fogt2011},
-\citet{cai2020a}]{cazes-boezio2003}.
+the ENSO teleconnection is more active
+\citep{cazes-boezio2003, fogt2011, cai2020a}.
 
@@
 \begin{figure}
-\includegraphics{A1-1} \appendcaption{A1}{Lag-correlation between Symmetric and Asymmetric SAM at each level.}\label{fig:A1}
+\includegraphics{A1-1} \appendcaption{A1}{Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
+Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.}\label{fig:A1}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A2 ccf-levels-1} \caption[Cross-correlation functions for each index and two differnet base levels]{Cross-correlation functions for each index and two differnet base levels.}\label{fig:A2 ccf-levels}
+\includegraphics{A3-1} \appendcaption{A3}{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with withs 3 and 5.
+The shading indicates de 95\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\% of the simulated sampels had an amplitude equal or lower).
+The light line indicates the theoretical expected amplitude from the autorregressive model.}\label{fig:A3}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A3-1} \appendcaption{A3}{Fourier spectrum of each timeseries.
-The shading indicates de 95\% area derived by fitting an AR process to each series and bootstrapping 5000 simulated samples.}\label{fig:A3}
+\includegraphics{A9-1} \caption[700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold]{700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold.
+Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A9}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A8-1} \appendcaption{A8}{Pattern cross-correlation \#FIXME!}\label{fig:A8}
+\includegraphics{A10-1} \caption[Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently]{Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently.}\label{fig:A10}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A9-1} \caption[700 hPa Geopotetnial height zonal anomalies of composites of positive and negative SAM months selected using 1 standard deviation as threshhold]{700 hPa Geopotetnial height zonal anomalies of composites of positive and negative SAM months selected using 1 standard deviation as threshhold.
-Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A9}
-\end{figure}
-
-\begin{verbatim}
-#> # A tibble: 12 x 3
-#>    item1 item2 correlation
-#>    <fct> <fct>       <dbl>
-#>  1 MAM   DJF     -1.16e-33
-#>  2 JJA   DJF      3.07e-34
-#>  3 SON   DJF      2.55e-33
-#>  4 DJF   MAM     -1.16e-33
-#>  5 JJA   MAM      1.84e-33
-#>  6 SON   MAM      1.53e-32
-#>  7 DJF   JJA      3.07e-34
-#>  8 MAM   JJA      1.84e-33
-#>  9 SON   JJA     -4.03e-33
-#> 10 DJF   SON      2.55e-33
-#> 11 MAM   SON      1.53e-32
-#> 12 JJA   SON     -4.03e-33
-\end{verbatim}
-
-\begin{figure}
-\includegraphics{A10-1} \caption[Zonal of projection of 700 hPa onto the first EOF of each season]{Zonal of projection of 700 hPa onto the first EOF of each season.}\label{fig:A10}
-\end{figure}
-
-\begin{figure}
-\includegraphics{A11-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height computed for the periods 1979 to 1998 and 1999 to 2018]{Spatial patterns of the first EOF of 700 hPa geopotential height computed for the periods 1979 to 1998 and 1999 to 2018.
-Full field (left), zonally asymmetric component (middle) and zonally symmetric component (right).
-Pattern correlation between both fields is 0.77.
-Arbitrary units.}\label{fig:A11}
+\includegraphics{A11-1} \caption[Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 to 1998 and 1999 to 2018]{Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 to 1998 and 1999 to 2018.
+Pattern correlation between both fields is 0.76.}\label{fig:A11}
 \end{figure}

### 2020-09-29 analysis/paper/paper.Rmd
Intro: Comentarios Lean

@@
 twocol: FALSE  #  Use FALSE for submission. 
-title: New method to describe the zonal symmetries and asymmetries of the Southern Annular Mode
+title: Asssesment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
 author1: 
@@
 
-The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a; fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [@fogt2020].
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a; @fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [@fogt2020].
 Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
 
+#LEAN: Antes de empezar este parráfo, me parece que deberías describir las maneras en que se suele calcular el SAM en la literatura (algo así creo que habías incluido para la AGU).
+
 However, computed as the leading Empirical Orthogonal Function (EOF) of Sea Level Pressure or low-level geopotential height, the SAM spatial structure contains noticeable deviations from this zonally symmetric description, particularly in the Pacific Ocean region.
-These zonal asymmetries are not widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009].
+These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009].
 At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availabilty of dense observations in the Southern Hemisphere [@jones2009].
@@
 
-Our objective is, then, to systematically characterise the zonally asymmetric component of the SAM variability by constructing two indices which aim to capture exclusively the variability of the symmetric and asymmetric component each.
-We then analyse their temporal variability, trends and vertical coherence.
-We study the spatial patterns described by the variability exclusive to each index.
-Finally, we investigate theri relationship with temperature and recipitation anomalies.
+#LEAN: También me parece que faltaría agregar dos parráfos más, uno relacionado con las tendencias del SAM y otro relacionado con los impactos conocidos del SAM.
+
+#LEAN: Otra cosa que no veo y que me parece que sería importante decir es sobre la estructura vertical del SAM (sería bueno también agregar la referencia de Baldwin que es la que inspira un poco el estudio).
+
+Our objective is, then, to systematically characterise the zonally asymmetric and symmetric components of the SAM variability by constructing two indices which aim to capture exclusively the variability of the symmetric and asymmetric component each.
+These indices are constructed for each vertical level, so that vertical coherence could be assessed.
+We analyse their temporal variability and trends.
+
+#acá puede ser confuso porque cuando hablas de "cada indice" a veces te referis a los indices de todos los niveles verticales, en otro de solo uno que describa la estratósfera y otro la tropósfera y en otro solo al de la troposfera.
+
+We then study the spatial patterns described by the variability exclusive to each index. Finally, we investigate their relationship with temperature and precipitation anomalies.
+
+#LEAN: Faltaría acá después describir las secciones del trabajo
 

### 2020-09-29 analysis/paper/paper.Rmd
Arregla comentarios introducción de Leandro

@@
 
-The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a; @fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [@fogt2020].
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a;@fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [@fogt2020].
 Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
 
-#LEAN: Antes de empezar este parráfo, me parece que deberías describir las maneras en que se suele calcular el SAM en la literatura (algo así creo que habías incluido para la AGU).
-
-However, computed as the leading Empirical Orthogonal Function (EOF) of Sea Level Pressure or low-level geopotential height, the SAM spatial structure contains noticeable deviations from this zonally symmetric description, particularly in the Pacific Ocean region.
+Most authors describe the SAM as a zonally symmetric pattern, a fact that is reflected not only in its name, but also in the various methods used to characterise it.
+Of the several different indices presented in the literature, many of them are based on zonal means of sea level pressure or geopotential height [@ho2012].
+@gong1999 defined the SAM index as the zonal mean sea level pressure differene between 40\degree S and 60\degree S, which is also the definition used by the station-based index in @marshall2003.
+@baldwin2009 proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
+
+Even though these indices are based on zonal averages, the spatial the spatial structure of the SAM computed from them contains noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
 These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009].
@@
 
+At least some of the variabilty associated with the zonal asymmetries of the SAM is probably forced by the tropics.
+In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987; @kidson1988;@karoly1989], whose wave train proyects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
+And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed [@fan2007; @fogt2011; @clem2013].
+
+Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn [@fogt2020 and references therein].
+These trends are thought of driven primarily by stratospheric ozone depletion and understood in the context of zonal mean variables.
+However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variabilty could be masking influencing the observed trends independendly.
+
+Similarly unclear are the specific impacts of the zonally asymmetric component of the SAM.
+Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes [@jones2019] (and viceversa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012].
+The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
+The importance of zonal asymmetries of the SAM in these impacts have been studied in certain regions.
+For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymetric circulation associated with the SAM [@silvestry2009; @rosso2018].
+@fan2007 also found that precipitation in East Asia was impacted by the variabilty of only the Western Hemisphere part of the SAM.
+
 We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of @fogt2012.
 However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988).
-This is particularly important due to the inhomogeneities in renalysis products prior to the sattelite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].
-
-#LEAN: También me parece que faltaría agregar dos parráfos más, uno relacionado con las tendencias del SAM y otro relacionado con los impactos conocidos del SAM.
-
-#LEAN: Otra cosa que no veo y que me parece que sería importante decir es sobre la estructura vertical del SAM (sería bueno también agregar la referencia de Baldwin que es la que inspira un poco el estudio).
-
-Our objective is, then, to systematically characterise the zonally asymmetric and symmetric components of the SAM variability by constructing two indices which aim to capture exclusively the variability of the symmetric and asymmetric component each.
-These indices are constructed for each vertical level, so that vertical coherence could be assessed.
-We analyse their temporal variability and trends.
-
-#acá puede ser confuso porque cuando hablas de "cada indice" a veces te referis a los indices de todos los niveles verticales, en otro de solo uno que describa la estratósfera y otro la tropósfera y en otro solo al de la troposfera.
-
-We then study the spatial patterns described by the variability exclusive to each index. Finally, we investigate their relationship with temperature and precipitation anomalies.
-
-#LEAN: Faltaría acá después describir las secciones del trabajo
+This is particularly important due to the inhomogeneities in renalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].
+
+Moreover, @fogt2012 studied the zonal asymmetric component of the SAM only in sea level pressure.
+Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere [@baldwin2009].
+
+Our objective is, then, to systematically characterise the zonally asymmetric and symmetric components of the SAM variability.
+For each level, we construct two indices which aim to capture exclusively the variability of the symmetric and asymmetric component respectively.
+We assess their vertical structure and coherence, temporal variability and trends.
+We then study the spatial patterns described by the variability exclusive to each index.
+Finally, we investigate their relationship with temperature and precipitation anomalies.
+
+In the Section \ref{sec:methods} we describe the methods.
+In Section \ref{sec:results}.\ref{sec:temporal} we describe the temporal variability and vertical coherence of the inidices.
+In  Section \ref{sec:results}.\ref{sec:spatial}, we analyse the spatial patterns of geopotential height associated with them.
+In Section \ref{sec:results}.\ref{sec:impacts}, we study their relationship with surface-level tempreature and precipitation.
 
@@
 # Methods 
+\label{sec:methods}
 
@@
 
-To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
+To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
 We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.
@@
 We describe the relationship between the SAM indices and temperature and precipitation.
-We use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree lagitude by 5\degree longitude.
-For precipitation, we use monthly, 0.5\degree latitude by 0.5\degree longitude data from the Global Precipitation Climatology Centre [@schneider2015].
+We use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude For precipitation, we use monthly, 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015].
 
@@
 At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions.
-In such case, using naive p-values to test for significance leads to misleading results [@walker1914;@katz1991].
+In such case, using naive p-values to test for significance leads to misleading results [@walker1914; @katz1991].
 While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjunting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue.
@@
 # Results
+\label{sec:results}
 
 ## Temporal evolution
+\label{sec:temporal}
 
@@
 Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
-annual_trend <- indices %>%
+
+indices %>% 
   copy() %>% 
+  .[, season := season(time)] %>% 
+  rbind(., copy(.)[, season := "Annual"]) %>%
     .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
-  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
+  .[, FitLm(estimate_norm, time, se = TRUE), by = .(type = term, lev, season)] %>% 
   rm_intercept() %>% 
   .[, term := NULL]  %>%
-  .[, season := "Annual"]
-
-indices %>% 
-  copy() %>% 
-  # .[time > as.Date("1990-01-01")] %>%
-  .[, season := season(time)] %>% 
-  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
-  .[, FitLm(estimate_norm, time = time, se = TRUE), by = .(type = term, lev, season)] %>% 
-  rm_intercept() %>% 
-  .[, term := NULL] %>% 
-  rbind(annual_trend) %>% 
   .[, estimate := estimate*3600*24*365*10] %>%
   .[, std.error := std.error*3600*24*365*10] %>%
+  # .[, estimate := estimate/(3600*24*365*10)] %>%
+  # .[, std.error := std.error/(3600*24*365*10)] %>%
   .[, pval := Pvaluate(estimate, std.error, df, "fdr")] %>% 
@@
 ## Spatial patterns
+\label{sec:spatial}
 
@@
 ## Impacts
+\label{sec:impacts}
 
@@
 On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
-This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region [@silvestri2009, @zamboni2010].
+This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region [@silvestri2009; @zamboni2010].
 This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events [@rosso2018].
@@
 There is evidence of influence between the QBO and the Northern Annular Mode [e.g.
-@holton1980, @watson2014, @zhang2020], so it's not unlikely that the SAM would be similarly connected.
+@holton1980; @watson2014; @zhang2020], so it's not unlikely that the SAM would be similarly connected.
 However establishing this link would require further research.

### 2020-09-29 analysis/paper/paper.tex
Arregla comentarios introducción de Leandro

@@
 
-\title{New method to describe the zonal symmetries and asymmetries of the
-Southern Annular Mode}
+\title{Asssesment of zonal symmetric and asymmetric components of the Southern
+Annular Mode using a novel approach}
 
@@
 Southern Hemisphere extratropical circulation \citep{rogers1982} in
-daily, monthly, and decadal timescales {[}\citet{baldwin2001a};
-fogt2006{]} and exerts an important influence in weather conditions such
-as temperature and precipitation anomalies and sea ice concentration
+daily, monthly, and decadal timescales \citep{baldwin2001a, fogt2006}
+and exerts an important influence in weather conditions such as
+temperature and precipitation anomalies and sea ice concentration
 \citep{fogt2020}. Its positive phase is traditionally described as
@@
 
-However, computed as the leading Empirical Orthogonal Function (EOF) of
-Sea Level Pressure or low-level geopotential height, the SAM spatial
-structure contains noticeable deviations from this zonally symmetric
-description, particularly in the Pacific Ocean region. These zonal
-asymmetries are not widely studied, but previous work suggest that they
-strongly modulate the regional impacts of the SAM, going as far as
-reversing its relationship between precipitation in South America
-\citep{silvestri2009}. At the very least, the fact that the SAM is not
-entirely zonally symmetric hinders our ability to reconstruct its
-historical variability prior to the availabilty of dense observations in
-the Southern Hemisphere \citep{jones2009}.
+Most authors describe the SAM as a zonally symmetric pattern, a fact
+that is reflected not only in its name, but also in the various methods
+used to characterise it. Of the several different indices presented in
+the literature, many of them are based on zonal means of sea level
+pressure or geopotential height \citep{ho2012}. \citet{gong1999} defined
+the SAM index as the zonal mean sea level pressure differene between
+40\degree S and 60\degree S, which is also the definition used by the
+station-based index in \citet{marshall2003}. \citet{baldwin2009}
+proposed defining the Northern and Annular modes as the leading EOF of
+the zonally averaged geopotential height at each level.
+
+Even though these indices are based on zonal averages, the spatial the
+spatial structure of the SAM computed from them contains noticeable
+deviations from zonal symmetry, particularly in the Pacific Ocean
+region. These zonal asymmetries have not been widely studied, but
+previous work suggest that they strongly modulate the regional impacts
+of the SAM, going as far as reversing its relationship between
+precipitation in South America \citep{silvestri2009}. At the very least,
+the fact that the SAM is not entirely zonally symmetric hinders our
+ability to reconstruct its historical variability prior to the
+availabilty of dense observations in the Southern Hemisphere
+\citep{jones2009}.
+
+At least some of the variabilty associated with the zonal asymmetries of
+the SAM is probably forced by the tropics. In particular, ENSO-like
+variability affects the Southern Hemisphere extratopics through the
+Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989},
+whose wave train proyects strongly onto the zonal anomalies
+corresponding to the SAM in the Pacific sector. And although the
+relationship between ENSO and SAM is far from simple, tropical
+influences on the SAM have been observed
+\citep{fan2007, fogt2011, clem2013}.
+
+Positive trends in SAM index have been documented by various researchers
+using different indices mostly on boreal Summer and Autumn \citep[ and
+references therein]{fogt2020}. These trends are thought of driven
+primarily by stratospheric ozone depletion and understood in the context
+of zonal mean variables. However, it's not clear how or if the
+asymmetric component responds to this forcing, or whether its variabilty
+could be masking influencing the observed trends independendly.
+
+Similarly unclear are the specific impacts of the zonally asymmetric
+component of the SAM. Positive phase of the SAM is associated with
+generally colder temperatures over Antarctica and warmer temperatures at
+higher latitudes \citep{jones2019} (and viceversa for negative SAM), but
+there are significant deviations from this zonal mean response, notably
+in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}. The
+SAM signal in precipitation behaves similarly, although with even
+greater deviation from zonal symmetry \citep{lim2016}. The importance of
+zonal asymmetries of the SAM in these impacts have been studied in
+certain regions. For example, the SAM-precipitation relationship in
+Southeastern South America and Southern Brazil can be explained by the
+PSA-like zonally asymetric circulation associated with the SAM
+\citep{silvestry2009, rosso2018}. \citet{fan2007} also found that
+precipitation in East Asia was impacted by the variabilty of only the
+Western Hemisphere part of the SAM.
 
@@
 from earlier than 1988). This is particularly important due to the
-inhomogeneities in renalysis products prior to the sattelite era and the
+inhomogeneities in renalysis products prior to the satellite era and the
 possible change in the asymmetric structure of the SAM
@@
 
+Moreover, \citet{fogt2012} studied the zonal asymmetric component of the
+SAM only in sea level pressure. Zonal asymmetries in the SAM spatial
+pattern are fairly barotropic throughout the troposphere, but they
+change dramatically in the stratosphere \citep{baldwin2009}.
+
 Our objective is, then, to systematically characterise the zonally
-asymmetric component of the SAM variability by constructing two indices
-which aim to capture exclusively the variability of the symmetric and
-asymmetric component each. We then analyse their temporal variability,
-trends and vertical coherence. We study the spatial patterns described
-by the variability exclusive to each index. Finally, we investigate
-theri relationship with temperature and recipitation anomalies.
+asymmetric and symmetric components of the SAM variability. For each
+level, we construct two indices which aim to capture exclusively the
+variability of the symmetric and asymmetric component respectively. We
+assess their vertical structure and coherence, temporal variability and
+trends. We then study the spatial patterns described by the variability
+exclusive to each index. Finally, we investigate their relationship with
+temperature and precipitation anomalies.
+
+In the Section \ref{sec:methods} we describe the methods. In Section
+\ref{sec:results}.\ref{sec:temporal} we describe the temporal
+variability and vertical coherence of the inidices. In Section
+\ref{sec:results}.\ref{sec:spatial}, we analyse the spatial patterns of
+geopotential height associated with them. In Section
+\ref{sec:results}.\ref{sec:impacts}, we study their relationship with
+surface-level tempreature and precipitation.
 
 \section{Methods}
+
+\label{sec:methods}
 
@@
 To describe the Southern Annular Mode and its variability we used
-monthly geopotential height at 2.5\degree longitude by
-2.5\degree latitude of horizontal resolution and 37 vertical isobaric
+monthly geopotential height at 2.5\degree~longitude by
+2.5\degree~latitude of horizontal resolution and 37 vertical isobaric
 levels from ERA5 \citep{hersbach2020} for the period 1979 to 2018. We
@@
 which blends land and ocean temperature analysis into a monthly global
-grid 5\degree lagitude by 5\degree longitude. For precipitation, we use
-monthly, 0.5\degree latitude by 0.5\degree longitude data from the
+grid 5\degree~ longitude by 5\degree~ latitude For precipitation, we use
+monthly, 0.5\degree~ longitude by 0.5\degree~ latitude data from the
 Global Precipitation Climatology Centre \citep{schneider2015}.
@@
 
+\label{sec:results}
+
 \subsection{Temporal evolution}
+
+\label{sec:temporal}
 
@@
 
+\label{sec:spatial}
+
 \begin{figure*}
@@
 \subsection{Impacts}
+
+\label{sec:impacts}
 
@@
 Southeastern South America, which inhibits the flow of the Low Level Jet
-to the region \citep[\citet{zamboni2010}]{silvestri2009}. This same
-pattern was found to be associated with increased precipitation in
-Southern Brazil during South Atlantic Convergence Zone events
-\citep{rosso2018}.
+to the region \citep{silvestri2009, zamboni2010}. This same pattern was
+found to be associated with increased precipitation in Southern Brazil
+during South Atlantic Convergence Zone events \citep{rosso2018}.
 
@@
 There is evidence of influence between the QBO and the Northern Annular
-Mode \citep[e.g.][\citet{watson2014}, \citet{zhang2020}]{holton1980}, so
-it's not unlikely that the SAM would be similarly connected. However
-establishing this link would require further research.
+Mode \citep[e.g.][]{holton1980, watson2014, zhang2020}, so it's not
+unlikely that the SAM would be similarly connected. However establishing
+this link would require further research.
 

### 2020-09-30 analysis/paper/paper.Rmd
Agrega comentarios de Leandro

@@
 author2: 
-  name: Leandro Díaz
+  name: Leandro B. Díaz
 affiliation: |
@@
 # Methods
-\label{sec:methods}
 
@@
 
-To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
+To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
 We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.
 
-We describe the relationship between the SAM indices and temperature and precipitation.
-We use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude For precipitation, we use monthly, 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015].
+To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree lagitude by 5\degree longitude, and monthly rainfall 0.5\degree latitude by 0.5\degree longitude data from the Global Precipitation Climatology Centre [@schneider2015].
 
@@
 
-Traditionally the Souther Annular Mode (SAM) is defined as de leading Empiral Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels [@ho2012].
+Traditionally the Souther Annular Mode (SAM) is defined as de leading Empirical Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels [@ho2012].
 Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the the leading EOF of the monthly anomalies of geopotential field south of 20\degree S at each level.
@@
 
-To separate between the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure\ \ref{fig:method} for 700hPa.
+To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure\ \ref{fig:method} for 700hPa.
 The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
-We then compute the "Full SAM", "Asymmetric SAM" and "Symmetric SAM" indices as the regression coefficients of the regression ofeach monthly geopotential field on the respective patterns (weighting by the cosine of latitude).
-The three indices are normalised by dividing them by the standard deviation of the "Full" index at each level.
-As a result, the magnitude between indices is comparable.
+We then compute the "Full SAM", "Asymmetric SAM" and "Symmetric SAM" indices as the regression coefficients of the regression of each monthly geopotential field on the respective patterns (weighting by the cosine of latitude).
+The three indices are normalized by dividing them by the standard deviation of the "Full" index at each level.
+As a result, the magnitudes between indices are comparable.
 However, only "Full" index will have unit standard deviation per definition.
-From the regression, we also use the explained variance of each pattern as a indication of the degere of symmetry or asymmetry of each monthly field.
-
-```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height.
+From the regression, we also use the explained variance of each pattern as an indicator of the degere of zonally symmetry or asymmetry of each monthly field.
+
+```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 Our method assumes linearity in the asymmetric component of the SAM.
-That is, we assume that zonal symmetries associated with positive SAM are oposite and equal to the ones associated with negatie SAM.
+That is, we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negatie SAM.
 @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
-Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (smaller than negative 1 standard deviation) show relatively high pattern correlations all seasons and are visually very linear (Figure A9).
-Therefore, we belive that our method is at the very least a reasonable approximation of the phenomenon.
+Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear (Figure A9).
+Therefore, we belive that our method is a reasonable approximation of the phenomenon.
+
+#LEAN:  Obersvación general: Agrega en todos los captions el período que estás usando para graficar
 
@@
 
+#LEAN: Deberías en el texto decir qué son las siglas de las estaciones (December-February (DJF), ...)
+
 Finally, we assume that the zonally asymmetric pattern is stationary in time.
@@
 
+#LEAN: ¿Se pueden referenciar bien las figuras suplementarias?
+
 ### Regressions
 
-We perform linear regression to quantify the association between the SAM indices and other variables.
+We perform linear regressions to quantify the association between the SAM indices and other variables.
 Since the Asymmetric and Symmetric SAM indices are significantly correlated with each other, to capture the variability explained uniquely by each index we use one multiple linear regression instead of two simple linear regressions.
@@
 
-where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear coefficients, $X_0$ and $\epsilon$ are the constant and error terms.
-From this equarion, $\alpha$ represents the (linear) association of $X$ with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; in other words, it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and viceversa for $\beta$.
+where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and $\epsilon$ are the constant and error terms.
+From this equation, $\alpha$ represents the (linear) association of $X$ with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; i.e.
+it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and viceversa for $\beta$.
+When performing a separate regression for each trimester (DJF, MAM, JJA, SON), the seasonal average of the relevant variables is obtained for each year and trimester before computing the regression.
 
 At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions.
-In such case, using naive p-values to test for significance leads to misleading results [@walker1914; @katz1991].
+In such case, using p-values to test for significance leads to misleading results [@walker1914;@katz1991].
 While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjunting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue.
@@
 
-When performing a separate regression for each trimester (DJF, MAM, JJA, SON) we first average the relevant variables to obtain a single value for each year and each trimester.
-
 # Results
-\label{sec:results}
 
 ## Temporal evolution
-\label{sec:temporal}
 
@@
 
+#LEAN: En la figura 2 quedo un poco raro ahora el "density estimate".
+Creo que por lo menos ayudaría que prolongues la línea del 0 para que sea más clara la lectura.
+Algo que también deberías agregar en la metodología es cómo se calcula.
+
 Month-to-month variability is evident for both indices, with noisy variations in the low frequency.
 At first glance the series can be distinguished by their distributions.
-Compared to the stratospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound.
+Compared to the tropospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound.
 The Asymmetric series have both more variability in the higher frequencies than the Symmetric series.
@@
 In the troposphere the most significant peak of variability is found in the Asymmetric index at around `r 12*.3` months.
+
+#LEAN: Si va a ser la primera vez que nombra la QBO., habría que poner qué significa la sigla y la referencia correspondiente
+
+#LEAN: Un comentario general.
+Creo que es mejor que siempre que hagas referencia a los indices los nombres como "SAM ..." como los nombraste en la metoodología.
+Si te parece muy plomo, quizás se puede poner con algun acrónimo "SAM Sym/Asym".
 
@@
 We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}.
-Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctiating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
+Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
 One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`.
@@
 As a consequence, statospheric Symmetric index tend to precede corresponding Asymmetric index.
+
+#LEAN: Me resulto confuso en la figura 3 el caption con respecto a lo que decis.
+¿Puede ser que esté al revés lo del paréntesis o yo lo estoy entendiendo mal?
+También me queda la duda con respecto a la última oración.
 
@@
 
-```{r trends, fig.cap = "Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+```{r trends, fig.cap = "Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
 Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
+annual_trend <- indices %>%
+  copy() %>% 
+  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
+  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
+  rm_intercept() %>% 
+  .[, term := NULL] %>% 
+  .[, season := "Annual"]
 
@@
   copy() %>% 
+  # .[time > as.Date("1990-01-01")] %>%
   .[, season := season(time)] %>% 
-  rbind(., copy(.)[, season := "Annual"]) %>%
   .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
-  .[, FitLm(estimate_norm, time, se = TRUE), by = .(type = term, lev, season)] %>% 
+  .[, FitLm(estimate_norm, time = time, se = TRUE), by = .(type = term, lev, season)] %>% 
   rm_intercept() %>% 
   .[, term := NULL] %>% 
+  rbind(annual_trend) %>% 
   .[, estimate := estimate*3600*24*365*10] %>% 
   .[, std.error := std.error*3600*24*365*10] %>% 
-  # .[, estimate := estimate/(3600*24*365*10)] %>%
-  # .[, std.error := std.error/(3600*24*365*10)] %>%
   .[, pval := Pvaluate(estimate, std.error, df, "fdr")] %>% 
@@
 
-The trends for each of the indices (Full, Symmetric, Assymetric) were evaluate for the whole period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separed by trimesters.
+The linear trends for each of the indices (Full, Symmetric, Assymetric) were evaluated for the complete period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separed by trimesters.
 The Full SAM index presents a statistically significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throught the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
-The seasonal trends (rest of column a) indicate that positive trends are present in  Autumn and particularly in in Summer, where the 100 hPa maximum is much more defined.
+The seasonal trends (rest of column a) indicate that positive trends are present in  Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
 Positive trends have been documented by previous studies [e.g.
 @fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.
+
+#LEAN: Esto tabień deberías incluirlo en los métodos. Cómo calculaste la tendencia (¿lineal ordinaria?) y cómo estimas el error.
+#LEAN: ¿Qué pasa en invierno y primavera? Deberías indicarlo también.
+#LEAN: Deberías ser más claro con la referencia en lo qué refiere a qué sucede con cada una de las estaciones y si coincide con tus resultados.
 
@@
 In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full index; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
-In Autumn, using the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full index.
+In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full index.
+#LEAN: ¿Qué pasa en invierno y primavera? Deberías indicarlo también.
 
@@
 There is no evidencie of a significant trend in the stratosphere.
-In the troposphere, there is a positive trend for the Asymmetric SAM and no significant trend for the Symmetric SAM.
+In the troposphere, there is a positive trend for the Asymmetric SAM and not significant trend for the Symmetric SAM.
 This suggest that the SAM has become more asymmetric in the period from 1979 to 2018.
-The change is slight, though; of the order of 1% icreased explained variance per decade.
+However, the change is slight, around 1% increased explained variance per decade.
 
@@
 ## Spatial patterns
-\label{sec:spatial}
 
@@
 
-```{r 2d-regr, fig.cap = "Regression patterns of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.
+```{r 2d-regr, fig.cap = "Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM.
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices.
 Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index.
@@
 
-To show if, and to what extent, the Asymmetric and Symmetric SAM inidices indeed capture the asymmetric and symmetric compoment of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index.
+To show if, and to what extent, the Asymmetric and Symmetric SAM inidices indeed capture the asymmetric and symmetric compoment of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa levels.
 Figure \ref{fig:2d-regr} shows these regressions.
@@
 In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the South Pole.
-The monopole obtained by multiple regression with the Asymmetric and Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel b.1).
+The monopole obtained by the regression pattern for Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel b.1).
 
@@
 The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
-The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is assosiated with an trully annular mode, almost devoid of zonal asymmetries (panel c.2).
+The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is associated with an annular mode, almost devoid of zonal asymmetries (panel c.2).
 The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern asociated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure.
 Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
+
+#LEAN: Deberías incluir alguna/s referencia para la primera oración
 
@@
 Looking at panel `r get_tag(term == "asym" & lev == 700)` from Figure \ref{fig:2d-regr}, it becomes apparent that zonal wavess 1 and 2 modulate the amplitude of zonal wave 3, which --as mentioned before-- is larger in the Western Hemisphere than in the Easten Hemisphere.
+
+#LEAN: En la metodología deberías aclarar cómo caluclas también los números de onda zonal
+
+#LEAN: Este parráfo deberías ampliarlo un poco, primero describiendo lo que se ve en el full, las diferencias entre los niveles y luego, cómo se descompone eso entre ambas componentes.
 
@@
 
-To analyse the vertical structure of the geopotential anomalies asociated with the asymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
+To analyse the vertical structure of the geopotential anomalies asociated with the assymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
 The geopotential anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
@@
 Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
+
+#LEAN: ¿Sería geopotential height en realidad, no?
 
@@
 So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
-
-The structures shown in panels a and b in  Figure \ref{fig:vertical-regression} are surprisignly robust to the choice of index level.
+The structures shown in panels a and b in  Figure \ref{fig:vertical-regression} are robust to the choice of index level.
 For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in panel a.
 Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b.
-The patterns mainly change in amplitude.
+The patterns mainly change in amplitude (not shown).
 
@@
 The correlation between the full SAM and the ENSO as measured by the Oceanic Niño Index [@bamston1997] (ONI) is `r cor_texts[index == "oni"]$full`.
-Consistent with @fan2007, we show that this relationship is captured entirely the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni"]$asym` with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni"]$sym`).
+Consistent with @fan2007, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni"]$asym` with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni"]$sym`).
 We performed the same analysis using the Multivariate Enso Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987] to conclude that these results do not depend on the ENSO index used.
 
+#LEAN: No me queda claro acá que sería consistente con Fan (2007). Sería bueno seguramente que aparezca en la introducción
+
+#LEAN: Me parece que sería bueno decir algo sobre qué tan significativos son cada una de esas relaciones.
+
 ## Impacts
-\label{sec:impacts}
 
@@
 
-```{r regr-air-season, fig.cap = "Regression pattern of season mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM.
+```{r regr-air-season, fig.cap = "Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 
+#Para la parte de los estudios de impacto sería bueno enfatizar el nivel qué utilizas y por qué (considerando que antes venis mostrando siempre lo que pasa también en la estratósfera)
+
+#LEAN: Con lo que empezas esta parte, lo mandaría a la introducción
+
 Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface temperature for each trimester.
-In Summer positive values of the Full SAM index (panel `r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
+In Summer, positive values of the Full SAM index (panel `r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
 The ring is not zonally symmetric, as there are three clear local maximums around 30\degree W, 15\degree E and 50\degree E and a local minimum (with negative sign) around 120\degree W.
@@
 Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the Asymetric SAM regression map, while temperature patterns linked to positive Symmetric SAM show a more zonally consistent ring and less relation to the tropics.
-Noticeable, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to positive values of Asymmetric SAM.
+Noticeable, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to Asymmetric SAM.
 This signal is not present in the regression pattern with the Full SAM.
-Spring (row `r get_row(season == "SON")`) features very similar patterns but of generally smaller in magnitude and statistical significance.
+Spring (row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with less regions where regressed anoamlies have statistical significance.
+
+#LEAN: Me están resultando confusas las referencias espaciales que hace en "The ring is not zonally symmetric, as there are three clear local maximums around 30\degree W, 15\degree E and 50\degree E and a local minimum (with negative sign) around 120\degree W".
+¿Son correctas?
 
@@
 These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following @wilks2016.
-However, repeating this analysis with 2-meter temperature from ERA5 resulted in similar patterns that were statistically significant.
+However, repeating this analysis with 2-meter temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
 Moreover, similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
@@
 
+# LEAN: Agregaría alguna referencia para la primer oración
+
 These results suggests that Asymmetric and Symmetric SAM indices are associated with overall distinct temperature patterns which may not be apparent when using the Full SAM index.
 
+# LEAN: "These results" no me queda del todo claro a qué te estás refiriendo
+
 Figure \ref{fig:regr-air-season} column b can be partially compared with Figure 11 from @fogt2012.
-Although they used station data from 1958 to 2001, a lot of the characteristics are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.
-
-Regression of the SAM indicies with seasonal mean precipitation and 700 hPa geopotential height are shown Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
+Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.
+
+Regression of the SAM indicies with seasonal mean precipitation and 700 hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
 South Africa is not shown because no significant signal was detected there.
@@
 
-```{r pp-regr-africa,  fig.cap = "Regression pattern of precipiation with Asymmetric and Symmetric SAM.
+```{r pp-regr-africa,  fig.cap = "Regression of precipiation anomalies with Asymmetric and Symmetric SAM.
 P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
@@
 
-```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression pattern of (row 1) annual and (rows 2 to 5) season mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM.
+```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM.
 thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
@@
 ```
+
+#LEAN: Una observación sobre los mapas de precipitación es que pueden ser contraintuitivos por la escala de colores que elegiste.
+En general se suele asociar preciptiación positiva con colores más azulados/verdosos.
+Creo que en ese sentido lo mejor sería tener un colormap distinto a las otras figuras (quizás uno que vaya de marrones a verdes o algo así)
 
@@
 The Asymmetric SAM appears related to increased precipitation in the West coast of Southeastern Australia (panel b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
-
-This Spring signal is broadly consistent with @hendon2007, but whereas @hendon2007 also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coeffcients have the consistent sign).
 
@@
 In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the Symmetric SAM.
-None of these regression coefficients are statistically signifincat at the 95% level
+None of these regression coefficients are statistically signifincat at the 95% level.
+The Spring signal is broadly consistent with @hendon2007, but whereas @hendon2007 also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coeffcients have the consistent sign).
 
@@
 
-In South America (Figure \ref{fig:pp-regr-america}), the annual-level regression shows that the SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1).
+In South America (Figure \ref{fig:pp-regr-america}), the annual-level regression shows that positive SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1).
 Panels b.1 and c.1 show a remarkably clean separation between the Asymmeric SAM --associated with the Southeastern South American and Southern Brazilian signals-- and the Symmetric SAM --associated with the signal in Southern Chile.
@@
 On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
-This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region [@silvestri2009; @zamboni2010].
+This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region [@silvestri2009, @zamboni2010].
 This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events [@rosso2018].
-
 There is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 and that is explained by the Asymmmetric SAM.
 
+#LEAN: Podes agregar esta cita creo para la segunda oración https://journals.ametsoc.org/jcli/article/20/18/4789/31713
+
 ## Conclusions
 
-In this study we tried to systematically characterise the variability of the zonally symmetric and zonally asymmetric structure of the SAM.
-By projecting monthly geopotential fields at each level with the corresponding asymmetric and symmetric pattern, we created two indices representing the zonally asymmetric and zonally symmetric contributions of the SAM respectively.
+In this study we characterise the temporal and spatial variability of the zonally symmetric and asymmetric structure of the SAM.
+By projecting monthly geopotential fields at each level with the corresponding asymmetric and symmetric pattern, we created two indices for representing the zonally asymmetric and symmetric contributions of the SAM respectively.
+
+#LEAN: Cambié el "tried" porque en la conclusión hay que ser un poco más afirmativo.
 
@@
 There is evidence of influence between the QBO and the Northern Annular Mode [e.g.
-@holton1980; @watson2014; @zhang2020], so it's not unlikely that the SAM would be similarly connected.
+@holton1980, @watson2014, @zhang2020], so it's not unlikely that the SAM would be similarly connected.
 However establishing this link would require further research.
 
-As documented by previous studies, such as @fogt2020 (and references therein), we observe a positive trend towards positive SAM in Summer and Autumn.
+we observe a positive trend towards positive SAM in Summer and Autumn, As was documented by previous studies, such as @fogt2020 (and references therein) for surface levels.
 We show that these trends are maximised at the 100 hPa level and are explained by the zonally symmetric component.
@@
 
-In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM is much closer to being trully annular than the patterns associated with the Full SAM index.
-The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific region and whose phae is rotated a quarter wavelength from the mean zonal wave 3 described by @raphael2004's index.
+In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being trully annular than the patterns associated with the Full SAM index.
+The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific region and whose phase is rotated a quarter wavelength from the mean zonal wave 3 described by @raphael2004's index.
 This pattern extends in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimum for capturing this variability.
@@
 The Asymemtric SAM index explains a big portion of these deviations.
-In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship delineated above.
+In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship.
 These are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [@cazes-boezio2003; @fogt2011; @cai2020a].
 
+#LEAN: Acá también el "these" no me queda del todo claro
+
 In Australia the Full SAM is associated with positive precipitation anomalies in South East and this is explained by the Symemtric SAM.
-However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
-
-In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
+However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
 This features are cleanly separated between the Asymmetric and Symmetric components.
@@
 The correlation between ENSO and SAM is similarly non-stationary, also disapearing before 1973.
-
 Seeing as both the ENSO-SAM relationship and most of the precipitation imacts in South America are captured by the Asymmetric SAM, the results presented here are most likely period-dependent.
-Therefore, is very likely that if we were to repeat this analysis using pre-satellite data, the resulting Asymmetric SAM would look very different.
+
+#LEAN: Revisaste más bibliografía con respecto a la relación ENSO-SAM?
+
+#LEAN: Fatlaría un último parráfo de cierre.
+Lo que se me ocurre que se puede decir es lo positivo del estudio en cuanto a qué permite separar la señales simétricas/asimétricas, y ahi también ponerlas en contexto del cambio climático.
+También podes agregar alguna cosa que pueda quedar como futuras investigaciones.
 
@@
 NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ 
+The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
 
@@
 
-```{r A3, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with withs 3 and 5.
+```{r A3, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
 The shading indicates de 95\\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\\% of the simulated sampels had an amplitude equal or lower).

### 2020-10-01 analysis/paper/paper.Rmd
segudno borrador

@@
   This is significant becasue I wrote it.
-bibliography: AsymSAM.bib
+bibliography: "AsymSAM.bib,packages.bib"
 output: 
@@
 Of the several different indices presented in the literature, many of them are based on zonal means of sea level pressure or geopotential height [@ho2012].
-@gong1999 defined the SAM index as the zonal mean sea level pressure differene between 40\degree S and 60\degree S, which is also the definition used by the station-based index in @marshall2003.
+@gong1999 defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 60\degree S, which is also the definition used by the station-based index in @marshall2003.
 @baldwin2009 proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
@@
 These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009].
-At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availabilty of dense observations in the Southern Hemisphere [@jones2009].
-
-At least some of the variabilty associated with the zonal asymmetries of the SAM is probably forced by the tropics.
-In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987; @kidson1988;@karoly1989], whose wave train proyects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
+At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].
+
+At least some of the variability associated with the zonal asymmetries of the SAM is probably forced by the tropics.
+In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987; @kidson1988;@karoly1989], whose wave train protects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
 And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed [@fan2007; @fogt2011; @clem2013].
+In particular, @fan2007 computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated if the (linear) signal of the ENSO was removed.
+While this relates to temporally coherent variability of the two hemisphere and not necessarily to zonal asymmetries in the associated spatial patterns, it is nonetheless consistent of the ENSO playing a crucial role in zonal asymmetries of the SAM.
 
@@
 These trends are thought of driven primarily by stratospheric ozone depletion and understood in the context of zonal mean variables.
-However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variabilty could be masking influencing the observed trends independendly.
+However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variability could be masking influencing the observed trends independently.
 
 Similarly unclear are the specific impacts of the zonally asymmetric component of the SAM.
-Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes [@jones2019] (and viceversa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012].
+Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012].
 The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
 The importance of zonal asymmetries of the SAM in these impacts have been studied in certain regions.
-For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymetric circulation associated with the SAM [@silvestry2009; @rosso2018].
-@fan2007 also found that precipitation in East Asia was impacted by the variabilty of only the Western Hemisphere part of the SAM.
+For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
+@fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
 
@@
 However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988).
-This is particularly important due to the inhomogeneities in renalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].
+This is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].
 
@@
 In the Section \ref{sec:methods} we describe the methods.
-In Section \ref{sec:results}.\ref{sec:temporal} we describe the temporal variability and vertical coherence of the inidices.
+In Section \ref{sec:results}.\ref{sec:temporal} we describe the temporal variability and vertical coherence of the indices.
 In  Section \ref{sec:results}.\ref{sec:spatial}, we analyse the spatial patterns of geopotential height associated with them.
-In Section \ref{sec:results}.\ref{sec:impacts}, we study their relationship with surface-level tempreature and precipitation.
+In Section \ref{sec:results}.\ref{sec:impacts}, we study their relationship with surface-level temperature and precipitation.
 
@@
 # Methods
+\label{sec:methods}
 
@@
 
-To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
+To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
 We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.
 
-To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree lagitude by 5\degree longitude, and monthly rainfall 0.5\degree latitude by 0.5\degree longitude data from the Global Precipitation Climatology Centre [@schneider2015].
+To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude, and monthly rainfall 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015].
 
@@
 
-To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure\ \ref{fig:method} for 700hPa.
+To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \ref{fig:method} for 700 hPa.
 The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
@@
 However, only "Full" index will have unit standard deviation per definition.
-From the regression, we also use the explained variance of each pattern as an indicator of the degere of zonally symmetry or asymmetry of each monthly field.
+From the regression, we also use the explained variance of each pattern as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
 
@@
 Our method assumes linearity in the asymmetric component of the SAM.
-That is, we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negatie SAM.
+That is, we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM.
 @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
 Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear (Figure A9).
-Therefore, we belive that our method is a reasonable approximation of the phenomenon.
-
-#LEAN:  Obersvación general: Agrega en todos los captions el período que estás usando para graficar
-
-By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons.
+Therefore, we believe that our method is a reasonable approximation of the phenomenon.
+
+By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons --December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
 Geopotential zonal anomalies computed by projecting the first EOF *of each season* are very similar to each other (Figure A10) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (between MAM and SON).
 Based on this, we believe that our initial assumption is not unreasonable.
-
-#LEAN: Deberías en el texto decir qué son las siglas de las estaciones (December-February (DJF), ...)
 
@@
 
-#LEAN: ¿Se pueden referenciar bien las figuras suplementarias?
-
 ### Regressions
@@
 Since the Asymmetric and Symmetric SAM indices are significantly correlated with each other, to capture the variability explained uniquely by each index we use one multiple linear regression instead of two simple linear regressions.
-To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Asymmetric SAM ($SAM_a$) and Symmetric SAM ($SAM_s$) we fit the ecuation
+To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Asymmetric SAM ($SAM_a$) and Symmetric SAM ($SAM_s$) we fit the equation
 
@@
 From this equation, $\alpha$ represents the (linear) association of $X$ with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; i.e.
-it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and viceversa for $\beta$.
-When performing a separate regression for each trimester (DJF, MAM, JJA, SON), the seasonal average of the relevant variables is obtained for each year and trimester before computing the regression.
+it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and vice versa for $\beta$.
+When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we average of the relevant variables seasonally for each year and trimester before computing the regression.
 
@@
 In such case, using p-values to test for significance leads to misleading results [@walker1914;@katz1991].
-While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjunting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue.
+While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjusting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue.
 Therefore, p-values showed in regression fields are all adjusted following @benjamini1995.
 
+We computed linear trends by Ordinary Least Squares and the 95% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
+
+To compute the amplitude of the zonal waves we computed the Fourier transform of the spatial field at each latitude circle.
+
+### Computation procedures
+
+We performed all analysis in this paper using the R programming language [@rcoreteam2020], using the data.table package [@dowle2020] and the metR package [@campitelli2020].
+All graphics are made using ggplot2 [@wickham2009].
+We downloaded data from ranalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
+The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].
+
+A version-controlled repository of the code used to create this analysis can be found at https://github.com/eliocamp/asymsam.
+
 # Results
+\label{sec:results}
 
 ## Temporal evolution
+\label{sec:temporal}
 
@@
   indices <- readRDS(indices_file)
-  indices[, `:=`(estimate = -estimate, estimate_norm = -estimate_norm)]
+  indices[, `:=`(estimate = -estimate, 
+                 estimate_norm = -estimate_norm)]
 } else {
@@
 ```{r asymsam-timeseries, fig.cap = "Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
-To the left, probability density estimate of each index.
+To the left, probability density estimate of each index computed using a gaussian kernel of optimal bindwidth according to \\cite{sheather1991}.
 Series are standarised by the standard deviation of the Full SAM at each level.", fig.width=6, fig.height=3, fig.env="figure*"}
@@
   ggplot(aes(estimate_norm)) +
-  geom_density(aes(color = term)) +
+  geom_density(aes(color = term), bw = "SJ") +
   facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
@@
   scale_y_continuous(NULL, trans = scales::reverse_trans(), labels = NULL, breaks = NULL) +
-  scale_x_continuous(NULL, labels = NULL, breaks = NULL) +
+  scale_x_continuous(NULL, labels = NULL, breaks = 0) +
   coord_flip()  +
@@
 
-The temporal evolution of the Assymmetric and Symmetric SAM was firstly asssesed.
+We first asses the temporal evolution of the Asymmetric SAM and Symmetric SAM.
 Figure \ref{fig:asymsam-timeseries} shows the corresponding time series for 700 hPa and 50 hPa and their corresponding density estimates.
@@
 
-#LEAN: En la figura 2 quedo un poco raro ahora el "density estimate".
-Creo que por lo menos ayudaría que prolongues la línea del 0 para que sea más clara la lectura.
-Algo que también deberías agregar en la metodología es cómo se calcula.
-
 Month-to-month variability is evident for both indices, with noisy variations in the low frequency.
@@
 Compared to the tropospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound.
-The Asymmetric series have both more variability in the higher frequencies than the Symmetric series.
+The Asymmetric SAM series have both more variability in the higher frequencies than the Symmetric SAM series.
 
 The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A3).
-This might suggests a link between stratospheric SAM variability and QBO.
-There is a local peak at 2 years in the periodigram of the tropospheric Symmetric SAM also, although it's not statistically significant.
+This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b].
+There is a local peak at 2 years in the periodogram of the tropospheric Symmetric SAM also, although it's not statistically significant.
 In the troposphere the most significant peak of variability is found in the Asymmetric index at around `r 12*.3` months.
-
-#LEAN: Si va a ser la primera vez que nombra la QBO., habría que poner qué significa la sigla y la referencia correspondiente
-
-#LEAN: Un comentario general.
-Creo que es mejor que siempre que hagas referencia a los indices los nombres como "SAM ..." como los nombraste en la metoodología.
-Si te parece muy plomo, quizás se puede poner con algun acrónimo "SAM Sym/Asym".
 
@@
 
-From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric and Symmetric time series appear to be correlated.
-Moreover, looking at the extremes in the stratosphere, the Symmetric serie appears to lag the Asymmetric series (see, for example, the positive events on late 1987).
-We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric index leading the Symmetric index), in Figure \ref{fig:cor-lev}.
-Zero-lag correlations between the Asymmetric and Symmetric series are relatively constant throught the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
+From Figure \ref{fig:asymsam-timeseries} we can see that the Asymmetric SAM and Symmetric SAM time series appear to be correlated.
+Moreover, looking at the extremes in the stratosphere, the Symmetric SAM series appears to lag the Asymmetric SAM series (see, for example, the positive events on late 1987).
+We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric SAM index leading the Symmetric SAM index), in Figure \ref{fig:cor-lev}.
+Zero-lag correlations between the Asymmetric SAM and Symmetric SAM series are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
 One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`.
@@
 At the same time, one-month-lag correlations increase with height.
-As a consequence, statospheric Symmetric index tend to precede corresponding Asymmetric index.
-
-#LEAN: Me resulto confuso en la figura 3 el caption con respecto a lo que decis.
-¿Puede ser que esté al revés lo del paréntesis o yo lo estoy entendiendo mal?
-También me queda la duda con respecto a la última oración.
-
-```{r cor-lev, fig.cap = "Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Symmetric leads Asymmetric).", fig.width=3, fig.height=3}
+As a consequence, stratospheric Asymmetric SAM index tend to precede corresponding Symmetric SAM index.
+
+```{r cor-lev, fig.cap = "Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979–2018 period.", fig.width=3, fig.height=3}
 cors %>% 
@@
   scale_y_continuous("Correlation", limits = c(0, 1)) + 
-  scale_linetype_manual(NULL, values = c(1, 5), labels = c("0" = "Zero-lag", 
-                                                           "-1" = "Symmetric leads\nAsymmetric by 1 month")) +
-  scale_color_brewer(NULL, palette = "Dark2", labels = c("0" = "Zero-lag", 
-                                                         "-1" = "Symmetric leads\nAsymmetric by 1 month")) +
+  scale_linetype_manual(NULL, values = c(1, 5), 
+                        labels = c("0" = "Zero-lag", 
+                                   "-1" = "Asymmetric SAM leads\nSymmetric SAM by 1 month")) +
+  scale_color_brewer(NULL, palette = "Dark2", 
+                     labels = c("0" = "Zero-lag",
+                                "-1" = "Asymmetric SAM leads\nSymmetric SAM by 1 month")) +
   coord_flip()
@@
 
-```{r cross-correlation, fig.cap = "Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM.", fig.width = 6, fig.height=3, fig.env="figure*"}
+```{r cross-correlation, fig.cap = "Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979–2018 period.", fig.width = 6, fig.height=3, fig.env="figure*"}
 tri <- matrix(c(1000, 1000, 1000, 1, 1, 1, 1000, 1000), ncol = 2, byrow = TRUE) %>% 
@@
 Figure \ref{fig:cross-correlation}a shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
-For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposfere.
+For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
 Above 100 hPa correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
 Therefore, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
-Examining panels b and c, we see that the Asymemtric and Symmetric SAM  share the same high level of coherency in the troposphere but they differ in their stratospheric behaviour.
-Stratospheric coherency is stronger for the Asymmetric SAM than the Symemtric SAM.
-The stratospheric Symmetric SAM seems to connect more strongly to the trosposphere than the Asymmetric SAM.
+Examining panels b and c, we see that the Asymmetric and Symmetric SAM  share the same high level of coherency in the troposphere but they differ in their stratospheric behaviour.
+Stratospheric coherency is stronger for the Asymmetric SAM than the Symmetric SAM.
+The stratospheric Symmetric SAM seems to connect more strongly to the troposphere than the Asymmetric SAM.
 
@@
 ```{r trends, fig.cap = "Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
-Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
-annual_trend <- indices %>%
-  copy() %>% 
-  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
-  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
-  rm_intercept() %>% 
-  .[, term := NULL] %>% 
-  .[, season := "Annual"]
-
+Shading indicates the 95\\% confidence interval from a t-distribution.", fig.width = 6, fig.height=4, fig.env="figure*"}
 indices %>% 
   copy() %>% 
-  # .[time > as.Date("1990-01-01")] %>%
+  .[, lapply(.SD, mean), by = .(lev, PC, term, time = seasonally(time))] %>% 
   .[, season := season(time)] %>% 
+  rbind(., copy(indices)[, season := "Annual"]) %>% 
   .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
@@
   .[, term := NULL] %>%  
-  rbind(annual_trend) %>% 
   .[, estimate := estimate*3600*24*365*10] %>% 
@@
   .[, season := forcats::fct_relevel(season, "Annual")] %>% 
+  .[, t := qt(.975, df)] %>% 
   ggplot(aes(lev, estimate)) +
   geom_hline(yintercept = 0) +
-  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), 
+  geom_ribbon(aes(ymax = estimate + std.error*t, ymin = estimate - std.error*t), 
               fill = "#95a3ab", color = "black", 
@@
 
-The linear trends for each of the indices (Full, Symmetric, Assymetric) were evaluated for the complete period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separed by trimesters.
-The Full SAM index presents a statistically significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throught the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
+The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separated by trimesters.
+The Full SAM index presents a statistically significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
 The seasonal trends (rest of column a) indicate that positive trends are present in  Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
-Positive trends have been documented by previous studies [e.g.
+In Winter and Spring, we detect no statistically significant trend.
+This is consistent with the results of previous studies, which find clear positive trends in Summer, weaker in Autumn and no trends in the other seasons [e.g.
 @fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.
 
-#LEAN: Esto tabień deberías incluirlo en los métodos. Cómo calculaste la tendencia (¿lineal ordinaria?) y cómo estimas el error.
-#LEAN: ¿Qué pasa en invierno y primavera? Deberías indicarlo también.
-#LEAN: Deberías ser más claro con la referencia en lo qué refiere a qué sucede con cada una de las estaciones y si coincide con tus resultados.
-
-By separating the SAM signal in its Asymmetric and Symmetric parts, we can not only see that these trends are almost entirely due to the Symmetric component (colum b vs.
+By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.
 column c), but in some cases the trends become more clear.
-In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full index; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
-In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full index.
-#LEAN: ¿Qué pasa en invierno y primavera? Deberías indicarlo también.
+In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
+In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full SAM index.
 
@@
 
-We stress that these are only linear trends during the whole period and the absence of a statistically significant signal should not be taken as evidence of no sistematic change.
-In particular, going back to Figure \ref{fig:asymsam-timeseries}, we can see an evident change in the stratospheric Asymemtric component (red line in panel a) between the 90's, when we see a dominance of extreme negative values, and the 00's, when we see the inverse.
-This change is restricted to the Winter months: the linear trend for JJA starting in 1990 for the Asymmetric component at 50hPa is $`r trends_text[term == "asym" & lev == 50 & season == "JJA"]$V1`$.
+We stress that these are only linear trends during the whole period and the absence of a statistically significant signal should not be taken as evidence of no systematic change.
+In particular, going back to Figure \ref{fig:asymsam-timeseries}, we can see an evident change in the stratospheric Asymmetric component (red line in panel a) between the 90's, when we see a dominance of extreme negative values, and the 00's, when we see the inverse.
+This change is restricted to the Winter months: the linear trend for Winter starting in 1990 for the Asymmetric component at 50hPa is $`r trends_text[term == "asym" & lev == 50 & season == "JJA"]$V1`$.
 
 Figure \ref{fig:r-squared-trend} shows decadal trends for the explained variance of each index.
-There is no evidencie of a significant trend in the stratosphere.
+There is no evidence of a significant trend in the stratosphere.
 In the troposphere, there is a positive trend for the Asymmetric SAM and not significant trend for the Symmetric SAM.
@@
 
-```{r 2d-regr, fig.cap = "Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM.
+```{r 2d-regr, fig.cap = "Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979–2018 period.
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices.
 Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index.
-", fig.env = "figure*", fig.width=7, fig.height=5}
+", fig.env = "figure*", fig.width=6, fig.height=5}
 indices_2d_yearly %>% 
@@
 
-To show if, and to what extent, the Asymmetric and Symmetric SAM inidices indeed capture the asymmetric and symmetric compoment of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa levels.
+To show if, and to what extent, the Asymmetric and Symmetric SAM indices indeed capture the asymmetric and symmetric component of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa levels.
 Figure \ref{fig:2d-regr} shows these regressions.
@@
 
-In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centered in the South Pole.
-The monopole obtained by the regression pattern for Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel b.1).
-
-In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3.
+In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centred in the South Pole.
+The monopole obtained by the regression pattern for Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davis Sea in the Eastern Hemisphere (panel b.1).
+
+In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 [@fogt2012].
 The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
-The Asymmetric component gives rise to a cleaner zonal wave (panel b.2) and the Symemtric component is associated with an annular mode, almost devoid of zonal asymmetries (panel c.2).
-The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern asociated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure.
+The Asymmetric SAM index gives rise to a cleaner zonal wave (panel b.2) and the Symmetric SAM index is associated with an annular mode, almost devoid of zonal asymmetries (panel c.2).
+The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern associated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure.
 Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
-
-#LEAN: Deberías incluir alguna/s referencia para la primera oración
 
@@
 
-```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.
-Note the differnet x axis.", fig.width=3, fig.height=3}
+```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.", fig.width=6, fig.height=3, fig.env = "figure*"}
 indices_2d_yearly %>% 
@@
   coord_flip() +
-  facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab), scales = "free") +
+  facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
   tag_facets("cr", position = "tr")  +
-  no_grid +
-  theme(strip.text = element_text(size = 7))
-```
-
-The amplitude of first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
-Column `r get_col(term == "asym")` shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel `r get_col(term== "asym" & lev == 50)`), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel `r get_col(term== "asym" & lev == 700)`).
-Looking at panel `r get_tag(term == "asym" & lev == 700)` from Figure \ref{fig:2d-regr}, it becomes apparent that zonal wavess 1 and 2 modulate the amplitude of zonal wave 3, which --as mentioned before-- is larger in the Western Hemisphere than in the Easten Hemisphere.
-
-#LEAN: En la metodología deberías aclarar cómo caluclas también los números de onda zonal
-
-#LEAN: Este parráfo deberías ampliarlo un poco, primero describiendo lo que se ve en el full, las diferencias entre los niveles y luego, cómo se descompone eso entre ambas componentes.
+  no_grid 
+```
+
+The amplitude of the first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
+Zonal wave amplitudes of the spatial pattern described by the Full SAM index (column `r get_col(term == "full")`) are dominated by the zonal mean (wave-number 0) at both levels.
+However, zonal waves are important, particularly North of 50\degree S, with wave-number 1 clearly dominating at 50 hPa (panel `r get_tag(term == "full" & lev == 50)`) and a more equal mix of waves at 700 hPa (panel `r get_tag(term == "full" & lev == 700)`).
+Column `r get_col(term == "asym")` shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel `r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel `r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean.
+The Symmetric SAM, on the other hand, it's almost entirely composed of zonal mean at both levels (column `r get_col(term == "sym")`), with little to now contribution from zonal waves with wave-numbers 1 to 3.
+
+We can see that the amplitude and latitudinal distribution zonal waves in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric SAM on the other correspond almost exactly to the amplitude and latitudinal distribution in the Full SAM.
+This confirms the correct decomposition of the SAM in its symmetric and asymmetric components.
+
+Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which --as mentioned before-- is larger in the Western Hemisphere than in the Eastern Hemisphere.
 
@@
 
-```{r vertical-regression, fig.cap = "Regression between monthly geopotential anomalies (meters) averaged betweeen 65\\degree and 40\\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
-(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa.", fig.width = 3, fig.height=3}
+```{r vertical-regression, fig.cap = "Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\\degree and 40\\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
+(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979–2018 period.", fig.width = 3, fig.height=3}
 vertical_regr %>% 
@@
 
-To analyse the vertical structure of the geopotential anomalies asociated with the assymetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
-The geopotential anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
+To analyse the vertical structure of the geopotential height anomalies associated with the asymmetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
+The geopotential height anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
 The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes.
 Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
-
-#LEAN: ¿Sería geopotential height en realidad, no?
 
@@
 The anomalies are much more intense in the Western hemisphere, where they extent into the stratosphere.
-In the Eastern hemisphere the wave-3 signal is weaker and confined to the troosphere while negative anomalies dominate in the stratosphere.
+In the Eastern hemisphere the wave-3 signal is weaker and confined to the troposphere while negative anomalies dominate in the stratosphere.
 So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
@@
            list(cor = estimate, 
+                p.value = p.value, 
                 low = conf.int[1], 
@@
 
-cor_texts <- enso_sam_cor[lev == 700, 
-                         paste0(signif(cor, 2)), # " (CI: ", signif(low, 2), "–", signif(upp, 2), ")"), 
-                         by = .(term, index)] %>% 
-  dcast(index~term)
+cor_texts <- enso_sam_cor[lev == 700 & index == "oni"] %>% 
+  .[, `:=`(cor = signif(cor, 2), 
+           p.value = signif(p.value, 2)
+  )]
+
+enso_sam_partial <- enso %>% 
+  .[indices_wide[lev == 700], on = "time"] %>% 
+  .[, partial_cor(value, asym, sym), by = .(season(time), index)] %>% 
+  .[, partial_correlation := signif(partial_correlation, 2)] %>% 
+  .[, p.value := signif(p.value, 2)]
 
@@
   .[, partial_correlation := signif(partial_correlation, 2)] %>% 
-  dcast(index~term)
+  .[, p.value := signif(p.value, 2)]
 ```
@@
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the full SAM and the ENSO as measured by the Oceanic Niño Index [@bamston1997] (ONI) is `r cor_texts[index == "oni"]$full`.
-Consistent with @fan2007, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni"]$asym` with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni"]$sym`).
-We performed the same analysis using the Multivariate Enso Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987] to conclude that these results do not depend on the ENSO index used.
-
-#LEAN: No me queda claro acá que sería consistente con Fan (2007). Sería bueno seguramente que aparezca en la introducción
-
-#LEAN: Me parece que sería bueno decir algo sobre qué tan significativos son cada una de esas relaciones.
+The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index [@bamston1997] (ONI) is `r cor_texts[term == "full"]$cor` (p-value = `r cor_texts[term == "full"]$p.value`).
+Consistent with @fan2007, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value = `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value = `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
+We performed the same analysis using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987] to conclude that these results do not depend on the ENSO index used.
 
 ## Impacts
+\label{sec:impacts}
 
@@
 
-```{r regr-air-season, fig.cap = "Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM.
+```{r regr-air-season, fig.cap = "Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979–2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 
-The SAM has been shown to be associated with important surface variables such as temperature and precipitacion [e.g.
-@gillett2006, and see @fogt2020 for a review].
-Naturally, most studies on the surface impacts of the SAM are based on an index identical or analogous to what we call Full SAM index (@fogt2012 being the only exception that we are aware of).
-We regress surface temperature and precipitation onto each of the three SAM indices to see if there are diferent surface impact associated with the asymmetric and symmetric SAM circulation.
-
-#Para la parte de los estudios de impacto sería bueno enfatizar el nivel qué utilizas y por qué (considerando que antes venis mostrando siempre lo que pasa también en la estratósfera)
-
-#LEAN: Con lo que empezas esta parte, lo mandaría a la introducción
+To see if there are different surface impacts associated with the asymmetric and symmetric SAM circulation we regress surface temperature and precipitation onto each of the three SAM indices at 700 hPa.
+As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
 
@@
 In Summer, positive values of the Full SAM index (panel `r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
-The ring is not zonally symmetric, as there are three clear local maximums around 30\degree W, 15\degree E and 50\degree E and a local minimum (with negative sign) around 120\degree W.
+The ring is not zonally symmetric, as there are four clear local maximums around 30\degree W,  120\degree W, 150\degree E and 90\degree E.
 In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
 Panels `r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of the Asymmetric and Symmetric SAM, respectively.
-Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the Asymetric SAM regression map, while temperature patterns linked to positive Symmetric SAM show a more zonally consistent ring and less relation to the tropics.
+Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the Asymmetric SAM regression map, while temperature patterns linked to positive Symmetric SAM show a more zonally consistent ring and less relation to the tropics.
 Noticeable, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to Asymmetric SAM.
 This signal is not present in the regression pattern with the Full SAM.
-Spring (row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with less regions where regressed anoamlies have statistical significance.
-
-#LEAN: Me están resultando confusas las referencias espaciales que hace en "The ring is not zonally symmetric, as there are three clear local maximums around 30\degree W, 15\degree E and 50\degree E and a local minimum (with negative sign) around 120\degree W".
-¿Son correctas?
-
-In Autumn and Winter (rows `r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))`) the postive ring is only present through its local maximums in the regression with the Full SAM.
+Spring (row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with less regions where regressed anomalies have statistical significance.
+
+In Autumn and Winter (rows `r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))`) the positive ring is only present through its local maximums in the regression with the Full SAM.
 There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
@@
 
-The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons --although with varying intensity and small-scale details-- is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM.
-It's then not surprising to see it more clearly in association with the Symemtric SAM (at least in Summer and Spring).
-
-# LEAN: Agregaría alguna referencia para la primer oración
-
-These results suggests that Asymmetric and Symmetric SAM indices are associated with overall distinct temperature patterns which may not be apparent when using the Full SAM index.
-
-# LEAN: "These results" no me queda del todo claro a qué te estás refiriendo
+The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons --although with varying intensity and small-scale details-- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
+It's then not surprising to see it more clearly in association with the Symmetric SAM (at least in Summer and Spring).
 
@@
 
-Regression of the SAM indicies with seasonal mean precipitation and 700 hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
+Regression of the SAM indices with seasonal mean precipitation and 700 hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
 South Africa is not shown because no significant signal was detected there.
@@
   .[, season := "Annual"]
+
 pp_regr <- rbind(pp_regr_season, pp_regr_annual) %>% 
@@
                          oob = scales::squish,
+                         low = "#8E521C",
+                         high = "#00665E",
                          # label = scales::percent_format(),
@@
                          breaks = AnchorBreaks(0, 0.1, 0)) +
+
     scale_linetype(guide = "none") +
@@
 
-```{r pp-regr-africa,  fig.cap = "Regression of precipiation anomalies with Asymmetric and Symmetric SAM.
+```{r pp-regr-africa,  fig.cap = "Regression of precipiation anomalies with Asymmetric and Symmetric SAM for the 1979–2018 period.
 P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
@@
 
-```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM.
-thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
+```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979–2018 period.
+Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
 plot_pp("oceania") 
 ```
-
-#LEAN: Una observación sobre los mapas de precipitación es que pueden ser contraintuitivos por la escala de colores que elegiste.
-En general se suele asociar preciptiación positiva con colores más azulados/verdosos.
-Creo que en ese sentido lo mejor sería tener un colormap distinto a las otras figuras (quizás uno que vaya de marrones a verdes o algo así)
 
@@
 The seasonal-level regressions show statistically significant anomalies only in Spring, when positive Full SAM is associated with positive precipitation anomalies in Eastern Australia (panel a.5).
-In this trimester the Symmetric SAM seems to be associated with precipitation in a relatively reduced area of the East Coast (panel c.5) while the positive precipitation anomalies related with positive Asymmetric SAM affect all Estern Australia (panel b.5).
+In this trimester the Symmetric SAM seems to be associated with precipitation in a relatively reduced area of the East Coast (panel c.5) while the positive precipitation anomalies related with positive Asymmetric SAM affect all Eastern Australia (panel b.5).
 
 In Summer, positive Full SAM index is associated with with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (panel a.2).
-The Eastern part being dominated by the relationship with the Symmetric SAM and the Western, by the Asymemtric SAM.
+The Eastern part being dominated by the relationship with the Symmetric SAM and the Western, by the Asymmetric SAM.
 In Autumn, the regression with Full SAM shows positive values in the North, similar to Summer, and a broad area of positive values in the North-East to South-West direction.
@@
 In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the Symmetric SAM.
-None of these regression coefficients are statistically signifincat at the 95% level.
-The Spring signal is broadly consistent with @hendon2007, but whereas @hendon2007 also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coeffcients have the consistent sign).
+None of these regression coefficients are statistically significant at the 95% level.
+The Spring signal is broadly consistent with @hendon2007, but whereas @hendon2007 also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coefficients have the consistent sign).
 
@@
 In South America (Figure \ref{fig:pp-regr-america}), the annual-level regression shows that positive SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1).
-Panels b.1 and c.1 show a remarkably clean separation between the Asymmeric SAM --associated with the Southeastern South American and Southern Brazilian signals-- and the Symmetric SAM --associated with the signal in Southern Chile.
+Panels b.1 and c.1 show a remarkably clean separation between the Asymmetric SAM --associated with the Southeastern South American and Southern Brazilian signals-- and the Symmetric SAM --associated with the signal in Southern Chile.
 
@@
 The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous Easterly flow over Southern Chile.
-This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region.
+This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region [@garreaud2007].
 On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
@@
 This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events [@rosso2018].
-There is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 and that is explained by the Asymmmetric SAM.
-
-#LEAN: Podes agregar esta cita creo para la segunda oración https://journals.ametsoc.org/jcli/article/20/18/4789/31713
+There is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 and that is explained by the Asymmetric SAM.
 
@@
 
-#LEAN: Cambié el "tried" porque en la conclusión hay que ser un poco más afirmativo.
-
 As expected, the Asymmetric SAM index correlates strongly with the Symmetric SAM index.
 In the troposphere, this correlation is maximum at zero lag, while in the stratosphere is maximised with the Asymmetric SAM leading the Symmetric SAM by one month.
-Since most indices of the SAM are calculated using suftace or near-surface conditions, this result would suggest that they might not be sensitive to the most dramatic changes in SAM variability.
-
-The two-year periodcty we found in the stratospheric Symmetric SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
+Since most indices of the SAM are calculated using surface or near-surface conditions, this result would suggest that they might not be sensitive to the most dramatic changes in SAM variability.
+
+The two-year periodicity we found in the stratospheric Symmetric SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
 There is evidence of influence between the QBO and the Northern Annular Mode [e.g.
@@
 
-we observe a positive trend towards positive SAM in Summer and Autumn, As was documented by previous studies, such as @fogt2020 (and references therein) for surface levels.
+We observe a positive trend towards positive SAM in Summer and Autumn, As was documented by previous studies, such as @fogt2020 (and references therein) for surface levels.
 We show that these trends are maximised at the 100 hPa level and are explained by the zonally symmetric component.
@@
 
-In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being trully annular than the patterns associated with the Full SAM index.
+In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being truly annular than the patterns associated with the Full SAM index.
 The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific region and whose phase is rotated a quarter wavelength from the mean zonal wave 3 described by @raphael2004's index.
@@
 We found that the significant correlation that exists between the Full SAM index and the Oceanic Niño Index is captured entirely by the Asymmetric SAM index.
-This suggests that ENSO is linked to SAM exclusively through the variability in the latter's Asymmetric compoment.
+This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the Asymmetric SAM index could be a useful measure to further study that relationship.
 
 Temperature anomalies associated with the Full SAM broadly show a pattern of negative anomalies at polar latitudes surrounded by positive anomalies, but with many deviations from symmetry.
-The Asymemtric SAM index explains a big portion of these deviations.
+The Asymmetric SAM index explains a big portion of these deviations.
 In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship.
-These are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [@cazes-boezio2003; @fogt2011; @cai2020a].
-
-#LEAN: Acá también el "these" no me queda del todo claro
-
-In Australia the Full SAM is associated with positive precipitation anomalies in South East and this is explained by the Symemtric SAM.
+These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [@cazes-boezio2003; @fogt2011; @cai2020a].
+
+In Australia the Full SAM is associated with positive precipitation anomalies in South East and this is explained by the Symmetric SAM.
 However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
@@
 @silvestri2009 suggests that precipitation impacts linked to the SAM changed rather dramatically before and after 1980.
-In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in other in the earlier period.
-The correlation between ENSO and SAM is similarly non-stationary, also disapearing before 1973.
-Seeing as both the ENSO-SAM relationship and most of the precipitation imacts in South America are captured by the Asymmetric SAM, the results presented here are most likely period-dependent.
-
-#LEAN: Revisaste más bibliografía con respecto a la relación ENSO-SAM?
-
-#LEAN: Fatlaría un último parráfo de cierre.
-Lo que se me ocurre que se puede decir es lo positivo del estudio en cuanto a qué permite separar la señales simétricas/asimétricas, y ahi también ponerlas en contexto del cambio climático.
-También podes agregar alguna cosa que pueda quedar como futuras investigaciones.
+In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in others in the earlier period.
+The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s [@fogt2006;@clem2013].
+Seeing as both the ENSO-SAM relationship and most of the precipitation impacts in South America are captured by the Asymmetric SAM, the results presented here are most likely period-dependent.
+
+By succesfully separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
+This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component [@arblaster2006;@simpkins2012].
 
@@
 NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/
+
 The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
 
-\bibliography{AsymSAM}
+\bibliography{AsymSAM,packages}
 

### 2020-10-01 analysis/paper/paper.tex
segudno borrador

@@
 \correspondingauthor{Elio Campitelli,elio.campitelli@cima.fcen.uba.ar}
-and Leandro Díaz
+and Leandro B. Díaz
 }
@@
 pressure or geopotential height \citep{ho2012}. \citet{gong1999} defined
-the SAM index as the zonal mean sea level pressure differene between
+the SAM index as the zonal mean sea level pressure difference between
 40\degree S and 60\degree S, which is also the definition used by the
@@
 ability to reconstruct its historical variability prior to the
-availabilty of dense observations in the Southern Hemisphere
+availability of dense observations in the Southern Hemisphere
 \citep{jones2009}.
 
-At least some of the variabilty associated with the zonal asymmetries of
-the SAM is probably forced by the tropics. In particular, ENSO-like
+At least some of the variability associated with the zonal asymmetries
+of the SAM is probably forced by the tropics. In particular, ENSO-like
 variability affects the Southern Hemisphere extratopics through the
 Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989},
-whose wave train proyects strongly onto the zonal anomalies
+whose wave train protects strongly onto the zonal anomalies
 corresponding to the SAM in the Pacific sector. And although the
@@
 influences on the SAM have been observed
-\citep{fan2007, fogt2011, clem2013}.
+\citep{fan2007, fogt2011, clem2013}. In particular, \citet{fan2007}
+computed SAM indices of the Western and the Eastern Hemisphere
+separately and found that they were much more correlated if the (linear)
+signal of the ENSO was removed. While this relates to temporally
+coherent variability of the two hemisphere and not necessarily to zonal
+asymmetries in the associated spatial patterns, it is nonetheless
+consistent of the ENSO playing a crucial role in zonal asymmetries of
+the SAM.
 
@@
 of zonal mean variables. However, it's not clear how or if the
-asymmetric component responds to this forcing, or whether its variabilty
-could be masking influencing the observed trends independendly.
+asymmetric component responds to this forcing, or whether its
+variability could be masking influencing the observed trends
+independently.
 
@@
 generally colder temperatures over Antarctica and warmer temperatures at
-higher latitudes \citep{jones2019} (and viceversa for negative SAM), but
-there are significant deviations from this zonal mean response, notably
-in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}. The
-SAM signal in precipitation behaves similarly, although with even
-greater deviation from zonal symmetry \citep{lim2016}. The importance of
-zonal asymmetries of the SAM in these impacts have been studied in
-certain regions. For example, the SAM-precipitation relationship in
-Southeastern South America and Southern Brazil can be explained by the
-PSA-like zonally asymetric circulation associated with the SAM
-\citep{silvestry2009, rosso2018}. \citet{fan2007} also found that
-precipitation in East Asia was impacted by the variabilty of only the
-Western Hemisphere part of the SAM.
+higher latitudes \citep{jones2019} (and vice versa for negative SAM),
+but there are significant deviations from this zonal mean response,
+notably in the Antarctic Peninsula and the South Atlantic
+\citep{fogt2012}. The SAM signal in precipitation behaves similarly,
+although with even greater deviation from zonal symmetry
+\citep{lim2016}. The importance of zonal asymmetries of the SAM in these
+impacts have been studied in certain regions. For example, the
+SAM-precipitation relationship in Southeastern South America and
+Southern Brazil can be explained by the PSA-like zonally asymmetric
+circulation associated with the SAM \citep{silvestri2009, rosso2018}.
+\citet{fan2007} also found that precipitation in East Asia was impacted
+by the variability of only the Western Hemisphere part of the SAM.
 
@@
 from earlier than 1988). This is particularly important due to the
-inhomogeneities in renalysis products prior to the satellite era and the
-possible change in the asymmetric structure of the SAM
+inhomogeneities in reanalysis products prior to the satellite era and
+the possible change in the asymmetric structure of the SAM
 \citep{silvestri2009}.
@@
 \ref{sec:results}.\ref{sec:temporal} we describe the temporal
-variability and vertical coherence of the inidices. In Section
+variability and vertical coherence of the indices. In Section
 \ref{sec:results}.\ref{sec:spatial}, we analyse the spatial patterns of
@@
 \ref{sec:results}.\ref{sec:impacts}, we study their relationship with
-surface-level tempreature and precipitation.
+surface-level temperature and precipitation.
 
@@
 
-We describe the relationship between the SAM indices and temperature and
-precipitation. We use temperature data from NOAA's Merged Land Ocean
+To describe the relationship between the SAM indices and temperature and
+precipitation, we use temperature data from NOAA's Merged Land Ocean
 Global Surface Temperature Analysis V4.0.1 \citep{vose2012, smith2008},
 which blends land and ocean temperature analysis into a monthly global
-grid 5\degree~ longitude by 5\degree~ latitude For precipitation, we use
-monthly, 0.5\degree~ longitude by 0.5\degree~ latitude data from the
-Global Precipitation Climatology Centre \citep{schneider2015}.
+grid 5\degree~longitude by 5\degree~latitude, and monthly rainfall
+0.5\degree~longitude by 0.5\degree~latitude data from the Global
+Precipitation Climatology Centre \citep{schneider2015}.
 
@@
 Traditionally the Souther Annular Mode (SAM) is defined as de leading
-Empiral Orthogonal Mode (EOF) of sea level pressure or geopotential
+Empirical Orthogonal Mode (EOF) of sea level pressure or geopotential
 height at lower levels \citep{ho2012}. Following \citet{baldwin2001}, we
@@
 
-To separate between the zonally symmetric and asymmetric components of
-the SAM, we computed the zonal mean and anomalies of the full SAM
-spatial pattern, as shown in Figure~\ref{fig:method} for 700hPa. The
-full spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is the sum of
-the zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\)) and
-symmetric (\([\mathrm{EOF_1}](\lambda, \phi)\)) components. We then
-compute the ``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM''
-indices as the regression coefficients of the regression ofeach monthly
-geopotential field on the respective patterns (weighting by the cosine
-of latitude). The three indices are normalised by dividing them by the
-standard deviation of the ``Full'' index at each level. As a result, the
-magnitude between indices is comparable. However, only ``Full'' index
+To separate the zonally symmetric and asymmetric components of the SAM,
+we computed the zonal mean and anomalies of the full SAM spatial
+pattern, as shown in Figure \ref{fig:method} for 700 hPa. The full
+spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is the sum of the
+zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\)) and symmetric
+(\([\mathrm{EOF_1}](\lambda, \phi)\)) components. We then compute the
+``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM'' indices as the
+regression coefficients of the regression of each monthly geopotential
+field on the respective patterns (weighting by the cosine of latitude).
+The three indices are normalized by dividing them by the standard
+deviation of the ``Full'' index at each level. As a result, the
+magnitudes between indices are comparable. However, only ``Full'' index
 will have unit standard deviation per definition. From the regression,
-we also use the explained variance of each pattern as a indication of
-the degere of symmetry or asymmetry of each monthly field.
+we also use the explained variance of each pattern as an indicator of
+the degree of zonally symmetry or asymmetry of each monthly field.
 
 \begin{figure*}
-\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height]{Spatial patterns of the first EOF of 700 hPa geopotential height.
+\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period]{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 That is, we assume that zonal symmetries associated with positive SAM
-are oposite and equal to the ones associated with negatie SAM.
-\citet{fogt2012}'s composites (their Figure 4) suggest that this might
-not be entirely valid, although we argue that much of that apparent
-non-linearity is due to the heterogenous nature of the selected years
-for constructing the composites. Using our data (from 1979 to 2018),
-seasonal composites of zonal anomalies of 700 hPa geopotential height
-for SAM+ (Full SAM index greater than 1 standard deviation) and SAM-
-(smaller than negative 1 standard deviation) show relatively high
-pattern correlations all seasons and are visually very linear (Figure
-A9). Therefore, we belive that our method is at the very least a
+are almost opposite and of the same magnitude to the ones associated
+with negative SAM. \citet{fogt2012}'s composites (their Figure 4)
+suggest that this might not be entirely valid, although we argue that
+much of that apparent non-linearity is due to the heterogenous nature of
+the selected years for constructing the composites. Using our data (from
+1979 to 2018), seasonal composites of zonal anomalies of 700 hPa
+geopotential height for SAM+ (Full SAM index greater than 1 standard
+deviation) and SAM- (less than minus 1 standard deviation) show
+relatively high pattern correlations for all seasons and are visually
+very linear (Figure A9). Therefore, we believe that our method is a
 reasonable approximation of the phenomenon.
@@
 By computing a single EOF pattern using data for all months we are
-assuming that the zonal anomalies of the SAM are the same in all
-seasons. Geopotential zonal anomalies computed by projecting the first
-EOF \emph{of each season} are very similar to each other (Figure A10)
-and show pattern correlations between 0.65 (DJF with JJA) and 0.9
-(between MAM and SON). Based on this, we believe that our initial
-assumption is not unreasonable.
+assuming that the zonal anomalies of the SAM are the same in all seasons
+--December to February (DJF), March to May (MAM), June to August (JJA)
+and September to November (SON). Geopotential zonal anomalies computed
+by projecting the first EOF \emph{of each season} are very similar to
+each other (Figure A10) and show pattern correlations between 0.65 (DJF
+with JJA) and 0.9 (between MAM and SON). Based on this, we believe that
+our initial assumption is not unreasonable.
 
@@
 
-We perform linear regression to quantify the association between the SAM
-indices and other variables. Since the Asymmetric and Symmetric SAM
+We perform linear regressions to quantify the association between the
+SAM indices and other variables. Since the Asymmetric and Symmetric SAM
 indices are significantly correlated with each other, to capture the
@@
 precipitation, etc\ldots{}) with the Asymmetric SAM (\(SAM_a\)) and
-Symmetric SAM (\(SAM_s\)) we fit the ecuation
+Symmetric SAM (\(SAM_s\)) we fit the equation
 
@@
 where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is
-the time, \(\alpha\) and \(\beta\) are the linear coefficients, \(X_0\)
-and \(\epsilon\) are the constant and error terms. From this equarion,
-\(\alpha\) represents the (linear) association of \(X\) with the
-variability of the Asymmetric SAM that is not explained by the
-variability of the Symmetric SAM; in other words, it is proportional to
-the partial correlation of \(X\) and the Asymmetric SAM, controlling for
-the effect of the Symmetric SAM and viceversa for \(\beta\).
+the time, \(\alpha\) and \(\beta\) are the linear regression
+coefficients, \(X_0\) and \(\epsilon\) are the constant and error terms.
+From this equation, \(\alpha\) represents the (linear) association of
+\(X\) with the variability of the Asymmetric SAM that is not explained
+by the variability of the Symmetric SAM; i.e.~it is proportional to the
+partial correlation of \(X\) and the Asymmetric SAM, controlling for the
+effect of the Symmetric SAM and vice versa for \(\beta\). When
+performing a separate regression for each trimester (DJF, MAM, JJA,
+SON), we average of the relevant variables seasonally for each year and
+trimester before computing the regression.
 
 At 2.5\degree by 2.5\degree resolution, a single regression field is
-composed of thousands of regressions. In such case, using naive p-values
-to test for significance leads to misleading results
+composed of thousands of regressions. In such case, using p-values to
+test for significance leads to misleading results
 \citep{walker1914, katz1991}. While there are multiple proposed
-solutions in the literature, \citet{wilks2016} suggests that adjunting
+solutions in the literature, \citet{wilks2016} suggests that adjusting
 p-values by controlling for the False Discovery Rate
@@
 
-When performing a separate regression for each trimester (DJF, MAM, JJA,
-SON) we first average the relevant variables to obtain a single value
-for each year and each trimester.
+We computed linear trends by Ordinary Least Squares and the 95\%
+confidence interval assuming a t-distribution of the appropriate
+residual degrees of freedom.
+
+To compute the amplitude of the zonal waves we computed the Fourier
+transform of the spatial field at each latitude circle.
+
+\subsubsection{Computation procedures}
+
+We performed all analysis in this paper using the R programming language
+\citep{rcoreteam2020}, using the data.table package \citep{dowle2020}
+and the metR package \citep{campitelli2020}. All graphics are made using
+ggplot2 \citep{wickham2009}. We downloaded data from ranalysis using the
+ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi
+package \citep{albers2020}. The paper was rendered using knitr and
+rmarkdown \citep{xie2015, allaire2019}.
+
+A version-controlled repository of the code used to create this analysis
+can be found at https://github.com/eliocamp/asymsam.
 
@@
 \includegraphics{asymsam-timeseries-1} \caption[Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa]{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
-To the left, probability density estimate of each index.
+To the left, probability density estimate of each index computed using a gaussian kernel of optimal bindwidth according to \cite{sheather1991}.
 Series are standarised by the standard deviation of the Full SAM at each level.}\label{fig:asymsam-timeseries}
@@
 
-The temporal evolution of the Assymmetric and Symmetric SAM was firstly
-asssesed. Figure \ref{fig:asymsam-timeseries} shows the corresponding
-time series for 700 hPa and 50 hPa and their corresponding density
-estimates. We selected these two levels as representative of the
+We first asses the temporal evolution of the Asymmetric SAM and
+Symmetric SAM. Figure \ref{fig:asymsam-timeseries} shows the
+corresponding time series for 700 hPa and 50 hPa and their corresponding
+density estimates. We selected these two levels as representative of the
 tropospheric and stratospheric variability respectively. As will be
@@
 variations in the low frequency. At first glance the series can be
-distinguished by their distributions. Compared to the stratospheric
+distinguished by their distributions. Compared to the tropospheric
 indices, the stratospheric indices are much more long-tailed; that is,
-extreme values (both negative and positive) abound. The Asymmetric
+extreme values (both negative and positive) abound. The Asymmetric SAM
 series have both more variability in the higher frequencies than the
-Symmetric series.
+Symmetric SAM series.
 
@@
 which can be seen by spectral analysis (Figure A3). This might suggests
-a link between stratospheric SAM variability and QBO. There is a local
-peak at 2 years in the periodigram of the tropospheric Symmetric SAM
-also, although it's not statistically significant. In the troposphere
-the most significant peak of variability is found in the Asymmetric
-index at around 3.6 months.
+a link between stratospheric SAM variability and the Quasi-Biennial
+Oscillation \citep{baldwin2001b}. There is a local peak at 2 years in
+the periodogram of the tropospheric Symmetric SAM also, although it's
+not statistically significant. In the troposphere the most significant
+peak of variability is found in the Asymmetric index at around 3.6
+months.
 
-From Figure \ref{fig:asymsam-timeseries} we can see that the Asymemtric
-and Symmetric time series appear to be correlated. Moreover, looking at
-the extremes in the stratosphere, the Symmetric serie appears to lag the
-Asymmetric series (see, for example, the positive events on late 1987).
-We show these correlations, across all the levels of the reanalysis for
-zero and -1 lag (Asymmetric index leading the Symmetric index), in
-Figure \ref{fig:cor-lev}. Zero-lag correlations between the Asymmetric
-and Symmetric series are relatively constant throught the troposphere,
-fluctiating between 0.39 and 0.45. One-month-lag correlations are
-similarly constant but significantly reduced to around 0.17. In the
-stratosphere, zero-lag correlations drop to a minimum of 0.21 at 20 hPa
-and then it increases again monotonically with height up to the
-uppermost level of the reanalysis (although results near the top of the
-models are to be interpreted with care). At the same time, one-month-lag
-correlations increase with height. As a consequence, statospheric
-Symmetric index tend to precede corresponding Asymmetric index.
+From Figure \ref{fig:asymsam-timeseries} we can see that the Asymmetric
+SAM and Symmetric SAM time series appear to be correlated. Moreover,
+looking at the extremes in the stratosphere, the Symmetric SAM series
+appears to lag the Asymmetric SAM series (see, for example, the positive
+events on late 1987). We show these correlations, across all the levels
+of the reanalysis for zero and -1 lag (Asymmetric SAM index leading the
+Symmetric SAM index), in Figure \ref{fig:cor-lev}. Zero-lag correlations
+between the Asymmetric SAM and Symmetric SAM series are relatively
+constant throughout the troposphere, fluctuating between 0.39 and 0.45.
+One-month-lag correlations are similarly constant but significantly
+reduced to around 0.17. In the stratosphere, zero-lag correlations drop
+to a minimum of 0.21 at 20 hPa and then it increases again monotonically
+with height up to the uppermost level of the reanalysis (although
+results near the top of the models are to be interpreted with care). At
+the same time, one-month-lag correlations increase with height. As a
+consequence, stratospheric Asymmetric SAM index tend to precede
+corresponding Symmetric SAM index.
 
 \begin{figure}
-\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Symmetric leads Asymmetric)]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Symmetric leads Asymmetric).}\label{fig:cor-lev}
+\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979–2018 period]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979–2018 period.}\label{fig:cor-lev}
 \end{figure}
@@
 \begin{figure*}
-\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM.}\label{fig:cross-correlation}
+\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979–2018 period]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979–2018 period.}\label{fig:cross-correlation}
 \end{figure*}
@@
 the Full SAM (panel a), high values below 100 hPa reflect the vertical
-(zero-lag) coherency throughout the troposfere. Above 100 hPa
+(zero-lag) coherency throughout the troposphere. Above 100 hPa
 correlation between levels falls off more rapidly, indicating less
@@
 correlation between the troposphere and the lower-to-middle
-stratosphere. Examining panels b and c, we see that the Asymemtric and
+stratosphere. Examining panels b and c, we see that the Asymmetric and
 Symmetric SAM share the same high level of coherency in the troposphere
 but they differ in their stratospheric behaviour. Stratospheric
-coherency is stronger for the Asymmetric SAM than the Symemtric SAM. The
+coherency is stronger for the Asymmetric SAM than the Symmetric SAM. The
 stratospheric Symmetric SAM seems to connect more strongly to the
-trosposphere than the Asymmetric SAM.
+troposphere than the Asymmetric SAM.
 
 \begin{figure*}
-\includegraphics{trends-1} \caption[Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
-Shading indicates the 95\% confidence interval.}\label{fig:trends}
+\includegraphics{trends-1} \caption[Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+Shading indicates the 95\% confidence interval from a t-distribution.}\label{fig:trends}
 \end{figure*}
 
-The trends for each of the indices (Full, Symmetric, Assymetric) were
-evaluate for the whole period 1979-2018 at each level (Figure
-\ref{fig:trends}) for the whole year and separed by trimesters. The Full
-SAM index presents a statistically significant trend (panel a.1) that
-extends throught the troposphere up to about 50 hPa and reaches its
-maximum value at 100 hPa. The seasonal trends (rest of column a)
-indicate that positive trends are present in Autumn and particularly in
-in Summer, where the 100 hPa maximum is much more defined. Positive
-trends have been documented by previous studies \citep[e.g.][ and
-references therein]{fogt2020} using indices of the SAM based on surface
-or near-surface circulation.
+The linear trends for each of the indices (Full SAM, Symmetric SAM and
+Asymmetric SAM) were evaluated for the complete period 1979-2018 at each
+level (Figure \ref{fig:trends}) for the whole year and separated by
+trimesters. The Full SAM index presents a statistically significant
+trend (panel a.1) that extends throughout the troposphere up to about 50
+hPa and reaches its maximum value at 100 hPa. The seasonal trends (rest
+of column a) indicate that positive trends are present in Autumn and
+particularly in Summer, where the 100 hPa maximum is much more defined.
+In Winter and Spring, we detect no statistically significant trend. This
+is consistent with the results of previous studies, which find clear
+positive trends in Summer, weaker in Autumn and no trends in the other
+seasons \citep[e.g.][ and references therein]{fogt2020} using indices of
+the SAM based on surface or near-surface circulation.
 
-By separating the SAM signal in its Asymmetric and Symmetric parts, we
+By separating the SAM signal in its asymmetric and symmetric parts, we
 can not only see that these trends are almost entirely due to the
-Symmetric component (colum b vs.~column c), but in some cases the trends
-become more clear. In Summer, the Asymmetric SAM has a statistically non
-significant negative trend in the middle troposphere that obscures the
-trend in the Full index; as a result, trends computed using only the
-Symmetric component are more clear (compare the shading region in panel
-a.2 and c.2). In Autumn, using the Symmetric SAM reveals a statistically
-significant positive trend in the stratosphere that is not significant
-using the Full index.
+symmetric component (column b vs.~column c), but in some cases the
+trends become more clear. In Summer, the Asymmetric SAM has a
+statistically non significant negative trend in the middle troposphere
+that obscures the trend in the Full SAM index; as a result, trends
+computed using only the Symmetric component are more clear (compare the
+shading region in panel a.2 and c.2). In Autumn, the Symmetric SAM
+reveals a statistically significant positive trend in the stratosphere
+that is not significant using the Full SAM index.
 
@@
 the absence of a statistically significant signal should not be taken as
-evidence of no sistematic change. In particular, going back to Figure
+evidence of no systematic change. In particular, going back to Figure
 \ref{fig:asymsam-timeseries}, we can see an evident change in the
-stratospheric Asymemtric component (red line in panel a) between the
+stratospheric Asymmetric component (red line in panel a) between the
 90's, when we see a dominance of extreme negative values, and the 00's,
 when we see the inverse. This change is restricted to the Winter months:
-the linear trend for JJA starting in 1990 for the Asymmetric component
-at 50hPa is \(0.37 \pm 0.22\).
+the linear trend for Winter starting in 1990 for the Asymmetric
+component at 50hPa is \(0.37 \pm 0.22\).
 
 Figure \ref{fig:r-squared-trend} shows decadal trends for the explained
-variance of each index. There is no evidencie of a significant trend in
+variance of each index. There is no evidence of a significant trend in
 the stratosphere. In the troposphere, there is a positive trend for the
-Asymmetric SAM and no significant trend for the Symmetric SAM. This
+Asymmetric SAM and not significant trend for the Symmetric SAM. This
 suggest that the SAM has become more asymmetric in the period from 1979
-to 2018. The change is slight, though; of the order of 1\% icreased
-explained variance per decade.
+to 2018. However, the change is slight, around 1\% increased explained
+variance per decade.
 
@@
 \begin{figure*}
-\includegraphics{2d-regr-1} \caption[Regression patterns of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM]{Regression patterns of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices, not of two simple regressions involving each index by itsef.
+\includegraphics{2d-regr-1} \caption[Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979–2018 period]{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979–2018 period.
+The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices.
 Points marked on panel b.2 are the location of the reference points used by \cite{raphael2004} for their Zonal Wave 3 index.
@@
 
-To show if, and to what extent, the Asymmetric and Symmetric SAM
-inidices indeed capture the asymmetric and symmetric compoment of the
-SAM respectively, we computed the spatial regression of geopotential
-height anomalies on these indices and the Full SAM index. Figure
-\ref{fig:2d-regr} shows these regressions. Regression coefficients in
-column a are computed using the Full SAM. Regression coefficients in
-columns b and c are computed using multiple regression using the
-Asymmetric and Symmetric indices at the same time. Thus, they are to be
-interpreted as the patterns associated with each index, removing the
-variability (linearly) explained by the other index.
+To show if, and to what extent, the Asymmetric and Symmetric SAM indices
+indeed capture the asymmetric and symmetric component of the SAM
+respectively, we computed the spatial regression of geopotential height
+anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa
+levels. Figure \ref{fig:2d-regr} shows these regressions. Regression
+coefficients in column a are computed using the Full SAM. Regression
+coefficients in columns b and c are computed using multiple regression
+using the Asymmetric and Symmetric indices at the same time. Thus, they
+are to be interpreted as the patterns associated with each index,
+removing the variability (linearly) explained by the other index.
 
@@
 more clearly dominated by a zonally symmetric, monopolar structure
-(panel a.1) which is, however, not perfectly centered in the South Pole.
-The monopole obtained by multiple regression with the Asymmetric and
-Symmetric SAM (panel c.1) is much more symmetric and the shift from
-total symmetry is captured by the regression pattern of the Asymmetric
-SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the
-Western Hemisphere and and Davids Sea in the Eastern Hemisphere (panel
-b.1).
+(panel a.1) which is, however, not perfectly centred in the South Pole.
+The monopole obtained by the regression pattern for Symmetric SAM (panel
+c.1) is much more symmetric and the shift from total symmetry is
+captured by the regression pattern of the Asymmetric SAM as a wave-1
+with maximum anomalies above the Belinghausen Sea on the Western
+Hemisphere and and Davis Sea in the Eastern Hemisphere (panel b.1).
 
@@
 zonally symmetrical annular mode with zonal asymmetries in the form of a
-wave-3. The regression using the Asymmetric and Symmetric SAM indices
-successfully disentangle both structures. The Asymmetric component gives
-rise to a cleaner zonal wave (panel b.2) and the Symemtric component is
-assosiated with an trully annular mode, almost devoid of zonal
-asymmetries (panel c.2). The wave-3 pattern observed in panel b.2 is
-rotated by half a wavelength from the average position of the mean
-wave-3 pattern asociated with \citet{raphael2004}'s ZW3 index, whose
-reference locations are marked with points in the figure. Thus, the
-tropospheric Asymmetric SAM index represents a zonal displacement in the
-position of the climatological wave-3 pattern.
+wave-3 \citep{fogt2012}. The regression using the Asymmetric and
+Symmetric SAM indices successfully disentangle both structures. The
+Asymmetric SAM index gives rise to a cleaner zonal wave (panel b.2) and
+the Symmetric SAM index is associated with an annular mode, almost
+devoid of zonal asymmetries (panel c.2). The wave-3 pattern observed in
+panel b.2 is rotated by half a wavelength from the average position of
+the mean wave-3 pattern associated with \citet{raphael2004}'s ZW3 index,
+whose reference locations are marked with points in the figure. Thus,
+the tropospheric Asymmetric SAM index represents a zonal displacement in
+the position of the climatological wave-3 pattern.
+
+\begin{figure*}
+\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
+\end{figure*}
+
+The amplitude of the first zonal wave numbers at each latitude at 50 hPa
+and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave
+number zero represents the amplitude of the zonal mean. Zonal wave
+amplitudes of the spatial pattern described by the Full SAM index
+(column a) are dominated by the zonal mean (wave-number 0) at both
+levels. However, zonal waves are important, particularly North of
+50\degree S, with wave-number 1 clearly dominating at 50 hPa (panel a.1)
+and a more equal mix of waves at 700 hPa (panel a.2). Column b shows
+that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the
+stratosphere (panel b.1), while in the troposphere it is composed of
+zonal waves 3 to 1 in decreasing level of importance (panel b.2) with
+negligible amplitude of the zonal mean. The Symmetric SAM, on the other
+hand, it's almost entirely composed of zonal mean at both levels (column
+c), with little to now contribution from zonal waves with wave-numbers 1
+to 3.
+
+We can see that the amplitude and latitudinal distribution zonal waves
+in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric
+SAM on the other correspond almost exactly to the amplitude and
+latitudinal distribution in the Full SAM. This confirms the correct
+decomposition of the SAM in its symmetric and asymmetric components.
+
+Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent
+that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which
+--as mentioned before-- is larger in the Western Hemisphere than in the
+Eastern Hemisphere.
 
 \begin{figure}
-\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.
-Note the differnet x axis.}\label{fig:wave-amplitude}
+\includegraphics{vertical-regression-1} \caption[Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM)]{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
+(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979–2018 period.}\label{fig:vertical-regression}
 \end{figure}
 
-The amplitude of first zonal wave numbers at each latitude at 50 hPa and
-700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number
-zero represents the amplitude of the zonal mean. Column b shows that the
-Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere
-(panel b), while in the troposphere it is composed of zonal waves 3 to 1
-in decreasing level of importance (panel b). Looking at panel b.2 from
-Figure \ref{fig:2d-regr}, it becomes apparent that zonal wavess 1 and 2
-modulate the amplitude of zonal wave 3, which --as mentioned before-- is
-larger in the Western Hemisphere than in the Easten Hemisphere.
-
-\begin{figure}
-\includegraphics{vertical-regression-1} \caption[Regression between monthly geopotential anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM)]{Regression between monthly geopotential anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
-(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa.}\label{fig:vertical-regression}
-\end{figure}
-
-To analyse the vertical structure of the geopotential anomalies
-asociated with the asymetric SAM index, we show a vertical cross section
-of regressions of mean geopotential height between 65\degree S and
-40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the
-700 hPa Asymmetric SAM index (panel b) (Figure
-\ref{fig:vertical-regression}). The geopotential anomalies associated
-with the stratospheric Asymmetric SAM (panel a) are clearly constrained
-to the stratosphere, which underscores the uncoupling between the
-stratospheric and tropospheric Asymmetric SAM. The vertical structure of
-this signal tilts about 60\degree to the West between 100 hPa and 1 hPa,
-suggesting baroclinic processes. Interestingly, the signal in the
-stratosphere maximises near 10 hPa despite using the 50 hPa index for
-the regression.
+To analyse the vertical structure of the geopotential height anomalies
+associated with the asymmetric SAM index, we show a vertical cross
+section of regressions of mean geopotential height between 65\degree S
+and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for
+the 700 hPa Asymmetric SAM index (panel b) (Figure
+\ref{fig:vertical-regression}). The geopotential height anomalies
+associated with the stratospheric Asymmetric SAM (panel a) are clearly
+constrained to the stratosphere, which underscores the uncoupling
+between the stratospheric and tropospheric Asymmetric SAM. The vertical
+structure of this signal tilts about 60\degree to the West between 100
+hPa and 1 hPa, suggesting baroclinic processes. Interestingly, the
+signal in the stratosphere maximises near 10 hPa despite using the 50
+hPa index for the regression.
 
@@
 Eastern hemisphere the wave-3 signal is weaker and confined to the
-troosphere while negative anomalies dominate in the stratosphere. So,
+troposphere while negative anomalies dominate in the stratosphere. So,
 while the tropospheric Asymmetric SAM index is associated with
 stratospheric geopotential anomalies, these do not project strongly onto
-the stratospheric Asymmetric SAM.
-
-The structures shown in panels a and b in Figure
-\ref{fig:vertical-regression} are surprisignly robust to the choice of
+the stratospheric Asymmetric SAM. The structures shown in panels a and b
+in Figure \ref{fig:vertical-regression} are robust to the choice of
 index level. For any stratospheric (above 100 hPa) index, the resulting
@@
 the result is very similar to panel b. The patterns mainly change in
-amplitude.
+amplitude (not shown).
 
@@
 significant relationship between the SAM and the ENSO. The correlation
-between the full SAM and the ENSO as measured by the Oceanic Niño Index
-\citep{bamston1997} (ONI) is -0.16. Consistent with \citet{fan2007}, we
-show that this relationship is captured entirely the Asymmetric SAM, as
-this index has a partial correlation of -0.26 with the ONI controlling
-for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial
-correlation with the ONI is essentially null (0.019). We performed the
-same analysis using the Multivariate Enso Index \citep{wolter2011} and
-the Southern Oscillation Index \citep{ropelewski1987} to conclude that
-these results do not depend on the ENSO index used.
+between the Full SAM and the ENSO as measured by the Oceanic Niño Index
+\citep{bamston1997} (ONI) is -0.16 (p-value =
+\ensuremath{2.8\times 10^{-4}}). Consistent with \citet{fan2007}, we
+show that this relationship is captured entirely by the Asymmetric SAM,
+as this index has a partial correlation of -0.26 (p-value =
+\ensuremath{6.3\times 10^{-9}}) with the ONI controlling for the effect
+of the Symmetric SAM, whereas the Symmetric SAM's partial correlation
+with the ONI is essentially null (0.019; p-value = 0.67). We performed
+the same analysis using the Multivariate ENSO Index \citep{wolter2011}
+and the Southern Oscillation Index \citep{ropelewski1987} to conclude
+that these results do not depend on the ENSO index used.
 
@@
 \begin{figure*}
-\includegraphics{regr-air-season-1} \caption[Regression pattern of season mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM]{Regression pattern of season mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM.
+\includegraphics{regr-air-season-1} \caption[Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979–2018 period]{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979–2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 
-The SAM has been shown to be associated with important surface variables
-such as temperature and precipitacion \citep[e.g.][and see
-\citet{fogt2020} for a review]{gillett2006}. Naturally, most studies on
-the surface impacts of the SAM are based on an index identical or
-analogous to what we call Full SAM index (\citet{fogt2012} being the
-only exception that we are aware of). We regress surface temperature and
-precipitation onto each of the three SAM indices to see if there are
-diferent surface impact associated with the asymmetric and symmetric SAM
-circulation.
+To see if there are different surface impacts associated with the
+asymmetric and symmetric SAM circulation we regress surface temperature
+and precipitation onto each of the three SAM indices at 700 hPa. As
+shown in previous sections, the three indices are highly coherent in the
+troposphere, so we select this level to represent the tropospheric
+circulation for consistency with previous studies.
 
 Figure \ref{fig:regr-air-season} shows regression coefficients of each
-index at 700 hPa with surface temperature for each trimester. In Summer
+index at 700 hPa with surface temperature for each trimester. In Summer,
 positive values of the Full SAM index (panel a.1) are associated with
@@
 ring of positive anomalies. The ring is not zonally symmetric, as there
-are three clear local maximums around 30\degree W, 15\degree E and
-50\degree E and a local minimum (with negative sign) around
-120\degree W. In the tropics, there are negative anomalies in the
-equatorial Pacific, consistent with the negative correlation between SAM
-and ENSO. Panels b.1 and c.1 show temperature anomalies associated with
-positive values of the Asymmetric and Symmetric SAM, respectively. Both
-the local maximums in the ring and the anomalies in the Pacific regions
-are present mostly on the Asymetric SAM regression map, while
-temperature patterns linked to positive Symmetric SAM show a more
-zonally consistent ring and less relation to the tropics. Noticeable,
-temperature anomalies in the Indian ocean, South Africa and Australia
-are strongly related to positive values of Asymmetric SAM. This signal
-is not present in the regression pattern with the Full SAM. Spring (row
-4) features very similar patterns but of generally smaller in magnitude
-and statistical significance.
+are four clear local maximums around 30\degree W, 120\degree W,
+150\degree E and 90\degree E. In the tropics, there are negative
+anomalies in the equatorial Pacific, consistent with the negative
+correlation between SAM and ENSO. Panels b.1 and c.1 show temperature
+anomalies associated with positive values of the Asymmetric and
+Symmetric SAM, respectively. Both the local maximums in the ring and the
+anomalies in the Pacific regions are present mostly on the Asymmetric
+SAM regression map, while temperature patterns linked to positive
+Symmetric SAM show a more zonally consistent ring and less relation to
+the tropics. Noticeable, temperature anomalies in the Indian ocean,
+South Africa and Australia are strongly related to Asymmetric SAM. This
+signal is not present in the regression pattern with the Full SAM.
+Spring (row 4) features similar patterns but of smaller magnitude, with
+less regions where regressed anomalies have statistical significance.
 
-In Autumn and Winter (rows 2 and 3) the postive ring is only present
+In Autumn and Winter (rows 2 and 3) the positive ring is only present
 through its local maximums in the regression with the Full SAM. There
@@
 However, repeating this analysis with 2-meter temperature from ERA5
-resulted in similar patterns that were statistically significant.
-Moreover, similar features were observed in station measurements by
-\citet{jones2019}, although using data from 1957 to 2016.
+resulted in similar patterns that were statistically significant (not
+shown). Moreover, similar features were observed in station measurements
+by \citet{jones2019}, although using data from 1957 to 2016.
 
@@
 anomalies roughly seen in all seasons --although with varying intensity
-and small-scale details-- is consistent with the intensification and
-poleward migration of the westerlies commonly linked to the SAM. It's
-then not surprising to see it more clearly in association with the
-Symemtric SAM (at least in Summer and Spring).
-
-These results suggests that Asymmetric and Symmetric SAM indices are
-associated with overall distinct temperature patterns which may not be
-apparent when using the Full SAM index.
+and small-scale details-- translates to enhanced meridional temperature
+gradient maximised in the zero line, which is consistent with the
+intensification and poleward migration of the westerlies commonly linked
+to the SAM through thermal wind balance. It's then not surprising to see
+it more clearly in association with the Symmetric SAM (at least in
+Summer and Spring).
 
@@
 Figure 11 from \citet{fogt2012}. Although they used station data from
-1958 to 2001, a lot of the characteristics are reproduced here, such as
-the strong signal in New Zealand and Australia in Summer and Spring.
+1958 to 2001, main features are reproduced here, such as the strong
+signal in New Zealand and Australia in Summer and Spring.
 
-Regression of the SAM indicies with seasonal mean precipitation and 700
-hPa geopotential height are shown Figures \ref{fig:pp-regr-oceania} and
-\ref{fig:pp-regr-america} for Australasia and South America
+Regression of the SAM indices with seasonal mean precipitation and 700
+hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania}
+and \ref{fig:pp-regr-america} for Australasia and South America
 respectively. South Africa is not shown because no significant signal
@@
 \begin{figure*}
-\includegraphics{pp-regr-oceania-1} \caption[Regression pattern of (row 1) annual and (rows 2 to 5) season mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM]{Regression pattern of (row 1) annual and (rows 2 to 5) season mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM.
-thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
+\includegraphics{pp-regr-oceania-1} \caption[Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979–2018 period]{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979–2018 period.
+Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
 \end{figure*}
@@
 
-This Spring signal is broadly consistent with \citet{hendon2007}, but
-whereas \citet{hendon2007} also detected a strong signal in Summer,
-panel a.2 shows no statistically significant association (although the
-coeffcients have the consistent sign).
-
 The seasonal-level regressions show statistically significant anomalies
@@
 positive precipitation anomalies related with positive Asymmetric SAM
-affect all Estern Australia (panel b.5).
+affect all Eastern Australia (panel b.5).
 
@@
 in the North East (panel a.2). The Eastern part being dominated by the
-relationship with the Symmetric SAM and the Western, by the Asymemtric
+relationship with the Symmetric SAM and the Western, by the Asymmetric
 SAM. In Autumn, the regression with Full SAM shows positive values in
@@
 present only in relation with the Symmetric SAM. None of these
-regression coefficients are statistically signifincat at the 95\% level
+regression coefficients are statistically significant at the 95\% level.
+The Spring signal is broadly consistent with \citet{hendon2007}, but
+whereas \citet{hendon2007} also detected a strong signal in Summer,
+panel a.2 shows no statistically significant association (although the
+coefficients have the consistent sign).
 
@@
 In South America (Figure \ref{fig:pp-regr-america}), the annual-level
-regression shows that the SAM is associated with statistically
+regression shows that positive SAM is associated with statistically
 significant precipitation decrease in Southeastern South America (SESA)
@@
 the South Atlantic Convergence Zone (SACZ) (panel a.1). Panels b.1 and
-c.1 show a remarkably clean separation between the Asymmeric SAM
+c.1 show a remarkably clean separation between the Asymmetric SAM
 --associated with the Southeastern South American and Southern Brazilian
@@
 leads to reduced influx of moist air from the Pacific Ocean which, is
-the main source of precipitable water in that region. On the other hand,
-the anomalous circulation associated with positive values of Asymmetric
-SAM (panel b.1) in the Atlantic is anticyclonic in the South and
-cyclonic in the North. This creates anomalous South-Easterly flow over
-Southeastern South America, which inhibits the flow of the Low Level Jet
-to the region \citep{silvestri2009, zamboni2010}. This same pattern was
-found to be associated with increased precipitation in Southern Brazil
-during South Atlantic Convergence Zone events \citep{rosso2018}.
-
-There is a small area of increased precipitation with SAM near central
-Argentina which is also present in the station-based analysis by
-\citet{gillett2006} and that is explained by the Asymmmetric SAM.
+the main source of precipitable water in that region
+\citep{garreaud2007}. On the other hand, the anomalous circulation
+associated with positive values of Asymmetric SAM (panel b.1) in the
+Atlantic is anticyclonic in the South and cyclonic in the North. This
+creates anomalous South-Easterly flow over Southeastern South America,
+which inhibits the flow of the Low Level Jet to the region
+\citep[\citet{zamboni2010}]{silvestri2009}. This same pattern was found
+to be associated with increased precipitation in Southern Brazil during
+South Atlantic Convergence Zone events \citep{rosso2018}. There is a
+small area of increased precipitation with SAM near central Argentina
+which is also present in the station-based analysis by
+\citet{gillett2006} and that is explained by the Asymmetric SAM.
 
@@
 
-In this study we tried to systematically characterise the variability of
-the zonally symmetric and zonally asymmetric structure of the SAM. By
-projecting monthly geopotential fields at each level with the
-corresponding asymmetric and symmetric pattern, we created two indices
-representing the zonally asymmetric and zonally symmetric contributions
-of the SAM respectively.
+In this study we characterise the temporal and spatial variability of
+the zonally symmetric and asymmetric structure of the SAM. By projecting
+monthly geopotential fields at each level with the corresponding
+asymmetric and symmetric pattern, we created two indices for
+representing the zonally asymmetric and symmetric contributions of the
+SAM respectively.
 
@@
 leading the Symmetric SAM by one month. Since most indices of the SAM
-are calculated using suftace or near-surface conditions, this result
+are calculated using surface or near-surface conditions, this result
 would suggest that they might not be sensitive to the most dramatic
@@
 
-The two-year periodcty we found in the stratospheric Symmetric SAM might
-point to a link between the SAM and the Quasi Biennial Oscillation.
-There is evidence of influence between the QBO and the Northern Annular
-Mode \citep[e.g.][]{holton1980, watson2014, zhang2020}, so it's not
-unlikely that the SAM would be similarly connected. However establishing
-this link would require further research.
+The two-year periodicity we found in the stratospheric Symmetric SAM
+might point to a link between the SAM and the Quasi Biennial
+Oscillation. There is evidence of influence between the QBO and the
+Northern Annular Mode \citep[e.g.][\citet{watson2014},
+\citet{zhang2020}]{holton1980}, so it's not unlikely that the SAM would
+be similarly connected. However establishing this link would require
+further research.
 
-As documented by previous studies, such as \citet{fogt2020} (and
-references therein), we observe a positive trend towards positive SAM in
-Summer and Autumn. We show that these trends are maximised at the 100
-hPa level and are explained by the zonally symmetric component. We also
-find a statistically significant positive trend in the Symmetric
-component of the SAM in the stratosphere that is not apparent in the
-Full SAM index. In contrast to \citet{fogt2012} we find some evidence of
-the SAM becoming more zonally asymmetric, as there is a slight positive
-trend in the variance explained by the as the Asymmetric SAM explains an
-increasingly proportion of the total variance.
+We observe a positive trend towards positive SAM in Summer and Autumn,
+As was documented by previous studies, such as \citet{fogt2020} (and
+references therein) for surface levels. We show that these trends are
+maximised at the 100 hPa level and are explained by the zonally
+symmetric component. We also find a statistically significant positive
+trend in the Symmetric component of the SAM in the stratosphere that is
+not apparent in the Full SAM index. In contrast to \citet{fogt2012} we
+find some evidence of the SAM becoming more zonally asymmetric, as there
+is a slight positive trend in the variance explained by the as the
+Asymmetric SAM explains an increasingly proportion of the total
+variance.
 
 In the troposphere, the spatial patterns of geopotential associated with
-the Symmetric SAM is much closer to being trully annular than the
+the Symmetric SAM are much closer to being truly annular than the
 patterns associated with the Full SAM index. The Asymmetric SAM, on the
 other hand, describes a wave-3 pattern with maximum amplitude in the
-Pacific region and whose phae is rotated a quarter wavelength from the
+Pacific region and whose phase is rotated a quarter wavelength from the
 mean zonal wave 3 described by \citet{raphael2004}'s index. This pattern
@@
 index. This suggests that ENSO is linked to SAM exclusively through the
-variability in the latter's Asymmetric compoment.
+variability in the latter's asymmetric component and thus, the
+Asymmetric SAM index could be a useful measure to further study that
+relationship.
 
@@
 pattern of negative anomalies at polar latitudes surrounded by positive
-anomalies, but with many deviations from symmetry. The Asymemtric SAM
+anomalies, but with many deviations from symmetry. The Asymmetric SAM
 index explains a big portion of these deviations. In particular, the
@@
 as well as the negative anomalies in the equatorial Pacific consistent
-with the ENSO-SAM relationship delineated above. These are particularly
-clear in the DJF and SON trimesters, which include the months in which
-the ENSO teleconnection is more active
+with the ENSO-SAM relationship. These negative anomalies are
+particularly clear in the DJF and SON trimesters, which include the
+months in which the ENSO teleconnection is more active
 \citep{cazes-boezio2003, fogt2011, cai2020a}.
@@
 In Australia the Full SAM is associated with positive precipitation
-anomalies in South East and this is explained by the Symemtric SAM.
+anomalies in South East and this is explained by the Symmetric SAM.
 However, the Asymmetric SAM is associated with a small area of positive
 precipitation anomalies in the Eastern Coast of West Australia, maybe
-due to advection of moist air from the Indian Ocean.
-
-In South America, precipitation anomalies associated with the Full SAM
-are negative both in Southern Chile and Southeastern South America, and
-positive in Southern Brazil. This features are cleanly separated between
-the Asymmetric and Symmetric components. The Symmetric SAM explains the
+due to advection of moist air from the Indian Ocean.In South America,
+precipitation anomalies associated with the Full SAM are negative both
+in Southern Chile and Southeastern South America, and positive in
+Southern Brazil. This features are cleanly separated between the
+Asymmetric and Symmetric components. The Symmetric SAM explains the
 negative anomalies in Southern Chile and the Asymmetric SAM, the
@@
 the negative relationship with precipitation in South America was absent
-in some areas and switched sign in other in the earlier period. The
+in some areas and switched sign in others in the earlier period. The
 correlation between ENSO and SAM is similarly non-stationary, also
-disapearing before 1973.
+changing sign before the 1980s \citep{fogt2006, clem2013}. Seeing as
+both the ENSO-SAM relationship and most of the precipitation impacts in
+South America are captured by the Asymmetric SAM, the results presented
+here are most likely period-dependent.
 
-Seeing as both the ENSO-SAM relationship and most of the precipitation
-imacts in South America are captured by the Asymmetric SAM, the results
-presented here are most likely period-dependent. Therefore, is very
-likely that if we were to repeat this analysis using pre-satellite data,
-the resulting Asymmetric SAM would look very different.
+By succesfully separating the zonally symmetric and zonally symmetric
+SAM signals, we show that the asymmetric component of the SAM has its
+unique variability, trends and impacts. This is particularly important
+in the context of a changing climate, as the impact on the SAM of ozone
+recovery is modeled as highly zonally symmetric, while the impact of
+increased concentration of greenhose gases has also a zonally asymmetric
+component \citep{arblaster2006, simpkins2012}.
 
@@
 
-\bibliography{AsymSAM}
+The research was supported by UBACyT20020170100428BA and the CLIMAX
+Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was
+supported by a PhD grant from CONICET, Argentina.
+
+\bibliography{AsymSAM,packages}
 
@@
 \begin{figure}
-\includegraphics{A3-1} \appendcaption{A3}{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with withs 3 and 5.
+\includegraphics{A3-1} \appendcaption{A3}{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
 The shading indicates de 95\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\% of the simulated sampels had an amplitude equal or lower).

### 2020-10-01 R/partial_cor.R

@@ -2,19 +2,27 @@
 #'
 #' @param y variable to which partial correlation will be computed
 #' @param x1,x2 variables that will be correlated with y
-#' @param weights vector of weights
+# #' @param weights vector of weights
 #'
 #' @export
-partial_cor <- function(y, x1, x2, weights = rep(1, length(y))) {
-  rho_x1x2 <- stats::cov.wt(cbind(x1, x2), wt = weights, cor = TRUE)$cor[1, 2]
-  rho_x1y <- stats::cov.wt(cbind(x1, y), wt = weights, cor = TRUE)$cor[1, 2]
-  rho_x2y <- stats::cov.wt(cbind(x2, y), wt = weights, cor = TRUE)$cor[1, 2]
+partial_cor <- function(y, x1, x2) {
 
-  estimate1 <- (rho_x1y - rho_x2y*rho_x1x2 ) / (sqrt(1 - rho_x2y^2) * sqrt(1 - rho_x1x2^2))
-  estimate2 <- (rho_x2y - rho_x1y*rho_x1x2 ) / (sqrt(1 - rho_x1y^2) * sqrt(1 - rho_x1x2^2))
+
+  estimate1 <- ppcor::pcor.test(y, x1, x2)
+  estimate2 <- ppcor::pcor.test(y, x2, x1)
+
+
+  # rho_x1x2 <- stats::cov.wt(cbind(x1, x2), wt = weights, cor = TRUE)$cor[1, 2]
+  # rho_x1y <- stats::cov.wt(cbind(x1, y), wt = weights, cor = TRUE)$cor[1, 2]
+  # rho_x2y <- stats::cov.wt(cbind(x2, y), wt = weights, cor = TRUE)$cor[1, 2]
+  #
+  # estimate1 <- (rho_x1y - rho_x2y*rho_x1x2 ) / (sqrt(1 - rho_x2y^2) * sqrt(1 - rho_x1x2^2))
+  # estimate2 <- (rho_x2y - rho_x1y*rho_x1x2 ) / (sqrt(1 - rho_x1y^2) * sqrt(1 - rho_x1x2^2))
 
   list(term = c(deparse(substitute(x1)),
                 deparse(substitute(x2))),
-       partial_correlation = c(estimate1, estimate2))
+       partial_correlation = c(estimate1$estimate, estimate2$estimate),
+       p.value = c(estimate1$p.value, estimate2$p.value)
+  )
 }
 

### 2020-10-02 analysis/paper/paper.Rmd
agrega datastatement

@@
 The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].
-
-A version-controlled repository of the code used to create this analysis can be found at https://github.com/eliocamp/asymsam.
 
@@
 
+\datastatement
+All data used in this paper is freely available in their respective sources.
+ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means).
+NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html).
+The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt
+
+A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at https://github.com/eliocamp/asymsam.
+
 \bibliography{AsymSAM,packages}

### 2020-10-02 analysis/paper/paper.tex
agrega datastatement

@@
 rmarkdown \citep{xie2015, allaire2019}.
-
-A version-controlled repository of the code used to create this analysis
-can be found at https://github.com/eliocamp/asymsam.
 
@@
 
+\datastatement
+
+All data used in this paper is freely available in their respective
+sources. ERA5 data can be obtained via the Copernicus Climate Data Store
+(https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means).
+NOAAGlobalTemp and GPCC precipitation data can be obtained through the
+NOAA Physical Sicences Laboratory website
+(https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and
+https://psl.noaa.gov/data/gridded/data.gpcc.html). The Oceanic Niño
+Index is avaiable via NOAA's Climate Precidction Center:
+https://www.cpc.ncep.noaa.gov/products/analysis\_monitoring/ensostuff/detrend.nino34.ascii.txt
+
+A version-controlled repository of the code used to create this
+analysis, including the code used to download the data can be found at
+https://github.com/eliocamp/asymsam.
+
 \bibliography{AsymSAM,packages}

### 2020-10-09 analysis/paper/paper.Rmd
Agrega comentarios leandro

@@
   - name: Carolina Vera
+  # LEAN: A Carolina la podes agrgear como author3 con nuestra misma afiliación 
 abstract: |
@@
 
-The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a;@fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [@fogt2020].
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a;@fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [e.g.
+@fogt2020].
 Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
@@
 
-Even though these indices are based on zonal averages, the spatial the spatial structure of the SAM computed from them contains noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
+Even though these indices are based on zonal averages, the spatial structure of the SAM computed from them contains noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
 These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009].
@@
 
+#LEAN: En el "from them" no me queda claro exactamente a qué se refiere el them, me parece que podrías ser un poco más especifíco.
+
+#LEAN:  Creo que deberían agregarse algunas otras referencias para "previous work suggest that they strongly modulate the regional impacts of the SAM"
+
 At least some of the variability associated with the zonal asymmetries of the SAM is probably forced by the tropics.
-In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987; @kidson1988;@karoly1989], whose wave train protects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
+In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987; @kidson1988;@karoly1989], whose wave train projects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
 And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed [@fan2007; @fogt2011; @clem2013].
@@
 
-Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn [@fogt2020 and references therein].
+Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn [e.g.
+@fogt2020 and references therein].
 These trends are thought of driven primarily by stratospheric ozone depletion and understood in the context of zonal mean variables.
 However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variability could be masking influencing the observed trends independently.
+
+#LEAN: Te faltaría alguna referencia de la segunda oración "These trends are thought of driven primarily by stratospheric ozone depletion and understood in the context of zonal mean variables.".
+Se me ocurre alguno de Julie arblaster que tiene varios trabajos en esa línea.
+También agregaría que los GHG también tienen efecto.
 
@@
 The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
-The importance of zonal asymmetries of the SAM in these impacts have been studied in certain regions.
+The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
 For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
@@
 This is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].
-
 Moreover, @fogt2012 studied the zonal asymmetric component of the SAM only in sea level pressure.
@@
 Finally, we investigate their relationship with temperature and precipitation anomalies.
+
+#LEAN: Te había hecho el comentario antes pero quizás no había sido del todo claro.
+Me parece que acá hay que distinguir más en cómo se trabaja metodológicamente cada parte con respecto a los niveles atmosféricos.
+Por ej.
+para la evaluación de los impactos solamente se está considerando el ínidice troposférico.
 
@@
 
-To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude, and monthly rainfall 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015].
+To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude, and monthly rainfall 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015].
+The rainfall product uses station-based records, and thus it only has continental coverage.
+
+#LEAN: Se suele citar esto tambień para GPCC, sería el paper correspondiente de la base https://doi.org/10.3390/atmos8030052
 
@@
 
+#LEAN: Hay algún problema con los guiones en los figure captions, revisa que en varios el PDF quedo todo pegado 19792018
+
 Our method assumes linearity in the asymmetric component of the SAM.
-That is, we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM.
+That means we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM.
 @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
@@
 
+#Para la Figure A9, A10 también aclara el período
+
+#Un detalle que habría que aclarar mejor en algún lado es por qué utilizas 700 hPa para esta discusión y otras que siguen.
+Sin la aclaración puede zonar un poco arbitrario (y acá solo consideras ese nivel para la justificación metodológica).
+
 By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons --December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
-Geopotential zonal anomalies computed by projecting the first EOF *of each season* are very similar to each other (Figure A10) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (between MAM and SON).
+Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other (Figure A10) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (between MAM and SON).
 Based on this, we believe that our initial assumption is not unreasonable.
@@
 it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and vice versa for $\beta$.
-When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we average of the relevant variables seasonally for each year and trimester before computing the regression.
+When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we average the relevant variables seasonally for each year and trimester before computing the regression.
 
@@
 We computed linear trends by Ordinary Least Squares and the 95% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
-
 To compute the amplitude of the zonal waves we computed the Fourier transform of the spatial field at each latitude circle.
@@
 The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].
+
+A version-controlled repository of the code used to create this analysis can be found at https://github.com/eliocamp/asymsam.
+#LEAN: Ver de mover a dónde debería ir
 
@@
 As will be shown later, both indices are highly coherent within each atmospheric layer, therefore is reasonable to take one level as representative of each layer.
+
+#LEAN: La aclaración metodológica de los "density estimates" no debería aparecer en el caption sino en la metodología del trabajo.
+Después en el caption ya no es necesario volverlo a aclarar.
+Algo que me resulta raro en la figura es la posición de las letras.
+Creo que es mejor darla vuelta, poner la aclaración de los niveles a la izquierda y los density estimates del lado derecho, eso la va a hacer bastante más clara y también más chorente con el caption.
 
@@
 This confirms the correct decomposition of the SAM in its symmetric and asymmetric components.
-
 Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which --as mentioned before-- is larger in the Western Hemisphere than in the Eastern Hemisphere.
@@
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index [@bamston1997] (ONI) is `r cor_texts[term == "full"]$cor` (p-value = `r cor_texts[term == "full"]$p.value`).
+The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value = `r cor_texts[term == "full"]$p.value`).
 Consistent with @fan2007, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value = `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value = `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
@@
 
-## Impacts
+## Surface Impacts
 \label{sec:impacts}
@@
 
+#LEAN: En las figuras 11 y 12 deberías decir en el caption algo sobre cuál es el intervalo de los contornos de la altura geopotencial
+
 ```{r calc-indices2d-seasonal}
@@
 
+#LEAN: Te quedo distinto el estilo en que se citan  @silvestri2009 y @zamboni2010
+
 ## Conclusions
@@
 In Australia the Full SAM is associated with positive precipitation anomalies in South East and this is explained by the Symmetric SAM.
-However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
+However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
+In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
 This features are cleanly separated between the Asymmetric and Symmetric components.

### 2020-10-10 analysis/paper/paper.Rmd
Borra repo
casi casi final draft

@@
 twocol: FALSE  #  Use FALSE for submission. 
-title: Asssesment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
+title: Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
 author1: 
@@
   - name: Carolina Vera
-  # LEAN: A Carolina la podes agrgear como author3 con nuestra misma afiliación 
+    affiliation: |
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+     CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+     CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
 abstract: |
@@
   This is significant becasue I wrote it.
-bibliography: "AsymSAM.bib,packages.bib"
+bibliography: ["AsymSAM.bib","packages.bib"]
 output: 
-  rticles::ams_article:
+  bookdown::word_document2: default
+  bookdown::pdf_book: 
+    base_format: rticles::ams_article
     latex_engine: "xelatex"
@@
 
-Even though these indices are based on zonal averages, the spatial structure of the SAM computed from them contains noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
-These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009].
+Even though these indices are based on zonal averages, spatial anomalies of geopotential associated with them (either by regression or composition) contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
+These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM [@fan2007, @fogt2012, @rosso2018], going as far as reversing its relationship between precipitation in South America [@silvestri2009].
 At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].
-
-#LEAN: En el "from them" no me queda claro exactamente a qué se refiere el them, me parece que podrías ser un poco más especifíco.
-
-#LEAN:  Creo que deberían agregarse algunas otras referencias para "previous work suggest that they strongly modulate the regional impacts of the SAM"
 
@@
 @fogt2020 and references therein].
-These trends are thought of driven primarily by stratospheric ozone depletion and understood in the context of zonal mean variables.
+These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013].
 However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variability could be masking influencing the observed trends independently.
-
-#LEAN: Te faltaría alguna referencia de la segunda oración "These trends are thought of driven primarily by stratospheric ozone depletion and understood in the context of zonal mean variables.".
-Se me ocurre alguno de Julie arblaster que tiene varios trabajos en esa línea.
-También agregaría que los GHG también tienen efecto.
 
@@
 We then study the spatial patterns described by the variability exclusive to each index.
-Finally, we investigate their relationship with temperature and precipitation anomalies.
+Finally, we investigate the relationship of the SAM at 700 hPa with temperature and precipitation anomalies.
 
@@
 
-In the Section \ref{sec:methods} we describe the methods.
-In Section \ref{sec:results}.\ref{sec:temporal} we describe the temporal variability and vertical coherence of the indices.
-In  Section \ref{sec:results}.\ref{sec:spatial}, we analyse the spatial patterns of geopotential height associated with them.
-In Section \ref{sec:results}.\ref{sec:impacts}, we study their relationship with surface-level temperature and precipitation.
+In the Section \@ref(methods) we describe the methods.
+In Section \@ref(temporal) we describe the temporal variability and vertical coherence of the indices.
+In  Section \@ref(spatial), we analyse the spatial patterns of geopotential height associated with them.
+In Section \@ref(impacts), we study their relationship with surface-level temperature and precipitation.
 
@@
 
-# Methods
-\label{sec:methods}
+# Methods {#methods}
 
@@
 
-To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude, and monthly rainfall 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015].
+To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude, and monthly rainfall 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015;@schneider2017].
 The rainfall product uses station-based records, and thus it only has continental coverage.
-
-#LEAN: Se suele citar esto tambień para GPCC, sería el paper correspondiente de la base https://doi.org/10.3390/atmos8030052
 
@@
 
-To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \ref{fig:method} for 700 hPa.
+To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \@ref(fig:method) for 700 hPa.
 The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
@@
 
-```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period.
+```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 
-#LEAN: Hay algún problema con los guiones en los figure captions, revisa que en varios el PDF quedo todo pegado 19792018
-
 Our method assumes linearity in the asymmetric component of the SAM.
@@
 @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
-Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of 700 hPa geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear (Figure A9).
+Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures A3 and A4).
 Therefore, we believe that our method is a reasonable approximation of the phenomenon.
 
-#Para la Figure A9, A10 también aclara el período
-
-#Un detalle que habría que aclarar mejor en algún lado es por qué utilizas 700 hPa para esta discusión y otras que siguen.
-Sin la aclaración puede zonar un poco arbitrario (y acá solo consideras ese nivel para la justificación metodológica).
-
-By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons --December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
-Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other (Figure A10) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (between MAM and SON).
-Based on this, we believe that our initial assumption is not unreasonable.
+By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
+Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure A5, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
+In the stratosphere (Figure A5, row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree\ in comparison with the rest of the year.
+Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
+Based on this, we believe that our initial assumption is not unreasonable throught most of the year with the exception of DJF in the stratosphere.
 
 Finally, we assume that the zonally asymmetric pattern is stationary in time.
-@silvestri2009 suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979-2018) so it's unlikely that we could observe significant changes.
-Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change (Figure A11).
+@silvestri2009 suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes.
+Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure A6).
 
@@
 
+We computed density probability estimates using a gaussian kernel of optimal bindwidth according to @sheather1991.
+
 ### Computation procedures
@@
 
-A version-controlled repository of the code used to create this analysis can be found at https://github.com/eliocamp/asymsam.
-#LEAN: Ver de mover a dónde debería ir
-
-# Results
-\label{sec:results}
-
-## Temporal evolution
-\label{sec:temporal}
+# Results {#results}
+
+## Temporal evolution {#temporal}
 
@@
 ```{r asymsam-timeseries, fig.cap = "Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
-To the left, probability density estimate of each index computed using a gaussian kernel of optimal bindwidth according to \\cite{sheather1991}.
+To the right, probability density estimate of each index.
 Series are standarised by the standard deviation of the Full SAM at each level.", fig.width=6, fig.height=3, fig.env="figure*"}
-indices %>% 
-  .[lev %in% c(50, 700) & term != "full"] %>% 
-  ggplot(aes(estimate_norm)) +
-  geom_density(aes(color = term), bw = "SJ") +
-  facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
-  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam, guide = "none") +
-  scale_y_continuous(NULL, trans = scales::reverse_trans(), labels = NULL, breaks = NULL) +
-  scale_x_continuous(NULL, labels = NULL, breaks = 0) +
-  coord_flip()  +
-  theme(strip.text = element_blank()) +
-
   indices %>% 
@@
   scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam) +
-  scale_y_continuous(NULL) +
+  scale_y_continuous(NULL, position = "right") +
   scale_x_datetime(NULL, expand = c(0, 0)) +
+  facet_grid(lev~., labeller = labeller(lev = lev.lab), switch = "y") +
+  tag_facets() +
+
+indices %>% 
+  .[lev %in% c(50, 700) & term != "full"] %>% 
+  ggplot(aes(estimate_norm)) +
+  geom_density(aes(color = term), bw = "SJ") +
   facet_grid(lev~., labeller = labeller(lev = lev.lab)) +
-  tag_facets() +
-  plot_layout(widths = c(1, 4))
+  scale_color_brewer(NULL, palette = "Dark2", labels = lab_sam, guide = "none") +
+  scale_y_continuous(NULL, labels = NULL, breaks = NULL) +
+  scale_x_continuous(NULL, labels = NULL, breaks = 0) +
+  coord_flip()  +
+  theme(strip.text = element_blank()) +
+  plot_layout(widths = c(4, 1)) 
 ```
@@
 We first asses the temporal evolution of the Asymmetric SAM and Symmetric SAM.
-Figure \ref{fig:asymsam-timeseries} shows the corresponding time series for 700 hPa and 50 hPa and their corresponding density estimates.
+Figure \@ref(fig:asymsam-timeseries) shows the corresponding time series for 700 hPa and 50 hPa and their corresponding density estimates.
 We selected these two levels as representative of the tropospheric and stratospheric variability respectively.
 As will be shown later, both indices are highly coherent within each atmospheric layer, therefore is reasonable to take one level as representative of each layer.
-
-#LEAN: La aclaración metodológica de los "density estimates" no debería aparecer en el caption sino en la metodología del trabajo.
-Después en el caption ya no es necesario volverlo a aclarar.
-Algo que me resulta raro en la figura es la posición de las letras.
-Creo que es mejor darla vuelta, poner la aclaración de los niveles a la izquierda y los density estimates del lado derecho, eso la va a hacer bastante más clara y también más chorente con el caption.
 
@@
 
-The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A3).
+The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A2).
 This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b].
@@
 
-From Figure \ref{fig:asymsam-timeseries} we can see that the Asymmetric SAM and Symmetric SAM time series appear to be correlated.
+From Figure \@ref(fig:asymsam-timeseries) we can see that the Asymmetric SAM and Symmetric SAM time series appear to be correlated.
 Moreover, looking at the extremes in the stratosphere, the Symmetric SAM series appears to lag the Asymmetric SAM series (see, for example, the positive events on late 1987).
-We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric SAM index leading the Symmetric SAM index), in Figure \ref{fig:cor-lev}.
+We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric SAM index leading the Symmetric SAM index), in Figure \@ref(fig:cor-lev).
 Zero-lag correlations between the Asymmetric SAM and Symmetric SAM series are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
@@
 
-```{r cor-lev, fig.cap = "Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979–2018 period.", fig.width=3, fig.height=3}
+```{r cor-lev, fig.cap = "Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.", fig.width=3, fig.height=3}
 cors %>% 
@@
 
-```{r cross-correlation, fig.cap = "Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979–2018 period.", fig.width = 6, fig.height=3, fig.env="figure*"}
+```{r cross-correlation, fig.cap = "Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.", fig.width = 6, fig.height=3, fig.env="figure*"}
 tri <- matrix(c(1000, 1000, 1000, 1, 1, 1, 1000, 1000), ncol = 2, byrow = TRUE) %>% 
@@
 
-Figure \ref{fig:cross-correlation}a shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
+Figure \@ref(fig:cross-correlation)a shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
 For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
@@
 
-```{r trends, fig.cap = "Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+```{r trends, fig.cap = "Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
 Shading indicates the 95\\% confidence interval from a t-distribution.", fig.width = 6, fig.height=4, fig.env="figure*"}
@@
 
-The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979-2018 at each level (Figure \ref{fig:trends}) for the whole year and separated by trimesters.
+The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979 -- 2018 at each level (Figure \@ref(fig:trends)) for the whole year and separated by trimesters.
 The Full SAM index presents a statistically significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
@@
 We stress that these are only linear trends during the whole period and the absence of a statistically significant signal should not be taken as evidence of no systematic change.
-In particular, going back to Figure \ref{fig:asymsam-timeseries}, we can see an evident change in the stratospheric Asymmetric component (red line in panel a) between the 90's, when we see a dominance of extreme negative values, and the 00's, when we see the inverse.
+In particular, going back to Figure \@ref(fig:asymsam-timeseries), we can see an evident change in the stratospheric Asymmetric component (red line in panel a) between the 90's, when we see a dominance of extreme negative values, and the 00's, when we see the inverse.
 This change is restricted to the Winter months: the linear trend for Winter starting in 1990 for the Asymmetric component at 50hPa is $`r trends_text[term == "asym" & lev == 50 & season == "JJA"]$V1`$.
 
-Figure \ref{fig:r-squared-trend} shows decadal trends for the explained variance of each index.
+Figure \@ref(fig:r-squared-trend) shows decadal trends for the explained variance of each index.
 There is no evidence of a significant trend in the stratosphere.
@@
 
-```{r r-squared-trend, fig.height=3, fig.width=3, fig.cap = "Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979-2018.
+```{r r-squared-trend, fig.height=3, fig.width=3, fig.cap = "Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018.
 Shading indicates the 95\\% confidence interval."}
@@
 
-## Spatial patterns
-\label{sec:spatial}
+## Spatial patterns {#spatial}
 
@@
 
-```{r 2d-regr, fig.cap = "Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979–2018 period.
+```{r 2d-regr, fig.cap = "Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period.
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices.
@@
 To show if, and to what extent, the Asymmetric and Symmetric SAM indices indeed capture the asymmetric and symmetric component of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa levels.
-Figure \ref{fig:2d-regr} shows these regressions.
+Figure \@ref(fig:2d-regr) shows these regressions.
 Regression coefficients in column a are computed using the Full SAM.
@@
 
-```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.", fig.width=6, fig.height=3, fig.env = "figure*"}
+```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\@ref(fig:2d-regr) for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.", fig.width=6, fig.height=3, fig.env = "figure*"}
 indices_2d_yearly %>% 
@@
 
-The amplitude of the first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
+The amplitude of the first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \@ref(fig:wave-amplitude), where wave number zero represents the amplitude of the zonal mean.
 Zonal wave amplitudes of the spatial pattern described by the Full SAM index (column `r get_col(term == "full")`) are dominated by the zonal mean (wave-number 0) at both levels.
@@
 This confirms the correct decomposition of the SAM in its symmetric and asymmetric components.
-Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which --as mentioned before-- is larger in the Western Hemisphere than in the Eastern Hemisphere.
+Looking at panel b.2 from Figure \@ref(fig:2d-regr), it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
 
@@
 ```{r vertical-regression, fig.cap = "Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\\degree and 40\\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
-(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979–2018 period.", fig.width = 3, fig.height=3}
+(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979 -- 2018 period.", fig.width = 3, fig.height=3}
 vertical_regr %>% 
@@
 
-To analyse the vertical structure of the geopotential height anomalies associated with the asymmetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
+To analyse the vertical structure of the geopotential height anomalies associated with the asymmetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \@ref(fig:vertical-regression)).
 The geopotential height anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
@@
 So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
-The structures shown in panels a and b in  Figure \ref{fig:vertical-regression} are robust to the choice of index level.
+The structures shown in panels a and b in  Figure \@ref(fig:vertical-regression) are robust to the choice of index level.
 For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in panel a.
@@
 
-The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
+The wave-3 pattern from Figure \@ref(fig:2d-regr) panel b.2 is very similar to the Pacific-South American Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
@@
 
-## Surface Impacts
-\label{sec:impacts}
+## Surface Impacts {#impacts}
 
@@
 
-```{r regr-air-season, fig.cap = "Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979–2018 period.
+```{r regr-air-season, fig.cap = "Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 
-Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface temperature for each trimester.
+Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface temperature for each trimester.
 In Summer, positive values of the Full SAM index (panel `r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
@@
 
-The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons --although with varying intensity and small-scale details-- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
+The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons -- although with varying intensity and small-scale details -- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
 It's then not surprising to see it more clearly in association with the Symmetric SAM (at least in Summer and Spring).
 
-Figure \ref{fig:regr-air-season} column b can be partially compared with Figure 11 from @fogt2012.
+Figure \@ref(fig:regr-air-season) column b can be partially compared with Figure 11 from @fogt2012.
 Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.
 
-Regression of the SAM indices with seasonal mean precipitation and 700 hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
+Regression of the SAM indices with seasonal mean precipitation and 700 hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for Australasia and South America respectively.
 South Africa is not shown because no significant signal was detected there.
-
-#LEAN: En las figuras 11 y 12 deberías decir en el caption algo sobre cuál es el intervalo de los contornos de la altura geopotencial
 
@@
     # stat_subset(aes(subset = p.value <= 0.05 & is.cross(lon, lat, 2)), size = 0.4, alpha = 0.8) +
-    geom_contour2(data = gh,
+    geom_contour2(data = gh, breaks = MakeBreaks(bins = 15, exclude = 0),
                   color = "gray50", size = 0.2, global.breaks = FALSE, 
                   aes(z = estimate, linetype = factor(-sign(..level..)))) +
+    geom_text_contour(data = gh, breaks = MakeBreaks(bins = 15, exclude = 0),
+                      global.breaks = FALSE,
+                      size = 3,
+                      stroke = 0.15, stroke.color = "white",
+                      aes(z = estimate), rotate = FALSE, skip = 1) +
     geom_qmap(~.x[lat <= 10], keep = 0.015, weighting = 0.9) +
@@
 
-```{r pp-regr-africa,  fig.cap = "Regression of precipiation anomalies with Asymmetric and Symmetric SAM for the 1979–2018 period.
+```{r pp-regr-africa,  fig.cap = "Regression of precipiation anomalies with Asymmetric and Symmetric SAM for the 1979 -- 2018 period.
 P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
@@
 
-```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979–2018 period.
+```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
@@
 
-In Australia, the annual regression shows that the Full SAM index is positively associated with precipitation in the Southeastern region (Figure \ref{fig:pp-regr-oceania} panel a.1), which reproduces the results from @gillett2006.
+In Australia, the annual regression shows that the Full SAM index is positively associated with precipitation in the Southeastern region (Figure \@ref(fig:pp-regr-oceania) panel a.1), which reproduces the results from @gillett2006.
 The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (panel c.1).
@@
 
-```{r pp-regr-america, dependson = "def-plot_pp", fig.cap = "Same as Figure \\ref{fig:pp-regr-oceania} but for South America.", fig.width = 6, fig.height = 7.6, fig.env = "figure*"}
+```{r pp-regr-america, dependson = "def-plot_pp", fig.cap = "Same as Figure \\@ref(fig:pp-regr-oceania) but for South America.", fig.width = 6, fig.height = 7.6, fig.env = "figure*"}
 plot_pp("america")
@@
 
-In South America (Figure \ref{fig:pp-regr-america}), the annual-level regression shows that positive SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1).
-Panels b.1 and c.1 show a remarkably clean separation between the Asymmetric SAM --associated with the Southeastern South American and Southern Brazilian signals-- and the Symmetric SAM --associated with the signal in Southern Chile.
+In South America (Figure \@ref(fig:pp-regr-america)), the annual-level regression shows that positive SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1).
+Panels b.1 and c.1 show a remarkably clean separation between the Asymmetric SAM -- associated with the Southeastern South American and Southern Brazilian signals -- and the Symmetric SAM -- associated with the signal in Southern Chile.
 
@@
 On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
-This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region [@silvestri2009, @zamboni2010].
+This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region [@silvestri2009;@zamboni2010].
 This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events [@rosso2018].
 There is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 and that is explained by the Asymmetric SAM.
-
-#LEAN: Te quedo distinto el estilo en que se citan  @silvestri2009 y @zamboni2010
 
@@
 There is evidence of influence between the QBO and the Northern Annular Mode [e.g.
-@holton1980, @watson2014, @zhang2020], so it's not unlikely that the SAM would be similarly connected.
+@holton1980;@watson2014;@zhang2020], so it's not unlikely that the SAM would be similarly connected.
 However establishing this link would require further research.
@@
 In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship.
-These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [@cazes-boezio2003; @fogt2011; @cai2020a].
+These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [@cazes-boezio2003;@fogt2011;@cai2020a].
 
@@
 ```{r A1, fig.cap = "Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
-Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.", fig.width = 4, fig.height=4, appendix = TRUE}
+Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.
+For the 1979 -- 2018 period.", fig.width = 4, fig.height=4, appendix = TRUE}
 ccf_dt <- function(..., plot = FALSE) {
@@
 
-```{r A3, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
+```{r A2, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
 The shading indicates de 95\\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\\% of the simulated sampels had an amplitude equal or lower).
-The light line indicates the theoretical expected amplitude from the autorregressive model.", appendix = TRUE, fig.height=4, fig.width=7}
+The light line indicates the theoretical expected amplitude from the autorregressive model.
+For the 1979 -- 2018 period.", appendix = TRUE, fig.height=4, fig.width=7}
 spec %>%
@@
 
-```{r A9, fig.cap = "700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\\pm1$ standard deviation as threshhold.
+```{r A3, fig.cap = "50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
 Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.", apendix = TRUE, fig.height = 5, fig.width = 7}
-composites <- hgt[lev == 700] %>%
+composites <- hgt[lev %in% c(50, 700)] %>%
   indices_wide[., on = .NATURAL] %>% 
-  .[, sign := ifelse(abs(full) > sd(full), sign(full), 0), by = month(time)] %>% 
-  .[, .(hgt = mean(hgt_a), n = .N), by = .(lon, lat, sign, season(time))] %>% 
-  .[, z := Anomaly(hgt), by = .(lat, sign, season)]
+  .[, sign := ifelse(abs(full) > sd(full), sign(full), 0), by = .(lev, month(time))] %>% 
+  .[, .(hgt = mean(hgt_a), n = .N), by = .(lon, lat, sign, lev, season(time))] %>% 
+  .[, z := Anomaly(hgt), by = .(lat, sign, season, lev)]
+
+plot_composites <- function(this_lev) {
+  composites <- composites[lev == this_lev]
 
@@
     dcast(season + lat + lon  ~ sign, value.var = "z") %>% 
-  .[, .(cor = cor(`-1`, `1`)), by = season]
+    .[, .(cor = cor(`-1`, `1`)), by = .(season)]
 
@@
     .[sign != 0] %>% 
-  composites_cors[., on = "season"] %>% 
+    composites_cors[., on = c("season")] %>% 
     .[order(season)] %>% 
@@
     scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
-                       limits = c(-74, 74),
+                         # limits = c(-74, 74),
                          breaks = AnchorBreaks(0, binwidth = 15, exclude = 0)) +
@@
     tag_facets(tag_pool = tags$V1, tag_suffix = "")
-
-```
-
-```{r A10, fig.cap = "Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently.",  apendix = TRUE, fig.width=7, fig.height=2.5}
+}
+plot_composites(50)
+```
+
+```{r A4, fig.cap = "700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.", apendix = TRUE, fig.height = 5, fig.width = 7}
+plot_composites(700)
+```
+
+```{r A5, fig.cap = "Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.",  apendix = TRUE, fig.width=7, fig.height=6}
 set.seed(42)
 SAM_season <- hgt[lat <= -20] %>% 
-  .[lev == 700] %>% 
+  .[lev %in% c(700, 50)] %>% 
   .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
@@
   .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
-    by = .(season)]
-
-patterns <- SAM_season[, eof[[1]]$left, by = season] %>% 
+    by = .(season, lev)]
+
+patterns <- SAM_season[, eof[[1]]$left, by = .(lev, season)] %>% 
   copy() %>% 
   setnames("hgt", "EOF") %>% 
-  .[, EOF := EOF/sd(EOF), by = .(season)] %>% 
-  hgt[., on = "time"] %>% 
-  .[, FitLm(hgt_a, EOF), by = .(lon, lat, season)] %>%
+  .[, EOF := EOF/sd(EOF), by = .(season, lev)] %>% 
+  hgt[., on = c("time", "lev")] %>%
+  .[, FitLm(hgt_a, EOF), by = .(lon, lat, lev, season)] %>%
   rm_intercept() %>% 
   .[season == "DJF", estimate := -estimate] %>% 
-  .[, estimate_a := Anomaly(estimate), by = .(lat, season)]
+  .[, estimate_a := Anomaly(estimate), by = .(lat, lev, season)]
 
 cors <- patterns %>% 
-  .[ , f := interaction(lon, lat)] %>% 
-  widyr::pairwise_cor(season, f, estimate_a, upper = FALSE)
-
+  .[ , f := as.character(interaction(lon, lat)), by = lev] %>% 
+  .[, widyr::pairwise_cor(.SD, season, f, estimate_a, upper = FALSE), by = lev]
+
+plot_patterns <- function(this_lev) {
+  limits <- NULL
+  breaks <- if (this_lev == 50) AnchorBreaks(0, 20, exclude = 0) else AnchorBreaks(0, 5, exclude = 0)
   patterns %>% 
+    .[lev == this_lev] %>% 
     periodic(lon = c(0, 360)) %>% 
+    copy() %>% 
+    # .[, estimate_a := estimate_a/sd(estimate_a), by = .( lev)] %>% 
     ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate_a), breaks = AnchorBreaks(0, exclude = 0), global.breaks = FALSE) +
-  geom_contour_tanaka(aes(z = estimate_a), breaks = AnchorBreaks(0, exclude = 0), global.breaks = FALSE) +
+    geom_contour_fill(aes(z = estimate_a), breaks = breaks) +
+    geom_contour_tanaka(aes(z = estimate_a), breaks = breaks) +
     geom_qmap(~.x[lat <= 0]) +
@@
                          guide = guide_colorstrip_bottom(),
-                       breaks = AnchorBreaks(0, 10, exclude = 0)) +
+                         limits = limits,
+                         breaks = breaks) +
     coord_polar() +
@@
     no_grid +
-  facet_grid(.~season)
-```
-
-```{r A11, fig.cap = "Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 to 1998 and 1999 to 2018.
-Pattern correlation between both fields is 0.76.", fig.width=6, fig.height=5, apendix = TRUE}
+    facet_grid(lev~season, labeller = labeller(lev = lev.lab))
+}
+
+plot_patterns(50) +
+  plot_patterns(700) +
+  plot_layout(ncol = 1) 
+```
+
+```{r A6, fig.cap = "Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
+Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.", fig.width=6, fig.height=5, apendix = TRUE}
+set.seed(42)
 SAM_prepost <- hgt[lat <= -20] %>% 
-  .[lev == 700] %>%
+  .[lev %in% c(700, 50)] %>%
   .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
@@
   .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
-    by = .(period)]
-
-sam_sep <- SAM_prepost[, eof[[1]]$right, by = .(period)] %>% 
+    by = .(period, lev)]
+
+sam_sep <- SAM_prepost[, eof[[1]]$right, by = .(lev, period)] %>% 
   setnames("hgt", "full") %>% 
-  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, period)]
-
-patterns <- SAM_prepost[, eof[[1]]$left, by = .(period)] %>% 
+  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, period, lev)]
+
+patterns <- SAM_prepost[, eof[[1]]$left, by = .(period, lev)] %>% 
   copy() %>% 
   setnames("hgt", "EOF") %>% 
-  .[, EOF := EOF/sd(EOF), by = .(period)] %>% 
+  .[, EOF := EOF/sd(EOF), by = .(period, lev)] %>% 
+  .[period == "1979-1998" & lev == 50, EOF := -EOF] %>% 
   rm_singleton() %>% 
   hgt[., on = .NATURAL] %>% 
-  .[lev == 700] %>% 
-  .[, FitLm(hgt_a, EOF), by = .(lon, lat, period)] %>%
+  # .[lev == 700] %>% 
+  .[, FitLm(hgt_a, EOF), by = .(lon, lat, period, lev)] %>%
   rm_intercept() %>% 
   # .[season == "DJF", estimate := -estimate] %>% 
-  .[, estimate_a := Anomaly(estimate), by = .(lat, period)]
+  .[, estimate_a := Anomaly(estimate), by = .(lat, period, lev)]
 
 cor <- patterns %>% 
-  dcast(lon + lat ~ period, value.var = "estimate_a") %>% 
-  .[, cor(`1979-1998`, `1999-2018`)]
+  dcast(lon + lat + lev ~ period, value.var = "estimate_a") %>% 
+  .[, cor(`1979-1998`, `1999-2018`), by = lev]
 
@@
   coord_polar() +
-  facet_wrap(period~., labeller = labeller(variable = label_parsed)) +
+  facet_grid(lev~period, labeller = labeller(lev = lev.lab)) +
   axis_labs_smol +

### 2020-10-10 analysis/paper/paper.tex
casi casi final draft

@@
 
-\title{Asssesment of zonal symmetric and asymmetric components of the Southern
-Annular Mode using a novel approach}
+\title{Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach}
 
@@
 
-\affiliation{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales,
-Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires,
-Argentina CONICET -- Universidad de Buenos Aires, Centro de
-Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el
-Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires,
-Argentina}
+\affiliation{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina}
 
@@
 }
+\extraaffil{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina}
 
@@
 
-\abstract{Enter the text of your abstract here. This is a sample American
-Meteorological Society (AMS) \LaTeX~template. This document provides
-authors with instructions on the use of the AMS \LaTeX~template. Authors
-should refer to the file amspaper.tex to review the actual \LaTeX~code
-used to create this document. The template.tex file should be modified
-by authors for their own manuscript.}
+\abstract{Enter the text of your abstract here.
+This is a sample American Meteorological Society (AMS) \LaTeX~template.
+This document provides authors with instructions on the use of the AMS \LaTeX~template.
+Authors should refer to the file amspaper.tex to review the actual \LaTeX~code used to create this document.
+The template.tex file should be modified by authors for their own manuscript.}
 
@@
 
-The Southern Annular Mode (SAM) is the main mode of variability in the
-Southern Hemisphere extratropical circulation \citep{rogers1982} in
-daily, monthly, and decadal timescales \citep{baldwin2001a, fogt2006}
-and exerts an important influence in weather conditions such as
-temperature and precipitation anomalies and sea ice concentration
-\citep{fogt2020}. Its positive phase is traditionally described as
-anomalously low pressures over Antarctica surrounded by a ring of
-anomalous high pressures in middle-to-high latitudes.
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation \citep{rogers1982} in daily, monthly, and decadal timescales \citep{baldwin2001a, fogt2006} and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration \citep[e.g.][]{fogt2020}.
+Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
 
-Most authors describe the SAM as a zonally symmetric pattern, a fact
-that is reflected not only in its name, but also in the various methods
-used to characterise it. Of the several different indices presented in
-the literature, many of them are based on zonal means of sea level
-pressure or geopotential height \citep{ho2012}. \citet{gong1999} defined
-the SAM index as the zonal mean sea level pressure difference between
-40\degree S and 60\degree S, which is also the definition used by the
-station-based index in \citet{marshall2003}. \citet{baldwin2009}
-proposed defining the Northern and Annular modes as the leading EOF of
-the zonally averaged geopotential height at each level.
+Most authors describe the SAM as a zonally symmetric pattern, a fact that is reflected not only in its name, but also in the various methods used to characterise it.
+Of the several different indices presented in the literature, many of them are based on zonal means of sea level pressure or geopotential height \citep{ho2012}.
+\citet{gong1999} defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 60\degree S, which is also the definition used by the station-based index in \citet{marshall2003}.
+\citet{baldwin2009} proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
 
-Even though these indices are based on zonal averages, the spatial the
-spatial structure of the SAM computed from them contains noticeable
-deviations from zonal symmetry, particularly in the Pacific Ocean
-region. These zonal asymmetries have not been widely studied, but
-previous work suggest that they strongly modulate the regional impacts
-of the SAM, going as far as reversing its relationship between
-precipitation in South America \citep{silvestri2009}. At the very least,
-the fact that the SAM is not entirely zonally symmetric hinders our
-ability to reconstruct its historical variability prior to the
-availability of dense observations in the Southern Hemisphere
-\citep{jones2009}.
+Even though these indices are based on zonal averages, spatial anomalies of geopotential associated with these indices (either by regressino or composition) contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
+These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM \citep[\citet{fogt2012}, \citet{rosso2018}]{fan2007}, going as far as reversing its relationship between precipitation in South America \citep{silvestri2009}.
+At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere \citep{jones2009}.
 
-At least some of the variability associated with the zonal asymmetries
-of the SAM is probably forced by the tropics. In particular, ENSO-like
-variability affects the Southern Hemisphere extratopics through the
-Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989},
-whose wave train protects strongly onto the zonal anomalies
-corresponding to the SAM in the Pacific sector. And although the
-relationship between ENSO and SAM is far from simple, tropical
-influences on the SAM have been observed
-\citep{fan2007, fogt2011, clem2013}. In particular, \citet{fan2007}
-computed SAM indices of the Western and the Eastern Hemisphere
-separately and found that they were much more correlated if the (linear)
-signal of the ENSO was removed. While this relates to temporally
-coherent variability of the two hemisphere and not necessarily to zonal
-asymmetries in the associated spatial patterns, it is nonetheless
-consistent of the ENSO playing a crucial role in zonal asymmetries of
-the SAM.
+At least some of the variability associated with the zonal asymmetries of the SAM is probably forced by the tropics.
+In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989}, whose wave train projects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
+And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed \citep{fan2007, fogt2011, clem2013}.
+In particular, \citet{fan2007} computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated if the (linear) signal of the ENSO was removed.
+While this relates to temporally coherent variability of the two hemisphere and not necessarily to zonal asymmetries in the associated spatial patterns, it is nonetheless consistent of the ENSO playing a crucial role in zonal asymmetries of the SAM.
 
-Positive trends in SAM index have been documented by various researchers
-using different indices mostly on boreal Summer and Autumn \citep[ and
-references therein]{fogt2020}. These trends are thought of driven
-primarily by stratospheric ozone depletion and understood in the context
-of zonal mean variables. However, it's not clear how or if the
-asymmetric component responds to this forcing, or whether its
-variability could be masking influencing the observed trends
-independently.
+Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn \citep[e.g.][ and references therein]{fogt2020}.
+These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables \citep{marshall2004, gillett2005, arblaster2006, gillett2013}.
+However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variability could be masking influencing the observed trends independently.
 
-Similarly unclear are the specific impacts of the zonally asymmetric
-component of the SAM. Positive phase of the SAM is associated with
-generally colder temperatures over Antarctica and warmer temperatures at
-higher latitudes \citep{jones2019} (and vice versa for negative SAM),
-but there are significant deviations from this zonal mean response,
-notably in the Antarctic Peninsula and the South Atlantic
-\citep{fogt2012}. The SAM signal in precipitation behaves similarly,
-although with even greater deviation from zonal symmetry
-\citep{lim2016}. The importance of zonal asymmetries of the SAM in these
-impacts have been studied in certain regions. For example, the
-SAM-precipitation relationship in Southeastern South America and
-Southern Brazil can be explained by the PSA-like zonally asymmetric
-circulation associated with the SAM \citep{silvestri2009, rosso2018}.
-\citet{fan2007} also found that precipitation in East Asia was impacted
-by the variability of only the Western Hemisphere part of the SAM.
+Similarly unclear are the specific impacts of the zonally asymmetric component of the SAM.
+Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes \citep{jones2019} (and vice versa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}.
+The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry \citep{lim2016}.
+The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
+For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}.
+\citet{fan2007} also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
 
-We are not aware of any previous work which quantifies the temporal
-variability of the asymmetric component of the SAM with the exception of
-\citet{fogt2012}. However, their methods based on composites of positive
-and negative SAM events leads to some issues, such as spatial patterns
-derived from as little as 4 cases and from imbalanced periods (for
-example, 5 of the 7 cases in their DJF SAM+ composite are from later
-than 1988, whereas all of the 8 years in their DJF SAM- composite are
-from earlier than 1988). This is particularly important due to the
-inhomogeneities in reanalysis products prior to the satellite era and
-the possible change in the asymmetric structure of the SAM
-\citep{silvestri2009}.
+We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of \citet{fogt2012}.
+However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988).
+This is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM \citep{silvestri2009}.
+Moreover, \citet{fogt2012} studied the zonal asymmetric component of the SAM only in sea level pressure.
+Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere \citep{baldwin2009}.
 
-Moreover, \citet{fogt2012} studied the zonal asymmetric component of the
-SAM only in sea level pressure. Zonal asymmetries in the SAM spatial
-pattern are fairly barotropic throughout the troposphere, but they
-change dramatically in the stratosphere \citep{baldwin2009}.
+Our objective is, then, to systematically characterise the zonally asymmetric and symmetric components of the SAM variability.
+For each level, we construct two indices which aim to capture exclusively the variability of the symmetric and asymmetric component respectively.
+We assess their vertical structure and coherence, temporal variability and trends.
+We then study the spatial patterns described by the variability exclusive to each index.
+Finally, we investigate the relationship of the SAM at 700 hPa with temperature and precipitation anomalies.
 
-Our objective is, then, to systematically characterise the zonally
-asymmetric and symmetric components of the SAM variability. For each
-level, we construct two indices which aim to capture exclusively the
-variability of the symmetric and asymmetric component respectively. We
-assess their vertical structure and coherence, temporal variability and
-trends. We then study the spatial patterns described by the variability
-exclusive to each index. Finally, we investigate their relationship with
-temperature and precipitation anomalies.
+\#LEAN: Te había hecho el comentario antes pero quizás no había sido del todo claro.
+Me parece que acá hay que distinguir más en cómo se trabaja metodológicamente cada parte con respecto a los niveles atmosféricos.
+Por ej.
+para la evaluación de los impactos solamente se está considerando el ínidice troposférico.
 
-In the Section \ref{sec:methods} we describe the methods. In Section
-\ref{sec:results}.\ref{sec:temporal} we describe the temporal
-variability and vertical coherence of the indices. In Section
-\ref{sec:results}.\ref{sec:spatial}, we analyse the spatial patterns of
-geopotential height associated with them. In Section
-\ref{sec:results}.\ref{sec:impacts}, we study their relationship with
-surface-level temperature and precipitation.
+In the Section \ref{sec:methods} we describe the methods.
+In Section \ref{sec:results}.\ref{sec:temporal} we describe the temporal variability and vertical coherence of the indices.
+In Section \ref{sec:results}.\ref{sec:spatial}, we analyse the spatial patterns of geopotential height associated with them.
+In Section \ref{sec:results}.\ref{sec:impacts}, we study their relationship with surface-level temperature and precipitation.
 
@@
 
-To describe the Southern Annular Mode and its variability we used
-monthly geopotential height at 2.5\degree~longitude by
-2.5\degree~latitude of horizontal resolution and 37 vertical isobaric
-levels from ERA5 \citep{hersbach2020} for the period 1979 to 2018. We
-restrict our analysis to the post-satellite era to avoid any confounding
-factors arising from the introduction of satellite observations.
+To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{hersbach2020} for the period 1979 to 2018.
+We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.
 
-To describe the relationship between the SAM indices and temperature and
-precipitation, we use temperature data from NOAA's Merged Land Ocean
-Global Surface Temperature Analysis V4.0.1 \citep{vose2012, smith2008},
-which blends land and ocean temperature analysis into a monthly global
-grid 5\degree~longitude by 5\degree~latitude, and monthly rainfall
-0.5\degree~longitude by 0.5\degree~latitude data from the Global
-Precipitation Climatology Centre \citep{schneider2015}.
+To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land and ocean temperature analysis into a monthly global grid 5\degree~longitude by 5\degree~latitude, and monthly rainfall 0.5\degree~longitude by 0.5\degree~latitude data from the Global Precipitation Climatology Centre \citep{schneider2015, schneider2017}.
+The rainfall product uses station-based records, and thus it only has continental coverage.
 
@@
 
-Traditionally the Souther Annular Mode (SAM) is defined as de leading
-Empirical Orthogonal Mode (EOF) of sea level pressure or geopotential
-height at lower levels \citep{ho2012}. Following \citet{baldwin2001}, we
-extend that definition vertically and use the term SAM to refer to the
-the leading EOF of the monthly anomalies of geopotential field south of
-20\degree S at each level. We performed EOFs by computing the Singular
-Value Decomposition of the data matrix consisting in 481 rows and 4176
-columns (144 points of longitude and 29 points of latitude). We weighted
-the values by the square root of the cosine of latitude to account for
-the non-equal area of each gridpoint \citep{chung1999}.
+Traditionally the Souther Annular Mode (SAM) is defined as de leading Empirical Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels \citep{ho2012}.
+Following \citet{baldwin2001}, we extend that definition vertically and use the term SAM to refer to the the leading EOF of the monthly anomalies of geopotential field south of 20\degree S at each level.
+We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in 481 rows and 4176 columns (144 points of longitude and 29 points of latitude).
+We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint \citep{chung1999}.
 
-To separate the zonally symmetric and asymmetric components of the SAM,
-we computed the zonal mean and anomalies of the full SAM spatial
-pattern, as shown in Figure \ref{fig:method} for 700 hPa. The full
-spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is the sum of the
-zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\)) and symmetric
-(\([\mathrm{EOF_1}](\lambda, \phi)\)) components. We then compute the
-``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM'' indices as the
-regression coefficients of the regression of each monthly geopotential
-field on the respective patterns (weighting by the cosine of latitude).
-The three indices are normalized by dividing them by the standard
-deviation of the ``Full'' index at each level. As a result, the
-magnitudes between indices are comparable. However, only ``Full'' index
-will have unit standard deviation per definition. From the regression,
-we also use the explained variance of each pattern as an indicator of
-the degree of zonally symmetry or asymmetry of each monthly field.
+To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \ref{fig:method} for 700 hPa.
+The full spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is the sum of the zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\)) and symmetric (\([\mathrm{EOF_1}](\lambda, \phi)\)) components.
+We then compute the ``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM'' indices as the regression coefficients of the regression of each monthly geopotential field on the respective patterns (weighting by the cosine of latitude).
+The three indices are normalized by dividing them by the standard deviation of the ``Full'' index at each level.
+As a result, the magnitudes between indices are comparable.
+However, only ``Full'' index will have unit standard deviation per definition.
+From the regression, we also use the explained variance of each pattern as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
 
 \begin{figure*}
-\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period]{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979–2018 period.
+\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period]{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 Our method assumes linearity in the asymmetric component of the SAM.
-That is, we assume that zonal symmetries associated with positive SAM
-are almost opposite and of the same magnitude to the ones associated
-with negative SAM. \citet{fogt2012}'s composites (their Figure 4)
-suggest that this might not be entirely valid, although we argue that
-much of that apparent non-linearity is due to the heterogenous nature of
-the selected years for constructing the composites. Using our data (from
-1979 to 2018), seasonal composites of zonal anomalies of 700 hPa
-geopotential height for SAM+ (Full SAM index greater than 1 standard
-deviation) and SAM- (less than minus 1 standard deviation) show
-relatively high pattern correlations for all seasons and are visually
-very linear (Figure A9). Therefore, we believe that our method is a
-reasonable approximation of the phenomenon.
+That means we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM.
+\citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
+Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures A3 and A4).
+Therefore, we believe that our method is a reasonable approximation of the phenomenon.
 
-By computing a single EOF pattern using data for all months we are
-assuming that the zonal anomalies of the SAM are the same in all seasons
---December to February (DJF), March to May (MAM), June to August (JJA)
-and September to November (SON). Geopotential zonal anomalies computed
-by projecting the first EOF \emph{of each season} are very similar to
-each other (Figure A10) and show pattern correlations between 0.65 (DJF
-with JJA) and 0.9 (between MAM and SON). Based on this, we believe that
-our initial assumption is not unreasonable.
+By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
+Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure A5, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
+In the stratosphere (Figure A5, row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree~in comparison with the rest of the year.
+Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
+Based on this, we believe that our initial assumption is not unreasonable throught most of the year with the exception of DJF in the stratosphere.
 
-Finally, we assume that the zonally asymmetric pattern is stationary in
-time. \citet{silvestri2009} suggest that this might not be the case
-between 1958 and 2004 but the period we analyse is much shorter
-(1979-2018) so it's unlikely that we could observe significant changes.
-Moreover, zonal asymmetry of the spatial patterns for the two halves of
-the period (1979 to 1998 and 1999 to 2018) show no systematic change
-(Figure A11).
+Finally, we assume that the zonally asymmetric pattern is stationary in time.
+\citet{silvestri2009} suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes.
+Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure A6).
 
@@
 
-We perform linear regressions to quantify the association between the
-SAM indices and other variables. Since the Asymmetric and Symmetric SAM
-indices are significantly correlated with each other, to capture the
-variability explained uniquely by each index we use one multiple linear
-regression instead of two simple linear regressions. To obtain the
-linear coefficients of a variable \(X\) (geopotential, temperature,
-precipitation, etc\ldots{}) with the Asymmetric SAM (\(SAM_a\)) and
-Symmetric SAM (\(SAM_s\)) we fit the equation
+We perform linear regressions to quantify the association between the SAM indices and other variables.
+Since the Asymmetric and Symmetric SAM indices are significantly correlated with each other, to capture the variability explained uniquely by each index we use one multiple linear regression instead of two simple linear regressions.
+To obtain the linear coefficients of a variable \(X\) (geopotential, temperature, precipitation, etc\ldots{}) with the Asymmetric SAM (\(SAM_a\)) and Symmetric SAM (\(SAM_s\)) we fit the equation
 
@@
 
-where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is
-the time, \(\alpha\) and \(\beta\) are the linear regression
-coefficients, \(X_0\) and \(\epsilon\) are the constant and error terms.
-From this equation, \(\alpha\) represents the (linear) association of
-\(X\) with the variability of the Asymmetric SAM that is not explained
-by the variability of the Symmetric SAM; i.e.~it is proportional to the
-partial correlation of \(X\) and the Asymmetric SAM, controlling for the
-effect of the Symmetric SAM and vice versa for \(\beta\). When
-performing a separate regression for each trimester (DJF, MAM, JJA,
-SON), we average of the relevant variables seasonally for each year and
-trimester before computing the regression.
+where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients, \(X_0\) and \(\epsilon\) are the constant and error terms.
+From this equation, \(\alpha\) represents the (linear) association of \(X\) with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; i.e.~it is proportional to the partial correlation of \(X\) and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and vice versa for \(\beta\).
+When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we average the relevant variables seasonally for each year and trimester before computing the regression.
 
-At 2.5\degree by 2.5\degree resolution, a single regression field is
-composed of thousands of regressions. In such case, using p-values to
-test for significance leads to misleading results
-\citep{walker1914, katz1991}. While there are multiple proposed
-solutions in the literature, \citet{wilks2016} suggests that adjusting
-p-values by controlling for the False Discovery Rate
-\citep{benjamini1995} is a simple and effective method to ameliorate
-this issue. Therefore, p-values showed in regression fields are all
-adjusted following \citet{benjamini1995}.
+At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions.
+In such case, using p-values to test for significance leads to misleading results \citep{walker1914, katz1991}.
+While there are multiple proposed solutions in the literature, \citet{wilks2016} suggests that adjusting p-values by controlling for the False Discovery Rate \citep{benjamini1995} is a simple and effective method to ameliorate this issue.
+Therefore, p-values showed in regression fields are all adjusted following \citet{benjamini1995}.
 
-We computed linear trends by Ordinary Least Squares and the 95\%
-confidence interval assuming a t-distribution of the appropriate
-residual degrees of freedom.
+We computed linear trends by Ordinary Least Squares and the 95\% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
+To compute the amplitude of the zonal waves we computed the Fourier transform of the spatial field at each latitude circle.
 
-To compute the amplitude of the zonal waves we computed the Fourier
-transform of the spatial field at each latitude circle.
+We computed density probability estimates using a gaussian kernel of optimal bindwidth according to \citet{sheather1991}.
 
@@
 
-We performed all analysis in this paper using the R programming language
-\citep{rcoreteam2020}, using the data.table package \citep{dowle2020}
-and the metR package \citep{campitelli2020}. All graphics are made using
-ggplot2 \citep{wickham2009}. We downloaded data from ranalysis using the
-ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi
-package \citep{albers2020}. The paper was rendered using knitr and
-rmarkdown \citep{xie2015, allaire2019}.
+We performed all analysis in this paper using the R programming language \citep{rcoreteam2020}, using the data.table package \citep{dowle2020} and the metR package \citep{campitelli2020}.
+All graphics are made using ggplot2 \citep{wickham2009}.
+We downloaded data from ranalysis using the ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi package \citep{albers2020}.
+The paper was rendered using knitr and rmarkdown \citep{xie2015, allaire2019}.
 
@@
 \includegraphics{asymsam-timeseries-1} \caption[Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa]{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
-To the left, probability density estimate of each index computed using a gaussian kernel of optimal bindwidth according to \cite{sheather1991}.
+To the right, probability density estimate of each index.
 Series are standarised by the standard deviation of the Full SAM at each level.}\label{fig:asymsam-timeseries}
@@
 
-We first asses the temporal evolution of the Asymmetric SAM and
-Symmetric SAM. Figure \ref{fig:asymsam-timeseries} shows the
-corresponding time series for 700 hPa and 50 hPa and their corresponding
-density estimates. We selected these two levels as representative of the
-tropospheric and stratospheric variability respectively. As will be
-shown later, both indices are highly coherent within each atmospheric
-layer, therefore is reasonable to take one level as representative of
-each layer.
+We first asses the temporal evolution of the Asymmetric SAM and Symmetric SAM.
+Figure \ref{fig:asymsam-timeseries} shows the corresponding time series for 700 hPa and 50 hPa and their corresponding density estimates.
+We selected these two levels as representative of the tropospheric and stratospheric variability respectively.
+As will be shown later, both indices are highly coherent within each atmospheric layer, therefore is reasonable to take one level as representative of each layer.
 
-Month-to-month variability is evident for both indices, with noisy
-variations in the low frequency. At first glance the series can be
-distinguished by their distributions. Compared to the tropospheric
-indices, the stratospheric indices are much more long-tailed; that is,
-extreme values (both negative and positive) abound. The Asymmetric SAM
-series have both more variability in the higher frequencies than the
-Symmetric SAM series.
+Month-to-month variability is evident for both indices, with noisy variations in the low frequency.
+At first glance the series can be distinguished by their distributions.
+Compared to the tropospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound.
+The Asymmetric SAM series have both more variability in the higher frequencies than the Symmetric SAM series.
 
-The stratospheric Symmetric SAM varies strongly with a two-year period,
-which can be seen by spectral analysis (Figure A3). This might suggests
-a link between stratospheric SAM variability and the Quasi-Biennial
-Oscillation \citep{baldwin2001b}. There is a local peak at 2 years in
-the periodogram of the tropospheric Symmetric SAM also, although it's
-not statistically significant. In the troposphere the most significant
-peak of variability is found in the Asymmetric index at around 3.6
-months.
+The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A2).
+This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation \citep{baldwin2001b}.
+There is a local peak at 2 years in the periodogram of the tropospheric Symmetric SAM also, although it's not statistically significant.
+In the troposphere the most significant peak of variability is found in the Asymmetric index at around 3.6 months.
 
-From Figure \ref{fig:asymsam-timeseries} we can see that the Asymmetric
-SAM and Symmetric SAM time series appear to be correlated. Moreover,
-looking at the extremes in the stratosphere, the Symmetric SAM series
-appears to lag the Asymmetric SAM series (see, for example, the positive
-events on late 1987). We show these correlations, across all the levels
-of the reanalysis for zero and -1 lag (Asymmetric SAM index leading the
-Symmetric SAM index), in Figure \ref{fig:cor-lev}. Zero-lag correlations
-between the Asymmetric SAM and Symmetric SAM series are relatively
-constant throughout the troposphere, fluctuating between 0.39 and 0.45.
-One-month-lag correlations are similarly constant but significantly
-reduced to around 0.17. In the stratosphere, zero-lag correlations drop
-to a minimum of 0.21 at 20 hPa and then it increases again monotonically
-with height up to the uppermost level of the reanalysis (although
-results near the top of the models are to be interpreted with care). At
-the same time, one-month-lag correlations increase with height. As a
-consequence, stratospheric Asymmetric SAM index tend to precede
-corresponding Symmetric SAM index.
+From Figure \ref{fig:asymsam-timeseries} we can see that the Asymmetric SAM and Symmetric SAM time series appear to be correlated.
+Moreover, looking at the extremes in the stratosphere, the Symmetric SAM series appears to lag the Asymmetric SAM series (see, for example, the positive events on late 1987).
+We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric SAM index leading the Symmetric SAM index), in Figure \ref{fig:cor-lev}.
+Zero-lag correlations between the Asymmetric SAM and Symmetric SAM series are relatively constant throughout the troposphere, fluctuating between 0.39 and 0.45.
+One-month-lag correlations are similarly constant but significantly reduced to around 0.17.
+In the stratosphere, zero-lag correlations drop to a minimum of 0.21 at 20 hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care).
+At the same time, one-month-lag correlations increase with height.
+As a consequence, stratospheric Asymmetric SAM index tend to precede corresponding Symmetric SAM index.
 
 \begin{figure}
-\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979–2018 period]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979–2018 period.}\label{fig:cor-lev}
+\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.}\label{fig:cor-lev}
 \end{figure}
@@
 \begin{figure*}
-\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979–2018 period]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979–2018 period.}\label{fig:cross-correlation}
+\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.}\label{fig:cross-correlation}
 \end{figure*}
 
-Figure \ref{fig:cross-correlation}a shows (zero-lag) cross-correlation
-across levels for the Full, Symmetric and Asymmetric SAM indices. For
-the Full SAM (panel a), high values below 100 hPa reflect the vertical
-(zero-lag) coherency throughout the troposphere. Above 100 hPa
-correlation between levels falls off more rapidly, indicating less
-coherent (zero-lag) variability. Therefore, there is a non negligible
-correlation between the troposphere and the lower-to-middle
-stratosphere. Examining panels b and c, we see that the Asymmetric and
-Symmetric SAM share the same high level of coherency in the troposphere
-but they differ in their stratospheric behaviour. Stratospheric
-coherency is stronger for the Asymmetric SAM than the Symmetric SAM. The
-stratospheric Symmetric SAM seems to connect more strongly to the
-troposphere than the Asymmetric SAM.
+Figure \ref{fig:cross-correlation}a shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
+For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
+Above 100 hPa correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
+Therefore, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
+Examining panels b and c, we see that the Asymmetric and Symmetric SAM share the same high level of coherency in the troposphere but they differ in their stratospheric behaviour.
+Stratospheric coherency is stronger for the Asymmetric SAM than the Symmetric SAM.
+The stratospheric Symmetric SAM seems to connect more strongly to the troposphere than the Asymmetric SAM.
 
 \begin{figure*}
-\includegraphics{trends-1} \caption[Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+\includegraphics{trends-1} \caption[Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
 Shading indicates the 95\% confidence interval from a t-distribution.}\label{fig:trends}
@@
 
-The linear trends for each of the indices (Full SAM, Symmetric SAM and
-Asymmetric SAM) were evaluated for the complete period 1979-2018 at each
-level (Figure \ref{fig:trends}) for the whole year and separated by
-trimesters. The Full SAM index presents a statistically significant
-trend (panel a.1) that extends throughout the troposphere up to about 50
-hPa and reaches its maximum value at 100 hPa. The seasonal trends (rest
-of column a) indicate that positive trends are present in Autumn and
-particularly in Summer, where the 100 hPa maximum is much more defined.
-In Winter and Spring, we detect no statistically significant trend. This
-is consistent with the results of previous studies, which find clear
-positive trends in Summer, weaker in Autumn and no trends in the other
-seasons \citep[e.g.][ and references therein]{fogt2020} using indices of
-the SAM based on surface or near-surface circulation.
+The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979 -- 2018 at each level (Figure \ref{fig:trends}) for the whole year and separated by trimesters.
+The Full SAM index presents a statistically significant trend (panel a.1) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
+The seasonal trends (rest of column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
+In Winter and Spring, we detect no statistically significant trend.
+This is consistent with the results of previous studies, which find clear positive trends in Summer, weaker in Autumn and no trends in the other seasons \citep[e.g.][ and references therein]{fogt2020} using indices of the SAM based on surface or near-surface circulation.
 
-By separating the SAM signal in its asymmetric and symmetric parts, we
-can not only see that these trends are almost entirely due to the
-symmetric component (column b vs.~column c), but in some cases the
-trends become more clear. In Summer, the Asymmetric SAM has a
-statistically non significant negative trend in the middle troposphere
-that obscures the trend in the Full SAM index; as a result, trends
-computed using only the Symmetric component are more clear (compare the
-shading region in panel a.2 and c.2). In Autumn, the Symmetric SAM
-reveals a statistically significant positive trend in the stratosphere
-that is not significant using the Full SAM index.
+By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.~column c), but in some cases the trends become more clear.
+In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel a.2 and c.2).
+In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full SAM index.
 
-We stress that these are only linear trends during the whole period and
-the absence of a statistically significant signal should not be taken as
-evidence of no systematic change. In particular, going back to Figure
-\ref{fig:asymsam-timeseries}, we can see an evident change in the
-stratospheric Asymmetric component (red line in panel a) between the
-90's, when we see a dominance of extreme negative values, and the 00's,
-when we see the inverse. This change is restricted to the Winter months:
-the linear trend for Winter starting in 1990 for the Asymmetric
-component at 50hPa is \(0.37 \pm 0.22\).
+We stress that these are only linear trends during the whole period and the absence of a statistically significant signal should not be taken as evidence of no systematic change.
+In particular, going back to Figure \ref{fig:asymsam-timeseries}, we can see an evident change in the stratospheric Asymmetric component (red line in panel a) between the 90's, when we see a dominance of extreme negative values, and the 00's, when we see the inverse.
+This change is restricted to the Winter months: the linear trend for Winter starting in 1990 for the Asymmetric component at 50hPa is \(0.37 \pm 0.22\).
 
-Figure \ref{fig:r-squared-trend} shows decadal trends for the explained
-variance of each index. There is no evidence of a significant trend in
-the stratosphere. In the troposphere, there is a positive trend for the
-Asymmetric SAM and not significant trend for the Symmetric SAM. This
-suggest that the SAM has become more asymmetric in the period from 1979
-to 2018. However, the change is slight, around 1\% increased explained
-variance per decade.
+Figure \ref{fig:r-squared-trend} shows decadal trends for the explained variance of each index.
+There is no evidence of a significant trend in the stratosphere.
+In the troposphere, there is a positive trend for the Asymmetric SAM and not significant trend for the Symmetric SAM.
+This suggest that the SAM has become more asymmetric in the period from 1979 to 2018.
+However, the change is slight, around 1\% increased explained variance per decade.
 
 \begin{figure}
-\includegraphics{r-squared-trend-1} \caption[Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979-2018]{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979-2018.
+\includegraphics{r-squared-trend-1} \caption[Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018]{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018.
 Shading indicates the 95\% confidence interval.}\label{fig:r-squared-trend}
@@
 \begin{figure*}
-\includegraphics{2d-regr-1} \caption[Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979–2018 period]{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979–2018 period.
+\includegraphics{2d-regr-1} \caption[Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period]{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period.
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices.
@@
 
-To show if, and to what extent, the Asymmetric and Symmetric SAM indices
-indeed capture the asymmetric and symmetric component of the SAM
-respectively, we computed the spatial regression of geopotential height
-anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa
-levels. Figure \ref{fig:2d-regr} shows these regressions. Regression
-coefficients in column a are computed using the Full SAM. Regression
-coefficients in columns b and c are computed using multiple regression
-using the Asymmetric and Symmetric indices at the same time. Thus, they
-are to be interpreted as the patterns associated with each index,
-removing the variability (linearly) explained by the other index.
+To show if, and to what extent, the Asymmetric and Symmetric SAM indices indeed capture the asymmetric and symmetric component of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa levels.
+Figure \ref{fig:2d-regr} shows these regressions.
+Regression coefficients in column a are computed using the Full SAM.
+Regression coefficients in columns b and c are computed using multiple regression using the Asymmetric and Symmetric indices at the same time.
+Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other index.
 
-In the stratosphere, the spatial pattern associated with the Full SAM is
-more clearly dominated by a zonally symmetric, monopolar structure
-(panel a.1) which is, however, not perfectly centred in the South Pole.
-The monopole obtained by the regression pattern for Symmetric SAM (panel
-c.1) is much more symmetric and the shift from total symmetry is
-captured by the regression pattern of the Asymmetric SAM as a wave-1
-with maximum anomalies above the Belinghausen Sea on the Western
-Hemisphere and and Davis Sea in the Eastern Hemisphere (panel b.1).
+In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centred in the South Pole.
+The monopole obtained by the regression pattern for Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davis Sea in the Eastern Hemisphere (panel b.1).
 
-In the troposphere, panel a.2 shows the well known combination of
-zonally symmetrical annular mode with zonal asymmetries in the form of a
-wave-3 \citep{fogt2012}. The regression using the Asymmetric and
-Symmetric SAM indices successfully disentangle both structures. The
-Asymmetric SAM index gives rise to a cleaner zonal wave (panel b.2) and
-the Symmetric SAM index is associated with an annular mode, almost
-devoid of zonal asymmetries (panel c.2). The wave-3 pattern observed in
-panel b.2 is rotated by half a wavelength from the average position of
-the mean wave-3 pattern associated with \citet{raphael2004}'s ZW3 index,
-whose reference locations are marked with points in the figure. Thus,
-the tropospheric Asymmetric SAM index represents a zonal displacement in
-the position of the climatological wave-3 pattern.
+In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 \citep{fogt2012}.
+The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
+The Asymmetric SAM index gives rise to a cleaner zonal wave (panel b.2) and the Symmetric SAM index is associated with an annular mode, almost devoid of zonal asymmetries (panel c.2).
+The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern associated with \citet{raphael2004}'s ZW3 index, whose reference locations are marked with points in the figure.
+Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
 
 \begin{figure*}
-\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
+\includegraphics{wave-amplitude-1} \caption[Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean]{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
 \end{figure*}
 
-The amplitude of the first zonal wave numbers at each latitude at 50 hPa
-and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave
-number zero represents the amplitude of the zonal mean. Zonal wave
-amplitudes of the spatial pattern described by the Full SAM index
-(column a) are dominated by the zonal mean (wave-number 0) at both
-levels. However, zonal waves are important, particularly North of
-50\degree S, with wave-number 1 clearly dominating at 50 hPa (panel a.1)
-and a more equal mix of waves at 700 hPa (panel a.2). Column b shows
-that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the
-stratosphere (panel b.1), while in the troposphere it is composed of
-zonal waves 3 to 1 in decreasing level of importance (panel b.2) with
-negligible amplitude of the zonal mean. The Symmetric SAM, on the other
-hand, it's almost entirely composed of zonal mean at both levels (column
-c), with little to now contribution from zonal waves with wave-numbers 1
-to 3.
+The amplitude of the first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
+Zonal wave amplitudes of the spatial pattern described by the Full SAM index (column a) are dominated by the zonal mean (wave-number 0) at both levels.
+However, zonal waves are important, particularly North of 50\degree S, with wave-number 1 clearly dominating at 50 hPa (panel a.1) and a more equal mix of waves at 700 hPa (panel a.2).
+Column b shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b.1), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel b.2) with negligible amplitude of the zonal mean.
+The Symmetric SAM, on the other hand, it's almost entirely composed of zonal mean at both levels (column c), with little to now contribution from zonal waves with wave-numbers 1 to 3.
 
-We can see that the amplitude and latitudinal distribution zonal waves
-in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric
-SAM on the other correspond almost exactly to the amplitude and
-latitudinal distribution in the Full SAM. This confirms the correct
-decomposition of the SAM in its symmetric and asymmetric components.
-
-Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent
-that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which
---as mentioned before-- is larger in the Western Hemisphere than in the
-Eastern Hemisphere.
+We can see that the amplitude and latitudinal distribution zonal waves in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric SAM on the other correspond almost exactly to the amplitude and latitudinal distribution in the Full SAM.
+This confirms the correct decomposition of the SAM in its symmetric and asymmetric components.
+Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
 
@@
 \includegraphics{vertical-regression-1} \caption[Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM)]{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
-(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979–2018 period.}\label{fig:vertical-regression}
+(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979 -- 2018 period.}\label{fig:vertical-regression}
 \end{figure}
 
-To analyse the vertical structure of the geopotential height anomalies
-associated with the asymmetric SAM index, we show a vertical cross
-section of regressions of mean geopotential height between 65\degree S
-and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for
-the 700 hPa Asymmetric SAM index (panel b) (Figure
-\ref{fig:vertical-regression}). The geopotential height anomalies
-associated with the stratospheric Asymmetric SAM (panel a) are clearly
-constrained to the stratosphere, which underscores the uncoupling
-between the stratospheric and tropospheric Asymmetric SAM. The vertical
-structure of this signal tilts about 60\degree to the West between 100
-hPa and 1 hPa, suggesting baroclinic processes. Interestingly, the
-signal in the stratosphere maximises near 10 hPa despite using the 50
-hPa index for the regression.
+To analyse the vertical structure of the geopotential height anomalies associated with the asymmetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
+The geopotential height anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
+The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes.
+Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
 
-The tropospheric Asymmetric SAM (panel b) has significant signals that
-extend upwards to the uppermost levels of the reanalysis. In the
-troposphere, the wave-3 structure is equivalent barotropic with maximum
-amplitude at roughly 250 hPa. The anomalies are much more intense in the
-Western hemisphere, where they extent into the stratosphere. In the
-Eastern hemisphere the wave-3 signal is weaker and confined to the
-troposphere while negative anomalies dominate in the stratosphere. So,
-while the tropospheric Asymmetric SAM index is associated with
-stratospheric geopotential anomalies, these do not project strongly onto
-the stratospheric Asymmetric SAM. The structures shown in panels a and b
-in Figure \ref{fig:vertical-regression} are robust to the choice of
-index level. For any stratospheric (above 100 hPa) index, the resulting
-anomalies are very similar to the wave-1 structure with maximum near 10
-hPa in panel a. Conversely, for any tropospheric (below 100 hPa) index,
-the result is very similar to panel b. The patterns mainly change in
-amplitude (not shown).
+The tropospheric Asymmetric SAM (panel b) has significant signals that extend upwards to the uppermost levels of the reanalysis.
+In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
+The anomalies are much more intense in the Western hemisphere, where they extent into the stratosphere.
+In the Eastern hemisphere the wave-3 signal is weaker and confined to the troposphere while negative anomalies dominate in the stratosphere.
+So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
+The structures shown in panels a and b in Figure \ref{fig:vertical-regression} are robust to the choice of index level.
+For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in panel a.
+Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b.
+The patterns mainly change in amplitude (not shown).
 
-The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very
-similar to the Pacific-South American Pattern \citep{mo1987, kidson1988}
-which is a teleconnection pattern associated with the ENSO
-\citep{karoly1989}. Indeed, \citet{fogt2011} showed that there is a
-significant relationship between the SAM and the ENSO. The correlation
-between the Full SAM and the ENSO as measured by the Oceanic Niño Index
-\citep{bamston1997} (ONI) is -0.16 (p-value =
-\ensuremath{2.8\times 10^{-4}}). Consistent with \citet{fan2007}, we
-show that this relationship is captured entirely by the Asymmetric SAM,
-as this index has a partial correlation of -0.26 (p-value =
-\ensuremath{6.3\times 10^{-9}}) with the ONI controlling for the effect
-of the Symmetric SAM, whereas the Symmetric SAM's partial correlation
-with the ONI is essentially null (0.019; p-value = 0.67). We performed
-the same analysis using the Multivariate ENSO Index \citep{wolter2011}
-and the Southern Oscillation Index \citep{ropelewski1987} to conclude
-that these results do not depend on the ENSO index used.
+The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}.
+Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO.
+The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is -0.16 (p-value = \ensuremath{2.8\times 10^{-4}}).
+Consistent with \citet{fan2007}, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of -0.26 (p-value = \ensuremath{6.3\times 10^{-9}}) with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67).
+We performed the same analysis using the Multivariate ENSO Index \citep{wolter2011} and the Southern Oscillation Index \citep{ropelewski1987} to conclude that these results do not depend on the ENSO index used.
 
-\subsection{Impacts}
+\subsection{Surface Impacts}
 
@@
 \begin{figure*}
-\includegraphics{regr-air-season-1} \caption[Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979–2018 period]{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979–2018 period.
+\includegraphics{regr-air-season-1} \caption[Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period]{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 
-To see if there are different surface impacts associated with the
-asymmetric and symmetric SAM circulation we regress surface temperature
-and precipitation onto each of the three SAM indices at 700 hPa. As
-shown in previous sections, the three indices are highly coherent in the
-troposphere, so we select this level to represent the tropospheric
-circulation for consistency with previous studies.
+To see if there are different surface impacts associated with the asymmetric and symmetric SAM circulation we regress surface temperature and precipitation onto each of the three SAM indices at 700 hPa.
+As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
 
-Figure \ref{fig:regr-air-season} shows regression coefficients of each
-index at 700 hPa with surface temperature for each trimester. In Summer,
-positive values of the Full SAM index (panel a.1) are associated with
-negative temperature anomalies near Antarctica which are surrounded by a
-ring of positive anomalies. The ring is not zonally symmetric, as there
-are four clear local maximums around 30\degree W, 120\degree W,
-150\degree E and 90\degree E. In the tropics, there are negative
-anomalies in the equatorial Pacific, consistent with the negative
-correlation between SAM and ENSO. Panels b.1 and c.1 show temperature
-anomalies associated with positive values of the Asymmetric and
-Symmetric SAM, respectively. Both the local maximums in the ring and the
-anomalies in the Pacific regions are present mostly on the Asymmetric
-SAM regression map, while temperature patterns linked to positive
-Symmetric SAM show a more zonally consistent ring and less relation to
-the tropics. Noticeable, temperature anomalies in the Indian ocean,
-South Africa and Australia are strongly related to Asymmetric SAM. This
-signal is not present in the regression pattern with the Full SAM.
-Spring (row 4) features similar patterns but of smaller magnitude, with
-less regions where regressed anomalies have statistical significance.
+Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface temperature for each trimester.
+In Summer, positive values of the Full SAM index (panel a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
+The ring is not zonally symmetric, as there are four clear local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E.
+In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
+Panels b.1 and c.1 show temperature anomalies associated with positive values of the Asymmetric and Symmetric SAM, respectively.
+Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the Asymmetric SAM regression map, while temperature patterns linked to positive Symmetric SAM show a more zonally consistent ring and less relation to the tropics.
+Noticeable, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to Asymmetric SAM.
+This signal is not present in the regression pattern with the Full SAM.
+Spring (row 4) features similar patterns but of smaller magnitude, with less regions where regressed anomalies have statistical significance.
 
-In Autumn and Winter (rows 2 and 3) the positive ring is only present
-through its local maximums in the regression with the Full SAM. There
-are also negative anomalies in Southern Australia, and positive
-anomalies over New Zealand and Southern South America. These patterns
-are not significant in the sense that there are no areas with p-values
-below 0.05 when controlling for FDR following \citet{wilks2016}.
-However, repeating this analysis with 2-meter temperature from ERA5
-resulted in similar patterns that were statistically significant (not
-shown). Moreover, similar features were observed in station measurements
-by \citet{jones2019}, although using data from 1957 to 2016.
+In Autumn and Winter (rows 2 and 3) the positive ring is only present through its local maximums in the regression with the Full SAM.
+There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
+These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following \citet{wilks2016}.
+However, repeating this analysis with 2-meter temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
+Moreover, similar features were observed in station measurements by \citet{jones2019}, although using data from 1957 to 2016.
 
-The pattern of negative anomalies in the pole surrounded by positive
-anomalies roughly seen in all seasons --although with varying intensity
-and small-scale details-- translates to enhanced meridional temperature
-gradient maximised in the zero line, which is consistent with the
-intensification and poleward migration of the westerlies commonly linked
-to the SAM through thermal wind balance. It's then not surprising to see
-it more clearly in association with the Symmetric SAM (at least in
-Summer and Spring).
+The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons -- although with varying intensity and small-scale details -- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
+It's then not surprising to see it more clearly in association with the Symmetric SAM (at least in Summer and Spring).
 
-Figure \ref{fig:regr-air-season} column b can be partially compared with
-Figure 11 from \citet{fogt2012}. Although they used station data from
-1958 to 2001, main features are reproduced here, such as the strong
-signal in New Zealand and Australia in Summer and Spring.
+Figure \ref{fig:regr-air-season} column b can be partially compared with Figure 11 from \citet{fogt2012}.
+Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.
 
-Regression of the SAM indices with seasonal mean precipitation and 700
-hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania}
-and \ref{fig:pp-regr-america} for Australasia and South America
-respectively. South Africa is not shown because no significant signal
-was detected there.
+Regression of the SAM indices with seasonal mean precipitation and 700 hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
+South Africa is not shown because no significant signal was detected there.
 
 \begin{figure*}
-\includegraphics{pp-regr-oceania-1} \caption[Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979–2018 period]{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979–2018 period.
+\includegraphics{pp-regr-oceania-1} \caption[Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period]{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
@@
 
-In Australia, the annual regression shows that the Full SAM index is
-positively associated with precipitation in the Southeastern region
-(Figure \ref{fig:pp-regr-oceania} panel a.1), which reproduces the
-results from \citet{gillett2006}. The separation between Asymmetric and
-Symmetric SAM suggest that this positive anomaly is explained by the
-Symmetric SAM only in the East coast (panel c.1). Geopotential anomalies
-associated with this index (black contours) are indicative of easterly
-flow from the Tasman Sea, which could explain the positive anomalies in
-precipitation as found by \citet{hendon2007}. The Asymmetric SAM appears
-related to increased precipitation in the West coast of Southeastern
-Australia (panel b.2), which could similarly be explained by the
-anomalous westerly circulation transporting moist air to the continent
-from the Indian Ocean.
+In Australia, the annual regression shows that the Full SAM index is positively associated with precipitation in the Southeastern region (Figure \ref{fig:pp-regr-oceania} panel a.1), which reproduces the results from \citet{gillett2006}.
+The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (panel c.1).
+Geopotential anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by \citet{hendon2007}.
+The Asymmetric SAM appears related to increased precipitation in the West coast of Southeastern Australia (panel b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
 
-The seasonal-level regressions show statistically significant anomalies
-only in Spring, when positive Full SAM is associated with positive
-precipitation anomalies in Eastern Australia (panel a.5). In this
-trimester the Symmetric SAM seems to be associated with precipitation in
-a relatively reduced area of the East Coast (panel c.5) while the
-positive precipitation anomalies related with positive Asymmetric SAM
-affect all Eastern Australia (panel b.5).
+The seasonal-level regressions show statistically significant anomalies only in Spring, when positive Full SAM is associated with positive precipitation anomalies in Eastern Australia (panel a.5).
+In this trimester the Symmetric SAM seems to be associated with precipitation in a relatively reduced area of the East Coast (panel c.5) while the positive precipitation anomalies related with positive Asymmetric SAM affect all Eastern Australia (panel b.5).
 
-In Summer, positive Full SAM index is associated with with positive
-precipitation anomalies in Western and Eastern Australia, particularly
-in the North East (panel a.2). The Eastern part being dominated by the
-relationship with the Symmetric SAM and the Western, by the Asymmetric
-SAM. In Autumn, the regression with Full SAM shows positive values in
-the North, similar to Summer, and a broad area of positive values in the
-North-East to South-West direction. This structure seems to be
-associated with the Symmetric SAM, while the Northern positive values
-are associated with the Asymmetric SAM. In Winter we see the same NE to
-SW aligned anomaly (although with much reduced amplitude) that is also
-present only in relation with the Symmetric SAM. None of these
-regression coefficients are statistically significant at the 95\% level.
-The Spring signal is broadly consistent with \citet{hendon2007}, but
-whereas \citet{hendon2007} also detected a strong signal in Summer,
-panel a.2 shows no statistically significant association (although the
-coefficients have the consistent sign).
+In Summer, positive Full SAM index is associated with with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (panel a.2).
+The Eastern part being dominated by the relationship with the Symmetric SAM and the Western, by the Asymmetric SAM.
+In Autumn, the regression with Full SAM shows positive values in the North, similar to Summer, and a broad area of positive values in the North-East to South-West direction.
+This structure seems to be associated with the Symmetric SAM, while the Northern positive values are associated with the Asymmetric SAM.
+In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the Symmetric SAM.
+None of these regression coefficients are statistically significant at the 95\% level.
+The Spring signal is broadly consistent with \citet{hendon2007}, but whereas \citet{hendon2007} also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coefficients have the consistent sign).
 
 \begin{figure*}
-\includegraphics{pp-regr-america-1} \caption{Same as Figure \ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
+\includegraphics{pp-regr-america-1} \caption[Same as Figure \\ref{fig:pp-regr-oceania} but for South America]{Same as Figure \\ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
 \end{figure*}
 
-In South America (Figure \ref{fig:pp-regr-america}), the annual-level
-regression shows that positive SAM is associated with statistically
-significant precipitation decrease in Southeastern South America (SESA)
-and Southern Chile and non-significant increase in South Brazil, near
-the South Atlantic Convergence Zone (SACZ) (panel a.1). Panels b.1 and
-c.1 show a remarkably clean separation between the Asymmetric SAM
---associated with the Southeastern South American and Southern Brazilian
-signals-- and the Symmetric SAM --associated with the signal in Southern
-Chile.
+In South America (Figure \ref{fig:pp-regr-america}), the annual-level regression shows that positive SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1).
+Panels b.1 and c.1 show a remarkably clean separation between the Asymmetric SAM -- associated with the Southeastern South American and Southern Brazilian signals -- and the Symmetric SAM -- associated with the signal in Southern Chile.
 
-Except Winter, seasonal-level regressions mirror this same pattern. Even
-if not statistically significant, they all show negative values in
-Southeastern South America and Southern Chile along with positive values
-in Southern Brazil in relation with the Full SAM. The separation of
-these features between the Asymmetric SAM and Symmetric SAM regression
-maps is also rather consistent.
+Except Winter, seasonal-level regressions mirror this same pattern.
+Even if not statistically significant, they all show negative values in Southeastern South America and Southern Chile along with positive values in Southern Brazil in relation with the Full SAM.
+The separation of these features between the Asymmetric SAM and Symmetric SAM regression maps is also rather consistent.
 
-The anomalous circulation at 700 hPa associated with the Symmetric SAM
-(panel c.1) indicate anomalous Easterly flow over Southern Chile. This
-leads to reduced influx of moist air from the Pacific Ocean which, is
-the main source of precipitable water in that region
-\citep{garreaud2007}. On the other hand, the anomalous circulation
-associated with positive values of Asymmetric SAM (panel b.1) in the
-Atlantic is anticyclonic in the South and cyclonic in the North. This
-creates anomalous South-Easterly flow over Southeastern South America,
-which inhibits the flow of the Low Level Jet to the region
-\citep[\citet{zamboni2010}]{silvestri2009}. This same pattern was found
-to be associated with increased precipitation in Southern Brazil during
-South Atlantic Convergence Zone events \citep{rosso2018}. There is a
-small area of increased precipitation with SAM near central Argentina
-which is also present in the station-based analysis by
-\citet{gillett2006} and that is explained by the Asymmetric SAM.
+The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous Easterly flow over Southern Chile.
+This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region \citep{garreaud2007}.
+On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
+This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region \citep{silvestri2009, zamboni2010}.
+This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events \citep{rosso2018}.
+There is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} and that is explained by the Asymmetric SAM.
 
@@
 
-In this study we characterise the temporal and spatial variability of
-the zonally symmetric and asymmetric structure of the SAM. By projecting
-monthly geopotential fields at each level with the corresponding
-asymmetric and symmetric pattern, we created two indices for
-representing the zonally asymmetric and symmetric contributions of the
-SAM respectively.
+In this study we characterise the temporal and spatial variability of the zonally symmetric and asymmetric structure of the SAM.
+By projecting monthly geopotential fields at each level with the corresponding asymmetric and symmetric pattern, we created two indices for representing the zonally asymmetric and symmetric contributions of the SAM respectively.
 
-As expected, the Asymmetric SAM index correlates strongly with the
-Symmetric SAM index. In the troposphere, this correlation is maximum at
-zero lag, while in the stratosphere is maximised with the Asymmetric SAM
-leading the Symmetric SAM by one month. Since most indices of the SAM
-are calculated using surface or near-surface conditions, this result
-would suggest that they might not be sensitive to the most dramatic
-changes in SAM variability.
+As expected, the Asymmetric SAM index correlates strongly with the Symmetric SAM index.
+In the troposphere, this correlation is maximum at zero lag, while in the stratosphere is maximised with the Asymmetric SAM leading the Symmetric SAM by one month.
+Since most indices of the SAM are calculated using surface or near-surface conditions, this result would suggest that they might not be sensitive to the most dramatic changes in SAM variability.
 
-The two-year periodicity we found in the stratospheric Symmetric SAM
-might point to a link between the SAM and the Quasi Biennial
-Oscillation. There is evidence of influence between the QBO and the
-Northern Annular Mode \citep[e.g.][\citet{watson2014},
-\citet{zhang2020}]{holton1980}, so it's not unlikely that the SAM would
-be similarly connected. However establishing this link would require
-further research.
+The two-year periodicity we found in the stratospheric Symmetric SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
+There is evidence of influence between the QBO and the Northern Annular Mode \citep[e.g.][]{holton1980, watson2014, zhang2020}, so it's not unlikely that the SAM would be similarly connected.
+However establishing this link would require further research.
 
-We observe a positive trend towards positive SAM in Summer and Autumn,
-As was documented by previous studies, such as \citet{fogt2020} (and
-references therein) for surface levels. We show that these trends are
-maximised at the 100 hPa level and are explained by the zonally
-symmetric component. We also find a statistically significant positive
-trend in the Symmetric component of the SAM in the stratosphere that is
-not apparent in the Full SAM index. In contrast to \citet{fogt2012} we
-find some evidence of the SAM becoming more zonally asymmetric, as there
-is a slight positive trend in the variance explained by the as the
-Asymmetric SAM explains an increasingly proportion of the total
-variance.
+We observe a positive trend towards positive SAM in Summer and Autumn, As was documented by previous studies, such as \citet{fogt2020} (and references therein) for surface levels.
+We show that these trends are maximised at the 100 hPa level and are explained by the zonally symmetric component.
+We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not apparent in the Full SAM index.
+In contrast to \citet{fogt2012} we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
 
-In the troposphere, the spatial patterns of geopotential associated with
-the Symmetric SAM are much closer to being truly annular than the
-patterns associated with the Full SAM index. The Asymmetric SAM, on the
-other hand, describes a wave-3 pattern with maximum amplitude in the
-Pacific region and whose phase is rotated a quarter wavelength from the
-mean zonal wave 3 described by \citet{raphael2004}'s index. This pattern
-extends in the troposphere but its maximum is located at 250 hPa, which
-also could suggest that surface-based indices are not optimum for
-capturing this variability.
+In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being truly annular than the patterns associated with the Full SAM index.
+The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific region and whose phase is rotated a quarter wavelength from the mean zonal wave 3 described by \citet{raphael2004}'s index.
+This pattern extends in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimum for capturing this variability.
 
-This wave-3 pattern is similar to the Pacific-South American Pattern,
-which is a teleconnection pattern linked to ENSO variability. We found
-that the significant correlation that exists between the Full SAM index
-and the Oceanic Niño Index is captured entirely by the Asymmetric SAM
-index. This suggests that ENSO is linked to SAM exclusively through the
-variability in the latter's asymmetric component and thus, the
-Asymmetric SAM index could be a useful measure to further study that
-relationship.
+This wave-3 pattern is similar to the Pacific-South American Pattern, which is a teleconnection pattern linked to ENSO variability.
+We found that the significant correlation that exists between the Full SAM index and the Oceanic Niño Index is captured entirely by the Asymmetric SAM index.
+This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the Asymmetric SAM index could be a useful measure to further study that relationship.
 
-Temperature anomalies associated with the Full SAM broadly show a
-pattern of negative anomalies at polar latitudes surrounded by positive
-anomalies, but with many deviations from symmetry. The Asymmetric SAM
-index explains a big portion of these deviations. In particular, the
-positive phase of the Asymmetric SAM is associated with colder
-temperatures over Southern Brazil, South Africa and Southern Australia,
-as well as the negative anomalies in the equatorial Pacific consistent
-with the ENSO-SAM relationship. These negative anomalies are
-particularly clear in the DJF and SON trimesters, which include the
-months in which the ENSO teleconnection is more active
-\citep{cazes-boezio2003, fogt2011, cai2020a}.
+Temperature anomalies associated with the Full SAM broadly show a pattern of negative anomalies at polar latitudes surrounded by positive anomalies, but with many deviations from symmetry.
+The Asymmetric SAM index explains a big portion of these deviations.
+In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship.
+These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active \citep{cazes-boezio2003, fogt2011, cai2020a}.
 
-In Australia the Full SAM is associated with positive precipitation
-anomalies in South East and this is explained by the Symmetric SAM.
-However, the Asymmetric SAM is associated with a small area of positive
-precipitation anomalies in the Eastern Coast of West Australia, maybe
-due to advection of moist air from the Indian Ocean.In South America,
-precipitation anomalies associated with the Full SAM are negative both
-in Southern Chile and Southeastern South America, and positive in
-Southern Brazil. This features are cleanly separated between the
-Asymmetric and Symmetric components. The Symmetric SAM explains the
-negative anomalies in Southern Chile and the Asymmetric SAM, the
-negative-positive dipole between Southeastern South America and Southern
-Brazil. Individual seasons mostly follow this pattern.
+In Australia the Full SAM is associated with positive precipitation anomalies in South East and this is explained by the Symmetric SAM.
+However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
+In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
+This features are cleanly separated between the Asymmetric and Symmetric components.
+The Symmetric SAM explains the negative anomalies in Southern Chile and the Asymmetric SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
+Individual seasons mostly follow this pattern.
 
-\citet{silvestri2009} suggests that precipitation impacts linked to the
-SAM changed rather dramatically before and after 1980. In particular,
-the negative relationship with precipitation in South America was absent
-in some areas and switched sign in others in the earlier period. The
-correlation between ENSO and SAM is similarly non-stationary, also
-changing sign before the 1980s \citep{fogt2006, clem2013}. Seeing as
-both the ENSO-SAM relationship and most of the precipitation impacts in
-South America are captured by the Asymmetric SAM, the results presented
-here are most likely period-dependent.
+\citet{silvestri2009} suggests that precipitation impacts linked to the SAM changed rather dramatically before and after 1980.
+In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in others in the earlier period.
+The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s \citep{fogt2006, clem2013}.
+Seeing as both the ENSO-SAM relationship and most of the precipitation impacts in South America are captured by the Asymmetric SAM, the results presented here are most likely period-dependent.
 
-By succesfully separating the zonally symmetric and zonally symmetric
-SAM signals, we show that the asymmetric component of the SAM has its
-unique variability, trends and impacts. This is particularly important
-in the context of a changing climate, as the impact on the SAM of ozone
-recovery is modeled as highly zonally symmetric, while the impact of
-increased concentration of greenhose gases has also a zonally asymmetric
-component \citep{arblaster2006, simpkins2012}.
+By succesfully separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
+This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component \citep{arblaster2006, simpkins2012}.
 
@@
 
-NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the
-NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at
-https://psl.noaa.gov/
+NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/
 
-The research was supported by UBACyT20020170100428BA and the CLIMAX
-Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was
-supported by a PhD grant from CONICET, Argentina.
+The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
 
@@
 
-All data used in this paper is freely available in their respective
-sources. ERA5 data can be obtained via the Copernicus Climate Data Store
-(https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means).
-NOAAGlobalTemp and GPCC precipitation data can be obtained through the
-NOAA Physical Sicences Laboratory website
-(https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and
-https://psl.noaa.gov/data/gridded/data.gpcc.html). The Oceanic Niño
-Index is avaiable via NOAA's Climate Precidction Center:
-https://www.cpc.ncep.noaa.gov/products/analysis\_monitoring/ensostuff/detrend.nino34.ascii.txt
+All data used in this paper is freely available in their respective sources.
+ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means).
+NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html).
+The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: https://www.cpc.ncep.noaa.gov/products/analysis\_monitoring/ensostuff/detrend.nino34.ascii.txt
 
-A version-controlled repository of the code used to create this
-analysis, including the code used to download the data can be found at
-https://github.com/eliocamp/asymsam.
+A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at https://github.com/eliocamp/asymsam.
 
@@
 \includegraphics{A1-1} \appendcaption{A1}{Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
-Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.}\label{fig:A1}
+Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.
+For the 1979 -- 2018 period.}\label{fig:A1}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A3-1} \appendcaption{A3}{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
+\includegraphics{A2-1} \appendcaption{A2}{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
 The shading indicates de 95\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\% of the simulated sampels had an amplitude equal or lower).
-The light line indicates the theoretical expected amplitude from the autorregressive model.}\label{fig:A3}
+The light line indicates the theoretical expected amplitude from the autorregressive model.
+For the 1979 -- 2018 period.}\label{fig:A2}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A9-1} \caption[700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold]{700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold.
-Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A9}
+\includegraphics{A3-1} \caption[50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period]{50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A3}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A10-1} \caption[Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently]{Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently.}\label{fig:A10}
+\includegraphics{A4-1} \caption[700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period]{700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A4}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A11-1} \caption[Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 to 1998 and 1999 to 2018]{Regression of 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 to 1998 and 1999 to 2018.
-Pattern correlation between both fields is 0.76.}\label{fig:A11}
+\includegraphics{A5-1} \caption[Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period]{Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
+\end{figure}
+
+\begin{figure}
+\includegraphics{A6-1} \caption[Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018]{Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
+Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.}\label{fig:A6}
 \end{figure}

### 2020-10-10 analysis/paper/paper.Rmd
Corrige todo

@@
 output: 
-  bookdown::word_document2: default
   bookdown::pdf_book: 
@@
     latex_engine: "xelatex"
+  bookdown::word_document2: default
 header-includes: 
@@
 Even though these indices are based on zonal averages, spatial anomalies of geopotential associated with them (either by regression or composition) contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
-These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM [@fan2007, @fogt2012, @rosso2018], going as far as reversing its relationship between precipitation in South America [@silvestri2009].
+These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM [@fan2007;@fogt2012;@rosso2018], going as far as reversing its relationship between precipitation in South America [@silvestri2009].
 At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].
@@
 At least some of the variability associated with the zonal asymmetries of the SAM is probably forced by the tropics.
-In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987; @kidson1988;@karoly1989], whose wave train projects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
-And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed [@fan2007; @fogt2011; @clem2013].
+In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987;@kidson1988;@karoly1989], whose wave train projects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
+And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed [@fan2007;@fogt2011; @clem2013].
 In particular, @fan2007 computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated if the (linear) signal of the ENSO was removed.
@@
 We assess their vertical structure and coherence, temporal variability and trends.
-We then study the spatial patterns described by the variability exclusive to each index.
+We then study the spatial patterns described by the variability exclusive to each index focusing on 50 hPa as representing the stratosphere and 700 hPa as representing the troposphere.
 Finally, we investigate the relationship of the SAM at 700 hPa with temperature and precipitation anomalies.
-
-#LEAN: Te había hecho el comentario antes pero quizás no había sido del todo claro.
-Me parece que acá hay que distinguir más en cómo se trabaja metodológicamente cada parte con respecto a los niveles atmosféricos.
-Por ej.
-para la evaluación de los impactos solamente se está considerando el ínidice troposférico.
 

### 2020-10-10 analysis/paper/paper.Rmd
render

@@
  - \usepackage[inline]{showlabels}
+ - \usepackage[hidelinks]{hyperref}
 ---

### 2020-10-10 analysis/paper/paper.tex
render

@@
 \usepackage[inline]{showlabels}
+\usepackage[hidelinks]{hyperref}
 
@@
 
-Even though these indices are based on zonal averages, spatial anomalies of geopotential associated with these indices (either by regressino or composition) contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
-These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM \citep[\citet{fogt2012}, \citet{rosso2018}]{fan2007}, going as far as reversing its relationship between precipitation in South America \citep{silvestri2009}.
+Even though these indices are based on zonal averages, spatial anomalies of geopotential associated with them (either by regression or composition) contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
+These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM \citep{fan2007, fogt2012, rosso2018}, going as far as reversing its relationship between precipitation in South America \citep{silvestri2009}.
 At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere \citep{jones2009}.
@@
 We assess their vertical structure and coherence, temporal variability and trends.
-We then study the spatial patterns described by the variability exclusive to each index.
+We then study the spatial patterns described by the variability exclusive to each index focusing on 50 hPa as representing the stratosphere and 700 hPa as representing the troposphere.
 Finally, we investigate the relationship of the SAM at 700 hPa with temperature and precipitation anomalies.
 
-\#LEAN: Te había hecho el comentario antes pero quizás no había sido del todo claro.
-Me parece que acá hay que distinguir más en cómo se trabaja metodológicamente cada parte con respecto a los niveles atmosféricos.
-Por ej.
-para la evaluación de los impactos solamente se está considerando el ínidice troposférico.
+In the Section \ref{methods} we describe the methods.
+In Section \ref{temporal} we describe the temporal variability and vertical coherence of the indices.
+In Section \ref{spatial}, we analyse the spatial patterns of geopotential height associated with them.
+In Section \ref{impacts}, we study their relationship with surface-level temperature and precipitation.
 
-In the Section \ref{sec:methods} we describe the methods.
-In Section \ref{sec:results}.\ref{sec:temporal} we describe the temporal variability and vertical coherence of the indices.
-In Section \ref{sec:results}.\ref{sec:spatial}, we analyse the spatial patterns of geopotential height associated with them.
-In Section \ref{sec:results}.\ref{sec:impacts}, we study their relationship with surface-level temperature and precipitation.
-
-\section{Methods}
-
-\label{sec:methods}
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
@@
 
-\section{Results}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-\label{sec:results}
-
-\subsection{Temporal evolution}
-
-\label{sec:temporal}
+\hypertarget{temporal}{%
+\subsection{Temporal evolution}\label{temporal}}
 
@@
 
-\subsection{Spatial patterns}
-
-\label{sec:spatial}
+\hypertarget{spatial}{%
+\subsection{Spatial patterns}\label{spatial}}
 
@@
 
-\subsection{Surface Impacts}
-
-\label{sec:impacts}
+\hypertarget{impacts}{%
+\subsection{Surface Impacts}\label{impacts}}
 

### 2020-11-25 analysis/paper2/paper2.Rmd
Adds climate dynamics paper

--- 2020-10-10 analysis/paper/paper.Rmd
@@
 ---
-twocol: FALSE  #  Use FALSE for submission. 
 title: Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
-author1: 
-  name: Elio Campitelli
-  email: elio.campitelli@cima.fcen.uba.ar
-author2: 
-  name: Leandro B. Díaz
-affiliation: |
+# subtitle: Do you have a subtitle? If so, write it here
+titlerunning:  Assessment of zonal symmetric and asymmetric components of the SAM
+# authorrunning: Short form of author list if too long for running head
+thanks: | 
+    The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+
+authors: 
+- name: Elio Campitelli
+  address: |
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
@@
       CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
-extra_authors: 
-  - name: Carolina Vera
-    affiliation: |
+  email: elio.campitelli@cima.fcen.uba.ar
+
+- name:  Leandro B. Díaz
+  address: |
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
@@
       CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+
+- name:  Carolina Vera
+  address: |
+    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+      CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+      CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+
+keywords:
+  - Southern Annular Mode
+  - general circulation
+  - zonally asymmetric circulation
+  - El Niño Southern Oscillation
+
+#PACS: 
+#- PAC1
+#- superPAC
+
+MSC:
+- MSC code 1
+- MSC code 2
+
 abstract: |
-  Enter the text of your abstract here.
-This is a sample American Meteorological Society (AMS) \LaTeX\ template.
-This document provides authors with instructions on the use of the AMS \LaTeX\ template.
-Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document.
-The template.tex file should be modified by authors for their own manuscript.
-statement: |
-  This is significant becasue I wrote it.
-bibliography: ["AsymSAM.bib","packages.bib"]
+  The text of your abstract.  150 -- 250 words.
+
+bibliography: 
+  - AsymSAM.bib
+  - packages.bib
+bibstyle: spphys
+# bibstyle options spbasic(default), spphys, spmpsci
 output: 
   bookdown::pdf_book: 
-    base_format: rticles::ams_article
+      base_format: rticles::springer_article
       latex_engine: "xelatex"
-  bookdown::word_document2: default
+      # citation_package: "natbib"
 header-includes: 
@@
  - \usepackage[inline]{showlabels}
+ - \usepackage{chngcntr}
+ - \usepackage{natbib}
 ---
@@
   comment = "#>",
-  dpi = 300,
+  dev = "pdf",
+  # dpi = 300,
   out.extra = ""
@@
 @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
-Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures A3 and A4).
+Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures \@ref(fig:A3) and \@ref(fig:A4)).
 Therefore, we believe that our method is a reasonable approximation of the phenomenon.
@@
 By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
-Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure A5, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
-In the stratosphere (Figure A5, row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree\ in comparison with the rest of the year.
+Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
+In the stratosphere (Figure \@ref(fig:A5), row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree\ in comparison with the rest of the year.
 Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
@@
 @silvestri2009 suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes.
-Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure A6).
+Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure \@ref(fig:A6)).
 
@@
 
-The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A2).
+The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \@ref(fig:A2)).
 This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b].
@@
 
-\acknowledgments
+\begin{acknowledgements}
 NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ 
-
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
-
-\datastatement
+\end{acknowledgements}
+
+### atastatement
 All data used in this paper is freely available in their respective sources.
@@
 
-\bibliography{AsymSAM,packages}
+<!-- \bibliography{AsymSAM,packages} -->
 
@@
 
-\appendixtitle{Extra figures}
+\counterwithin{figure}{section}
+
+# Extra figures
+
+\newpage
 
@@
 Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.
-For the 1979 -- 2018 period.", fig.width = 4, fig.height=4, appendix = TRUE}
+For the 1979 -- 2018 period.", fig.width = 4, fig.height=4, fig.pos="ht"}
 ccf_dt <- function(..., plot = FALSE) {
@@
 The light line indicates the theoretical expected amplitude from the autorregressive model.
-For the 1979 -- 2018 period.", appendix = TRUE, fig.height=4, fig.width=7}
+For the 1979 -- 2018 period.", fig.height=4, fig.width=7}
 spec %>%
@@
 ```{r A6, fig.cap = "Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
-Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.", fig.width=6, fig.height=5, apendix = TRUE}
+Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.", fig.width=6, fig.height=5}
 set.seed(42)

### 2020-11-25 analysis/paper2/paper2.tex
Adds climate dynamics paper

--- 2020-10-10 analysis/paper/paper.tex
@@
-%% Version 5.0, 2 January 2020
+% !TeX program = pdfLaTeX
+\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+%\documentclass[twocolumn]{svjour3}          % twocolumn
 %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% TemplateV5.tex --  LaTeX-based template for submissions to the
-% American Meteorological Society
+\smartqed  % flush right qed marks, e.g. at end of proof
 %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% PREAMBLE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage[utf8]{inputenc}
 
-%% Start with one of the following:
-% DOUBLE-SPACED VERSION FOR SUBMISSION TO THE AMS
-\documentclass[]{ametsocV5}
+\usepackage[hyphens]{url} % not crucial - just used below for the URL
+\usepackage{hyperref}
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
-% TWO-COLUMN JOURNAL PAGE LAYOUT---FOR AUTHOR USE ONLY
-% \documentclass[twocol]{ametsocV5}
+%
+% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+%
+% insert here the call for the packages your document requires
+%\usepackage{latexsym}
+% etc.
+%
+% please place your own definitions here and don't use \def but
+% \newcommand{}{}
+%
+% Insert the name of "your journal" with
+% \journalname{myjournal}
+%
 
-% Enter packages here. If too many math alphabets are used,
-% remove unnecessary packages or define hmmax and bmmax as necessary.
+%% load any required packages here
 
-\newcommand{\hmmax}{0}
-\newcommand{\bmmax}{0}
-\usepackage{amsmath,amsfonts,amssymb,bm}
-\usepackage{mathptmx}%{times}
-\usepackage{newtxtext}
-\usepackage{newtxmath}
+% Pandoc citation processing
 
@@
 \usepackage[inline]{showlabels}
-\usepackage[hidelinks]{hyperref}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%% To be entered by author:
-
-%% May use \\ to break lines in title:
-
-\title{Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach}
-
-\authors{Elio Campitelli
-\correspondingauthor{Elio Campitelli,elio.campitelli@cima.fcen.uba.ar}
-and Leandro B. Díaz
-}
-
-\affiliation{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina}
-
-\extraauthor{Carolina Vera
-}
-\extraaffil{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% ABSTRACT
-%
-% Enter your abstract here
-% Abstracts should not exceed 250 words in length!
-%
-
-\abstract{Enter the text of your abstract here.
-This is a sample American Meteorological Society (AMS) \LaTeX~template.
-This document provides authors with instructions on the use of the AMS \LaTeX~template.
-Authors should refer to the file amspaper.tex to review the actual \LaTeX~code used to create this document.
-The template.tex file should be modified by authors for their own manuscript.}
+\usepackage{chngcntr}
+\usepackage{natbib}
 
@@
 
-%% Necessary!
+\title{Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
+Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }
+
+    \titlerunning{Assessment of zonal symmetric and asymmetric components of the SAM}
+
+\author{  Elio Campitelli \and  Leandro B. Díaz \and  Carolina Vera \and  }
+
+\institute{
+        Elio Campitelli \at
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+ CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+ CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     \email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}  %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+    \and
+        Leandro B. Díaz \at
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+ CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+ CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+    \and
+        Carolina Vera \at
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+ CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+ CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+    \and
+    }
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
 \maketitle
 
-\bibliographystyle{ametsoc2014}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% SIGNIFICANCE STATEMENT/CAPSULE SUMMARY
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% If you are including an optional significance statement for a journal article or a required capsule summary for BAMS
-% (see www.ametsoc.org/ams/index.cfm/publications/authors/journal-and-bams-authors/formatting-and-manuscript-components for details),
-% please apply the necessary command as shown below:
-%
-\statement
-This is significant becasue I wrote it.
+\begin{abstract}
+The text of your abstract. 150 -- 250 words.
+\\
+\keywords{
+        Southern Annular Mode \and
+        general circulation \and
+        zonally asymmetric circulation \and
+        El Niño Southern Oscillation \and
+    }
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MAIN BODY OF PAPER
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
+    \subclass{
+                    MSC code 1 \and
+                    MSC code 2 \and
+            }
 
-\section{Introduction}
+\end{abstract}
 
-The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation \citep{rogers1982} in daily, monthly, and decadal timescales \citep{baldwin2001a, fogt2006} and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration \citep[e.g.][]{fogt2020}.
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation (Rogers and van Loon 1982) in daily, monthly, and decadal timescales (Mark P.
+Baldwin 2001; Fogt and Bromwich 2006) and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration (e.g.
+Fogt and Marshall 2020).
 Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
@@
 Most authors describe the SAM as a zonally symmetric pattern, a fact that is reflected not only in its name, but also in the various methods used to characterise it.
-Of the several different indices presented in the literature, many of them are based on zonal means of sea level pressure or geopotential height \citep{ho2012}.
-\citet{gong1999} defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 60\degree S, which is also the definition used by the station-based index in \citet{marshall2003}.
-\citet{baldwin2009} proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
+Of the several different indices presented in the literature, many of them are based on zonal means of sea level pressure or geopotential height (Ho, Kiem, and Verdon-Kidd 2012).
+Gong and Wang (1999) defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 60\degree S, which is also the definition used by the station-based index in Marshall (2003).
+Baldwin and Thompson (2009) proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
 
 Even though these indices are based on zonal averages, spatial anomalies of geopotential associated with them (either by regression or composition) contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
-These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM \citep{fan2007, fogt2012, rosso2018}, going as far as reversing its relationship between precipitation in South America \citep{silvestri2009}.
-At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere \citep{jones2009}.
+These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM (Fan 2007; Fogt, Jones, and Renwick 2012; Rosso et al.
+2018), going as far as reversing its relationship between precipitation in South America (Silvestri and Vera 2009).
+At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere (Jones et al.
+2009).
 
 At least some of the variability associated with the zonal asymmetries of the SAM is probably forced by the tropics.
-In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989}, whose wave train projects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
-And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed \citep{fan2007, fogt2011, clem2013}.
-In particular, \citet{fan2007} computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated if the (linear) signal of the ENSO was removed.
+In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern (Mo and Ghil 1987; Kidson 1988; Karoly 1989), whose wave train projects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
+And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed (Fan 2007; Fogt, Bromwich, and Hines 2011; Clem and Fogt 2013).
+In particular, Fan (2007) computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated if the (linear) signal of the ENSO was removed.
 While this relates to temporally coherent variability of the two hemisphere and not necessarily to zonal asymmetries in the associated spatial patterns, it is nonetheless consistent of the ENSO playing a crucial role in zonal asymmetries of the SAM.
 
-Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn \citep[e.g.][ and references therein]{fogt2020}.
-These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables \citep{marshall2004, gillett2005, arblaster2006, gillett2013}.
+Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn (e.g.
+Fogt and Marshall 2020 and references therein).
+These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables (Marshall et al.
+2004; Gillett, Allan, and Ansell 2005; Arblaster and Meehl 2006; Gillett, Fyfe, and Parker 2013).
 However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variability could be masking influencing the observed trends independently.
@@
 Similarly unclear are the specific impacts of the zonally asymmetric component of the SAM.
-Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes \citep{jones2019} (and vice versa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}.
-The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry \citep{lim2016}.
+Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes (Jones et al.
+2019) (and vice versa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic (Fogt, Jones, and Renwick 2012).
+The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry (Lim et al.
+2016).
 The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
-For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}.
-\citet{fan2007} also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
+For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymmetric circulation associated with the SAM (Silvestri and Vera 2009; Rosso et al.
+2018).
+Fan (2007) also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
 
-We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of \citet{fogt2012}.
+We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of Fogt, Jones, and Renwick (2012).
 However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988).
-This is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM \citep{silvestri2009}.
-Moreover, \citet{fogt2012} studied the zonal asymmetric component of the SAM only in sea level pressure.
-Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere \citep{baldwin2009}.
+This is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM (Silvestri and Vera 2009).
+Moreover, Fogt, Jones, and Renwick (2012) studied the zonal asymmetric component of the SAM only in sea level pressure.
+Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere (Baldwin and Thompson 2009).
 
@@
 
-\subsubsection{Data}
+\hypertarget{data}{%
+\subsubsection{Data}\label{data}}
 
-To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{hersbach2020} for the period 1979 to 2018.
+To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 (Hersbach et al.
+2020) for the period 1979 to 2018.
 We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.
 
-To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land and ocean temperature analysis into a monthly global grid 5\degree~longitude by 5\degree~latitude, and monthly rainfall 0.5\degree~longitude by 0.5\degree~latitude data from the Global Precipitation Climatology Centre \citep{schneider2015, schneider2017}.
+To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 (Smith et al.
+2008; Vose et al.
+2012), which blends land and ocean temperature analysis into a monthly global grid 5\degree~longitude by 5\degree~latitude, and monthly rainfall 0.5\degree~longitude by 0.5\degree~latitude data from the Global Precipitation Climatology Centre (Schneider et al.
+2015, 2017).
 The rainfall product uses station-based records, and thus it only has continental coverage.
 
-\subsubsection{Definition of indices}
+\hypertarget{definition-of-indices}{%
+\subsubsection{Definition of indices}\label{definition-of-indices}}
 
-Traditionally the Souther Annular Mode (SAM) is defined as de leading Empirical Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels \citep{ho2012}.
-Following \citet{baldwin2001}, we extend that definition vertically and use the term SAM to refer to the the leading EOF of the monthly anomalies of geopotential field south of 20\degree S at each level.
+Traditionally the Souther Annular Mode (SAM) is defined as de leading Empirical Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels (Ho, Kiem, and Verdon-Kidd 2012).
+Following Mark P.
+Baldwin and Dunkerton (2001), we extend that definition vertically and use the term SAM to refer to the the leading EOF of the monthly anomalies of geopotential field south of 20\degree S at each level.
 We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in 481 rows and 4176 columns (144 points of longitude and 29 points of latitude).
-We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint \citep{chung1999}.
+We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint (Chung and Nigam 1999).
 
@@
 \begin{figure*}
-\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period]{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
+\includegraphics{figures/method-1} \caption{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 That means we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM.
-\citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
-Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures A3 and A4).
+Fogt, Jones, and Renwick (2012)'s composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
+Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures \ref{fig:A3} and \ref{fig:A4}).
 Therefore, we believe that our method is a reasonable approximation of the phenomenon.
@@
 By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
-Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure A5, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
-In the stratosphere (Figure A5, row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree~in comparison with the rest of the year.
+Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure \ref{fig:A5}, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
+In the stratosphere (Figure \ref{fig:A5}, row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree~in comparison with the rest of the year.
 Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
@@
 Finally, we assume that the zonally asymmetric pattern is stationary in time.
-\citet{silvestri2009} suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes.
-Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure A6).
+Silvestri and Vera (2009) suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes.
+Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure \ref{fig:A6}).
 
-\subsubsection{Regressions}
+\hypertarget{regressions}{%
+\subsubsection{Regressions}\label{regressions}}
 
@@
 At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions.
-In such case, using p-values to test for significance leads to misleading results \citep{walker1914, katz1991}.
-While there are multiple proposed solutions in the literature, \citet{wilks2016} suggests that adjusting p-values by controlling for the False Discovery Rate \citep{benjamini1995} is a simple and effective method to ameliorate this issue.
-Therefore, p-values showed in regression fields are all adjusted following \citet{benjamini1995}.
+In such case, using p-values to test for significance leads to misleading results (Walker 1914; Katz and Brown 1991).
+While there are multiple proposed solutions in the literature, Wilks (2016) suggests that adjusting p-values by controlling for the False Discovery Rate (Benjamini and Hochberg 1995) is a simple and effective method to ameliorate this issue.
+Therefore, p-values showed in regression fields are all adjusted following Benjamini and Hochberg (1995).
 
@@
 
-We computed density probability estimates using a gaussian kernel of optimal bindwidth according to \citet{sheather1991}.
+We computed density probability estimates using a gaussian kernel of optimal bindwidth according to Sheather and Jones (1991).
 
-\subsubsection{Computation procedures}
+\hypertarget{computation-procedures}{%
+\subsubsection{Computation procedures}\label{computation-procedures}}
 
-We performed all analysis in this paper using the R programming language \citep{rcoreteam2020}, using the data.table package \citep{dowle2020} and the metR package \citep{campitelli2020}.
-All graphics are made using ggplot2 \citep{wickham2009}.
-We downloaded data from ranalysis using the ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi package \citep{albers2020}.
-The paper was rendered using knitr and rmarkdown \citep{xie2015, allaire2019}.
+We performed all analysis in this paper using the R programming language (R Core Team 2020), using the data.table package (Dowle et al.
+2020) and the metR package (Campitelli 2020).
+All graphics are made using ggplot2 (Wickham 2009).
+We downloaded data from ranalysis using the ecmwfr package (Hufkens, Stauffer, and Campitelli 2020) and indices of the ENSO with the rsoi package (Albers and Campitelli 2020).
+The paper was rendered using knitr and rmarkdown (Xie 2015; Allaire et al.
+2019).
 
@@
 \begin{figure*}
-\includegraphics{asymsam-timeseries-1} \caption[Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa]{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
+\includegraphics{figures/asymsam-timeseries-1} \caption{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
 To the right, probability density estimate of each index.
@@
 
-The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A2).
-This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation \citep{baldwin2001b}.
+The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \ref{fig:A2}).
+This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation (M.
+P.
+Baldwin et al.
+2001).
 There is a local peak at 2 years in the periodogram of the tropospheric Symmetric SAM also, although it's not statistically significant.
@@
 \begin{figure}
-\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.}\label{fig:cor-lev}
+\includegraphics{figures/cor-lev-1} \caption{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.}\label{fig:cor-lev}
 \end{figure}
@@
 \begin{figure*}
-\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.}\label{fig:cross-correlation}
+\includegraphics{figures/cross-correlation-1} \caption{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.}\label{fig:cross-correlation}
 \end{figure*}
@@
 \begin{figure*}
-\includegraphics{trends-1} \caption[Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+\includegraphics{figures/trends-1} \caption{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
 Shading indicates the 95\% confidence interval from a t-distribution.}\label{fig:trends}
@@
 In Winter and Spring, we detect no statistically significant trend.
-This is consistent with the results of previous studies, which find clear positive trends in Summer, weaker in Autumn and no trends in the other seasons \citep[e.g.][ and references therein]{fogt2020} using indices of the SAM based on surface or near-surface circulation.
+This is consistent with the results of previous studies, which find clear positive trends in Summer, weaker in Autumn and no trends in the other seasons (e.g.
+Fogt and Marshall 2020 and references therein) using indices of the SAM based on surface or near-surface circulation.
 
@@
 \begin{figure}
-\includegraphics{r-squared-trend-1} \caption[Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018]{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018.
+\includegraphics{figures/r-squared-trend-1} \caption{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018.
 Shading indicates the 95\% confidence interval.}\label{fig:r-squared-trend}
@@
 \begin{figure*}
-\includegraphics{2d-regr-1} \caption[Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period]{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{figures/2d-regr-1} \caption{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period.
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices.
@@
 
-In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 \citep{fogt2012}.
+In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 (Fogt, Jones, and Renwick 2012).
 The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
 The Asymmetric SAM index gives rise to a cleaner zonal wave (panel b.2) and the Symmetric SAM index is associated with an annular mode, almost devoid of zonal asymmetries (panel c.2).
-The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern associated with \citet{raphael2004}'s ZW3 index, whose reference locations are marked with points in the figure.
+The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern associated with Raphael (2004)'s ZW3 index, whose reference locations are marked with points in the figure.
 Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
@@
 \begin{figure*}
-\includegraphics{wave-amplitude-1} \caption[Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean]{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
+\includegraphics{figures/wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
 \end{figure*}
@@
 \begin{figure}
-\includegraphics{vertical-regression-1} \caption[Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM)]{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
+\includegraphics{figures/vertical-regression-1} \caption{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
 (a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979 -- 2018 period.}\label{fig:vertical-regression}
@@
 
-The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}.
-Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is -0.16 (p-value = \ensuremath{2.8\times 10^{-4}}).
-Consistent with \citet{fan2007}, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of -0.26 (p-value = \ensuremath{6.3\times 10^{-9}}) with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67).
-We performed the same analysis using the Multivariate ENSO Index \citep{wolter2011} and the Southern Oscillation Index \citep{ropelewski1987} to conclude that these results do not depend on the ENSO index used.
+The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern (Mo and Ghil 1987; Kidson 1988) which is a teleconnection pattern associated with the ENSO (Karoly 1989).
+Indeed, Fogt, Bromwich, and Hines (2011) showed that there is a significant relationship between the SAM and the ENSO.
+The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index (ONI, Bamston, Chelliah, and Goldenberg 1997) is -0.16 (p-value = \ensuremath{2.8\times 10^{-4}}).
+Consistent with Fan (2007), we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of -0.26 (p-value = \ensuremath{6.3\times 10^{-9}}) with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67).
+We performed the same analysis using the Multivariate ENSO Index (Wolter and Timlin 2011) and the Southern Oscillation Index (Ropelewski and Jones 1987) to conclude that these results do not depend on the ENSO index used.
 
@@
 \begin{figure*}
-\includegraphics{regr-air-season-1} \caption[Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period]{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{figures/regr-air-season-1} \caption{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
-These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following \citet{wilks2016}.
+These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following Wilks (2016).
 However, repeating this analysis with 2-meter temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
-Moreover, similar features were observed in station measurements by \citet{jones2019}, although using data from 1957 to 2016.
+Moreover, similar features were observed in station measurements by Jones et al.
+(2019), although using data from 1957 to 2016.
 
@@
 
-Figure \ref{fig:regr-air-season} column b can be partially compared with Figure 11 from \citet{fogt2012}.
+Figure \ref{fig:regr-air-season} column b can be partially compared with Figure 11 from Fogt, Jones, and Renwick (2012).
 Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.
@@
 \begin{figure*}
-\includegraphics{pp-regr-oceania-1} \caption[Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period]{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{figures/pp-regr-oceania-1} \caption{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
@@
 
-In Australia, the annual regression shows that the Full SAM index is positively associated with precipitation in the Southeastern region (Figure \ref{fig:pp-regr-oceania} panel a.1), which reproduces the results from \citet{gillett2006}.
+In Australia, the annual regression shows that the Full SAM index is positively associated with precipitation in the Southeastern region (Figure \ref{fig:pp-regr-oceania} panel a.1), which reproduces the results from Gillett, Kell, and Jones (2006).
 The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (panel c.1).
-Geopotential anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by \citet{hendon2007}.
+Geopotential anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by Hendon, Thompson, and Wheeler (2007).
 The Asymmetric SAM appears related to increased precipitation in the West coast of Southeastern Australia (panel b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
@@
 None of these regression coefficients are statistically significant at the 95\% level.
-The Spring signal is broadly consistent with \citet{hendon2007}, but whereas \citet{hendon2007} also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coefficients have the consistent sign).
+The Spring signal is broadly consistent with Hendon, Thompson, and Wheeler (2007), but whereas Hendon, Thompson, and Wheeler (2007) also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coefficients have the consistent sign).
 
 \begin{figure*}
-\includegraphics{pp-regr-america-1} \caption[Same as Figure \\ref{fig:pp-regr-oceania} but for South America]{Same as Figure \\ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
+\includegraphics{figures/pp-regr-america-1} \caption{Same as Figure \\ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
 \end{figure*}
@@
 The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous Easterly flow over Southern Chile.
-This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region \citep{garreaud2007}.
+This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region (Garreaud 2007).
 On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
-This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region \citep{silvestri2009, zamboni2010}.
-This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events \citep{rosso2018}.
-There is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} and that is explained by the Asymmetric SAM.
+This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region (Silvestri and Vera 2009; Zamboni, Mechoso, and Kucharski 2010).
+This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events (Rosso et al.
+2018).
+There is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by Gillett, Kell, and Jones (2006) and that is explained by the Asymmetric SAM.
 
-\subsection{Conclusions}
+\hypertarget{conclusions}{%
+\subsection{Conclusions}\label{conclusions}}
 
@@
 The two-year periodicity we found in the stratospheric Symmetric SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
-There is evidence of influence between the QBO and the Northern Annular Mode \citep[e.g.][]{holton1980, watson2014, zhang2020}, so it's not unlikely that the SAM would be similarly connected.
+There is evidence of influence between the QBO and the Northern Annular Mode (e.g.
+Holton and Tan 1980; Watson and Gray 2014; Zhang, Tian, and Wang 2020), so it's not unlikely that the SAM would be similarly connected.
 However establishing this link would require further research.
 
-We observe a positive trend towards positive SAM in Summer and Autumn, As was documented by previous studies, such as \citet{fogt2020} (and references therein) for surface levels.
+We observe a positive trend towards positive SAM in Summer and Autumn, As was documented by previous studies, such as Fogt and Marshall (2020) (and references therein) for surface levels.
 We show that these trends are maximised at the 100 hPa level and are explained by the zonally symmetric component.
 We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not apparent in the Full SAM index.
-In contrast to \citet{fogt2012} we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
+In contrast to Fogt, Jones, and Renwick (2012) we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
 
 In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being truly annular than the patterns associated with the Full SAM index.
-The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific region and whose phase is rotated a quarter wavelength from the mean zonal wave 3 described by \citet{raphael2004}'s index.
+The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific region and whose phase is rotated a quarter wavelength from the mean zonal wave 3 described by Raphael (2004)'s index.
 This pattern extends in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimum for capturing this variability.
@@
 In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship.
-These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active \citep{cazes-boezio2003, fogt2011, cai2020a}.
+These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active (Cazes-Boezio, Robertson, and Mechoso 2003; Fogt, Bromwich, and Hines 2011; Cai et al.
+2020).
 
@@
 
-\citet{silvestri2009} suggests that precipitation impacts linked to the SAM changed rather dramatically before and after 1980.
+Silvestri and Vera (2009) suggests that precipitation impacts linked to the SAM changed rather dramatically before and after 1980.
 In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in others in the earlier period.
-The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s \citep{fogt2006, clem2013}.
+The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s (Fogt and Bromwich 2006; Clem and Fogt 2013).
 Seeing as both the ENSO-SAM relationship and most of the precipitation impacts in South America are captured by the Asymmetric SAM, the results presented here are most likely period-dependent.
@@
 By succesfully separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
-This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component \citep{arblaster2006, simpkins2012}.
+This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component (Arblaster and Meehl 2006; Simpkins and Karpechko 2012).
 
-\acknowledgments
+\begin{acknowledgements}
+NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ 
+\end{acknowledgements}
 
-NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/
-
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
-
-\datastatement
+\hypertarget{atastatement}{%
+\subsubsection{atastatement}\label{atastatement}}
 
 All data used in this paper is freely available in their respective sources.
-ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means).
-NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html).
-The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: https://www.cpc.ncep.noaa.gov/products/analysis\_monitoring/ensostuff/detrend.nino34.ascii.txt
+ERA5 data can be obtained via the Copernicus Climate Data Store (\url{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means}).
+NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (\url{https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html} and \url{https://psl.noaa.gov/data/gridded/data.gpcc.html}).
+The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}
 
-A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at https://github.com/eliocamp/asymsam.
-
-\bibliography{AsymSAM,packages}
+A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at \url{https://github.com/eliocamp/asymsam}.
 
@@
 
-\appendixtitle{Extra figures}
+\counterwithin{figure}{section}
 
-\begin{figure}
-\includegraphics{A1-1} \appendcaption{A1}{Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
+\hypertarget{extra-figures}{%
+\section{Extra figures}\label{extra-figures}}
+
+\newpage
+
+\begin{figure}[ht]
+\includegraphics{figures/A1-1} \caption{Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
 Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.
@@
 \begin{figure}
-\includegraphics{A2-1} \appendcaption{A2}{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
+\includegraphics{figures/A2-1} \caption{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
 The shading indicates de 95\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\% of the simulated sampels had an amplitude equal or lower).
@@
 \begin{figure}
-\includegraphics{A3-1} \caption[50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period]{50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+\includegraphics{figures/A3-1} \caption{50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
 Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A3}
@@
 \begin{figure}
-\includegraphics{A4-1} \caption[700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period]{700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+\includegraphics{figures/A4-1} \caption{700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
 Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A4}
@@
 \begin{figure}
-\includegraphics{A5-1} \caption[Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period]{Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
+\includegraphics{figures/A5-1} \caption{Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A6-1} \caption[Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018]{Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
+\includegraphics{figures/A6-1} \caption{Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
 Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.}\label{fig:A6}
@@
 
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-albers2020}{}%
+Albers, Sam, and Elio Campitelli. 2020. ``Rsoi: Import Various Northern and Southern Hemisphere Climate Indices.''
+
+\leavevmode\hypertarget{ref-allaire2019}{}%
+Allaire, JJ, Jeffrey Horner, Yihui Xie, Vicent Marti, and Natacha Porte. 2019. \emph{Markdown: Render Markdown with the c Library 'Sundown'}. Manual.
+
+\leavevmode\hypertarget{ref-arblaster2006}{}%
+Arblaster, Julie M., and Gerald A.
+Meehl.
+2006.
+``Contributions of External Forcings to Southern Annular Mode Trends.'' \emph{J.
+Climate} 19 (12).
+American Meteorological Society: 2896--2905.
+\url{https://doi.org/10.1175/JCLI3774.1}.
+
+\leavevmode\hypertarget{ref-baldwin2001a}{}%
+Baldwin, Mark P. 2001. ``Annular Modes in Global Daily Surface Pressure.'' \emph{Geophysical Research Letters} 28 (21): 4115--8. \url{https://doi.org/10.1029/2001GL013564}.
+
+\leavevmode\hypertarget{ref-baldwin2001}{}%
+Baldwin, Mark P., and Timothy J.
+Dunkerton.
+2001.
+``Stratospheric Harbingers of Anomalous Weather Regimes.'' \emph{Science} 294 (5542).
+American Association for the Advancement of Science: 581--84.
+\url{https://doi.org/10.1126/science.1063315}.
+
+\leavevmode\hypertarget{ref-baldwin2009}{}%
+Baldwin, Mark P., and David W.
+J.
+Thompson.
+2009.
+``A Critical Comparison of StratosphereTroposphere Coupling Indices.'' \emph{Quarterly Journal of the Royal Meteorological Society} 135 (644): 1661--72.
+\url{https://doi.org/10.1002/qj.479}.
+
+\leavevmode\hypertarget{ref-baldwin2001b}{}%
+Baldwin, M.
+P., L.
+J.
+Gray, T.
+J.
+Dunkerton, K.
+Hamilton, P.
+H.
+Haynes, W.
+J.
+Randel, J.
+R.
+Holton, et al.
+2001.
+``The Quasi-Biennial Oscillation.'' \emph{Reviews of Geophysics} 39 (2): 179--229.
+\url{https://doi.org/10.1029/1999RG000073}.
+
+\leavevmode\hypertarget{ref-bamston1997}{}%
+Bamston, Anthony G., Muthuvel Chelliah, and Stanley B.
+Goldenberg.
+1997.
+``Documentation of a Highly ENSO-Related Sst Region in the Equatorial Pacific: Research Note.'' \emph{Atmosphere-Ocean} 35 (3).
+Taylor \& Francis: 367--83.
+\url{https://doi.org/10.1080/07055900.1997.9649597}.
+
+\leavevmode\hypertarget{ref-benjamini1995}{}%
+Benjamini, Yoav, and Yosef Hochberg.
+1995.
+``Controlling the False Discovery Rate: A Practical and Powerful Approach to Multiple Testing.'' \emph{Journal of the Royal Statistical Society: Series B (Methodological)} 57 (1): 289--300.
+\url{https://doi.org/10.1111/j.2517-6161.1995.tb02031.x}.
+
+\leavevmode\hypertarget{ref-cai2020a}{}%
+Cai, Wenju, Michael J.
+McPhaden, Alice M.
+Grimm, Regina R.
+Rodrigues, Andréa S.
+Taschetto, René D.
+Garreaud, Boris Dewitte, et al.
+2020.
+``Climate Impacts of the El NiñoSouthern Oscillation on South America.'' \emph{Nature Reviews Earth \& Environment} 1 (4).
+Nature Publishing Group: 215--31.
+\url{https://doi.org/10.1038/s43017-020-0040-3}.
+
+\leavevmode\hypertarget{ref-campitelli2020}{}%
+Campitelli, Elio. 2020. ``metR: Tools for Easier Analysis of Meteorological Fields.''
+
+\leavevmode\hypertarget{ref-cazes-boezio2003}{}%
+Cazes-Boezio, Gabriel, Andrew W.
+Robertson, and Carlos R.
+Mechoso.
+2003.
+``Seasonal Dependence of ENSO Teleconnections over South America and Relationships with Precipitation in Uruguay.'' \emph{J.
+Climate} 16 (8).
+American Meteorological Society: 1159--76.
+\href{https://doi.org/10.1175/1520-0442(2003)16\%3C1159:SDOETO\%3E2.0.CO;2}{https://doi.org/10.1175/1520-0442(2003)16\textless{}1159:SDOETO\textgreater{}2.0.CO;2}.
+
+\leavevmode\hypertarget{ref-chung1999}{}%
+Chung, Chul, and Sumant Nigam.
+1999.
+``Weighting of Geophysical Data in Principal Component Analysis.'' \emph{Journal of Geophysical Research: Atmospheres} 104 (D14): 16925--8.
+\url{https://doi.org/10.1029/1999JD900234}.
+
+\leavevmode\hypertarget{ref-clem2013}{}%
+Clem, Kyle R., and Ryan L.
+Fogt.
+2013.
+``Varying Roles of ENSO and SAM on the Antarctic Peninsula Climate in Austral Spring.'' \emph{Journal of Geophysical Research: Atmospheres} 118 (20): 11, 481--11, 492.
+\url{https://doi.org/10.1002/jgrd.50860}.
+
+\leavevmode\hypertarget{ref-dowle2020}{}%
+Dowle, Matt, Arun Srinivasan, Jan Gorecki, Michael Chirico, Pasha Stetsenko, Tom Short, Steve Lianoglou, et al. 2020. ``Data.Table: Extension of 'Data.frame'.''
+
+\leavevmode\hypertarget{ref-fan2007}{}%
+Fan, Ke. 2007. ``Zonal Asymmetry of the Antarctic Oscillation.'' \emph{Geophysical Research Letters} 34 (2). John Wiley \& Sons, Ltd. \url{https://doi.org/10.1029/2006GL028045}.
+
+\leavevmode\hypertarget{ref-fogt2006}{}%
+Fogt, Ryan L., and David H.
+Bromwich.
+2006.
+``Decadal Variability of the ENSO Teleconnection to the High-Latitude South Pacific Governed by Coupling with the Southern Annular Mode.'' \emph{J.
+Climate} 19 (6).
+American Meteorological Society: 979--97.
+\url{https://doi.org/10.1175/JCLI3671.1}.
+
+\leavevmode\hypertarget{ref-fogt2011}{}%
+Fogt, Ryan L., David H.
+Bromwich, and Keith M.
+Hines.
+2011.
+``Understanding the SAM Influence on the South Pacific ENSO Teleconnection.'' \emph{Clim Dyn} 36 (7): 1555--76.
+\url{https://doi.org/10.1007/s00382-010-0905-0}.
+
+\leavevmode\hypertarget{ref-fogt2012}{}%
+Fogt, Ryan L., Julie M.
+Jones, and James Renwick.
+2012.
+``Seasonal Zonal Asymmetries in the Southern Annular Mode and Their Impact on Regional Temperature Anomalies.'' \emph{J.
+Climate} 25 (18).
+American Meteorological Society: 6253--70.
+\url{https://doi.org/10.1175/JCLI-D-11-00474.1}.
+
+\leavevmode\hypertarget{ref-fogt2020}{}%
+Fogt, Ryan L., and Gareth J.
+Marshall.
+2020.
+``The Southern Annular Mode: Variability, Trends, and Climate Impacts Across the Southern Hemisphere.'' \emph{WIREs Climate Change} 11 (4): e652.
+\url{https://doi.org/10.1002/wcc.652}.
+
+\leavevmode\hypertarget{ref-garreaud2007}{}%
+Garreaud, RenéD.
+2007.
+``Precipitation and Circulation Covariability in the Extratropics.'' \emph{J.
+Climate} 20 (18).
+American Meteorological Society: 4789--97.
+\url{https://doi.org/10.1175/JCLI4257.1}.
+
+\leavevmode\hypertarget{ref-gillett2013}{}%
+Gillett, Nathan P., John C.
+Fyfe, and David E.
+Parker.
+2013.
+``Attribution of Observed Sea Level Pressure Trends to Greenhouse Gas, Aerosol, and Ozone Changes.'' \emph{Geophysical Research Letters} 40 (10): 2302--6.
+\url{https://doi.org/10.1002/grl.50500}.
+
+\leavevmode\hypertarget{ref-gillett2005}{}%
+Gillett, N.
+P., R.
+J.
+Allan, and T.
+J.
+Ansell.
+2005.
+``Detection of External Influence on Sea Level Pressure with a Multi-Model Ensemble.'' \emph{Geophysical Research Letters} 32 (19).
+\url{https://doi.org/10.1029/2005GL023640}.
+
+\leavevmode\hypertarget{ref-gillett2006}{}%
+Gillett, N. P., T. D. Kell, and P. D. Jones. 2006. ``Regional Climate Impacts of the Southern Annular Mode.'' \emph{Geophysical Research Letters} 33 (23). \url{https://doi.org/10.1029/2006GL027721}.
+
+\leavevmode\hypertarget{ref-gong1999}{}%
+Gong, Daoyi, and Shaowu Wang. 1999. ``Definition of Antarctic Oscillation Index.'' \emph{Geophysical Research Letters} 26 (4): 459--62. \url{https://doi.org/10.1029/1999GL900003}.
+
+\leavevmode\hypertarget{ref-hendon2007}{}%
+Hendon, Harry H., David W.
+J.
+Thompson, and Matthew C.
+Wheeler.
+2007.
+``Australian Rainfall and Surface Temperature Variations Associated with the Southern Hemisphere Annular Mode.'' \emph{J.
+Climate} 20 (11).
+American Meteorological Society: 2452--67.
+\url{https://doi.org/10.1175/JCLI4134.1}.
+
+\leavevmode\hypertarget{ref-hersbach2020}{}%
+Hersbach, Hans, Bill Bell, Paul Berrisford, Shoji Hirahara, András Horányi, Joaquín Muñoz-Sabater, Julien Nicolas, et al.
+2020.
+``The ERA5 Global Reanalysis.'' \emph{Quarterly Journal of the Royal Meteorological Society} 146 (730): 1999--2049.
+\url{https://doi.org/10.1002/qj.3803}.
+
+\leavevmode\hypertarget{ref-ho2012}{}%
+Ho, M., A.
+S.
+Kiem, and D.
+C.
+Verdon-Kidd.
+2012.
+``The Southern Annular Mode: A Comparison of Indices.'' \emph{Hydrology and Earth System Sciences} 16 (3).
+Copernicus GmbH: 967--82.
+\url{https://doi.org/10.5194/hess-16-967-2012}.
+
+\leavevmode\hypertarget{ref-holton1980}{}%
+Holton, James R., and Hsiu-Chi Tan.
+1980.
+``The Influence of the Equatorial Quasi-Biennial Oscillation on the Global Circulation at 50 Mb.'' \emph{J.
+Atmos.
+Sci.} 37 (10).
+American Meteorological Society: 2200--2208.
+\href{https://doi.org/10.1175/1520-0469(1980)037\%3C2200:TIOTEQ\%3E2.0.CO;2}{https://doi.org/10.1175/1520-0469(1980)037\textless{}2200:TIOTEQ\textgreater{}2.0.CO;2}.
+
+\leavevmode\hypertarget{ref-hufkens2020}{}%
+Hufkens, Koen, Reto Stauffer, and Elio Campitelli. 2020. ``Ecmwfr: Interface to 'ECMWF' and 'CDS' Data Web Services.''
+
+\leavevmode\hypertarget{ref-jones2009}{}%
+Jones, Julie M., Ryan L.
+Fogt, Martin Widmann, Gareth J.
+Marshall, Phil D.
+Jones, and Martin Visbeck.
+2009.
+``Historical SAM Variability.
+Part I: Century-Length Seasonal Reconstructions.'' \emph{J.
+Climate} 22 (20).
+American Meteorological Society: 5319--45.
+\url{https://doi.org/10.1175/2009JCLI2785.1}.
+
+\leavevmode\hypertarget{ref-jones2019}{}%
+Jones, Megan E., David H.
+Bromwich, Julien P.
+Nicolas, Jorge Carrasco, Eva Plavcová, Xun Zou, and Sheng-Hung Wang.
+2019.
+``Sixty Years of Widespread Warming in the Southern Middle and High Latitudes (19572016).'' \emph{J.
+Climate} 32 (20).
+American Meteorological Society: 6875--98.
+\url{https://doi.org/10.1175/JCLI-D-18-0565.1}.
+
+\leavevmode\hypertarget{ref-karoly1989}{}%
+Karoly, David J.
+1989.
+``Southern Hemisphere Circulation Features Associated with El Niño-Southern Oscillation Events.'' \emph{J.
+Climate} 2 (11): 1239--52.
+\href{https://doi.org/10.1175/1520-0442(1989)002\%3C1239:SHCFAW\%3E2.0.CO;2}{https://doi.org/10.1175/1520-0442(1989)002\textless{}1239:SHCFAW\textgreater{}2.0.CO;2}.
+
+\leavevmode\hypertarget{ref-katz1991}{}%
+Katz, Richard W., and Barbara G.
+Brown.
+1991.
+``The Problem of Multiplicity in Research on Teleconnections.'' \emph{International Journal of Climatology} 11 (5): 505--13.
+\url{https://doi.org/10.1002/joc.3370110504}.
+
+\leavevmode\hypertarget{ref-kidson1988}{}%
+Kidson, John W.
+1988.
+``Interannual Variations in the Southern Hemisphere Circulation.'' \emph{J.
+Climate} 1 (12): 1177--98.
+\href{https://doi.org/10.1175/1520-0442(1988)001\%3C1177:IVITSH\%3E2.0.CO;2}{https://doi.org/10.1175/1520-0442(1988)001\textless{}1177:IVITSH\textgreater{}2.0.CO;2}.
+
+\leavevmode\hypertarget{ref-lim2016}{}%
+Lim, Eun-Pa, Harry H.
+Hendon, Julie M.
+Arblaster, Francois Delage, Hanh Nguyen, Seung-Ki Min, and Matthew C.
+Wheeler.
+2016.
+``The Impact of the Southern Annular Mode on Future Changes in Southern Hemisphere Rainfall.'' \emph{Geophysical Research Letters} 43 (13): 7160--7.
+\url{https://doi.org/10.1002/2016GL069453}.
+
+\leavevmode\hypertarget{ref-marshall2003}{}%
+Marshall, Gareth J.
+2003.
+``Trends in the Southern Annular Mode from Observations and Reanalyses.'' \emph{J.
+Climate} 16 (24).
+American Meteorological Society: 4134--43.
+\href{https://doi.org/10.1175/1520-0442(2003)016\%3C4134:TITSAM\%3E2.0.CO;2}{https://doi.org/10.1175/1520-0442(2003)016\textless{}4134:TITSAM\textgreater{}2.0.CO;2}.
+
+\leavevmode\hypertarget{ref-marshall2004}{}%
+Marshall, Gareth J., Peter A.
+Stott, John Turner, William M.
+Connolley, John C.
+King, and Thomas A.
+Lachlan-Cope.
+2004.
+``Causes of Exceptional Atmospheric Circulation Changes in the Southern Hemisphere.'' \emph{Geophysical Research Letters} 31 (14).
+\url{https://doi.org/10.1029/2004GL019952}.
+
+\leavevmode\hypertarget{ref-mo1987}{}%
+Mo, Kingtse C., and Michael Ghil.
+1987.
+``Statistics and Dynamics of Persistent Anomalies.'' \emph{J.
+Atmos.
+Sci.} 44 (5): 877--902.
+\href{https://doi.org/10.1175/1520-0469(1987)044\%3C0877:SADOPA\%3E2.0.CO;2}{https://doi.org/10.1175/1520-0469(1987)044\textless{}0877:SADOPA\textgreater{}2.0.CO;2}.
+
+\leavevmode\hypertarget{ref-raphael2004}{}%
+Raphael, M. N. 2004. ``A Zonal Wave 3 Index for the Southern Hemisphere.'' \emph{Geophysical Research Letters} 31 (23). \url{https://doi.org/10.1029/2004GL020365}.
+
+\leavevmode\hypertarget{ref-rcoreteam2020}{}%
+R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Manual. Vienna, Austria: R Foundation for Statistical Computing.
+
+\leavevmode\hypertarget{ref-rogers1982}{}%
+Rogers, Jeffery C., and Harry van Loon.
+1982.
+``Spatial Variability of Sea Level Pressure and 500 Mb Height Anomalies over the Southern Hemisphere.'' \emph{Mon.
+Wea.
+Rev.} 110 (10).
+American Meteorological Society: 1375--92.
+\href{https://doi.org/10.1175/1520-0493(1982)110\%3C1375:SVOSLP\%3E2.0.CO;2}{https://doi.org/10.1175/1520-0493(1982)110\textless{}1375:SVOSLP\textgreater{}2.0.CO;2}.
+
+\leavevmode\hypertarget{ref-ropelewski1987}{}%
+Ropelewski, C.
+F., and P.
+D.
+Jones.
+1987.
+``An Extension of the TahitiDarwin Southern Oscillation Index.'' \emph{Mon.
+Wea.
+Rev.} 115 (9).
+American Meteorological Society: 2161--5.
+\href{https://doi.org/10.1175/1520-0493(1987)115\%3C2161:AEOTTS\%3E2.0.CO;2}{https://doi.org/10.1175/1520-0493(1987)115\textless{}2161:AEOTTS\textgreater{}2.0.CO;2}.
+
+\leavevmode\hypertarget{ref-rosso2018}{}%
+Rosso, Flávia Venturini, Nathalie Tissot Boiaski, Simone Erotildes Teleginski Ferraz, and Tiago Capello Robles.
+2018.
+``Influence of the Antarctic Oscillation on the South Atlantic Convergence Zone.'' \emph{Atmosphere} 9 (11).
+Multidisciplinary Digital Publishing Institute: 431.
+\url{https://doi.org/10.3390/atmos9110431}.
+
+\leavevmode\hypertarget{ref-schneider2015}{}%
+Schneider, Udo, Andreas Becker, Peter Finger, Anja Meyer-Christoffer, Bruno Rudolf, and Markus Ziese.
+2015.
+``GPCC Full Data Reanalysis Version 7.0 at 0.5: Monthly Land-Surface Precipitation from Rain-Gauges Built on GTS-Based and Historic Data: Gridded Monthly Totals.'' Global Precipitation Climatology Centre (GPCC) at Deutscher Wetterdienst.
+\url{https://doi.org/10.5676/DWD_GPCC/FD_M_V7_050}.
+
+\leavevmode\hypertarget{ref-schneider2017}{}%
+Schneider, Udo, Peter Finger, Anja Meyer-Christoffer, Elke Rustemeier, Markus Ziese, and Andreas Becker.
+2017.
+``Evaluating the Hydrological Cycle over Land Using the Newly-Corrected Precipitation Climatology from the Global Precipitation Climatology Centre (GPCC).'' \emph{Atmosphere} 8 (3).
+Multidisciplinary Digital Publishing Institute: 52.
+\url{https://doi.org/10.3390/atmos8030052}.
+
+\leavevmode\hypertarget{ref-sheather1991}{}%
+Sheather, S.
+J., and M.
+C.
+Jones.
+1991.
+``A Reliable Data-Based Bandwidth Selection Method for Kernel Density Estimation.'' \emph{Journal of the Royal Statistical Society.
+Series B (Methodological)} 53 (3).
+{[}Royal Statistical Society, Wiley{]}: 683--90.
+
+\leavevmode\hypertarget{ref-silvestri2009}{}%
+Silvestri, Gabriel, and Carolina Vera.
+2009.
+``Nonstationary Impacts of the Southern Annular Mode on Southern Hemisphere Climate.'' \emph{J.
+Climate} 22 (22).
+American Meteorological Society: 6142--8.
+\url{https://doi.org/10.1175/2009JCLI3036.1}.
+
+\leavevmode\hypertarget{ref-simpkins2012}{}%
+Simpkins, Graham R., and Alexey Yu.
+Karpechko.
+2012.
+``Sensitivity of the Southern Annular Mode to Greenhouse Gas Emission Scenarios.'' \emph{Clim Dyn} 38 (3): 563--72.
+\url{https://doi.org/10.1007/s00382-011-1121-2}.
+
+\leavevmode\hypertarget{ref-smith2008}{}%
+Smith, Thomas M., Richard W.
+Reynolds, Thomas C.
+Peterson, and Jay Lawrimore.
+2008.
+``Improvements to NOAA's Historical Merged LandOcean Surface Temperature Analysis (18802006).'' \emph{J.
+Climate} 21 (10).
+American Meteorological Society: 2283--96.
+\url{https://doi.org/10.1175/2007JCLI2100.1}.
+
+\leavevmode\hypertarget{ref-vose2012}{}%
+Vose, Russell S., Derek Arndt, Viva F.
+Banzon, David R.
+Easterling, Byron Gleason, Boyin Huang, Ed Kearns, et al.
+2012.
+``NOAA's Merged LandOcean Surface Temperature Analysis.'' \emph{Bull.
+Amer.
+Meteor.
+Soc.} 93 (11).
+American Meteorological Society: 1677--85.
+\url{https://doi.org/10.1175/BAMS-D-11-00241.1}.
+
+\leavevmode\hypertarget{ref-walker1914}{}%
+Walker, Sir Gilbert Thomas. 1914. \emph{Correlation in Seasonal Variations of Weather, III: On the Criterion for the Reality of Relationships or Periodicities}. Meteorological Office.
+
+\leavevmode\hypertarget{ref-watson2014}{}%
+Watson, Peter A.
+G., and Lesley J.
+Gray.
+2014.
+``How Does the Quasi-Biennial Oscillation Affect the Stratospheric Polar Vortex?'' \emph{J.
+Atmos.
+Sci.} 71 (1).
+American Meteorological Society: 391--409.
+\url{https://doi.org/10.1175/JAS-D-13-096.1}.
+
+\leavevmode\hypertarget{ref-wickham2009}{}%
+Wickham, Hadley. 2009. \emph{Ggplot2: Elegant Graphics for Data Analysis}. Use R! New York: Springer-Verlag. \url{https://doi.org/10.1007/978-0-387-98141-3}.
+
+\leavevmode\hypertarget{ref-wilks2016}{}%
+Wilks, D.
+S.
+2016.
+```The Stippling Shows Statistically Significant Grid Points': How Research Results Are Routinely Overstated and Overinterpreted, and What to Do About It.'' \emph{Bull.
+Amer.
+Meteor.
+Soc.} 97 (12): 2263--73.
+\url{https://doi.org/10.1175/BAMS-D-15-00267.1}.
+
+\leavevmode\hypertarget{ref-wolter2011}{}%
+Wolter, Klaus, and Michael S.
+Timlin.
+2011.
+``El Niño/Southern Oscillation Behaviour Since 1871 as Diagnosed in an Extended Multivariate ENSO Index (MEI.Ext).'' \emph{International Journal of Climatology} 31 (7): 1074--87.
+\url{https://doi.org/10.1002/joc.2336}.
+
+\leavevmode\hypertarget{ref-xie2015}{}%
+Xie, Yihui. 2015. \emph{Dynamic Documents with R and Knitr}. Second. Boca Raton, Florida: Chapman and Hall/CRC.
+
+\leavevmode\hypertarget{ref-zamboni2010}{}%
+Zamboni, Laura, Carlos R.
+Mechoso, and Fred Kucharski.
+2010.
+``Relationships Between Upper-Level Circulation over South America and Rainfall over Southeastern South America: A Physical Base for Seasonal Predictions.'' \emph{J.
+Climate} 23 (12).
+American Meteorological Society: 3300--3315.
+\url{https://doi.org/10.1175/2009JCLI3129.1}.
+
+\leavevmode\hypertarget{ref-zhang2020}{}%
+Zhang, Ruhua, Wenshou Tian, and Tao Wang.
+2020.
+``Role of the Quasi-Biennial Oscillation in the Downward Extension of Stratospheric Northern Annular Mode Anomalies.'' \emph{Clim Dyn} 55 (3): 595--612.
+\url{https://doi.org/10.1007/s00382-020-05285-4}.
+
+\bibliographystyle{spphys}
+\bibliography{AsymSAM.bib,packages.bib}
+
 \end{document}

### 2020-11-25 analysis/paper2/paper2.tex
Agrega figuras climdyn

@@
 \begin{figure*}
-\includegraphics{figures/method-1} \caption{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
+\includegraphics{method-1} \caption{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 \begin{figure*}
-\includegraphics{figures/asymsam-timeseries-1} \caption{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
+\includegraphics{asymsam-timeseries-1} \caption{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
 To the right, probability density estimate of each index.
@@
 \begin{figure}
-\includegraphics{figures/cor-lev-1} \caption{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.}\label{fig:cor-lev}
+\includegraphics{cor-lev-1} \caption{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.}\label{fig:cor-lev}
 \end{figure}
@@
 \begin{figure*}
-\includegraphics{figures/cross-correlation-1} \caption{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.}\label{fig:cross-correlation}
+\includegraphics{cross-correlation-1} \caption{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.}\label{fig:cross-correlation}
 \end{figure*}
@@
 \begin{figure*}
-\includegraphics{figures/trends-1} \caption{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+\includegraphics{trends-1} \caption{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
 Shading indicates the 95\% confidence interval from a t-distribution.}\label{fig:trends}
@@
 \begin{figure}
-\includegraphics{figures/r-squared-trend-1} \caption{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018.
+\includegraphics{r-squared-trend-1} \caption{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018.
 Shading indicates the 95\% confidence interval.}\label{fig:r-squared-trend}
@@
 \begin{figure*}
-\includegraphics{figures/2d-regr-1} \caption{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{2d-regr-1} \caption{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period.
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices.
@@
 \begin{figure*}
-\includegraphics{figures/wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
+\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
 \end{figure*}
@@
 \begin{figure}
-\includegraphics{figures/vertical-regression-1} \caption{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
+\includegraphics{vertical-regression-1} \caption{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
 (a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979 -- 2018 period.}\label{fig:vertical-regression}
@@
 \begin{figure*}
-\includegraphics{figures/regr-air-season-1} \caption{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{regr-air-season-1} \caption{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 \begin{figure*}
-\includegraphics{figures/pp-regr-oceania-1} \caption{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{pp-regr-oceania-1} \caption{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
@@
 \begin{figure*}
-\includegraphics{figures/pp-regr-america-1} \caption{Same as Figure \\ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
+\includegraphics{pp-regr-america-1} \caption{Same as Figure \\ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
 \end{figure*}
@@
 \begin{figure}[ht]
-\includegraphics{figures/A1-1} \caption{Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
+\includegraphics{A1-1} \caption{Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
 Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.
@@
 \begin{figure}
-\includegraphics{figures/A2-1} \caption{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
+\includegraphics{A2-1} \caption{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
 The shading indicates de 95\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\% of the simulated sampels had an amplitude equal or lower).
@@
 \begin{figure}
-\includegraphics{figures/A3-1} \caption{50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+\includegraphics{A3-1} \caption{50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
 Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A3}
@@
 \begin{figure}
-\includegraphics{figures/A4-1} \caption{700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+\includegraphics{A4-1} \caption{700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
 Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A4}
@@
 \begin{figure}
-\includegraphics{figures/A5-1} \caption{Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
+\includegraphics{A5-1} \caption{Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{figures/A6-1} \caption{Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
+\includegraphics{A6-1} \caption{Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
 Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.}\label{fig:A6}

### 2020-11-25 analysis/paper2/paper2.Rmd
Funciona localmente

@@
 bibliography: 
-  - AsymSAM.bib
-  - packages.bib
+  - AsymSAM
+  - packages
 bibstyle: spphys

### 2020-11-25 analysis/paper2/paper2.Rmd
use natbib

@@
       latex_engine: "xelatex"
-      # citation_package: "natbib"
+      citation_package: "natbib"
 header-includes: 

### 2020-11-25 analysis/paper2/paper2.tex
Funciona localmente
use natbib

--- 2020-10-10 analysis/paper/paper.tex
@@
 
-%% Necessary!
+\title{Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
+Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }
+
+    \titlerunning{Assessment of zonal symmetric and asymmetric components of the SAM}
+
+\author{  Elio Campitelli \and  Leandro B. Díaz \and  Carolina Vera \and  }
+
+\institute{
+        Elio Campitelli \at
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+ CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+ CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     \email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}  %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+    \and
+        Leandro B. Díaz \at
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+ CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+ CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+    \and
+        Carolina Vera \at
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+ CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+ CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+    \and
+    }
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
 \maketitle
 
-\bibliographystyle{ametsoc2014}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% SIGNIFICANCE STATEMENT/CAPSULE SUMMARY
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% If you are including an optional significance statement for a journal article or a required capsule summary for BAMS
-% (see www.ametsoc.org/ams/index.cfm/publications/authors/journal-and-bams-authors/formatting-and-manuscript-components for details),
-% please apply the necessary command as shown below:
-%
-\statement
-This is significant becasue I wrote it.
+\begin{abstract}
+The text of your abstract. 150 -- 250 words.
+\\
+\keywords{
+        Southern Annular Mode \and
+        general circulation \and
+        zonally asymmetric circulation \and
+        El Niño Southern Oscillation \and
+    }
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MAIN BODY OF PAPER
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
+    \subclass{
+                    MSC code 1 \and
+                    MSC code 2 \and
+            }
 
-\section{Introduction}
+\end{abstract}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 
-\subsubsection{Data}
+\hypertarget{data}{%
+\subsubsection{Data}\label{data}}
 
@@
 
-\subsubsection{Definition of indices}
+\hypertarget{definition-of-indices}{%
+\subsubsection{Definition of indices}\label{definition-of-indices}}
 
@@
 \begin{figure*}
-\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period]{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
+\includegraphics{method-1} \caption{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 \citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
-Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures A3 and A4).
+Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures \ref{fig:A3} and \ref{fig:A4}).
 Therefore, we believe that our method is a reasonable approximation of the phenomenon.
@@
 \citet{silvestri2009} suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes.
-Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure A6).
+Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure \ref{fig:A6}).
 
-\subsubsection{Regressions}
+\hypertarget{regressions}{%
+\subsubsection{Regressions}\label{regressions}}
 
@@
 
-\subsubsection{Computation procedures}
+\hypertarget{computation-procedures}{%
+\subsubsection{Computation procedures}\label{computation-procedures}}
 
@@
 \begin{figure*}
-\includegraphics{asymsam-timeseries-1} \caption[Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa]{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
+\includegraphics{asymsam-timeseries-1} \caption{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
 To the right, probability density estimate of each index.
@@
 
-The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A2).
+The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \ref{fig:A2}).
 This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation \citep{baldwin2001b}.
@@
 \begin{figure}
-\includegraphics{cor-lev-1} \caption[Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period]{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.}\label{fig:cor-lev}
+\includegraphics{cor-lev-1} \caption{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.}\label{fig:cor-lev}
 \end{figure}
@@
 \begin{figure*}
-\includegraphics{cross-correlation-1} \caption[Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period]{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.}\label{fig:cross-correlation}
+\includegraphics{cross-correlation-1} \caption{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.}\label{fig:cross-correlation}
 \end{figure*}
@@
 \begin{figure*}
-\includegraphics{trends-1} \caption[Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index]{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+\includegraphics{trends-1} \caption{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
 Shading indicates the 95\% confidence interval from a t-distribution.}\label{fig:trends}
@@
 \begin{figure}
-\includegraphics{r-squared-trend-1} \caption[Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018]{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018.
+\includegraphics{r-squared-trend-1} \caption{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018.
 Shading indicates the 95\% confidence interval.}\label{fig:r-squared-trend}
@@
 \begin{figure*}
-\includegraphics{2d-regr-1} \caption[Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period]{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{2d-regr-1} \caption{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period.
 The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices.
@@
 \begin{figure*}
-\includegraphics{wave-amplitude-1} \caption[Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean]{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
+\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
 \end{figure*}
@@
 \begin{figure}
-\includegraphics{vertical-regression-1} \caption[Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM)]{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
+\includegraphics{vertical-regression-1} \caption{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
 (a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979 -- 2018 period.}\label{fig:vertical-regression}
@@
 \begin{figure*}
-\includegraphics{regr-air-season-1} \caption[Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period]{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{regr-air-season-1} \caption{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 \begin{figure*}
-\includegraphics{pp-regr-oceania-1} \caption[Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period]{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{pp-regr-oceania-1} \caption{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
@@
 \begin{figure*}
-\includegraphics{pp-regr-america-1} \caption[Same as Figure \\ref{fig:pp-regr-oceania} but for South America]{Same as Figure \\ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
+\includegraphics{pp-regr-america-1} \caption{Same as Figure \\ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
 \end{figure*}
@@
 
-\subsection{Conclusions}
+\hypertarget{conclusions}{%
+\subsection{Conclusions}\label{conclusions}}
 
@@
 
-\acknowledgments
+\begin{acknowledgements}
+NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ 
+\end{acknowledgements}
 
-NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/
-
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
-
-\datastatement
+\hypertarget{atastatement}{%
+\subsubsection{atastatement}\label{atastatement}}
 
 All data used in this paper is freely available in their respective sources.
-ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means).
-NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html).
-The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: https://www.cpc.ncep.noaa.gov/products/analysis\_monitoring/ensostuff/detrend.nino34.ascii.txt
+ERA5 data can be obtained via the Copernicus Climate Data Store (\url{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means}).
+NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (\url{https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html} and \url{https://psl.noaa.gov/data/gridded/data.gpcc.html}).
+The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}
 
-A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at https://github.com/eliocamp/asymsam.
-
-\bibliography{AsymSAM,packages}
+A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at \url{https://github.com/eliocamp/asymsam}.
 
@@
 
-\appendixtitle{Extra figures}
+\counterwithin{figure}{section}
 
-\begin{figure}
-\includegraphics{A1-1} \appendcaption{A1}{Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
+\hypertarget{extra-figures}{%
+\section{Extra figures}\label{extra-figures}}
+
+\newpage
+
+\begin{figure}[ht]
+\includegraphics{A1-1} \caption{Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
 Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.
@@
 \begin{figure}
-\includegraphics{A2-1} \appendcaption{A2}{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
+\includegraphics{A2-1} \caption{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
 The shading indicates de 95\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\% of the simulated sampels had an amplitude equal or lower).
@@
 \begin{figure}
-\includegraphics{A3-1} \caption[50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period]{50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+\includegraphics{A3-1} \caption{50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
 Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A3}
@@
 \begin{figure}
-\includegraphics{A4-1} \caption[700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period]{700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+\includegraphics{A4-1} \caption{700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
 Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A4}
@@
 \begin{figure}
-\includegraphics{A5-1} \caption[Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period]{Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
+\includegraphics{A5-1} \caption{Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A6-1} \caption[Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018]{Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
+\includegraphics{A6-1} \caption{Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
 Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.}\label{fig:A6}
@@
 
+\bibliographystyle{spphys}
+\bibliography{AsymSAM,packages}
+
 \end{document}

### 2020-11-26 analysis/paper/paper2.Rmd
Mueve todo a paper

--- 2020-11-25 analysis/paper2/paper2.Rmd
@@
   - packages
-bibstyle: spphys
+bibstyle: spbasic
 # bibstyle options spbasic(default), spphys, spmpsci
@@
 ```
+
+\newpage

### 2020-11-26 analysis/paper/paper.Rmd
Agrega correcciones de Carolina

--- 2020-10-10 analysis/paper/paper.Rmd
@@
       CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+
+- name:  Carolina Vera
+  address: |
+    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+      CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+      CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+
+keywords:
+  - Southern Annular Mode
+  - general circulation
+  - zonally asymmetric circulation
+  - El Niño Southern Oscillation
+
+#PACS: 
+#- PAC1
+#- superPAC
+
+MSC:
+- MSC code 1
+- MSC code 2
+
 abstract: |
-  Enter the text of your abstract here.
-This is a sample American Meteorological Society (AMS) \LaTeX\ template.
-This document provides authors with instructions on the use of the AMS \LaTeX\ template.
-Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document.
-The template.tex file should be modified by authors for their own manuscript.
-statement: |
-  This is significant becasue I wrote it.
-bibliography: ["AsymSAM.bib","packages.bib"]
+  The text of your abstract.  150 -- 250 words.
+
+bibliography: 
+  - AsymSAM
+  - packages
+bibstyle: spbasic
+# bibstyle options spbasic(default), spphys, spmpsci
 output: 
   bookdown::pdf_book: 
-    base_format: rticles::ams_article
+      base_format: rticles::springer_article
       latex_engine: "xelatex"
-  bookdown::word_document2: default
+      citation_package: "natbib"
 header-includes: 
@@
 
-The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a;@fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [e.g.
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] on daily, monthly, and decadal timescales [@baldwin2001a;@fogt2006] and exerts an important influence on temperature and precipitation anomalies, and sea ice concentration [e.g.
 @fogt2020].
-Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
+Its positive phase is usually described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
 
@@
 
-Even though these indices are based on zonal averages, spatial anomalies of geopotential associated with them (either by regression or composition) contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
-These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM [@fan2007;@fogt2012;@rosso2018], going as far as reversing its relationship between precipitation in South America [@silvestri2009].
-At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].
-
-At least some of the variability associated with the zonal asymmetries of the SAM is probably forced by the tropics.
-In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987;@kidson1988;@karoly1989], whose wave train projects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
-And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed [@fan2007;@fogt2011; @clem2013].
-In particular, @fan2007 computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated if the (linear) signal of the ENSO was removed.
-While this relates to temporally coherent variability of the two hemisphere and not necessarily to zonal asymmetries in the associated spatial patterns, it is nonetheless consistent of the ENSO playing a crucial role in zonal asymmetries of the SAM.
-
-Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn [e.g.
+Even though these indices are based on zonal averages, their associated geopotential height spatial anomalies contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
+The zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM [@fan2007;@silvestri2009;@fogt2012;@rosso2018].
+The fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].
+
+Some of the variability associated with the zonal asymmetries of the SAM seems to be forced by the tropics.
+ENSO-like variability affects the Southern Hemisphere extratopics through the Rossby wave trains [@mo1987;@kidson1988;@karoly1989] which project strongly onto the zonal anomalies associated with the SAM in the Pacific sector.
+Moreover, tropical influences on the SAM have been observed [@fan2007;@fogt2011; @clem2013].
+@fan2007 computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated to each other if the (linear) signal of the ENSO was removed.
+
+Positive trends in SAM index have been documented by various researchers using different indices mostly on austral Summer and Autumn [e.g.
 @fogt2020 and references therein].
-These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013].
-However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variability could be masking influencing the observed trends independently.
-
-Similarly unclear are the specific impacts of the zonally asymmetric component of the SAM.
-Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012].
-The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
+These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013].
+However, it’s not clear yet how or if the asymmetric SAM component responds to this forcing, or how its variability alters the observed trends.
+
+The impact of the zonally asymmetric component of the SAM at regional scales has not been studied in detail yet.
+Positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM).
+But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012].
+SAM signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
 The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
-For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
+For example, the SAM-precipitation relationshipin Southeastern South America can be explained by the PSA-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
 @fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
 
-We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of @fogt2012.
-However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988).
-This is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].
+The study of the temporal variability of the asymmetric component of the SAM has not received much attention except for @fogt2012.
+This study provides evidences of the relevance of SAM asymmetric component.
+However, their conclusions  are based on composites of positive and negative SAM events including a small number of cases unevenly distributed among years with and without satellite information.
+The latter  is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].
 Moreover, @fogt2012 studied the zonal asymmetric component of the SAM only in sea level pressure.
@@
 
-Our objective is, then, to systematically characterise the zonally asymmetric and symmetric components of the SAM variability.
-For each level, we construct two indices which aim to capture exclusively the variability of the symmetric and asymmetric component respectively.
+Our objective is, then, to describe the zonally asymmetric and symmetric components of the SAM variability.
+We first propose a methodology that provides for each level, two indices which aim to capture independently the variability of the symmetric and asymmetric SAM component respectively.
 We assess their vertical structure and coherence, temporal variability and trends.
@@
 
-### Data
-
-To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
-We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.
-
-To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude, and monthly rainfall 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015;@schneider2017].
-The rainfall product uses station-based records, and thus it only has continental coverage.
-
-### Definition of indices
-
-Traditionally the Souther Annular Mode (SAM) is defined as de leading Empirical Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels [@ho2012].
-Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the the leading EOF of the monthly anomalies of geopotential field south of 20\degree S at each level.
+## Data
+
+We used monthly geopotential height at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
+We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations.
+
+Temperature data used are from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude.
+Monthly rainfall fields at 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015;@schneider2017] are also considered.
+The rainfall dataset is based on station-based records, and thus it only has continental coverage.
+
+## Definition of indices
+
+Traditionally the SAM is defined as de leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels [@ho2012].
+Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the leading EOF of the monthly anomalies of geopotential height  south of 20\degree S at each level.
 We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in `r dims$t` rows and  `r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points of latitude).
 We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999].
-
-To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \@ref(fig:method) for 700 hPa.
+We consider in the EOF analysis all months together without separating in seasons.
+
+To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \@ref(fig:method) at 700 hPa.
 The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
-We then compute the "Full SAM", "Asymmetric SAM" and "Symmetric SAM" indices as the regression coefficients of the regression of each monthly geopotential field on the respective patterns (weighting by the cosine of latitude).
-The three indices are normalized by dividing them by the standard deviation of the "Full" index at each level.
+We then compute the "Full SAM", "Asymmetric SAM" and "Symmetric SAM" indices as the regression coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude).
+The three indices are then normalized by dividing them by the standard deviation of the "Full SAM" index at each level.
 As a result, the magnitudes between indices are comparable.
-However, only "Full" index will have unit standard deviation per definition.
-From the regression, we also use the explained variance of each pattern as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
+However, only Full SAM index has unit standard deviation per definition.
+We also use the explained variance of each pattern as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
 
@@
 
-Our method assumes linearity in the asymmetric component of the SAM.
-That means we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM.
-@fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
-Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures A3 and A4).
-Therefore, we believe that our method is a reasonable approximation of the phenomenon.
-
-By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
-Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure A5, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
-In the stratosphere (Figure A5, row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree\ in comparison with the rest of the year.
+The method assumes linearity in the asymmetric component of the SAM.
+That means that zonal symmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
+@fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
+To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the Full SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation) for the period from 1979 to 2018 at the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures \@ref(fig:A3) and \@ref(fig:A4)).
+In all seasons and both levels, SAM+ composites are similar to SAM- in structure but the sign switched.
+Pattern correlations (computed as the Pearson correlation between fields) between SAM+ and SAM- composites for each season are high.
+The method considered in this study seems then a reasonable approximation of the phenomenon.
+
+By performing the EOF analysis using data for all months we are assuming that the zonal asymmetric structure of the SAM is the same at all seasons.
+The latter was assessed by computing geopotential height zonal anomalies by projecting the first EOF of each season independently.
+The following seasons were considered – December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
+Results are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
+In the stratosphere (Figure \@ref(fig:A5), row 2), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree\ in comparison with the rest of the year.
 Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
-Based on this, we believe that our initial assumption is not unreasonable throught most of the year with the exception of DJF in the stratosphere.
-
-Finally, we assume that the zonally asymmetric pattern is stationary in time.
-@silvestri2009 suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes.
-Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure A6).
-
-### Regressions
+Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year with the exception of DJF in the stratosphere.
+
+The method also implies that the zonally asymmetric pattern of SAM remains stationary along the period considered.
+@silvestri2009 suggest that this might not be the case between 1958 and 2004.
+Zonal asymmetric patterns of SAM were computed for the two halves of the period (1979 to 1998 and 1999 to 2018) respectively.
+They show no systematic change neither in the stratosphere nor in the troposphere (Figure \@ref(fig:A6)).
+
+## Regressions
 
@@
 
-### Computation procedures
+## Computation procedures
 
@@
 
-\acknowledgments
+\begin{acknowledgements}
 NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ 
-
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
-
-\datastatement
+\end{acknowledgements}
+
+## datastatement
 All data used in this paper is freely available in their respective sources.

### 2020-11-26 analysis/paper/paper.tex
Agrega correcciones de Carolina

@@
 
-%% Necessary!
+\title{Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
+Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }
+
+    \titlerunning{Assessment of zonal symmetric and asymmetric components of the SAM}
+
+\author{  Elio Campitelli \and  Leandro B. Díaz \and  Carolina Vera \and  }
+
+\institute{
+        Elio Campitelli \at
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+ CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+ CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     \email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}  %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+    \and
+        Leandro B. Díaz \at
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+ CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+ CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+    \and
+        Carolina Vera \at
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
+ CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
+ CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+    \and
+    }
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
 \maketitle
 
-\bibliographystyle{ametsoc2014}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% SIGNIFICANCE STATEMENT/CAPSULE SUMMARY
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% If you are including an optional significance statement for a journal article or a required capsule summary for BAMS
-% (see www.ametsoc.org/ams/index.cfm/publications/authors/journal-and-bams-authors/formatting-and-manuscript-components for details),
-% please apply the necessary command as shown below:
-%
-\statement
-This is significant becasue I wrote it.
+\begin{abstract}
+The text of your abstract. 150 -- 250 words.
+\\
+\keywords{
+        Southern Annular Mode \and
+        general circulation \and
+        zonally asymmetric circulation \and
+        El Niño Southern Oscillation \and
+    }
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MAIN BODY OF PAPER
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
+    \subclass{
+                    MSC code 1 \and
+                    MSC code 2 \and
+            }
 
-\section{Introduction}
+\end{abstract}
 
-The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation \citep{rogers1982} in daily, monthly, and decadal timescales \citep{baldwin2001a, fogt2006} and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration \citep[e.g.][]{fogt2020}.
-Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation \citep{rogers1982} on daily, monthly, and decadal timescales \citep{baldwin2001a, fogt2006} and exerts an important influence on temperature and precipitation anomalies, and sea ice concentration \citep[e.g.][]{fogt2020}.
+Its positive phase is usually described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
 
@@
 
-Even though these indices are based on zonal averages, spatial anomalies of geopotential associated with them (either by regression or composition) contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
-These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM \citep{fan2007, fogt2012, rosso2018}, going as far as reversing its relationship between precipitation in South America \citep{silvestri2009}.
-At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere \citep{jones2009}.
+Even though these indices are based on zonal averages, their associated geopotential height spatial anomalies contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
+The zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM \citep{fan2007, silvestri2009, fogt2012, rosso2018}.
+The fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere \citep{jones2009}.
 
-At least some of the variability associated with the zonal asymmetries of the SAM is probably forced by the tropics.
-In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989}, whose wave train projects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector.
-And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed \citep{fan2007, fogt2011, clem2013}.
-In particular, \citet{fan2007} computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated if the (linear) signal of the ENSO was removed.
-While this relates to temporally coherent variability of the two hemisphere and not necessarily to zonal asymmetries in the associated spatial patterns, it is nonetheless consistent of the ENSO playing a crucial role in zonal asymmetries of the SAM.
+Some of the variability associated with the zonal asymmetries of the SAM seems to be forced by the tropics.
+ENSO-like variability affects the Southern Hemisphere extratopics through the Rossby wave trains \citep{mo1987, kidson1988, karoly1989} which project strongly onto the zonal anomalies associated with the SAM in the Pacific sector.
+Moreover, tropical influences on the SAM have been observed \citep{fan2007, fogt2011, clem2013}.
+\citet{fan2007} computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated to each other if the (linear) signal of the ENSO was removed.
 
-Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn \citep[e.g.][ and references therein]{fogt2020}.
-These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables \citep{marshall2004, gillett2005, arblaster2006, gillett2013}.
-However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variability could be masking influencing the observed trends independently.
+Positive trends in SAM index have been documented by various researchers using different indices mostly on austral Summer and Autumn \citep[e.g.][ and references therein]{fogt2020}.
+These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases and understood in the context of zonal mean variables \citep{marshall2004, gillett2005, arblaster2006, gillett2013}.
+However, it's not clear yet how or if the asymmetric SAM component responds to this forcing, or how its variability alters the observed trends.
 
-Similarly unclear are the specific impacts of the zonally asymmetric component of the SAM.
-Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes \citep{jones2019} (and vice versa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}.
-The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry \citep{lim2016}.
+The impact of the zonally asymmetric component of the SAM at regional scales has not been studied in detail yet.
+Positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at higher latitudes \citep{jones2019} (and vice versa for negative SAM).
+But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}.
+SAM signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry \citep{lim2016}.
 The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
-For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}.
+For example, the SAM-precipitation relationshipin Southeastern South America can be explained by the PSA-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}.
 \citet{fan2007} also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
 
-We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of \citet{fogt2012}.
-However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988).
-This is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM \citep{silvestri2009}.
+The study of the temporal variability of the asymmetric component of the SAM has not received much attention except for \citet{fogt2012}.
+This study provides evidences of the relevance of SAM asymmetric component.
+However, their conclusions are based on composites of positive and negative SAM events including a small number of cases unevenly distributed among years with and without satellite information.
+The latter is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM \citep{silvestri2009}.
 Moreover, \citet{fogt2012} studied the zonal asymmetric component of the SAM only in sea level pressure.
@@
 
-\subsubsection{Data}
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
 
-To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{hersbach2020} for the period 1979 to 2018.
-We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.
+We used monthly geopotential height at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{hersbach2020} for the period 1979 to 2018.
+We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations.
 
-To describe the relationship between the SAM indices and temperature and precipitation, we use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land and ocean temperature analysis into a monthly global grid 5\degree~longitude by 5\degree~latitude, and monthly rainfall 0.5\degree~longitude by 0.5\degree~latitude data from the Global Precipitation Climatology Centre \citep{schneider2015, schneider2017}.
-The rainfall product uses station-based records, and thus it only has continental coverage.
+Temperature data used are from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land and ocean temperature analysis into a monthly global grid 5\degree~longitude by 5\degree~latitude.
+Monthly rainfall fields at 0.5\degree~longitude by 0.5\degree~latitude data from the Global Precipitation Climatology Centre \citep{schneider2015, schneider2017} are also considered.
+The rainfall dataset is based on station-based records, and thus it only has continental coverage.
 
-\subsubsection{Definition of indices}
+\hypertarget{definition-of-indices}{%
+\subsection{Definition of indices}\label{definition-of-indices}}
 
-Traditionally the Souther Annular Mode (SAM) is defined as de leading Empirical Orthogonal Mode (EOF) of sea level pressure or geopotential height at lower levels \citep{ho2012}.
-Following \citet{baldwin2001}, we extend that definition vertically and use the term SAM to refer to the the leading EOF of the monthly anomalies of geopotential field south of 20\degree S at each level.
+Traditionally the SAM is defined as de leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels \citep{ho2012}.
+Following \citet{baldwin2001}, we extend that definition vertically and use the term SAM to refer to the leading EOF of the monthly anomalies of geopotential height south of 20\degree S at each level.
 We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in 481 rows and 4176 columns (144 points of longitude and 29 points of latitude).
 We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint \citep{chung1999}.
+We consider in the EOF analysis all months together without separating in seasons.
 
-To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \ref{fig:method} for 700 hPa.
+To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \ref{fig:method} at 700 hPa.
 The full spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is the sum of the zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\)) and symmetric (\([\mathrm{EOF_1}](\lambda, \phi)\)) components.
-We then compute the ``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM'' indices as the regression coefficients of the regression of each monthly geopotential field on the respective patterns (weighting by the cosine of latitude).
-The three indices are normalized by dividing them by the standard deviation of the ``Full'' index at each level.
+We then compute the ``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM'' indices as the regression coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude).
+The three indices are then normalized by dividing them by the standard deviation of the ``Full SAM'' index at each level.
 As a result, the magnitudes between indices are comparable.
-However, only ``Full'' index will have unit standard deviation per definition.
-From the regression, we also use the explained variance of each pattern as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
+However, only Full SAM index has unit standard deviation per definition.
+We also use the explained variance of each pattern as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
 
 \begin{figure*}
-\includegraphics{method-1} \caption[Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period]{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
+\includegraphics{method-1} \caption{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 
-Our method assumes linearity in the asymmetric component of the SAM.
-That means we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM.
-\citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites.
-Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures A3 and A4).
-Therefore, we believe that our method is a reasonable approximation of the phenomenon.
+The method assumes linearity in the asymmetric component of the SAM.
+That means that zonal symmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
+\citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
+To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the Full SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation) for the period from 1979 to 2018 at the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures \ref{fig:A3} and \ref{fig:A4}).
+In all seasons and both levels, SAM+ composites are similar to SAM- in structure but the sign switched.
+Pattern correlations (computed as the Pearson correlation between fields) between SAM+ and SAM- composites for each season are high.
+The method considered in this study seems then a reasonable approximation of the phenomenon.
 
-By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
-Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure A5, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
-In the stratosphere (Figure A5, row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree~in comparison with the rest of the year.
+By performing the EOF analysis using data for all months we are assuming that the zonal asymmetric structure of the SAM is the same at all seasons.
+The latter was assessed by computing geopotential height zonal anomalies by projecting the first EOF of each season independently.
+The following seasons were considered -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
+Results are very similar to each other in the troposphere (Figure \ref{fig:A5}, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
+In the stratosphere (Figure \ref{fig:A5}, row 2), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree~in comparison with the rest of the year.
 Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
-Based on this, we believe that our initial assumption is not unreasonable throught most of the year with the exception of DJF in the stratosphere.
+Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year with the exception of DJF in the stratosphere.
 
-Finally, we assume that the zonally asymmetric pattern is stationary in time.
-\citet{silvestri2009} suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes.
-Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure A6).
+The method also implies that the zonally asymmetric pattern of SAM remains stationary along the period considered.
+\citet{silvestri2009} suggest that this might not be the case between 1958 and 2004.
+Zonal asymmetric patterns of SAM were computed for the two halves of the period (1979 to 1998 and 1999 to 2018) respectively.
+They show no systematic change neither in the stratosphere nor in the troposphere (Figure \ref{fig:A6}).
 
-\subsubsection{Regressions}
+\hypertarget{regressions}{%
+\subsection{Regressions}\label{regressions}}
 
@@
 
-\subsubsection{Computation procedures}
+\hypertarget{computation-procedures}{%
+\subsection{Computation procedures}\label{computation-procedures}}
 
@@
 
-\acknowledgments
+\begin{acknowledgements}
+NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ 
+\end{acknowledgements}
 
-NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/
-
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
-
-\datastatement
+\hypertarget{datastatement}{%
+\subsection{datastatement}\label{datastatement}}
 
 All data used in this paper is freely available in their respective sources.
-ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means).
-NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html).
-The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: https://www.cpc.ncep.noaa.gov/products/analysis\_monitoring/ensostuff/detrend.nino34.ascii.txt
+ERA5 data can be obtained via the Copernicus Climate Data Store (\url{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means}).
+NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (\url{https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html} and \url{https://psl.noaa.gov/data/gridded/data.gpcc.html}).
+The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}
 
-A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at https://github.com/eliocamp/asymsam.
-
-\bibliography{AsymSAM,packages}
+A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at \url{https://github.com/eliocamp/asymsam}.
 
@@
 
+\newpage
+
+\bibliographystyle{spbasic}
+\bibliography{AsymSAM,packages}
+
 \end{document}

### 2020-11-30 analysis/paper/paper.Rmd
Correcciones superficiales carolina

@@
 
-knitr::opts_hooks$set(fig.cap = function(options) {
-  options$cache <- FALSE
-  options
-})
+# knitr::opts_hooks$set(fig.cap = function(options) {
+#   options$cache <- FALSE
+#   options
+# })
 
@@
 library(tagger)
+options(tagger.cache = TRUE)
 library(patchwork)
@@
 We then compute the "Full SAM", "Asymmetric SAM" and "Symmetric SAM" indices as the regression coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude).
-The three indices are then normalized by dividing them by the standard deviation of the "Full SAM" index at each level.
+The three indices are then normalized by dividing them by the standard deviation of the Full SAM index at each level.
 As a result, the magnitudes between indices are comparable.
@@
 We perform linear regressions to quantify the association between the SAM indices and other variables.
-Since the Asymmetric and Symmetric SAM indices are significantly correlated with each other, to capture the variability explained uniquely by each index we use one multiple linear regression instead of two simple linear regressions.
+Moreover, we apply multiple linear regression analysis to describe the combined influence of both Asymmetric and Symmetric SAM indices.
 To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Asymmetric SAM ($SAM_a$) and Symmetric SAM ($SAM_s$) we fit the equation
@@
 In such case, using p-values to test for significance leads to misleading results [@walker1914;@katz1991].
-While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjusting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue.
-Therefore, p-values showed in regression fields are all adjusted following @benjamini1995.
+Therefore, p-values showed in regression fields are all adjusted  for the False Discovery Rate following @benjamini1995, which @wilks2016 suggests as a simple and effective method to ameliorate this issue.
 
 We computed linear trends by Ordinary Least Squares and the 95% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
-To compute the amplitude of the zonal waves we computed the Fourier transform of the spatial field at each latitude circle.
+To the amplitude of the zonal waves is defined through computing  the Fourier transform of the spatial field at each latitude circle.
 
@@
 All graphics are made using ggplot2 [@wickham2009].
-We downloaded data from ranalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
+We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
 The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].
@@
 We selected these two levels as representative of the tropospheric and stratospheric variability respectively.
-As will be shown later, both indices are highly coherent within each atmospheric layer, therefore is reasonable to take one level as representative of each layer.
+As it is shown below, the variabilities of both indices are highly coherent within each atmospheric layer, therefore is reasonable to take one level as representative of each layer.
 
@@
 The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \@ref(fig:A2)).
-This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b].
-There is a local peak at 2 years in the periodogram of the tropospheric Symmetric SAM also, although it's not statistically significant.
+This might suggest a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b].
+A local peak at 2 years  years is discernible in the periodogram of the tropospheric Symmetric SAM, although it's not statistically significant.
 In the troposphere the most significant peak of variability is found in the Asymmetric index at around `r 12*.3` months.
@@
 
-From Figure \@ref(fig:asymsam-timeseries) we can see that the Asymmetric SAM and Symmetric SAM time series appear to be correlated.
+From a visual inspection, the Asymmetric SAM and Symmetric SAM time series appear to be correlated.
 Moreover, looking at the extremes in the stratosphere, the Symmetric SAM series appears to lag the Asymmetric SAM series (see, for example, the positive events on late 1987).
-We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric SAM index leading the Symmetric SAM index), in Figure \@ref(fig:cor-lev).
-Zero-lag correlations between the Asymmetric SAM and Symmetric SAM series are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
+Figure \@ref(fig:cor-lev) shows these correlations along all levels considered, for zero and -1 lags.
+Values of zero-lag correlations between the Asymmetric SAM and Symmetric SAM series are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
 One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`.
@@
 At the same time, one-month-lag correlations increase with height.
-As a consequence, stratospheric Asymmetric SAM index tend to precede corresponding Symmetric SAM index.
+Therefore, stratospheric Asymmetric SAM index tends to precede the Symmetric SAM index.
 
@@
 For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
-Above 100 hPa correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
+Above 100 hPa, correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
 Therefore, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
-Examining panels b and c, we see that the Asymmetric and Symmetric SAM  share the same high level of coherency in the troposphere but they differ in their stratospheric behaviour.
+Asymmetric and Symmetric SAM (panels b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
 Stratospheric coherency is stronger for the Asymmetric SAM than the Symmetric SAM.
@@
 ```{r trends, fig.cap = "Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
-Shading indicates the 95\\% confidence interval from a t-distribution.", fig.width = 6, fig.height=4, fig.env="figure*"}
+Shading indicates the 95\\% confidence interval from a t-distribution.", fig.width = 6, fig.height=4, fig.env="figure*", cache = FALSE}
 indices %>% 
@@
 
-The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979 -- 2018 at each level (Figure \@ref(fig:trends)) for the whole year and separated by trimesters.
-The Full SAM index presents a statistically significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
+The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \@ref(fig:trends)).
+The Full SAM index presents a statistically positive significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
 The seasonal trends (rest of column a) indicate that positive trends are present in  Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
 In Winter and Spring, we detect no statistically significant trend.
-This is consistent with the results of previous studies, which find clear positive trends in Summer, weaker in Autumn and no trends in the other seasons [e.g.
+This is consistent with the results of previous studies, which find large positive trends in Summer, smaller in Autumn and no trends in the other seasons [e.g.
 @fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.
@@
 By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.
-column c), but in some cases the trends become more clear.
-In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
+column c), but in some cases the trends become clearer.
+In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component have a higher signal-to-noise ratio (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
 In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full SAM index.
@@
 
-To show if, and to what extent, the Asymmetric and Symmetric SAM indices indeed capture the asymmetric and symmetric component of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa levels.
-Figure \@ref(fig:2d-regr) shows these regressions.
-Regression coefficients in column a are computed using the Full SAM.
-Regression coefficients in columns b and c are computed using multiple regression using the Asymmetric and Symmetric indices at the same time.
+We then computed the spatial regression of geopotential height anomalies on Full, Asymmetric and Symmetric indices at 700 hPa and 50 hPa levels (Figure \@ref(fig:2d-regr)).
+While, regression coefficients in column a are computed using the Full SAM, the regression coefficients in columns b and c of Figure \@ref(fig:2d-regr) are computed using multiple regression using the Asymmetric and Symmetric indices at the same time.
 Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other index.
 
-In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centred in the South Pole.
-The monopole obtained by the regression pattern for Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davis Sea in the Eastern Hemisphere (panel b.1).
-
-In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 [@fogt2012].
-The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
-The Asymmetric SAM index gives rise to a cleaner zonal wave (panel b.2) and the Symmetric SAM index is associated with an annular mode, almost devoid of zonal asymmetries (panel c.2).
-The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern associated with @raphael2004's ZW3 index, whose reference locations are marked with points in the figure.
+In the stratosphere, the spatial pattern associated with the Full SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \@ref(fig:2d-regr)a.1) which is, not centred in the South Pole.
+On the other hand, the monopole associated with the Symmetric SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric.
+Furthermore, the regression pattern of the Asymmetric SAM is characterized by a wave-1 structure with centers over the Bellinghausen Sea on the
+Western Hemisphere and Davis Sea in the Eastern Hemisphere
+
+In the troposphere, the regression pattern associated with Full SAM shows the well-known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 (Figure \@ref(fig:2d-regr)a.2, [@fogt2012]).
+The regression patterns associated with the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
+While, the Asymmetric SAM index gives rise to a cleaner zonal wave (Figure \@ref(fig:2d-regr)b.2), the Symmetric SAM index is associated with an annular structure, almost devoid of zonal asymmetries (Figure \@ref(fig:2d-regr)c.2).
+The wave-3 pattern observed in Figure \@ref(fig:2d-regr)b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern described by Raphael (2004), whose reference locations are marked with points in the figure.
 Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
@@
 
-```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\@ref(fig:2d-regr) for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.", fig.width=6, fig.height=3, fig.env = "figure*"}
+```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\@ref(fig:2d-regr) for zonal waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the amplitude of the zonal mean.", fig.width=6, fig.height=3, fig.env = "figure*", cache = FALSE}
 indices_2d_yearly %>% 
@@
   geom_line(aes(color = factor(k))) +
-  scale_color_brewer("Wave-number", palette = "Dark2") +
+  scale_color_brewer("Wavenumber", palette = "Dark2") +
   scale_y_continuous("Amplitude") +
@@
 The amplitude of the first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \@ref(fig:wave-amplitude), where wave number zero represents the amplitude of the zonal mean.
-Zonal wave amplitudes of the spatial pattern described by the Full SAM index (column `r get_col(term == "full")`) are dominated by the zonal mean (wave-number 0) at both levels.
-However, zonal waves are important, particularly North of 50\degree S, with wave-number 1 clearly dominating at 50 hPa (panel `r get_tag(term == "full" & lev == 50)`) and a more equal mix of waves at 700 hPa (panel `r get_tag(term == "full" & lev == 700)`).
-Column `r get_col(term == "asym")` shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel `r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel `r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean.
-The Symmetric SAM, on the other hand, it's almost entirely composed of zonal mean at both levels (column `r get_col(term == "sym")`), with little to now contribution from zonal waves with wave-numbers 1 to 3.
+Zonal wave amplitudes of the spatial pattern described by the Full SAM index (Figure \@ref(fig:wave-amplitude) column `r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at both levels.
+However, zonal waves are important, particularly North of 50\degree S, with wavenumber 1 clearly dominating at 50 hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of waves of similar amplitude at 700 hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`).
+Figure \@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean.
+On the other hand, the Symmetric SAM it is almost entirely explained by the zonal mean at both levels (Figure \@ref(fig:wave-amplitude) column c), with little to no contribution from the zonal waves with wavenumbers 1 to 3.
 
 We can see that the amplitude and latitudinal distribution zonal waves in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric SAM on the other correspond almost exactly to the amplitude and latitudinal distribution in the Full SAM.
-This confirms the correct decomposition of the SAM in its symmetric and asymmetric components.
 Looking at panel b.2 from Figure \@ref(fig:2d-regr), it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
@@
 The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes.
-Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
-
-The tropospheric Asymmetric SAM (panel b) has significant signals that extend upwards to the uppermost levels of the reanalysis.
+The signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
+
+The tropospheric Asymmetric SAM (Figure \@ref(fig:vertical-regression) panel b)  has significant signals that extend upwards to the uppermost levels considered.
 In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
-The anomalies are much more intense in the Western hemisphere, where they extent into the stratosphere.
-In the Eastern hemisphere the wave-3 signal is weaker and confined to the troposphere while negative anomalies dominate in the stratosphere.
+The anomalies are larger in the Western hemisphere, where they extent into the stratosphere.
+In the Eastern hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere.
 So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
@@
 
-The wave-3 pattern from Figure \@ref(fig:2d-regr) panel b.2 is very similar to the Pacific-South American Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
+The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the Pacific-South American Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
 The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value = `r cor_texts[term == "full"]$p.value`).
-Consistent with @fan2007, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value = `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value = `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
-We performed the same analysis using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987] to conclude that these results do not depend on the ENSO index used.
-
-## Surface Impacts {#impacts}
+This relationship is captured mainly by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value = `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI, consistent with @fan2007's results, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value = `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
+The same analysis was performed using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987], obtaining similar results.
+The latter allows to conclude that these results do not depend on the ENSO index used.
+
+## Impacts {#impacts}
 
@@
 
-```{r regr-air-season, fig.cap = "Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
+```{r regr-air-season, fig.cap = "Regression of seasonal mean 2-meter air temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Gray areas in Antarctica are areas with have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*"}
+Gray areas in Antarctica are areas with have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*", cache = FALSE}
 air_regr %>% 
@@
                    minor_breaks = NULL) +
-  scale_fill_divergent("Surface temperature",  
+  scale_fill_divergent("2-meter air tempreature",  
                        limits = c(-0.6, .6), 
@@
 
-To see if there are different surface impacts associated with the asymmetric and symmetric SAM circulation we regress surface temperature and precipitation onto each of the three SAM indices at 700 hPa.
+To assess the differences in the impacts associated with full, asymmetric and symmetric SAM, we regress 2-meter air temperature and precipitation onto each of the three SAM indices at 700 hPa.
 As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
 
-Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface temperature for each trimester.
+Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface air temperature for each trimester.
 In Summer, positive values of the Full SAM index (panel `r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
-The ring is not zonally symmetric, as there are four clear local maximums around 30\degree W,  120\degree W, 150\degree E and 90\degree E.
+The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W,  120\degree W, 150\degree E and 90\degree E respectively.
 In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
@@
 Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the Asymmetric SAM regression map, while temperature patterns linked to positive Symmetric SAM show a more zonally consistent ring and less relation to the tropics.
-Noticeable, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to Asymmetric SAM.
+Temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to Asymmetric SAM.
 This signal is not present in the regression pattern with the Full SAM.
@@
 These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following @wilks2016.
-However, repeating this analysis with 2-meter temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
+However, repeating this analysis with 2-meter air temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
 Moreover, similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
@@
 
-In Australia, the annual regression shows that the Full SAM index is positively associated with precipitation in the Southeastern region (Figure \@ref(fig:pp-regr-oceania) panel a.1), which reproduces the results from @gillett2006.
-The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (panel c.1).
+In Australia, the annual regression shows that the Full SAM index is associated with positive precipitation anomalies in the southeastern region (Figure \@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006.
+The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (Figure \@ref(fig:pp-regr-oceania)c.1).
 Geopotential anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by @hendon2007.
-The Asymmetric SAM appears related to increased precipitation in the West coast of Southeastern Australia (panel b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
-
-The seasonal-level regressions show statistically significant anomalies only in Spring, when positive Full SAM is associated with positive precipitation anomalies in Eastern Australia (panel a.5).
-In this trimester the Symmetric SAM seems to be associated with precipitation in a relatively reduced area of the East Coast (panel c.5) while the positive precipitation anomalies related with positive Asymmetric SAM affect all Eastern Australia (panel b.5).
-
-In Summer, positive Full SAM index is associated with with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (panel a.2).
+The Asymmetric SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
+
+The seasonal-based regressions show statistically significant anomalies only in Spring, when positive Full SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
+In this trimester, the Symmetric SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related with positive Asymmetric SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).
+
+In Summer, positive Full SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \@ref(fig:pp-regr-oceania)a.2).
 The Eastern part being dominated by the relationship with the Symmetric SAM and the Western, by the Asymmetric SAM.
-In Autumn, the regression with Full SAM shows positive values in the North, similar to Summer, and a broad area of positive values in the North-East to South-West direction.
+In Autumn, the regression with Full SAM shows positive anomalies in the North, similar to Summer, and a broad area of positive anomalies in the North-East to South-West direction.
 This structure seems to be associated with the Symmetric SAM, while the Northern positive values are associated with the Asymmetric SAM.
@@
 None of these regression coefficients are statistically significant at the 95% level.
-The Spring signal is broadly consistent with @hendon2007, but whereas @hendon2007 also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coefficients have the consistent sign).
+The Spring signal is broadly consistent with @hendon2007, but whereas they also detected a strong signal in Summer, Figure \@ref(fig:pp-regr-oceania)a.2 shows no statistically significant association (although the coefficients have the consistent sign).
 
@@
 
-In South America (Figure \@ref(fig:pp-regr-america)), the annual-level regression shows that positive SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1).
-Panels b.1 and c.1 show a remarkably clean separation between the Asymmetric SAM -- associated with the Southeastern South American and Southern Brazilian signals -- and the Symmetric SAM -- associated with the signal in Southern Chile.
-
-Except Winter, seasonal-level regressions mirror this same pattern.
-Even if not statistically significant, they all show negative values in Southeastern South America and Southern Chile along with positive values in Southern Brazil in relation with the Full SAM.
+In South America (Figure \@ref(fig:pp-regr-america)), the annual-based regression shows that positive SAM is associated with statistically significant  negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
+Figure \@ref(fig:pp-regr-america)b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with the Asymmetric SAM, that in southern Chile is related to the Symmetric SAM.
+
+Except Winter, seasonal-based regressions mirror this same pattern.
+Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the Full SAM.
 The separation of these features between the Asymmetric SAM and Symmetric SAM regression maps is also rather consistent.
 
-The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous Easterly flow over Southern Chile.
-This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region [@garreaud2007].
+The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous easterly flow over Southern Chile.
+This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region [e.g.
+@garreaud2007].
 On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
-This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region [@silvestri2009;@zamboni2010].
-This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events [@rosso2018].
-There is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 and that is explained by the Asymmetric SAM.
+This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region [@silvestri2009;@zamboni2010].
+This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events [@rosso2018].
+There is a small area of positive statistically significant  precipitation anomalies with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the Asymmetric SAM.
 
@@
 
-As expected, the Asymmetric SAM index correlates strongly with the Symmetric SAM index.
+The Asymmetric SAM index correlates strongly with the Symmetric SAM index.
 In the troposphere, this correlation is maximum at zero lag, while in the stratosphere is maximised with the Asymmetric SAM leading the Symmetric SAM by one month.
@@
 There is evidence of influence between the QBO and the Northern Annular Mode [e.g.
-@holton1980;@watson2014;@zhang2020], so it's not unlikely that the SAM would be similarly connected.
-However establishing this link would require further research.
-
-We observe a positive trend towards positive SAM in Summer and Autumn, As was documented by previous studies, such as @fogt2020 (and references therein) for surface levels.
-We show that these trends are maximised at the 100 hPa level and are explained by the zonally symmetric component.
-We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not apparent in the Full SAM index.
+@holton1980;@watson2014;@zhang2020], so it is not unlikely that the SAM would be similarly connected.
+However, establishing this link would require further research.
+We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies [e.g.
+@fogt2020; and references therein] for low levels.
+We show that these trends maximise at 100 hPa, and are explained by the zonally symmetric component.
+We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the Full SAM index.
 In contrast to @fogt2012 we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
-
-In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being truly annular than the patterns associated with the Full SAM index.
-The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific region and whose phase is rotated a quarter wavelength from the mean zonal wave 3 described by @raphael2004's index.
-This pattern extends in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimum for capturing this variability.
-
-This wave-3 pattern is similar to the Pacific-South American Pattern, which is a teleconnection pattern linked to ENSO variability.
-We found that the significant correlation that exists between the Full SAM index and the Oceanic Niño Index is captured entirely by the Asymmetric SAM index.
-This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the Asymmetric SAM index could be a useful measure to further study that relationship.
+This might be due either due to differneces in methodology or analysed period.
+
+In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being fully annular than the patterns associated with the Full SAM index.
+The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific This pattern extends in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimum for capturing this variability.
+This wave-3 pattern is similar to the Pacific-South American Pattern, linked to ENSO variability.
+We found that the significant correlation that exists between the Full SAM index and the Oceanic Niño Index over this period, is captured entirely by the Asymmetric SAM index.
+This suggests that ENSO is linked to SAM exclusively through the variability in the latter’s asymmetric component and thus, the Asymmetric SAM index could be a useful measure to further study that relationship.
 
@@
 In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship.
-These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [@cazes-boezio2003;@fogt2011;@cai2020a].
+These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [e.g.
+@cai2020a].
 
@@
 In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
-This features are cleanly separated between the Asymmetric and Symmetric components.
+These features are cleanly separated between the Asymmetric and Symmetric components.
 The Symmetric SAM explains the negative anomalies in Southern Chile and the Asymmetric SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
@@
 
-@silvestri2009 suggests that precipitation impacts linked to the SAM changed rather dramatically before and after 1980.
+@silvestri2009 suggests that precipitation impacts linked to the SAM changed strongly before and after 1980.
 In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in others in the earlier period.
@@
 Seeing as both the ENSO-SAM relationship and most of the precipitation impacts in South America are captured by the Asymmetric SAM, the results presented here are most likely period-dependent.
-
-By succesfully separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
-This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component [@arblaster2006;@simpkins2012].
+The decadal variations of the Asymmetric SAM should be focus of future studies.
+
+By separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
+This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component [@arblaster2006].
 

### 2020-11-30 analysis/paper/paper.tex
Correcciones superficiales carolina

@@
 We then compute the ``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM'' indices as the regression coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude).
-The three indices are then normalized by dividing them by the standard deviation of the ``Full SAM'' index at each level.
+The three indices are then normalized by dividing them by the standard deviation of the Full SAM index at each level.
 As a result, the magnitudes between indices are comparable.
@@
 We perform linear regressions to quantify the association between the SAM indices and other variables.
-Since the Asymmetric and Symmetric SAM indices are significantly correlated with each other, to capture the variability explained uniquely by each index we use one multiple linear regression instead of two simple linear regressions.
+Moreover, we apply multiple linear regression analysis to describe the combined influence of both Asymmetric and Symmetric SAM indices.
 To obtain the linear coefficients of a variable \(X\) (geopotential, temperature, precipitation, etc\ldots{}) with the Asymmetric SAM (\(SAM_a\)) and Symmetric SAM (\(SAM_s\)) we fit the equation
@@
 In such case, using p-values to test for significance leads to misleading results \citep{walker1914, katz1991}.
-While there are multiple proposed solutions in the literature, \citet{wilks2016} suggests that adjusting p-values by controlling for the False Discovery Rate \citep{benjamini1995} is a simple and effective method to ameliorate this issue.
-Therefore, p-values showed in regression fields are all adjusted following \citet{benjamini1995}.
+Therefore, p-values showed in regression fields are all adjusted for the False Discovery Rate following \citet{benjamini1995}, which \citet{wilks2016} suggests as a simple and effective method to ameliorate this issue.
 
 We computed linear trends by Ordinary Least Squares and the 95\% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
-To compute the amplitude of the zonal waves we computed the Fourier transform of the spatial field at each latitude circle.
+To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
 
@@
 All graphics are made using ggplot2 \citep{wickham2009}.
-We downloaded data from ranalysis using the ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi package \citep{albers2020}.
+We downloaded data from reanalysis using the ecmwfr package \citep{hufkens2020} and indices of the ENSO with the rsoi package \citep{albers2020}.
 The paper was rendered using knitr and rmarkdown \citep{xie2015, allaire2019}.
@@
 The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \ref{fig:A2}).
-This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation \citep{baldwin2001b}.
-There is a local peak at 2 years in the periodogram of the tropospheric Symmetric SAM also, although it's not statistically significant.
+This might suggest a link between stratospheric SAM variability and the Quasi-Biennial Oscillation \citep{baldwin2001b}.
+A local peak at 2 years years is discernible in the periodogram of the tropospheric Symmetric SAM, although it's not statistically significant.
 In the troposphere the most significant peak of variability is found in the Asymmetric index at around 3.6 months.
 
-From Figure \ref{fig:asymsam-timeseries} we can see that the Asymmetric SAM and Symmetric SAM time series appear to be correlated.
+From a visual inspection, the Asymmetric SAM and Symmetric SAM time series appear to be correlated.
 Moreover, looking at the extremes in the stratosphere, the Symmetric SAM series appears to lag the Asymmetric SAM series (see, for example, the positive events on late 1987).
-We show these correlations, across all the levels of the reanalysis for zero and -1 lag (Asymmetric SAM index leading the Symmetric SAM index), in Figure \ref{fig:cor-lev}.
-Zero-lag correlations between the Asymmetric SAM and Symmetric SAM series are relatively constant throughout the troposphere, fluctuating between 0.39 and 0.45.
+Figure \ref{fig:cor-lev} shows these correlations along all levels considered, for zero and -1 lags.
+Values of zero-lag correlations between the Asymmetric SAM and Symmetric SAM series are relatively constant throughout the troposphere, fluctuating between 0.39 and 0.45.
 One-month-lag correlations are similarly constant but significantly reduced to around 0.17.
@@
 For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
-Above 100 hPa correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
+Above 100 hPa, correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
 Therefore, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
-Examining panels b and c, we see that the Asymmetric and Symmetric SAM share the same high level of coherency in the troposphere but they differ in their stratospheric behaviour.
+Asymmetric and Symmetric SAM (panels b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
 Stratospheric coherency is stronger for the Asymmetric SAM than the Symmetric SAM.
@@
 
-The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979 -- 2018 at each level (Figure \ref{fig:trends}) for the whole year and separated by trimesters.
-The Full SAM index presents a statistically significant trend (panel a.1) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
+The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \ref{fig:trends}).
+The Full SAM index presents a statistically positive significant trend (panel a.1) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
 The seasonal trends (rest of column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
 In Winter and Spring, we detect no statistically significant trend.
-This is consistent with the results of previous studies, which find clear positive trends in Summer, weaker in Autumn and no trends in the other seasons \citep[e.g.][ and references therein]{fogt2020} using indices of the SAM based on surface or near-surface circulation.
+This is consistent with the results of previous studies, which find large positive trends in Summer, smaller in Autumn and no trends in the other seasons \citep[e.g.][ and references therein]{fogt2020} using indices of the SAM based on surface or near-surface circulation.
 
-By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.~column c), but in some cases the trends become more clear.
-In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component are more clear (compare the shading region in panel a.2 and c.2).
+By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.~column c), but in some cases the trends become clearer.
+In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component have a higher signal-to-noise ratio (compare the shading region in panel a.2 and c.2).
 In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full SAM index.
@@
 
-To show if, and to what extent, the Asymmetric and Symmetric SAM indices indeed capture the asymmetric and symmetric component of the SAM respectively, we computed the spatial regression of geopotential height anomalies on these indices and the Full SAM index for 700 hPa and 50 hPa levels.
-Figure \ref{fig:2d-regr} shows these regressions.
-Regression coefficients in column a are computed using the Full SAM.
-Regression coefficients in columns b and c are computed using multiple regression using the Asymmetric and Symmetric indices at the same time.
+We then computed the spatial regression of geopotential height anomalies on Full, Asymmetric and Symmetric indices at 700 hPa and 50 hPa levels (Figure \ref{fig:2d-regr}).
+While, regression coefficients in column a are computed using the Full SAM, the regression coefficients in columns b and c of Figure \ref{fig:2d-regr} are computed using multiple regression using the Asymmetric and Symmetric indices at the same time.
 Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other index.
 
-In the stratosphere, the spatial pattern associated with the Full SAM is more clearly dominated by a zonally symmetric, monopolar structure (panel a.1) which is, however, not perfectly centred in the South Pole.
-The monopole obtained by the regression pattern for Symmetric SAM (panel c.1) is much more symmetric and the shift from total symmetry is captured by the regression pattern of the Asymmetric SAM as a wave-1 with maximum anomalies above the Belinghausen Sea on the Western Hemisphere and and Davis Sea in the Eastern Hemisphere (panel b.1).
+In the stratosphere, the spatial pattern associated with the Full SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \ref{fig:2d-regr}a.1) which is, not centred in the South Pole.
+On the other hand, the monopole associated with the Symmetric SAM (Figure \ref{fig:2d-regr}c.1) is more symmetric.
+Furthermore, the regression pattern of the Asymmetric SAM is characterized by a wave-1 structure with centers over the Bellinghausen Sea on the
+Western Hemisphere and Davis Sea in the Eastern Hemisphere
 
-In the troposphere, panel a.2 shows the well known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 \citep{fogt2012}.
-The regression using the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
-The Asymmetric SAM index gives rise to a cleaner zonal wave (panel b.2) and the Symmetric SAM index is associated with an annular mode, almost devoid of zonal asymmetries (panel c.2).
-The wave-3 pattern observed in panel b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern associated with \citet{raphael2004}'s ZW3 index, whose reference locations are marked with points in the figure.
+In the troposphere, the regression pattern associated with Full SAM shows the well-known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 (Figure \ref{fig:2d-regr}a.2, \citep{fogt2012}).
+The regression patterns associated with the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
+While, the Asymmetric SAM index gives rise to a cleaner zonal wave (Figure \ref{fig:2d-regr}b.2), the Symmetric SAM index is associated with an annular structure, almost devoid of zonal asymmetries (Figure \ref{fig:2d-regr}c.2).
+The wave-3 pattern observed in Figure \ref{fig:2d-regr}b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern described by Raphael (2004), whose reference locations are marked with points in the figure.
 Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
@@
 \begin{figure*}
-\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wave-number 0, 1, 2, and 3, where wave-number 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
+\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
 \end{figure*}
@@
 The amplitude of the first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
-Zonal wave amplitudes of the spatial pattern described by the Full SAM index (column a) are dominated by the zonal mean (wave-number 0) at both levels.
-However, zonal waves are important, particularly North of 50\degree S, with wave-number 1 clearly dominating at 50 hPa (panel a.1) and a more equal mix of waves at 700 hPa (panel a.2).
-Column b shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (panel b.1), while in the troposphere it is composed of zonal waves 3 to 1 in decreasing level of importance (panel b.2) with negligible amplitude of the zonal mean.
-The Symmetric SAM, on the other hand, it's almost entirely composed of zonal mean at both levels (column c), with little to now contribution from zonal waves with wave-numbers 1 to 3.
+Zonal wave amplitudes of the spatial pattern described by the Full SAM index (Figure \ref{fig:wave-amplitude} column a) are dominated by the zonal mean (wavenumber 0) at both levels.
+However, zonal waves are important, particularly North of 50\degree S, with wavenumber 1 clearly dominating at 50 hPa (Figure \ref{fig:wave-amplitude}a.1) and a mix of waves of similar amplitude at 700 hPa (Figure \ref{fig:wave-amplitude}a.2).
+Figure \ref{fig:wave-amplitude} column b shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \ref{fig:wave-amplitude}b.1), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \ref{fig:wave-amplitude}b.2) with negligible amplitude of the zonal mean.
+On the other hand, the Symmetric SAM it is almost entirely explained by the zonal mean at both levels (Figure \ref{fig:wave-amplitude} column c), with little to no contribution from the zonal waves with wavenumbers 1 to 3.
 
 We can see that the amplitude and latitudinal distribution zonal waves in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric SAM on the other correspond almost exactly to the amplitude and latitudinal distribution in the Full SAM.
-This confirms the correct decomposition of the SAM in its symmetric and asymmetric components.
 Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
@@
 The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes.
-Interestingly, the signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
+The signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
 
-The tropospheric Asymmetric SAM (panel b) has significant signals that extend upwards to the uppermost levels of the reanalysis.
+The tropospheric Asymmetric SAM (Figure \ref{fig:vertical-regression} panel b) has significant signals that extend upwards to the uppermost levels considered.
 In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
-The anomalies are much more intense in the Western hemisphere, where they extent into the stratosphere.
-In the Eastern hemisphere the wave-3 signal is weaker and confined to the troposphere while negative anomalies dominate in the stratosphere.
+The anomalies are larger in the Western hemisphere, where they extent into the stratosphere.
+In the Eastern hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere.
 So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
@@
 
-The wave-3 pattern from Figure \ref{fig:2d-regr} panel b.2 is very similar to the Pacific-South American Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}.
+The wave-3 pattern from Figure \ref{fig:2d-regr}b.2 is very similar to the Pacific-South American Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}.
 Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO.
 The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is -0.16 (p-value = \ensuremath{2.8\times 10^{-4}}).
-Consistent with \citet{fan2007}, we show that this relationship is captured entirely by the Asymmetric SAM, as this index has a partial correlation of -0.26 (p-value = \ensuremath{6.3\times 10^{-9}}) with the ONI controlling for the effect of the Symmetric SAM, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67).
-We performed the same analysis using the Multivariate ENSO Index \citep{wolter2011} and the Southern Oscillation Index \citep{ropelewski1987} to conclude that these results do not depend on the ENSO index used.
+This relationship is captured mainly by the Asymmetric SAM, as this index has a partial correlation of -0.26 (p-value = \ensuremath{6.3\times 10^{-9}}) with the ONI, consistent with \citet{fan2007}'s results, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67).
+The same analysis was performed using the Multivariate ENSO Index \citep{wolter2011} and the Southern Oscillation Index \citep{ropelewski1987}, obtaining similar results.
+The latter allows to conclude that these results do not depend on the ENSO index used.
 
 \hypertarget{impacts}{%
-\subsection{Surface Impacts}\label{impacts}}
+\subsection{Impacts}\label{impacts}}
 
 \begin{figure*}
-\includegraphics{regr-air-season-1} \caption{Regression of seasonal mean surface temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{regr-air-season-1} \caption{Regression of seasonal mean 2-meter air temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 
-To see if there are different surface impacts associated with the asymmetric and symmetric SAM circulation we regress surface temperature and precipitation onto each of the three SAM indices at 700 hPa.
+To assess the differences in the impacts associated with full, asymmetric and symmetric SAM, we regress 2-meter air temperature and precipitation onto each of the three SAM indices at 700 hPa.
 As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
 
-Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface temperature for each trimester.
+Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface air temperature for each trimester.
 In Summer, positive values of the Full SAM index (panel a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
-The ring is not zonally symmetric, as there are four clear local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E.
+The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E respectively.
 In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
@@
 These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following \citet{wilks2016}.
-However, repeating this analysis with 2-meter temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
+However, repeating this analysis with 2-meter air temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
 Moreover, similar features were observed in station measurements by \citet{jones2019}, although using data from 1957 to 2016.
@@
 
-In Australia, the annual regression shows that the Full SAM index is positively associated with precipitation in the Southeastern region (Figure \ref{fig:pp-regr-oceania} panel a.1), which reproduces the results from \citet{gillett2006}.
-The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (panel c.1).
+In Australia, the annual regression shows that the Full SAM index is associated with positive precipitation anomalies in the southeastern region (Figure \ref{fig:pp-regr-oceania}a.1), in agreement with \citet{gillett2006}.
+The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (Figure \ref{fig:pp-regr-oceania}c.1).
 Geopotential anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by \citet{hendon2007}.
-The Asymmetric SAM appears related to increased precipitation in the West coast of Southeastern Australia (panel b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
+The Asymmetric SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \ref{fig:pp-regr-oceania}b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
 
-The seasonal-level regressions show statistically significant anomalies only in Spring, when positive Full SAM is associated with positive precipitation anomalies in Eastern Australia (panel a.5).
-In this trimester the Symmetric SAM seems to be associated with precipitation in a relatively reduced area of the East Coast (panel c.5) while the positive precipitation anomalies related with positive Asymmetric SAM affect all Eastern Australia (panel b.5).
+The seasonal-based regressions show statistically significant anomalies only in Spring, when positive Full SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \ref{fig:pp-regr-oceania}a.5).
+In this trimester, the Symmetric SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \ref{fig:pp-regr-oceania}c.5) while the positive precipitation anomalies related with positive Asymmetric SAM affect all Eastern Australia (Figure \ref{fig:pp-regr-oceania}b.5).
 
-In Summer, positive Full SAM index is associated with with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (panel a.2).
+In Summer, positive Full SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \ref{fig:pp-regr-oceania}a.2).
 The Eastern part being dominated by the relationship with the Symmetric SAM and the Western, by the Asymmetric SAM.
-In Autumn, the regression with Full SAM shows positive values in the North, similar to Summer, and a broad area of positive values in the North-East to South-West direction.
+In Autumn, the regression with Full SAM shows positive anomalies in the North, similar to Summer, and a broad area of positive anomalies in the North-East to South-West direction.
 This structure seems to be associated with the Symmetric SAM, while the Northern positive values are associated with the Asymmetric SAM.
@@
 None of these regression coefficients are statistically significant at the 95\% level.
-The Spring signal is broadly consistent with \citet{hendon2007}, but whereas \citet{hendon2007} also detected a strong signal in Summer, panel a.2 shows no statistically significant association (although the coefficients have the consistent sign).
+The Spring signal is broadly consistent with \citet{hendon2007}, but whereas they also detected a strong signal in Summer, Figure \ref{fig:pp-regr-oceania}a.2 shows no statistically significant association (although the coefficients have the consistent sign).
 
@@
 
-In South America (Figure \ref{fig:pp-regr-america}), the annual-level regression shows that positive SAM is associated with statistically significant precipitation decrease in Southeastern South America (SESA) and Southern Chile and non-significant increase in South Brazil, near the South Atlantic Convergence Zone (SACZ) (panel a.1).
-Panels b.1 and c.1 show a remarkably clean separation between the Asymmetric SAM -- associated with the Southeastern South American and Southern Brazilian signals -- and the Symmetric SAM -- associated with the signal in Southern Chile.
+In South America (Figure \ref{fig:pp-regr-america}), the annual-based regression shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \ref{fig:pp-regr-america}a.1).
+Figure \ref{fig:pp-regr-america}b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with the Asymmetric SAM, that in southern Chile is related to the Symmetric SAM.
 
-Except Winter, seasonal-level regressions mirror this same pattern.
-Even if not statistically significant, they all show negative values in Southeastern South America and Southern Chile along with positive values in Southern Brazil in relation with the Full SAM.
+Except Winter, seasonal-based regressions mirror this same pattern.
+Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the Full SAM.
 The separation of these features between the Asymmetric SAM and Symmetric SAM regression maps is also rather consistent.
 
-The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous Easterly flow over Southern Chile.
-This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region \citep{garreaud2007}.
+The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous easterly flow over Southern Chile.
+This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region \citep[e.g.][]{garreaud2007}.
 On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
-This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region \citep{silvestri2009, zamboni2010}.
-This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events \citep{rosso2018}.
-There is a small area of increased precipitation with SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} and that is explained by the Asymmetric SAM.
+This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region \citep{silvestri2009, zamboni2010}.
+This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events \citep{rosso2018}.
+There is a small area of positive statistically significant precipitation anomalies with SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} that is explained by the Asymmetric SAM.
 
@@
 The two-year periodicity we found in the stratospheric Symmetric SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
-There is evidence of influence between the QBO and the Northern Annular Mode \citep[e.g.][]{holton1980, watson2014, zhang2020}, so it's not unlikely that the SAM would be similarly connected.
-However establishing this link would require further research.
+There is evidence of influence between the QBO and the Northern Annular Mode \citep[e.g.][]{holton1980, watson2014, zhang2020}, so it is not unlikely that the SAM would be similarly connected.
+However, establishing this link would require further research.
+We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies {[}e.g.
+\citet{fogt2020}; and references therein{]} for low levels.
+We show that these trends maximise at 100 hPa, and are explained by the zonally symmetric component.
+We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the Full SAM index.
+In contrast to \citet{fogt2012} we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
+This might be due either due to differneces in methodology or analysed period.
 
-We observe a positive trend towards positive SAM in Summer and Autumn, As was documented by previous studies, such as \citet{fogt2020} (and references therein) for surface levels.
-We show that these trends are maximised at the 100 hPa level and are explained by the zonally symmetric component.
-We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not apparent in the Full SAM index.
-In contrast to \citet{fogt2012} we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
-
-In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being truly annular than the patterns associated with the Full SAM index.
-The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific region and whose phase is rotated a quarter wavelength from the mean zonal wave 3 described by \citet{raphael2004}'s index.
-This pattern extends in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimum for capturing this variability.
-
-This wave-3 pattern is similar to the Pacific-South American Pattern, which is a teleconnection pattern linked to ENSO variability.
-We found that the significant correlation that exists between the Full SAM index and the Oceanic Niño Index is captured entirely by the Asymmetric SAM index.
+In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being fully annular than the patterns associated with the Full SAM index.
+The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific This pattern extends in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimum for capturing this variability.
+This wave-3 pattern is similar to the Pacific-South American Pattern, linked to ENSO variability.
+We found that the significant correlation that exists between the Full SAM index and the Oceanic Niño Index over this period, is captured entirely by the Asymmetric SAM index.
 This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the Asymmetric SAM index could be a useful measure to further study that relationship.
@@
 In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship.
-These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active \citep{cazes-boezio2003, fogt2011, cai2020a}.
+These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active \citep[e.g.][]{cai2020a}.
 
@@
 
-\citet{silvestri2009} suggests that precipitation impacts linked to the SAM changed rather dramatically before and after 1980.
+\citet{silvestri2009} suggests that precipitation impacts linked to the SAM changed strongly before and after 1980.
 In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in others in the earlier period.
@@
 Seeing as both the ENSO-SAM relationship and most of the precipitation impacts in South America are captured by the Asymmetric SAM, the results presented here are most likely period-dependent.
+The decadal variations of the Asymmetric SAM should be focus of future studies.
 
-By succesfully separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
-This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component \citep{arblaster2006, simpkins2012}.
+By separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
+This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component \citep{arblaster2006}.
 

### 2020-12-04 analysis/paper/paper.Rmd
Corrige los gráficos.

@@
 
-MSC:
-- MSC code 1
-- MSC code 2
+# MSC:
+# - MSC code 1
+# - MSC code 2
 
@@
       citation_package: "natbib"
+      # pandoc_args: ["--lua-filter=abstract-to-meta.lua"]
 header-includes: 
@@
 library(tagger)
-options(tagger.cache = TRUE)
 library(patchwork)
@@
 combine_lists <- function(...) {
-  # browser(expr = length(x) != length(y))
   X <- list(...)
@@
 @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
-To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the Full SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation) for the period from 1979 to 2018 at the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures \@ref(fig:A3) and \@ref(fig:A4)).
-In all seasons and both levels, SAM+ composites are similar to SAM- in structure but the sign switched.
+To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the Full SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700 hPa and 50 HPa levels (Figures \@ref(fig:A3) and \@ref(fig:A4)).
+In all seasons and both levels, SAM+ composites are similar to SAM- in structure but with the opposite sign.
 Pattern correlations (computed as the Pearson correlation between fields) between SAM+ and SAM- composites for each season are high.
@@
 
+N <- indices %>% 
+  .[, .N, by = .(lev, term)] %>% 
+  .[, unique(N)]
+
 crosscor <- indices %>% 
@@
   .[, widyr::pairwise_cor(.SD, lev, time, estimate, diag = TRUE), by = .(term)]
+
+crosscor %>% 
+  .[, p.value := 2*pnorm(atanh(correlation), 0, sd = 1/sqrt(N), lower.tail = FALSE)] %>% 
+  .[, p.value := p.adjust(p.value, method = "bonferroni")]
+
 crosscor %>% 
   ggplot(aes(item2, item1)) +
-  geom_contour_fill(aes(z = correlation), breaks = MakeBreaks(binwidth = 0.1)) +
-  geom_contour_tanaka(aes(z = correlation), size = 0.1, na.fill = .99, breaks = MakeBreaks(binwidth = 0.1)) +
+  geom_contour_fill(aes(z = correlation, fill = ..level..), breaks = MakeBreaks(binwidth = 0.1)) +
+  geom_contour_tanaka(aes(z = correlation), size = 0.1, na.fill = .99, 
+                      breaks = MakeBreaks(binwidth = 0.1)) +
   geom_text_contour(aes(z = correlation), breaks = MakeBreaks(binwidth = 0.1), 
@@
   geom_polygon(data = tri, fill = "white") +
-  scale_fill_divergent("Cross-correlation", 
-                       breaks = MakeBreaks(binwidth = 0.1),
+  scale_fill_divergent_discretised("Cross-correlation",
                                    guide = guide_colorstrip_bottom()) +
@@
 
-Figure \@ref(fig:cross-correlation)a shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
-For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
-Above 100 hPa, correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
-Therefore, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
-Asymmetric and Symmetric SAM (panels b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
+Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
+For the Full SAM (Figure \@ref(fig:cross-correlation)a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
+Above 100 hPa, correlation between levels does falls off more rapidly, indicating less coherent (zero-lag) variability.
+But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g.
+grater than 0.4 between tropospheric levels and levels below 30 hPa).
+Asymmetric and Symmetric SAM (Figure \@ref(fig:cross-correlation)b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
 Stratospheric coherency is stronger for the Asymmetric SAM than the Symmetric SAM.
@@
 The Full SAM index presents a statistically positive significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
-The seasonal trends (rest of column a) indicate that positive trends are present in  Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
+The seasonal trends (rest of Figure \@ref(fig:trends) column a) indicate that positive trends are present in  Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
 In Winter and Spring, we detect no statistically significant trend.
@@
 By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.
-column c), but in some cases the trends become clearer.
+column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer.
 In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component have a higher signal-to-noise ratio (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
@@
 
-We stress that these are only linear trends during the whole period and the absence of a statistically significant signal should not be taken as evidence of no systematic change.
-In particular, going back to Figure \@ref(fig:asymsam-timeseries), we can see an evident change in the stratospheric Asymmetric component (red line in panel a) between the 90's, when we see a dominance of extreme negative values, and the 00's, when we see the inverse.
-This change is restricted to the Winter months: the linear trend for Winter starting in 1990 for the Asymmetric component at 50hPa is $`r trends_text[term == "asym" & lev == 50 & season == "JJA"]$V1`$.
-
-Figure \@ref(fig:r-squared-trend) shows decadal trends for the explained variance of each index.
+Figure \@ref(fig:r-squared-trend) shows trends for the explained variance of each index.
 There is no evidence of a significant trend in the stratosphere.
@@
 
-```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\@ref(fig:2d-regr) for zonal waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the amplitude of the zonal mean.", fig.width=6, fig.height=3, fig.env = "figure*", cache = FALSE}
+```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\@ref(fig:2d-regr) for zonal waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the amplitude of the zonal mean.", fig.width=4, fig.height=3, fig.env = "figure*", cache = FALSE}
 indices_2d_yearly %>% 
@@
 
-We can see that the amplitude and latitudinal distribution zonal waves in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric SAM on the other correspond almost exactly to the amplitude and latitudinal distribution in the Full SAM.
-Looking at panel b.2 from Figure \@ref(fig:2d-regr), it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
+Looking at Figure \@ref(fig:2d-regr)b.2, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
 
@@
 ```{r vertical-regression, fig.cap = "Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\\degree and 40\\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
-(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979 -- 2018 period.", fig.width = 3, fig.height=3}
+(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979 -- 2018 period.", fig.width = 4, fig.height=3}
 vertical_regr %>% 
@@
   ggplot(aes(lon, lev)) +
-  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, 10, 0)) +
+  geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(0, 10, 0)) +
   geom_contour_tanaka(aes(z = estimate),
@@
   # geom_contour2(aes(z = p.val), breaks = 0.01) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(13), 
-                       oob = scales::squish,
-                       breaks = MakeBreaks(binwidth = 10, exclude = 0)) +
+  scale_fill_divergent_discretised(NULL, guide = guide_colorstrip_bottom(13)) +
+  # oob = scales::squish,
   scale_y_level() +
@@
 
-To analyse the vertical structure of the geopotential height anomalies associated with the asymmetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \@ref(fig:vertical-regression)).
-The geopotential height anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
+To analyse the vertical structure of the geopotential height anomalies associated with the asymmetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (Figure \@ref(fig:vertical-regression)a) and for the 700 hPa Asymmetric SAM index (Figure \@ref(fig:vertical-regression)b).
+The geopotential height anomalies associated with the stratospheric Asymmetric SAM (Figure \@ref(fig:vertical-regression)a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
 The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes.
@@
 
-The tropospheric Asymmetric SAM (Figure \@ref(fig:vertical-regression) panel b)  has significant signals that extend upwards to the uppermost levels considered.
+The tropospheric Asymmetric SAM (Figure \@ref(fig:vertical-regression)b)  has significant signals that extend upwards to the uppermost levels considered.
 In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
@@
 So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
-The structures shown in panels a and b in  Figure \@ref(fig:vertical-regression) are robust to the choice of index level.
-For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in panel a.
-Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b.
+The structures shown in Figure \@ref(fig:vertical-regression)a Figure \@ref(fig:vertical-regression)b in  Figure \@ref(fig:vertical-regression) are robust to the choice of index level.
+For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in Figure \@ref(fig:vertical-regression)a.
+Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to Figure \@ref(fig:vertical-regression)b.
 The patterns mainly change in amplitude (not shown).
@@
 enso <- rbindlist(list(oni = oni, soi = soi, mei = mei), idcol = "index")
+
+report_p <- function(p.value) {
+  ifelse(p.value < 0.001, " < 0.001", paste0(" = ", signif(p.value, 2)))
+}
 
@@
   .[, `:=`(cor = signif(cor, 2), 
-           p.value = signif(p.value, 2)
-  )]
+           p.value = report_p(p.value))]
 
@@
   .[, partial_correlation := signif(partial_correlation, 2)] %>% 
-  .[, p.value := signif(p.value, 2)]
+  .[, p.value := report_p(p.value)]
 
@@
   .[, partial_correlation := signif(partial_correlation, 2)] %>% 
-  .[, p.value := signif(p.value, 2)]
+  .[, p.value := report_p(p.value)]
 ```
@@
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value = `r cor_texts[term == "full"]$p.value`).
-This relationship is captured mainly by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value = `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI, consistent with @fan2007's results, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value = `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
+The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value `r cor_texts[term == "full"]$p.value`).
+This relationship is captured mainly by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI, consistent with @fan2007's results, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
 The same analysis was performed using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987], obtaining similar results.
@@
 Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface air temperature for each trimester.
-In Summer, positive values of the Full SAM index (panel `r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
+In Summer, positive values of the Full SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
 The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W,  120\degree W, 150\degree E and 90\degree E respectively.
 In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
-Panels `r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of the Asymmetric and Symmetric SAM, respectively.
+Figure \@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of the Asymmetric and Symmetric SAM, respectively.
 Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the Asymmetric SAM regression map, while temperature patterns linked to positive Symmetric SAM show a more zonally consistent ring and less relation to the tropics.
@@
 This signal is not present in the regression pattern with the Full SAM.
-Spring (row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with less regions where regressed anomalies have statistical significance.
-
-In Autumn and Winter (rows `r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))`) the positive ring is only present through its local maximums in the regression with the Full SAM.
+Spring (Figure \@ref(fig:regr-air-season), row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with less regions where regressed anomalies have statistical significance.
+
+In Autumn and Winter (rows `r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the Full SAM.
 There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
@@
 
-In South America (Figure \@ref(fig:pp-regr-america)), the annual-based regression shows that positive SAM is associated with statistically significant  negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
+In South America (Figure \@ref(fig:pp-regr-america)), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
 Figure \@ref(fig:pp-regr-america)b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with the Asymmetric SAM, that in southern Chile is related to the Symmetric SAM.
@@
 Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.
-For the 1979 -- 2018 period.", fig.width = 4, fig.height=4, fig.pos="ht"}
+For the 1979 -- 2018 period.", fig.width = 5, fig.height=5, fig.pos="ht"}
 ccf_dt <- function(..., plot = FALSE) {
@@
   geom_vline(xintercept = 0) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(15),
+  scale_fill_divergent(NULL,
+                       guide = guide_colorstrip_bottom(20),
                        breaks = AnchorBreaks(0, 0.05),
                        limits = c(NA, .59)) +
-  scale_x_continuous("Lag (months)\nSymmetric leads Asymmetric\t \tAsymmetric leads Symmetric", breaks = seq(-5, 5, 1), 
+  scale_x_continuous("Lag (months)\nSymmetric leads Asymmetric\t \tAsymmetric leads Symmetric", 
+                     breaks = seq(-5, 5, 1), 
                      expand = c(0, 0)) +
@@
 The light line indicates the theoretical expected amplitude from the autorregressive model.
-For the 1979 -- 2018 period.", fig.height=4, fig.width=7}
+For the 1979 -- 2018 period.", fig.height=4, fig.width=6}
 spec %>%
@@
 ```{r A3, fig.cap = "50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
-Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.", apendix = TRUE, fig.height = 5, fig.width = 7}
+Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.", apendix = TRUE, fig.height = 5, fig.width = 6}
 composites <- hgt[lev %in% c(50, 700)] %>%
@@
   tags <- composites[order(sign, season)][sign != 0, paste0("n: ", unique(n)), by = .(sign, season)]
+  if (this_lev == 50){
+    limits <- c(-119, 179) 
+  } else {
+    limits <- c(-59, 74)
+  }
 
@@
     scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
-                         # limits = c(-74, 74),
+                         limits = limits,
                          breaks = AnchorBreaks(0, binwidth = 15, exclude = 0)) +
@@
 ```{r A4, fig.cap = "700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
-Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.", apendix = TRUE, fig.height = 5, fig.width = 7}
+Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.", apendix = TRUE, fig.height = 4.5, fig.width = 6}
 plot_composites(700)
@@
 
-```{r A5, fig.cap = "Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.",  apendix = TRUE, fig.width=7, fig.height=6}
+```{r A5, fig.cap = "Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.",  apendix = TRUE, fig.width=6, fig.height=4.5}
 set.seed(42)

### 2020-12-04 analysis/paper/paper.tex
Corrige los gráficos.

@@
         El Niño Southern Oscillation \and
-    }
-
-    \subclass{
-                    MSC code 1 \and
-                    MSC code 2 \and
     }
@@
 \citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
-To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the Full SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation) for the period from 1979 to 2018 at the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures \ref{fig:A3} and \ref{fig:A4}).
-In all seasons and both levels, SAM+ composites are similar to SAM- in structure but the sign switched.
+To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the Full SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700 hPa and 50 HPa levels (Figures \ref{fig:A3} and \ref{fig:A4}).
+In all seasons and both levels, SAM+ composites are similar to SAM- in structure but with the opposite sign.
 Pattern correlations (computed as the Pearson correlation between fields) between SAM+ and SAM- composites for each season are high.
@@
 
-Figure \ref{fig:cross-correlation}a shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
-For the Full SAM (panel a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
-Above 100 hPa, correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
-Therefore, there is a non negligible correlation between the troposphere and the lower-to-middle stratosphere.
-Asymmetric and Symmetric SAM (panels b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
+Figure \ref{fig:cross-correlation} shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
+For the Full SAM (Figure \ref{fig:cross-correlation}a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
+Above 100 hPa, correlation between levels does falls off more rapidly, indicating less coherent (zero-lag) variability.
+But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g.~grater than 0.4 between tropospheric levels and levels below 30 hPa).
+Asymmetric and Symmetric SAM (Figure \ref{fig:cross-correlation}b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
 Stratospheric coherency is stronger for the Asymmetric SAM than the Symmetric SAM.
@@
 The Full SAM index presents a statistically positive significant trend (panel a.1) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
-The seasonal trends (rest of column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
+The seasonal trends (rest of Figure \ref{fig:trends} column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
 In Winter and Spring, we detect no statistically significant trend.
@@
 
-By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.~column c), but in some cases the trends become clearer.
+By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.~column c in Figure \ref{fig:trends}), but in some cases the trends become clearer.
 In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component have a higher signal-to-noise ratio (compare the shading region in panel a.2 and c.2).
@@
 
-We stress that these are only linear trends during the whole period and the absence of a statistically significant signal should not be taken as evidence of no systematic change.
-In particular, going back to Figure \ref{fig:asymsam-timeseries}, we can see an evident change in the stratospheric Asymmetric component (red line in panel a) between the 90's, when we see a dominance of extreme negative values, and the 00's, when we see the inverse.
-This change is restricted to the Winter months: the linear trend for Winter starting in 1990 for the Asymmetric component at 50hPa is \(0.37 \pm 0.22\).
-
-Figure \ref{fig:r-squared-trend} shows decadal trends for the explained variance of each index.
+Figure \ref{fig:r-squared-trend} shows trends for the explained variance of each index.
 There is no evidence of a significant trend in the stratosphere.
@@
 
-We can see that the amplitude and latitudinal distribution zonal waves in the Asymmetric SAM on one hand, and the zonal mean in the Symmetric SAM on the other correspond almost exactly to the amplitude and latitudinal distribution in the Full SAM.
-Looking at panel b.2 from Figure \ref{fig:2d-regr}, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
+Looking at Figure \ref{fig:2d-regr}b.2, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
 
@@
 
-To analyse the vertical structure of the geopotential height anomalies associated with the asymmetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (panel a) and for the 700 hPa Asymmetric SAM index (panel b) (Figure \ref{fig:vertical-regression}).
-The geopotential height anomalies associated with the stratospheric Asymmetric SAM (panel a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
+To analyse the vertical structure of the geopotential height anomalies associated with the asymmetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (Figure \ref{fig:vertical-regression}a) and for the 700 hPa Asymmetric SAM index (Figure \ref{fig:vertical-regression}b).
+The geopotential height anomalies associated with the stratospheric Asymmetric SAM (Figure \ref{fig:vertical-regression}a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
 The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes.
@@
 
-The tropospheric Asymmetric SAM (Figure \ref{fig:vertical-regression} panel b) has significant signals that extend upwards to the uppermost levels considered.
+The tropospheric Asymmetric SAM (Figure \ref{fig:vertical-regression}b) has significant signals that extend upwards to the uppermost levels considered.
 In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
@@
 So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
-The structures shown in panels a and b in Figure \ref{fig:vertical-regression} are robust to the choice of index level.
-For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in panel a.
-Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to panel b.
+The structures shown in Figure \ref{fig:vertical-regression}a Figure \ref{fig:vertical-regression}b in Figure \ref{fig:vertical-regression} are robust to the choice of index level.
+For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in Figure \ref{fig:vertical-regression}a.
+Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to Figure \ref{fig:vertical-regression}b.
 The patterns mainly change in amplitude (not shown).
@@
 Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is -0.16 (p-value = \ensuremath{2.8\times 10^{-4}}).
-This relationship is captured mainly by the Asymmetric SAM, as this index has a partial correlation of -0.26 (p-value = \ensuremath{6.3\times 10^{-9}}) with the ONI, consistent with \citet{fan2007}'s results, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67).
+The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is -0.16 (p-value \textless{} 0.001).
+This relationship is captured mainly by the Asymmetric SAM, as this index has a partial correlation of -0.26 (p-value \textless{} 0.001) with the ONI, consistent with \citet{fan2007}'s results, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67).
 The same analysis was performed using the Multivariate ENSO Index \citep{wolter2011} and the Southern Oscillation Index \citep{ropelewski1987}, obtaining similar results.
@@
 Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface air temperature for each trimester.
-In Summer, positive values of the Full SAM index (panel a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
+In Summer, positive values of the Full SAM index (Figure \ref{fig:regr-air-season}a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
 The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E respectively.
 In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
-Panels b.1 and c.1 show temperature anomalies associated with positive values of the Asymmetric and Symmetric SAM, respectively.
+Figure \ref{fig:regr-air-season}b.1 and c.1 show temperature anomalies associated with positive values of the Asymmetric and Symmetric SAM, respectively.
 Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the Asymmetric SAM regression map, while temperature patterns linked to positive Symmetric SAM show a more zonally consistent ring and less relation to the tropics.
@@
 This signal is not present in the regression pattern with the Full SAM.
-Spring (row 4) features similar patterns but of smaller magnitude, with less regions where regressed anomalies have statistical significance.
+Spring (Figure \ref{fig:regr-air-season}, row 4) features similar patterns but of smaller magnitude, with less regions where regressed anomalies have statistical significance.
 
-In Autumn and Winter (rows 2 and 3) the positive ring is only present through its local maximums in the regression with the Full SAM.
+In Autumn and Winter (rows 2 and 3 in Figure \ref{fig:regr-air-season}) the positive ring is only present through its local maximums in the regression with the Full SAM.
 There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
@@
 
-In South America (Figure \ref{fig:pp-regr-america}), the annual-based regression shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \ref{fig:pp-regr-america}a.1).
+In South America (Figure \ref{fig:pp-regr-america}), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \ref{fig:pp-regr-america}a.1).
 Figure \ref{fig:pp-regr-america}b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with the Asymmetric SAM, that in southern Chile is related to the Symmetric SAM.

### 2020-12-10 analysis/paper/paper.Rmd
Correcciones de leandro

@@
  - \usepackage{subfig}
- - \usepackage[inline]{showlabels}
+#  - \usepackage[inline]{showlabels}
  - \usepackage{chngcntr}
@@
 The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
-For example, the SAM-precipitation relationshipin Southeastern South America can be explained by the PSA-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
+For example, the SAM-precipitation relationshipin Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
 @fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
@@
 We first propose a methodology that provides for each level, two indices which aim to capture independently the variability of the symmetric and asymmetric SAM component respectively.
-We assess their vertical structure and coherence, temporal variability and trends.
+Their vertical structure and coherence, temporal variability and trends are consequently assesed.
 We then study the spatial patterns described by the variability exclusive to each index focusing on 50 hPa as representing the stratosphere and 700 hPa as representing the troposphere.
-Finally, we investigate the relationship of the SAM at 700 hPa with temperature and precipitation anomalies.
+Finally, the relationships of the SAM at 700 hPa with temperature and precipitation anomalies are investigated.
 
@@
 
-Temperature data used are from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude.
+Temperature data used are from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude.
 Monthly rainfall fields at 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015;@schneider2017] are also considered.
@@
 
-Traditionally the SAM is defined as de leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels [@ho2012].
+Traditionally the SAM is defined as the leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels [@ho2012].
 Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the leading EOF of the monthly anomalies of geopotential height  south of 20\degree S at each level.
@@
 However, only Full SAM index has unit standard deviation per definition.
-We also use the explained variance of each pattern as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
+The explained variance of each pattern is used as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
+To quantify the coherence between temporal series corresponding to different indices or the same index at different levels, we computed the temporal correlation between them.
 
@@
 In all seasons and both levels, SAM+ composites are similar to SAM- in structure but with the opposite sign.
-Pattern correlations (computed as the Pearson correlation between fields) between SAM+ and SAM- composites for each season are high.
+Spatial correlations between composites for each season are high.
 The method considered in this study seems then a reasonable approximation of the phenomenon.
@@
 The following seasons were considered – December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
-Results are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
+Results are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 1) and show spatial correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
 In the stratosphere (Figure \@ref(fig:A5), row 2), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree\ in comparison with the rest of the year.
-Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
+Spatial correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
 Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year with the exception of DJF in the stratosphere.
@@
 
-At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions.
-In such case, using p-values to test for significance leads to misleading results [@walker1914;@katz1991].
-Therefore, p-values showed in regression fields are all adjusted for the False Discovery Rate following @benjamini1995, which @wilks2016 suggests as a simple and effective method to ameliorate this issue.
-
-We computed linear trends by Ordinary Least Squares and the 95% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
+Statistical significance for regression fields were evaluated adjusting p-values by controlling for the False Discovery Rate [@benjamini1995;@wilks2016] to avoid misleading results from a high number of regressions [@walker1914;@katz1991].
+
+Linear trends were computed by Ordinary Least Squares and the 95% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
 To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
@@
 
-```{r trends, fig.cap = "Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+```{r trends, fig.cap = "Linear trends (in standard deviations per decade) at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
 Shading indicates the 95\\% confidence interval from a t-distribution.", fig.width = 6, fig.height=4, fig.env="figure*", cache = FALSE}
@@
 column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer.
-In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component have a higher signal-to-noise ratio (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
+In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
 In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full SAM index.
@@
 
-The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the Pacific-South American Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
+The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the PSA Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
@@
 
-```{r regr-air-season, fig.cap = "Regression of seasonal mean 2-meter air temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
+```{r regr-air-season, fig.cap = "Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 
-To assess the differences in the impacts associated with full, asymmetric and symmetric SAM, we regress 2-meter air temperature and precipitation onto each of the three SAM indices at 700 hPa.
+To assess the differences in the impacts associated with full, asymmetric and symmetric SAM, we regress s surface land air and sea temperature and precipitation onto each of the three SAM indices at 700 hPa.
 As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
 
-Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface air temperature for each trimester.
+Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface land air and sea temperature for each trimester.
 In Summer, positive values of the Full SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
@@
 
-```{r pp-regr-africa,  fig.cap = "Regression of precipiation anomalies with Asymmetric and Symmetric SAM for the 1979 -- 2018 period.
-P-values smaller than 0.05 (controlling for Flase Detection Rate) as hatched areas.
-Africa", fig.width = 6, fig.height = 4.5, fig.env = "figure*"}
-# plot_pp("africa") 
-```
-
-```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
+```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height anomalies (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
@@
 The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (Figure \@ref(fig:pp-regr-oceania)c.1).
-Geopotential anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by @hendon2007.
+Geopotential height anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by @hendon2007.
 The Asymmetric SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
@@
 This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events [@rosso2018].
-There is a small area of positive statistically significant  precipitation anomalies with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the Asymmetric SAM.
+There is a small area of significative positive precipitation anomalies with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the Asymmetric SAM.
 
@@
 We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the Full SAM index.
-In contrast to @fogt2012 we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
+In contrast to @fogt2012 who made the evaluation for the period 158 -- 2001, we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
 This might be due either due to differneces in methodology or analysed period.
@@
 
-## datastatement
-All data used in this paper is freely available in their respective sources.
+## Data statement
+
+All data used in this paper is freely available from their respective sources.
 ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means).

### 2020-12-10 analysis/paper/paper.tex
Correcciones de leandro

@@
 \usepackage{subfig}
-\usepackage[inline]{showlabels}
 \usepackage{chngcntr}
@@
 The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
-For example, the SAM-precipitation relationshipin Southeastern South America can be explained by the PSA-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}.
+For example, the SAM-precipitation relationshipin Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}.
 \citet{fan2007} also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
@@
 
-Temperature data used are from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land and ocean temperature analysis into a monthly global grid 5\degree~longitude by 5\degree~latitude.
+Temperature data used are from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid 5\degree~longitude by 5\degree~latitude.
 Monthly rainfall fields at 0.5\degree~longitude by 0.5\degree~latitude data from the Global Precipitation Climatology Centre \citep{schneider2015, schneider2017} are also considered.
@@
 
-Traditionally the SAM is defined as de leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels \citep{ho2012}.
+Traditionally the SAM is defined as the leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels \citep{ho2012}.
 Following \citet{baldwin2001}, we extend that definition vertically and use the term SAM to refer to the leading EOF of the monthly anomalies of geopotential height south of 20\degree S at each level.
@@
 The following seasons were considered -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
-Results are very similar to each other in the troposphere (Figure \ref{fig:A5}, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
+Results are very similar to each other in the troposphere (Figure \ref{fig:A5}, row 1) and show spatial correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
 In the stratosphere (Figure \ref{fig:A5}, row 2), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree~in comparison with the rest of the year.
-Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
+Spatial correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
 Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year with the exception of DJF in the stratosphere.
@@
 
-At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions.
-In such case, using p-values to test for significance leads to misleading results \citep{walker1914, katz1991}.
-Therefore, p-values showed in regression fields are all adjusted for the False Discovery Rate following \citet{benjamini1995}, which \citet{wilks2016} suggests as a simple and effective method to ameliorate this issue.
+Statistical significance for regression fields were evaluated adjusting p-values by controlling for the False Discovery Rate \citep{benjamini1995, wilks2016} to avoid misleading results from a high number of regressions \citep{walker1914, katz1991}.
 
-We computed linear trends by Ordinary Least Squares and the 95\% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
+Linear trends were computed by Ordinary Least Squares and the 95\% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
 To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
@@
 \begin{figure*}
-\includegraphics{trends-1} \caption{Decadal linear trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+\includegraphics{trends-1} \caption{Linear trends (in standard deviations per decade) at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
 Shading indicates the 95\% confidence interval from a t-distribution.}\label{fig:trends}
@@
 By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.~column c in Figure \ref{fig:trends}), but in some cases the trends become clearer.
-In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component have a higher signal-to-noise ratio (compare the shading region in panel a.2 and c.2).
+In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in panel a.2 and c.2).
 In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full SAM index.
@@
 
-The wave-3 pattern from Figure \ref{fig:2d-regr}b.2 is very similar to the Pacific-South American Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}.
+The wave-3 pattern from Figure \ref{fig:2d-regr}b.2 is very similar to the PSA Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}.
 Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO.
@@
 \begin{figure*}
-\includegraphics{regr-air-season-1} \caption{Regression of seasonal mean 2-meter air temperature (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{regr-air-season-1} \caption{Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 
-To assess the differences in the impacts associated with full, asymmetric and symmetric SAM, we regress 2-meter air temperature and precipitation onto each of the three SAM indices at 700 hPa.
+To assess the differences in the impacts associated with full, asymmetric and symmetric SAM, we regress s surface land air and sea temperature and precipitation onto each of the three SAM indices at 700 hPa.
 As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
 
-Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface air temperature for each trimester.
+Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface land air and sea temperature for each trimester.
 In Summer, positive values of the Full SAM index (Figure \ref{fig:regr-air-season}a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
@@
 \begin{figure*}
-\includegraphics{pp-regr-oceania-1} \caption{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{pp-regr-oceania-1} \caption{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height anomalies (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
@@
 The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (Figure \ref{fig:pp-regr-oceania}c.1).
-Geopotential anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by \citet{hendon2007}.
+Geopotential height anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by \citet{hendon2007}.
 The Asymmetric SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \ref{fig:pp-regr-oceania}b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
@@
 This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events \citep{rosso2018}.
-There is a small area of positive statistically significant precipitation anomalies with SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} that is explained by the Asymmetric SAM.
+There is a small area of significative positive precipitation anomalies with SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} that is explained by the Asymmetric SAM.
 
@@
 We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the Full SAM index.
-In contrast to \citet{fogt2012} we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
+In contrast to \citet{fogt2012} who made the evaluation for the period 158 -- 2001, we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
 This might be due either due to differneces in methodology or analysed period.
@@
 
-\hypertarget{datastatement}{%
-\subsection{datastatement}\label{datastatement}}
+\hypertarget{data-statement}{%
+\subsection{Data statement}\label{data-statement}}
 
-All data used in this paper is freely available in their respective sources.
+All data used in this paper is freely available from their respective sources.
 ERA5 data can be obtained via the Copernicus Climate Data Store (\url{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means}).

### 2020-12-11 analysis/paper/paper.Rmd
Abstract

@@
 abstract: |
-  The text of your abstract.  150 -- 250 words.
+  The main mode of variability in the Southern Hemisphere extratropical circulation is the Southern Annular Mode (SAM); so called because of it's zonally symmetric ring-like shape.
+However, the SAM pattern actually contains noticeable deviations from zonal symmetry.
+We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices.
+In the troposphere, the Asymmetric SAM, is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure.
+The Symmetric SAM, on the other hand, is associated with a negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
+The observed relationship between the El Niño Southern Oscillation and the SAM is explained fully by the Asymmetric SAM index.
+The positive trend of the SAM is present only in its asymmetric component.
+Despite this, the SAM is becoming more zonally asymmetric.
+The regional impacts of the SAM in temperature and precipitation are strongly affected by its asymmetric component.
 
@@
 <!-- The actual document text starts here: -->
+
+<!-- # Abstract -->
+
+<!-- The main mode of variability in the Southern Hemisphere extratropical circulation is the Southern Annular Mode (SAM); so called because of it's zonally symmetric ring-like shape.
+However, the SAM pattern actually contains noticeable deviations from zonal symmetry.
+We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices.
+In the troposphere, the Asymmetric SAM, is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure.
+The Symmetric SAM, on the other hand, is associated with a negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
+The observed relationship between the El Niño Southern Oscillation and the SAM is explained fully by the Asymmetric SAM index.
+The positive trend of the SAM is present only in its asymmetric component.
+Despite this, the SAM is becoming more zonally asymmetric.
+The regional impacts of the SAM in temperature and precipitation are strongly affected by its asymmetric nature.
+-->
 

### 2020-12-11 analysis/paper/paper.tex
Render abstract

@@
 \begin{abstract}
-The text of your abstract. 150 -- 250 words.
+The main mode of variability in the Southern Hemisphere extratropical circulation is the Southern Annular Mode (SAM); so called because of it's zonally symmetric ring-like shape.
+However, the SAM pattern actually contains noticeable deviations from zonal symmetry.
+We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices.
+In the troposphere, the Asymmetric SAM, is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure.
+The Symmetric SAM, on the other hand, is associated with a negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
+The observed relationship between the El Niño Southern Oscillation and the SAM is explained fully by the Asymmetric SAM index.
+The positive trend of the SAM is present only in its asymmetric component.
+Despite this, the SAM is becoming more zonally asymmetric.
+The regional impacts of the SAM in temperature and precipitation are strongly affected by its asymmetric component.
 \\

### 2020-12-11 analysis/paper/paper.Rmd
Arregla epígrafes

@@
 
-```{r wave-amplitude, fig.cap = "Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\@ref(fig:2d-regr) for zonal waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the amplitude of the zonal mean.", fig.width=4, fig.height=3, fig.env = "figure*", cache = FALSE}
+(ref:wave-amplitude-caption) Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \@ref(fig:2d-regr) for zonal waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the amplitude of the zonal mean.
+
+```{r wave-amplitude, fig.cap = "(ref:wave-amplitude-caption)", fig.width=4, fig.height=3, fig.env = "figure*", cache = FALSE}
 indices_2d_yearly %>% 
@@
 
-```{r pp-regr-america, dependson = "def-plot_pp", fig.cap = "Same as Figure \\@ref(fig:pp-regr-oceania) but for South America.", fig.width = 6, fig.height = 7.6, fig.env = "figure*"}
+(ref:pp-rer-america-cap) Same as Figure \@ref(fig:pp-regr-oceania) but for South America.
+
+```{r pp-regr-america, dependson = "def-plot_pp", fig.cap = "(ref:pp-rer-america-cap)", fig.width = 6, fig.height = 7.6, fig.env = "figure*"}
 plot_pp("america")
@@
 
-The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous easterly flow over Southern Chile.
+The anomalous circulation at 700 hPa associated with the Symmetric SAM (Figure \@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over Southern Chile.
 This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region [e.g.
 @garreaud2007].
-On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
+On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
 This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region [@silvestri2009;@zamboni2010].

### 2020-12-11 analysis/paper/paper.tex
Arregla epígrafes

@@
 \begin{figure*}
-\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \\ref{fig:2d-regr} for zonal waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
+\includegraphics{wave-amplitude-1} \caption{Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \ref{fig:2d-regr} for zonal waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the amplitude of the zonal mean.}\label{fig:wave-amplitude}
 \end{figure*}
@@
 \begin{figure*}
-\includegraphics{pp-regr-america-1} \caption{Same as Figure \\ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
+\includegraphics{pp-regr-america-1} \caption{Same as Figure \ref{fig:pp-regr-oceania} but for South America.}\label{fig:pp-regr-america}
 \end{figure*}
@@
 
-The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous easterly flow over Southern Chile.
+The anomalous circulation at 700 hPa associated with the Symmetric SAM (Figure \ref{fig:pp-regr-america}c.1) indicate anomalous easterly flow over Southern Chile.
 This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region \citep[e.g.][]{garreaud2007}.
-On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
+On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (Figure \ref{fig:pp-regr-america}b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
 This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region \citep{silvestri2009, zamboni2010}.

### 2020-12-11 analysis/paper/paper.Rmd
Última revision. Que el resto de los errores los encuentren les revisores.

@@
 ---
-title: Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
+title: Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach
 # subtitle: Do you have a subtitle? If so, write it here
-titlerunning:  Assessment of zonal symmetric and asymmetric components of the SAM
+titlerunning:  Assessment of zonally symmetric and asymmetric components of the SAM
 # authorrunning: Short form of author list if too long for running head
@@
 The Symmetric SAM, on the other hand, is associated with a negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
-The observed relationship between the El Niño Southern Oscillation and the SAM is explained fully by the Asymmetric SAM index.
+The observed relationship between the El Niño Southern Oscillation and the SAM is explained fully by the A\nobreakdash-SAM index.
 The positive trend of the SAM is present only in its asymmetric component.
@@
 
-theme_set(theme_asymsam(base_size = 9))
+theme_set(theme_asymsam(base_size = 9) + theme(tagger.panel.tag.text = element_text(size = 8)))
 ZeroBreaks <- AnchorBreaks(0, NULL, 0)
@@
 
-lab_sam <-  c(full = "Full", 
-              asym = "Asymmetric",
-              sym  = "Symmetric")
+report_p <- function(p.value) {
+  ifelse(p.value < 0.001, " < 0.001", paste0(" = ", signif(p.value, 2)))
+}
+
+lab_sam <-  c(full = "SAM", 
+              asym = "A-SAM",
+              sym  = "S-SAM")
 ```
@@
 Moreover, tropical influences on the SAM have been observed [@fan2007;@fogt2011; @clem2013].
-@fan2007 computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated to each other if the (linear) signal of the ENSO was removed.
-
-Positive trends in SAM index have been documented by various researchers using different indices mostly on austral Summer and Autumn [e.g.
+@fan2007 computed SAM indices of the Western and the Eastern Hemispheres separately and found that they were much more correlated to each other if the (linear) signal of the ENSO was removed.
+
+Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn [e.g.
 @fogt2020 and references therein].
-These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013].
-However, it’s not clear yet how or if the asymmetric SAM component responds to this forcing, or how its variability alters the observed trends.
+It is though that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013].
+However, it’s not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.
 
 The impact of the zonally asymmetric component of the SAM at regional scales has not been studied in detail yet.
-Positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM).
+The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM).
 But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012].
-SAM signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
+The SAM-related signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
 The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
@@
 The study of the temporal variability of the asymmetric component of the SAM has not received much attention except for @fogt2012.
-This study provides evidences of the relevance of SAM asymmetric component.
+This study provides evidences for the relevance of the SAM's asymmetric component.
 However, their conclusions  are based on composites of positive and negative SAM events including a small number of cases unevenly distributed among years with and without satellite information.
@@
 Their vertical structure and coherence, temporal variability and trends are consequently assesed.
-We then study the spatial patterns described by the variability exclusive to each index focusing on 50 hPa as representing the stratosphere and 700 hPa as representing the troposphere.
-Finally, the relationships of the SAM at 700 hPa with temperature and precipitation anomalies are investigated.
+We then study the spatial patterns described by the variability exclusive to each index focusing on 50\ hPa as representing the stratosphere and 700\ hPa as representing the troposphere.
+Finally, the relationships of the SAM at 700\ hPa with temperature and precipitation anomalies are investigated.
 
@@
 hgt <- ReadNetCDF(ERA5(), 
-                  subset = list(time = c("1979-01-01", "2018-12-31"),
+                  subset = list(time = c("1979-01-01", "2018-12-01"),
                                 latitude = c(-90, 10),
@@
 
-```{r compute-SAM}
+```{r compute\nobreakdash-SAM}
 SAM <- hgt[lat <= -20] %>% 
@@
   .[, .(x = uniqueN(lon), y = uniqueN(lat), t = uniqueN(time))]
+
 ```
@@
 
-Traditionally the SAM is defined as the leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels [@ho2012].
+Traditionally, the SAM is defined as the leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels [@ho2012].
 Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the leading EOF of the monthly anomalies of geopotential height  south of 20\degree S at each level.
@@
 We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999].
-We consider in the EOF analysis all months together without separating in seasons.
-
-To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \@ref(fig:method) at 700 hPa.
+We consider in the EOF analysis all months together without dividing by seasons.
+
+To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \@ref(fig:method) at 700\ hPa.
 The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
-We then compute the "Full SAM", "Asymmetric SAM" and "Symmetric SAM" indices as the regression coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude).
-The three indices are then normalized by dividing them by the standard deviation of the Full SAM index at each level.
+We then compute the SAM index, Asymmetric SAM index (A\nobreakdash-SAM) and Symmetric SAM (S\nobreakdash-SAM) indices as the coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude).
+The three indices are then normalized by dividing them by the standard deviation of the SAM index at each level.
 As a result, the magnitudes between indices are comparable.
-However, only Full SAM index has unit standard deviation per definition.
+However, only SAM index has unit standard deviation per definition.
 The explained variance of each pattern is used as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
@@
 
-```{r method, fig.cap = "Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
+```{r method, fig.cap = "Spatial patterns of the first EOF of 700~hPa geopotential height for 1979 -– 2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 The method assumes linearity in the asymmetric component of the SAM.
-That means that zonal symmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
+That means that zonal asymmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
 @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
-To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the Full SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700 hPa and 50 HPa levels (Figures \@ref(fig:A3) and \@ref(fig:A4)).
+To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700\ hPa and 50\ hPa levels (Figures \@ref(fig:A3) and \@ref(fig:A4)).
 In all seasons and both levels, SAM+ composites are similar to SAM- in structure but with the opposite sign.
@@
 
-By performing the EOF analysis using data for all months we are assuming that the zonal asymmetric structure of the SAM is the same at all seasons.
+By performing the EOF analysis using data for all months we are assuming that the zonally asymmetric structure of the SAM is the same at all seasons.
 The latter was assessed by computing geopotential height zonal anomalies by projecting the first EOF of each season independently.
@@
 
-The method also implies that the zonally asymmetric pattern of SAM remains stationary along the period considered.
+The method also assumes that the zonally asymmetric pattern of the SAM remains stationary along the period considered.
 @silvestri2009 suggest that this might not be the case between 1958 and 2004.
@@
 
-We perform linear regressions to quantify the association between the SAM indices and other variables.
-Moreover, we apply multiple linear regression analysis to describe the combined influence of both Asymmetric and Symmetric SAM indices.
-To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with the Asymmetric SAM ($SAM_a$) and Symmetric SAM ($SAM_s$) we fit the equation
+We performed linear regressions to quantify the association between the SAM indices and other variables.
+Moreover, we apply multiple linear regression analysis to describe the combined influence of both A\nobreakdash-SAM and S\nobreakdash-SAM.
+To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with A\nobreakdash-SAM and S\nobreakdash-SAM we fit the equation
 
 $$
-X(\lambda, \phi, t) = \alpha(\lambda, \phi) SAM_a + \beta(\lambda, \phi) SAM_s + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{A-SAM} + \beta(\lambda, \phi) \operatorname{S-SAM} + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
 $$
@@
 where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and $\epsilon$ are the constant and error terms.
-From this equation, $\alpha$ represents the (linear) association of $X$ with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; i.e.
-it is proportional to the partial correlation of $X$ and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and vice versa for $\beta$.
-When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we average the relevant variables seasonally for each year and trimester before computing the regression.
-
-Statistical significance for regression fields were evaluated adjusting p-values by controlling for the False Discovery Rate [@benjamini1995;@wilks2016] to avoid misleading results from a high number of regressions [@walker1914;@katz1991].
-
-Linear trends were computed by Ordinary Least Squares and the 95% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
+From this equation, $\alpha$ represents the (linear) association of $X$ with the variability of A\nobreakdash-SAM that is not explained by the variability of S\nobreakdash-SAM; i.e.
+it is proportional to the partial correlation of $X$ and A\nobreakdash-SAM, controlling for the effect of S\nobreakdash-SAM, and vice versa for $\beta$.
+When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we averaged the relevant variables seasonally for each year and trimester before computing the regression.
+
+Statistical significance for regression fields were evaluated adjusting p-values by controlling for the False Discovery Rate [@benjamini1995;@wilks2016] to avoid misleading results from the high number of regressions [@walker1914;@katz1991].
+
+Linear trends were computed by Ordinary Least Squares and the 95% confidence interval  was computed assuming a t-distribution with the appropriate residual degrees of freedom.
 To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
@@
 
-```{r asymsam-timeseries, fig.cap = "Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
+```{r asymsam-timeseries, fig.cap = "Time series for A\\nobreakdash-SAM and S\\nobreakdash-SAM at (a) 50~hPa and (b) 700~hPa.
 To the right, probability density estimate of each index.
-Series are standarised by the standard deviation of the Full SAM at each level.", fig.width=6, fig.height=3, fig.env="figure*"}
+Series are standarised by the standard deviation of the SAM at each level.", fig.width=6, fig.height=3, fig.env="figure*"}
 indices %>% 
@@
   scale_y_continuous(NULL, position = "right") +
-  scale_x_datetime(NULL, expand = c(0, 0)) +
+  scale_x_datetime(NULL, expand = c(0, 0), 
+                   date_labels = "%Y", date_minor_breaks = "1 year",
+                   breaks = lubridate::as_datetime(seq(as.Date("1980-01-01"), as.Date("2019-01-01"), 
+                                                       by = "5 year"))) +
   facet_grid(lev~., labeller = labeller(lev = lev.lab), switch = "y") +
@@
 
-We first asses the temporal evolution of the Asymmetric SAM and Symmetric SAM.
-Figure \@ref(fig:asymsam-timeseries) shows the corresponding time series for 700 hPa and 50 hPa and their corresponding density estimates.
+We first asses the temporal evolution of A\nobreakdash-SAM and S\nobreakdash-SAM.
+Figure \@ref(fig:asymsam-timeseries) shows the corresponding time series for 700\ hPa and 50\ hPa and their corresponding density estimates.
 We selected these two levels as representative of the tropospheric and stratospheric variability respectively.
@@
 
-Month-to-month variability is evident for both indices, with noisy variations in the low frequency.
+Month-to-month variability is evident for both indices, with noisy variations in the low frequencies.
 At first glance the series can be distinguished by their distributions.
 Compared to the tropospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound.
-The Asymmetric SAM series have both more variability in the higher frequencies than the Symmetric SAM series.
-
-The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \@ref(fig:A2)).
+The A\nobreakdash-SAM series have both more variability in the higher frequencies than the S\nobreakdash-SAM series.
+
+The stratospheric S\nobreakdash-SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \@ref(fig:A2)).
 This might suggest a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b].
-A local peak at 2 years  years is discernible in the periodogram of the tropospheric Symmetric SAM, although it's not statistically significant.
-In the troposphere the most significant peak of variability is found in the Asymmetric index at around `r 12*.3` months.
+A local peak at 2 years  years is discernible in the periodogram of the tropospheric S\nobreakdash-SAM, although it's not statistically significant.
+In the troposphere the most significant peak of variability is found in A\nobreakdash-SAM at around `r 12*.3` months.
 
@@
 
-From a visual inspection, the Asymmetric SAM and Symmetric SAM time series appear to be correlated.
-Moreover, looking at the extremes in the stratosphere, the Symmetric SAM series appears to lag the Asymmetric SAM series (see, for example, the positive events on late 1987).
+From a visual inspection, the A\nobreakdash-SAM and S\nobreakdash-SAM time series appear to be correlated.
+Moreover, looking at the extremes in the stratosphere, the S\nobreakdash-SAM series appears to lag the A\nobreakdash-SAM series (see, for example, the positive events on late 1987).
 Figure \@ref(fig:cor-lev) shows these correlations along all levels considered, for zero and -1 lags.
-Values of zero-lag correlations between the Asymmetric SAM and Symmetric SAM series are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
+Values of zero-lag correlations between A\nobreakdash-SAM and S\nobreakdash-SAM are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
 One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`.
-In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care).
+In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev`\ hPa and then increase again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care).
 At the same time, one-month-lag correlations increase with height.
-Therefore, stratospheric Asymmetric SAM index tends to precede the Symmetric SAM index.
-
-```{r cor-lev, fig.cap = "Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.", fig.width=3, fig.height=3}
+Therefore, stratospheric A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index.
+
+```{r cor-lev, fig.cap = "Correlation between S\\nobreakdash-SAM and A\\nobreakdash-SAM at each level for lag zero and lag -1 (A\\nobreakdash-SAM leads S\\nobreakdash-SAM) for the 1979 -- 2018 period.", fig.width=3, fig.height=3}
 cors %>% 
@@
                         labels = c("0" = "Zero-lag", 
-                                   "-1" = "Asymmetric SAM leads\nSymmetric SAM by 1 month")) +
+                                   "-1" = "A-SAM leads\nS-SAM by 1 month")) +
   scale_color_brewer(NULL, palette = "Dark2", 
                      labels = c("0" = "Zero-lag",
-                                "-1" = "Asymmetric SAM leads\nSymmetric SAM by 1 month")) +
+                                "-1" = "A-SAM leads\nS-SAM by 1 month")) +
   coord_flip()
@@
 
-```{r cross-correlation, fig.cap = "Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.", fig.width = 6, fig.height=3, fig.env="figure*"}
+```{r cross-correlation, fig.cap = "Cross correlation between levels for the (a) SAM, (b) A\\nobreakdash-SAM, and (c) S\\nobreakdash-SAM for the 1979 -- 2018 period.", fig.width = 6, fig.height=3, fig.env="figure*"}
 tri <- matrix(c(1000, 1000, 1000, 1, 1, 1, 1000, 1000), ncol = 2, byrow = TRUE) %>% 
@@
 
-Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
-For the Full SAM (Figure \@ref(fig:cross-correlation)a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
-Above 100 hPa, correlation between levels does falls off more rapidly, indicating less coherent (zero-lag) variability.
+Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across levels for the SAM, A\nobreakdash-SAM  and S\nobreakdash-SAM indices.
+For the SAM (Figure \@ref(fig:cross-correlation)a), high values below 100\ hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
+Above 100\ hPa, correlation between levels does falls off more rapidly, indicating less coherent (zero-lag) variability.
 But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g.
-grater than 0.4 between tropospheric levels and levels below 30 hPa).
-Asymmetric and Symmetric SAM (Figure \@ref(fig:cross-correlation)b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
-Stratospheric coherency is stronger for the Asymmetric SAM than the Symmetric SAM.
-The stratospheric Symmetric SAM seems to connect more strongly to the troposphere than the Asymmetric SAM.
+grater than 0.4 between tropospheric levels and levels below 30\ hPa).
+A\nobreakdash-SAM and S\nobreakdash-SAM (Figure \@ref(fig:cross-correlation)b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
+Stratospheric coherency is stronger for the A\nobreakdash-SAM than the S\nobreakdash-SAM.
+The stratospheric S\nobreakdash-SAM seems to connect more strongly to the troposphere than the stratospheric A\nobreakdash-SAM.
 
@@
 
-```{r trends, fig.cap = "Linear trends (in standard deviations per decade) at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+```{r trends, fig.cap = "Linear trends (in standard deviations per decade) at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) SAM index, (column b) A\\nobreakdash-SAM index, and (column c) S\\nobreakdash-SAM index.
 Shading indicates the 95\\% confidence interval from a t-distribution.", fig.width = 6, fig.height=4, fig.env="figure*", cache = FALSE}
@@
 
-The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \@ref(fig:trends)).
-The Full SAM index presents a statistically positive significant trend (panel `r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
-The seasonal trends (rest of Figure \@ref(fig:trends) column a) indicate that positive trends are present in  Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
+The linear trends for each of the indices (SAM, S\nobreakdash-SAM and A\nobreakdash-SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \@ref(fig:trends)).
+The SAM index presents a statistically positive significant trend (Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50\ hPa and reaches its maximum value at 100\ hPa.
+The seasonal trends (rest of Figure \@ref(fig:trends) column a) indicate that positive trends are present in  Autumn and particularly in Summer, where the 100\ hPa maximum is much more defined.
 In Winter and Spring, we detect no statistically significant trend.
@@
 
-By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.
+By separating the SAM signal in its asymmetric and symmetric parts, notonly we can see that these trends are almost entirely due to the symmetric component (column b vs.
 column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer.
-In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in panel `r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
-In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full SAM index.
-
-```{r regression-90}
-local <- indices %>%
-  .[time >= as.Date("1990-01-01")] %>% 
-  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
-  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term, season(time))] %>% 
-  rm_intercept() %>% 
-  .[, term := NULL] %>% 
-  .[, estimate := estimate*3600*24*365*10] %>% 
-  .[, std.error := std.error*3600*24*365*10]
-
-trends_text <- local[, paste0(signif(estimate, 2), " \\pm ", signif(std.error*1.96, 2)), 
-                     by = .(term = type, season, lev)]
-```
+In Summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
+In Autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.
 
@@
 There is no evidence of a significant trend in the stratosphere.
-In the troposphere, there is a positive trend for the Asymmetric SAM and not significant trend for the Symmetric SAM.
+In the troposphere, there is a positive trend for A\nobreakdash-SAM and no significant trend for S\nobreakdash-SAM.
 This suggest that the SAM has become more asymmetric in the period from 1979 to 2018.
@@
 
-```{r r-squared-trend, fig.height=3, fig.width=3, fig.cap = "Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018.
+```{r r-squared-trend, fig.height=3, fig.width=3, fig.cap = "Linear trends (in percent per decade) of the variance explained by A\\nobreakdash-SAM and S\\nobreakdash-SAM at each level for the period 1979 -- 2018.
 Shading indicates the 95\\% confidence interval."}
@@
   scale_x_level() +
-  scale_y_continuous("Change in explained variance per decade", label = scales::percent_format()) +
+  scale_y_continuous("Change in explained variance per decade", 
+                     label = scales::percent_format(1)) +
   scale_fill_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill"), labels = lab_sam) +
@@
 
-```{r 2d-regr, fig.cap = "Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices.
+```{r 2d-regr, fig.cap = "Regression of geopotential height (meters) at (row 1) 50~hPa and (row 2) 700~hPa with (column a) SAM, (column b) A\\nobreakdash-SAM, and (column c) S\\nobreakdash-SAM for the 1979 -- 2018 period.
+The regression patterns for A\\nobreakdash-SAM and S\\nobreakdash-SAM are the result of one multiple regression using both indices.
 Points marked on panel b.2 are the location of the reference points used by \\cite{raphael2004} for their Zonal Wave 3 index.
@@
 
-We then computed the spatial regression of geopotential height anomalies on Full, Asymmetric and Symmetric indices at 700 hPa and 50 hPa levels (Figure \@ref(fig:2d-regr)).
-While, regression coefficients in column a are computed using the Full SAM, the regression coefficients in columns b and c of Figure \@ref(fig:2d-regr) are computed using multiple regression using the Asymmetric and Symmetric indices at the same time.
-Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other index.
-
-In the stratosphere, the spatial pattern associated with the Full SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \@ref(fig:2d-regr)a.1) which is, not centred in the South Pole.
-On the other hand, the monopole associated with the Symmetric SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric.
-Furthermore, the regression pattern of the Asymmetric SAM is characterized by a wave-1 structure with centers over the Bellinghausen Sea on the
-Western Hemisphere and Davis Sea in the Eastern Hemisphere
-
-In the troposphere, the regression pattern associated with Full SAM shows the well-known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 (Figure \@ref(fig:2d-regr)a.2, [@fogt2012]).
-The regression patterns associated with the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
-While, the Asymmetric SAM index gives rise to a cleaner zonal wave (Figure \@ref(fig:2d-regr)b.2), the Symmetric SAM index is associated with an annular structure, almost devoid of zonal asymmetries (Figure \@ref(fig:2d-regr)c.2).
-The wave-3 pattern observed in Figure \@ref(fig:2d-regr)b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern described by Raphael (2004), whose reference locations are marked with points in the figure.
-Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
+We then computed the spatial regression of geopotential height anomalies on the A\nobreakdash-SAM and S\nobreakdash-SAM indices at 700\ hPa and 50\ hPa levels (Figure \@ref(fig:2d-regr)).
+While regression coefficients in Figure \@ref(fig:2d-regr) column a are computed using SAM, the regression coefficients in columns b and c of Figure \@ref(fig:2d-regr) are computed using multiple regression using the A\nobreakdash-SAM  and S\nobreakdash-SAM indices at the same time.
+Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other one.
+
+In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \@ref(fig:2d-regr)a.1) which is not centred in the South Pole.
+On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric.
+Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centers over the Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern Hemisphere
 
@@
 raphael <- ReadNetCDF(ERA5(), 
-                      subset = list(time = c("1979-01-01", "2018-12-31"),
+                      subset = list(time = c("1979-01-01", "2019-01-01"),
                                     latitude = -49,
@@
   .[, .(raphael = mean(hgt_s)), by = time]
-```
+
+cor_raphael <- raphael[indices_wide[lev == 500], on = "time"] %>% 
+  na.omit() %>% 
+  .[, cor.test(asym, raphael)]
+```
+
+In the troposphere, the regression pattern associated with the SAM shows the well-known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 (Figure \@ref(fig:2d-regr)a.2, [@fogt2012]).
+The regression patterns associated with the A\nobreakdash-SAM and S\nobreakdash-SAM indices successfully disentangle both structures.
+While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \@ref(fig:2d-regr)b.2), the S\nobreakdash-SAM index is associated with an annular structure, almost devoid of zonal asymmetries (Figure \@ref(fig:2d-regr)c.2).
+The wave-3 pattern observed in Figure \@ref(fig:2d-regr)b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern described by @raphael2004, whose reference locations are marked with points in the figure.
+Indeed, there is no correlation between @raphael2004's index and  A\nobreakdash-SAM (cor = `r signif(cor_raphael$estimate, 2)`, p-value `r report_p(cor_raphael$p.value)`).
+Thus, the tropospheric A\nobreakdash-SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
 
@@
 
-The amplitude of the first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \@ref(fig:wave-amplitude), where wave number zero represents the amplitude of the zonal mean.
-Zonal wave amplitudes of the spatial pattern described by the Full SAM index (Figure \@ref(fig:wave-amplitude) column `r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at both levels.
-However, zonal waves are important, particularly North of 50\degree S, with wavenumber 1 clearly dominating at 50 hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of waves of similar amplitude at 700 hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`).
-Figure \@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean.
-On the other hand, the Symmetric SAM it is almost entirely explained by the zonal mean at both levels (Figure \@ref(fig:wave-amplitude) column c), with little to no contribution from the zonal waves with wavenumbers 1 to 3.
+The amplitude of the first zonal wave numbers at each latitude at 50\ hPa and 700\ hPa are shown in Figure \@ref(fig:wave-amplitude), where wave number zero represents the amplitude of the zonal mean.
+Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column `r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at both levels.
+However, zonal waves are important, particularly South of 50\degree S, with wavenumber 1 clearly dominating at 50\ hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of waves of similar amplitude at 700\ hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`).
+Figure \@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean.
+On the other hand, the S\nobreakdash-SAM it is almost entirely explained by the zonal mean at both levels (Figure \@ref(fig:wave-amplitude) column c), with little to no contribution from zonal waves with wavenumbers 1 to 3.
 
@@
 
-```{r vertical-regression, fig.cap = "Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\\degree and 40\\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
-(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979 -- 2018 period.", fig.width = 4, fig.height=3}
+```{r vertical-regression, fig.cap = "Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\\degree and 40\\degree S and the A\\nobreakdash-SAM index (extracted from a multiple regression which included the S\\nobreakdash-SAM index).
+(a) With the A\\nobreakdash-SAM in 50~hPa and (b) in 700~hPa for the 1979 -- 2018 period.", fig.width = 4, fig.height=3}
 vertical_regr %>% 
@@
 
-To analyse the vertical structure of the geopotential height anomalies associated with the asymmetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (Figure \@ref(fig:vertical-regression)a) and for the 700 hPa Asymmetric SAM index (Figure \@ref(fig:vertical-regression)b).
-The geopotential height anomalies associated with the stratospheric Asymmetric SAM (Figure \@ref(fig:vertical-regression)a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
-The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes.
-The signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
-
-The tropospheric Asymmetric SAM (Figure \@ref(fig:vertical-regression)b) has significant signals that extend upwards to the uppermost levels considered.
-In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
-The anomalies are larger in the Western hemisphere, where they extent into the stratosphere.
+To analyse the vertical structure of the geopotential height anomalies associated with the A\nobreakdash-SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50\ hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)a) and for the 700\ hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)b).
+The geopotential height anomalies associated with the stratospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric A\nobreakdash-SAM.
+The vertical structure of this signal tilts about 60\degree\ to the West between 100\ hPa and 1\ hPa, suggesting baroclinic processes.
+The signal in the stratosphere maximises near 10\ hPa despite using the 50\ hPa index for the regression.
+
+The tropospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)b) has significant signals that extend upwards to the uppermost levels considered.
+In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250\ hPa.
+The anomalies are larger in the Western hemisphere, where they extend into the stratosphere.
 In the Eastern hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere.
-So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
-The structures shown in Figure \@ref(fig:vertical-regression)a Figure \@ref(fig:vertical-regression)b in  Figure \@ref(fig:vertical-regression) are robust to the choice of index level.
-For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in Figure \@ref(fig:vertical-regression)a.
-Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to Figure \@ref(fig:vertical-regression)b.
+So, while the tropospheric A\nobreakdash-SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric A\nobreakdash-SAM.
+The structures shown in  Figure \@ref(fig:vertical-regression) are robust to the choice of index level.
+For any stratospheric (above 100\ hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10\ hPa in Figure \@ref(fig:vertical-regression)a.
+Conversely, for any tropospheric (below 100\ hPa) index, the result is very similar to Figure \@ref(fig:vertical-regression)b.
 The patterns mainly change in amplitude (not shown).
@@
 enso <- rbindlist(list(oni = oni, soi = soi, mei = mei), idcol = "index")
-
-report_p <- function(p.value) {
-  ifelse(p.value < 0.001, " < 0.001", paste0(" = ", signif(p.value, 2)))
-}
 
@@
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value `r cor_texts[term == "full"]$p.value`).
-This relationship is captured mainly by the Asymmetric SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI, consistent with @fan2007's results, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
+The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value `r cor_texts[term == "full"]$p.value`).
+This relationship is captured mainly by the A\nobreakdash-SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI, consistent with @fan2007's results, whereas the S\nobreakdash-SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
 The same analysis was performed using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987], obtaining similar results.
@@
 
-```{r regr-air-season, fig.cap = "Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
+```{r regr-air-season, fig.cap = "Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with SAM, A\\nobreakdash-SAM and S\\nobreakdash-SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Gray areas in Antarctica are areas with have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*", cache = FALSE}
+Gray areas in Antarctica have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*", cache = FALSE}
 air_regr %>% 
-  # .[season == "DJF"] %>% 
   copy() %>% 
-  # .[lat <= -61, estimate := NA] %>%
   .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
-  # .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] %>%
   periodic(lon = c(0, 360)) %>% 
@@
 
-To assess the differences in the impacts associated with full, asymmetric and symmetric SAM, we regress s surface land air and sea temperature and precipitation onto each of the three SAM indices at 700 hPa.
+To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices we regressed surface land air and sea temperature and precipitation onto each of the three SAM indices at 700\ hPa.
 As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
 
-Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface land air and sea temperature for each trimester.
-In Summer, positive values of the Full SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
+Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700\ hPa with surface land air and sea temperature for each trimester.
+In Summer, positive values of the SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
 The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W,  120\degree W, 150\degree E and 90\degree E respectively.
 In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
-Figure \@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of the Asymmetric and Symmetric SAM, respectively.
-Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the Asymmetric SAM regression map, while temperature patterns linked to positive Symmetric SAM show a more zonally consistent ring and less relation to the tropics.
-Temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to Asymmetric SAM.
-This signal is not present in the regression pattern with the Full SAM.
-Spring (Figure \@ref(fig:regr-air-season), row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with less regions where regressed anomalies have statistical significance.
-
-In Autumn and Winter (rows `r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the Full SAM.
+Figure \@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of A\nobreakdash-SAM and S\nobreakdash-SAM, respectively.
+Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics.
+Temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM.
+This signal is not present in the regression pattern with the SAM.
+Spring (Figure \@ref(fig:regr-air-season), row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with fewer regions where regressed anomalies have statistical significance.
+
+In Autumn and Winter (rows `r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the SAM.
 There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
@@
 The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons -- although with varying intensity and small-scale details -- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
-It's then not surprising to see it more clearly in association with the Symmetric SAM (at least in Summer and Spring).
+It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in Summer and Spring).
 
@@
 
-Regression of the SAM indices with seasonal mean precipitation and 700 hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for Australasia and South America respectively.
+Regression of the SAM indices with seasonal mean precipitation and 700\ hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for Australasia and South America respectively.
 South Africa is not shown because no significant signal was detected there.
@@
 
-```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height anomalies (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
+```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700~hPa geopotential height anomalies (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) SAM, (column (b) A\\nobreakdash-SAM and (column c) S\\nobreakdash-SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
@@
 
-In Australia, the annual regression shows that the Full SAM index is associated with positive precipitation anomalies in the southeastern region (Figure \@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006.
-The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (Figure \@ref(fig:pp-regr-oceania)c.1).
+In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the southeastern region (Figure \@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006.
+The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure \@ref(fig:pp-regr-oceania)c.1).
 Geopotential height anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by @hendon2007.
-The Asymmetric SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
-
-The seasonal-based regressions show statistically significant anomalies only in Spring, when positive Full SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
-In this trimester, the Symmetric SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related with positive Asymmetric SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).
-
-In Summer, positive Full SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \@ref(fig:pp-regr-oceania)a.2).
-The Eastern part being dominated by the relationship with the Symmetric SAM and the Western, by the Asymmetric SAM.
-In Autumn, the regression with Full SAM shows positive anomalies in the North, similar to Summer, and a broad area of positive anomalies in the North-East to South-West direction.
-This structure seems to be associated with the Symmetric SAM, while the Northern positive values are associated with the Asymmetric SAM.
-In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the Symmetric SAM.
+A\nobreakdash-SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
+
+The seasonal-based regressions show statistically significant anomalies only in Spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
+In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related with positive A\nobreakdash-SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).
+
+In Summer, positive SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \@ref(fig:pp-regr-oceania)a.2).
+The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
+In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, and a broad area of positive anomalies in the North-East to South-West direction.
+This structure seems to be associated with the S\nobreakdash-SAM, while the Northern positive values are associated with the A\nobreakdash-SAM.
+In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the S\nobreakdash-SAM.
 None of these regression coefficients are statistically significant at the 95% level.
@@
 
-In South America (Figure \@ref(fig:pp-regr-america)), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
-Figure \@ref(fig:pp-regr-america)b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with the Asymmetric SAM, that in southern Chile is related to the Symmetric SAM.
+In South America (Figure \@ref(fig:pp-regr-america)), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile, and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
+Figure \@ref(fig:pp-regr-america)b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with A\nobreakdash-SAM, that in southern Chile is related to S\nobreakdash-SAM.
 
 Except Winter, seasonal-based regressions mirror this same pattern.
-Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the Full SAM.
-The separation of these features between the Asymmetric SAM and Symmetric SAM regression maps is also rather consistent.
-
-The anomalous circulation at 700 hPa associated with the Symmetric SAM (Figure \@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over Southern Chile.
-This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region [e.g.
+Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the SAM.
+The separation of these features between A\nobreakdash-SAM and S\nobreakdash-SAM regression maps is also rather consistent.
+
+The anomalous circulation at 700\ hPa associated with S\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over Southern Chile.
+This leads to reduced influx of moist air from the Pacific Ocean which is the main source of precipitable water in that region [e.g.
 @garreaud2007].
-On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
+On the other hand, the anomalous circulation associated with positive values of A\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
 This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region [@silvestri2009;@zamboni2010].
 This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events [@rosso2018].
-There is a small area of significative positive precipitation anomalies with SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the Asymmetric SAM.
+There is a small area of significative positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the A\nobreakdash-SAM.
 
@@
 
-The Asymmetric SAM index correlates strongly with the Symmetric SAM index.
-In the troposphere, this correlation is maximum at zero lag, while in the stratosphere is maximised with the Asymmetric SAM leading the Symmetric SAM by one month.
+The A\nobreakdash-SAM index correlates strongly with the S\nobreakdash-SAM index.
+In the troposphere, this correlation is maximum at zero lag, while in the stratosphere is maximised with the A\nobreakdash-SAM leading the S\nobreakdash-SAM by one month.
 Since most indices of the SAM are calculated using surface or near-surface conditions, this result would suggest that they might not be sensitive to the most dramatic changes in SAM variability.
 
-The two-year periodicity we found in the stratospheric Symmetric SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
+The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
 There is evidence of influence between the QBO and the Northern Annular Mode [e.g.
-@holton1980;@watson2014;@zhang2020], so it is not unlikely that the SAM would be similarly connected.
+@holton1980;@watson2014;@zhang2020], so it is not unlikely that the SAM would be similarly affected.
 However, establishing this link would require further research.
 We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies [e.g.
-@fogt2020; and references therein] for low levels.
-We show that these trends maximise at 100 hPa, and are explained by the zonally symmetric component.
-We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the Full SAM index.
-In contrast to @fogt2012 who made the evaluation for the period 158 -- 2001, we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
-This might be due either due to differneces in methodology or analysed period.
-
-In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being fully annular than the patterns associated with the Full SAM index.
-The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific This pattern extends in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimum for capturing this variability.
+@fogt2020 and references therein] for low levels.
+We show that these trends maximise at 100\ hPa, and are explained by the zonally symmetric component.
+We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the SAM index.
+We find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
+This is in contrast to @fogt2012 who made the evaluation for the period 158 -- 2001.
+This discrepancy might be due either due to differneces in methodology or analysed period.
+
+In the troposphere, the spatial patterns of geopotential associated with the S\nobreakdash-SAM index are much closer to being fully annular than the patterns associated with the SAM index.
+The A\nobreakdash-SAM index, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific.
+This pattern extends vertically in the troposphere but its maximum is located at 250\ hPa, which also could suggest that surface-based indices are not opitimal for capturing this variability.
 This wave-3 pattern is similar to the Pacific-South American Pattern, linked to ENSO variability.
-We found that the significant correlation that exists between the Full SAM index and the Oceanic Niño Index over this period, is captured entirely by the Asymmetric SAM index.
-This suggests that ENSO is linked to SAM exclusively through the variability in the latter’s asymmetric component and thus, the Asymmetric SAM index could be a useful measure to further study that relationship.
-
-Temperature anomalies associated with the Full SAM broadly show a pattern of negative anomalies at polar latitudes surrounded by positive anomalies, but with many deviations from symmetry.
-The Asymmetric SAM index explains a big portion of these deviations.
-In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship.
+We found that the significant correlation that exists between the SAM index and the Oceanic Niño Index over this period is captured entirely by the A\nobreakdash-SAM index.
+This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the A\nobreakdash-SAM index could be a useful measure to further study that relationship.
+
+Temperature anomalies associated with the SAM broadly show a pattern of negative anomalies at polar latitudes surrounded by positive anomalies, but with many deviations from symmetry.
+The A\nobreakdash-SAM index explains a big portion of these deviations.
+In particular, the positive phase of A\nobreakdash-SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO\nobreakdash-SAM relationship.
 These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [e.g.
@@
 
-In Australia the Full SAM is associated with positive precipitation anomalies in South East and this is explained by the Symmetric SAM.
-However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
-In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
-These features are cleanly separated between the Asymmetric and Symmetric components.
-The Symmetric SAM explains the negative anomalies in Southern Chile and the Asymmetric SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
+In Australia the SAM is associated with positive precipitation anomalies in South East and this is explained by the S\nobreakdash-SAM.
+However, the A\nobreakdash-SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
+In South America, precipitation anomalies associated with the SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
+These features are cleanly separated between A\nobreakdash-SAM and S\nobreakdash-SAM.
+S\nobreakdash-SAM explains the negative anomalies in Southern Chile and A\nobreakdash-SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
 Individual seasons mostly follow this pattern.
@@
 The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s [@fogt2006;@clem2013].
-Seeing as both the ENSO-SAM relationship and most of the precipitation impacts in South America are captured by the Asymmetric SAM, the results presented here are most likely period-dependent.
-The decadal variations of the Asymmetric SAM should be focus of future studies.
+Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the results presented here are most likely period-dependent.
+The decadal variations of the A\nobreakdash-SAM should be focus of future studies.
 
@@
 
-```{r A1, fig.cap = "Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
-Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.
+```{r A1, fig.cap = "Lag-correlation between the A\\nobreakdash-SAM and the S\\nobreakdash-SAM index at each level.
+Negative lags imply A\\nobreakdash-SAM leading S\\nobreakdash-SAM and vice versa.
 For the 1979 -- 2018 period.", fig.width = 5, fig.height=5, fig.pos="ht"}
@@
 ```{r A2, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
-The shading indicates de 95\\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\\% of the simulated sampels had an amplitude equal or lower).
+The shading indicates de 95\\% confidence area derived by computing the spectrum for 5000 simulated samples from a fitted autorregressive model and (95\\% of the simulated sampels had an amplitude equal or lower).
 The light line indicates the theoretical expected amplitude from the autorregressive model.
@@
 
-```{r A3, fig.cap = "50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
-Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.", apendix = TRUE, fig.height = 5, fig.width = 6}
+```{r A3, fig.cap = "50~hPa geopotential height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+Numbers in the column headers are the spatial correlation between SAM+ and SAM- composites and number of monthly fields used to construct each composite.", apendix = TRUE, fig.height = 5, fig.width = 6}
 composites <- hgt[lev %in% c(50, 700)] %>%
@@
 
-```{r A4, fig.cap = "700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
-Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.", apendix = TRUE, fig.height = 4.5, fig.width = 6}
+(ref:a4-cap) Same as \@ref(fig:A3) but for 700\ hPa geopotential height.
+
+```{r A4, fig.cap = "(ref:a4-cap)", apendix = TRUE, fig.height = 4.5, fig.width = 6}
 plot_composites(700)
@@
 
-```{r A5, fig.cap = "Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.",  apendix = TRUE, fig.width=6, fig.height=4.5}
+```{r A5, fig.cap = "Regression coefficients of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.",  apendix = TRUE, fig.width=6, fig.height=4.5}
 set.seed(42)
@@
 
-```{r A6, fig.cap = "Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
-Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.", fig.width=6, fig.height=5}
+```{r A6, fig.cap = "Regression of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
+Spatial correlation between both fields is 0.86 for the 50~hPa fields and 0.76 for the 700~hPa fields.", fig.width=6, fig.height=5}
 set.seed(42)

### 2020-12-11 analysis/paper/paper.tex
Última revision. Que el resto de los errores los encuentren les revisores.

@@
 
-\title{Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
+\title{Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
 Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }
 
-    \titlerunning{Assessment of zonal symmetric and asymmetric components of the SAM}
+    \titlerunning{Assessment of zonally symmetric and asymmetric components of the SAM}
 
@@
 Moreover, tropical influences on the SAM have been observed \citep{fan2007, fogt2011, clem2013}.
-\citet{fan2007} computed SAM indices of the Western and the Eastern Hemisphere separately and found that they were much more correlated to each other if the (linear) signal of the ENSO was removed.
+\citet{fan2007} computed SAM indices of the Western and the Eastern Hemispheres separately and found that they were much more correlated to each other if the (linear) signal of the ENSO was removed.
 
-Positive trends in SAM index have been documented by various researchers using different indices mostly on austral Summer and Autumn \citep[e.g.][ and references therein]{fogt2020}.
-These trends are thought of driven primarily by stratospheric ozone depletion and the increase in greenhouse gases and understood in the context of zonal mean variables \citep{marshall2004, gillett2005, arblaster2006, gillett2013}.
-However, it's not clear yet how or if the asymmetric SAM component responds to this forcing, or how its variability alters the observed trends.
+Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn \citep[e.g.][ and references therein]{fogt2020}.
+It is though that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables \citep{marshall2004, gillett2005, arblaster2006, gillett2013}.
+However, it's not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.
 
 The impact of the zonally asymmetric component of the SAM at regional scales has not been studied in detail yet.
-Positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at higher latitudes \citep{jones2019} (and vice versa for negative SAM).
+The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at higher latitudes \citep{jones2019} (and vice versa for negative SAM).
 But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}.
-SAM signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry \citep{lim2016}.
+The SAM-related signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry \citep{lim2016}.
 The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
@@
 The study of the temporal variability of the asymmetric component of the SAM has not received much attention except for \citet{fogt2012}.
-This study provides evidences of the relevance of SAM asymmetric component.
+This study provides evidences for the relevance of the SAM's asymmetric component.
 However, their conclusions are based on composites of positive and negative SAM events including a small number of cases unevenly distributed among years with and without satellite information.
@@
 Their vertical structure and coherence, temporal variability and trends are consequently assesed.
-We then study the spatial patterns described by the variability exclusive to each index focusing on 50 hPa as representing the stratosphere and 700 hPa as representing the troposphere.
-Finally, the relationships of the SAM at 700 hPa with temperature and precipitation anomalies are investigated.
+We then study the spatial patterns described by the variability exclusive to each index focusing on 50~hPa as representing the stratosphere and 700~hPa as representing the troposphere.
+Finally, the relationships of the SAM at 700~hPa with temperature and precipitation anomalies are investigated.
 
@@
 
-Traditionally the SAM is defined as the leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels \citep{ho2012}.
+Traditionally, the SAM is defined as the leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels \citep{ho2012}.
 Following \citet{baldwin2001}, we extend that definition vertically and use the term SAM to refer to the leading EOF of the monthly anomalies of geopotential height south of 20\degree S at each level.
-We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in 481 rows and 4176 columns (144 points of longitude and 29 points of latitude).
+We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in 480 rows and 4176 columns (144 points of longitude and 29 points of latitude).
 We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint \citep{chung1999}.
-We consider in the EOF analysis all months together without separating in seasons.
+We consider in the EOF analysis all months together without dividing by seasons.
 
-To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \ref{fig:method} at 700 hPa.
+To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \ref{fig:method} at 700~hPa.
 The full spatial signal (\(\mathrm{EOF_1}(\lambda, \phi)\)) is the sum of the zonally asymmetric (\(\mathrm{EOF_1^*}(\lambda, \phi)\)) and symmetric (\([\mathrm{EOF_1}](\lambda, \phi)\)) components.
-We then compute the ``Full SAM'', ``Asymmetric SAM'' and ``Symmetric SAM'' indices as the regression coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude).
-The three indices are then normalized by dividing them by the standard deviation of the Full SAM index at each level.
+We then compute the SAM index, Asymmetric SAM index (A\nobreakdash-SAM) and Symmetric SAM (S\nobreakdash-SAM) indices as the coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude).
+The three indices are then normalized by dividing them by the standard deviation of the SAM index at each level.
 As a result, the magnitudes between indices are comparable.
-However, only Full SAM index has unit standard deviation per definition.
+However, only SAM index has unit standard deviation per definition.
 The explained variance of each pattern is used as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
@@
 \begin{figure*}
-\includegraphics{method-1} \caption{Spatial patterns of the first EOF of 700 hPa geopotential height for 1979 -– 2018 period.
+\includegraphics{method-1} \caption{Spatial patterns of the first EOF of 700~hPa geopotential height for 1979 -– 2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 The method assumes linearity in the asymmetric component of the SAM.
-That means that zonal symmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
+That means that zonal asymmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
 \citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
-To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the Full SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700 hPa and 50 HPa levels (Figures \ref{fig:A3} and \ref{fig:A4}).
+To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700~hPa and 50~hPa levels (Figures \ref{fig:A3} and \ref{fig:A4}).
 In all seasons and both levels, SAM+ composites are similar to SAM- in structure but with the opposite sign.
@@
 
-The method also implies that the zonally asymmetric pattern of SAM remains stationary along the period considered.
+The method also assumes that the zonally asymmetric pattern of the SAM remains stationary along the period considered.
 \citet{silvestri2009} suggest that this might not be the case between 1958 and 2004.
@@
 
-We perform linear regressions to quantify the association between the SAM indices and other variables.
-Moreover, we apply multiple linear regression analysis to describe the combined influence of both Asymmetric and Symmetric SAM indices.
-To obtain the linear coefficients of a variable \(X\) (geopotential, temperature, precipitation, etc\ldots{}) with the Asymmetric SAM (\(SAM_a\)) and Symmetric SAM (\(SAM_s\)) we fit the equation
+We performed linear regressions to quantify the association between the SAM indices and other variables.
+Moreover, we apply multiple linear regression analysis to describe the combined influence of both A\nobreakdash-SAM and S\nobreakdash-SAM.
+To obtain the linear coefficients of a variable \(X\) (geopotential, temperature, precipitation, etc\ldots{}) with A\nobreakdash-SAM and S\nobreakdash-SAM we fit the equation
 
 \[
-X(\lambda, \phi, t) = \alpha(\lambda, \phi) SAM_a + \beta(\lambda, \phi) SAM_s + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
+X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{A-SAM} + \beta(\lambda, \phi) \operatorname{S-SAM} + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
 \]
@@
 where \(\lambda\) and \(\phi\) are the longitude and latitude, \(t\) is the time, \(\alpha\) and \(\beta\) are the linear regression coefficients, \(X_0\) and \(\epsilon\) are the constant and error terms.
-From this equation, \(\alpha\) represents the (linear) association of \(X\) with the variability of the Asymmetric SAM that is not explained by the variability of the Symmetric SAM; i.e.~it is proportional to the partial correlation of \(X\) and the Asymmetric SAM, controlling for the effect of the Symmetric SAM and vice versa for \(\beta\).
-When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we average the relevant variables seasonally for each year and trimester before computing the regression.
+From this equation, \(\alpha\) represents the (linear) association of \(X\) with the variability of A\nobreakdash-SAM that is not explained by the variability of S\nobreakdash-SAM; i.e.~it is proportional to the partial correlation of \(X\) and A\nobreakdash-SAM, controlling for the effect of S\nobreakdash-SAM, and vice versa for \(\beta\).
+When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we averaged the relevant variables seasonally for each year and trimester before computing the regression.
 
-Statistical significance for regression fields were evaluated adjusting p-values by controlling for the False Discovery Rate \citep{benjamini1995, wilks2016} to avoid misleading results from a high number of regressions \citep{walker1914, katz1991}.
+Statistical significance for regression fields were evaluated adjusting p-values by controlling for the False Discovery Rate \citep{benjamini1995, wilks2016} to avoid misleading results from the high number of regressions \citep{walker1914, katz1991}.
 
-Linear trends were computed by Ordinary Least Squares and the 95\% confidence interval assuming a t-distribution of the appropriate residual degrees of freedom.
+Linear trends were computed by Ordinary Least Squares and the 95\% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom.
 To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
@@
 \begin{figure*}
-\includegraphics{asymsam-timeseries-1} \caption{Time series for the Asymmetric SAM and Symmetric SAM indices at (a) 50 hPa and (b) 700 hPa.
+\includegraphics{asymsam-timeseries-1} \caption{Time series for A\nobreakdash-SAM and S\nobreakdash-SAM at (a) 50~hPa and (b) 700~hPa.
 To the right, probability density estimate of each index.
-Series are standarised by the standard deviation of the Full SAM at each level.}\label{fig:asymsam-timeseries}
+Series are standarised by the standard deviation of the SAM at each level.}\label{fig:asymsam-timeseries}
 \end{figure*}
 
-We first asses the temporal evolution of the Asymmetric SAM and Symmetric SAM.
-Figure \ref{fig:asymsam-timeseries} shows the corresponding time series for 700 hPa and 50 hPa and their corresponding density estimates.
+We first asses the temporal evolution of A\nobreakdash-SAM and S\nobreakdash-SAM.
+Figure \ref{fig:asymsam-timeseries} shows the corresponding time series for 700~hPa and 50~hPa and their corresponding density estimates.
 We selected these two levels as representative of the tropospheric and stratospheric variability respectively.
@@
 
-Month-to-month variability is evident for both indices, with noisy variations in the low frequency.
+Month-to-month variability is evident for both indices, with noisy variations in the low frequencies.
 At first glance the series can be distinguished by their distributions.
 Compared to the tropospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound.
-The Asymmetric SAM series have both more variability in the higher frequencies than the Symmetric SAM series.
+The A\nobreakdash-SAM series have both more variability in the higher frequencies than the S\nobreakdash-SAM series.
 
-The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \ref{fig:A2}).
+The stratospheric S\nobreakdash-SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \ref{fig:A2}).
 This might suggest a link between stratospheric SAM variability and the Quasi-Biennial Oscillation \citep{baldwin2001b}.
-A local peak at 2 years years is discernible in the periodogram of the tropospheric Symmetric SAM, although it's not statistically significant.
-In the troposphere the most significant peak of variability is found in the Asymmetric index at around 3.6 months.
+A local peak at 2 years years is discernible in the periodogram of the tropospheric S\nobreakdash-SAM, although it's not statistically significant.
+In the troposphere the most significant peak of variability is found in A\nobreakdash-SAM at around 3.6 months.
 
-From a visual inspection, the Asymmetric SAM and Symmetric SAM time series appear to be correlated.
-Moreover, looking at the extremes in the stratosphere, the Symmetric SAM series appears to lag the Asymmetric SAM series (see, for example, the positive events on late 1987).
+From a visual inspection, the A\nobreakdash-SAM and S\nobreakdash-SAM time series appear to be correlated.
+Moreover, looking at the extremes in the stratosphere, the S\nobreakdash-SAM series appears to lag the A\nobreakdash-SAM series (see, for example, the positive events on late 1987).
 Figure \ref{fig:cor-lev} shows these correlations along all levels considered, for zero and -1 lags.
-Values of zero-lag correlations between the Asymmetric SAM and Symmetric SAM series are relatively constant throughout the troposphere, fluctuating between 0.39 and 0.45.
+Values of zero-lag correlations between A\nobreakdash-SAM and S\nobreakdash-SAM are relatively constant throughout the troposphere, fluctuating between 0.39 and 0.45.
 One-month-lag correlations are similarly constant but significantly reduced to around 0.17.
-In the stratosphere, zero-lag correlations drop to a minimum of 0.21 at 20 hPa and then it increases again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care).
+In the stratosphere, zero-lag correlations drop to a minimum of 0.21 at 20~hPa and then increase again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care).
 At the same time, one-month-lag correlations increase with height.
-Therefore, stratospheric Asymmetric SAM index tends to precede the Symmetric SAM index.
+Therefore, stratospheric A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index.
 
 \begin{figure}
-\includegraphics{cor-lev-1} \caption{Correlation between the Symmetric SAM and Asymmetric SAM index at each level for lag zero and lag -1 (Asymmetric SAM leads Symmetric SAM) for the 1979 -- 2018 period.}\label{fig:cor-lev}
+\includegraphics{cor-lev-1} \caption{Correlation between S\nobreakdash-SAM and A\nobreakdash-SAM at each level for lag zero and lag -1 (A\nobreakdash-SAM leads S\nobreakdash-SAM) for the 1979 -- 2018 period.}\label{fig:cor-lev}
 \end{figure}
@@
 \begin{figure*}
-\includegraphics{cross-correlation-1} \caption{Cross correlation between levels of the (a) Full SAM, (b) Asymmetric SAM, and (c) Symmetric SAM for the 1979 -- 2018 period.}\label{fig:cross-correlation}
+\includegraphics{cross-correlation-1} \caption{Cross correlation between levels for the (a) SAM, (b) A\nobreakdash-SAM, and (c) S\nobreakdash-SAM for the 1979 -- 2018 period.}\label{fig:cross-correlation}
 \end{figure*}
 
-Figure \ref{fig:cross-correlation} shows (zero-lag) cross-correlation across levels for the Full, Symmetric and Asymmetric SAM indices.
-For the Full SAM (Figure \ref{fig:cross-correlation}a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
-Above 100 hPa, correlation between levels does falls off more rapidly, indicating less coherent (zero-lag) variability.
-But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g.~grater than 0.4 between tropospheric levels and levels below 30 hPa).
-Asymmetric and Symmetric SAM (Figure \ref{fig:cross-correlation}b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
-Stratospheric coherency is stronger for the Asymmetric SAM than the Symmetric SAM.
-The stratospheric Symmetric SAM seems to connect more strongly to the troposphere than the Asymmetric SAM.
+Figure \ref{fig:cross-correlation} shows (zero-lag) cross-correlation across levels for the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices.
+For the SAM (Figure \ref{fig:cross-correlation}a), high values below 100~hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
+Above 100~hPa, correlation between levels does falls off more rapidly, indicating less coherent (zero-lag) variability.
+But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g.~grater than 0.4 between tropospheric levels and levels below 30~hPa).
+A\nobreakdash-SAM and S\nobreakdash-SAM (Figure \ref{fig:cross-correlation}b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
+Stratospheric coherency is stronger for the A\nobreakdash-SAM than the S\nobreakdash-SAM.
+The stratospheric S\nobreakdash-SAM seems to connect more strongly to the troposphere than the stratospheric A\nobreakdash-SAM.
 
 \begin{figure*}
-\includegraphics{trends-1} \caption{Linear trends (in standard deviations per decade) at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index.
+\includegraphics{trends-1} \caption{Linear trends (in standard deviations per decade) at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979 -- 2018 and for the (column a) SAM index, (column b) A\nobreakdash-SAM index, and (column c) S\nobreakdash-SAM index.
 Shading indicates the 95\% confidence interval from a t-distribution.}\label{fig:trends}
@@
 
-The linear trends for each of the indices (Full SAM, Symmetric SAM and Asymmetric SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \ref{fig:trends}).
-The Full SAM index presents a statistically positive significant trend (panel a.1) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
-The seasonal trends (rest of Figure \ref{fig:trends} column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
+The linear trends for each of the indices (SAM, S\nobreakdash-SAM and A\nobreakdash-SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \ref{fig:trends}).
+The SAM index presents a statistically positive significant trend (Figure \ref{fig:trends}a.1) that extends throughout the troposphere up to about 50~hPa and reaches its maximum value at 100~hPa.
+The seasonal trends (rest of Figure \ref{fig:trends} column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100~hPa maximum is much more defined.
 In Winter and Spring, we detect no statistically significant trend.
@@
 
-By separating the SAM signal in its asymmetric and symmetric parts, we can not only see that these trends are almost entirely due to the symmetric component (column b vs.~column c in Figure \ref{fig:trends}), but in some cases the trends become clearer.
-In Summer, the Asymmetric SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the Full SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in panel a.2 and c.2).
-In Autumn, the Symmetric SAM reveals a statistically significant positive trend in the stratosphere that is not significant using the Full SAM index.
+By separating the SAM signal in its asymmetric and symmetric parts, notonly we can see that these trends are almost entirely due to the symmetric component (column b vs.~column c in Figure \ref{fig:trends}), but in some cases the trends become clearer.
+In Summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \ref{fig:trends}a.2 and c.2).
+In Autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.
 
@@
 \begin{figure}
-\includegraphics{r-squared-trend-1} \caption{Decadal trends of the variance explained by the Asymmetric and Symmetric SAM at each level for the period 1979 -- 2018.
+\includegraphics{r-squared-trend-1} \caption{Linear trends (in percent per decade) of the variance explained by A\nobreakdash-SAM and S\nobreakdash-SAM at each level for the period 1979 -- 2018.
 Shading indicates the 95\% confidence interval.}\label{fig:r-squared-trend}
@@
 \begin{figure*}
-\includegraphics{2d-regr-1} \caption{Regression of geopotential height (meters) at (row 1) 50 hPa and (row 2) 700 hPa with the (column a) Full SAM, (column b) Asymmetric SAM, and (column c) Symmetric SAM for the 1979 -- 2018 period.
-The regression patterns for Asymmetric and Symmetric SAM are the result of one multiple regression using both indices.
+\includegraphics{2d-regr-1} \caption{Regression of geopotential height (meters) at (row 1) 50~hPa and (row 2) 700~hPa with (column a) SAM, (column b) A\nobreakdash-SAM, and (column c) S\nobreakdash-SAM for the 1979 -- 2018 period.
+The regression patterns for A\nobreakdash-SAM and S\nobreakdash-SAM are the result of one multiple regression using both indices.
 Points marked on panel b.2 are the location of the reference points used by \cite{raphael2004} for their Zonal Wave 3 index.
@@
 
-We then computed the spatial regression of geopotential height anomalies on Full, Asymmetric and Symmetric indices at 700 hPa and 50 hPa levels (Figure \ref{fig:2d-regr}).
-While, regression coefficients in column a are computed using the Full SAM, the regression coefficients in columns b and c of Figure \ref{fig:2d-regr} are computed using multiple regression using the Asymmetric and Symmetric indices at the same time.
-Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other index.
+We then computed the spatial regression of geopotential height anomalies on the A\nobreakdash-SAM and S\nobreakdash-SAM indices at 700~hPa and 50~hPa levels (Figure \ref{fig:2d-regr}).
+While regression coefficients in Figure \ref{fig:2d-regr} column a are computed using SAM, the regression coefficients in columns b and c of Figure \ref{fig:2d-regr} are computed using multiple regression using the A\nobreakdash-SAM and S\nobreakdash-SAM indices at the same time.
+Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other one.
 
-In the stratosphere, the spatial pattern associated with the Full SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \ref{fig:2d-regr}a.1) which is, not centred in the South Pole.
-On the other hand, the monopole associated with the Symmetric SAM (Figure \ref{fig:2d-regr}c.1) is more symmetric.
-Furthermore, the regression pattern of the Asymmetric SAM is characterized by a wave-1 structure with centers over the Bellinghausen Sea on the
-Western Hemisphere and Davis Sea in the Eastern Hemisphere
+In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \ref{fig:2d-regr}a.1) which is not centred in the South Pole.
+On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \ref{fig:2d-regr}c.1) is more symmetric.
+Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centers over the Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern Hemisphere
 
-In the troposphere, the regression pattern associated with Full SAM shows the well-known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 (Figure \ref{fig:2d-regr}a.2, \citep{fogt2012}).
-The regression patterns associated with the Asymmetric and Symmetric SAM indices successfully disentangle both structures.
-While, the Asymmetric SAM index gives rise to a cleaner zonal wave (Figure \ref{fig:2d-regr}b.2), the Symmetric SAM index is associated with an annular structure, almost devoid of zonal asymmetries (Figure \ref{fig:2d-regr}c.2).
-The wave-3 pattern observed in Figure \ref{fig:2d-regr}b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern described by Raphael (2004), whose reference locations are marked with points in the figure.
-Thus, the tropospheric Asymmetric SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
+In the troposphere, the regression pattern associated with the SAM shows the well-known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 (Figure \ref{fig:2d-regr}a.2, \citep{fogt2012}).
+The regression patterns associated with the A\nobreakdash-SAM and S\nobreakdash-SAM indices successfully disentangle both structures.
+While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \ref{fig:2d-regr}b.2), the S\nobreakdash-SAM index is associated with an annular structure, almost devoid of zonal asymmetries (Figure \ref{fig:2d-regr}c.2).
+The wave-3 pattern observed in Figure \ref{fig:2d-regr}b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern described by \citet{raphael2004}, whose reference locations are marked with points in the figure.
+Indeed, there is no correlation between \citet{raphael2004}'s index and A\nobreakdash-SAM (cor = 0.0076, p-value = 0.87).
+Thus, the tropospheric A\nobreakdash-SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
 
@@
 
-The amplitude of the first zonal wave numbers at each latitude at 50 hPa and 700 hPa is shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
-Zonal wave amplitudes of the spatial pattern described by the Full SAM index (Figure \ref{fig:wave-amplitude} column a) are dominated by the zonal mean (wavenumber 0) at both levels.
-However, zonal waves are important, particularly North of 50\degree S, with wavenumber 1 clearly dominating at 50 hPa (Figure \ref{fig:wave-amplitude}a.1) and a mix of waves of similar amplitude at 700 hPa (Figure \ref{fig:wave-amplitude}a.2).
-Figure \ref{fig:wave-amplitude} column b shows that the Asymmetric SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \ref{fig:wave-amplitude}b.1), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \ref{fig:wave-amplitude}b.2) with negligible amplitude of the zonal mean.
-On the other hand, the Symmetric SAM it is almost entirely explained by the zonal mean at both levels (Figure \ref{fig:wave-amplitude} column c), with little to no contribution from the zonal waves with wavenumbers 1 to 3.
+The amplitude of the first zonal wave numbers at each latitude at 50~hPa and 700~hPa are shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
+Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \ref{fig:wave-amplitude} column a) are dominated by the zonal mean (wavenumber 0) at both levels.
+However, zonal waves are important, particularly South of 50\degree S, with wavenumber 1 clearly dominating at 50~hPa (Figure \ref{fig:wave-amplitude}a.1) and a mix of waves of similar amplitude at 700~hPa (Figure \ref{fig:wave-amplitude}a.2).
+Figure \ref{fig:wave-amplitude} column b shows that the A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \ref{fig:wave-amplitude}b.1), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \ref{fig:wave-amplitude}b.2) with negligible amplitude of the zonal mean.
+On the other hand, the S\nobreakdash-SAM it is almost entirely explained by the zonal mean at both levels (Figure \ref{fig:wave-amplitude} column c), with little to no contribution from zonal waves with wavenumbers 1 to 3.
 
@@
 \begin{figure}
-\includegraphics{vertical-regression-1} \caption{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the Asymmetric SAM index (extracted from multiple regression including the Symmetric SAM).
-(a) With the Asymmetric SAM in 50 hPa and (b) in 700 hPa for the 1979 -- 2018 period.}\label{fig:vertical-regression}
+\includegraphics{vertical-regression-1} \caption{Regression between monthly geopotential height anomalies (meters) averaged betweeen 65\degree and 40\degree S and the A\nobreakdash-SAM index (extracted from a multiple regression which included the S\nobreakdash-SAM index).
+(a) With the A\nobreakdash-SAM in 50~hPa and (b) in 700~hPa for the 1979 -- 2018 period.}\label{fig:vertical-regression}
 \end{figure}
 
-To analyse the vertical structure of the geopotential height anomalies associated with the asymmetric SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa Asymmetric SAM index (Figure \ref{fig:vertical-regression}a) and for the 700 hPa Asymmetric SAM index (Figure \ref{fig:vertical-regression}b).
-The geopotential height anomalies associated with the stratospheric Asymmetric SAM (Figure \ref{fig:vertical-regression}a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric Asymmetric SAM.
-The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes.
-The signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
+To analyse the vertical structure of the geopotential height anomalies associated with the A\nobreakdash-SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50~hPa A\nobreakdash-SAM index (Figure \ref{fig:vertical-regression}a) and for the 700~hPa A\nobreakdash-SAM index (Figure \ref{fig:vertical-regression}b).
+The geopotential height anomalies associated with the stratospheric A\nobreakdash-SAM (Figure \ref{fig:vertical-regression}a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric A\nobreakdash-SAM.
+The vertical structure of this signal tilts about 60\degree~to the West between 100~hPa and 1~hPa, suggesting baroclinic processes.
+The signal in the stratosphere maximises near 10~hPa despite using the 50~hPa index for the regression.
 
-The tropospheric Asymmetric SAM (Figure \ref{fig:vertical-regression}b) has significant signals that extend upwards to the uppermost levels considered.
-In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
-The anomalies are larger in the Western hemisphere, where they extent into the stratosphere.
+The tropospheric A\nobreakdash-SAM (Figure \ref{fig:vertical-regression}b) has significant signals that extend upwards to the uppermost levels considered.
+In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250~hPa.
+The anomalies are larger in the Western hemisphere, where they extend into the stratosphere.
 In the Eastern hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere.
-So, while the tropospheric Asymmetric SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric Asymmetric SAM.
-The structures shown in Figure \ref{fig:vertical-regression}a Figure \ref{fig:vertical-regression}b in Figure \ref{fig:vertical-regression} are robust to the choice of index level.
-For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in Figure \ref{fig:vertical-regression}a.
-Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to Figure \ref{fig:vertical-regression}b.
+So, while the tropospheric A\nobreakdash-SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric A\nobreakdash-SAM.
+The structures shown in Figure \ref{fig:vertical-regression} are robust to the choice of index level.
+For any stratospheric (above 100~hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10~hPa in Figure \ref{fig:vertical-regression}a.
+Conversely, for any tropospheric (below 100~hPa) index, the result is very similar to Figure \ref{fig:vertical-regression}b.
 The patterns mainly change in amplitude (not shown).
@@
 Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the Full SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is -0.16 (p-value \textless{} 0.001).
-This relationship is captured mainly by the Asymmetric SAM, as this index has a partial correlation of -0.26 (p-value \textless{} 0.001) with the ONI, consistent with \citet{fan2007}'s results, whereas the Symmetric SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67).
+The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is -0.16 (p-value \textless{} 0.001).
+This relationship is captured mainly by the A\nobreakdash-SAM, as this index has a partial correlation of -0.26 (p-value \textless{} 0.001) with the ONI, consistent with \citet{fan2007}'s results, whereas the S\nobreakdash-SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67).
 The same analysis was performed using the Multivariate ENSO Index \citep{wolter2011} and the Southern Oscillation Index \citep{ropelewski1987}, obtaining similar results.
@@
 \begin{figure*}
-\includegraphics{regr-air-season-1} \caption{Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with Asymmetric SAM and Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{regr-air-season-1} \caption{Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with SAM, A\nobreakdash-SAM and S\nobreakdash-SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Gray areas in Antarctica are areas with have more than 15\% of missing data.}\label{fig:regr-air-season}
+Gray areas in Antarctica have more than 15\% of missing data.}\label{fig:regr-air-season}
 \end{figure*}
 
-To assess the differences in the impacts associated with full, asymmetric and symmetric SAM, we regress s surface land air and sea temperature and precipitation onto each of the three SAM indices at 700 hPa.
+To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices we regressed surface land air and sea temperature and precipitation onto each of the three SAM indices at 700~hPa.
 As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
 
-Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700 hPa with surface land air and sea temperature for each trimester.
-In Summer, positive values of the Full SAM index (Figure \ref{fig:regr-air-season}a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
+Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700~hPa with surface land air and sea temperature for each trimester.
+In Summer, positive values of the SAM index (Figure \ref{fig:regr-air-season}a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
 The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E respectively.
 In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
-Figure \ref{fig:regr-air-season}b.1 and c.1 show temperature anomalies associated with positive values of the Asymmetric and Symmetric SAM, respectively.
-Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the Asymmetric SAM regression map, while temperature patterns linked to positive Symmetric SAM show a more zonally consistent ring and less relation to the tropics.
-Temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to Asymmetric SAM.
-This signal is not present in the regression pattern with the Full SAM.
-Spring (Figure \ref{fig:regr-air-season}, row 4) features similar patterns but of smaller magnitude, with less regions where regressed anomalies have statistical significance.
+Figure \ref{fig:regr-air-season}b.1 and c.1 show temperature anomalies associated with positive values of A\nobreakdash-SAM and S\nobreakdash-SAM, respectively.
+Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics.
+Temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM.
+This signal is not present in the regression pattern with the SAM.
+Spring (Figure \ref{fig:regr-air-season}, row 4) features similar patterns but of smaller magnitude, with fewer regions where regressed anomalies have statistical significance.
 
-In Autumn and Winter (rows 2 and 3 in Figure \ref{fig:regr-air-season}) the positive ring is only present through its local maximums in the regression with the Full SAM.
+In Autumn and Winter (rows 2 and 3 in Figure \ref{fig:regr-air-season}) the positive ring is only present through its local maximums in the regression with the SAM.
 There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
@@
 
-Regression of the SAM indices with seasonal mean precipitation and 700 hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
+Regression of the SAM indices with seasonal mean precipitation and 700~hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
 South Africa is not shown because no significant signal was detected there.
@@
 \begin{figure*}
-\includegraphics{pp-regr-oceania-1} \caption{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700 hPa geopotential height anomalies (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) Full SAM, (column (b) Asymmetric SAM and (column c) Symmetric SAM for the 1979 -- 2018 period.
+\includegraphics{pp-regr-oceania-1} \caption{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700~hPa geopotential height anomalies (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) SAM, (column (b) A\nobreakdash-SAM and (column c) S\nobreakdash-SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
@@
 
-In Australia, the annual regression shows that the Full SAM index is associated with positive precipitation anomalies in the southeastern region (Figure \ref{fig:pp-regr-oceania}a.1), in agreement with \citet{gillett2006}.
-The separation between Asymmetric and Symmetric SAM suggest that this positive anomaly is explained by the Symmetric SAM only in the East coast (Figure \ref{fig:pp-regr-oceania}c.1).
+In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the southeastern region (Figure \ref{fig:pp-regr-oceania}a.1), in agreement with \citet{gillett2006}.
+The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure \ref{fig:pp-regr-oceania}c.1).
 Geopotential height anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by \citet{hendon2007}.
-The Asymmetric SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \ref{fig:pp-regr-oceania}b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
+A\nobreakdash-SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \ref{fig:pp-regr-oceania}b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
 
-The seasonal-based regressions show statistically significant anomalies only in Spring, when positive Full SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \ref{fig:pp-regr-oceania}a.5).
-In this trimester, the Symmetric SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \ref{fig:pp-regr-oceania}c.5) while the positive precipitation anomalies related with positive Asymmetric SAM affect all Eastern Australia (Figure \ref{fig:pp-regr-oceania}b.5).
+The seasonal-based regressions show statistically significant anomalies only in Spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \ref{fig:pp-regr-oceania}a.5).
+In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \ref{fig:pp-regr-oceania}c.5) while the positive precipitation anomalies related with positive A\nobreakdash-SAM affect all Eastern Australia (Figure \ref{fig:pp-regr-oceania}b.5).
 
-In Summer, positive Full SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \ref{fig:pp-regr-oceania}a.2).
-The Eastern part being dominated by the relationship with the Symmetric SAM and the Western, by the Asymmetric SAM.
-In Autumn, the regression with Full SAM shows positive anomalies in the North, similar to Summer, and a broad area of positive anomalies in the North-East to South-West direction.
-This structure seems to be associated with the Symmetric SAM, while the Northern positive values are associated with the Asymmetric SAM.
-In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the Symmetric SAM.
+In Summer, positive SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \ref{fig:pp-regr-oceania}a.2).
+The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
+In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, and a broad area of positive anomalies in the North-East to South-West direction.
+This structure seems to be associated with the S\nobreakdash-SAM, while the Northern positive values are associated with the A\nobreakdash-SAM.
+In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the S\nobreakdash-SAM.
 None of these regression coefficients are statistically significant at the 95\% level.
@@
 
-In South America (Figure \ref{fig:pp-regr-america}), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \ref{fig:pp-regr-america}a.1).
-Figure \ref{fig:pp-regr-america}b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with the Asymmetric SAM, that in southern Chile is related to the Symmetric SAM.
+In South America (Figure \ref{fig:pp-regr-america}), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile, and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \ref{fig:pp-regr-america}a.1).
+Figure \ref{fig:pp-regr-america}b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with A\nobreakdash-SAM, that in southern Chile is related to S\nobreakdash-SAM.
 
 Except Winter, seasonal-based regressions mirror this same pattern.
-Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the Full SAM.
-The separation of these features between the Asymmetric SAM and Symmetric SAM regression maps is also rather consistent.
+Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the SAM.
+The separation of these features between A\nobreakdash-SAM and S\nobreakdash-SAM regression maps is also rather consistent.
 
-The anomalous circulation at 700 hPa associated with the Symmetric SAM (Figure \ref{fig:pp-regr-america}c.1) indicate anomalous easterly flow over Southern Chile.
-This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region \citep[e.g.][]{garreaud2007}.
-On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (Figure \ref{fig:pp-regr-america}b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
+The anomalous circulation at 700~hPa associated with S\nobreakdash-SAM (Figure \ref{fig:pp-regr-america}c.1) indicate anomalous easterly flow over Southern Chile.
+This leads to reduced influx of moist air from the Pacific Ocean which is the main source of precipitable water in that region \citep[e.g.][]{garreaud2007}.
+On the other hand, the anomalous circulation associated with positive values of A\nobreakdash-SAM (Figure \ref{fig:pp-regr-america}b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
 This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region \citep{silvestri2009, zamboni2010}.
 This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events \citep{rosso2018}.
-There is a small area of significative positive precipitation anomalies with SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} that is explained by the Asymmetric SAM.
+There is a small area of significative positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} that is explained by the A\nobreakdash-SAM.
 
@@
 
-The Asymmetric SAM index correlates strongly with the Symmetric SAM index.
-In the troposphere, this correlation is maximum at zero lag, while in the stratosphere is maximised with the Asymmetric SAM leading the Symmetric SAM by one month.
+The A\nobreakdash-SAM index correlates strongly with the S\nobreakdash-SAM index.
+In the troposphere, this correlation is maximum at zero lag, while in the stratosphere is maximised with the A\nobreakdash-SAM leading the S\nobreakdash-SAM by one month.
 Since most indices of the SAM are calculated using surface or near-surface conditions, this result would suggest that they might not be sensitive to the most dramatic changes in SAM variability.
 
-The two-year periodicity we found in the stratospheric Symmetric SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
-There is evidence of influence between the QBO and the Northern Annular Mode \citep[e.g.][]{holton1980, watson2014, zhang2020}, so it is not unlikely that the SAM would be similarly connected.
+The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
+There is evidence of influence between the QBO and the Northern Annular Mode \citep[e.g.][]{holton1980, watson2014, zhang2020}, so it is not unlikely that the SAM would be similarly affected.
 However, establishing this link would require further research.
-We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies {[}e.g.
-\citet{fogt2020}; and references therein{]} for low levels.
-We show that these trends maximise at 100 hPa, and are explained by the zonally symmetric component.
-We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the Full SAM index.
-In contrast to \citet{fogt2012} who made the evaluation for the period 158 -- 2001, we find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the Asymmetric SAM explains an increasingly proportion of the total variance.
-This might be due either due to differneces in methodology or analysed period.
+We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies \citep[e.g.][ and references therein]{fogt2020} for low levels.
+We show that these trends maximise at 100~hPa, and are explained by the zonally symmetric component.
+We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the SAM index.
+We find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
+This is in contrast to \citet{fogt2012} who made the evaluation for the period 158 -- 2001.
+This discrepancy might be due either due to differneces in methodology or analysed period.
 
-In the troposphere, the spatial patterns of geopotential associated with the Symmetric SAM are much closer to being fully annular than the patterns associated with the Full SAM index.
-The Asymmetric SAM, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific This pattern extends in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimum for capturing this variability.
+In the troposphere, the spatial patterns of geopotential associated with the S\nobreakdash-SAM index are much closer to being fully annular than the patterns associated with the SAM index.
+The A\nobreakdash-SAM index, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific.
+This pattern extends vertically in the troposphere but its maximum is located at 250~hPa, which also could suggest that surface-based indices are not opitimal for capturing this variability.
 This wave-3 pattern is similar to the Pacific-South American Pattern, linked to ENSO variability.
-We found that the significant correlation that exists between the Full SAM index and the Oceanic Niño Index over this period, is captured entirely by the Asymmetric SAM index.
-This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the Asymmetric SAM index could be a useful measure to further study that relationship.
+We found that the significant correlation that exists between the SAM index and the Oceanic Niño Index over this period is captured entirely by the A\nobreakdash-SAM index.
+This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the A\nobreakdash-SAM index could be a useful measure to further study that relationship.
 
-Temperature anomalies associated with the Full SAM broadly show a pattern of negative anomalies at polar latitudes surrounded by positive anomalies, but with many deviations from symmetry.
-The Asymmetric SAM index explains a big portion of these deviations.
-In particular, the positive phase of the Asymmetric SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO-SAM relationship.
+Temperature anomalies associated with the SAM broadly show a pattern of negative anomalies at polar latitudes surrounded by positive anomalies, but with many deviations from symmetry.
+The A\nobreakdash-SAM index explains a big portion of these deviations.
+In particular, the positive phase of A\nobreakdash-SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO\nobreakdash-SAM relationship.
 These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active \citep[e.g.][]{cai2020a}.
 
-In Australia the Full SAM is associated with positive precipitation anomalies in South East and this is explained by the Symmetric SAM.
-However, the Asymmetric SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
-In South America, precipitation anomalies associated with the Full SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
-These features are cleanly separated between the Asymmetric and Symmetric components.
-The Symmetric SAM explains the negative anomalies in Southern Chile and the Asymmetric SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
+In Australia the SAM is associated with positive precipitation anomalies in South East and this is explained by the S\nobreakdash-SAM.
+However, the A\nobreakdash-SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
+In South America, precipitation anomalies associated with the SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
+These features are cleanly separated between A\nobreakdash-SAM and S\nobreakdash-SAM.
+S\nobreakdash-SAM explains the negative anomalies in Southern Chile and A\nobreakdash-SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
 Individual seasons mostly follow this pattern.
@@
 The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s \citep{fogt2006, clem2013}.
-Seeing as both the ENSO-SAM relationship and most of the precipitation impacts in South America are captured by the Asymmetric SAM, the results presented here are most likely period-dependent.
-The decadal variations of the Asymmetric SAM should be focus of future studies.
+Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the results presented here are most likely period-dependent.
+The decadal variations of the A\nobreakdash-SAM should be focus of future studies.
 
@@
 \begin{figure}[ht]
-\includegraphics{A1-1} \caption{Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level.
-Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa.
+\includegraphics{A1-1} \caption{Lag-correlation between the A\nobreakdash-SAM and the S\nobreakdash-SAM index at each level.
+Negative lags imply A\nobreakdash-SAM leading S\nobreakdash-SAM and vice versa.
 For the 1979 -- 2018 period.}\label{fig:A1}
@@
 \includegraphics{A2-1} \caption{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
-The shading indicates de 95\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\% of the simulated sampels had an amplitude equal or lower).
+The shading indicates de 95\% confidence area derived by computing the spectrum for 5000 simulated samples from a fitted autorregressive model and (95\% of the simulated sampels had an amplitude equal or lower).
 The light line indicates the theoretical expected amplitude from the autorregressive model.
@@
 \begin{figure}
-\includegraphics{A3-1} \caption{50 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
-Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A3}
+\includegraphics{A3-1} \caption{50~hPa geopotential height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
+Numbers in the column headers are the spatial correlation between SAM+ and SAM- composites and number of monthly fields used to construct each composite.}\label{fig:A3}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A4-1} \caption{700 hPa Geopotetnial height zonal anomalies (meters) of composites of positive and negative SAM months selected using $\pm1$ standard deviation as threshhold for the 1979 -- 2018 period.
-Numbers in the column headers are pattern correlation between SAM+ and SAM- composites and number of monthly fields used to construct the composites.}\label{fig:A4}
+\includegraphics{A4-1} \caption{Same as \ref{fig:A3} but for 700~hPa geopotential height.}\label{fig:A4}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A5-1} \caption{Regression coefficients of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
+\includegraphics{A5-1} \caption{Regression coefficients of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A6-1} \caption{Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
-Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.}\label{fig:A6}
+\includegraphics{A6-1} \caption{Regression of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
+Spatial correlation between both fields is 0.86 for the 50~hPa fields and 0.76 for the 700~hPa fields.}\label{fig:A6}
 \end{figure}

### 2020-12-11 analysis/paper/paper.Rmd
JA! ¿Quién te creía que el anterior commit era la última revisión? (Esta sí)

@@
 We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices.
-In the troposphere, the Asymmetric SAM, is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure.
-The Symmetric SAM, on the other hand, is associated with a negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
+In the troposphere, the Asymmetric SAM (A\nobreakdash-SAM), is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure.
+The Symmetric SAM (S\nobreakdash-SAM), on the other hand, is associated with a negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
 The observed relationship between the El Niño Southern Oscillation and the SAM is explained fully by the A\nobreakdash-SAM index.

### 2020-12-11 analysis/paper/paper.tex
Declaraciones

@@
 
-\hypertarget{data-statement}{%
-\subsection{Data statement}\label{data-statement}}
+\hypertarget{declarations}{%
+\section*{Declarations}\label{declarations}}
+\addcontentsline{toc}{section}{Declarations}
+
+\hypertarget{funding}{%
+\subsection*{Funding}\label{funding}}
+\addcontentsline{toc}{subsection}{Funding}
+
+The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+
+\hypertarget{conflicts-of-interest}{%
+\subsection*{Conflicts of interest}\label{conflicts-of-interest}}
+\addcontentsline{toc}{subsection}{Conflicts of interest}
+
+The authors declare that they have no conflict of interest.
+
+\hypertarget{availability-of-data-and-material}{%
+\subsection*{Availability of data and material}\label{availability-of-data-and-material}}
+\addcontentsline{toc}{subsection}{Availability of data and material}
 
@@
 The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}
+
+\hypertarget{code-availability}{%
+\subsection*{Code availability}\label{code-availability}}
+\addcontentsline{toc}{subsection}{Code availability}
 

### 2020-12-15 analysis/paper/paper.Rmd
Declaraciones
Correcciones de Leandro

--- 2020-12-11 analysis/paper/paper.Rmd
@@
 abstract: |
-  The main mode of variability in the Southern Hemisphere extratropical circulation is the Southern Annular Mode (SAM); so called because of it's zonally symmetric ring-like shape.
+  The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extra-tropical circulation and it is so called because of its zonally symmetric ring-like shape.
 However, the SAM pattern actually contains noticeable deviations from zonal symmetry.
-We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices.
-In the troposphere, the Asymmetric SAM, is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure.
-The Symmetric SAM, on the other hand, is associated with a negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
-The observed relationship between the El Niño Southern Oscillation and the SAM is explained fully by the A\nobreakdash-SAM index.
-The positive trend of the SAM is present only in its asymmetric component.
+Thus, the purpose of this study is to describe the zonally asymmetric and symmetric components of the SAM variability and their impacts.
+We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices: Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM).
+In the troposphere, the A-SAM is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure.
+On the other hand, the S-SAM is associated with negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
+The observed relationship between the El Niño Southern Oscillation and the SAM is fully explained by the A-SAM index.
+The positive trend of the SAM is present only in its symmetric component.
 Despite this, the SAM is becoming more zonally asymmetric.
@@
 @fogt2020 and references therein].
-It is though that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013].
+It is thought that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013].
 However, it’s not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.
@@
 The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
-For example, the SAM-precipitation relationshipin Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
+For example, the SAM-precipitation relationship in Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
 @fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
@@
 We first propose a methodology that provides for each level, two indices which aim to capture independently the variability of the symmetric and asymmetric SAM component respectively.
-Their vertical structure and coherence, temporal variability and trends are consequently assesed.
+Their vertical structure and coherence, temporal variability and trends are consequently assessed.
 We then study the spatial patterns described by the variability exclusive to each index focusing on 50\ hPa as representing the stratosphere and 700\ hPa as representing the troposphere.
@@
 
-We computed density probability estimates using a gaussian kernel of optimal bindwidth according to @sheather1991.
+We computed density probability estimates using a Gaussian kernel of optimal bin width according to @sheather1991.
 
@@
 Therefore, stratospheric A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index.
+(Correlations at lags -5 to 5 are shown in Figure \@ref(fig:A1).)
 
@@
 
-By separating the SAM signal in its asymmetric and symmetric parts, notonly we can see that these trends are almost entirely due to the symmetric component (column b vs.
+By separating the SAM signal in its asymmetric and symmetric parts, not only we can see that these trends are almost entirely due to the symmetric component (column b vs.
 column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer.
@@
 On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric.
-Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centers over the Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern Hemisphere
+Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern Hemisphere
 
@@
   geom_line(aes(color = factor(k))) +
-  scale_color_brewer("Wavenumber", palette = "Dark2") +
+  scale_color_brewer("wavenumber", palette = "Dark2") +
   scale_y_continuous("Amplitude") +
@@
 
-The amplitude of the first zonal wave numbers at each latitude at 50\ hPa and 700\ hPa are shown in Figure \@ref(fig:wave-amplitude), where wave number zero represents the amplitude of the zonal mean.
+The amplitude of the first zonal wavenumbers at each latitude at 50\ hPa and 700\ hPa are shown in Figure \@ref(fig:wave-amplitude), where wavenumber zero represents the amplitude of the zonal mean.
 Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column `r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at both levels.
@@
                    minor_breaks = NULL) +
-  scale_fill_divergent("2-meter air tempreature",  
+  scale_fill_divergent("2-meter air temperature",  
                        limits = c(-0.6, .6), 
@@
 
-Regression of the SAM indices with seasonal mean precipitation and 700\ hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for Australasia and South America respectively.
+Regression of the SAM indices with seasonal mean precipitation and 700\ hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for New Zealand and neighbouring islands, and South America respectively.
 South Africa is not shown because no significant signal was detected there.
@@
 
-In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the southeastern region (Figure \@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006.
+In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the Southeastern region (Figure \@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006.
 The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure \@ref(fig:pp-regr-oceania)c.1).
@@
 This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events [@rosso2018].
-There is a small area of significative positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the A\nobreakdash-SAM.
+There is a small area of significant positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the A\nobreakdash-SAM.
 
@@
 However, establishing this link would require further research.
+
 We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies [e.g.
@@
 We find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
-This is in contrast to @fogt2012 who made the evaluation for the period 158 -- 2001.
-This discrepancy might be due either due to differneces in methodology or analysed period.
+This is in contrast to @fogt2012 who made the evaluation for the period 1958 -- 2001.
+This discrepancy might be due either due to differences in methodology or analysed period.
 
@@
 The A\nobreakdash-SAM index, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific.
-This pattern extends vertically in the troposphere but its maximum is located at 250\ hPa, which also could suggest that surface-based indices are not opitimal for capturing this variability.
+This pattern extends vertically in the troposphere but its maximum is located at 250\ hPa, which also could suggest that surface-based indices are not optimal for capturing this variability.
 This wave-3 pattern is similar to the Pacific-South American Pattern, linked to ENSO variability.
@@
 By separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
-This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component [@arblaster2006].
+This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modelled as highly zonally symmetric, while the impact of increased concentration of greenhouse gases has also a zonally asymmetric component [@arblaster2006].
 
@@
 
-## Data statement
+# Declarations {-}
+
+## Funding {-}
+
+The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+
+## Conflicts of interest {-}
+
+The authors declare that they have no conflict of interest.
+
+## Availability of data and material {-}
 
@@
 ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means).
-NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html).
-The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt
+NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sciences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html).
+The Oceanic Niño Index is available via NOAA's Climate Prediction Center: https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt
+
+## Code availability {-}
 
@@
 \newpage
-
-```{r A1, fig.cap = "Lag-correlation between the A\\nobreakdash-SAM and the S\\nobreakdash-SAM index at each level.
-Negative lags imply A\\nobreakdash-SAM leading S\\nobreakdash-SAM and vice versa.
-For the 1979 -- 2018 period.", fig.width = 5, fig.height=5, fig.pos="ht"}
-ccf_dt <- function(..., plot = FALSE) {
-  stat <- ccf(..., plot = plot)
-  list(lag = stat$lag[, 1, 1], 
-       acf = stat$acf[, 1, 1])
-}
-
-indices_wide %>% 
-  .[, ccf_dt(asym, sym, lag.max = 5), by = .(lev)] %>% 
-  ggplot(aes(lag, lev)) +
-  geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
-  geom_contour_tanaka(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
-  geom_vline(xintercept = 0) +
-  scale_fill_divergent(NULL,
-                       guide = guide_colorstrip_bottom(20),
-                       breaks = AnchorBreaks(0, 0.05),
-                       limits = c(NA, .59)) +
-  scale_x_continuous("Lag (months)\nSymmetric leads Asymmetric\t \tAsymmetric leads Symmetric", 
-                     breaks = seq(-5, 5, 1), 
-                     expand = c(0, 0)) +
-  scale_y_level()
-```
-
-```{r compute-spectrums}
-indices %>% 
-  .[lev %in% c(50, 700)] %>% 
-  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
-  .[, fftspectrum(estimate, c(3, 5), B = 5000, probs = c(0, 0.95)), by = .(term, lev)] -> spec
-```
-
-```{r A2, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
-The shading indicates de 95\\% confidence area derived by computing the spectrum for 5000 simulated samples from a fitted autorregressive model and (95\\% of the simulated sampels had an amplitude equal or lower).
-The light line indicates the theoretical expected amplitude from the autorregressive model.
-For the 1979 -- 2018 period.", fig.height=4, fig.width=6}
-spec %>%
-  # .[term != "full"] %>% 
-  ggplot(aes(1/freq/12, spec)) +
-  geom_ribbon(aes(ymin = 0, ymax = `95%`), position = "dodge", alpha = 0.1) +
-  geom_line(aes(y = ar_spectrum), alpha = 0.15) +
-  geom_line() +
-  annotation_logticks(sides = "b") +
-  scale_color_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill")) +
-  scale_x_log10("Period (years)", sec.axis = sec_axis(~.*12, name = "Period (months)")) +
-  scale_y_continuous("Spectrum") +
-  facet_grid(lev ~ term,labeller = labeller(lev = lev.lab, term = lab_sam), scales = "free") +
-  tag_facets() 
-```
 
@@
 
+```{r A1, fig.cap = "Lag-correlation between the A\\nobreakdash-SAM and the S\\nobreakdash-SAM index at each level.
+Negative lags imply A\\nobreakdash-SAM leading S\\nobreakdash-SAM and vice versa.
+For the 1979 -- 2018 period.", fig.width = 5, fig.height=5, fig.pos="ht"}
+ccf_dt <- function(..., plot = FALSE) {
+  stat <- ccf(..., plot = plot)
+  list(lag = stat$lag[, 1, 1], 
+       acf = stat$acf[, 1, 1])
+}
+
+indices_wide %>% 
+  .[, ccf_dt(asym, sym, lag.max = 5), by = .(lev)] %>% 
+  ggplot(aes(lag, lev)) +
+  geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
+  geom_contour_tanaka(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
+  geom_vline(xintercept = 0) +
+  scale_fill_divergent(NULL,
+                       guide = guide_colorstrip_bottom(20),
+                       breaks = AnchorBreaks(0, 0.05),
+                       limits = c(NA, .59)) +
+  scale_x_continuous("Lag (months)\nA-SAM leads S-SAM              S-SAM leads A-SAM", 
+                     breaks = seq(-5, 5, 1), 
+                     expand = c(0, 0)) +
+  scale_y_level()
+```
+
+```{r compute-spectrums}
+indices %>% 
+  .[lev %in% c(50, 700)] %>% 
+  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
+  .[, fftspectrum(estimate, c(3, 5), B = 5000, probs = c(0, 0.95)), by = .(term, lev)] -> spec
+```
+
+```{r A2, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
+The shading indicates de 95\\% confidence area derived by computing the spectrum for 5000 simulated samples from a fitted autorregressive model and (95\\% of the simulated sampels had an amplitude equal or lower).
+The light line indicates the theoretical expected amplitude from the autorregressive model.
+For the 1979 -- 2018 period.", fig.height=4, fig.width=6}
+spec %>%
+  # .[term != "full"] %>% 
+  ggplot(aes(1/freq/12, spec)) +
+  geom_ribbon(aes(ymin = 0, ymax = `95%`), position = "dodge", alpha = 0.1) +
+  geom_line(aes(y = ar_spectrum), alpha = 0.15) +
+  geom_line() +
+  annotation_logticks(sides = "b") +
+  scale_color_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill")) +
+  scale_x_log10("Period (years)", sec.axis = sec_axis(~.*12, name = "Period (months)")) +
+  scale_y_continuous("Spectrum") +
+  facet_grid(lev ~ term,labeller = labeller(lev = lev.lab, term = lab_sam), scales = "free") +
+  tag_facets() 
+```
+
 \newpage

### 2020-12-15 analysis/paper/paper.tex
Correcciones de Leandro

@@
 \begin{abstract}
-The main mode of variability in the Southern Hemisphere extratropical circulation is the Southern Annular Mode (SAM); so called because of it's zonally symmetric ring-like shape.
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extra-tropical circulation and it is so called because of its zonally symmetric ring-like shape.
 However, the SAM pattern actually contains noticeable deviations from zonal symmetry.
-We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices.
-In the troposphere, the Asymmetric SAM (A\nobreakdash-SAM), is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure.
-The Symmetric SAM (S\nobreakdash-SAM), on the other hand, is associated with a negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
-The observed relationship between the El Niño Southern Oscillation and the SAM is explained fully by the A\nobreakdash-SAM index.
-The positive trend of the SAM is present only in its asymmetric component.
+Thus, the purpose of this study is to describe the zonally asymmetric and symmetric components of the SAM variability and their impacts.
+We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices: Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM).
+In the troposphere, the A-SAM is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure.
+On the other hand, the S-SAM is associated with negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
+The observed relationship between the El Niño Southern Oscillation and the SAM is fully explained by the A-SAM index.
+The positive trend of the SAM is present only in its symmetric component.
 Despite this, the SAM is becoming more zonally asymmetric.
@@
 Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn \citep[e.g.][ and references therein]{fogt2020}.
-It is though that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables \citep{marshall2004, gillett2005, arblaster2006, gillett2013}.
+It is thought that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables \citep{marshall2004, gillett2005, arblaster2006, gillett2013}.
 However, it's not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.
@@
 The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
-For example, the SAM-precipitation relationshipin Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}.
+For example, the SAM-precipitation relationship in Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}.
 \citet{fan2007} also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
@@
 We first propose a methodology that provides for each level, two indices which aim to capture independently the variability of the symmetric and asymmetric SAM component respectively.
-Their vertical structure and coherence, temporal variability and trends are consequently assesed.
+Their vertical structure and coherence, temporal variability and trends are consequently assessed.
 We then study the spatial patterns described by the variability exclusive to each index focusing on 50~hPa as representing the stratosphere and 700~hPa as representing the troposphere.
@@
 
-We computed density probability estimates using a gaussian kernel of optimal bindwidth according to \citet{sheather1991}.
+We computed density probability estimates using a Gaussian kernel of optimal bin width according to \citet{sheather1991}.
 
@@
 Therefore, stratospheric A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index.
+(Correlations at lags -5 to 5 are shown in Figure \ref{fig:A1}.)
 
@@
 
-By separating the SAM signal in its asymmetric and symmetric parts, notonly we can see that these trends are almost entirely due to the symmetric component (column b vs.~column c in Figure \ref{fig:trends}), but in some cases the trends become clearer.
+By separating the SAM signal in its asymmetric and symmetric parts, not only we can see that these trends are almost entirely due to the symmetric component (column b vs.~column c in Figure \ref{fig:trends}), but in some cases the trends become clearer.
 In Summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \ref{fig:trends}a.2 and c.2).
@@
 On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \ref{fig:2d-regr}c.1) is more symmetric.
-Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centers over the Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern Hemisphere
+Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern Hemisphere
 
@@
 
-The amplitude of the first zonal wave numbers at each latitude at 50~hPa and 700~hPa are shown in Figure \ref{fig:wave-amplitude}, where wave number zero represents the amplitude of the zonal mean.
+The amplitude of the first zonal wavenumbers at each latitude at 50~hPa and 700~hPa are shown in Figure \ref{fig:wave-amplitude}, where wavenumber zero represents the amplitude of the zonal mean.
 Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \ref{fig:wave-amplitude} column a) are dominated by the zonal mean (wavenumber 0) at both levels.
@@
 
-Regression of the SAM indices with seasonal mean precipitation and 700~hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for Australasia and South America respectively.
+Regression of the SAM indices with seasonal mean precipitation and 700~hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for New Zealand and neighbouring islands, and South America respectively.
 South Africa is not shown because no significant signal was detected there.
@@
 
-In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the southeastern region (Figure \ref{fig:pp-regr-oceania}a.1), in agreement with \citet{gillett2006}.
+In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the Southeastern region (Figure \ref{fig:pp-regr-oceania}a.1), in agreement with \citet{gillett2006}.
 The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure \ref{fig:pp-regr-oceania}c.1).
@@
 This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events \citep{rosso2018}.
-There is a small area of significative positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} that is explained by the A\nobreakdash-SAM.
+There is a small area of significant positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} that is explained by the A\nobreakdash-SAM.
 
@@
 However, establishing this link would require further research.
+
 We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies \citep[e.g.][ and references therein]{fogt2020} for low levels.
@@
 We find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
-This is in contrast to \citet{fogt2012} who made the evaluation for the period 158 -- 2001.
-This discrepancy might be due either due to differneces in methodology or analysed period.
+This is in contrast to \citet{fogt2012} who made the evaluation for the period 1958 -- 2001.
+This discrepancy might be due either due to differences in methodology or analysed period.
 
@@
 The A\nobreakdash-SAM index, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific.
-This pattern extends vertically in the troposphere but its maximum is located at 250~hPa, which also could suggest that surface-based indices are not opitimal for capturing this variability.
+This pattern extends vertically in the troposphere but its maximum is located at 250~hPa, which also could suggest that surface-based indices are not optimal for capturing this variability.
 This wave-3 pattern is similar to the Pacific-South American Pattern, linked to ENSO variability.
@@
 By separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
-This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modeled as highly zonally symmetric, while the impact of increased concentration of greenhose gases has also a zonally asymmetric component \citep{arblaster2006}.
+This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modelled as highly zonally symmetric, while the impact of increased concentration of greenhouse gases has also a zonally asymmetric component \citep{arblaster2006}.
 
@@
 ERA5 data can be obtained via the Copernicus Climate Data Store (\url{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means}).
-NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (\url{https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html} and \url{https://psl.noaa.gov/data/gridded/data.gpcc.html}).
-The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}
+NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sciences Laboratory website (\url{https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html} and \url{https://psl.noaa.gov/data/gridded/data.gpcc.html}).
+The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}
 
@@
 \newpage
-
-\begin{figure}[ht]
-\includegraphics{A1-1} \caption{Lag-correlation between the A\nobreakdash-SAM and the S\nobreakdash-SAM index at each level.
-Negative lags imply A\nobreakdash-SAM leading S\nobreakdash-SAM and vice versa.
-For the 1979 -- 2018 period.}\label{fig:A1}
-\end{figure}
-
-\begin{figure}
-\includegraphics{A2-1} \caption{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
-The shading indicates de 95\% confidence area derived by computing the spectrum for 5000 simulated samples from a fitted autorregressive model and (95\% of the simulated sampels had an amplitude equal or lower).
-The light line indicates the theoretical expected amplitude from the autorregressive model.
-For the 1979 -- 2018 period.}\label{fig:A2}
-\end{figure}
 
@@
 
+\begin{figure}[ht]
+\includegraphics{A1-1} \caption{Lag-correlation between the A\nobreakdash-SAM and the S\nobreakdash-SAM index at each level.
+Negative lags imply A\nobreakdash-SAM leading S\nobreakdash-SAM and vice versa.
+For the 1979 -- 2018 period.}\label{fig:A1}
+\end{figure}
+
+\begin{figure}
+\includegraphics{A2-1} \caption{Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.
+The shading indicates de 95\% confidence area derived by computing the spectrum for 5000 simulated samples from a fitted autorregressive model and (95\% of the simulated sampels had an amplitude equal or lower).
+The light line indicates the theoretical expected amplitude from the autorregressive model.
+For the 1979 -- 2018 period.}\label{fig:A2}
+\end{figure}
+
 \newpage

### 2020-12-15 analysis/paper/paper.Rmd
Mayúscula en wavenumber

@@
   geom_line(aes(color = factor(k))) +
-  scale_color_brewer("wavenumber", palette = "Dark2") +
+  scale_color_brewer("Wavenumber", palette = "Dark2") +
   scale_y_continuous("Amplitude") +

### 2020-12-15 analysis/pre-print/pre-print.Rmd
Agrega preprint

--- 2020-12-15 analysis/paper/paper.Rmd
@@
 title: Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach
-# subtitle: Do you have a subtitle? If so, write it here
-titlerunning:  Assessment of zonally symmetric and asymmetric components of the SAM
-# authorrunning: Short form of author list if too long for running head
-thanks: | 
-  The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
 
@@
   - name: Elio Campitelli
-  address: |
-    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
     email: elio.campitelli@cima.fcen.uba.ar
-
   - name:  Leandro B. Díaz
-  address: |
-    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
-
   - name:  Carolina Vera
-  address: |
-    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
-
-keywords:
-  - Southern Annular Mode
-  - general circulation
-  - zonally asymmetric circulation
-  - El Niño Southern Oscillation
-
-#PACS: 
-#- PAC1
-#- superPAC
-
-# MSC:
-# - MSC code 1
-# - MSC code 2
 
@@
 
-bibliography: 
-  - AsymSAM
-  - packages
-bibstyle: spbasic
-# bibstyle options spbasic(default), spphys, spmpsci
+keywords:
+  - Southern Annular Mode
+  - general circulation
+  - zonally asymmetric circulation
+  - El Niño Southern Oscillation
+bibliography: "AsymSAM,packages"
+bibstyle: "plainnat"
 output: 
  bookdown::pdf_book: 
-      base_format: rticles::springer_article
+    base_format: rticles::arxiv_article
     latex_engine: "xelatex"
     citation_package: "natbib"
-      # pandoc_args: ["--lua-filter=abstract-to-meta.lua"]
+    keep_tex: true
 header-includes: 
@@
  - \usepackage{subfig}
+ - \usepackage{amsmath}
 #  - \usepackage[inline]{showlabels}
@@
   cache.extra = 41,
-  cache.path = "../cache/",
+  cache.path = "../cache/pre-print",
   collapse = TRUE,
@@
 
-\begin{acknowledgements}
+# Acknowledgements
+
 NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/
-\end{acknowledgements}
 

### 2020-12-15 analysis/pre-print/pre-print.tex
Agrega preprint

--- 2020-12-15 analysis/paper/paper.tex
@@
-% !TeX program = pdfLaTeX
-\documentclass[smallextended]{svjour3}       % onecolumn (second format)
-%\documentclass[twocolumn]{svjour3}          % twocolumn
-%
-\smartqed  % flush right qed marks, e.g. at end of proof
-%
-\usepackage{amsmath}
+\documentclass{article}
+
+\usepackage{arxiv}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{lipsum}
 \usepackage{graphicx}
-\usepackage[utf8]{inputenc}
 
-\usepackage[hyphens]{url} % not crucial - just used below for the URL
-\usepackage{hyperref}
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\title{Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach}
 
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-%\usepackage{latexsym}
-% etc.
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-%
-% Insert the name of "your journal" with
-% \journalname{myjournal}
-%
-
-%% load any required packages here
+\author{
+    Elio Campitelli
+   \\
+     \\
+   \\
+  \texttt{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}} \\
+   \And
+    Leandro B. Díaz
+   \\
+     \\
+   \\
+  \texttt{} \\
+   \And
+    Carolina Vera
+   \\
+     \\
+   \\
+  \texttt{} \\
+  }
 
@@
 \usepackage{subfig}
+\usepackage{amsmath}
 \usepackage{chngcntr}
@@
 \begin{document}
+\maketitle
 
-\title{Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
-Elio Campitelli was supported by a PhD grant from CONICET, Argentina.} }
-
-    \titlerunning{Assessment of zonally symmetric and asymmetric components of the SAM}
-
-\author{  Elio Campitelli \and  Leandro B. Díaz \and  Carolina Vera \and  }
-
-\institute{
-        Elio Campitelli \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
- CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
- CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
-     \email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}  %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-    \and
-        Leandro B. Díaz \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
- CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
- CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
-     %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-    \and
-        Carolina Vera \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
- CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
- CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
-     %  \\
-%             \emph{Present address:} of F. Author  %  if needed
-    \and
-    }
-
-\date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
-
-\maketitle
+\def\tightlist{}
 
@@
 The regional impacts of the SAM in temperature and precipitation are strongly affected by its asymmetric component.
-\\
-\keywords{
-        Southern Annular Mode \and
-        general circulation \and
-        zonally asymmetric circulation \and
-        El Niño Southern Oscillation \and
-    }
-
 \end{abstract}
 
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
+\keywords{
+    Southern Annular Mode
+   \and
+    general circulation
+   \and
+    zonally asymmetric circulation
+   \and
+    El Niño Southern Oscillation
+  }
 
@@
 
-\begin{acknowledgements}
-NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ 
-\end{acknowledgements}
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at \url{https://psl.noaa.gov/}
 
@@
 
-\bibliographystyle{spbasic}
+\bibliographystyle{plainnat}
 \bibliography{AsymSAM,packages}

### 2020-12-16 analysis/paper/paper.Rmd
Agrega números de línea

@@
  - \usepackage{natbib}
+ - \usepackage{lineno}
+ - \linenumbers 
 ---

### 2020-12-16 analysis/pre-print/pre-print.Rmd
Pre-print

@@
 title: Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach
-
 authors:
   - name:  Elio Campitelli
+    affiliation: |
+      UBA - FCEN - DCAO
     email: elio.campitelli@cima.fcen.uba.ar
-  - name:  Leandro B. Díaz
+  - name:  Leandro Díaz
+    affiliation: |
+      UBA - FCEN - DCAO
   - name:  Carolina Vera
+    affiliation: |
+      UBA - FCEN - DCAO
 
@@
   - El Niño Southern Oscillation
+
+notice: "This is a non-peer reviewed preprint submitted to EarthArXiv. The manuscript is under review at Climate Dynamics. Subsequent versions of this article may have different content"
+
 bibliography: "AsymSAM,packages"
-bibstyle: "plainnat"
+biblio-style: plainnat
 output: 
+  bookdown::pdf_document2:
+      template: "template.tex"
+      latex_engine: "xelatex"
+      citation_package: "natbib"
+      keep_tex: true
   bookdown::pdf_book: 
@@
       keep_tex: true
+      # pandoc_args: ["--lua-filter=abstract-to-meta.lua"]
 header-includes: 
@@
  - \usepackage{chngcntr}
- - \usepackage{natbib}    
+ - \usepackage[round]{natbib}
+ - \hypersetup{hidelinks}
 ---
@@
   cache = TRUE,
-  cache.extra = 41,
+  cache.extra = 42,
   cache.path = "../cache/pre-print",
@@
 
-# Acknowledgements
+# Acknowledgements {-}
 

### 2021-04-26 analysis/paper/paper.Rmd
Actualiza afiliación.
Closes #25

@@
   address: |
-    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.  
+    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.  
+    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
   email: elio.campitelli@cima.fcen.uba.ar
@@
   address: |
-    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
-    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
-    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
+    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.  
+    CONICET – Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.  
+    CNRS – IRD – CONICET – UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina.
 

### 2021-04-26 analysis/paper/paper.Rmd
Cambia a CMAP
#21

@@
 
-Temperature data used are from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude.
-Monthly rainfall fields at 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015;@schneider2017] are also considered.
-The rainfall dataset is based on station-based records, and thus it only has continental coverage.
+Temperature data used are from NOAA's Merged Land Ocean Global Surface
+Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface
+air temperature and sea surface (water) temperature analysis into a monthly
+global grid 5\degree longitude by 5\degree latitude. For precipitation data we
+used monthly data from the CPC Merged Analysis of Precipitation [\@xie1997],
+with a 2.5\\degree\\ resolution in latitude and longitude. This rainfall gridded
+dataset is based on information from different sources such as rain gauge
+observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and
+it is available since 1979 to present.
 
@@
 
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+The research was supported by UBACyT20020170100428BA and the CLIMAX Project
+funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a
+PhD grant from CONICET, Argentina.
 
@@
 
-## Availability of data and material {-}
+## Availability of data and material {.unnumbered}
 
 All data used in this paper is freely available from their respective sources.
-ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means).
-NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sciences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html).
-The Oceanic Niño Index is available via NOAA's Climate Prediction Center: https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt
-
-## Code availability {-}
+ERA5 data can be obtained via the Copernicus Climate Data Store
+([<https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means>](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means){.uri}).
+CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado,
+USA, from their Web site at <https://psl.noaa.gov/data/gridded/data.cmap.html>.
+The Oceanic Niño Index is available via NOAA's Climate Prediction Center:
+<https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>
+
+## Code availability {.unnumbered}
 
 A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at https://github.com/eliocamp/asymsam.
+including the code used to download the data can be found at
+<https://github.com/eliocamp/asymsam>.
 

### 2021-04-26 analysis/paper/paper.Rmd
Usa visual editor (corta líneas cada 80 caracteres)

@@
 
-The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] on daily, monthly, and decadal timescales [@baldwin2001a;@fogt2006] and exerts an important influence on temperature and precipitation anomalies, and sea ice concentration [e.g.
-@fogt2020].
-Its positive phase is usually described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
-
-Most authors describe the SAM as a zonally symmetric pattern, a fact that is reflected not only in its name, but also in the various methods used to characterise it.
-Of the several different indices presented in the literature, many of them are based on zonal means of sea level pressure or geopotential height [@ho2012].
-@gong1999 defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 60\degree S, which is also the definition used by the station-based index in @marshall2003.
-@baldwin2009 proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
-
-Even though these indices are based on zonal averages, their associated geopotential height spatial anomalies contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
-The zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM [@fan2007;@silvestri2009;@fogt2012;@rosso2018].
-The fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].
-
-Some of the variability associated with the zonal asymmetries of the SAM seems to be forced by the tropics.
-ENSO-like variability affects the Southern Hemisphere extratopics through the Rossby wave trains [@mo1987;@kidson1988;@karoly1989] which project strongly onto the zonal anomalies associated with the SAM in the Pacific sector.
-Moreover, tropical influences on the SAM have been observed [@fan2007;@fogt2011; @clem2013].
-@fan2007 computed SAM indices of the Western and the Eastern Hemispheres separately and found that they were much more correlated to each other if the (linear) signal of the ENSO was removed.
-
-Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn [e.g.
-@fogt2020 and references therein].
-It is thought that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004;@gillett2005;@arblaster2006;@gillett2013].
-However, it’s not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.
-
-The impact of the zonally asymmetric component of the SAM at regional scales has not been studied in detail yet.
-The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM).
-But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012].
-The SAM-related signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
-The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
-For example, the SAM-precipitation relationship in Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
-@fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
-
-The study of the temporal variability of the asymmetric component of the SAM has not received much attention except for @fogt2012.
-This study provides evidences for the relevance of the SAM's asymmetric component.
-However, their conclusions  are based on composites of positive and negative SAM events including a small number of cases unevenly distributed among years with and without satellite information.
-The latter is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].
-Moreover, @fogt2012 studied the zonal asymmetric component of the SAM only in sea level pressure.
-Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere [@baldwin2009].
-
-Our objective is, then, to describe the zonally asymmetric and symmetric components of the SAM variability.
-We first propose a methodology that provides for each level, two indices which aim to capture independently the variability of the symmetric and asymmetric SAM component respectively.
-Their vertical structure and coherence, temporal variability and trends are consequently assessed.
-We then study the spatial patterns described by the variability exclusive to each index focusing on 50\ hPa as representing the stratosphere and 700\ hPa as representing the troposphere.
-Finally, the relationships of the SAM at 700\ hPa with temperature and precipitation anomalies are investigated.
-
-In the Section \@ref(methods) we describe the methods.
-In Section \@ref(temporal) we describe the temporal variability and vertical coherence of the indices.
-In Section \@ref(spatial), we analyse the spatial patterns of geopotential height associated with them.
-In Section \@ref(impacts), we study their relationship with surface-level temperature and precipitation.
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern
+Hemisphere extratropical circulation [@rogers1982] on daily, monthly, and
+decadal timescales [@baldwin2001a; @fogt2006] and exerts an important influence
+on temperature and precipitation anomalies, and sea ice concentration [e.g.
+@fogt2020]. Its positive phase is usually described as anomalously low pressures
+over Antarctica surrounded by a ring of anomalous high pressures in
+middle-to-high latitudes.
+
+Most authors describe the SAM as a zonally symmetric pattern, a fact that is
+reflected not only in its name, but also in the various methods used to
+characterise it. Of the several different indices presented in the literature,
+many of them are based on zonal means of sea level pressure or geopotential
+height [@ho2012]. @gong1999 defined the SAM index as the zonal mean sea level
+pressure difference between 40\degree S and 60\degree S, which is also the
+definition used by the station-based index in @marshall2003. @baldwin2009
+proposed defining the Northern and Annular modes as the leading EOF of the
+zonally averaged geopotential height at each level.
+
+Even though these indices are based on zonal averages, their associated
+geopotential height spatial anomalies contain noticeable deviations from zonal
+symmetry, particularly in the Pacific Ocean region. The zonal asymmetries have
+not been widely studied, but previous work suggest that they strongly modulate
+the regional impacts of the SAM [@fan2007; @silvestri2009; @fogt2012;
+@rosso2018]. The fact that the SAM is not entirely zonally symmetric hinders our
+ability to reconstruct its historical variability prior to the availability of
+dense observations in the Southern Hemisphere [@jones2009].
+
+Some of the variability associated with the zonal asymmetries of the SAM seems
+to be forced by the tropics. ENSO-like variability affects the Southern
+Hemisphere extratopics through the Rossby wave trains [@mo1987; @kidson1988;
+@karoly1989] which project strongly onto the zonal anomalies associated with the
+SAM in the Pacific sector. Moreover, tropical influences on the SAM have been
+observed [@fan2007; @fogt2011; @clem2013]. @fan2007 computed SAM indices of the
+Western and the Eastern Hemispheres separately and found that they were much
+more correlated to each other if the (linear) signal of the ENSO was removed.
+
+Positive trends in the SAM have been documented by various researchers using
+different indices, mostly on austral Summer and Autumn [e.g. @fogt2020 and
+references therein]. It is thought that these trends are driven primarily by
+stratospheric ozone depletion and the increase in greenhouse gases, and
+understood in the context of zonal mean variables [@marshall2004; @gillett2005;
+@arblaster2006; @gillett2013]. However, it's not clear yet how or if the
+asymmetric SAM component responds to these forcings, or how its variability
+alters the observed trends.
+
+The impact of the zonally asymmetric component of the SAM at regional scales has
+not been studied in detail yet. The positive phase of the SAM is associated with
+colder-than-normal temperatures over Antarctica and warmer-than-normal
+temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM).
+But there are significant deviations from this zonal mean response, notably in
+the Antarctic Peninsula and the South Atlantic [@fogt2012]. The SAM-related
+signal on precipitation anomalies behaves similarly, although with even greater
+deviation from zonal symmetry [@lim2016]. The importance of zonal asymmetries of
+the SAM on these impacts have been studied in certain regions. For example, the
+SAM-precipitation relationship in Southeastern South America can be explained by
+the Pacific-South American (PSA)-like zonally asymmetric circulation associated
+with the SAM [@silvestri2009; @rosso2018]. @fan2007 also found that
+precipitation in East Asia was impacted by the variability of only the Western
+Hemisphere part of the SAM.
+
+The study of the temporal variability of the asymmetric component of the SAM has
+not received much attention except for @fogt2012. This study provides evidences
+for the relevance of the SAM's asymmetric component. However, their conclusions
+are based on composites of positive and negative SAM events including a small
+number of cases unevenly distributed among years with and without satellite
+information. The latter is particularly important due to the inhomogeneities in
+reanalysis products prior to the satellite era and the possible change in the
+asymmetric structure of the SAM [@silvestri2009]. Moreover, @fogt2012 studied
+the zonal asymmetric component of the SAM only in sea level pressure. Zonal
+asymmetries in the SAM spatial pattern are fairly barotropic throughout the
+troposphere, but they change dramatically in the stratosphere [@baldwin2009].
+
+Our objective is, then, to describe the zonally asymmetric and symmetric
+components of the SAM variability. We first propose a methodology that provides
+for each level, two indices which aim to capture independently the variability
+of the symmetric and asymmetric SAM component respectively. Their vertical
+structure and coherence, temporal variability and trends are consequently
+assessed. We then study the spatial patterns described by the variability
+exclusive to each index focusing on 50 hPa as representing the stratosphere and
+700 hPa as representing the troposphere. Finally, the relationships of the SAM
+at 700 hPa with temperature and precipitation anomalies are investigated.
+
+In the Section \@ref(methods) we describe the methods. In Section
+\@ref(temporal) we describe the temporal variability and vertical coherence of
+the indices. In Section \@ref(spatial), we analyse the spatial patterns of
+geopotential height associated with them. In Section \@ref(impacts), we study
+their relationship with surface-level temperature and precipitation.
 
@@
 
-We used monthly geopotential height at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
-We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations.
+We used monthly geopotential height at 2.5\degree longitude by
+2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels
+from ERA5 [@hersbach2020] for the period 1979 to 2018. We restrict our analysis
+to the post-satellite era to avoid any confounding factors arising from the
+incorporation of satellite observations.
 
@@
 
-Traditionally, the SAM is defined as the leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels [@ho2012].
-Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the leading EOF of the monthly anomalies of geopotential height  south of 20\degree S at each level.
-We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in `r dims$t` rows and  `r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points of latitude).
-We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999].
-We consider in the EOF analysis all months together without dividing by seasons.
-
-To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \@ref(fig:method) at 700\ hPa.
-The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
-We then compute the SAM index, Asymmetric SAM index (A\nobreakdash-SAM) and Symmetric SAM (S\nobreakdash-SAM) indices as the coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude).
-The three indices are then normalized by dividing them by the standard deviation of the SAM index at each level.
-As a result, the magnitudes between indices are comparable.
-However, only SAM index has unit standard deviation per definition.
-The explained variance of each pattern is used as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
-To quantify the coherence between temporal series corresponding to different indices or the same index at different levels, we computed the temporal correlation between them.
+Traditionally, the SAM is defined as the leading empirical orthogonal mode (EOF)
+of sea-level pressure or geopotential height anomalies at low levels [@ho2012].
+Following @baldwin2001, we extend that definition vertically and use the term
+SAM to refer to the leading EOF of the monthly anomalies of geopotential height
+south of 20\degree S at each level. We performed EOFs by computing the Singular
+Value Decomposition of the data matrix consisting in `r dims$t` rows and
+`r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points
+of latitude). We weighted the values by the square root of the cosine of
+latitude to account for the non-equal area of each gridpoint [@chung1999]. We
+consider in the EOF analysis all months together without dividing by seasons.
+
+To separate the zonally symmetric and asymmetric components of the SAM, we
+computed the zonal mean and anomalies of the full SAM spatial pattern, as shown
+in Figure \@ref(fig:method) at 700 hPa. The full spatial signal
+($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric
+($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric
+($[\mathrm{EOF_1}](\lambda, \phi)$) components. We then compute the SAM index,
+Asymmetric SAM index (A\nobreakdash-SAM) and Symmetric SAM (S\nobreakdash-SAM)
+indices as the coefficients of the regression of each monthly geopotential
+height field on the respective patterns (weighting by the cosine of latitude).
+The three indices are then normalized by dividing them by the standard deviation
+of the SAM index at each level. As a result, the magnitudes between indices are
+comparable. However, only SAM index has unit standard deviation per definition.
+The explained variance of each pattern is used as an indicator of the degree of
+zonally symmetry or asymmetry of each monthly field. To quantify the coherence
+between temporal series corresponding to different indices or the same index at
+different levels, we computed the temporal correlation between them.
 
@@
 
-The method assumes linearity in the asymmetric component of the SAM.
-That means that zonal asymmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
-@fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
-To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700\ hPa and 50\ hPa levels (Figures \@ref(fig:A3) and \@ref(fig:A4)).
-In all seasons and both levels, SAM+ composites are similar to SAM- in structure but with the opposite sign.
-Spatial correlations between composites for each season are high.
-The method considered in this study seems then a reasonable approximation of the phenomenon.
-
-By performing the EOF analysis using data for all months we are assuming that the zonally asymmetric structure of the SAM is the same at all seasons.
-The latter was assessed by computing geopotential height zonal anomalies by projecting the first EOF of each season independently.
-The following seasons were considered – December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
-Results are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 1) and show spatial correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
-In the stratosphere (Figure \@ref(fig:A5), row 2), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree\ in comparison with the rest of the year.
-Spatial correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
-Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year with the exception of DJF in the stratosphere.
-
-The method also assumes that the zonally asymmetric pattern of the SAM remains stationary along the period considered.
-@silvestri2009 suggest that this might not be the case between 1958 and 2004.
-Zonal asymmetric patterns of SAM were computed for the two halves of the period (1979 to 1998 and 1999 to 2018) respectively.
-They show no systematic change neither in the stratosphere nor in the troposphere (Figure \@ref(fig:A6)).
+The method assumes linearity in the asymmetric component of the SAM. That means
+that zonal asymmetries associated with positive SAM phase (SAM+) are almost
+opposite in sign and of the same magnitude to the ones associated with negative
+SAM phase (SAM-). @fogt2012's composites (their Figure 4) suggest that this
+might not be entirely valid, although much of that apparent non-linearity could
+be due to the heterogenous nature of the selected years for constructing the
+composites. To test this assumption, we computed seasonal composites of zonal
+anomalies of geopotential height for SAM+ and SAM- (defined as months in which
+the SAM index is greater than 1 standard deviation and lower than minus 1
+standard deviation, respectively) for the period from 1979 to 2018 at the
+700 hPa and 50 hPa levels (Figures \@ref(fig:A3) and \@ref(fig:A4)). In all
+seasons and both levels, SAM+ composites are similar to SAM- in structure but
+with the opposite sign. Spatial correlations between composites for each season
+are high. The method considered in this study seems then a reasonable
+approximation of the phenomenon.
+
+By performing the EOF analysis using data for all months we are assuming that
+the zonally asymmetric structure of the SAM is the same at all seasons. The
+latter was assessed by computing geopotential height zonal anomalies by
+projecting the first EOF of each season independently. The following seasons
+were considered -- December to February (DJF), March to May (MAM), June to
+August (JJA) and September to November (SON). Results are very similar to each
+other in the troposphere (Figure \@ref(fig:A5), row 1) and show spatial
+correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON). In the
+stratosphere (Figure \@ref(fig:A5), row 2), patterns are similar for all seasons
+except DJF, when the wave-1 zonal anomalies are rotated 90\degree in comparison
+with the rest of the year. Spatial correlations in the stratosphere are between
+-0.24 (DJF with SON) and 0.95 (MAM with JJA). Therefore, the results confirm
+that the zonal asymmetric structure of the SAM is very similar throughout most
+of the year with the exception of DJF in the stratosphere.
+
+The method also assumes that the zonally asymmetric pattern of the SAM remains
+stationary along the period considered. @silvestri2009 suggest that this might
+not be the case between 1958 and 2004. Zonal asymmetric patterns of SAM were
+computed for the two halves of the period (1979 to 1998 and 1999 to 2018)
+respectively. They show no systematic change neither in the stratosphere nor in
+the troposphere (Figure \@ref(fig:A6)).
 
@@
 
-We performed linear regressions to quantify the association between the SAM indices and other variables.
-Moreover, we apply multiple linear regression analysis to describe the combined influence of both A\nobreakdash-SAM and S\nobreakdash-SAM.
-To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with A\nobreakdash-SAM and S\nobreakdash-SAM we fit the equation
+We performed linear regressions to quantify the association between the SAM
+indices and other variables. Moreover, we apply multiple linear regression
+analysis to describe the combined influence of both A\nobreakdash-SAM and
+S\nobreakdash-SAM. To obtain the linear coefficients of a variable $X$
+(geopotential, temperature, precipitation, etc...) with A\nobreakdash-SAM and
+S\nobreakdash-SAM we fit the equation
 
@@
 
-where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and $\epsilon$ are the constant and error terms.
-From this equation, $\alpha$ represents the (linear) association of $X$ with the variability of A\nobreakdash-SAM that is not explained by the variability of S\nobreakdash-SAM; i.e.
-it is proportional to the partial correlation of $X$ and A\nobreakdash-SAM, controlling for the effect of S\nobreakdash-SAM, and vice versa for $\beta$.
-When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we averaged the relevant variables seasonally for each year and trimester before computing the regression.
-
-Statistical significance for regression fields were evaluated adjusting p-values by controlling for the False Discovery Rate [@benjamini1995;@wilks2016] to avoid misleading results from the high number of regressions [@walker1914;@katz1991].
-
-Linear trends were computed by Ordinary Least Squares and the 95% confidence interval  was computed assuming a t-distribution with the appropriate residual degrees of freedom.
-To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
-
-We computed density probability estimates using a Gaussian kernel of optimal bin width according to @sheather1991.
+where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time,
+$\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and
+$\epsilon$ are the constant and error terms. From this equation, $\alpha$
+represents the (linear) association of $X$ with the variability of
+A\nobreakdash-SAM that is not explained by the variability of S\nobreakdash-SAM;
+i.e. it is proportional to the partial correlation of $X$ and A\nobreakdash-SAM,
+controlling for the effect of S\nobreakdash-SAM, and vice versa for $\beta$.
+When performing a separate regression for each trimester (DJF, MAM, JJA, SON),
+we averaged the relevant variables seasonally for each year and trimester before
+computing the regression.
+
+Statistical significance for regression fields were evaluated adjusting p-values
+by controlling for the False Discovery Rate [@benjamini1995; @wilks2016] to
+avoid misleading results from the high number of regressions [@walker1914;
+@katz1991].
+
+Linear trends were computed by Ordinary Least Squares and the 95% confidence
+interval was computed assuming a t-distribution with the appropriate residual
+degrees of freedom. To the amplitude of the zonal waves is defined through
+computing the Fourier transform of the spatial field at each latitude circle.
+
+We computed density probability estimates using a Gaussian kernel of optimal bin
+width according to @sheather1991.
 
@@
 
-We performed all analysis in this paper using the R programming language [@rcoreteam2020], using the data.table package [@dowle2020] and the metR package [@campitelli2020].
-All graphics are made using ggplot2 [@wickham2009].
-We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
-The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].
+We performed all analysis in this paper using the R programming language
+[@rcoreteam2020], using the data.table package [@dowle2020] and the metR package
+[@campitelli2020]. All graphics are made using ggplot2 [@wickham2009]. We
+downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and
+indices of the ENSO with the rsoi package [@albers2020]. The paper was rendered
+using knitr and rmarkdown [@xie2015; @allaire2019].
 
@@
 
-We first asses the temporal evolution of A\nobreakdash-SAM and S\nobreakdash-SAM.
-Figure \@ref(fig:asymsam-timeseries) shows the corresponding time series for 700\ hPa and 50\ hPa and their corresponding density estimates.
-We selected these two levels as representative of the tropospheric and stratospheric variability respectively.
-As it is shown below, the variabilities of both indices are highly coherent within each atmospheric layer, therefore is reasonable to take one level as representative of each layer.
-
-Month-to-month variability is evident for both indices, with noisy variations in the low frequencies.
-At first glance the series can be distinguished by their distributions.
-Compared to the tropospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound.
-The A\nobreakdash-SAM series have both more variability in the higher frequencies than the S\nobreakdash-SAM series.
-
-The stratospheric S\nobreakdash-SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \@ref(fig:A2)).
-This might suggest a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b].
-A local peak at 2 years  years is discernible in the periodogram of the tropospheric S\nobreakdash-SAM, although it's not statistically significant.
-In the troposphere the most significant peak of variability is found in A\nobreakdash-SAM at around `r 12*.3` months.
+We first asses the temporal evolution of A\nobreakdash-SAM and
+S\nobreakdash-SAM. Figure \@ref(fig:asymsam-timeseries) shows the corresponding
+time series for 700 hPa and 50 hPa and their corresponding density estimates. We
+selected these two levels as representative of the tropospheric and
+stratospheric variability respectively. As it is shown below, the variabilities
+of both indices are highly coherent within each atmospheric layer, therefore is
+reasonable to take one level as representative of each layer.
+
+Month-to-month variability is evident for both indices, with noisy variations in
+the low frequencies. At first glance the series can be distinguished by their
+distributions. Compared to the tropospheric indices, the stratospheric indices
+are much more long-tailed; that is, extreme values (both negative and positive)
+abound. The A\nobreakdash-SAM series have both more variability in the higher
+frequencies than the S\nobreakdash-SAM series.
+
+The stratospheric S\nobreakdash-SAM varies strongly with a two-year period,
+which can be seen by spectral analysis (Figure \@ref(fig:A2)). This might
+suggest a link between stratospheric SAM variability and the Quasi-Biennial
+Oscillation [@baldwin2001b]. A local peak at 2 years years is discernible in the
+periodogram of the tropospheric S\nobreakdash-SAM, although it's not
+statistically significant. In the troposphere the most significant peak of
+variability is found in A\nobreakdash-SAM at around `r 12*.3` months.
 
@@
 
-From a visual inspection, the A\nobreakdash-SAM and S\nobreakdash-SAM time series appear to be correlated.
-Moreover, looking at the extremes in the stratosphere, the S\nobreakdash-SAM series appears to lag the A\nobreakdash-SAM series (see, for example, the positive events on late 1987).
-Figure \@ref(fig:cor-lev) shows these correlations along all levels considered, for zero and -1 lags.
-Values of zero-lag correlations between A\nobreakdash-SAM and S\nobreakdash-SAM are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
-One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`.
-In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev`\ hPa and then increase again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care).
-At the same time, one-month-lag correlations increase with height.
-Therefore, stratospheric A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index.
+From a visual inspection, the A\nobreakdash-SAM and S\nobreakdash-SAM time
+series appear to be correlated. Moreover, looking at the extremes in the
+stratosphere, the S\nobreakdash-SAM series appears to lag the A\nobreakdash-SAM
+series (see, for example, the positive events on late 1987). Figure
+\@ref(fig:cor-lev) shows these correlations along all levels considered, for
+zero and -1 lags. Values of zero-lag correlations between A\nobreakdash-SAM and
+S\nobreakdash-SAM are relatively constant throughout the troposphere,
+fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and
+`r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations
+are similarly constant but significantly reduced to around
+`r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere,
+zero-lag correlations drop to a minimum of
+`r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at
+`r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then increase again
+monotonically with height up to the uppermost level of the reanalysis (although
+results near the top of the models are to be interpreted with care). At the same
+time, one-month-lag correlations increase with height. Therefore, stratospheric
+A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index.
 (Correlations at lags -5 to 5 are shown in Figure \@ref(fig:A1).)
@@
 
-Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across levels for the SAM, A\nobreakdash-SAM  and S\nobreakdash-SAM indices.
-For the SAM (Figure \@ref(fig:cross-correlation)a), high values below 100\ hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
-Above 100\ hPa, correlation between levels does falls off more rapidly, indicating less coherent (zero-lag) variability.
-But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g.
-grater than 0.4 between tropospheric levels and levels below 30\ hPa).
-A\nobreakdash-SAM and S\nobreakdash-SAM (Figure \@ref(fig:cross-correlation)b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
-Stratospheric coherency is stronger for the A\nobreakdash-SAM than the S\nobreakdash-SAM.
-The stratospheric S\nobreakdash-SAM seems to connect more strongly to the troposphere than the stratospheric A\nobreakdash-SAM.
+Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across
+levels for the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices. For the SAM
+(Figure \@ref(fig:cross-correlation)a), high values below 100 hPa reflect the
+vertical (zero-lag) coherency throughout the troposphere. Above 100 hPa,
+correlation between levels does falls off more rapidly, indicating less coherent
+(zero-lag) variability. But correlations between tropospheric and
+lower-to-middle stratospheric levels are still relatively high (e.g. grater than
+0.4 between tropospheric levels and levels below 30 hPa). A\nobreakdash-SAM and
+S\nobreakdash-SAM (Figure \@ref(fig:cross-correlation)b and c, respectively)
+share similar high level of coherency in the troposphere but they differ in
+their stratospheric behaviour. Stratospheric coherency is stronger for the
+A\nobreakdash-SAM than the S\nobreakdash-SAM. The stratospheric
+S\nobreakdash-SAM seems to connect more strongly to the troposphere than the
+stratospheric A\nobreakdash-SAM.
 
 <!-- ```{r} -->
+
 <!-- indices_wide %>%  -->
+
 <!--   .[, FitLm(as.numeric(time), asym, sym, se = TRUE), by = .(lev)] %>%  -->
+
 <!--   rm_intercept() %>%  -->
+
 <!--   # .[, term := NULL] %>%  -->
+
 <!--   .[, estimate := estimate/(3600*24*365*10)] %>% -->
+
 <!--   .[, std.error := std.error/(3600*24*365*10)] %>% -->
+
 <!--   # setnames("type", "term") %>%  -->
+
 <!--   ggplot(aes(lev, estimate)) + -->
+
 <!--   geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) + -->
+
 <!--   geom_line() + -->
+
 <!--   scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) + -->
+
 <!--   geom_hline(yintercept = 0) + -->
+
 <!--   scale_x_level() + -->
+
 <!--   scale_y_continuous("Standard deviations per decade") + -->
+
 <!--   coord_flip() + -->
+
 <!--   facet_grid(~ term,labeller = labeller(term = lab_sam)) + -->
+
 <!--   tag_facets()  -->
+
 <!-- ``` -->
@@
 
-The linear trends for each of the indices (SAM, S\nobreakdash-SAM and A\nobreakdash-SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \@ref(fig:trends)).
-The SAM index presents a statistically positive significant trend (Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50\ hPa and reaches its maximum value at 100\ hPa.
-The seasonal trends (rest of Figure \@ref(fig:trends) column a) indicate that positive trends are present in  Autumn and particularly in Summer, where the 100\ hPa maximum is much more defined.
-In Winter and Spring, we detect no statistically significant trend.
-This is consistent with the results of previous studies, which find large positive trends in Summer, smaller in Autumn and no trends in the other seasons [e.g.
-@fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.
-
-By separating the SAM signal in its asymmetric and symmetric parts, not only we can see that these trends are almost entirely due to the symmetric component (column b vs.
-column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer.
-In Summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
-In Autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.
-
-Figure \@ref(fig:r-squared-trend) shows trends for the explained variance of each index.
-There is no evidence of a significant trend in the stratosphere.
-In the troposphere, there is a positive trend for A\nobreakdash-SAM and no significant trend for S\nobreakdash-SAM.
-This suggest that the SAM has become more asymmetric in the period from 1979 to 2018.
-However, the change is slight, around 1% increased explained variance per decade.
+The linear trends for each of the indices (SAM, S\nobreakdash-SAM and
+A\nobreakdash-SAM) were evaluated for the complete period 1979 -- 2018 at each
+level for the whole year and separated by trimesters (Figure \@ref(fig:trends)).
+The SAM index presents a statistically positive significant trend (Figure
+\@ref(fig:trends)`r get_tag(term == "full" & season == "Annual")`) that extends
+throughout the troposphere up to about 50 hPa and reaches its maximum value at
+100 hPa. The seasonal trends (rest of Figure \@ref(fig:trends) column a)
+indicate that positive trends are present in Autumn and particularly in Summer,
+where the 100 hPa maximum is much more defined. In Winter and Spring, we detect
+no statistically significant trend. This is consistent with the results of
+previous studies, which find large positive trends in Summer, smaller in Autumn
+and no trends in the other seasons [e.g. @fogt2020 and references therein] using
+indices of the SAM based on surface or near-surface circulation.
+
+By separating the SAM signal in its asymmetric and symmetric parts, not only we
+can see that these trends are almost entirely due to the symmetric component
+(column b vs. column c in Figure \@ref(fig:trends)), but in some cases the
+trends become clearer. In Summer, A\nobreakdash-SAM has a statistically non
+significant negative trend in the middle troposphere that obscures the trend in
+the SAM index; as a result, trends computed using only the Symmetric component
+are stronger (compare the shading region in Figure
+\@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and
+`r get_tag(term == "sym" & season == "DJF")`). In Autumn, the S\nobreakdash-SAM
+index reveals a statistically significant positive trend in the stratosphere
+that is not significant using the SAM index.
+
+Figure \@ref(fig:r-squared-trend) shows trends for the explained variance of
+each index. There is no evidence of a significant trend in the stratosphere. In
+the troposphere, there is a positive trend for A\nobreakdash-SAM and no
+significant trend for S\nobreakdash-SAM. This suggest that the SAM has become
+more asymmetric in the period from 1979 to 2018. However, the change is slight,
+around 1% increased explained variance per decade.
 
@@
 
-We then computed the spatial regression of geopotential height anomalies on the A\nobreakdash-SAM and S\nobreakdash-SAM indices at 700\ hPa and 50\ hPa levels (Figure \@ref(fig:2d-regr)).
-While regression coefficients in Figure \@ref(fig:2d-regr) column a are computed using SAM, the regression coefficients in columns b and c of Figure \@ref(fig:2d-regr) are computed using multiple regression using the A\nobreakdash-SAM  and S\nobreakdash-SAM indices at the same time.
-Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other one.
-
-In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \@ref(fig:2d-regr)a.1) which is not centred in the South Pole.
-On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric.
-Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern Hemisphere
+We then computed the spatial regression of geopotential height anomalies on the
+A\nobreakdash-SAM and S\nobreakdash-SAM indices at 700 hPa and 50 hPa levels
+(Figure \@ref(fig:2d-regr)). While regression coefficients in Figure
+\@ref(fig:2d-regr) column a are computed using SAM, the regression coefficients
+in columns b and c of Figure \@ref(fig:2d-regr) are computed using multiple
+regression using the A\nobreakdash-SAM and S\nobreakdash-SAM indices at the same
+time. Thus, they are to be interpreted as the patterns associated with each
+index, removing the variability (linearly) explained by the other one.
+
+In the stratosphere, the spatial pattern associated with the SAM is clearly
+dominated by a zonally symmetric, monopolar structure (Figure
+\@ref(fig:2d-regr)a.1) which is not centred in the South Pole. On the other
+hand, the monopole associated with S\nobreakdash-SAM (Figure
+\@ref(fig:2d-regr)c.1) is more symmetric. Furthermore, the regression pattern of
+A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the
+Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern
+Hemisphere
 
@@
 
-In the troposphere, the regression pattern associated with the SAM shows the well-known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 (Figure \@ref(fig:2d-regr)a.2, [@fogt2012]).
-The regression patterns associated with the A\nobreakdash-SAM and S\nobreakdash-SAM indices successfully disentangle both structures.
-While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \@ref(fig:2d-regr)b.2), the S\nobreakdash-SAM index is associated with an annular structure, almost devoid of zonal asymmetries (Figure \@ref(fig:2d-regr)c.2).
-The wave-3 pattern observed in Figure \@ref(fig:2d-regr)b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern described by @raphael2004, whose reference locations are marked with points in the figure.
-Indeed, there is no correlation between @raphael2004's index and  A\nobreakdash-SAM (cor = `r signif(cor_raphael$estimate, 2)`, p-value `r report_p(cor_raphael$p.value)`).
-Thus, the tropospheric A\nobreakdash-SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
-
-(ref:wave-amplitude-caption) Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \@ref(fig:2d-regr) for zonal waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the amplitude of the zonal mean.
+In the troposphere, the regression pattern associated with the SAM shows the
+well-known combination of zonally symmetrical annular mode with zonal
+asymmetries in the form of a wave-3 (Figure \@ref(fig:2d-regr)a.2, [@fogt2012]).
+The regression patterns associated with the A\nobreakdash-SAM and
+S\nobreakdash-SAM indices successfully disentangle both structures. While the
+A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure
+\@ref(fig:2d-regr)b.2), the S\nobreakdash-SAM index is associated with an
+annular structure, almost devoid of zonal asymmetries (Figure
+\@ref(fig:2d-regr)c.2). The wave-3 pattern observed in Figure
+\@ref(fig:2d-regr)b.2 is rotated by half a wavelength from the average position
+of the mean wave-3 pattern described by @raphael2004, whose reference locations
+are marked with points in the figure. Indeed, there is no correlation between
+@raphael2004's index and A\nobreakdash-SAM (cor =
+`r signif(cor_raphael$estimate, 2)`, p-value `r report_p(cor_raphael$p.value)`).
+Thus, the tropospheric A\nobreakdash-SAM index represents a zonal displacement
+in the position of the climatological wave-3 pattern.
+
+(ref:wave-amplitude-caption) Amplitude (meters) of zonal waves of the
+geopotential height regression patterns in Figure \@ref(fig:2d-regr) for zonal
+waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the
+amplitude of the zonal mean.
 
@@
 
-The amplitude of the first zonal wavenumbers at each latitude at 50\ hPa and 700\ hPa are shown in Figure \@ref(fig:wave-amplitude), where wavenumber zero represents the amplitude of the zonal mean.
-Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column `r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at both levels.
-However, zonal waves are important, particularly South of 50\degree S, with wavenumber 1 clearly dominating at 50\ hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of waves of similar amplitude at 700\ hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`).
-Figure \@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean.
-On the other hand, the S\nobreakdash-SAM it is almost entirely explained by the zonal mean at both levels (Figure \@ref(fig:wave-amplitude) column c), with little to no contribution from zonal waves with wavenumbers 1 to 3.
-
-Looking at Figure \@ref(fig:2d-regr)b.2, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
+The amplitude of the first zonal wavenumbers at each latitude at 50 hPa and
+700 hPa are shown in Figure \@ref(fig:wave-amplitude), where wavenumber zero
+represents the amplitude of the zonal mean. Zonal wave amplitudes of the spatial
+pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column
+`r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at
+both levels. However, zonal waves are important, particularly South of 50\degree
+S, with wavenumber 1 clearly dominating at 50 hPa (Figure
+\@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of
+waves of similar amplitude at 700 hPa (Figure
+\@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`). Figure
+\@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the
+A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere
+(Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while
+in the troposphere it is explained by a combination of zonal waves 3 to 1 in
+decreasing level of importance (Figure
+\@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with
+negligible amplitude of the zonal mean. On the other hand, the S\nobreakdash-SAM
+it is almost entirely explained by the zonal mean at both levels (Figure
+\@ref(fig:wave-amplitude) column c), with little to no contribution from zonal
+waves with wavenumbers 1 to 3.
+
+Looking at Figure \@ref(fig:2d-regr)b.2, it becomes apparent that zonal waves 1
+and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is
+larger in the Western Hemisphere than in the Eastern Hemisphere.
 
@@
 
-To analyse the vertical structure of the geopotential height anomalies associated with the A\nobreakdash-SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50\ hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)a) and for the 700\ hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)b).
-The geopotential height anomalies associated with the stratospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric A\nobreakdash-SAM.
-The vertical structure of this signal tilts about 60\degree\ to the West between 100\ hPa and 1\ hPa, suggesting baroclinic processes.
-The signal in the stratosphere maximises near 10\ hPa despite using the 50\ hPa index for the regression.
-
-The tropospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)b) has significant signals that extend upwards to the uppermost levels considered.
-In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250\ hPa.
-The anomalies are larger in the Western hemisphere, where they extend into the stratosphere.
-In the Eastern hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere.
-So, while the tropospheric A\nobreakdash-SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric A\nobreakdash-SAM.
-The structures shown in  Figure \@ref(fig:vertical-regression) are robust to the choice of index level.
-For any stratospheric (above 100\ hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10\ hPa in Figure \@ref(fig:vertical-regression)a.
-Conversely, for any tropospheric (below 100\ hPa) index, the result is very similar to Figure \@ref(fig:vertical-regression)b.
-The patterns mainly change in amplitude (not shown).
+To analyse the vertical structure of the geopotential height anomalies
+associated with the A\nobreakdash-SAM index, we show a vertical cross section of
+regressions of mean geopotential height between 65\degree S and 40\degree S for
+the 50 hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)a) and
+for the 700 hPa A\nobreakdash-SAM index (Figure
+\@ref(fig:vertical-regression)b). The geopotential height anomalies associated
+with the stratospheric A\nobreakdash-SAM (Figure
+\@ref(fig:vertical-regression)a) are clearly constrained to the stratosphere,
+which underscores the uncoupling between the stratospheric and tropospheric
+A\nobreakdash-SAM. The vertical structure of this signal tilts about
+60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic
+processes. The signal in the stratosphere maximises near 10 hPa despite using
+the 50 hPa index for the regression.
+
+The tropospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)b) has
+significant signals that extend upwards to the uppermost levels considered. In
+the troposphere, the wave-3 structure is equivalent barotropic with maximum
+amplitude at roughly 250 hPa. The anomalies are larger in the Western
+hemisphere, where they extend into the stratosphere. In the Eastern hemisphere
+the wave-3 signal is smaller and confined to the troposphere while negative
+anomalies dominate in the stratosphere. So, while the tropospheric
+A\nobreakdash-SAM index is associated with stratospheric geopotential anomalies,
+these do not project strongly onto the stratospheric A\nobreakdash-SAM. The
+structures shown in Figure \@ref(fig:vertical-regression) are robust to the
+choice of index level. For any stratospheric (above 100 hPa) index, the
+resulting anomalies are very similar to the wave-1 structure with maximum near
+10 hPa in Figure \@ref(fig:vertical-regression)a. Conversely, for any
+tropospheric (below 100 hPa) index, the result is very similar to Figure
+\@ref(fig:vertical-regression)b. The patterns mainly change in amplitude (not
+shown).
 
@@
 
-The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the PSA Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
-Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value `r cor_texts[term == "full"]$p.value`).
-This relationship is captured mainly by the A\nobreakdash-SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI, consistent with @fan2007's results, whereas the S\nobreakdash-SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
-The same analysis was performed using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987], obtaining similar results.
-The latter allows to conclude that these results do not depend on the ENSO index used.
+The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the PSA
+Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with
+the ENSO [@karoly1989]. Indeed, @fogt2011 showed that there is a significant
+relationship between the SAM and the ENSO. The correlation between the SAM and
+the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is
+`r cor_texts[term == "full"]$cor` (p-value
+`r cor_texts[term == "full"]$p.value`). This relationship is captured mainly by
+the A\nobreakdash-SAM, as this index has a partial correlation of
+`r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation`
+(p-value `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the
+ONI, consistent with @fan2007's results, whereas the S\nobreakdash-SAM's partial
+correlation with the ONI is essentially null
+(`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`;
+p-value `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`). The same
+analysis was performed using the Multivariate ENSO Index [@wolter2011] and the
+Southern Oscillation Index [@ropelewski1987], obtaining similar results. The
+latter allows to conclude that these results do not depend on the ENSO index
+used.
 
@@
 
-To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices we regressed surface land air and sea temperature and precipitation onto each of the three SAM indices at 700\ hPa.
-As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
-
-Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700\ hPa with surface land air and sea temperature for each trimester.
-In Summer, positive values of the SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
-The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W,  120\degree W, 150\degree E and 90\degree E respectively.
-In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
-Figure \@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of A\nobreakdash-SAM and S\nobreakdash-SAM, respectively.
-Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics.
-Temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM.
-This signal is not present in the regression pattern with the SAM.
-Spring (Figure \@ref(fig:regr-air-season), row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with fewer regions where regressed anomalies have statistical significance.
-
-In Autumn and Winter (rows `r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the SAM.
-There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
-These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following @wilks2016.
-However, repeating this analysis with 2-meter air temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
-Moreover, similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
-
-The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons -- although with varying intensity and small-scale details -- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
-It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in Summer and Spring).
-
-Figure \@ref(fig:regr-air-season) column b can be partially compared with Figure 11 from @fogt2012.
-Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.
-
-Regression of the SAM indices with seasonal mean precipitation and 700\ hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for New Zealand and neighbouring islands, and South America respectively.
-South Africa is not shown because no significant signal was detected there.
+To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM
+and S\nobreakdash-SAM indices we regressed surface land air and sea temperature
+and precipitation onto each of the three SAM indices at 700 hPa. As shown in
+previous sections, the three indices are highly coherent in the troposphere, so
+we select this level to represent the tropospheric circulation for consistency
+with previous studies.
+
+Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at
+700 hPa with surface land air and sea temperature for each trimester. In Summer,
+positive values of the SAM index (Figure
+\@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are
+associated with negative temperature anomalies near Antarctica which are
+surrounded by a ring of positive anomalies. The ring is not zonally symmetric,
+as there are four distinctive local maximums around 30\degree W, 120\degree W,
+150\degree E and 90\degree E respectively. In the tropics, there are negative
+anomalies in the equatorial Pacific, consistent with the negative correlation
+between SAM and ENSO. Figure
+\@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))`
+show temperature anomalies associated with positive values of A\nobreakdash-SAM
+and S\nobreakdash-SAM, respectively. Both the local maximums in the ring and the
+anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM
+regression map, while temperature patterns linked to positive S\nobreakdash-SAM
+show a more zonally consistent ring and less relation to the tropics.
+Temperature anomalies in the Indian ocean, South Africa and Australia are
+strongly related to A\nobreakdash-SAM. This signal is not present in the
+regression pattern with the SAM. Spring (Figure \@ref(fig:regr-air-season), row
+`r get_row(season == "SON")`) features similar patterns but of smaller
+magnitude, with fewer regions where regressed anomalies have statistical
+significance.
+
+In Autumn and Winter (rows
+`r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))` in Figure
+\@ref(fig:regr-air-season)) the positive ring is only present through its local
+maximums in the regression with the SAM. There are also negative anomalies in
+Southern Australia, and positive anomalies over New Zealand and Southern South
+America. These patterns are not significant in the sense that there are no areas
+with p-values below 0.05 when controlling for FDR following @wilks2016. However,
+repeating this analysis with 2-meter air temperature from ERA5 resulted in
+similar patterns that were statistically significant (not shown). Moreover,
+similar features were observed in station measurements by @jones2019, although
+using data from 1957 to 2016.
+
+The pattern of negative anomalies in the pole surrounded by positive anomalies
+roughly seen in all seasons -- although with varying intensity and small-scale
+details -- translates to enhanced meridional temperature gradient maximised in
+the zero line, which is consistent with the intensification and poleward
+migration of the westerlies commonly linked to the SAM through thermal wind
+balance. It's then not surprising to see it more clearly in association with
+S\nobreakdash-SAM (at least in Summer and Spring).
+
+Figure \@ref(fig:regr-air-season) column b can be partially compared with Figure
+11 from @fogt2012. Although they used station data from 1958 to 2001, main
+features are reproduced here, such as the strong signal in New Zealand and
+Australia in Summer and Spring.
+
+Regression of the SAM indices with seasonal mean precipitation and 700 hPa
+geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and
+\@ref(fig:pp-regr-america) for New Zealand and neighbouring islands, and South
+America respectively. South Africa is not shown because no significant signal
+was detected there.
 
@@
 
-pp <- GPCC() %>% 
+pp <- CMAP() %>% 
   ReadNetCDF(subset = list(lat = c(-90, 0),
@@
   normalise_coords() %>%
-  .[, days := lubridate::days_in_month(time[1]), by = time] %>% 
-  .[, precip := precip/days]
+  .[, days := lubridate::days_in_month(time[1]), by = time]
 
@@
 
-In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the Southeastern region (Figure \@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006.
-The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure \@ref(fig:pp-regr-oceania)c.1).
-Geopotential height anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by @hendon2007.
-A\nobreakdash-SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
-
-The seasonal-based regressions show statistically significant anomalies only in Spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
-In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related with positive A\nobreakdash-SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).
-
-In Summer, positive SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \@ref(fig:pp-regr-oceania)a.2).
-The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
-In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, and a broad area of positive anomalies in the North-East to South-West direction.
-This structure seems to be associated with the S\nobreakdash-SAM, while the Northern positive values are associated with the A\nobreakdash-SAM.
-In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the S\nobreakdash-SAM.
-None of these regression coefficients are statistically significant at the 95% level.
-The Spring signal is broadly consistent with @hendon2007, but whereas they also detected a strong signal in Summer, Figure \@ref(fig:pp-regr-oceania)a.2 shows no statistically significant association (although the coefficients have the consistent sign).
-
-(ref:pp-rer-america-cap) Same as Figure \@ref(fig:pp-regr-oceania) but for South America.
+In Australia, the annual regression shows that the SAM index is associated with
+positive precipitation anomalies in the Southeastern region (Figure
+\@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006. The separation
+between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive
+anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure
+\@ref(fig:pp-regr-oceania)c.1). Geopotential height anomalies associated with
+this index (black contours) are indicative of easterly flow from the Tasman Sea,
+which could explain the positive anomalies in precipitation as found by
+@hendon2007. A\nobreakdash-SAM appears related to positive precipitation
+anomalies in the West coast of Southeastern Australia (Figure
+\@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the
+anomalous westerly circulation transporting moist air to the continent from the
+Indian Ocean.
+
+The seasonal-based regressions show statistically significant anomalies only in
+Spring, when positive SAM is associated with positive precipitation anomalies in
+Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5). In this trimester,
+S\nobreakdash-SAM seems to be associated with positive precipitation anomalies
+in a relatively reduced area of the East Coast (Figure
+\@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies
+related with positive A\nobreakdash-SAM affect all Eastern Australia (Figure
+\@ref(fig:pp-regr-oceania)b.5).
+
+In Summer, positive SAM index is associated with positive precipitation
+anomalies in Western and Eastern Australia, particularly in the North East
+(Figure \@ref(fig:pp-regr-oceania)a.2). The Eastern part being dominated by the
+relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM. In
+Autumn, the regression with SAM shows positive anomalies in the North, similar
+to Summer, and a broad area of positive anomalies in the North-East to
+South-West direction. This structure seems to be associated with the
+S\nobreakdash-SAM, while the Northern positive values are associated with the
+A\nobreakdash-SAM. In Winter we see the same NE to SW aligned anomaly (although
+with much reduced amplitude) that is also present only in relation with the
+S\nobreakdash-SAM. None of these regression coefficients are statistically
+significant at the 95% level. The Spring signal is broadly consistent with
+@hendon2007, but whereas they also detected a strong signal in Summer, Figure
+\@ref(fig:pp-regr-oceania)a.2 shows no statistically significant association
+(although the coefficients have the consistent sign).
+
+(ref:pp-rer-america-cap) Same as Figure \@ref(fig:pp-regr-oceania) but for South
+America.
 
@@
 
-In South America (Figure \@ref(fig:pp-regr-america)), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile, and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
-Figure \@ref(fig:pp-regr-america)b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with A\nobreakdash-SAM, that in southern Chile is related to S\nobreakdash-SAM.
-
-Except Winter, seasonal-based regressions mirror this same pattern.
-Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the SAM.
-The separation of these features between A\nobreakdash-SAM and S\nobreakdash-SAM regression maps is also rather consistent.
-
-The anomalous circulation at 700\ hPa associated with S\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over Southern Chile.
-This leads to reduced influx of moist air from the Pacific Ocean which is the main source of precipitable water in that region [e.g.
-@garreaud2007].
-On the other hand, the anomalous circulation associated with positive values of A\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
-This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region [@silvestri2009;@zamboni2010].
-This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events [@rosso2018].
-There is a small area of significant positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the A\nobreakdash-SAM.
+In South America (Figure \@ref(fig:pp-regr-america)), regression using all
+seasons shows that positive SAM is associated with statistically significant
+negative precipitation anomalies in Southeastern South America (SESA) and
+Southern Chile, and non-significant positive anomalies in South Brazil, near the
+South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
+Figure \@ref(fig:pp-regr-america)b.1 and c.1 show that while the signal over
+SESA and southern Brazil is associated with A\nobreakdash-SAM, that in southern
+Chile is related to S\nobreakdash-SAM.
+
+Except Winter, seasonal-based regressions mirror this same pattern. Even if not
+statistically significant, they all show negative values in SESA and southern
+Chile along with positive values in southern Brazil in relation with the SAM.
+The separation of these features between A\nobreakdash-SAM and S\nobreakdash-SAM
+regression maps is also rather consistent.
+
+The anomalous circulation at 700 hPa associated with S\nobreakdash-SAM (Figure
+\@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over Southern
+Chile. This leads to reduced influx of moist air from the Pacific Ocean which is
+the main source of precipitable water in that region [e.g. @garreaud2007]. On
+the other hand, the anomalous circulation associated with positive values of
+A\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is
+anticyclonic in the South and cyclonic in the North. This promotes anomalous
+South-Easterly flow over SESA, which inhibits the flow of the South America
+Low-Level Jet to the region [@silvestri2009; @zamboni2010]. This same pattern
+was found to be associated with increased precipitation in Southern Brazil
+during SACZ events [@rosso2018]. There is a small area of significant positive
+precipitation anomalies with the SAM near central Argentina which is also
+present in the station-based analysis by @gillett2006 that is explained by the
+A\nobreakdash-SAM.
 
@@
 
-In this study we characterise the temporal and spatial variability of the zonally symmetric and asymmetric structure of the SAM.
-By projecting monthly geopotential fields at each level with the corresponding asymmetric and symmetric pattern, we created two indices for representing the zonally asymmetric and symmetric contributions of the SAM respectively.
-
-The A\nobreakdash-SAM index correlates strongly with the S\nobreakdash-SAM index.
-In the troposphere, this correlation is maximum at zero lag, while in the stratosphere is maximised with the A\nobreakdash-SAM leading the S\nobreakdash-SAM by one month.
-Since most indices of the SAM are calculated using surface or near-surface conditions, this result would suggest that they might not be sensitive to the most dramatic changes in SAM variability.
-
-The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
-There is evidence of influence between the QBO and the Northern Annular Mode [e.g.
-@holton1980;@watson2014;@zhang2020], so it is not unlikely that the SAM would be similarly affected.
-However, establishing this link would require further research.
-
-We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies [e.g.
-@fogt2020 and references therein] for low levels.
-We show that these trends maximise at 100\ hPa, and are explained by the zonally symmetric component.
-We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the SAM index.
-We find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
-This is in contrast to @fogt2012 who made the evaluation for the period 1958 -- 2001.
-This discrepancy might be due either due to differences in methodology or analysed period.
-
-In the troposphere, the spatial patterns of geopotential associated with the S\nobreakdash-SAM index are much closer to being fully annular than the patterns associated with the SAM index.
-The A\nobreakdash-SAM index, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific.
-This pattern extends vertically in the troposphere but its maximum is located at 250\ hPa, which also could suggest that surface-based indices are not optimal for capturing this variability.
-This wave-3 pattern is similar to the Pacific-South American Pattern, linked to ENSO variability.
-We found that the significant correlation that exists between the SAM index and the Oceanic Niño Index over this period is captured entirely by the A\nobreakdash-SAM index.
-This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the A\nobreakdash-SAM index could be a useful measure to further study that relationship.
-
-Temperature anomalies associated with the SAM broadly show a pattern of negative anomalies at polar latitudes surrounded by positive anomalies, but with many deviations from symmetry.
-The A\nobreakdash-SAM index explains a big portion of these deviations.
-In particular, the positive phase of A\nobreakdash-SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO\nobreakdash-SAM relationship.
-These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [e.g.
-@cai2020a].
-
-In Australia the SAM is associated with positive precipitation anomalies in South East and this is explained by the S\nobreakdash-SAM.
-However, the A\nobreakdash-SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
-In South America, precipitation anomalies associated with the SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
-These features are cleanly separated between A\nobreakdash-SAM and S\nobreakdash-SAM.
-S\nobreakdash-SAM explains the negative anomalies in Southern Chile and A\nobreakdash-SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
+In this study we characterise the temporal and spatial variability of the
+zonally symmetric and asymmetric structure of the SAM. By projecting monthly
+geopotential fields at each level with the corresponding asymmetric and
+symmetric pattern, we created two indices for representing the zonally
+asymmetric and symmetric contributions of the SAM respectively.
+
+The A\nobreakdash-SAM index correlates strongly with the S\nobreakdash-SAM
+index. In the troposphere, this correlation is maximum at zero lag, while in the
+stratosphere is maximised with the A\nobreakdash-SAM leading the
+S\nobreakdash-SAM by one month. Since most indices of the SAM are calculated
+using surface or near-surface conditions, this result would suggest that they
+might not be sensitive to the most dramatic changes in SAM variability.
+
+The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might
+point to a link between the SAM and the Quasi Biennial Oscillation. There is
+evidence of influence between the QBO and the Northern Annular Mode [e.g.
+@holton1980; @watson2014; @zhang2020], so it is not unlikely that the SAM would
+be similarly affected. However, establishing this link would require further
+research.
+
+We observe a positive trend of SAM in Summer and Autumn, as was documented by
+previous studies [e.g. @fogt2020 and references therein] for low levels. We show
+that these trends maximise at 100 hPa, and are explained by the zonally
+symmetric component. We also find a statistically significant positive trend in
+the Symmetric component of the SAM in the stratosphere that is not evident in
+the SAM index. We find some evidence of the SAM becoming more zonally
+asymmetric, as there is a slight positive trend in the variance explained by the
+as the A\nobreakdash-SAM. This is in contrast to @fogt2012 who made the
+evaluation for the period 1958 -- 2001. This discrepancy might be due either due
+to differences in methodology or analysed period.
+
+In the troposphere, the spatial patterns of geopotential associated with the
+S\nobreakdash-SAM index are much closer to being fully annular than the patterns
+associated with the SAM index. The A\nobreakdash-SAM index, on the other hand,
+describes a wave-3 pattern with maximum amplitude in the Pacific. This pattern
+extends vertically in the troposphere but its maximum is located at 250 hPa,
+which also could suggest that surface-based indices are not optimal for
+capturing this variability. This wave-3 pattern is similar to the Pacific-South
+American Pattern, linked to ENSO variability. We found that the significant
+correlation that exists between the SAM index and the Oceanic Niño Index over
+this period is captured entirely by the A\nobreakdash-SAM index. This suggests
+that ENSO is linked to SAM exclusively through the variability in the latter's
+asymmetric component and thus, the A\nobreakdash-SAM index could be a useful
+measure to further study that relationship.
+
+Temperature anomalies associated with the SAM broadly show a pattern of negative
+anomalies at polar latitudes surrounded by positive anomalies, but with many
+deviations from symmetry. The A\nobreakdash-SAM index explains a big portion of
+these deviations. In particular, the positive phase of A\nobreakdash-SAM is
+associated with colder temperatures over Southern Brazil, South Africa and
+Southern Australia, as well as the negative anomalies in the equatorial Pacific
+consistent with the ENSO\nobreakdash-SAM relationship. These negative anomalies
+are particularly clear in the DJF and SON trimesters, which include the months
+in which the ENSO teleconnection is more active [e.g. @cai2020a].
+
+In Australia the SAM is associated with positive precipitation anomalies in
+South East and this is explained by the S\nobreakdash-SAM. However, the
+A\nobreakdash-SAM is associated with a small area of positive precipitation
+anomalies in the Eastern Coast of West Australia, maybe due to advection of
+moist air from the Indian Ocean. In South America, precipitation anomalies
+associated with the SAM are negative both in Southern Chile and Southeastern
+South America, and positive in Southern Brazil. These features are cleanly
+separated between A\nobreakdash-SAM and S\nobreakdash-SAM. S\nobreakdash-SAM
+explains the negative anomalies in Southern Chile and A\nobreakdash-SAM, the
+negative-positive dipole between Southeastern South America and Southern Brazil.
 Individual seasons mostly follow this pattern.
 
-@silvestri2009 suggests that precipitation impacts linked to the SAM changed strongly before and after 1980.
-In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in others in the earlier period.
-The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s [@fogt2006;@clem2013].
-Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the results presented here are most likely period-dependent.
-The decadal variations of the A\nobreakdash-SAM should be focus of future studies.
-
-By separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
-This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modelled as highly zonally symmetric, while the impact of increased concentration of greenhouse gases has also a zonally asymmetric component [@arblaster2006].
-
+@silvestri2009 suggests that precipitation impacts linked to the SAM changed
+strongly before and after 1980. In particular, the negative relationship with
+precipitation in South America was absent in some areas and switched sign in
+others in the earlier period. The correlation between ENSO and SAM is similarly
+non-stationary, also changing sign before the 1980s [@fogt2006; @clem2013].
+Seeing as both the ENSO\nobreakdash-SAM relationship and most of the
+precipitation impacts in South America are captured by the A\nobreakdash-SAM,
+the results presented here are most likely period-dependent. The decadal
+variations of the A\nobreakdash-SAM should be focus of future studies.
+
+By separating the zonally symmetric and zonally symmetric SAM signals, we show
+that the asymmetric component of the SAM has its unique variability, trends and
+impacts. This is particularly important in the context of a changing climate, as
+the impact on the SAM of ozone recovery is modelled as highly zonally symmetric,
+while the impact of increased concentration of greenhouse gases has also a
+zonally asymmetric component [@arblaster2006].
+
+```{=tex}
 \begin{acknowledgements}
@@
 \end{acknowledgements}
-
-# Declarations {-}
-
-## Funding {-}
+```
+# Declarations {.unnumbered}
+
+## Funding {.unnumbered}
 
@@
 
-## Conflicts of interest {-}
+## Conflicts of interest {.unnumbered}
 
@@
 
-A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at https://github.com/eliocamp/asymsam.
+A version-controlled repository of the code used to create this analysis,
 including the code used to download the data can be found at
@@
 
-(ref:a4-cap) Same as \@ref(fig:A3) but for 700\ hPa geopotential height.
+(ref:a4-cap) Same as \@ref(fig:A3) but for 700 hPa geopotential height.
 

### 2021-04-27 analysis/paper/asymsam.Rmd
Cambia a line wraping por oración.

--- 2021-04-26 analysis/paper/paper.Rmd
@@
 
-The Southern Annular Mode (SAM) is the main mode of variability in the Southern
-Hemisphere extratropical circulation [@rogers1982] on daily, monthly, and
-decadal timescales [@baldwin2001a; @fogt2006] and exerts an important influence
-on temperature and precipitation anomalies, and sea ice concentration [e.g.
-@fogt2020]. Its positive phase is usually described as anomalously low pressures
-over Antarctica surrounded by a ring of anomalous high pressures in
-middle-to-high latitudes.
-
-Most authors describe the SAM as a zonally symmetric pattern, a fact that is
-reflected not only in its name, but also in the various methods used to
-characterise it. Of the several different indices presented in the literature,
-many of them are based on zonal means of sea level pressure or geopotential
-height [@ho2012]. @gong1999 defined the SAM index as the zonal mean sea level
-pressure difference between 40\degree S and 60\degree S, which is also the
-definition used by the station-based index in @marshall2003. @baldwin2009
-proposed defining the Northern and Annular modes as the leading EOF of the
-zonally averaged geopotential height at each level.
-
-Even though these indices are based on zonal averages, their associated
-geopotential height spatial anomalies contain noticeable deviations from zonal
-symmetry, particularly in the Pacific Ocean region. The zonal asymmetries have
-not been widely studied, but previous work suggest that they strongly modulate
-the regional impacts of the SAM [@fan2007; @silvestri2009; @fogt2012;
-@rosso2018]. The fact that the SAM is not entirely zonally symmetric hinders our
-ability to reconstruct its historical variability prior to the availability of
-dense observations in the Southern Hemisphere [@jones2009].
-
-Some of the variability associated with the zonal asymmetries of the SAM seems
-to be forced by the tropics. ENSO-like variability affects the Southern
-Hemisphere extratopics through the Rossby wave trains [@mo1987; @kidson1988;
-@karoly1989] which project strongly onto the zonal anomalies associated with the
-SAM in the Pacific sector. Moreover, tropical influences on the SAM have been
-observed [@fan2007; @fogt2011; @clem2013]. @fan2007 computed SAM indices of the
-Western and the Eastern Hemispheres separately and found that they were much
-more correlated to each other if the (linear) signal of the ENSO was removed.
-
-Positive trends in the SAM have been documented by various researchers using
-different indices, mostly on austral Summer and Autumn [e.g. @fogt2020 and
-references therein]. It is thought that these trends are driven primarily by
-stratospheric ozone depletion and the increase in greenhouse gases, and
-understood in the context of zonal mean variables [@marshall2004; @gillett2005;
-@arblaster2006; @gillett2013]. However, it's not clear yet how or if the
-asymmetric SAM component responds to these forcings, or how its variability
-alters the observed trends.
-
-The impact of the zonally asymmetric component of the SAM at regional scales has
-not been studied in detail yet. The positive phase of the SAM is associated with
-colder-than-normal temperatures over Antarctica and warmer-than-normal
-temperatures at higher latitudes [@jones2019] (and vice versa for negative SAM).
-But there are significant deviations from this zonal mean response, notably in
-the Antarctic Peninsula and the South Atlantic [@fogt2012]. The SAM-related
-signal on precipitation anomalies behaves similarly, although with even greater
-deviation from zonal symmetry [@lim2016]. The importance of zonal asymmetries of
-the SAM on these impacts have been studied in certain regions. For example, the
-SAM-precipitation relationship in Southeastern South America can be explained by
-the Pacific-South American (PSA)-like zonally asymmetric circulation associated
-with the SAM [@silvestri2009; @rosso2018]. @fan2007 also found that
-precipitation in East Asia was impacted by the variability of only the Western
-Hemisphere part of the SAM.
-
-The study of the temporal variability of the asymmetric component of the SAM has
-not received much attention except for @fogt2012. This study provides evidences
-for the relevance of the SAM's asymmetric component. However, their conclusions
-are based on composites of positive and negative SAM events including a small
-number of cases unevenly distributed among years with and without satellite
-information. The latter is particularly important due to the inhomogeneities in
-reanalysis products prior to the satellite era and the possible change in the
-asymmetric structure of the SAM [@silvestri2009]. Moreover, @fogt2012 studied
-the zonal asymmetric component of the SAM only in sea level pressure. Zonal
-asymmetries in the SAM spatial pattern are fairly barotropic throughout the
-troposphere, but they change dramatically in the stratosphere [@baldwin2009].
-
-Our objective is, then, to describe the zonally asymmetric and symmetric
-components of the SAM variability. We first propose a methodology that provides
-for each level, two indices which aim to capture independently the variability
-of the symmetric and asymmetric SAM component respectively. Their vertical
-structure and coherence, temporal variability and trends are consequently
-assessed. We then study the spatial patterns described by the variability
-exclusive to each index focusing on 50 hPa as representing the stratosphere and
-700 hPa as representing the troposphere. Finally, the relationships of the SAM
-at 700 hPa with temperature and precipitation anomalies are investigated.
-
-In the Section \@ref(methods) we describe the methods. In Section
-\@ref(temporal) we describe the temporal variability and vertical coherence of
-the indices. In Section \@ref(spatial), we analyse the spatial patterns of
-geopotential height associated with them. In Section \@ref(impacts), we study
-their relationship with surface-level temperature and precipitation.
+The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] on daily, monthly, and decadal timescales [@baldwin2001a; @fogt2006] and exerts an important influence on temperature and precipitation anomalies, and sea ice concentration [e.g.
+@fogt2020].
+Its positive phase is usually described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.
+
+Most authors describe the SAM as a zonally symmetric pattern, a fact that is reflected not only in its name, but also in the various methods used to characterise it.
+Of the several different indices presented in the literature, many of them are based on zonal means of sea level pressure or geopotential height [@ho2012].
+@gong1999 defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 65\degree S, which is also the definition used by the station-based index in @marshall2003.
+@baldwin2009 proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
+
+Even though these indices are based on zonal averages, their associated geopotential height spatial anomalies contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
+The zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM [@fan2007; @silvestri2009; @fogt2012; @rosso2018].
+The fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].
+
+Some of the variability associated with the zonal asymmetries of the SAM seems to be forced by the tropics.
+ENSO-like variability affects the Southern Hemisphere extratopics through the Rossby wave trains [@mo1987; @kidson1988; @karoly1989] which project strongly onto the zonal anomalies associated with the SAM in the Pacific sector.
+Moreover, tropical influences on the SAM have been observed [@fan2007; @fogt2011; @clem2013].
+@fan2007 computed SAM indices of the Western and the Eastern Hemispheres separately and found that they were much more correlated to each other if the (linear) signal of the ENSO was removed.
+
+Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn [e.g. @fogt2020 and references therein].
+It is thought that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004; @gillett2005; @arblaster2006; @gillett2013].
+However, it's not clear yet how or if the asymmetric SAM component responds to these forcings, or how its variability alters the observed trends.
+
+The impact of the zonally asymmetric component of the SAM at regional scales has not been studied in detail yet.
+The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at lower latitudes [@jones2019] (and vice versa for negative SAM).
+But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012].
+The SAM-related signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
+The SAM-precipitation relationship in Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM [@silvestri2009; @rosso2018].
+@fan2007 also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
+
+The study of the temporal variability of the asymmetric component of the SAM has not received much attention except for @fogt2012.
+This study provides evidences for the relevance of the SAM's asymmetric component.
+However, their conclusions are based on composites of positive and negative SAM events including a small number of cases unevenly distributed among years with and without satellite information.
+The latter is particularly important due to the inhomogeneities in reanalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].
+Moreover, @fogt2012 studied the zonal asymmetric component of the SAM only in sea level pressure.
+Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere [@baldwin2009].
+
+Our objective is, then, to describe the zonally asymmetric and symmetric components of the SAM variability.
+We first propose a methodology that provides for each level, two indices which aim to capture independently the variability of the symmetric and asymmetric SAM component respectively.
+Their vertical structure and coherence, temporal variability and trends are consequently assessed.
+We then study the spatial patterns described by the variability exclusive to each index focusing on 50 hPa as representing the stratosphere and 700 hPa as representing the troposphere.
+Finally, the relationships of the SAM at 700 hPa with temperature and precipitation anomalies are investigated.
+
+In the Section \@ref(methods) we describe the methods.
+In Section \@ref(temporal) we describe the temporal variability and vertical coherence of the indices.
+In Section \@ref(spatial), we analyse the spatial patterns of geopotential height associated with them.
+In Section \@ref(impacts), we study their relationship with surface-level temperature and precipitation.
 
@@
 
-```{r compute\nobreakdash-SAM}
+```{r compute-SAM}
+set.seed(43)
 SAM <- hgt[lat <= -20] %>% 
@@
 
+```{r test-aao1}
+# test that my SAM have the correct sign.
+aao <- rsoi::download_aao(TRUE, data_path("raw", "aao.csv")) %>% 
+  as.data.table() %>% 
+  .[, .(time = lubridate::as_datetime(Date), aao = AAO)] %>% 
+  .[time %between% as.Date(c("1979-01-01", "2018-12-01"))]
+
+cor_aao <- SAM$eof[[1]]$left %>% 
+  .[aao, on = .NATURAL] %>% 
+  na.omit() %>% 
+  .[, cor(hgt, aao)]
+
+if (sign(cor_aao) != 1) {
+  stop("AAO and full index are not positively correlated!")
+}
+```
+
 # Methods {#methods}
@@
 
-We used monthly geopotential height at 2.5\degree longitude by
-2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels
-from ERA5 [@hersbach2020] for the period 1979 to 2018. We restrict our analysis
-to the post-satellite era to avoid any confounding factors arising from the
-incorporation of satellite observations.
-
-Temperature data used are from NOAA's Merged Land Ocean Global Surface
-Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface
-air temperature and sea surface (water) temperature analysis into a monthly
-global grid 5\degree longitude by 5\degree latitude. For precipitation data we
-used monthly data from the CPC Merged Analysis of Precipitation [\@xie1997],
-with a 2.5\\degree\\ resolution in latitude and longitude. This rainfall gridded
-dataset is based on information from different sources such as rain gauge
-observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and
-it is available since 1979 to present.
+We used monthly geopotential height at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
+We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations.
+
+Temperature data used are from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid 5\degree longitude by 5\degree latitude.
+For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [\@xie1997], with a 2.5\\degree\\ resolution in latitude and longitude.
+This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
 
@@
 
-Traditionally, the SAM is defined as the leading empirical orthogonal mode (EOF)
-of sea-level pressure or geopotential height anomalies at low levels [@ho2012].
-Following @baldwin2001, we extend that definition vertically and use the term
-SAM to refer to the leading EOF of the monthly anomalies of geopotential height
-south of 20\degree S at each level. We performed EOFs by computing the Singular
-Value Decomposition of the data matrix consisting in `r dims$t` rows and
-`r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points
-of latitude). We weighted the values by the square root of the cosine of
-latitude to account for the non-equal area of each gridpoint [@chung1999]. We
-consider in the EOF analysis all months together without dividing by seasons.
-
-To separate the zonally symmetric and asymmetric components of the SAM, we
-computed the zonal mean and anomalies of the full SAM spatial pattern, as shown
-in Figure \@ref(fig:method) at 700 hPa. The full spatial signal
-($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric
-($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric
-($[\mathrm{EOF_1}](\lambda, \phi)$) components. We then compute the SAM index,
-Asymmetric SAM index (A\nobreakdash-SAM) and Symmetric SAM (S\nobreakdash-SAM)
-indices as the coefficients of the regression of each monthly geopotential
-height field on the respective patterns (weighting by the cosine of latitude).
-The three indices are then normalized by dividing them by the standard deviation
-of the SAM index at each level. As a result, the magnitudes between indices are
-comparable. However, only SAM index has unit standard deviation per definition.
-The explained variance of each pattern is used as an indicator of the degree of
-zonally symmetry or asymmetry of each monthly field. To quantify the coherence
-between temporal series corresponding to different indices or the same index at
-different levels, we computed the temporal correlation between them.
+Traditionally, the SAM is defined as the leading empirical orthogonal mode (EOF) of sea-level pressure or geopotential height anomalies at low levels [@ho2012].
+Following @baldwin2001, we extend that definition vertically and use the term SAM to refer to the leading EOF of the monthly anomalies of geopotential height south of 20\degree S at each level.
+We performed EOFs by computing the Singular Value Decomposition of the data matrix consisting in `r dims$t` rows and `r dims$x*dims$y` columns (`r dims$x` points of longitude and `r dims$y` points of latitude).
+We weighted the values by the square root of the cosine of latitude to account for the non-equal area of each gridpoint [@chung1999].
+We consider in the EOF analysis all months together without dividing by seasons.
+
+To separate the zonally symmetric and asymmetric components of the SAM, we computed the zonal mean and anomalies of the full SAM spatial pattern, as shown in Figure \@ref(fig:method) at 700 hPa.
+The full spatial signal ($\mathrm{EOF_1}(\lambda, \phi)$) is the sum of the zonally asymmetric ($\mathrm{EOF_1^*}(\lambda, \phi)$) and symmetric ($[\mathrm{EOF_1}](\lambda, \phi)$) components.
+We then compute the SAM index, Asymmetric SAM index (A\nobreakdash-SAM) and Symmetric SAM (S\nobreakdash-SAM) indices as the coefficients of the regression of each monthly geopotential height field on the respective patterns (weighting by the cosine of latitude).
+The three indices are then normalized by dividing them by the standard deviation of the SAM index at each level.
+As a result, the magnitudes between indices are comparable.
+However, only SAM index has unit standard deviation per definition.
+The explained variance of each pattern is used as an indicator of the degree of zonally symmetry or asymmetry of each monthly field.
+To quantify the coherence between temporal series corresponding to different indices or the same index at different levels, we computed the temporal correlation between them.
 
@@
 
-The method assumes linearity in the asymmetric component of the SAM. That means
-that zonal asymmetries associated with positive SAM phase (SAM+) are almost
-opposite in sign and of the same magnitude to the ones associated with negative
-SAM phase (SAM-). @fogt2012's composites (their Figure 4) suggest that this
-might not be entirely valid, although much of that apparent non-linearity could
-be due to the heterogenous nature of the selected years for constructing the
-composites. To test this assumption, we computed seasonal composites of zonal
-anomalies of geopotential height for SAM+ and SAM- (defined as months in which
-the SAM index is greater than 1 standard deviation and lower than minus 1
-standard deviation, respectively) for the period from 1979 to 2018 at the
-700 hPa and 50 hPa levels (Figures \@ref(fig:A3) and \@ref(fig:A4)). In all
-seasons and both levels, SAM+ composites are similar to SAM- in structure but
-with the opposite sign. Spatial correlations between composites for each season
-are high. The method considered in this study seems then a reasonable
-approximation of the phenomenon.
-
-By performing the EOF analysis using data for all months we are assuming that
-the zonally asymmetric structure of the SAM is the same at all seasons. The
-latter was assessed by computing geopotential height zonal anomalies by
-projecting the first EOF of each season independently. The following seasons
-were considered -- December to February (DJF), March to May (MAM), June to
-August (JJA) and September to November (SON). Results are very similar to each
-other in the troposphere (Figure \@ref(fig:A5), row 1) and show spatial
-correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON). In the
-stratosphere (Figure \@ref(fig:A5), row 2), patterns are similar for all seasons
-except DJF, when the wave-1 zonal anomalies are rotated 90\degree in comparison
-with the rest of the year. Spatial correlations in the stratosphere are between
--0.24 (DJF with SON) and 0.95 (MAM with JJA). Therefore, the results confirm
-that the zonal asymmetric structure of the SAM is very similar throughout most
-of the year with the exception of DJF in the stratosphere.
-
-The method also assumes that the zonally asymmetric pattern of the SAM remains
-stationary along the period considered. @silvestri2009 suggest that this might
-not be the case between 1958 and 2004. Zonal asymmetric patterns of SAM were
-computed for the two halves of the period (1979 to 1998 and 1999 to 2018)
-respectively. They show no systematic change neither in the stratosphere nor in
-the troposphere (Figure \@ref(fig:A6)).
+The method assumes linearity in the asymmetric component of the SAM.
+That means that zonal asymmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
+@fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
+To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700 hPa and 50 hPa levels (Figures \@ref(fig:A3) and \@ref(fig:A4)).
+In all seasons and both levels, SAM+ composites are similar to SAM- in structure but with the opposite sign.
+Spatial correlations between composites for each season are high.
+The method considered in this study seems then a reasonable approximation of the phenomenon.
+
+By performing the EOF analysis using data for all months we are assuming that the zonally asymmetric structure of the SAM is the same at all seasons.
+The latter was assessed by computing geopotential height zonal anomalies by projecting the first EOF of each season independently.
+The following seasons were considered -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
+Results are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 2) and show spatial correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
+In the stratosphere (Figure \@ref(fig:A5), row 1), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree in comparison with the rest of the year.
+Spatial correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
+Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year.
+The only outlier being DJF, which shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref{fig:A3}), which is consistent with @fogt2020.
+With this in mind, a
+
+The method also assumes that the zonally asymmetric pattern of the SAM remains stationary along the period considered.
+@silvestri2009 suggest that this might not be the case between 1958 and 2004.
+Zonal asymmetric patterns of SAM were computed for the two halves of the period (1979 to 1998 and 1999 to 2018) respectively.
+They show no systematic change neither in the stratosphere nor in the troposphere (Figure \@ref(fig:A6)).
 
@@
 
-We performed linear regressions to quantify the association between the SAM
-indices and other variables. Moreover, we apply multiple linear regression
-analysis to describe the combined influence of both A\nobreakdash-SAM and
-S\nobreakdash-SAM. To obtain the linear coefficients of a variable $X$
-(geopotential, temperature, precipitation, etc...) with A\nobreakdash-SAM and
-S\nobreakdash-SAM we fit the equation
+We performed linear regressions to quantify the association between the SAM indices and other variables.
+Moreover, we apply multiple linear regression analysis to describe the combined influence of both A\nobreakdash-SAM and S\nobreakdash-SAM.
+To obtain the linear coefficients of a variable $X$ (geopotential, temperature, precipitation, etc...) with A\nobreakdash-SAM and S\nobreakdash-SAM we fit the equation
 
@@
 
-where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time,
-$\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and
-$\epsilon$ are the constant and error terms. From this equation, $\alpha$
-represents the (linear) association of $X$ with the variability of
-A\nobreakdash-SAM that is not explained by the variability of S\nobreakdash-SAM;
-i.e. it is proportional to the partial correlation of $X$ and A\nobreakdash-SAM,
-controlling for the effect of S\nobreakdash-SAM, and vice versa for $\beta$.
-When performing a separate regression for each trimester (DJF, MAM, JJA, SON),
-we averaged the relevant variables seasonally for each year and trimester before
-computing the regression.
-
-Statistical significance for regression fields were evaluated adjusting p-values
-by controlling for the False Discovery Rate [@benjamini1995; @wilks2016] to
-avoid misleading results from the high number of regressions [@walker1914;
-@katz1991].
-
-Linear trends were computed by Ordinary Least Squares and the 95% confidence
-interval was computed assuming a t-distribution with the appropriate residual
-degrees of freedom. To the amplitude of the zonal waves is defined through
-computing the Fourier transform of the spatial field at each latitude circle.
-
-We computed density probability estimates using a Gaussian kernel of optimal bin
-width according to @sheather1991.
+where $\lambda$ and $\phi$ are the longitude and latitude, $t$ is the time, $\alpha$ and $\beta$ are the linear regression coefficients, $X_0$ and $\epsilon$ are the constant and error terms.
+From this equation, $\alpha$ represents the (linear) association of $X$ with the variability of A\nobreakdash-SAM that is not explained by the variability of S\nobreakdash-SAM; i.e.
+it is proportional to the partial correlation of $X$ and A\nobreakdash-SAM, controlling for the effect of S\nobreakdash-SAM, and vice versa for $\beta$.
+When performing a separate regression for each trimester (DJF, MAM, JJA, SON), we averaged the relevant variables seasonally for each year and trimester before computing the regression.
+
+Statistical significance for regression fields were evaluated adjusting p-values by controlling for the False Discovery Rate [@benjamini1995; @wilks2016] to avoid misleading results from the high number of regressions [@walker1914; @katz1991].
+
+Linear trends were computed by Ordinary Least Squares and the 95% confidence interval was computed assuming a t-distribution with the appropriate residual degrees of freedom.
+To the amplitude of the zonal waves is defined through computing the Fourier transform of the spatial field at each latitude circle.
+
+We computed density probability estimates using a Gaussian kernel of optimal bin width according to @sheather1991.
 
@@
 
-We performed all analysis in this paper using the R programming language
-[@rcoreteam2020], using the data.table package [@dowle2020] and the metR package
-[@campitelli2020]. All graphics are made using ggplot2 [@wickham2009]. We
-downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and
-indices of the ENSO with the rsoi package [@albers2020]. The paper was rendered
-using knitr and rmarkdown [@xie2015; @allaire2019].
+We performed all analysis in this paper using the R programming language [@rcoreteam2020], using the data.table package [@dowle2020] and the metR package [@campitelli2020].
+All graphics are made using ggplot2 [@wickham2009].
+We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
+The paper was rendered using knitr and rmarkdown [@xie2015; @allaire2019].
 
@@
   indices <- readRDS(indices_file)
-  indices[, `:=`(estimate = -estimate, 
-                 estimate_norm = -estimate_norm)]
+  indices[, `:=`(estimate = estimate, 
+                 estimate_norm = estimate_norm)]
 } else {
@@
 
-```{r test-aao}
+```{r test-aao2}
 # test that my indices hace the correct sign. The Full index needs to be **positively** correlated
 # with the AAO index.
-aao <- rsoi::download_aao(TRUE, data_path("raw", "aao.csv")) %>% 
-  as.data.table() %>% 
-  .[, .(time = lubridate::as_datetime(Date), aao = AAO)]
-
 cor_aao <- indices[term == "full" & lev == 700] %>% 
@@
 
-We first asses the temporal evolution of A\nobreakdash-SAM and
-S\nobreakdash-SAM. Figure \@ref(fig:asymsam-timeseries) shows the corresponding
-time series for 700 hPa and 50 hPa and their corresponding density estimates. We
-selected these two levels as representative of the tropospheric and
-stratospheric variability respectively. As it is shown below, the variabilities
-of both indices are highly coherent within each atmospheric layer, therefore is
-reasonable to take one level as representative of each layer.
-
-Month-to-month variability is evident for both indices, with noisy variations in
-the low frequencies. At first glance the series can be distinguished by their
-distributions. Compared to the tropospheric indices, the stratospheric indices
-are much more long-tailed; that is, extreme values (both negative and positive)
-abound. The A\nobreakdash-SAM series have both more variability in the higher
-frequencies than the S\nobreakdash-SAM series.
-
-The stratospheric S\nobreakdash-SAM varies strongly with a two-year period,
-which can be seen by spectral analysis (Figure \@ref(fig:A2)). This might
-suggest a link between stratospheric SAM variability and the Quasi-Biennial
-Oscillation [@baldwin2001b]. A local peak at 2 years years is discernible in the
-periodogram of the tropospheric S\nobreakdash-SAM, although it's not
-statistically significant. In the troposphere the most significant peak of
-variability is found in A\nobreakdash-SAM at around `r 12*.3` months.
+We first asses the temporal evolution of A\nobreakdash-SAM and S\nobreakdash-SAM.
+Figure \@ref(fig:asymsam-timeseries) shows the corresponding time series for 700 hPa and 50 hPa and their corresponding density estimates.
+We selected these two levels as representative of the tropospheric and stratospheric variability respectively.
+As it is shown below, the variabilities of both indices are highly coherent within each atmospheric layer, therefore is reasonable to take one level as representative of each layer.
+
+Month-to-month variability is evident for both indices, with noisy variations in the low frequencies.
+At first glance the series can be distinguished by their distributions.
+Compared to the tropospheric indices, the stratospheric indices are much more long-tailed; that is, extreme values (both negative and positive) abound.
+The A\nobreakdash-SAM series have both more variability in the higher frequencies than the S\nobreakdash-SAM series.
+
+The stratospheric S\nobreakdash-SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \@ref(fig:A2)).
+This might suggest a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b].
+A local peak at 2 years years is discernible in the periodogram of the tropospheric S\nobreakdash-SAM, although it's not statistically significant.
+In the troposphere the most significant peak of variability is found in A\nobreakdash-SAM at around `r 12*.3` months.
 
@@
 
-From a visual inspection, the A\nobreakdash-SAM and S\nobreakdash-SAM time
-series appear to be correlated. Moreover, looking at the extremes in the
-stratosphere, the S\nobreakdash-SAM series appears to lag the A\nobreakdash-SAM
-series (see, for example, the positive events on late 1987). Figure
-\@ref(fig:cor-lev) shows these correlations along all levels considered, for
-zero and -1 lags. Values of zero-lag correlations between A\nobreakdash-SAM and
-S\nobreakdash-SAM are relatively constant throughout the troposphere,
-fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and
-`r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`. One-month-lag correlations
-are similarly constant but significantly reduced to around
-`r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`. In the stratosphere,
-zero-lag correlations drop to a minimum of
-`r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at
-`r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then increase again
-monotonically with height up to the uppermost level of the reanalysis (although
-results near the top of the models are to be interpreted with care). At the same
-time, one-month-lag correlations increase with height. Therefore, stratospheric
-A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index.
+From a visual inspection, the A\nobreakdash-SAM and S\nobreakdash-SAM time series appear to be correlated.
+Moreover, looking at the extremes in the stratosphere, the S\nobreakdash-SAM series appears to lag the A\nobreakdash-SAM series (see, for example, the positive events on late 1987).
+Figure \@ref(fig:cor-lev) shows these correlations along all levels considered, for zero and -1 lags.
+Values of zero-lag correlations between A\nobreakdash-SAM and S\nobreakdash-SAM are relatively constant throughout the troposphere, fluctuating between `r signif(cors[lag == 0 & lev >= 100, min(acf)], 2)` and `r signif(cors[lag == 0 & lev >= 100, max(acf)], 2)`.
+One-month-lag correlations are similarly constant but significantly reduced to around `r signif(cors[lag == -1 &lev > 500][, mean(acf)], 2)`.
+In the stratosphere, zero-lag correlations drop to a minimum of `r signif(cors[lag == 0 & lev < 100][which.min(acf)]$acf, 2)` at `r cors[lag == 0 & lev < 100][which.min(acf)]$lev` hPa and then increase again monotonically with height up to the uppermost level of the reanalysis (although results near the top of the models are to be interpreted with care).
+At the same time, one-month-lag correlations increase with height.
+Therefore, stratospheric A\nobreakdash-SAM index tends to precede the S\nobreakdash-SAM index.
 (Correlations at lags -5 to 5 are shown in Figure \@ref(fig:A1).)
@@
 
-Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across
-levels for the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices. For the SAM
-(Figure \@ref(fig:cross-correlation)a), high values below 100 hPa reflect the
-vertical (zero-lag) coherency throughout the troposphere. Above 100 hPa,
-correlation between levels does falls off more rapidly, indicating less coherent
-(zero-lag) variability. But correlations between tropospheric and
-lower-to-middle stratospheric levels are still relatively high (e.g. grater than
-0.4 between tropospheric levels and levels below 30 hPa). A\nobreakdash-SAM and
-S\nobreakdash-SAM (Figure \@ref(fig:cross-correlation)b and c, respectively)
-share similar high level of coherency in the troposphere but they differ in
-their stratospheric behaviour. Stratospheric coherency is stronger for the
-A\nobreakdash-SAM than the S\nobreakdash-SAM. The stratospheric
-S\nobreakdash-SAM seems to connect more strongly to the troposphere than the
-stratospheric A\nobreakdash-SAM.
+Figure \@ref(fig:cross-correlation) shows (zero-lag) cross-correlation across levels for the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices.
+For the SAM (Figure \@ref(fig:cross-correlation)a), high values below 100 hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
+Above 100 hPa, correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
+But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g. greater than 0.4 between tropospheric levels and levels below 30 hPa).
+A\nobreakdash-SAM and S\nobreakdash-SAM (Figure \@ref(fig:cross-correlation)b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
+Stratospheric coherency is stronger for the A\nobreakdash-SAM than the S\nobreakdash-SAM.
+The stratospheric S\nobreakdash-SAM seems to connect more strongly to the troposphere than the stratospheric A\nobreakdash-SAM.
 
@@
 
-The linear trends for each of the indices (SAM, S\nobreakdash-SAM and
-A\nobreakdash-SAM) were evaluated for the complete period 1979 -- 2018 at each
-level for the whole year and separated by trimesters (Figure \@ref(fig:trends)).
-The SAM index presents a statistically positive significant trend (Figure
-\@ref(fig:trends)`r get_tag(term == "full" & season == "Annual")`) that extends
-throughout the troposphere up to about 50 hPa and reaches its maximum value at
-100 hPa. The seasonal trends (rest of Figure \@ref(fig:trends) column a)
-indicate that positive trends are present in Autumn and particularly in Summer,
-where the 100 hPa maximum is much more defined. In Winter and Spring, we detect
-no statistically significant trend. This is consistent with the results of
-previous studies, which find large positive trends in Summer, smaller in Autumn
-and no trends in the other seasons [e.g. @fogt2020 and references therein] using
-indices of the SAM based on surface or near-surface circulation.
-
-By separating the SAM signal in its asymmetric and symmetric parts, not only we
-can see that these trends are almost entirely due to the symmetric component
-(column b vs. column c in Figure \@ref(fig:trends)), but in some cases the
-trends become clearer. In Summer, A\nobreakdash-SAM has a statistically non
-significant negative trend in the middle troposphere that obscures the trend in
-the SAM index; as a result, trends computed using only the Symmetric component
-are stronger (compare the shading region in Figure
-\@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and
-`r get_tag(term == "sym" & season == "DJF")`). In Autumn, the S\nobreakdash-SAM
-index reveals a statistically significant positive trend in the stratosphere
-that is not significant using the SAM index.
-
-Figure \@ref(fig:r-squared-trend) shows trends for the explained variance of
-each index. There is no evidence of a significant trend in the stratosphere. In
-the troposphere, there is a positive trend for A\nobreakdash-SAM and no
-significant trend for S\nobreakdash-SAM. This suggest that the SAM has become
-more asymmetric in the period from 1979 to 2018. However, the change is slight,
-around 1% increased explained variance per decade.
+The linear trends for each of the indices (SAM, S\nobreakdash-SAM and A\nobreakdash-SAM) were evaluated for the complete period 1979 -- 2018 at each level for the whole year and separated by trimesters (Figure \@ref(fig:trends)).
+The SAM index presents a statistically positive significant trend (Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
+The seasonal trends (rest of Figure \@ref(fig:trends) column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
+In Winter and Spring, we detect no statistically significant trend.
+This is consistent with the results of previous studies, which find large positive trends in Summer, smaller in Autumn and no trends in the other seasons [e.g.
+@fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.
+
+By separating the SAM signal in its asymmetric and symmetric parts, not only we can see that these trends are almost entirely due to the symmetric component (column b vs.
+column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer.
+In Summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
+In Autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.
+
+Figure \@ref(fig:r-squared-trend) shows trends for the explained variance of each index.
+There is no evidence of a significant trend in the stratosphere.
+In the troposphere, there is a positive trend for A\nobreakdash-SAM and no significant trend for S\nobreakdash-SAM.
+This suggest that the SAM has become more asymmetric in the period from 1979 to 2018.
+However, the change is slight, around 1% increased explained variance per decade.
 
@@
 
-We then computed the spatial regression of geopotential height anomalies on the
-A\nobreakdash-SAM and S\nobreakdash-SAM indices at 700 hPa and 50 hPa levels
-(Figure \@ref(fig:2d-regr)). While regression coefficients in Figure
-\@ref(fig:2d-regr) column a are computed using SAM, the regression coefficients
-in columns b and c of Figure \@ref(fig:2d-regr) are computed using multiple
-regression using the A\nobreakdash-SAM and S\nobreakdash-SAM indices at the same
-time. Thus, they are to be interpreted as the patterns associated with each
-index, removing the variability (linearly) explained by the other one.
-
-In the stratosphere, the spatial pattern associated with the SAM is clearly
-dominated by a zonally symmetric, monopolar structure (Figure
-\@ref(fig:2d-regr)a.1) which is not centred in the South Pole. On the other
-hand, the monopole associated with S\nobreakdash-SAM (Figure
-\@ref(fig:2d-regr)c.1) is more symmetric. Furthermore, the regression pattern of
-A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the
-Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern
-Hemisphere
+We then computed the spatial regression of geopotential height anomalies on the A\nobreakdash-SAM and S\nobreakdash-SAM indices at 700 hPa and 50 hPa levels (Figure \@ref(fig:2d-regr)).
+While regression coefficients in Figure \@ref(fig:2d-regr) column a are computed using SAM, the regression coefficients in columns b and c of Figure \@ref(fig:2d-regr) are computed using multiple regression using the A\nobreakdash-SAM and S\nobreakdash-SAM indices at the same time.
+Thus, they are to be interpreted as the patterns associated with each index, removing the variability (linearly) explained by the other one.
+
+In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \@ref(fig:2d-regr)a.1) which is not centred in the South Pole.
+On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric.
+Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Drake Passage in the Western Hemisphere and Davis Sea in the Eastern Hemisphere
 
@@
 
-In the troposphere, the regression pattern associated with the SAM shows the
-well-known combination of zonally symmetrical annular mode with zonal
-asymmetries in the form of a wave-3 (Figure \@ref(fig:2d-regr)a.2, [@fogt2012]).
-The regression patterns associated with the A\nobreakdash-SAM and
-S\nobreakdash-SAM indices successfully disentangle both structures. While the
-A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure
-\@ref(fig:2d-regr)b.2), the S\nobreakdash-SAM index is associated with an
-annular structure, almost devoid of zonal asymmetries (Figure
-\@ref(fig:2d-regr)c.2). The wave-3 pattern observed in Figure
-\@ref(fig:2d-regr)b.2 is rotated by half a wavelength from the average position
-of the mean wave-3 pattern described by @raphael2004, whose reference locations
-are marked with points in the figure. Indeed, there is no correlation between
-@raphael2004's index and A\nobreakdash-SAM (cor =
-`r signif(cor_raphael$estimate, 2)`, p-value `r report_p(cor_raphael$p.value)`).
-Thus, the tropospheric A\nobreakdash-SAM index represents a zonal displacement
-in the position of the climatological wave-3 pattern.
-
-(ref:wave-amplitude-caption) Amplitude (meters) of zonal waves of the
-geopotential height regression patterns in Figure \@ref(fig:2d-regr) for zonal
-waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the
-amplitude of the zonal mean.
+In the troposphere, the regression pattern associated with the SAM shows the well-known combination of zonally symmetrical annular mode with zonal asymmetries in the form of a wave-3 (Figure \@ref(fig:2d-regr)a.2, [@fogt2012]).
+The regression patterns associated with the A\nobreakdash-SAM and S\nobreakdash-SAM indices successfully disentangle both structures.
+While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \@ref(fig:2d-regr)b.2), the S\nobreakdash-SAM index is associated with an annular structure, almost devoid of zonal asymmetries (Figure \@ref(fig:2d-regr)c.2).
+The wave-3 pattern observed in Figure \@ref(fig:2d-regr)b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern described by @raphael2004, whose reference locations are marked with points in the figure.
+Indeed, there is no correlation between @raphael2004's index and A\nobreakdash-SAM (cor = `r signif(cor_raphael$estimate, 2)`, p-value `r report_p(cor_raphael$p.value)`).
+Thus, the tropospheric A\nobreakdash-SAM index represents a zonal displacement in the position of the climatological wave-3 pattern.
+
+(ref:wave-amplitude-caption) Amplitude (meters) of zonal waves of the geopotential height regression patterns in Figure \@ref(fig:2d-regr) for zonal waves with wavenumber 0, 1, 2, and 3, where wavenumber 0 represents the amplitude of the zonal mean.
 
@@
   ggplot(aes(lat, amplitude)) +
-  geom_line(aes(color = factor(k))) +
-  scale_color_brewer("Wavenumber", palette = "Dark2") +
+  geom_line(aes(color = factor(k)), size = 1.2) +
+  scale_color_brewer("Wavenumber", palette = "Set1") +
   scale_y_continuous("Amplitude") +
@@
 
-The amplitude of the first zonal wavenumbers at each latitude at 50 hPa and
-700 hPa are shown in Figure \@ref(fig:wave-amplitude), where wavenumber zero
-represents the amplitude of the zonal mean. Zonal wave amplitudes of the spatial
-pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column
-`r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at
-both levels. However, zonal waves are important, particularly South of 50\degree
-S, with wavenumber 1 clearly dominating at 50 hPa (Figure
-\@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of
-waves of similar amplitude at 700 hPa (Figure
-\@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`). Figure
-\@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the
-A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere
-(Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while
-in the troposphere it is explained by a combination of zonal waves 3 to 1 in
-decreasing level of importance (Figure
-\@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with
-negligible amplitude of the zonal mean. On the other hand, the S\nobreakdash-SAM
-it is almost entirely explained by the zonal mean at both levels (Figure
-\@ref(fig:wave-amplitude) column c), with little to no contribution from zonal
-waves with wavenumbers 1 to 3.
-
-Looking at Figure \@ref(fig:2d-regr)b.2, it becomes apparent that zonal waves 1
-and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is
-larger in the Western Hemisphere than in the Eastern Hemisphere.
+The amplitude of the first zonal wavenumbers at each latitude at 50 hPa and 700 hPa are shown in Figure \@ref(fig:wave-amplitude), where wavenumber zero represents the amplitude of the zonal mean.
+Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column `r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at both levels.
+However, zonal waves are important, particularly South of 50\degree S, with wavenumber 1 clearly dominating at 50 hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of waves of similar amplitude at 700 hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`).
+Figure \@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean.
+On the other hand, the S\nobreakdash-SAM it is almost entirely explained by the zonal mean at both levels (Figure \@ref(fig:wave-amplitude) column c), with little to no contribution from zonal waves with wavenumbers 1 to 3.
+
+Looking at Figure \@ref(fig:2d-regr)b.2, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
 
@@
 
-To analyse the vertical structure of the geopotential height anomalies
-associated with the A\nobreakdash-SAM index, we show a vertical cross section of
-regressions of mean geopotential height between 65\degree S and 40\degree S for
-the 50 hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)a) and
-for the 700 hPa A\nobreakdash-SAM index (Figure
-\@ref(fig:vertical-regression)b). The geopotential height anomalies associated
-with the stratospheric A\nobreakdash-SAM (Figure
-\@ref(fig:vertical-regression)a) are clearly constrained to the stratosphere,
-which underscores the uncoupling between the stratospheric and tropospheric
-A\nobreakdash-SAM. The vertical structure of this signal tilts about
-60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic
-processes. The signal in the stratosphere maximises near 10 hPa despite using
-the 50 hPa index for the regression.
-
-The tropospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)b) has
-significant signals that extend upwards to the uppermost levels considered. In
-the troposphere, the wave-3 structure is equivalent barotropic with maximum
-amplitude at roughly 250 hPa. The anomalies are larger in the Western
-hemisphere, where they extend into the stratosphere. In the Eastern hemisphere
-the wave-3 signal is smaller and confined to the troposphere while negative
-anomalies dominate in the stratosphere. So, while the tropospheric
-A\nobreakdash-SAM index is associated with stratospheric geopotential anomalies,
-these do not project strongly onto the stratospheric A\nobreakdash-SAM. The
-structures shown in Figure \@ref(fig:vertical-regression) are robust to the
-choice of index level. For any stratospheric (above 100 hPa) index, the
-resulting anomalies are very similar to the wave-1 structure with maximum near
-10 hPa in Figure \@ref(fig:vertical-regression)a. Conversely, for any
-tropospheric (below 100 hPa) index, the result is very similar to Figure
-\@ref(fig:vertical-regression)b. The patterns mainly change in amplitude (not
-shown).
+To analyse the vertical structure of the geopotential height anomalies associated with the A\nobreakdash-SAM index, we show a vertical cross section of regressions of mean geopotential height between 65\degree S and 40\degree S for the 50 hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)a) and for the 700 hPa A\nobreakdash-SAM index (Figure \@ref(fig:vertical-regression)b).
+The geopotential height anomalies associated with the stratospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)a) are clearly constrained to the stratosphere, which underscores the uncoupling between the stratospheric and tropospheric A\nobreakdash-SAM.
+The vertical structure of this signal tilts about 60\degree to the West between 100 hPa and 1 hPa, suggesting baroclinic processes.
+The signal in the stratosphere maximises near 10 hPa despite using the 50 hPa index for the regression.
+
+The tropospheric A\nobreakdash-SAM (Figure \@ref(fig:vertical-regression)b) has significant signals that extend upwards to the uppermost levels considered.
+In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
+The anomalies are larger in the Western hemisphere, where they extend into the stratosphere.
+In the Eastern hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere.
+So, while the tropospheric A\nobreakdash-SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric A\nobreakdash-SAM.
+The structures shown in Figure \@ref(fig:vertical-regression) are robust to the choice of index level.
+For any stratospheric (above 100 hPa) index, the resulting anomalies are very similar to the wave-1 structure with maximum near 10 hPa in Figure \@ref(fig:vertical-regression)a.
+Conversely, for any tropospheric (below 100 hPa) index, the result is very similar to Figure \@ref(fig:vertical-regression)b.
+The patterns mainly change in amplitude (not shown).
 
@@
 
-The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the PSA
-Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with
-the ENSO [@karoly1989]. Indeed, @fogt2011 showed that there is a significant
-relationship between the SAM and the ENSO. The correlation between the SAM and
-the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is
-`r cor_texts[term == "full"]$cor` (p-value
-`r cor_texts[term == "full"]$p.value`). This relationship is captured mainly by
-the A\nobreakdash-SAM, as this index has a partial correlation of
-`r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation`
-(p-value `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the
-ONI, consistent with @fan2007's results, whereas the S\nobreakdash-SAM's partial
-correlation with the ONI is essentially null
-(`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`;
-p-value `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`). The same
-analysis was performed using the Multivariate ENSO Index [@wolter2011] and the
-Southern Oscillation Index [@ropelewski1987], obtaining similar results. The
-latter allows to conclude that these results do not depend on the ENSO index
-used.
+The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the PSA Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
+Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
+The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value `r cor_texts[term == "full"]$p.value`).
+This relationship is captured mainly by the A\nobreakdash-SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI, consistent with @fan2007's results, whereas the S\nobreakdash-SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
+The same analysis was performed using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987], obtaining similar results.
+The latter allows us to conclude that these results do not depend on the ENSO index used.
 
@@
 
-To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM
-and S\nobreakdash-SAM indices we regressed surface land air and sea temperature
-and precipitation onto each of the three SAM indices at 700 hPa. As shown in
-previous sections, the three indices are highly coherent in the troposphere, so
-we select this level to represent the tropospheric circulation for consistency
-with previous studies.
-
-Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at
-700 hPa with surface land air and sea temperature for each trimester. In Summer,
-positive values of the SAM index (Figure
-\@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are
-associated with negative temperature anomalies near Antarctica which are
-surrounded by a ring of positive anomalies. The ring is not zonally symmetric,
-as there are four distinctive local maximums around 30\degree W, 120\degree W,
-150\degree E and 90\degree E respectively. In the tropics, there are negative
-anomalies in the equatorial Pacific, consistent with the negative correlation
-between SAM and ENSO. Figure
-\@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))`
-show temperature anomalies associated with positive values of A\nobreakdash-SAM
-and S\nobreakdash-SAM, respectively. Both the local maximums in the ring and the
-anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM
-regression map, while temperature patterns linked to positive S\nobreakdash-SAM
-show a more zonally consistent ring and less relation to the tropics.
-Temperature anomalies in the Indian ocean, South Africa and Australia are
-strongly related to A\nobreakdash-SAM. This signal is not present in the
-regression pattern with the SAM. Spring (Figure \@ref(fig:regr-air-season), row
-`r get_row(season == "SON")`) features similar patterns but of smaller
-magnitude, with fewer regions where regressed anomalies have statistical
-significance.
-
-In Autumn and Winter (rows
-`r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))` in Figure
-\@ref(fig:regr-air-season)) the positive ring is only present through its local
-maximums in the regression with the SAM. There are also negative anomalies in
-Southern Australia, and positive anomalies over New Zealand and Southern South
-America. These patterns are not significant in the sense that there are no areas
-with p-values below 0.05 when controlling for FDR following @wilks2016. However,
-repeating this analysis with 2-meter air temperature from ERA5 resulted in
-similar patterns that were statistically significant (not shown). Moreover,
-similar features were observed in station measurements by @jones2019, although
-using data from 1957 to 2016.
-
-The pattern of negative anomalies in the pole surrounded by positive anomalies
-roughly seen in all seasons -- although with varying intensity and small-scale
-details -- translates to enhanced meridional temperature gradient maximised in
-the zero line, which is consistent with the intensification and poleward
-migration of the westerlies commonly linked to the SAM through thermal wind
-balance. It's then not surprising to see it more clearly in association with
-S\nobreakdash-SAM (at least in Summer and Spring).
-
-Figure \@ref(fig:regr-air-season) column b can be partially compared with Figure
-11 from @fogt2012. Although they used station data from 1958 to 2001, main
-features are reproduced here, such as the strong signal in New Zealand and
-Australia in Summer and Spring.
-
-Regression of the SAM indices with seasonal mean precipitation and 700 hPa
-geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and
-\@ref(fig:pp-regr-america) for New Zealand and neighbouring islands, and South
-America respectively. South Africa is not shown because no significant signal
-was detected there.
+To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices we regressed surface land air and sea temperature and precipitation onto each of the three SAM indices at 700 hPa.
+As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
+
+Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface land air and sea temperature for each trimester.
+In Summer, positive values of the SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
+The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E respectively.
+In the tropics, there are negative anomalies in the equatorial Pacific, consistent with the negative correlation between SAM and ENSO.
+Figure \@ref(fig:regr-air-season)`r combine_words(get_tag(season == "DJF" & term != "full", n = 2))` show temperature anomalies associated with positive values of A\nobreakdash-SAM and S\nobreakdash-SAM, respectively.
+Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics.
+Temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM.
+This signal is not present in the regression pattern with the SAM.
+Spring (Figure \@ref(fig:regr-air-season), row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with fewer regions where regressed anomalies have statistical significance.
+
+In Autumn and Winter (rows `r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the SAM.
+There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
+These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following @wilks2016.
+However, repeating this analysis with 2-meter air temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
+Moreover, similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
+
+The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons -- although with varying intensity and small-scale details -- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
+It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in Summer and Spring).
+
+Figure \@ref(fig:regr-air-season) column b can be partially compared with Figure 11 from @fogt2012.
+Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.
+
+Regression of the SAM indices with seasonal mean precipitation and 700 hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for New Zealand and neighbouring islands, and South America respectively.
+South Africa is not shown because no significant signal was detected there.
 
@@
 
-In Australia, the annual regression shows that the SAM index is associated with
-positive precipitation anomalies in the Southeastern region (Figure
-\@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006. The separation
-between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive
-anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure
-\@ref(fig:pp-regr-oceania)c.1). Geopotential height anomalies associated with
-this index (black contours) are indicative of easterly flow from the Tasman Sea,
-which could explain the positive anomalies in precipitation as found by
-@hendon2007. A\nobreakdash-SAM appears related to positive precipitation
-anomalies in the West coast of Southeastern Australia (Figure
-\@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the
-anomalous westerly circulation transporting moist air to the continent from the
-Indian Ocean.
-
-The seasonal-based regressions show statistically significant anomalies only in
-Spring, when positive SAM is associated with positive precipitation anomalies in
-Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5). In this trimester,
-S\nobreakdash-SAM seems to be associated with positive precipitation anomalies
-in a relatively reduced area of the East Coast (Figure
-\@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies
-related with positive A\nobreakdash-SAM affect all Eastern Australia (Figure
-\@ref(fig:pp-regr-oceania)b.5).
-
-In Summer, positive SAM index is associated with positive precipitation
-anomalies in Western and Eastern Australia, particularly in the North East
-(Figure \@ref(fig:pp-regr-oceania)a.2). The Eastern part being dominated by the
-relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM. In
-Autumn, the regression with SAM shows positive anomalies in the North, similar
-to Summer, and a broad area of positive anomalies in the North-East to
-South-West direction. This structure seems to be associated with the
-S\nobreakdash-SAM, while the Northern positive values are associated with the
-A\nobreakdash-SAM. In Winter we see the same NE to SW aligned anomaly (although
-with much reduced amplitude) that is also present only in relation with the
-S\nobreakdash-SAM. None of these regression coefficients are statistically
-significant at the 95% level. The Spring signal is broadly consistent with
-@hendon2007, but whereas they also detected a strong signal in Summer, Figure
-\@ref(fig:pp-regr-oceania)a.2 shows no statistically significant association
-(although the coefficients have the consistent sign).
-
-(ref:pp-rer-america-cap) Same as Figure \@ref(fig:pp-regr-oceania) but for South
-America.
+In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the Southeastern region (Figure \@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006.
+The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure \@ref(fig:pp-regr-oceania)c.1).
+Geopotential height anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by @hendon2007.
+A\nobreakdash-SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
+
+The seasonal-based regressions show statistically significant anomalies only in Spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
+In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related with positive A\nobreakdash-SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).
+
+In Summer, positive SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \@ref(fig:pp-regr-oceania)a.2).
+The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
+In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, and a broad area of positive anomalies in the North-East to South-West direction.
+This structure seems to be associated with the S\nobreakdash-SAM, while the Northern positive values are associated with the A\nobreakdash-SAM.
+In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the S\nobreakdash-SAM.
+None of these regression coefficients are statistically significant at the 95% level.
+The Spring signal is broadly consistent with @hendon2007, but whereas they also detected a strong signal in Summer, Figure \@ref(fig:pp-regr-oceania)a.2 shows no statistically significant association (although the coefficients have the consistent sign).
+
+(ref:pp-rer-america-cap) Same as Figure \@ref(fig:pp-regr-oceania) but for South America.
 
@@
 
-In South America (Figure \@ref(fig:pp-regr-america)), regression using all
-seasons shows that positive SAM is associated with statistically significant
-negative precipitation anomalies in Southeastern South America (SESA) and
-Southern Chile, and non-significant positive anomalies in South Brazil, near the
-South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
-Figure \@ref(fig:pp-regr-america)b.1 and c.1 show that while the signal over
-SESA and southern Brazil is associated with A\nobreakdash-SAM, that in southern
-Chile is related to S\nobreakdash-SAM.
-
-Except Winter, seasonal-based regressions mirror this same pattern. Even if not
-statistically significant, they all show negative values in SESA and southern
-Chile along with positive values in southern Brazil in relation with the SAM.
-The separation of these features between A\nobreakdash-SAM and S\nobreakdash-SAM
-regression maps is also rather consistent.
-
-The anomalous circulation at 700 hPa associated with S\nobreakdash-SAM (Figure
-\@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over Southern
-Chile. This leads to reduced influx of moist air from the Pacific Ocean which is
-the main source of precipitable water in that region [e.g. @garreaud2007]. On
-the other hand, the anomalous circulation associated with positive values of
-A\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is
-anticyclonic in the South and cyclonic in the North. This promotes anomalous
-South-Easterly flow over SESA, which inhibits the flow of the South America
-Low-Level Jet to the region [@silvestri2009; @zamboni2010]. This same pattern
-was found to be associated with increased precipitation in Southern Brazil
-during SACZ events [@rosso2018]. There is a small area of significant positive
-precipitation anomalies with the SAM near central Argentina which is also
-present in the station-based analysis by @gillett2006 that is explained by the
-A\nobreakdash-SAM.
+In South America (Figure \@ref(fig:pp-regr-america)), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile, and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
+Figure \@ref(fig:pp-regr-america)b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with A\nobreakdash-SAM, that in southern Chile is related to S\nobreakdash-SAM.
+
+Except Winter, seasonal-based regressions mirror this same pattern.
+Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the SAM.
+The separation of these features between A\nobreakdash-SAM and S\nobreakdash-SAM regression maps is also rather consistent.
+
+The anomalous circulation at 700 hPa associated with S\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over Southern Chile.
+This leads to reduced influx of moist air from the Pacific Ocean which is the main source of precipitable water in that region [e.g.
+@garreaud2007].
+On the other hand, the anomalous circulation associated with positive values of A\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
+This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region [@silvestri2009; @zamboni2010].
+This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events [@rosso2018].
+There is a small area of significant positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the A\nobreakdash-SAM.
 
@@
 
-In this study we characterise the temporal and spatial variability of the
-zonally symmetric and asymmetric structure of the SAM. By projecting monthly
-geopotential fields at each level with the corresponding asymmetric and
-symmetric pattern, we created two indices for representing the zonally
-asymmetric and symmetric contributions of the SAM respectively.
-
-The A\nobreakdash-SAM index correlates strongly with the S\nobreakdash-SAM
-index. In the troposphere, this correlation is maximum at zero lag, while in the
-stratosphere is maximised with the A\nobreakdash-SAM leading the
-S\nobreakdash-SAM by one month. Since most indices of the SAM are calculated
-using surface or near-surface conditions, this result would suggest that they
-might not be sensitive to the most dramatic changes in SAM variability.
-
-The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might
-point to a link between the SAM and the Quasi Biennial Oscillation. There is
-evidence of influence between the QBO and the Northern Annular Mode [e.g.
-@holton1980; @watson2014; @zhang2020], so it is not unlikely that the SAM would
-be similarly affected. However, establishing this link would require further
-research.
-
-We observe a positive trend of SAM in Summer and Autumn, as was documented by
-previous studies [e.g. @fogt2020 and references therein] for low levels. We show
-that these trends maximise at 100 hPa, and are explained by the zonally
-symmetric component. We also find a statistically significant positive trend in
-the Symmetric component of the SAM in the stratosphere that is not evident in
-the SAM index. We find some evidence of the SAM becoming more zonally
-asymmetric, as there is a slight positive trend in the variance explained by the
-as the A\nobreakdash-SAM. This is in contrast to @fogt2012 who made the
-evaluation for the period 1958 -- 2001. This discrepancy might be due either due
-to differences in methodology or analysed period.
-
-In the troposphere, the spatial patterns of geopotential associated with the
-S\nobreakdash-SAM index are much closer to being fully annular than the patterns
-associated with the SAM index. The A\nobreakdash-SAM index, on the other hand,
-describes a wave-3 pattern with maximum amplitude in the Pacific. This pattern
-extends vertically in the troposphere but its maximum is located at 250 hPa,
-which also could suggest that surface-based indices are not optimal for
-capturing this variability. This wave-3 pattern is similar to the Pacific-South
-American Pattern, linked to ENSO variability. We found that the significant
-correlation that exists between the SAM index and the Oceanic Niño Index over
-this period is captured entirely by the A\nobreakdash-SAM index. This suggests
-that ENSO is linked to SAM exclusively through the variability in the latter's
-asymmetric component and thus, the A\nobreakdash-SAM index could be a useful
-measure to further study that relationship.
-
-Temperature anomalies associated with the SAM broadly show a pattern of negative
-anomalies at polar latitudes surrounded by positive anomalies, but with many
-deviations from symmetry. The A\nobreakdash-SAM index explains a big portion of
-these deviations. In particular, the positive phase of A\nobreakdash-SAM is
-associated with colder temperatures over Southern Brazil, South Africa and
-Southern Australia, as well as the negative anomalies in the equatorial Pacific
-consistent with the ENSO\nobreakdash-SAM relationship. These negative anomalies
-are particularly clear in the DJF and SON trimesters, which include the months
-in which the ENSO teleconnection is more active [e.g. @cai2020a].
-
-In Australia the SAM is associated with positive precipitation anomalies in
-South East and this is explained by the S\nobreakdash-SAM. However, the
-A\nobreakdash-SAM is associated with a small area of positive precipitation
-anomalies in the Eastern Coast of West Australia, maybe due to advection of
-moist air from the Indian Ocean. In South America, precipitation anomalies
-associated with the SAM are negative both in Southern Chile and Southeastern
-South America, and positive in Southern Brazil. These features are cleanly
-separated between A\nobreakdash-SAM and S\nobreakdash-SAM. S\nobreakdash-SAM
-explains the negative anomalies in Southern Chile and A\nobreakdash-SAM, the
-negative-positive dipole between Southeastern South America and Southern Brazil.
+In this study we characterise the temporal and spatial variability of the zonally symmetric and asymmetric structure of the SAM.
+By projecting monthly geopotential fields at each level with the corresponding asymmetric and symmetric pattern, we created two indices for representing the zonally asymmetric and symmetric contributions of the SAM respectively.
+
+The A\nobreakdash-SAM index correlates strongly with the S\nobreakdash-SAM index.
+In the troposphere, this correlation is maximum at zero lag, while in the stratosphere is maximised with the A\nobreakdash-SAM leading the S\nobreakdash-SAM by one month.
+Since most indices of the SAM are calculated using surface or near-surface conditions, this result would suggest that they might not be sensitive to the most dramatic changes in SAM variability.
+
+The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
+There is evidence of influence between the QBO and the Northern Annular Mode [e.g. @holton1980; @watson2014; @zhang2020], so it is not unlikely that the SAM would be similarly affected.
+However, establishing this link would require further research.
+
+We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies [e.g. @fogt2020 and references therein] for low levels.
+We show that these trends maximise at 100 hPa, and are explained by the zonally symmetric component.
+We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the SAM index.
+We find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
+This is in contrast to @fogt2012 who made the evaluation for the period 1958 -- 2001.
+This discrepancy might be due either due to differences in methodology or analysed period.
+
+In the troposphere, the spatial patterns of geopotential associated with the S\nobreakdash-SAM index are much closer to being fully annular than the patterns associated with the SAM index.
+The A\nobreakdash-SAM index, on the other hand, describes a wave-3 pattern with maximum amplitude in the Pacific.
+This pattern extends vertically in the troposphere but its maximum is located at 250 hPa, which also could suggest that surface-based indices are not optimal for capturing this variability.
+This wave-3 pattern is similar to the Pacific-South American Pattern, linked to ENSO variability.
+We found that the significant correlation that exists between the SAM index and the Oceanic Niño Index over this period is captured entirely by the A\nobreakdash-SAM index.
+This suggests that ENSO is linked to SAM exclusively through the variability in the latter's asymmetric component and thus, the A\nobreakdash-SAM index could be a useful measure to further study that relationship.
+
+Temperature anomalies associated with the SAM broadly show a pattern of negative anomalies at polar latitudes surrounded by positive anomalies, but with many deviations from symmetry.
+The A\nobreakdash-SAM index explains a big portion of these deviations.
+In particular, the positive phase of A\nobreakdash-SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO\nobreakdash-SAM relationship.
+These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [e.g. @cai2020a].
+
+In Australia the SAM is associated with positive precipitation anomalies in South East and this is explained by the S\nobreakdash-SAM.
+However, the A\nobreakdash-SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
+In South America, precipitation anomalies associated with the SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
+These features are cleanly separated between A\nobreakdash-SAM and S\nobreakdash-SAM.
+S\nobreakdash-SAM explains the negative anomalies in Southern Chile and A\nobreakdash-SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
 Individual seasons mostly follow this pattern.
 
-@silvestri2009 suggests that precipitation impacts linked to the SAM changed
-strongly before and after 1980. In particular, the negative relationship with
-precipitation in South America was absent in some areas and switched sign in
-others in the earlier period. The correlation between ENSO and SAM is similarly
-non-stationary, also changing sign before the 1980s [@fogt2006; @clem2013].
-Seeing as both the ENSO\nobreakdash-SAM relationship and most of the
-precipitation impacts in South America are captured by the A\nobreakdash-SAM,
-the results presented here are most likely period-dependent. The decadal
-variations of the A\nobreakdash-SAM should be focus of future studies.
-
-By separating the zonally symmetric and zonally symmetric SAM signals, we show
-that the asymmetric component of the SAM has its unique variability, trends and
-impacts. This is particularly important in the context of a changing climate, as
-the impact on the SAM of ozone recovery is modelled as highly zonally symmetric,
-while the impact of increased concentration of greenhouse gases has also a
-zonally asymmetric component [@arblaster2006].
+@silvestri2009 suggest that precipitation impacts linked to the SAM changed strongly before and after 1980.
+In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in others in the earlier period.
+The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s [@fogt2006; @clem2013].
+Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the results presented here are most likely period-dependent.
+The decadal variations of the A\nobreakdash-SAM should be focus of future studies.
+
+By separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
+This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modelled as highly zonally symmetric, while the impact of increased concentration of greenhouse gases has also a zonally asymmetric component [@arblaster2006].
 
@@
 
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project
-funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a
-PhD grant from CONICET, Argentina.
+The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
+Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
 
@@
 All data used in this paper is freely available from their respective sources.
-ERA5 data can be obtained via the Copernicus Climate Data Store
-([<https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means>](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means){.uri}).
-CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado,
-USA, from their Web site at <https://psl.noaa.gov/data/gridded/data.cmap.html>.
-The Oceanic Niño Index is available via NOAA's Climate Prediction Center:
-<https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>
+ERA5 data can be obtained via the Copernicus Climate Data Store ([\<https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means\>](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means){.uri}).
+CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at <https://psl.noaa.gov/data/gridded/data.cmap.html>.
+The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>
 
@@
 
-A version-controlled repository of the code used to create this analysis,
-including the code used to download the data can be found at
-<https://github.com/eliocamp/asymsam>.
+A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at <https://github.com/eliocamp/asymsam>.
 

### 2021-04-27 analysis/paper/asymsam.Rmd
Expande un poco lo de QBO.
#17

@@
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
-Arbitrary units.", fig.width=6, fig.height=2.5, fig.env="figure*"}
+Arbitrary units; positive values in blue and negative values in red.", fig.width=6, fig.height=2.5, fig.env="figure*"}
 sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
@@
 
-The stratospheric S\nobreakdash-SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \@ref(fig:A2)).
-This might suggest a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b].
-A local peak at 2 years years is discernible in the periodogram of the tropospheric S\nobreakdash-SAM, although it's not statistically significant.
+```{r compute-spectrums}
+spec <- indices %>% 
+  .[lev %in% c(50, 700)] %>% 
+  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
+  .[, fftspectrum(estimate, c(3, 5), B = 5000, probs = c(0, 0.95)), by = .(term, lev)]
+```
+
+```{r spectrum-max}
+max_period <- spec[lev == 50 & term == "sym"] %>% 
+  .[which.max(spec - `95%`)] %>% 
+  .[, 1/freq]
+```
+
+The stratospheric S\nobreakdash-SAM varies strongly with a period between 15 and 30 months (the maximum spectrum is located at `r round(max_period)` months), which can be seen by spectral analysis (Figure \@ref(fig:A2)).
+A local peak at a similar frequency range is discernible in the periodogram of the tropospheric S\nobreakdash-SAM, although it's not statistically significant.
+This period band is around the range of periodicity of the Quasi-Biennial Oscillation [@baldwin2001b] and is consistent with [@vasconcellos2020], who found that the SAM and the QBO share significant common high power around the 2-year band.
+The fact that this periodicity is not evident on the A\nobreakdash-SAM index, is also consistent with their composites of geopotential height anomalies during easterly and westerly QBO, which show a rather symmetric monopole over Antarctica.
 In the troposphere the most significant peak of variability is found in A\nobreakdash-SAM at around `r 12*.3` months.
@@
 
-```{r compute-spectrums}
-indices %>% 
-  .[lev %in% c(50, 700)] %>% 
-  # .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>% 
-  .[, fftspectrum(estimate, c(3, 5), B = 5000, probs = c(0, 0.95)), by = .(term, lev)] -> spec
-```
-
 ```{r A2, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5.

### 2021-04-27 analysis/paper/asymsam.Rmd
Agrega comentarios sobre onda 3 en S-SAM
#18

@@
 The stratospheric S\nobreakdash-SAM seems to connect more strongly to the troposphere than the stratospheric A\nobreakdash-SAM.
-
-<!-- ```{r} -->
-
-<!-- indices_wide %>%  -->
-
-<!--   .[, FitLm(as.numeric(time), asym, sym, se = TRUE), by = .(lev)] %>%  -->
-
-<!--   rm_intercept() %>%  -->
-
-<!--   # .[, term := NULL] %>%  -->
-
-<!--   .[, estimate := estimate/(3600*24*365*10)] %>% -->
-
-<!--   .[, std.error := std.error/(3600*24*365*10)] %>% -->
-
-<!--   # setnames("type", "term") %>%  -->
-
-<!--   ggplot(aes(lev, estimate)) + -->
-
-<!--   geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), alpha = 0.3) + -->
-
-<!--   geom_line() + -->
-
-<!--   scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) + -->
-
-<!--   geom_hline(yintercept = 0) + -->
-
-<!--   scale_x_level() + -->
-
-<!--   scale_y_continuous("Standard deviations per decade") + -->
-
-<!--   coord_flip() + -->
-
-<!--   facet_grid(~ term,labeller = labeller(term = lab_sam)) + -->
-
-<!--   tag_facets()  -->
-
-<!-- ``` -->
 
@@
 The regression patterns associated with the A\nobreakdash-SAM and S\nobreakdash-SAM indices successfully disentangle both structures.
-While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \@ref(fig:2d-regr)b.2), the S\nobreakdash-SAM index is associated with an annular structure, almost devoid of zonal asymmetries (Figure \@ref(fig:2d-regr)c.2).
+While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \@ref(fig:2d-regr)b.2), the S\nobreakdash-SAM index is associated with an annular structure, with only vestigial zonal asymmetries (Figure \@ref(fig:2d-regr)c.2) in the shape of a wave-3 which is the inverse of the A\nobreakdash-SAM wave-3.
 The wave-3 pattern observed in Figure \@ref(fig:2d-regr)b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern described by @raphael2004, whose reference locations are marked with points in the figure.

### 2021-04-27 analysis/paper/asymsam.Rmd
Agrega tabla con correlaciones con ENSO por trimestres
#18

@@
 library(knitr)
+library(kableExtra)
 
@@
 
-report_p <- function(p.value) {
+report_p <- function(p.value, signs = TRUE) {
+  if (signs) {
   ifelse(p.value < 0.001, " < 0.001", paste0(" = ", signif(p.value, 2)))  
+  } else {
+  ifelse(p.value < 0.001, " < 0.001", paste0(signif(p.value, 2)))    
+  }
+
 }
@@
     by = .(lev)]
+
+# I need to change the sign
+SAM$eof[[1]]$left[, hgt := -hgt]
+SAM$eof[[1]]$right[, hgt := -hgt]
 
@@
   indices <- readRDS(indices_file)
-  indices[, `:=`(estimate = estimate, 
-                 estimate_norm = estimate_norm)]
+  indices[, `:=`(estimate = -estimate, 
+                 estimate_norm = -estimate_norm)]
 } else {
@@
   .[, env := metR::WaveEnvelope(estimate), by = .(lat, term, lev)] %>%
+  .[, term := factor(term, levels = c("full", "asym", "sym"), ordered = TRUE)] %>% 
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0), global.breaks = FALSE) +
+  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0),
+                    global.breaks = FALSE) +
   geom_contour_tanaka(aes(z = estimate),
@@
   .[indices, on = "time", allow.cartesian = TRUE] %>% 
+  .[lev == 700] %>% 
   .[, with(cor.test(estimate, value), 
            list(cor = estimate, 
-                p.value = p.value, 
-                low = conf.int[1], 
-                upp = conf.int[2])), by = .(lev, term, index)]
-
-# ggplot(enso_sam_cor, aes(lev, cor, group = interaction(term, index))) +
-#   geom_ribbon(aes(ymin = low, ymax = upp, fill = term), alpha  = .3) +
-#   geom_line(aes(color = term)) +
-#   scale_x_level() +
-#   coord_flip()
-
-cor_texts <- enso_sam_cor[lev == 700 & index == "oni"] %>% 
-  .[, `:=`(cor = signif(cor, 2), 
-           p.value = report_p(p.value))]
+                p.value = p.value)), by = .(lev, term, index)] %>% 
+  .[, type := "Correlation"] %>% 
+  .[, season := "Year"] %>% 
+  .[, lev := NULL]
+
+enso_sam_cor_season <- enso %>% 
+  .[indices, on = "time", allow.cartesian = TRUE] %>% 
+  .[lev == 700] %>% 
+  .[, with(cor.test(estimate, value), 
+           list(cor = estimate, 
+                p.value = p.value)), by = .(lev, term, season(time), index)] %>% 
+  .[, type := "Correlation"] %>% 
+  .[, lev := NULL]
 
@@
   .[indices_wide[lev == 700], on = "time"] %>% 
+  .[, partial_cor(value, asym, sym), by = .(index)] %>% 
+  setnames("partial_correlation", "cor") %>% 
+  .[, type := "Partial Correlation"] %>% 
+  .[, season := "Year-round"]
+
+enso_sam_partial_season <- enso %>% 
+  .[indices_wide[lev == 700], on = "time"] %>% 
   .[, partial_cor(value, asym, sym), by = .(season(time), index)] %>% 
-  .[, partial_correlation := signif(partial_correlation, 2)] %>% 
+  setnames("partial_correlation", "cor") %>% 
+  .[, type := "Partial Correlation"]
+
+enso_cors <- rbind(enso_sam_cor, 
+      enso_sam_cor_season, 
+      enso_sam_partial,
+      enso_sam_partial_season, use.names = TRUE) %>% 
+  .[type == "Partial Correlation" | term == "full"]
+
+cor_texts <- enso_cors[index == "oni"] %>% 
+  .[, p.value := p.adjust(p.value, "fdr")] %>% 
+  .[, cor := round(cor, 2)] %>% 
   .[, p.value := report_p(p.value)] 
-
-enso_sam_partial <- enso %>% 
-  .[indices_wide[lev == 700], on = "time"] %>% 
-  .[, partial_cor(value, asym, sym), by = index] %>% 
-  .[, partial_correlation := signif(partial_correlation, 2)] %>% 
-  .[, p.value := report_p(p.value)]
+```
+
+```{r enso-cor-table}
+table <- enso_cors %>% 
+  .[index == "oni"] %>% 
+  .[, p.value := p.adjust(p.value, "fdr")] %>% 
+  .[, p.value := scales::pvalue(p.value)] %>% 
+  .[, cor_text := paste0(round(cor, 2), "\n(", p.value, ")")] %>% 
+  .[, term := lab_sam[term]] %>% 
+  dcast(season ~ term, value.var = "cor_text") %>% 
+  setcolorder(c("season",lab_sam))
+
+# table2 <- enso_cors %>% 
+#   .[index == "oni"] %>% 
+#   .[, p.value := p.adjust(p.value, "fdr")] %>% 
+#   # .[, p.value := scales::pvalue(p.value)] %>% 
+#   # .[, cor_text := paste0(round(cor, 2), "\n(", p.value, ")")] %>% 
+#   dcast(type + term ~ season, value.var = "p.value") %>% 
+#   .[, term := lab_sam[term]] 
+table2 <- enso_cors %>% 
+  .[index == "oni"] %>% 
+  .[, p.value := p.adjust(p.value, "fdr")] %>% 
+  # .[, p.value := scales::pvalue(p.value)] %>% 
+  # .[, cor_text := paste0(round(cor, 2), "\n(", p.value, ")")] %>% 
+  .[, term := lab_sam[term]] %>% 
+  dcast(season ~ term, value.var = "p.value") %>% 
+  setcolorder(c("season",lab_sam))
+
+table %>% 
+  kbl(escape = FALSE,
+      caption = "Correlation between SAM indices and the Oceanic Niño Index. p-values corrected for False Detection Rate in parenthesis. In bold, correlations with p-value smaller than 0.05.",
+      col.names = c("", lab_sam)) %>% 
+  column_spec(2, bold = table2[, 2] < 0.05) %>%
+  column_spec(3, bold = table2[, 3] < 0.05) %>%
+  column_spec(4, bold = table2[, 4] < 0.05) %>%
+  kable_paper(full_width = F) %>% 
+  add_header_above(c(" " = 1, "Correlation" = 1 , "Partial correlation" = 2)) 
 ```
@@
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO. 
-The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is `r cor_texts[term == "full"]$cor` (p-value `r cor_texts[term == "full"]$p.value`).
-This relationship is captured mainly by the A\nobreakdash-SAM, as this index has a partial correlation of `r enso_sam_partial[index == "oni" & term == "asym"]$partial_correlation` (p-value `r enso_sam_partial[index == "oni" & term == "asym"]$p.value`) with the ONI, consistent with @fan2007's results, whereas the S\nobreakdash-SAM's partial correlation with the ONI is essentially null (`r enso_sam_partial[index == "oni" & term == "sym"]$partial_correlation`; p-value `r enso_sam_partial[index == "oni" & term == "sym"]$p.value`).
+The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is shown in Table\ \@ref{tab:enso-cor-table} for each SAM index and for each trimester. 
+There is a significant correlation between SAM and ENSO. 
+When divided in trimesters, this correlation is important only in DJF and SON. 
+This relationship is captured mainly by the A\nobreakdash-SAM, as this index has signifcant partial correlations with ENSO while correlations with S\nobreakdash-SAM are all smaller and non-significant.
+In MAM, ENSO is not significantly correlated with SAM, but it is significantly correlated with A\nobreakdash-SAM in a level comparable with the ENSO-SAM correlation in SON. 
 The same analysis was performed using the Multivariate ENSO Index [@wolter2011] and the Southern Oscillation Index [@ropelewski1987], obtaining similar results.

### 2021-04-28 analysis/paper/asymsam.Rmd
Agrega más comentarios de QBO en la conclusión
#18

@@
   .[, type := "Partial Correlation"] %>% 
-  .[, season := "Year-round"]
+  .[, season := "Year"]
 
@@
 
-The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the PSA Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989]. 
+The wave-3 pattern from Figure\ \@ref(fig:2d-regr)b.2 is very similar to the PSA Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO. 
@@
 
-The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
-There is evidence of influence between the QBO and the Northern Annular Mode [e.g. @holton1980; @watson2014; @zhang2020], so it is not unlikely that the SAM would be similarly affected.
-However, establishing this link would require further research.
+The periodicity of about two years we found in the stratospheric S\nobreakdash-SAM is consistent with the possible link between the SAM and the Quasi Biennial Oscillation as proposed by [@vasconcellos2020].
+There is evidence of influence between the QBO and the Northern Annular Mode [e.g.
+@holton1980; @watson2014; @zhang2020], so it is not unlikely that the SAM would be similarly affected.
+Establishing this link would require further research.
 

### 2021-04-28 analysis/paper/asymsam.Rmd
Cambia signo en Figura A6 y agrega que hay cambios.
#23

@@
 The only outlier being DJF, which shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref{fig:A3}), which is consistent with @fogt2020.
-With this in mind, a
 
@@
 Zonal asymmetric patterns of SAM were computed for the two halves of the period (1979 to 1998 and 1999 to 2018) respectively.
-They show no systematic change neither in the stratosphere nor in the troposphere (Figure \@ref(fig:A6)).
+The differences between the two periods appear to be relatively small in both the troposphere and stratosphere (Figure \@ref(fig:A6)).
 
@@
   .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>%
+
   .[, year := year(time)] %>% 
@@
 patterns %>% 
+  .[lev == 50 & period == "1999-2018", estimate_a := -estimate_a] %>% 
   periodic(lon = c(0, 360)) %>%

### 2021-04-28 analysis/paper/asymsam.Rmd
Hace los tanakas más finitos
#23

@@
               sym  = "S-SAM")
+
+geom_contour_tanaka2 <- purrr::partial(geom_contour_tanaka, range = c(0.01, 0.3))
 ```
@@
   geom_contour_fill(aes(z = value), global.breaks = FALSE, breaks = ZeroBreaks) +
-  geom_contour_tanaka(aes(z = value), global.breaks = FALSE, breaks = ZeroBreaks) +
+  geom_contour_tanaka2(aes(z = value), global.breaks = FALSE, breaks = ZeroBreaks) +
   geom_qmap() +
@@
   scale_fill_divergent(guide = "none", high = "#731C1F", low = "#323071") +
-  scale_fill_divergent(guide = "none") +
   coord_polar() +
@@
   geom_contour_fill(aes(z = correlation, fill = ..level..), breaks = MakeBreaks(binwidth = 0.1)) +
-  geom_contour_tanaka(aes(z = correlation), size = 0.1, na.fill = .99, 
+  geom_contour_tanaka2(aes(z = correlation), size = 0.1, na.fill = .99, 
                       breaks = MakeBreaks(binwidth = 0.1)) +
@@
   .[lev %in% c(50, 700)] %>%
-  .[, env := metR::WaveEnvelope(estimate), by = .(lat, term, lev)] %>%
   .[, term := factor(term, levels = c("full", "asym", "sym"), ordered = TRUE)] %>% 
@@
                     global.breaks = FALSE) +
-  geom_contour_tanaka(aes(z = estimate),
+  geom_contour_tanaka2(aes(z = estimate),
                       breaks = AnchorBreaks(0, exclude =  0),
@@
   geom_contour_fill(aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(0, 10, 0)) +
-  geom_contour_tanaka(aes(z = estimate),
+  geom_contour_tanaka2(aes(z = estimate),
                       breaks = AnchorBreaks(0, 10, 0),
@@
                     breaks = AnchorBreaks(0, 0.1, 0)) +
-  # geom_contour_tanaka(aes(z = estimate),
+  # geom_contour_tanaka2(aes(z = estimate),
   #                     breaks = AnchorBreaks(0, 0.1, 0)) +
@@
     geom_contour_fill(aes(z = z), breaks = AnchorBreaks(0,exclude = 0), global.breaks = FALSE) +
-    geom_contour_tanaka(aes(z = z), breaks = AnchorBreaks(0,exclude = 0), global.breaks = FALSE) +
+    geom_contour_tanaka2(aes(z = z), breaks = AnchorBreaks(0,exclude = 0), global.breaks = FALSE) +
 
@@
     geom_contour_fill(aes(z = estimate_a), breaks = breaks) +
-    geom_contour_tanaka(aes(z = estimate_a), breaks = breaks) +
+    geom_contour_tanaka2(aes(z = estimate_a), breaks = breaks) +
     geom_qmap(~.x[lat <= 0]) +
@@
   geom_contour_fill(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
-  geom_contour_tanaka(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
+  geom_contour_tanaka2(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
   geom_qmap() +
@@
   geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
-  geom_contour_tanaka(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
+  geom_contour_tanaka2(aes(z = acf), breaks = AnchorBreaks(0, 0.05),
+                      range = c(0.01, 0.3)) +
   geom_vline(xintercept = 0) +

### 2021-04-28 analysis/paper/asymsam.Rmd
Agrega ejes de latitud y longitud
#23

@@
 
+geom_coords <- function() {
+  list(
+    geom_segment(data = data.frame(xstart =  seq(0, 360-30, by = 30), 
+                                   xend =  seq(0, 360-30, by = 30),
+                                   ystart = -90+15,
+                                   yend = Inf),
+                 aes(x = xstart, xend = xend, y = ystart, yend = yend),
+                 size = 0.1, alpha = 0.5),
+
+    # geom_vline(xintercept = seq(0, 360-30, by = 30), size = 0.1, alpha = 0.5),
+
+      geom_hline(yintercept = seq(-90, 0, by = 15), size = 0.1, alpha = 0.5),
+      shadowtext::geom_shadowtext(data = data.frame(x = 0, y = seq(-90 + 15, 0, by = 15)),
+                                  aes(x, y, label = LatLabel(y)), size = 2.5, alpha = 0.5,
+                                  colour = "black",
+                                  bg.colour = "white" ) 
+  )
+}
+
 knitr::opts_hooks$set(invisible = function(options) {
@@
   geom_qmap() +
-  scale_x_longitude() +
+  geom_coords() +
+  scale_x_longitude(labels = NULL) +
   scale_y_latitude(limits = c(NA, -20), labels = NULL) +
@@
   geom_qmap(~.x[lat <= -20]) +
+  geom_coords() +
   geom_point(data = data.frame(lon = ConvertLongitude(c(50, 166, -76)),
@@
              shape = 21, size = 2.5, color = "black", fill = "#de3e80") +
-  scale_x_longitude() +
-  scale_y_latitude(limits = c(NA, -20), labels = NULL,
-                   minor_breaks = NULL) +
+  scale_x_longitude(labels = NULL) +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
   scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
@@
   tag_facets("cr") +
-  axis_labs_smol  +
-  no_grid +
-  theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
+  no_grid 
 ```
@@
   geom_qmap(~.x[lat <= 0]) +
-  scale_x_longitude() +
-  scale_y_latitude(labels = NULL, limits = c(NA, 0), 
-                   minor_breaks = NULL) +
+  geom_coords() +
+  scale_x_longitude(labels = NULL) +
+  scale_y_latitude(labels = NULL, limits = c(NA, 0)) +
   scale_fill_divergent("2-meter air temperature",  
@@
   tag_facets("cr", tag_suffix = "") +
-  no_grid +
-  axis_labs_smol +
-  theme(panel.grid.major.y = element_line(color = "gray60", size = 0.1))
+  no_grid 
 ```
@@
   geom_contour_fill(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
-  geom_contour_tanaka2(aes(z = acf), breaks = AnchorBreaks(0, 0.05),
-                      range = c(0.01, 0.3)) +
+  geom_contour_tanaka2(aes(z = acf), breaks = AnchorBreaks(0, 0.05)) +
   geom_vline(xintercept = 0) +

### 2021-04-29 analysis/paper/asymsam.Rmd
Agrega grilla en una figura
#23

@@
 theme_set(theme_asymsam(base_size = 9) + theme(tagger.panel.tag.text = element_text(size = 8)))
+
+panel_background <- theme(panel.background = element_rect(fill = "#fbfbfb", color = NA), 
+                          panel.ontop = FALSE,
+                          tagger.panel.tag.background = ggplot2::element_rect(color = NA, 
+                                                                              fill = "#fbfbfb"))
+
 ZeroBreaks <- AnchorBreaks(0, NULL, 0)
@@
   facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets("cr", position = "tr")  +
-  no_grid 
+  tag_facets("cr", position = "tr") 
 ```

### 2021-04-29 R/theme.R

@@ -22,6 +22,6 @@ theme_asymsam <- function(base_size = 16) {
       # # legend.title = element_blank(),
       # legend.box.spacing = grid::unit(3, "mm"),
       # legend.margin = ggplot2::margin(t = -5),
-      panel.grid = ggplot2::element_line(color = "gray10", size = 0.4, linetype = 3),
+      panel.grid = ggplot2::element_line(color = scales::alpha("black", 0.5), size = 0.2),
       panel.ontop = TRUE)
 }

### 2021-04-29 analysis/paper/asymsam.Rmd
Agrega aclaración sobre período y effectos
#23

@@
 The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s [@fogt2006; @clem2013].
-Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the results presented here are most likely period-dependent.
+Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the magnitude and sign these impacts are most likely period-dependent and represent the average signal between 1979 and 2018.
 The decadal variations of the A\nobreakdash-SAM should be focus of future studies.
@@
     geom_qmap(~.x[lat <= 0]) +
-    scale_x_longitude() +
+    geom_coords() +
+    scale_x_longitude(labels = NULL) +
     scale_y_latitude(limits = c(NA, -20), labels = NULL) +
@@
     geom_qmap(~.x[lat <= 0]) +
-    scale_x_longitude() +
+    scale_x_longitude(labels = NULL) +
     scale_y_latitude(limits = c(NA, -20), labels = NULL) +
@@
     coord_polar() +
+    geom_coords() +
     axis_labs_smol +
@@
   geom_qmap() +
-  scale_x_longitude() +
+  scale_x_longitude(labels = NULL) +
   scale_y_latitude(limits = c(NA, -20), labels = NULL) +
@@
   coord_polar() +
+  geom_coords() +
   facet_grid(lev~period, labeller = labeller(lev = lev.lab)) +

### 2021-04-30 R/theme.R
Cambia el theme

@@ -22,6 +22,6 @@ theme_asymsam <- function(base_size = 16) {
       # # legend.title = element_blank(),
       # legend.box.spacing = grid::unit(3, "mm"),
       # legend.margin = ggplot2::margin(t = -5),
-      panel.grid = ggplot2::element_line(color = scales::alpha("black", 0.5), size = 0.2),
+      panel.grid = ggplot2::element_line(color = scales::alpha("gray60", 0.5), size = 0.1),
       panel.ontop = TRUE)
 }

### 2021-04-30 analysis/scripts/paper_figs.Rmd
Agrega datos derivados
Closes #16

@@
 ```{r compute-asymsam}
-indexes_file <- data_path("derived", "indexes.Rds")
+indexes_file <- data_path("derived", "indices.Rds")
 if (file.exists(indexes_file)) {

### 2021-04-30 R/data_locations.R

@@ -14,7 +14,7 @@ data_path <- function(type = c("raw", "derived"), ...) {
 #' @export
 #' @rdname data_locations
 AAO <- function() {
-   file <- data_path("derived", "aao.Rds")
+   file <- data_path("raw", "aao.Rds")
    checkmate::assert_access(file, access = "r")
    return(file)
 }

### 2021-04-30 analysis/scripts/01-get-data.R

@@ -27,7 +27,7 @@ aao <- rsoi::download_aao(use_cache = !force, file = asymsam::data_path("raw", f
   data.table::as.data.table() %>%
   .[, .(time = lubridate::as_datetime(Date), aao = AAO)]
 
-saveRDS(aao, asymsam::data_path("derived", "aao.Rds"))
+saveRDS(aao, asymsam::data_path("raw", "aao.Rds"))
 
 # Download QBO
 qbourl <- "https://psl.noaa.gov/data/correlation/qbo.data"

### 2021-04-30 analysis/scripts/02-compute-eofs.R

@@ -5,7 +5,7 @@ library(data.table)
 
 indices_file <- data_path("derived", "indices.Rds")
 
-message("Hellow, you're running the script to compute symmetric and asymmetric SAM.\nI'll be your messenger for the day.")
+message("Hello, you're running the script to compute symmetric and asymmetric SAM.\nI'll be your messenger for the day.")
 
 if (!file.exists(indices_file)) {
   message("Reading data")
@@ -21,6 +21,7 @@ if (!file.exists(indices_file)) {
   indices <- hgt[lat <= -20, eof_asym(hgt_a, lon, lat, time, n = 1), by = lev]
   message("Done! Saving file.")
   saveRDS(indices, file = data_path("derived", "indices.Rds"))
+  fwrite(indices, file = data_path("derived", "indices.csv"), yaml = TRUE)
   message("Everything's ready. Have a nice day!")
 } else {
   message("The file ", indices_file, "already exists, so you don't need to do anything. Horray!")

### 2021-04-30 analysis/paper/asymsam.Rmd
Agrega aclaración de los datos al paper
#16

@@
 The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>
+The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at the public repository associated with this paper, at <https://github.com/eliocamp/asymsam/tree/main/analysis/data/derived_data>.
 

### 2021-04-30 analysis/paper/asymsam.Rmd
Achica grosor de amplitud de ondas
#17

@@
   ggplot(aes(lat, amplitude)) +
-  geom_line(aes(color = factor(k)), size = 1.2) +
+  geom_line(aes(color = factor(k)), size = 1) +
   scale_color_brewer("Wavenumber", palette = "Set1") +

### 2021-04-30 analysis/paper/asymsam.Rmd
Corrige opciones de datos

@@
 ```{r enso-cor}
-oni <- rsoi::download_oni(TRUE, data_path("derived", "oni.csv")) %>% 
+oni <- rsoi::download_oni(TRUE, data_path("raw", "oni.csv")) %>% 
   as.data.table() %>% 
@@
 
-mei <- rsoi::download_mei(TRUE, data_path("derived", "mei.csv")) %>% 
+mei <- rsoi::download_mei(TRUE, data_path("raw", "mei.csv")) %>% 
   as.data.table() %>% 
@@
 
-soi <- rsoi::download_soi(TRUE, data_path("derived", "soi.csv")) %>% 
+soi <- rsoi::download_soi(TRUE, data_path("raw", "soi.csv")) %>% 
   as.data.table() %>% 

### 2021-04-30 analysis/paper/asymsam.Rmd
Termina la tabla de cor ENSO-SAM

@@
   .[, p.value := p.adjust(p.value, "fdr")] %>% 
-  .[, p.value := scales::pvalue(p.value)] %>% 
-  .[, cor_text := paste0(round(cor, 2), "\n(", p.value, ")")] %>% 
+  .[, p.value := paste0("(", scales::pvalue(p.value), ")")] %>% 
+  .[, cor_text := scales::number(cor, 0.01)] %>% 
   .[, term := lab_sam[term]] %>% 
-  dcast(season ~ term, value.var = "cor_text") %>% 
-  setcolorder(c("season",lab_sam))
-
-# table2 <- enso_cors %>% 
-#   .[index == "oni"] %>% 
-#   .[, p.value := p.adjust(p.value, "fdr")] %>% 
-#   # .[, p.value := scales::pvalue(p.value)] %>% 
-#   # .[, cor_text := paste0(round(cor, 2), "\n(", p.value, ")")] %>% 
-#   dcast(type + term ~ season, value.var = "p.value") %>% 
-#   .[, term := lab_sam[term]] 
+  .[] %>% 
+  melt(id.vars = c("term", "index", "season" ), measure.vars = c("cor_text", "p.value")) %>% 
+  dcast(season + variable ~ term, value.var = "value") %>% 
+  .[, variable := NULL] %>% 
+  setcolorder(c("season", lab_sam))
+
 table2 <- enso_cors %>% 
@@
   .[, p.value := p.adjust(p.value, "fdr")] %>%
-  # .[, p.value := scales::pvalue(p.value)] %>% 
-  # .[, cor_text := paste0(round(cor, 2), "\n(", p.value, ")")] %>% 
+
   .[, term := lab_sam[term]] %>% 
@@
 table %>% 
-  kbl(escape = FALSE,
+  kbl(align = "c", 
       caption = "Correlation between SAM indices and the Oceanic Niño Index. p-values corrected for False Detection Rate in parenthesis. In bold, correlations with p-value smaller than 0.05.", 
-      col.names = c("", lab_sam)) %>% 
-  column_spec(2, bold = table2[, 2] < 0.05) %>%
-  column_spec(3, bold = table2[, 3] < 0.05) %>%
-  column_spec(4, bold = table2[, 4] < 0.05) %>%
-  kable_paper(full_width = F) %>% 
-  add_header_above(c(" " = 1, "Correlation" = 1 , "Partial correlation" = 2)) 
+      col.names = c("", lab_sam), booktabs = TRUE) %>%
+  kable_classic(full_width = F) %>%
+  column_spec(2, bold = rep(table2[, 2] < 0.05, each = 2)) %>%
+  column_spec(3, bold = rep(table2[, 3] < 0.05, each = 2)) %>%
+  column_spec(4, bold = rep(table2[, 4] < 0.05, each = 2)) %>%
+  collapse_rows(columns = 1:2, valign = "top") %>% 
+  add_header_above(c(" " = 1, "Correlation" = 1, 
+                     "Partial correlation" = 2)) 
 ```

### 2021-04-30 analysis/paper/asymsam.Rmd
Aclara mejor diferencias entre DJF y otros.
#17

@@
 Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year.
-The only outlier being DJF, which shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref{fig:A3}), which is consistent with @fogt2020.
+The outlier being DJF, which shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref{fig:A3}), which is consistent with @fogt2020.
+Therefore, even though we do our analysis for all months, we would expect for it to represent the other trimesters better.
 
@@
   .[season == "DJF", estimate := -estimate] %>% 
-  .[, estimate_a := Anomaly(estimate), by = .(lat, lev, season)]
+  .[, estimate_a := Anomaly(estimate), by = .(lat, lev, season)] %>% 
+  .[season == "DJF" & lev == 700, estimate_a := -estimate_a]
 

### 2021-04-30 analysis/paper/asymsam.Rmd
Separa trends de R2 en trimestres
#17

@@
 
-Figure \@ref(fig:r-squared-trend) shows trends for the explained variance of each index.
-There is no evidence of a significant trend in the stratosphere.
-In the troposphere, there is a positive trend for A\nobreakdash-SAM and no significant trend for S\nobreakdash-SAM.
-This suggest that the SAM has become more asymmetric in the period from 1979 to 2018.
-However, the change is slight, around 1% increased explained variance per decade.
-
-```{r r-squared-trend, fig.height=3, fig.width=3, fig.cap = "Linear trends (in percent per decade) of the variance explained by A\\nobreakdash-SAM and S\\nobreakdash-SAM at each level for the period 1979 -- 2018.
+A positive trend in the S\nobreakdash-SAM index and no trend in the A\nobreakdash-SAM index might at first suggest a trend towards a more symmetric SAM. 
+However, a very negative S\nobreakdash-SAM trending towards a less negative S\nobreakdash-SAM would be translated into a positive S\nobreakdash-SAM trend but a more asymmetric SAM.
+
+```{r r-squared-trend, fig.height=4, fig.width=4, fig.cap = "Linear trends (in percent per decade) of the variance explained by A\\nobreakdash-SAM and S\\nobreakdash-SAM at each level and for each trimester for the period 1979 -- 2018.
 Shading indicates the 95\\% confidence interval."}
@@
   .[term != "full"] %>% 
-  .[, FitLm(r.squared, time, se = TRUE), by = .(lev, type = term)] %>% 
+  .[, FitLm(r.squared, time, se = TRUE), by = .(lev, type = term, season(time))] %>% 
   rm_intercept() %>%
@@
   scale_fill_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill"), labels = lab_sam) +
-  coord_flip() 
-```
+  coord_flip() +
+  facet_wrap(~season)
+```
+To study the question of whether the SAM is becoming more or less asymmetric, we show trends for the  explained variance of each index for each trimester in Figure \@ref(fig:r-squared-trend). 
+In the troposphere the only significant trend is in DJF, in which the A\nobreakdash-SAM has a positive trend of about 2% per decade, suggesting, that the DJF SAM has become more asymmetric in the period from 1979 to 2018,
+However due to the atypical nature of the asymmetric component of the SAM during DJF (Section\ \ref(definition-of-indices)) this should be taken only as preliminary evidence. 
+The other significant trends is in the stratosphere during SON, where there is a positive trend in the variance explained by the S\nobreakdash-SAM of about 4% per decade. 
+This change could be the result of the forcing from ozone depletion.
 

### 2021-04-30 analysis/paper/asymsam.Rmd
Elimina párrafo de amplitud modulada.
#17
Agrega separación de trimestres R2 en conclusión
#17

@@
 
-Looking at Figure \@ref(fig:2d-regr)b.2, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
-
 ```{r compute-vertical-regression}
@@
 We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the SAM index.
-We find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
+We find some evidence of the SAM becoming more zonally asymmetric during DJF, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
 This is in contrast to @fogt2012 who made the evaluation for the period 1958 -- 2001.

### 2021-04-30 analysis/paper/asymsam.Rmd
Agrega comentario sobre ENSO en australia
#17

@@
 However, the A\nobreakdash-SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
+
 In South America, precipitation anomalies associated with the SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
@@
 Individual seasons mostly follow this pattern.
+
+Of note, even though the A\nobreakdash-SAM is significantly correlated with ENSO in all trimesters except JJA, precipitation patterns associated with the A\nobreakdash-SAM  are similar to precipitation patterns associated with ENSO only in SON in Australia, but in all trimesters in South America.
+This might indicate that the A\nobreakdash-SAM is closely related to the pathway between ENSO and precipitation patterns in South America but not so much with the pathway between ENSO and precipitation in Australia.
 

### 2021-04-30 analysis/paper/asymsam.Rmd
Agrega comentario de detrend
#17
Cambio menor y render.

@@
 As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
+Regressions were done without detrending either the variables nor the indices, but computing the regressions with detrended values doesn't change the results (not shown).
 
@@
 The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>
-The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at the public repository associated with this paper, at <https://github.com/eliocamp/asymsam/tree/main/analysis/data/derived_data>.
+The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at the public repository associated with this paper, at <https://github.com/eliocamp/asymsam/tree/master/analysis/data/derived_data>.
 

### 2021-05-04 analysis/scripts/01-get-data.R
Agrega temperatura de ERA5

@@ -16,7 +16,8 @@ files <- list(aao = "aao.csv",
               hadcrut = "HadCRUT.4.6.0.0.median.nc",
               gpcc = "precip.mon.total.v2018.nc",
               gpcp = "gpcp.precip.mon.mean.nc",
-              era5 =  "era5.z.mon.mean.nc")
+              era5 =  "era5.z.mon.mean.nc",
+              era52mt = "era5.2mt.mon.mean.nc")
 
 dir.create(asymsam::data_path("raw"), recursive = TRUE, showWarnings = FALSE)
 dir.create(asymsam::data_path("derived"), recursive = TRUE, showWarnings = FALSE)
@@ -144,85 +145,43 @@ if (!force && file.exists(asymsam::data_path("raw", file))) {
                               path = asymsam::data_path())
 }
 
-#
-# for (y in seq(1979, 2018)) {
-#   url <- paste0("ftp://ftp.cdc.noaa.gov/Datasets/ncep.reanalysis.dailyavgs/pressure/hgt.", y, ".nc")
-#   dir.create(asymsam::data_path("raw", "ncep"), )
-#   file <- asymsam::data_path("raw", "ncep", paste0("hgt.daily.", y, ".nc"))
-#   if (!file.exists(file)) {
-#     download.file(url, destfile = file)
-#   }
-# }
-#
-# ncepfiles <- list.files(asymsam::data_path("raw", "ncep"), full.names = TRUE)
-#
-# asymsam::nc_concatenate(ncepfiles, asymsam::data_path("raw", "ncep.daily.nc"), overwrite = TRUE)
 
-# # Antarctic sea ice
-# # from https://nsidc.org/data/G02135
-# months <- 1:12
-# years <- 1979:2019
-# dates <- data.table::CJ(years, months)
-# # There are no data from 3 December 1987 to 13 January 1988 due to satellite problems. :sob:
-# dates <- dates[!(paste0(years, months) %in% c(198712, 19881))]
-#
-# make_url <- function(year, month) {
-#   month_zero <- formatC(month, width = 2, flag = "0")
-#   month_Abb <- month.abb[month]
-#
-#   file_name <- paste0("S_", paste0(year, month_zero), "_concentration_v3.0.tif")
-#   url <- paste0("ftp://sidads.colorado.edu/DATASETS/NOAA/G02135/south/monthly/geotiff/",
-#                 month_zero, "_", month_Abb, "/", file_name)
-#
[…]

### 2021-05-04 analysis/paper/asymsam.Rmd
Regresiones con ERA5 t2m
Closes #26 Closes #20

@@
 
+guide_colorsteps_bottom <- function(width = 25, height = 0.5, ...) {
+  guide_colorsteps(title.position = "top", title.hjust = 0.5,
+                   barheight = height,
+                   barwidth = width, ...)
+}
+
 geom_coords <- function() {
@@
       shadowtext::geom_shadowtext(data = data.frame(x = 0, y = seq(-90 + 15, 0, by = 15)),
-                                  aes(x, y, label = LatLabel(y)), size = 2.5, alpha = 0.5,
+                                  aes(x, y, label = LatLabel(y)), size = 2, alpha = 0.5,
                                   colour = "black",
@@
 
-We used monthly geopotential height at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018.
+We used monthly geopotential height at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels as well as 2 metre temperature from ERA5 [@hersbach2020] for the period 1979 to 2018.
 We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations.
 
-Temperature data used are from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid 5\degree longitude by 5\degree latitude.
 For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [\@xie1997], with a 2.5\\degree\\ resolution in latitude and longitude.
@@
 time_subset <- list(time = range(indices$time))
-air <- NOAATemp() %>% 
-  ReadNetCDF(vars = "air", subset = c(time_subset, list(lat = c(-90, 10)))) %>% 
-  normalise_coords()  %>%
-  # Interpolate at lon = 0. Otherwise, can't close contours. 
-  qwrap(lon = c(0, 360) ~ c(-2.5, 360)) %>% 
-  .[ , Interpolate(air ~ lon + lat, x.out = c(0, unique(lon)), y.out = unique(lat)), by = time] %>% 
-  .[lon != -2.5]
-
-# air <- ReadNetCDF("~/DATOS/reanalysis/ERA5/mon/era5sl.mon.mean.nc", 
-#            vars = c(air = "t2m"), subset = c(time_subset, list(lat = c(-90, 10)))) %>% 
-#   normalise_coords()
+
+air <- ERA52mt() %>% 
+  ReadNetCDF(vars = c("air" = "t2m"), 
+             subset = c(time_subset, list(latitude = c(-90, 10)))) %>% 
+  normalise_coords()
 
@@
   rm_singleton() %>% 
+  .[year(time) %between% c(1981, 2018)] %>% 
   .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
-  .[, nas := mean(is.na(air)), by = .(lon, lat, season(time))] %>%
-  .[nas >= 0.15, air := NA] %>%
   .[, rbind(as.data.table(FitLm(air, asym, sym, se = TRUE)),
@@
 
-```{r regr-air-season, fig.cap = "Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with SAM, A\\nobreakdash-SAM and S\\nobreakdash-SAM for the 1979 -- 2018 period.
+(ref:regr-air-season-cap) Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with SAM, A\nobreakdash-SAM and S\nobreakdash-SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Gray areas in Antarctica have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*", cache = FALSE}
+Note that the colour scale cuts-off at $\pm0.6 \mathrm{K}$ so the areas of very high values in the Antarctic don't wash out detail in the mid-latitudes and tropics.
+
+```{r regr-air-season, fig.cap = "(ref:regr-air-season-cap)", fig.width = 6, fig.height = 8, fig.env = "figure*", cache = FALSE}
+
+anchor_limits <- function(anchor = 0, binwidth = NULL, exclude = NULL,  bins = 10,
+                          range = c(NA, NA)) {
+  force(range)
+  force(anchor)
+  force(binwidth)
+  force(exclude)
+  force(bins)
+  function(x, binwidth2 = NULL) {
+    if (!is.null(binwidth)) binwidth2 <- binwidth
+    if (is.null(binwidth2)) {
+      binwidth2 <- diff(pretty(x, bins))[1]
+    }
+
+    mult <- ceiling((x[1] - anchor)/binwidth2) - 1L
+    start <- anchor + mult*binwidth2
+    b <- seq(start, x[2] + binwidth2, binwidth2)
+    b <- b[!(b %in% exclude)]
+
+    if (!is.na(range[1])) {
+      m <- min(b)
+      b <- c(b[b >= range[1]], m)
+    }
+
+    if (!is.na(range[2])) {
+      m <- max(b)
+      b <- c(b[b <= range[2]], m)
+    }
+    sort(b)
+
+  }
+}
+
+discrete_div <- colorRampPalette(c(scales::muted("blue"), "white",
+                                   scales::muted("red")))
+
 air_regr %>% 
@@
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), 
-                    breaks = AnchorBreaks(0, 0.1, 0)) +
-  # geom_contour_tanaka2(aes(z = estimate),
-  #                     breaks = AnchorBreaks(0, 0.1, 0)) +
-  # geom_raster(aes(fill = estimate)) +
-  # geom_contour_fill(aes(z = p.value), breaks = c(0, 0.05), fill = "gray50", alpha = 0.2) +
+  geom_contour_fill(aes(z = estimate, fill = ..level..), 
+                    breaks = anchor_limits(0, 0.1, 0, range = c(-.6, .6))) +
   geom_contour2(aes(z = p.value), breaks = c(0, 0.05), color = "black", size = 0.3) +
-  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), fill = "gray") +
-  geom_contour2(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), color = "gray50") +
+
   geom_qmap(~.x[lat <= 0]) +
@@
   scale_y_latitude(labels = NULL, limits = c(NA, 0)) +
-  scale_fill_divergent("2-meter air temperature",  
-                       limits = c(-0.6, .6), 
-                       oob = scales::squish,
-                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
+  discrete_scale("fill", "name", name = "2-meter air temperature",
+                 palette = discrete_div,
+                 guide = guide_colorsteps_bottom(show.limits = TRUE)) +
   coord_polar() +
@@
 
-To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices we regressed surface land air and sea temperature and precipitation onto each of the three SAM indices at 700 hPa.
+To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices we regressed 2-meter air temperature and precipitation onto each of the three SAM indices at 700 hPa.
 As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
@@
 Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface land air and sea temperature for each trimester.
-In Summer, positive values of the SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
+In Summer, positive values of the SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies in the mid latitudes.
 The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E respectively.
@@
 Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics.
-Temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM.
-This signal is not present in the regression pattern with the SAM.
-Spring (Figure \@ref(fig:regr-air-season), row `r get_row(season == "SON")`) features similar patterns but of smaller magnitude, with fewer regions where regressed anomalies have statistical significance.
-
-In Autumn and Winter (rows `r combine_words(get_row(season %in% c("MAM", "JJA"), n = 2))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the SAM.
+Over Antarctica positive values of the SAM index are associated with negative temperature anomalies, particularly over the Eastern coast.
+These anomalies are associated only with the S\nobreakdash-SAM.
+On the other hand, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.
+
+In Autumn, Winter and Spring (rows `r combine_words(get_row(season %in% c("MAM", "JJA", "SON"), n = 3))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to Summer.
 There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America. 
-These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following @wilks2016.
-However, repeating this analysis with 2-meter air temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
-Moreover, similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
+Similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
+In Spring, the tropical signal of A\nobreakdash-SAM is similar to the one in Summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link.
+
+The same regressions using temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid are shown in Figure\ \@ref{fig:A10}.
+Where data is available, the general patterns are similar, although with reduced magnitude, perhaps due to the fact that sea surface temperatures vary less.
 
@@
 It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in Summer and Spring).
+Temperatures over Eastern Antarctica are most impacted by the S\nobreakdash-SAM, while on Western Antarctica they are more sensitive to the A\nobreakdash-SAM. 
+Considering the positive trend in the S\nobreakdash-SAM index, this might be consistent with temperature trends, which show warming in the West and no trend in the East [@nicolas2014].
 
@@
   tags <- composites[order(sign, season)][sign != 0, paste0("n: ", unique(n)), by = .(sign, season)]
-  if (this_lev == 50){
+  if (this_lev == 50) {
     limits <- c(-119, 179) 
@@
 
+```{r compute-air-regr-noaa}
+time_subset <- list(time = range(indices$time))
+air <- NOAATemp() %>% 
+  ReadNetCDF(vars = "air", subset = c(time_subset, list(lat = c(-90, 10)))) %>% 
+  normalise_coords()  %>%
+  # Interpolate at lon = 0. Otherwise, can't close contours. 
+  qwrap(lon = c(0, 360) ~ c(-2.5, 360)) %>% 
+  .[ , Interpolate(air ~ lon + lat, x.out = c(0, unique(lon)), y.out = unique(lat)), by = time] %>% 
+  .[lon != -2.5]
+
+# air <- ReadNetCDF("~/DATOS/reanalysis/ERA5/mon/era5sl.mon.mean.nc", 
+#            vars = c(air = "t2m"), subset = c(time_subset, list(lat = c(-90, 10)))) %>% 
+#   normalise_coords()
+
+air_regr <- air %>% 
+  # .[, air := Anomaly(air), by = .(lon, lat)] %>%
+  .[indices_wide[lev == 700], on = .NATURAL] %>% 
+  # mei[., on = .NATURAL] %>% 
+  rm_singleton() %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
+  .[, nas := mean(is.na(air)), by = .(lon, lat, season(time))] %>%
+  .[nas >= 0.15, air := NA] %>%
+  .[, rbind(as.data.table(FitLm(air, asym, sym, se = TRUE)),
+            as.data.table(FitLm(air, full, se = TRUE))),
+    by = .(lon, lat, season(time))] %>% 
+  rm_intercept() %>% 
+  .[, term := factor(term, levels = names(lab_sam), ordered = TRUE)] 
+```
+
+```{r A10, fig.cap = "Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with SAM, A\\nobreakdash-SAM and S\\nobreakdash-SAM for the 1979 -- 2018 period.
+Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
+Gray areas in Antarctica have more than 15\\% of missing data.", fig.width = 6, fig.height = 8, fig.env = "figure*", cache = FALSE}
+air_regr %>% 
+  copy() %>% 
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(term, season)] %>% 
+  periodic(lon = c(0, 360)) %>% 
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate), 
+                    breaks = AnchorBreaks(0, 0.1, 0)) +
+  geom_contour2(aes(z = p.value), breaks = c(0, 0.05), color = "black", size = 0.3) +
+  geom_contour_fill(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), fill = "gray") +
+  geom_contour2(aes(z = as.numeric(is.na(estimate))), breaks = c(0.5, 2), color = "gray50") +
+  geom_qmap(~.x[lat <= 0]) +
+  geom_coords() +
+  scale_x_longitude(labels = NULL) +
+  scale_y_latitude(labels = NULL, limits = c(NA, 0)) +
+  scale_fill_divergent("2-meter air temperature",  
+                       limits = c(-0.6, .6), 
+                       oob = scales::squish,
+                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
+  coord_polar() +
+  facet_grid(season ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  tag_facets("cr", tag_suffix = "") +
+  no_grid 
+```
+
 \newpage

### 2021-05-04 R/data_locations.R

@@ -45,6 +45,14 @@ ERA5 <- function() {
 }
 
 
+#' @export
+#' @rdname data_locations
+ERA52mt <- function() {
+   file <- data_path("raw", "era5.2mt.mon.mean.nc")
+   checkmate::assert_access(file, access = "r")
+   class(file) <- c("nc_file", class(file))
+   return(file)
+}
 
 #' @export
 #' @rdname data_locations

### 2021-05-04 analysis/paper/asymsam.Rmd
Agrega comentario de que 50 hPa no está centrado
#18

@@
   list(
-    geom_segment(data = data.frame(xstart =  seq(0, 360-30, by = 30), 
-                                   xend =  seq(0, 360-30, by = 30),
-                                   ystart = -90+15,
+    geom_segment(data = data.frame(xstart =  seq(0, 360 - 30, by = 30), 
+                                   xend =  seq(0, 360 - 30, by = 30),
+                                   ystart = -90 + 15,
                                    yend = Inf),
@@
 
-    # geom_vline(xintercept = seq(0, 360-30, by = 30), size = 0.1, alpha = 0.5),
-
       geom_hline(yintercept = seq(-90, 0, by = 15), size = 0.1, alpha = 0.5),
       shadowtext::geom_shadowtext(data = data.frame(x = 0, y = seq(-90 + 15, 0, by = 15)),
-                                  aes(x, y, label = LatLabel(y)), size = 2, alpha = 0.5,
+                                  aes(x, y, label = LatLabel(y)), size = 1.5, alpha = 0.5,
                                   colour = "black",
@@
 In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \@ref(fig:2d-regr)a.1) which is not centred in the South Pole.
-On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric.
+On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric, although still is not perfectly centred in the South Pole. 
 Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Drake Passage in the Western Hemisphere and Davis Sea in the Eastern Hemisphere

### 2021-05-04 R/eof_methods.R

@@ -88,19 +88,19 @@ as.data.table.eof <- function(x, keep.rownames = FALSE, n = NULL, ...) {
 #    r
 # }
 
-labeller <- function(...) {
-   UseMethod("labeller")
-}
-
-labeller.default <- function(...) {
-   ggplot2::labeller(...)
-}
-
-labeller.eof <- function(eof, sep = " - ") {
-   var <- attr(eof, "suffix")
-   with(eof$sdev, setNames(paste0(get(var), sep, scales::percent(r2)),
-                           get(var)))
-}
+# labeller <- function(...) {
+#    UseMethod("labeller")
+# }
+#
+# labeller.default <- function(...) {
+#    ggplot2::labeller(...)
+# }
+#
+# labeller.eof <- function(eof, sep = " - ") {
+#    var <- attr(eof, "suffix")
+#    with(eof$sdev, setNames(paste0(get(var), sep, scales::percent(r2)),
+#                            get(var)))
+# }
 
 # print.eof <- function(eof) {
 #    cat("right:\n")

### 2021-05-04 analysis/paper/asymsam.Rmd
Agrega comentario de representatividad DJF en fig 7
#17
Arregla orden en figura 7
Closes #28

@@
   .[lev %in% c(50, 700)] %>%
-  .[, term := factor(term, levels = c("full", "asym", "sym"), ordered = TRUE)] %>% 
+  .[, term := factor(term, levels = c("full", "asym", "sym"))] %>% 
   ggplot(aes(lon, lat)) +
@@
   geom_qmap(~.x[lat <= -20]) +
-  geom_coords() +
+  # geom_coords() +
   geom_point(data = data.frame(lon = ConvertLongitude(c(50, 166, -76)),
-                               lat = -49, term = "asym", lev = 700), 
+                               lat = -49, term = factor("asym", 
+                                                        levels = c("full", "asym", "sym")),
+                               lev = 700),
              shape = 21, size = 2.5, color = "black", fill = "#de3e80") +
@@
   coord_polar() +
-  facet_grid(lev ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
+  facet_grid(lev ~ term, labeller = labeller(lev = lev.lab, term = lab_sam)) +
   tag_facets("cr") +
@@
 The regression patterns associated with the A\nobreakdash-SAM and S\nobreakdash-SAM indices successfully disentangle both structures. 
+(Note that, in light of previous discussion about the atypical nature of DJF, this year-round average effect  will likely not represent DJF perfectly).
 While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \@ref(fig:2d-regr)b.2), the S\nobreakdash-SAM index is associated with an annular structure, with only vestigial zonal asymmetries (Figure \@ref(fig:2d-regr)c.2) in the shape of a wave-3 which is the inverse of the A\nobreakdash-SAM wave-3.

### 2021-05-04 R/lm_lite.R
Setea renv
Closes #24

@@ -35,14 +35,7 @@ var.wt <- function(x, w, na.rm = FALSE) {
   (sum(w*x^2) * sum.w - sum(w*x)^2) / (sum.w^2 - sum(w^2))
 }
 
-#' @export
-Detrend <- function(y, x) {
-   nas <- is.na(y)
-   m <- mean(y, na.rm = TRUE)
-   if (!hasArg(x)) x <- seq_along(y)
-   y[!nas] <- .lm.fit(cbind(1, x[!nas]), y[!nas])$residuals
-   return(y + m)
-}
+
 
 #' Compute p.values
 #'

### 2021-05-04 R/maps.R

@@ -5,6 +5,7 @@
 #' @param color,size,fill parameters passed to [ggplot2::geom_polygon()]
 #' @param wrap vector of length 2 interpreted as the longitude range of a global map (see [maps::map()])
 #' @param keep proportions of point to retain (passed to [rmapshaper::ms_simplify()])
+#' @param weighting weighting parameter passsed to [rmapshaper::ms_simplify()]
 #' @param ... other arguments passed to [ggplot2::geom_polygon()]
 #'
 #' @export
@@ -48,7 +49,7 @@ map_simple <- function(wrap = c(0, 360), keep = 0.015, weighting = 0.7) {
 fortify <- ggplot2::fortify
 
 # zzz.R
-.onLoad <- function(pkgname, libname) {
+.onLoad <- function(libname, pkgname) {
   map_simple <<- memoise::memoise(map_simple)
   fortify <<- memoise::memoise(fortify)
 }

### 2021-05-04 R/spectrum.R

@@ -7,15 +7,17 @@
 #' @param B number of bootstrap samples.
 #' @param spans vector of odd integers giving the widths of modified Daniell
 #' smoothers to be used to smooth the periodogram
+#' @param probs quantile for confidence interval.
 #' @param ... other arguments passed to [stats::spec.pgram()]
 #'
 #' @export
 fftspectrum <- function(x, spans = NULL, B = 10000, ..., probs = 0.95) {
-  mtm <- spec.pgram(x, spans = spans, ..., plot = FALSE)
+  ar_spectrum <- NULL
+  mtm <- stats::spec.pgram(x, spans = spans, ..., plot = FALSE)
 
-  out <- as.data.table(mtm[c("freq", "spec")])
+  out <- data.table::as.data.table(mtm[c("freq", "spec")])
 
-  ar <- ar(ts(x))
+  ar <- ar(stats::ts(x))
   # rho <- a$ar
   # var <- a$var.pred
   out[, ar_spectrum := arspectrum(mtm$freq, ar$ar, ar$var.pred)]
@@ -33,21 +35,21 @@ arspectrum <- function(freq, rho, var) {
 
 
 null_ar_spectrum_ <- function(B = 100, n, ar, spans = NULL, ..., probs = 0.95) {
-  y <- as.vector(arima.sim(model = list(ar = ar$ar), n = n))*sqrt(ar$var.pred)
-  nfreq <- length(spec.pgram(y, spans = spans, ..., plot = FALSE)$spec)
+  y <- as.vector(stats::arima.sim(model = list(ar = ar$ar), n = n))*sqrt(ar$var.pred)
+  nfreq <- length(stats::spec.pgram(y, spans = spans, ..., plot = FALSE)$spec)
   boots <- vapply(seq_len(B), function(b) {
-    y <- as.vector(arima.sim(model = list(ar = ar$ar), n = n))*sqrt(ar$var.pred)
-    spec.pgram(y, spans = spans, plot = FALSE)$spec
+    y <- as.vector(stats::arima.sim(model = list(ar = ar$ar), n = n))*sqrt(ar$var.pred)
+    stats::spec.pgram(y, spans = spans, plot = FALSE)$spec
 
   }, numeric(nfreq))
-  data.table::as.data.table(t(apply(boots, 1, quantile, probs = probs)))
+  data.table::as.data.table(t(apply(boots, 1, stats::quantile, probs = probs)))
 }
 
 null_ar_spectrum <- memoise::memoise(null_ar_spectrum_)
 
[…]

### 2021-05-04 R/theme.R

@@ -13,7 +13,7 @@ theme_asymsam <- function(base_size = 16) {
       # text = element_text(family = font_rc),
       legend.position = "bottom", legend.box = "vertical",
       tagger.panel.tag.background = ggplot2::element_rect(color = NA),
-      strip.text = element_text(size = 12),
+      strip.text = ggplot2::element_text(size = 12),
       # panel.spacing.y = grid::unit(5, "mm"),
       # panel.spacing.x = grid::unit(5, "mm"),
       # legend.spacing = grid::unit(2, "mm"),

### 2021-05-04 analysis/paper/asymsam.Rmd
Agrega regresión global de pp
Closes #22

@@
       shadowtext::geom_shadowtext(data = data.frame(x = 0, y = seq(-90 + 15, 0, by = 15)),
-                                  aes(x, y, label = LatLabel(y)), size = 1.5, alpha = 0.5,
+                                  aes(x, y, label = LatLabel(y)), size = 1.5, alpha = 0.7,
                                   colour = "black",
@@
   geom_qmap(~.x[lat <= -20]) +
-  # geom_coords() +
+  geom_coords() +
   geom_point(data = data.frame(lon = ConvertLongitude(c(50, 166, -76)),
@@
 
-Regression of the SAM indices with seasonal mean precipitation and 700 hPa geopotential height are shown in Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) for New Zealand and neighbouring islands, and South America respectively.
-South Africa is not shown because no significant signal was detected there.
-
 ```{r calc-indices2d-seasonal}
@@
                            time = range(indices$time))) %>% 
-  na.omit() %>% 
   normalise_coords() %>%
@@
 
-# mei <- rsoi::download_mei() %>% 
-#   as.data.table() %>% 
-#   .[, .(time = lubridate::as_datetime(Date), mei = MEI)]
-
 pp_regr_season <- pp %>% 
   .[indices_wide[lev == 700], on = .NATURAL] %>%
-  na.omit() %>% 
-  .[, precip_a := Anomaly(precip), by = .(lon, lat, month(time))] %>% 
+
+  .[, precip_a := Anomaly(precip,na.rm = TRUE), by = .(lon, lat, month(time))] %>% 
   .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
@@
   .[nas >= 0.15, precip := NA] %>% 
-  # na.omit() %>% 
-  .[is.finite(precip_a)] %>%
+  # .[is.finite(precip_a)] %>%
   .[, combine_lists(FitLm(precip_a, asym, sym, se = TRUE),
@@
   .[indices_wide[lev == 700], on = .NATURAL] %>% 
-  na.omit() %>% 
-  .[, precip_a := Anomaly(precip), by = .(lon, lat, month(time))] %>% 
+  .[, precip_a := Anomaly(precip, na.rm = TRUE), by = .(lon, lat, month(time))] %>% 
   .[, nas := mean(is.na(precip)), by = .(lon,  lat)] %>%
   .[nas >= 0.15, precip := NA] %>% 
-  .[is.finite(precip_a)] %>%
+  # .[is.finite(precip_a)] %>%
   .[, combine_lists(FitLm(precip_a, asym, sym, se = TRUE),
@@
 ```
+
+(ref:global-pp-cap) Regression of annual and seasonal mean precipitation anomalies (mm per day, shading)  with a) SAM, b) A\\nobreakdash-SAM and c) S\\nobreakdash-SAM for the 1979 -- 2018 period.
+Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
+Note that the colour scale cuts-off at $\pm0.25 \mathrm{K}$ so the areas of very high values in the Tropics don't wash out detail in the mid and high-latitudes.
+
+```{r global-pp, fig.cap = "(ref:global-pp-cap)", fig.width=6, fig.height=3.5}
+discrete_div_pp <- colorRampPalette(c("#8E521C", "white", "#00665E"))
+
+breaks_pp <- sort(c(seq(-0.25, 0.25, by = 0.05), -0.7, 0.75))
+breaks_pp <- breaks_pp[breaks_pp != 0]
+
+lons <- c(0, unique(pp_regr$lon), 360)
+pp_regr %>%
+  .[season == "Annual"] %>%
+  .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, term)] %>% 
+  # Need to interpolate 0 and 360 to get closed contours
+  qwrap(lon = c(1.25, 360 + 1.25) ~ c(-1.25, 360 + 1.25)) %>% 
+  .[, lapply(.SD, function(x) approx(lon, x, xout = lons)$y), by = .(lat, season, term)] %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate, fill = ..level..), 
+                    breaks = breaks_pp) +
+  stat_subset(aes(subset = is.na(estimate)), geom = "tile", color = "gray50") +
+  geom_contour2(aes(z = p.value), breaks = 0.05) +
+  geom_qmap(~.x[lat <= 10], keep = 0.015, weighting = 0.9) +
+  scale_x_longitude(breaks = NULL) +
+  scale_y_latitude(breaks = NULL, limits = c(-90, NA)) +
+  discrete_scale("fill", "name", name = "Precipitation",
+                 palette = discrete_div_pp,
+                 guide = guide_colorsteps_bottom(show.limits = TRUE)) +
+  scale_linetype(guide = "none") +
+  coord_polar() +
+  geom_coords() +
+  # coord_quickmap(ylim = c(NA, 0)) +
+  facet_grid(.~term, labeller = labeller(term = lab_sam)) +
+  tag_facets("panel", position = "tl")  +
+  no_grid 
+```
+
+Figure\ \@ref(fig:global-pp) shows regression of the SAM indices with precipitation for the Southern Hemisphere. 
+The precipitation signal associated with SAM shows generally decreased precipitation at around 45ºS, somewhat increase precipitation around 30ªS (Fig.\ \@ref(fig:global-pp)a) and increased precipitation over Antarctica, a pattern known from other studies [e.g.
+@hendon2014].
+This pattern is mostly unchanged between seasons albeit from varying in intensity. 
+Panels b and c in Figure\ \@ref(fig:global-pp) shows that the A\nobreakdash-SAM signal is only in the tropics and mid-latitudes, while the S\nobreakdash-SAM signal is strong in the higher latitudes. 
+In particular S\nobreakdash-SAM is associated with the increased precipitation over Antarctica and decrease precipitation around the Southern Ocean.
+
+To study in more detail the impacts over land, Figures \@ref(fig:pp-regr-oceania) and \@ref(fig:pp-regr-america) show regression of the SAM indices with seasonal mean precipitation and 700 hPa geopotential height for New Zealand and neighbouring islands, and South America respectively.
+South Africa is not shown because no significant signal was detected there.
 

### 2021-05-04 analysis/paper/asymsam.Rmd
Knitea

@@
   stat_subset(aes(subset = is.na(estimate)), geom = "tile", color = "gray50") +
-  geom_contour2(aes(z = p.value), breaks = 0.05) +
+  geom_contour2(aes(z = p.value), breaks = 0.05, size = 0.2) +
   geom_qmap(~.x[lat <= 10], keep = 0.015, weighting = 0.9) +

### 2021-05-04 analysis/paper/asymsam.tex
Knitea

--- 2020-12-15 analysis/paper/paper.tex
@@
 \usepackage{natbib}
+\usepackage{lineno}
+\linenumbers
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
@@
         Elio Campitelli \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
- CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
- CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.\\
+CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.\\
+CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
      \email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}  %  \\
@@
         Leandro B. Díaz \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
- CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
- CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.\\
+CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.\\
+CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
      %  \\
@@
         Carolina Vera \at
-     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
- CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
- CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
+     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.\\
+CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.\\
+CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
      %  \\
@@
 Of the several different indices presented in the literature, many of them are based on zonal means of sea level pressure or geopotential height \citep{ho2012}.
-\citet{gong1999} defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 60\degree S, which is also the definition used by the station-based index in \citet{marshall2003}.
+\citet{gong1999} defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 65\degree S, which is also the definition used by the station-based index in \citet{marshall2003}.
 \citet{baldwin2009} proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
@@
 The impact of the zonally asymmetric component of the SAM at regional scales has not been studied in detail yet.
-The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at higher latitudes \citep{jones2019} (and vice versa for negative SAM).
+The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at lower latitudes \citep{jones2019} (and vice versa for negative SAM).
 But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}.
 The SAM-related signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry \citep{lim2016}.
-The importance of zonal asymmetries of the SAM on these impacts have been studied in certain regions.
-For example, the SAM-precipitation relationship in Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}.
+The SAM-precipitation relationship in Southeastern South America can be explained by the Pacific-South American (PSA)-like zonally asymmetric circulation associated with the SAM \citep{silvestri2009, rosso2018}.
 \citet{fan2007} also found that precipitation in East Asia was impacted by the variability of only the Western Hemisphere part of the SAM.
@@
 
-We used monthly geopotential height at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{hersbach2020} for the period 1979 to 2018.
+We used monthly geopotential height at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels as well as 2 metre temperature from ERA5 \citep{hersbach2020} for the period 1979 to 2018.
 We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations.
 
-Temperature data used are from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid 5\degree~longitude by 5\degree~latitude.
-Monthly rainfall fields at 0.5\degree~longitude by 0.5\degree~latitude data from the Global Precipitation Climatology Centre \citep{schneider2015, schneider2017} are also considered.
-The rainfall dataset is based on station-based records, and thus it only has continental coverage.
+For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation {[}@xie1997{]}, with a 2.5\textbackslash degree\textbackslash{} resolution in latitude and longitude.
+This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
 
@@
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
-Arbitrary units.}\label{fig:method}
+Arbitrary units; positive values in blue and negative values in red.}\label{fig:method}
 \end{figure*}
@@
 The following seasons were considered -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON).
-Results are very similar to each other in the troposphere (Figure \ref{fig:A5}, row 1) and show spatial correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
-In the stratosphere (Figure \ref{fig:A5}, row 2), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree~in comparison with the rest of the year.
+Results are very similar to each other in the troposphere (Figure \ref{fig:A5}, row 2) and show spatial correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON).
+In the stratosphere (Figure \ref{fig:A5}, row 1), patterns are similar for all seasons except DJF, when the wave-1 zonal anomalies are rotated 90\degree~in comparison with the rest of the year.
 Spatial correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA).
-Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year with the exception of DJF in the stratosphere.
+Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year.
+The outlier being DJF, which shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure @ref\{fig:A3\}), which is consistent with \citet{fogt2020}.
+Therefore, even though we do our analysis for all months, we would expect for it to represent the other trimesters better.
 
@@
 Zonal asymmetric patterns of SAM were computed for the two halves of the period (1979 to 1998 and 1999 to 2018) respectively.
-They show no systematic change neither in the stratosphere nor in the troposphere (Figure \ref{fig:A6}).
+The differences between the two periods appear to be relatively small in both the troposphere and stratosphere (Figure \ref{fig:A6}).
 
@@
 Moreover, we apply multiple linear regression analysis to describe the combined influence of both A\nobreakdash-SAM and S\nobreakdash-SAM.
-To obtain the linear coefficients of a variable \(X\) (geopotential, temperature, precipitation, etc\ldots{}) with A\nobreakdash-SAM and S\nobreakdash-SAM we fit the equation
+To obtain the linear coefficients of a variable \(X\) (geopotential, temperature, precipitation, etc\ldots) with A\nobreakdash-SAM and S\nobreakdash-SAM we fit the equation
 
@@
 
-The stratospheric S\nobreakdash-SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \ref{fig:A2}).
-This might suggest a link between stratospheric SAM variability and the Quasi-Biennial Oscillation \citep{baldwin2001b}.
-A local peak at 2 years years is discernible in the periodogram of the tropospheric S\nobreakdash-SAM, although it's not statistically significant.
+The stratospheric S\nobreakdash-SAM varies strongly with a period between 15 and 30 months (the maximum spectrum is located at 20 months), which can be seen by spectral analysis (Figure \ref{fig:A2}).
+A local peak at a similar frequency range is discernible in the periodogram of the tropospheric S\nobreakdash-SAM, although it's not statistically significant.
+This period band is around the range of periodicity of the Quasi-Biennial Oscillation \citep{baldwin2001b} and is consistent with \citep{vasconcellos2020}, who found that the SAM and the QBO share significant common high power around the 2-year band.
+The fact that this periodicity is not evident on the A\nobreakdash-SAM index, is also consistent with their composites of geopotential height anomalies during easterly and westerly QBO, which show a rather symmetric monopole over Antarctica.
 In the troposphere the most significant peak of variability is found in A\nobreakdash-SAM at around 3.6 months.
@@
 For the SAM (Figure \ref{fig:cross-correlation}a), high values below 100~hPa reflect the vertical (zero-lag) coherency throughout the troposphere.
-Above 100~hPa, correlation between levels does falls off more rapidly, indicating less coherent (zero-lag) variability.
-But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g.~grater than 0.4 between tropospheric levels and levels below 30~hPa).
+Above 100~hPa, correlation between levels falls off more rapidly, indicating less coherent (zero-lag) variability.
+But correlations between tropospheric and lower-to-middle stratospheric levels are still relatively high (e.g.~greater than 0.4 between tropospheric levels and levels below 30~hPa).
 A\nobreakdash-SAM and S\nobreakdash-SAM (Figure \ref{fig:cross-correlation}b and c, respectively) share similar high level of coherency in the troposphere but they differ in their stratospheric behaviour.
@@
 
-Figure \ref{fig:r-squared-trend} shows trends for the explained variance of each index.
-There is no evidence of a significant trend in the stratosphere.
-In the troposphere, there is a positive trend for A\nobreakdash-SAM and no significant trend for S\nobreakdash-SAM.
-This suggest that the SAM has become more asymmetric in the period from 1979 to 2018.
-However, the change is slight, around 1\% increased explained variance per decade.
+A positive trend in the S\nobreakdash-SAM index and no trend in the A\nobreakdash-SAM index might at first suggest a trend towards a more symmetric SAM.
+However, a very negative S\nobreakdash-SAM trending towards a less negative S\nobreakdash-SAM would be translated into a positive S\nobreakdash-SAM trend but a more asymmetric SAM.
 
 \begin{figure}
-\includegraphics{r-squared-trend-1} \caption{Linear trends (in percent per decade) of the variance explained by A\nobreakdash-SAM and S\nobreakdash-SAM at each level for the period 1979 -- 2018.
+\includegraphics{r-squared-trend-1} \caption{Linear trends (in percent per decade) of the variance explained by A\nobreakdash-SAM and S\nobreakdash-SAM at each level and for each trimester for the period 1979 -- 2018.
 Shading indicates the 95\% confidence interval.}\label{fig:r-squared-trend}
 \end{figure}
+
+To study the question of whether the SAM is becoming more or less asymmetric, we show trends for the explained variance of each index for each trimester in Figure \ref{fig:r-squared-trend}.
+In the troposphere the only significant trend is in DJF, in which the A\nobreakdash-SAM has a positive trend of about 2\% per decade, suggesting, that the DJF SAM has become more asymmetric in the period from 1979 to 2018,
+However due to the atypical nature of the asymmetric component of the SAM during DJF (Section~\ref(definition-of-indices)) this should be taken only as preliminary evidence.
+The other significant trends is in the stratosphere during SON, where there is a positive trend in the variance explained by the S\nobreakdash-SAM of about 4\% per decade.
+This change could be the result of the forcing from ozone depletion.
 
@@
 In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \ref{fig:2d-regr}a.1) which is not centred in the South Pole.
-On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \ref{fig:2d-regr}c.1) is more symmetric.
-Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Bellinghausen Sea on the Western Hemisphere and Davis Sea in the Eastern Hemisphere
+On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \ref{fig:2d-regr}c.1) is more symmetric, although still is not perfectly centred in the South Pole.
+Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Drake Passage in the Western Hemisphere and Davis Sea in the Eastern Hemisphere
 
@@
 The regression patterns associated with the A\nobreakdash-SAM and S\nobreakdash-SAM indices successfully disentangle both structures.
-While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \ref{fig:2d-regr}b.2), the S\nobreakdash-SAM index is associated with an annular structure, almost devoid of zonal asymmetries (Figure \ref{fig:2d-regr}c.2).
+(Note that, in light of previous discussion about the atypical nature of DJF, this year-round average effect will likely not represent DJF perfectly).
+While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \ref{fig:2d-regr}b.2), the S\nobreakdash-SAM index is associated with an annular structure, with only vestigial zonal asymmetries (Figure \ref{fig:2d-regr}c.2) in the shape of a wave-3 which is the inverse of the A\nobreakdash-SAM wave-3.
 The wave-3 pattern observed in Figure \ref{fig:2d-regr}b.2 is rotated by half a wavelength from the average position of the mean wave-3 pattern described by \citet{raphael2004}, whose reference locations are marked with points in the figure.
@@
 On the other hand, the S\nobreakdash-SAM it is almost entirely explained by the zonal mean at both levels (Figure \ref{fig:wave-amplitude} column c), with little to no contribution from zonal waves with wavenumbers 1 to 3.
-
-Looking at Figure \ref{fig:2d-regr}b.2, it becomes apparent that zonal waves 1 and 2 modulate the amplitude of zonal wave 3, which -- as mentioned before -- is larger in the Western Hemisphere than in the Eastern Hemisphere.
 
@@
 
-The wave-3 pattern from Figure \ref{fig:2d-regr}b.2 is very similar to the PSA Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}.
+\begin{table}
+
+\caption{\label{tab:enso-cor-table}Correlation between SAM indices and the Oceanic Niño Index.
+p-values corrected for False Detection Rate in parenthesis.
+In bold, correlations with p-value smaller than 0.05.}
+\centering
+\begin{tabular}[t]{c>{}c>{}c>{}c}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{1}{c}{Correlation} & \multicolumn{2}{c}{Partial correlation} \\
+\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-4}
+ & SAM & A-SAM & S-SAM\\
+\midrule
+ & \textbf{-0.17} & \textbf{-0.26} & 0.02\\
+\cmidrule{2-4}
+\multirow[t]{-2}{*}{\centering\arraybackslash Year} & \textbf{(0.001)} & \textbf{(<0.001)} & (0.775)\\
+\cmidrule{1-4}
+ & \textbf{-0.31} & \textbf{-0.30} & -0.17\\
+\cmidrule{2-4}
+\multirow[t]{-2}{*}{\centering\arraybackslash DJF} & \textbf{(0.002)} & \textbf{(0.003)} & (0.115)\\
+\cmidrule{1-4}
+ & -0.07 & \textbf{-0.26} & 0.14\\
+\cmidrule{2-4}
+\multirow[t]{-2}{*}{\centering\arraybackslash MAM} & (0.530) & \textbf{(0.011)} & (0.192)\\
+\cmidrule{1-4}
+ & 0.01 & -0.14 & 0.11\\
+\cmidrule{2-4}
+\multirow[t]{-2}{*}{\centering\arraybackslash JJA} & (0.900) & (0.192) & (0.300)\\
+\cmidrule{1-4}
+ & \textbf{-0.25} & \textbf{-0.42} & 0.05\\
+\cmidrule{2-4}
+\multirow[t]{-2}{*}{\centering\arraybackslash SON} & \textbf{(0.014)} & \textbf{(<0.001)} & (0.686)\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+The wave-3 pattern from Figure~\ref{fig:2d-regr}b.2 is very similar to the PSA Pattern \citep{mo1987, kidson1988} which is a teleconnection pattern associated with the ENSO \citep{karoly1989}.
 Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is -0.16 (p-value \textless{} 0.001).
-This relationship is captured mainly by the A\nobreakdash-SAM, as this index has a partial correlation of -0.26 (p-value \textless{} 0.001) with the ONI, consistent with \citet{fan2007}'s results, whereas the S\nobreakdash-SAM's partial correlation with the ONI is essentially null (0.019; p-value = 0.67).
+The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is shown in Table~@ref\{tab:enso-cor-table\} for each SAM index and for each trimester.
+There is a significant correlation between SAM and ENSO.
+When divided in trimesters, this correlation is important only in DJF and SON.
+This relationship is captured mainly by the A\nobreakdash-SAM, as this index has signifcant partial correlations with ENSO while correlations with S\nobreakdash-SAM are all smaller and non-significant.
+In MAM, ENSO is not significantly correlated with SAM, but it is significantly correlated with A\nobreakdash-SAM in a level comparable with the ENSO-SAM correlation in SON.
 The same analysis was performed using the Multivariate ENSO Index \citep{wolter2011} and the Southern Oscillation Index \citep{ropelewski1987}, obtaining similar results.
-The latter allows to conclude that these results do not depend on the ENSO index used.
+The latter allows us to conclude that these results do not depend on the ENSO index used.
 
@@
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Gray areas in Antarctica have more than 15\% of missing data.}\label{fig:regr-air-season}
+Note that the colour scale cuts-off at \(\pm0.6 \mathrm{K}\) so the areas of very high values in the Antarctic don't wash out detail in the mid-latitudes and tropics.}\label{fig:regr-air-season}
 \end{figure*}
 
-To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices we regressed surface land air and sea temperature and precipitation onto each of the three SAM indices at 700~hPa.
+To assess differences in the impacts associated with the SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices we regressed 2-meter air temperature and precipitation onto each of the three SAM indices at 700~hPa.
 As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
+Regressions were done without detrending either the variables nor the indices, but computing the regressions with detrended values doesn't change the results (not shown).
 
 Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700~hPa with surface land air and sea temperature for each trimester.
-In Summer, positive values of the SAM index (Figure \ref{fig:regr-air-season}a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies.
+In Summer, positive values of the SAM index (Figure \ref{fig:regr-air-season}a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies in the mid latitudes.
 The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E respectively.
@@
 Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics.
-Temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM.
-This signal is not present in the regression pattern with the SAM.
-Spring (Figure \ref{fig:regr-air-season}, row 4) features similar patterns but of smaller magnitude, with fewer regions where regressed anomalies have statistical significance.
+Over Antarctica positive values of the SAM index are associated with negative temperature anomalies, particularly over the Eastern coast.
+These anomalies are associated only with the S\nobreakdash-SAM.
+On the other hand, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.
 
-In Autumn and Winter (rows 2 and 3 in Figure \ref{fig:regr-air-season}) the positive ring is only present through its local maximums in the regression with the SAM.
+In Autumn, Winter and Spring (rows 2, 3, and 4 in Figure \ref{fig:regr-air-season}) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to Summer.
 There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
-These patterns are not significant in the sense that there are no areas with p-values below 0.05 when controlling for FDR following \citet{wilks2016}.
-However, repeating this analysis with 2-meter air temperature from ERA5 resulted in similar patterns that were statistically significant (not shown).
-Moreover, similar features were observed in station measurements by \citet{jones2019}, although using data from 1957 to 2016.
+Similar features were observed in station measurements by \citet{jones2019}, although using data from 1957 to 2016.
+In Spring, the tropical signal of A\nobreakdash-SAM is similar to the one in Summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link.
+
+The same regressions using temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid are shown in Figure~@ref\{fig:A10\}.
+Where data is available, the general patterns are similar, although with reduced magnitude, perhaps due to the fact that sea surface temperatures vary less.
 
@@
 It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in Summer and Spring).
+Temperatures over Eastern Antarctica are most impacted by the S\nobreakdash-SAM, while on Western Antarctica they are more sensitive to the A\nobreakdash-SAM.
+Considering the positive trend in the S\nobreakdash-SAM index, this might be consistent with temperature trends, which show warming in the West and no trend in the East \citep{nicolas2014}.
 
@@
 
-Regression of the SAM indices with seasonal mean precipitation and 700~hPa geopotential height are shown in Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} for New Zealand and neighbouring islands, and South America respectively.
+\begin{figure}
+\includegraphics{global-pp-1} \caption{Regression of annual and seasonal mean precipitation anomalies (mm per day, shading) with a) SAM, b) A\textbackslash nobreakdash-SAM and c) S\textbackslash nobreakdash-SAM for the 1979 -- 2018 period.
+Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
+Note that the colour scale cuts-off at \(\pm0.25 \mathrm{K}\) so the areas of very high values in the Tropics don't wash out detail in the mid and high-latitudes.}\label{fig:global-pp}
+\end{figure}
+
+Figure~\ref{fig:global-pp} shows regression of the SAM indices with precipitation for the Southern Hemisphere.
+The precipitation signal associated with SAM shows generally decreased precipitation at around 45ºS, somewhat increase precipitation around 30ªS (Fig.~\ref{fig:global-pp}a) and increased precipitation over Antarctica, a pattern known from other studies \citep[e.g.][]{hendon2014}.
+This pattern is mostly unchanged between seasons albeit from varying in intensity.
+Panels b and c in Figure~\ref{fig:global-pp} shows that the A\nobreakdash-SAM signal is only in the tropics and mid-latitudes, while the S\nobreakdash-SAM signal is strong in the higher latitudes.
+In particular S\nobreakdash-SAM is associated with the increased precipitation over Antarctica and decrease precipitation around the Southern Ocean.
+
+To study in more detail the impacts over land, Figures \ref{fig:pp-regr-oceania} and \ref{fig:pp-regr-america} show regression of the SAM indices with seasonal mean precipitation and 700~hPa geopotential height for New Zealand and neighbouring islands, and South America respectively.
 South Africa is not shown because no significant signal was detected there.
@@
 
-The two-year periodicity we found in the stratospheric S\nobreakdash-SAM might point to a link between the SAM and the Quasi Biennial Oscillation.
+The periodicity of about two years we found in the stratospheric S\nobreakdash-SAM is consistent with the possible link between the SAM and the Quasi Biennial Oscillation as proposed by \citep{vasconcellos2020}.
 There is evidence of influence between the QBO and the Northern Annular Mode \citep[e.g.][]{holton1980, watson2014, zhang2020}, so it is not unlikely that the SAM would be similarly affected.
-However, establishing this link would require further research.
+Establishing this link would require further research.
 
@@
 We also find a statistically significant positive trend in the Symmetric component of the SAM in the stratosphere that is not evident in the SAM index.
-We find some evidence of the SAM becoming more zonally asymmetric, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
+We find some evidence of the SAM becoming more zonally asymmetric during DJF, as there is a slight positive trend in the variance explained by the as the A\nobreakdash-SAM.
 This is in contrast to \citet{fogt2012} who made the evaluation for the period 1958 -- 2001.
@@
 
-\citet{silvestri2009} suggests that precipitation impacts linked to the SAM changed strongly before and after 1980.
+Of note, even though the A\nobreakdash-SAM is significantly correlated with ENSO in all trimesters except JJA, precipitation patterns associated with the A\nobreakdash-SAM are similar to precipitation patterns associated with ENSO only in SON in Australia, but in all trimesters in South America.
+This might indicate that the A\nobreakdash-SAM is closely related to the pathway between ENSO and precipitation patterns in South America but not so much with the pathway between ENSO and precipitation in Australia.
+
+\citet{silvestri2009} suggest that precipitation impacts linked to the SAM changed strongly before and after 1980.
 In particular, the negative relationship with precipitation in South America was absent in some areas and switched sign in others in the earlier period.
 The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s \citep{fogt2006, clem2013}.
-Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the results presented here are most likely period-dependent.
+Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the magnitude and sign these impacts are most likely period-dependent and represent the average signal between 1979 and 2018.
 The decadal variations of the A\nobreakdash-SAM should be focus of future studies.
@@
 
-The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
+The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01.
+Elio Campitelli was supported by a PhD grant from CONICET, Argentina.
 
@@
 All data used in this paper is freely available from their respective sources.
-ERA5 data can be obtained via the Copernicus Climate Data Store (\url{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means}).
-NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sciences Laboratory website (\url{https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html} and \url{https://psl.noaa.gov/data/gridded/data.gpcc.html}).
+ERA5 data can be obtained via the Copernicus Climate Data Store (\href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means}{\textless https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means\textgreater{}}).
+CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at \url{https://psl.noaa.gov/data/gridded/data.cmap.html}.
 The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}
+The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at the public repository associated with this paper, at \url{https://github.com/eliocamp/asymsam/tree/master/analysis/data/derived_data}.
 
@@
 
+\begin{figure*}
+\includegraphics{A10-1} \caption{Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with SAM, A\nobreakdash-SAM and S\nobreakdash-SAM for the 1979 -- 2018 period.
+Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
+Gray areas in Antarctica have more than 15\% of missing data.}\label{fig:A10}
+\end{figure*}
+
 \newpage

### 2021-05-04 analysis/AGU/agu-abstract.R

@@ -1,8 +1,0 @@
----
-title: "Teasing out the asymmetric component of the Southern Annular Mode"
-output: pdf_document
----
-
-The conception of the Southern Annular Mode as a mostly zonally symmetric pattern is reflected not only in its name, but also on the various methods used to characterise it. Gong and Wang (1999) and many others define the SAM index based on zonal mean pressure anomalies. Baldwind & Thompson also put forward an index based on performin EOF on zonal mean geopotential height. However, composites, as well as regression patterns, based on these indinces un
-[@fogt2012 ]
-. Here, we propose a straightforward method of 

### 2021-05-06 analysis/paper/asymsam.Rmd
Achica espacio entre plots polares
Closes #29

@@
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
-Arbitrary units; positive values in blue and negative values in red.", fig.width=6, fig.height=2.5, fig.env="figure*"}
+Arbitrary units; positive values in blue and negative values in red.", fig.width=6, fig.height=2.7, fig.env="figure*"}
 sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
@@
   no_grid +
-  tag_facets()
+  tag_facets() +
+  theme(panel.spacing = grid::unit(-1, "lines"))
 ```
@@
   facet_grid(lev ~ term, labeller = labeller(lev = lev.lab, term = lab_sam)) +
-  tag_facets("cr") +
-  no_grid 
+  tag_facets("cr", position = list(x = 0.15, y = 0.85)) +
+  no_grid +
+  theme(panel.spacing = grid::unit(-1, "lines"))
 ```
@@
   facet_grid(season ~ term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets("cr", tag_suffix = "") +
-  no_grid 
+  tag_facets("cr", position = list(x = 0.15, y = 0.85)) +
+  no_grid +
+  theme(panel.spacing = grid::unit(-1, "lines"))
 ```
@@
   geom_contour2(aes(z = p.value), breaks = 0.05, size = 0.2) +
-  geom_qmap(~.x[lat <= 10], keep = 0.015, weighting = 0.9) +
+  geom_qmap(~.x[lat <= 0], keep = 0.015, weighting = 0.9) +
   scale_x_longitude(breaks = NULL) +
@@
   facet_grid(.~term, labeller = labeller(term = lab_sam)) +
-  tag_facets("panel", position = "tl")  +
-  no_grid 
+  # tag_facets("panel", position = "tl")  +
+  tag_facets("panel", position = list(x = 0.15, y = 0.85)) +
+  no_grid +
+  theme(panel.spacing = grid::unit(-1, "lines"))
 ```

### 2021-05-06 analysis/paper/asymsam.Rmd
Renderiza

@@
   comment = "#>",
-  dev = "pdf",
-  # dpi = 300,
+  dev = c("pdf", "png"),
+  dpi = 300,
   out.extra = ""
@@
   no_grid +
-  tag_facets() +
+  tag_facets(position = list(x = 0.15, y = 0.85)) +
   theme(panel.spacing = grid::unit(-1, "lines"))

### 2021-05-06 analysis/scripts/02-compute-eofs.R

@@ -18,7 +18,8 @@ if (!file.exists(indices_file)) {
     .[, hgt_a := hgt - mean(hgt), by = .(lon, lat, lev, month(time))]
 
   message("Computing eofs... (this might take a while, you can go make some coffee).")
-  indices <- hgt[lat <= -20, eof_asym(hgt_a, lon, lat, time, n = 1), by = lev]
+  indices <- hgt[lat <= -20, eof_asym(hgt_a, lon, lat, time, n = 1), by = lev] %>%
+    .[, PC := NULL]
   message("Done! Saving file.")
   saveRDS(indices, file = data_path("derived", "indices.Rds"))
   fwrite(indices, file = data_path("derived", "indices.csv"), yaml = TRUE)

### 2021-05-06 website/index.Rmd
Agrega source del sitio web
#27

---
title: "Asymmetric SAM"
description: |
  The Asymmetric and Symmetric SAM indices
site: distill::distill_website
bibliography: AsymSAM.bib
output: 
  distill::distill_article:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating websites with Distill at:
# https://rstudio.github.io/distill/website.html

# Learn more about publishing to GitHub Pages at:
# https://rstudio.github.io/distill/publish_website.html#github-pages

```

# What is the Asymmetric SAM

The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] on daily, monthly, and decadal timescales [@baldwin2001a; @fogt2006] and exerts an important influence on temperature and precipitation anomalies, and sea ice concentration [e.g. @fogt2020].
Its positive phase is usually described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes, but on top of that zonally symmetrical description, the SAM holds noticeable deviations from zonal symmetry (Figure \@ref(fig:method))

```{r method, fig.cap = "Spatial patterns of the first EOF of 700~hPa geopotential height for 1979 -– 2018 period. (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component. Arbitrary units; positive values in blue and negative values in red."}
knitr::include_graphics("method-1.png")
```

We derived a Asymmetric SAM index (A\nobreakdash-SAM) and a Symmetric SAM (S\nobreakdash-SAM) by computing the SAM as the leading EOF of geopotential height South of 20ºS (Fig. \@ref(fig:method)a), splitting that field into it's zonally asymmetric part (Fig. \@ref(fig:method)b) and zonally symmetric part (Fig. \@ref(fig:method)c) and finally projecting monthly geopotential fields onto each of these two new fields. 
The three indices are then normalized by dividing them by the standard deviation of the SAM index at each level.
As a result, the magnitudes between indices are comparable, but only SAM index has unit standard deviation per definition.

```{r 20-regr, fig.cap = "Regression of geopotential height (meters) at (row 1) 50~hPa and (row 2) 700~hPa with (column a) SAM, (column b) A\\nobreakdash-SAM, and (column c) S\\nobreakdash-SAM for the 1979 -- 2018 period. The regression patterns for A\\nobreakdash-SAM and S\\nobreakdash-SAM are the result of one multiple regression using both indices. Points marked on panel b.2 are the location of the reference points used by @raphael2004 for their Zonal Wave 3 index."}
knitr::include_graphics("2d-regr-1.png")
```

The resulting indices separate the zonally symmetric and zonally asymmetric part of the SAM (Figure \@ref(fig:20-regr)) and are associated with different impacts on temperature and precipitation.

For more details, see @campitelli2021.

# Get the data

You can grab the monthly indices for the 1979 -- 2018 period in [csvy](https://csvy.org/) format [here](indices.csv).

# How to cite

If you use this data on your research, please cite it with

Campitelli, E., Diaz, L., & Vera, C. (2021). Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach. Climate Dynamics, (under review).

Bibtex entry:

```
@article{campitelli2021,
  title = {Assessment of Zonally Symmetric and Asymmetric Components of the {{Southern Annular Mode}} Using a Novel Approach},
  author = {Campitelli, Elio and Diaz, Leandro and Vera, Carolina},
  year = {2021},
  volume = {(under review)},
  journal = {Climate Dynamics}
}
```

### 2021-05-13 analysis/paper/asymsam.Rmd
Agrega detalles

@@
 Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year.
-The outlier being DJF, which shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref{fig:A3}), which is consistent with @fogt2020.
-Therefore, even though we do our analysis for all months, we would expect for it to represent the other trimesters better.
+The DJF trimester shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref{fig:A3}), which is consistent with @fogt2020. 
+Therefore, we would expect that, even though the analysis is performed including all months, it represents more accurately the rest of the seasons.
 
@@
 ```
+
 To study the question of whether the SAM is becoming more or less asymmetric, we show trends for the  explained variance of each index for each trimester in Figure \@ref(fig:r-squared-trend). 
-In the troposphere the only significant trend is in DJF, in which the A\nobreakdash-SAM has a positive trend of about 2% per decade, suggesting, that the DJF SAM has become more asymmetric in the period from 1979 to 2018,
-However due to the atypical nature of the asymmetric component of the SAM during DJF (Section\ \ref(definition-of-indices)) this should be taken only as preliminary evidence. 
+In the troposphere the only significant trend is in DJF, in which the A\nobreakdash-SAM has a positive trend of about 2% per decade, suggesting that the DJF SAM has become more asymmetric in the period from 1979 to 2018.
+@fogt2012 observed a change from a more asymmetrical SAM before 1980 to a more symmetrical SAM after 1980, but our study period (1979 -- 2018) prevents us from detecting that change. 
+However due to the atypical nature of the asymmetric component of the SAM during DJF (Section\ \@ref(definition-of-indices)) this should be taken only as preliminary evidence. 
 The other significant trends is in the stratosphere during SON, where there is a positive trend in the variance explained by the S\nobreakdash-SAM of about 4% per decade. 
@@
 The regression patterns associated with the A\nobreakdash-SAM and S\nobreakdash-SAM indices successfully disentangle both structures. 
-(Note that, in light of previous discussion about the atypical nature of DJF, this year-round average effect  will likely not represent DJF perfectly).
+Note that, in light of previous discussion about the atypical nature of DJF, this year-round average effect  will likely not represent DJF perfectly.
 While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \@ref(fig:2d-regr)b.2), the S\nobreakdash-SAM index is associated with an annular structure, with only vestigial zonal asymmetries (Figure \@ref(fig:2d-regr)c.2) in the shape of a wave-3 which is the inverse of the A\nobreakdash-SAM wave-3.
@@
 As shown in previous sections, the three indices are highly coherent in the troposphere, so we select this level to represent the tropospheric circulation for consistency with previous studies.
-Regressions were done without detrending either the variables nor the indices, but computing the regressions with detrended values doesn't change the results (not shown).
+Regressions were done without detrending either the variables nor the indices, but computing the regressions with detrended values doesn't change the results considerably (not shown).
 
@@
   .[, season := forcats::fct_relevel(season, "Annual")] 
+```
+
+```{r pp-enso-patterns}
+
+pp_patterns <- enso[index == "oni"] %>% 
+  .[pp, on = "time"] %>% 
+  .[, precip_a := Anomaly(precip, na.rm = TRUE), by = .(lon, lat, month(time))] %>% 
+  .[, index := NULL] %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>% 
+  na.omit() %>% 
+  .[, FitLm(precip_a, value), by = .(lon, lat, season(time))] %>% 
+  rm_intercept() %>% 
+  add_continent() %>% 
+  .[continent %in% names(continents)[-1]]
+
+# pp_patterns %>% 
+#   .[continent == "america"] %>% 
+#   ggplot(aes(lon,lat)) + 
+#   geom_contour_fill(aes(z = estimate)) +
+#   facet_wrap(season~continent, scales = "free") +
+#   scale_fill_divergent()
+
+pp_enso_index <- pp %>% 
+  .[, precip_a := Anomaly(precip, na.rm = TRUE), by = .(lon, lat, month(time))] %>% 
+  .[, lapply(.SD, mean, na.rm = TRUE), by = .(lon, lat, time = seasonally(time))] %>%
+  add_continent() %>% 
+  .[continent %in% names(continents)[-1]] %>% 
+  na.omit() %>% 
+  .[, continent := NULL] %>% 
+  .[, season := season(time)] %>% 
+  .[pp_patterns, on = c("lon", "lat", "season")] %>% 
+  .[, mean(estimate*precip_a), by = .(time, continent)]
+
+pp_enso_index %>% 
+  dcast(time ~ continent) %>% 
+  ggplot(aes(oceania, america)) +
+  geom_point() +
+  facet_wrap(~season(time), scales = "free")
+
+# pp_enso_index %>% 
+#   dcast(time ~ continent) %>% 
+#   .[, cor(oceania, america), by = season(time)]
 ```
@@
 The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s [@fogt2006; @clem2013].
-Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the magnitude and sign these impacts are most likely period-dependent and represent the average signal between 1979 and 2018.
+Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the magnitude and sign of these impacts are most likely period-dependent and represent the average signal between 1979 and 2018.
 The decadal variations of the A\nobreakdash-SAM should be focus of future studies.

### 2021-05-13 analysis/paper/asymsam.Rmd
Agrega cómputo de los índices ENSO precipitación para responder a review

@@
 ```{r pp-enso-patterns}
-
+# This computation is not part of the paper, but was used as an answer to a reviewr. 
 pp_patterns <- enso[index == "oni"] %>% 
@@
 
-pp_enso_index %>% 
+cors_enso_pp <- pp_enso_index %>%
   dcast(time ~ continent) %>%
-  ggplot(aes(oceania, america)) +
-  geom_point() +
-  facet_wrap(~season(time), scales = "free")
+  .[, cor(oceania, america), by = season(time)] %>% 
+  .[, label := paste0(season, " (cor=", signif(V1, 2), ")")]
 
@@
 #   dcast(time ~ continent) %>% 
-#   .[, cor(oceania, america), by = season(time)]
+#   .[, season := season(time)] %>% 
+#   ggplot(aes(oceania, america)) +
+#   geom_point() +
+#   geom_smooth(method = "lm") +
+#   scale_x_continuous("Oceania index") +
+#   scale_y_continuous("America Index") +
+#   facet_wrap(~season, scales = "free", labeller = labeller(season= setNames(cors_enso_pp$label,
+#                                                                             cors_enso_pp$season)))
+
 ```

### 2021-05-13 analysis/paper/asymsam.Rmd
Aclara "positive values" en la regersión

@@
 Panels b and c in Figure\ \@ref(fig:global-pp) shows that the A\nobreakdash-SAM signal is only in the tropics and mid-latitudes, while the S\nobreakdash-SAM signal is strong in the higher latitudes. 
-In particular S\nobreakdash-SAM is associated with the increased precipitation over Antarctica and decrease precipitation around the Southern Ocean.
+In particular positive values of S\nobreakdash-SAM is associated with the increased precipitation over Antarctica and decrease precipitation around the Southern Ocean.
 

### 2021-05-13 analysis/paper/asymsam.Rmd
Cambios menores

@@
 
-In the Section \@ref(methods) we describe the methods.
+In Section \@ref(methods) we describe the methods.
 In Section \@ref(temporal) we describe the temporal variability and vertical coherence of the indices.
@@
 In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \@ref(fig:2d-regr)a.1) which is not centred in the South Pole.
-On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric, although still is not perfectly centred in the South Pole. 
+On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \@ref(fig:2d-regr)c.1) is more symmetric, although still not perfectly centred in the South Pole. 
 Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Drake Passage in the Western Hemisphere and Davis Sea in the Eastern Hemisphere
@@
 
-## Conclusions
+# Conclusions
 

### 2021-05-13 analysis/paper/asymsam.tex
render

@@
 
-In the Section \ref{methods} we describe the methods.
+In Section \ref{methods} we describe the methods.
 In Section \ref{temporal} we describe the temporal variability and vertical coherence of the indices.
@@
 Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year.
-The outlier being DJF, which shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure @ref\{fig:A3\}), which is consistent with \citet{fogt2020}.
-Therefore, even though we do our analysis for all months, we would expect for it to represent the other trimesters better.
+The DJF trimester shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure @ref\{fig:A3\}), which is consistent with \citet{fogt2020}.
+Therefore, we would expect that, even though the analysis is performed including all months, it represents more accurately the rest of the seasons.
 
@@
 To study the question of whether the SAM is becoming more or less asymmetric, we show trends for the explained variance of each index for each trimester in Figure \ref{fig:r-squared-trend}.
-In the troposphere the only significant trend is in DJF, in which the A\nobreakdash-SAM has a positive trend of about 2\% per decade, suggesting, that the DJF SAM has become more asymmetric in the period from 1979 to 2018,
-However due to the atypical nature of the asymmetric component of the SAM during DJF (Section~\ref(definition-of-indices)) this should be taken only as preliminary evidence.
+In the troposphere the only significant trend is in DJF, in which the A\nobreakdash-SAM has a positive trend of about 2\% per decade, suggesting that the DJF SAM has become more asymmetric in the period from 1979 to 2018.
+\citet{fogt2012} observed a change from a more asymmetrical SAM before 1980 to a more symmetrical SAM after 1980, but our study period (1979 -- 2018) prevents us from detecting that change.
+However due to the atypical nature of the asymmetric component of the SAM during DJF (Section~\ref{definition-of-indices}) this should be taken only as preliminary evidence.
 The other significant trends is in the stratosphere during SON, where there is a positive trend in the variance explained by the S\nobreakdash-SAM of about 4\% per decade.
@@
 In the stratosphere, the spatial pattern associated with the SAM is clearly dominated by a zonally symmetric, monopolar structure (Figure \ref{fig:2d-regr}a.1) which is not centred in the South Pole.
-On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \ref{fig:2d-regr}c.1) is more symmetric, although still is not perfectly centred in the South Pole.
+On the other hand, the monopole associated with S\nobreakdash-SAM (Figure \ref{fig:2d-regr}c.1) is more symmetric, although still not perfectly centred in the South Pole.
 Furthermore, the regression pattern of A\nobreakdash-SAM is characterized by a wave-1 structure with centres over the Drake Passage in the Western Hemisphere and Davis Sea in the Eastern Hemisphere
@@
 The regression patterns associated with the A\nobreakdash-SAM and S\nobreakdash-SAM indices successfully disentangle both structures.
-(Note that, in light of previous discussion about the atypical nature of DJF, this year-round average effect will likely not represent DJF perfectly).
+Note that, in light of previous discussion about the atypical nature of DJF, this year-round average effect will likely not represent DJF perfectly.
 While the A\nobreakdash-SAM index gives rise to a cleaner zonal wave (Figure \ref{fig:2d-regr}b.2), the S\nobreakdash-SAM index is associated with an annular structure, with only vestigial zonal asymmetries (Figure \ref{fig:2d-regr}c.2) in the shape of a wave-3 which is the inverse of the A\nobreakdash-SAM wave-3.
@@
 Panels b and c in Figure~\ref{fig:global-pp} shows that the A\nobreakdash-SAM signal is only in the tropics and mid-latitudes, while the S\nobreakdash-SAM signal is strong in the higher latitudes.
-In particular S\nobreakdash-SAM is associated with the increased precipitation over Antarctica and decrease precipitation around the Southern Ocean.
+In particular positive values of S\nobreakdash-SAM is associated with the increased precipitation over Antarctica and decrease precipitation around the Southern Ocean.
 
@@
 \hypertarget{conclusions}{%
-\subsection{Conclusions}\label{conclusions}}
+\section{Conclusions}\label{conclusions}}
 
@@
 The correlation between ENSO and SAM is similarly non-stationary, also changing sign before the 1980s \citep{fogt2006, clem2013}.
-Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the magnitude and sign these impacts are most likely period-dependent and represent the average signal between 1979 and 2018.
+Seeing as both the ENSO\nobreakdash-SAM relationship and most of the precipitation impacts in South America are captured by the A\nobreakdash-SAM, the magnitude and sign of these impacts are most likely period-dependent and represent the average signal between 1979 and 2018.
 The decadal variations of the A\nobreakdash-SAM should be focus of future studies.

### 2021-05-13 analysis/paper/asymsam.Rmd
Agrega recuadros a las leyendas

@@
 
-guide_colorsteps_bottom <- function(width = 25, height = 0.5, ...) {
+guide_colorsteps_bottom <- function(width = 25, height = 0.5, 
+                                    frame.colour = "black", ...) {
   guide_colorsteps(title.position = "top", title.hjust = 0.5,
                    barheight = height,
-                   barwidth = width, ...)
+                   barwidth = width, frame.colour = frame.colour, ...)
 }
@@
 To the right, probability density estimate of each index.
-Series are standarised by the standard deviation of the SAM at each level.", fig.width=6, fig.height=3, fig.env="figure*"}
+Series are standardised by the standard deviation of the SAM at each level.", fig.width=6, fig.height=3, fig.env="figure*"}
 indices %>% 
@@
   theme(strip.text = element_blank()) +
-  plot_layout(widths = c(4, 1)) 
+  plot_layout(widths = c(4, 1))  +
+  panel_background
 ```
@@
                                 "-1" = "A-SAM leads\nS-SAM by 1 month")) +
-  coord_flip()
+  coord_flip() +
+  panel_background
 ```
@@
   scale_fill_divergent_discretised("Cross-correlation",
-                                   guide = guide_colorstrip_bottom()) +
+                                   guide = guide_colorsteps_bottom()) +
   scale_x_level(minor_breaks = logbreaks, position = "top") +
@@
   theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside") +
-  tag_facets(position = "br")
+  tag_facets(position = "br") +
+  panel_background
 ```
@@
   facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
-  tag_facets("cr") 
+  tag_facets("cr") +
+  panel_background
 ```
@@
   coord_flip() +
-  facet_wrap(~season)
+  facet_wrap(~season) +
+  panel_background
 ```
@@
   scale_y_latitude(limits = c(NA, -20), labels = NULL) +
-  scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(), 
+  scale_fill_divergent(NULL, guide = guide_colorsteps_bottom(),
                        limits = c(-45, 45),
@@
   facet_grid(lev~term, labeller = labeller(term = lab_sam, lev = lev.lab)) +
-  tag_facets("cr", position = "tr") 
+  tag_facets("cr", position = "tr") +
+  panel_background
 ```
@@
                       breaks = AnchorBreaks(0, 10, 0),
-                      size = 0.3,
-                      global.breaks = FALSE) +
+                      size = 0.3) +
   # geom_contour2(aes(z = p.val), breaks = 0.01) +
-  scale_fill_divergent_discretised(NULL, guide = guide_colorstrip_bottom(13)) +
+  scale_fill_divergent_discretised(NULL, 
+                                   guide = guide_colorsteps_bottom(13, even.steps = FALSE)) +
   # oob = scales::squish,
@@
 
-```{r pp-enso-patterns}
+```{r pp-enso-patterns, eval=FALSE, include=FALSE}
 # This computation is not part of the paper, but was used as an answer to a reviewr. 
@@
 
-# pp_enso_index %>% 
-#   dcast(time ~ continent) %>% 
-#   .[, season := season(time)] %>% 
-#   ggplot(aes(oceania, america)) +
-#   geom_point() +
-#   geom_smooth(method = "lm") +
-#   scale_x_continuous("Oceania index") +
-#   scale_y_continuous("America Index") +
-#   facet_wrap(~season, scales = "free", labeller = labeller(season= setNames(cors_enso_pp$label,
-#                                                                             cors_enso_pp$season)))
+pp_enso_index %>%
+  dcast(time ~ continent) %>%
+  .[, season := season(time)] %>%
+  ggplot(aes(oceania, america)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  scale_x_continuous("Oceania index") +
+  scale_y_continuous("America Index") +
+  facet_wrap(~season, scales = "free", labeller = labeller(season= setNames(cors_enso_pp$label,
+                                                                            cors_enso_pp$season)))
 
@@
                  palette = discrete_div_pp,
-                 guide = guide_colorsteps_bottom(show.limits = TRUE)) +
+                 guide = guide_colorsteps_bottom(show.limits = TRUE, even.steps = FALSE)) +
   scale_linetype(guide = "none") +
@@
 
-  data %>%
+  gdata <- data %>%
     .[term != "mei"] %>% 
@@
     .[continent == this_continent] %>%
-    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, continent, term)] %>%
+    .[, p.value := Pvaluate(estimate, std.error, df, "fdr"), by = .(season, continent, term)]
+
+  breaks_pp <- sort(c(seq(-0.8, 0.8, by = 0.1), range(gdata$estimate)))
+  breaks_pp <- breaks_pp[breaks_pp != 0]
+
+  gdata %>%   
   ggplot(aes(lon, lat)) +
-    geom_contour_fill(aes(z = estimate), na.fill = 0, breaks = AnchorBreaks(0, 0.1, 0)) +
+    geom_contour_fill(aes(z = estimate, fill = ..level..), na.fill = 0, 
+                      breaks = breaks_pp) +
     stat_subset(aes(subset = is.na(estimate)), geom = "tile", color = "gray50") +
@@
     scale_y_latitude(breaks = NULL) +
-    scale_fill_divergent("Precipitation", 
-                         limits = c(-.8, .8),
-                         oob = scales::squish,
-                         low = "#8E521C",
-                         high = "#00665E",
-                         # label = scales::percent_format(),
-                         guide = guide_colorstrip_bottom(), 
-                         breaks = AnchorBreaks(0, 0.1, 0)) +
+     discrete_scale("fill", "name", name = "Precipitation",
+                 palette = discrete_div_pp,
+                 guide = guide_colorsteps_bottom(show.limits = TRUE, even.steps = FALSE)) +
+    # scale_fill_divergent("Precipitation", 
+    #                      limits = c(-.8, .8),
+    #                      oob = scales::squish,
+    #                      low = "#8E521C",
+    #                      high = "#00665E",
+    #                      # label = scales::percent_format(),
+    #                      guide = guide_colorsteps_bottom(), 
+    #                      breaks = AnchorBreaks(0, 0.1, 0)) +
 
@@
     scale_y_latitude(limits = c(NA, -20), labels = NULL) +
-    scale_fill_divergent(NULL, guide = guide_colorstrip_bottom(),
+    scale_fill_divergent(NULL, guide = guide_colorsteps_bottom(),
                          limits = limits,
@@
 
-```{r A5, fig.cap = "Regression coefficients of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.",  apendix = TRUE, fig.width=6, fig.height=4.5}
+```{r A5, fig.cap = "Regression coefficients of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standardised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.",  apendix = TRUE, fig.width=6, fig.height=4.5}
 set.seed(42)
@@
     scale_fill_divergent(NULL,
-                         guide = guide_colorstrip_bottom(),
+                         guide = guide_colorsteps_bottom(),
                          limits = limits,
@@
 
-```{r A6, fig.cap = "Regression of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
+```{r A6, fig.cap = "Regression of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standardised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
 Spatial correlation between both fields is 0.86 for the 50~hPa fields and 0.76 for the 700~hPa fields.", fig.width=6, fig.height=5}
@@
   # .[lev == 700] %>% 
+  .[, FitLm(hgt_a, EOF), by = .(lon, lat, period, lev)] %>%
+  rm_intercept() %>% 
+  # .[season == "DJF", estimate := -estimate] %>% 
+  .[, estimate_a := Anomaly(estimate), by = .(lat, period, lev)]
+
+cor <- patterns %>% 
+  dcast(lon + lat + lev ~ period, value.var = "estimate_a") %>% 
+  .[, cor(`1979-1998`, `1999-2018`), by = lev]
+
+patterns %>% 
+  .[lev == 50 & period == "1999-2018", estimate_a := -estimate_a] %>% 
+  periodic(lon = c(0, 360)) %>%
+  ggplot(aes(lon, lat)) +
+  geom_contour_fill(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
+  geom_contour_tanaka2(aes(z = estimate_a), breaks = AnchorBreaks(0, 5, 0)) +
+  geom_qmap() +
+  scale_x_longitude(labels = NULL) +
+  scale_y_latitude(limits = c(NA, -20), labels = NULL) +
+  scale_fill_divergent(NULL, 
+                       guide = guide_colorsteps_bottom(),
+                       breaks = AnchorBreaks(0, 5, exclude = 0)) +
+  coord_polar() +
+  geom_coords() +
+  facet_grid(lev~period, labeller = labeller(lev = lev.lab)) +
+  axis_labs_smol +
+  no_grid 
+```
+
+```{r A6b, eval=FALSE, fig.cap="Regression of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standardised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018 for the SON trimester only.
+Spatial correlation between both fields is 0.86 for the 50~hPa fields and 0.76 for the 700~hPa fields.", fig.height=5, fig.width=6, include=FALSE}
+# Code used to answer a reviewer comment
+set.seed(42)
+SAM_prepost <- hgt[lat <= -20] %>% 
+  .[season(time) == "SON"] %>% 
+  .[lev %in% c(700, 50)] %>%
+  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
+  .[, year := year(time)] %>% 
+  .[, period := factor(ifelse(year(time) <= 1998, "pre", "post"),
+                       levels = c("pre", "post"), ordered = TRUE,
+                       labels = c("1979-1998", "1999-2018"))] %>%
+  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
+    by = .(period, lev)]
+
+sam_sep <- SAM_prepost[, eof[[1]]$right, by = .(lev, period)] %>% 
+  setnames("hgt", "full") %>% 
+  .[, c("sym", "asym") := list(mean(full), Anomaly(full)), by = .(lat, period, lev)]
+
+patterns <- SAM_prepost[, eof[[1]]$left, by = .(period, lev)] %>% 
+  copy() %>% 
+  setnames("hgt", "EOF") %>% 
+  .[, EOF := EOF/sd(EOF), by = .(period, lev)] %>% 
+  .[period == "1979-1998" & lev == 50, EOF := -EOF] %>% 
+  rm_singleton() %>% 
+  hgt[., on = .NATURAL] %>% 
+  # .[lev == 700] %>% 
+  .[season(time) == "SON"] %>% 
   .[, FitLm(hgt_a, EOF), by = .(lon, lat, period, lev)] %>%
@@
   scale_fill_divergent(NULL,
-                       guide = guide_colorstrip_bottom(20),
+                       guide = guide_colorsteps_bottom(20),
                        breaks = AnchorBreaks(0, 0.05),
@@
                        oob = scales::squish,
-                       guide = guide_colorstrip_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
+                       guide = guide_colorsteps_bottom(), breaks = AnchorBreaks(0, 0.1, 0)) +
   coord_polar() +

### 2021-05-13 analysis/paper/asymsam.tex
Renderiza

@@
 To the right, probability density estimate of each index.
-Series are standarised by the standard deviation of the SAM at each level.}\label{fig:asymsam-timeseries}
+Series are standardised by the standard deviation of the SAM at each level.}\label{fig:asymsam-timeseries}
 \end{figure*}
@@
 \begin{figure}
-\includegraphics{A5-1} \caption{Regression coefficients of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
+\includegraphics{A5-1} \caption{Regression coefficients of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standardised timeseries of the leading EOF computed for each season independently for the 1979 -- 2018 period.}\label{fig:A5}
 \end{figure}
@@
 \begin{figure}
-\includegraphics{A6-1} \caption{Regression of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
+\includegraphics{A6-1} \caption{Regression of 50~hPa and 700~hPa geopotential height zonal anomalies (meters) onto the standardised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018.
 Spatial correlation between both fields is 0.86 for the 50~hPa fields and 0.76 for the 700~hPa fields.}\label{fig:A6}

### 2021-05-14 analysis/paper/asymsam.Rmd
Detalles de figuras

@@
                     data = ~.x[item2 > item1]) +
-  geom_polygon(data = tri, fill = "white") +
+  geom_polygon(data = tri, fill = "#fbfbfb") +
   scale_fill_divergent_discretised("Cross-correlation",
@@
   theme(panel.spacing.x = unit(2, "lines"), strip.placement = "outside") +
-  tag_facets(position = "br") +
-  panel_background
+  tag_facets(position = "br")
 ```
@@
   scale_y_latitude(limits = c(NA, -20), labels = NULL) +
-  scale_fill_divergent(NULL, guide = guide_colorsteps_bottom(),
+  scale_fill_divergent(NULL, guide = guide_colorsteps_bottom(even.steps = FALSE),
                        limits = c(-45, 45),
@@
   tag_facets("cr", position = "tr") +
-  panel_background
+  panel_background +
+  theme(panel.spacing = grid::unit(0.5, "lines"))
 ```
@@
   facet_grid(index_level~., labeller = labeller(index_level = lev.lab)) +
-  tag_facets(position = "bl") 
+  tag_facets(position = "bl") +
+  theme(panel.spacing = grid::unit(1, "lines"))
 ```
@@
       col.names = c("", lab_sam), booktabs = TRUE) %>%
-  kable_classic(full_width = F) %>%
+  kable_classic_2(full_width = F) %>%
   column_spec(2, bold = rep(table2[, 2] < 0.05, each = 2)) %>%
@@
 anchor_limits <- function(anchor = 0, binwidth = NULL, exclude = NULL,  bins = 10,
-                          range = c(NA, NA)) {
+                          range = c(NA, NA), sym = TRUE) {
   force(range)
@@
   function(x, binwidth2 = NULL) {
+    if (sym) {
+     D <- max(abs(x[1] - anchor), abs(x[2] - anchor))
+     x <- c(anchor - D, anchor + D)
+    }
+
     if (!is.null(binwidth)) binwidth2 <- binwidth
@@
       m <- min(b)
-      b <- c(b[b >= range[1]], m)
+      b <- c(b[b >= (range[1] + 1e-9)], -Inf)  # DAMN YOU FLOATING POINT ERRORS!!! 
     }
@@
       m <- max(b)
-      b <- c(b[b <= range[2]], m)
+      b <- c(b[b <= (range[2] - 1e-9)], Inf)
     }
@@
   geom_contour_fill(aes(z = estimate, fill = ..level..), 
-                    breaks = anchor_limits(0, 0.1, 0, range = c(-.6, .6))) +
+                    breaks = anchor_limits(0, 0.1, 0, range = c(-.6, 0.6))) +
   geom_contour2(aes(z = p.value), breaks = c(0, 0.05), color = "black", size = 0.3) +
@@
   scale_y_latitude(labels = NULL, limits = c(NA, 0)) +
-  discrete_scale("fill", "name", name = "2-meter air temperature",
-                 palette = discrete_div,
-                 guide = guide_colorsteps_bottom(show.limits = TRUE)) +
+  scale_fill_divergent_discretised("2-meter air temperature", 
+                                    guide = guide_colorsteps_bottom(even.steps = FALSE)) +
   coord_polar() +
@@
 
-(ref:global-pp-cap) Regression of annual and seasonal mean precipitation anomalies (mm per day, shading)  with a) SAM, b) A\\nobreakdash-SAM and c) S\\nobreakdash-SAM for the 1979 -- 2018 period.
+(ref:global-pp-cap) Regression of annual and seasonal mean precipitation anomalies (mm per day, shading)  with a) SAM, b) A\nobreakdash-SAM and c) S\nobreakdash-SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
   geom_contour_fill(aes(z = estimate, fill = ..level..), 
-                    breaks = breaks_pp) +
+                    breaks = anchor_limits(0, 0.05, 0, range = c(-0.25, 0.25))) +
   stat_subset(aes(subset = is.na(estimate)), geom = "tile", color = "gray50") +
@@
   scale_y_latitude(breaks = NULL, limits = c(-90, NA)) +
-  discrete_scale("fill", "name", name = "Precipitation",
-                 palette = discrete_div_pp,
-                 guide = guide_colorsteps_bottom(show.limits = TRUE, even.steps = FALSE)) +
+  # discrete_scale("fill", "name", name = "Precipitation",
+  #                palette = discrete_div_pp,
+  #                guide = guide_colorsteps_bottom(show.limits = TRUE, even.steps = FALSE)) +
+  scale_fill_divergent_discretised("Precipitation",
+                         low = "#8E521C", high = "#00665E",
+                         guide = guide_colorsteps_bottom(even.steps = FALSE)) +
   scale_linetype(guide = "none") +
@@
 
-  breaks_pp <- sort(c(seq(-0.8, 0.8, by = 0.1), range(gdata$estimate)))
-  breaks_pp <- breaks_pp[breaks_pp != 0]
-
   gdata %>%   
@@
     geom_contour_fill(aes(z = estimate, fill = ..level..), na.fill = 0, 
-                      breaks = breaks_pp) +
+                      breaks = anchor_limits(0, 0.1, 0, range = c(-.8, 0.8))) +
     stat_subset(aes(subset = is.na(estimate)), geom = "tile", color = "gray50") +
@@
     geom_qmap(~.x[lat <= 10], keep = 0.015, weighting = 0.9) +
-    scale_x_longitude(breaks = NULL) +
-    scale_y_latitude(breaks = NULL) +
-     discrete_scale("fill", "name", name = "Precipitation",
-                 palette = discrete_div_pp,
-                 guide = guide_colorsteps_bottom(show.limits = TRUE, even.steps = FALSE)) +
-    # scale_fill_divergent("Precipitation", 
+    scale_x_longitude(ticks = 15, minor_breaks = NULL) +
+    scale_y_latitude(ticks = 10,  minor_breaks = NULL) +
+     # discrete_scale("fill", "name", name = "Precipitation",
+     #             palette = discrete_div_pp,
+     #             guide = guide_colorsteps_bottom(show.limits = TRUE, even.steps = FALSE)) +
+    scale_fill_divergent_discretised("Precipitation",
                          # limits = c(-.8, .8),
                          # oob = scales::squish,
-    #                      low = "#8E521C",
-    #                      high = "#00665E",
-    #                      # label = scales::percent_format(),
-    #                      guide = guide_colorsteps_bottom(), 
-    #                      breaks = AnchorBreaks(0, 0.1, 0)) +
+                         low = "#8E521C",
+                         high = "#00665E",
+                         # label = scales::percent_format(),
+                         guide = guide_colorsteps_bottom(even.steps = FALSE)) +
 
@@
     tag_facets("cr", position = "bl")  +
-    no_grid +
     axis_labs_smol 
@@
 <!-- \bibliography{AsymSAM,packages} -->
+
+\begin{acknowledgements}
+We are grateful to the reviewers for their constructive comments that improved the manuscript.
+\end{acknowledgements}
 

### 2021-05-14 analysis/paper/asymsam.tex
Renderiza todo

@@
 \begin{figure}
-\includegraphics{global-pp-1} \caption{Regression of annual and seasonal mean precipitation anomalies (mm per day, shading) with a) SAM, b) A\textbackslash nobreakdash-SAM and c) S\textbackslash nobreakdash-SAM for the 1979 -- 2018 period.
+\includegraphics{global-pp-1} \caption{Regression of annual and seasonal mean precipitation anomalies (mm per day, shading) with a) SAM, b) A\nobreakdash-SAM and c) S\nobreakdash-SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
@@
 
+\begin{acknowledgements}
+We are grateful to the reviewers for their constructive comments that improved the manuscript.
+\end{acknowledgements}
+
 \newpage

### 2021-05-20 analysis/paper/asymsam.Rmd
Cosas de formato

@@
 @fogt2012 observed a change from a more asymmetrical SAM before 1980 to a more symmetrical SAM after 1980, but our study period (1979 -- 2018) prevents us from detecting that change.
-However due to the atypical nature of the asymmetric component of the SAM during DJF (Section\ \@ref(definition-of-indices)) this should be taken only as preliminary evidence. 
+However due to the atypical nature of the asymmetric component of the SAM during DJF (Section \@ref(definition-of-indices)) this should be taken only as preliminary evidence.
 The other significant trends is in the stratosphere during SON, where there is a positive trend in the variance explained by the S\nobreakdash-SAM of about 4% per decade.
@@
 
-The wave-3 pattern from Figure\ \@ref(fig:2d-regr)b.2 is very similar to the PSA Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
+The wave-3 pattern from Figure \@ref(fig:2d-regr)b.2 is very similar to the PSA Pattern [@mo1987; @kidson1988] which is a teleconnection pattern associated with the ENSO [@karoly1989].
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is shown in Table\ \@ref{tab:enso-cor-table} for each SAM index and for each trimester. 
+The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is shown in Table \@ref{tab:enso-cor-table} for each SAM index and for each trimester.
 There is a significant correlation between SAM and ENSO.
@@
 
-The same regressions using temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid are shown in Figure\ \@ref{fig:A10}.
+The same regressions using temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid are shown in Figure \@ref{fig:A10}.
 Where data is available, the general patterns are similar, although with reduced magnitude, perhaps due to the fact that sea surface temperatures vary less.
@@
 
-Figure\ \@ref(fig:global-pp) shows regression of the SAM indices with precipitation for the Southern Hemisphere. 
-The precipitation signal associated with SAM shows generally decreased precipitation at around 45ºS, somewhat increase precipitation around 30ªS (Fig.\ \@ref(fig:global-pp)a) and increased precipitation over Antarctica, a pattern known from other studies [e.g.
+Figure \@ref(fig:global-pp) shows regression of the SAM indices with precipitation for the Southern Hemisphere.
+The precipitation signal associated with SAM shows generally decreased precipitation at around 45ºS, somewhat increase precipitation around 30ªS (Fig. \@ref(fig:global-pp)a) and increased precipitation over Antarctica, a pattern known from other studies [e.g.
 @hendon2014].
 This pattern is mostly unchanged between seasons albeit from varying in intensity.
-Panels b and c in Figure\ \@ref(fig:global-pp) shows that the A\nobreakdash-SAM signal is only in the tropics and mid-latitudes, while the S\nobreakdash-SAM signal is strong in the higher latitudes. 
+Panels b and c in Figure \@ref(fig:global-pp) shows that the A\nobreakdash-SAM signal is only in the tropics and mid-latitudes, while the S\nobreakdash-SAM signal is strong in the higher latitudes.
 In particular positive values of S\nobreakdash-SAM is associated with the increased precipitation over Antarctica and decrease precipitation around the Southern Ocean.
@@
 The periodicity of about two years we found in the stratospheric S\nobreakdash-SAM is consistent with the possible link between the SAM and the Quasi Biennial Oscillation as proposed by [@vasconcellos2020].
-There is evidence of influence between the QBO and the Northern Annular Mode [e.g.
-@holton1980; @watson2014; @zhang2020], so it is not unlikely that the SAM would be similarly affected.
+There is evidence of influence between the QBO and the Northern Annular Mode [e.g. @holton1980; @watson2014; @zhang2020], so it is not unlikely that the SAM would be similarly affected.
 Establishing this link would require further research.
@@
 CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at <https://psl.noaa.gov/data/gridded/data.cmap.html>.
-The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>
+The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.
 The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at the public repository associated with this paper, at <https://github.com/eliocamp/asymsam/tree/master/analysis/data/derived_data>.
@@
 
+```{=tex}
 \begin{acknowledgements}
@@
 \end{acknowledgements}
-
+```
 \newpage

### 2021-05-20 analysis/paper/asymsam.Rmd
Agrega énfasis al "so what"

@@
 Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere [@baldwin2009].
+
+In summary, previous research strongly suggests that the zonally asymmetric component of the SAM can potentially be very different from the zonally symmetric component. 
+It might have different sources of variability, impacts and long-term response to radiative forcing. 
+A single SAM index that mixes the zonally symmetric and zonally asymmetric variability is only able to capture the combined effect of these two potentially distinct modes.
 
@@
 
-We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies [e.g. @fogt2020 and references therein] for low levels.
+We observe positive trends of the SAM in Summer and Autumn, as was documented by previous studies [e.g. @fogt2020 and references therein] for low levels.
 We show that these trends maximise at 100 hPa, and are explained by the zonally symmetric component.
@@
 The decadal variations of the A\nobreakdash-SAM should be focus of future studies. 
-
-By separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
 This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modelled as highly zonally symmetric, while the impact of increased concentration of greenhouse gases has also a zonally asymmetric component [@arblaster2006].
+
+By separating the zonally symmetric and zonally symmetric SAM patterns, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
+Some of these signals are only evident when the two SAM components are separated. 
+For those that are evident also using the traditional SAM index, the separation is still worthwhile, since it allows for a clearer view of the sources and processes involved.
 

### 2021-05-20 analysis/paper/asymsam.tex
Render

@@
 Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere \citep{baldwin2009}.
+
+In summary, previous research strongly suggests that the zonally asymmetric component of the SAM can potentially be very different from the zonally symmetric component.
+It might have different sources of variability, impacts and long-term response to radiative forcing.
+A single SAM index that mixes the zonally symmetric and zonally asymmetric variability is only able to capture the combined effect of these two potentially distinct modes.
 
@@
 
-We observe a positive trend of SAM in Summer and Autumn, as was documented by previous studies \citep[e.g.][ and references therein]{fogt2020} for low levels.
+We observe positive trends of the SAM in Summer and Autumn, as was documented by previous studies \citep[e.g.][ and references therein]{fogt2020} for low levels.
 We show that these trends maximise at 100~hPa, and are explained by the zonally symmetric component.
@@
 The decadal variations of the A\nobreakdash-SAM should be focus of future studies.
+This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modelled as highly zonally symmetric, while the impact of increased concentration of greenhouse gases has also a zonally asymmetric component \citep{arblaster2006}.
 
-By separating the zonally symmetric and zonally symmetric SAM signals, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
-This is particularly important in the context of a changing climate, as the impact on the SAM of ozone recovery is modelled as highly zonally symmetric, while the impact of increased concentration of greenhouse gases has also a zonally asymmetric component \citep{arblaster2006}.
+By separating the zonally symmetric and zonally symmetric SAM patterns, we show that the asymmetric component of the SAM has its unique variability, trends and impacts.
+Some of these signals are only evident when the two SAM components are separated.
+For those that are evident also using the traditional SAM index, the separation is still worthwhile, since it allows for a clearer view of the sources and processes involved.
 
@@
 CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at \url{https://psl.noaa.gov/data/gridded/data.cmap.html}.
-The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}
+The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}.
 The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at the public repository associated with this paper, at \url{https://github.com/eliocamp/asymsam/tree/master/analysis/data/derived_data}.

### 2021-05-24 analysis/paper/asymsam.Rmd
Cambia el sitio web de los datos.

@@
 The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.
-The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at the public repository associated with this paper, at <https://github.com/eliocamp/asymsam/tree/master/analysis/data/derived_data>.
+The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at <http://www.cima.fcen.uba.ar/~elio.campitelli/asymsam/>.
 

### 2021-05-26 analysis/paper/asymsam.Rmd
Cita a thompson
Closes #15

@@
 Even though these indices are based on zonal averages, their associated geopotential height spatial anomalies contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
-The zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM [@fan2007; @silvestri2009; @fogt2012; @rosso2018].
+The zonal asymmetries have not been widely studied, but previous work suggests that they strongly modulate the regional impacts of the SAM [@fan2007; @silvestri2009; @fogt2012; @rosso2018].
 The fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere [@jones2009].
@@
 These anomalies are associated only with the S\nobreakdash-SAM.
+Furthermore, the typical longitudinal variations of temperature anomalies along the Antarctic Peninsula are not evident in the regressions with the SAM, in agreement with previous works [e.g.
+marshall2016].
 On the other hand, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.

### 2021-05-26 analysis/paper/asymsam.Rmd
Arregla escala de colores en una figura

@@
 The regional impacts of the SAM in temperature and precipitation are strongly affected by its asymmetric component.
+We show that the asymmetric component of the SAM has its own unique variability, trends and impacts, some of these signals are only evident when the two SAM components are separated.
 
@@
   ggplot(aes(lon, lat)) +
-  geom_contour_fill(aes(z = estimate), breaks = AnchorBreaks(0, exclude =  0),
+  geom_contour_fill(aes(z = estimate), 
+                    breaks = AnchorBreaks(0, exclude =  0),
                     global.breaks = FALSE) +
@@
   scale_y_latitude(limits = c(NA, -20), labels = NULL) +
-  scale_fill_divergent(NULL, guide = guide_colorsteps_bottom(even.steps = FALSE),
+  scale_fill_divergent(NULL, 
+                       guide = guide_colorsteps_bottom(even.steps = TRUE),
                        limits = c(-45, 45),
                        oob = scales::squish,
-                       breaks = AnchorBreaks(0, 10, 0)) +
+                       breaks = AnchorBreaks(0, 10, 0)(c(-45, 45))) +
   scale_linetype(guide = "none") +
@@
 Furthermore, the typical longitudinal variations of temperature anomalies along the Antarctic Peninsula are not evident in the regressions with the SAM, in agreement with previous works [e.g.
-marshall2016].
+@marshall2016].
 On the other hand, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.
@@
 In Spring, the tropical signal of A\nobreakdash-SAM is similar to the one in Summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link. 
+Moreover, from Autumn to Spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM.
+In agreement, @marshall2016 found the same signal in association with SAM while that pattern is only evident in Autumm for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.
 

### 2021-05-26 analysis/paper/asymsam.tex
Renderiza todo

@@
 The regional impacts of the SAM in temperature and precipitation are strongly affected by its asymmetric component.
+We show that the asymmetric component of the SAM has its own unique variability, trends and impacts, some of these signals are only evident when the two SAM components are separated.
 \\
@@
 Even though these indices are based on zonal averages, their associated geopotential height spatial anomalies contain noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region.
-The zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM \citep{fan2007, silvestri2009, fogt2012, rosso2018}.
+The zonal asymmetries have not been widely studied, but previous work suggests that they strongly modulate the regional impacts of the SAM \citep{fan2007, silvestri2009, fogt2012, rosso2018}.
 The fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availability of dense observations in the Southern Hemisphere \citep{jones2009}.
@@
 These anomalies are associated only with the S\nobreakdash-SAM.
+Furthermore, the typical longitudinal variations of temperature anomalies along the Antarctic Peninsula are not evident in the regressions with the SAM, in agreement with previous works \citep[e.g.][]{marshall2016}.
 On the other hand, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.
@@
 In Spring, the tropical signal of A\nobreakdash-SAM is similar to the one in Summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link.
+Moreover, from Autumn to Spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM.
+In agreement, \citet{marshall2016} found the same signal in association with SAM while that pattern is only evident in Autumm for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.
 
@@
 The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}.
-The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at the public repository associated with this paper, at \url{https://github.com/eliocamp/asymsam/tree/master/analysis/data/derived_data}.
+The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at \url{http://www.cima.fcen.uba.ar/~elio.campitelli/asymsam/}.
 

### 2021-05-28 analysis/paper/asymsam.Rmd
Correcciones finales de LEandro

@@
 
-For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [\@xie1997], with a 2.5\\degree\\ resolution in latitude and longitude.
+For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@xie1997], with a 2.5\degree\ resolution in latitude and longitude.
 This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 
-```{r method, fig.cap = "Spatial patterns of the first EOF of 700~hPa geopotential height for 1979 -– 2018 period.
+(ref:method-cap) Spatial patterns of the first EOF of 700\ hPa geopotential height for 1979 -– 2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
-Arbitrary units; positive values in blue and negative values in red.", fig.width=6, fig.height=2.7, fig.env="figure*"}
+Arbitrary units; positive values in blue and negative values in red.
+
+```{r method, fig.cap = "(ref:method-cap", fig.width=6, fig.height=2.7, fig.env="figure*"}
 sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 
@@
 Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year.
-The DJF trimester shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref{fig:A3}), which is consistent with @fogt2020.
+The DJF trimester shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \@ref(fig:A3)), which is consistent with @fogt2020.
 Therefore, we would expect that, even though the analysis is performed including all months, it represents more accurately the rest of the seasons.
@@
   scale_y_continuous("Correlation", limits = c(0, 1)) + 
-  scale_linetype_manual(NULL, values = c(1, 5), 
+  scale_linetype_manual(name = NULL, values = c(1, 5), 
                         labels = c("0" = "Zero-lag", 
@@
 However due to the atypical nature of the asymmetric component of the SAM during DJF (Section \@ref(definition-of-indices)) this should be taken only as preliminary evidence.
-The other significant trends is in the stratosphere during SON, where there is a positive trend in the variance explained by the S\nobreakdash-SAM of about 4% per decade.
+The other significant trend is in the stratosphere during SON, where there is a positive trend in the variance explained by the S\nobreakdash-SAM of about 4% per decade.
 This change could be the result of the forcing from ozone depletion.
@@
 Indeed, @fogt2011 showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is shown in Table \@ref{tab:enso-cor-table} for each SAM index and for each trimester.
-There is a significant correlation between SAM and ENSO.
-When divided in trimesters, this correlation is important only in DJF and SON.
+The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index [ONI, @bamston1997] is shown in Table \@ref(tab:enso-cor-table) for each SAM index and for each trimester. 
+There is a significant correlation between SAM and ENSO. When divided in trimesters, this correlation is important only in DJF and SON. 
 This relationship is captured mainly by the A\nobreakdash-SAM, as this index has signifcant partial correlations with ENSO while correlations with S\nobreakdash-SAM are all smaller and non-significant.
@@
 
-(ref:regr-air-season-cap) Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with SAM, A\nobreakdash-SAM and S\nobreakdash-SAM for the 1979 -- 2018 period.
+(ref:regr-air-season-cap) Regression of seasonal mean 2-metre temperature anomalies (Kelvin) from ERA5 with SAM, A\nobreakdash-SAM and S\nobreakdash-SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Note that the colour scale cuts-off at $\pm0.6 \mathrm{K}$ so the areas of very high values in the Antarctic don't wash out detail in the mid-latitudes and tropics.
+Note that the colour scale cuts-off at $\pm0.6 \mathrm{K}$ to highlight mid-latitudes and tropics features at the expense of the higher values in Polar regions.
 
@@
 
-The same regressions using temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid are shown in Figure \@ref{fig:A10}.
+The same regressions using temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@smith2008; @vose2012], which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid are shown in Figure \@ref(fig:A10).
 Where data is available, the general patterns are similar, although with reduced magnitude, perhaps due to the fact that sea surface temperatures vary less.
@@
 
-(ref:global-pp-cap) Regression of annual and seasonal mean precipitation anomalies (mm per day, shading) with a) SAM, b) A\nobreakdash-SAM and c) S\nobreakdash-SAM for the 1979 -- 2018 period.
+(ref:global-pp-cap) Regression of monthly precipitation anomalies (mm per day, shading) with a) SAM, b) A\nobreakdash-SAM and c) S\nobreakdash-SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Note that the colour scale cuts-off at $\pm0.25 \mathrm{K}$ so the areas of very high values in the Tropics don't wash out detail in the mid and high-latitudes.
+Note that the colour scale cuts-off at $\pm0.25 \mathrm{K}$ to highlight mid- and high-latitude features at the expense of the very high values in the Tropics.
 
@@
 @hendon2014].
-This pattern is mostly unchanged between seasons albeit from varying in intensity.
+This pattern is mostly unchanged between seasons albeit varying in intensity (not shown).
 Panels b and c in Figure \@ref(fig:global-pp) shows that the A\nobreakdash-SAM signal is only in the tropics and mid-latitudes, while the S\nobreakdash-SAM signal is strong in the higher latitudes.
@@
                       global.breaks = FALSE,
-                      size = 3,
+                      size = 2.5,
                       stroke = 0.15, stroke.color = "white",
@@
 The seasonal-based regressions show statistically significant anomalies only in Spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
-In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related with positive A\nobreakdash-SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).
+In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related to positive A\nobreakdash-SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).
 
@@
 The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
-In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, and a broad area of positive anomalies in the North-East to South-West direction.
-This structure seems to be associated with the S\nobreakdash-SAM, while the Northern positive values are associated with the A\nobreakdash-SAM.
-In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the S\nobreakdash-SAM.
+In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, which are associated with the A\nobreakdash-SAM.
+In Winter regression coefficients are much weaker than during the rest of the year. 
 None of these regression coefficients are statistically significant at the 95% level.
@@
 \begin{acknowledgements}
-NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ 
+NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/.
+
+We are grateful to the reviewers for their constructive comments that improved the manuscript.
 \end{acknowledgements}
@@
 
-All data used in this paper is freely available from their respective sources.
-ERA5 data can be obtained via the Copernicus Climate Data Store ([\<https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means\>](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means){.uri}).
-CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at <https://psl.noaa.gov/data/gridded/data.cmap.html>.
-The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.
 The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at <http://www.cima.fcen.uba.ar/~elio.campitelli/asymsam/>.
 
+All data used in this paper is freely available from their respective sources:
+
+* ERA5 data can be obtained via the Copernicus Climate Data Store (<https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means\>).
+
+* CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at <https://psl.noaa.gov/data/gridded/data.cmap.html>.
+
+* The Oceanic Niño Index is available via NOAA's Climate Prediction Center: <https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt>.
+
 ## Code availability {.unnumbered}
@@
 
-```{=tex}
-\begin{acknowledgements}
-We are grateful to the reviewers for their constructive comments that improved the manuscript.
-\end{acknowledgements}
-```
 \newpage

### 2021-05-28 analysis/paper/asymsam.tex
Render

@@
 
-For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation {[}@xie1997{]}, with a 2.5\textbackslash degree\textbackslash{} resolution in latitude and longitude.
+For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation \citep{xie1997}, with a 2.5\degree~resolution in latitude and longitude.
 This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.
@@
 \begin{figure*}
-\includegraphics{method-1} \caption{Spatial patterns of the first EOF of 700~hPa geopotential height for 1979 -– 2018 period.
-(a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
-Arbitrary units; positive values in blue and negative values in red.}\label{fig:method}
+\includegraphics{method-1} \caption{(ref:method-cap}\label{fig:method}
 \end{figure*}
@@
 Therefore, the results confirm that the zonal asymmetric structure of the SAM is very similar throughout most of the year.
-The DJF trimester shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure @ref\{fig:A3\}), which is consistent with \citet{fogt2020}.
+The DJF trimester shows much lower correlations with the other seasons at both levels and the weakest zonal anomalies (Figure \ref{fig:A3}), which is consistent with \citet{fogt2020}.
 Therefore, we would expect that, even though the analysis is performed including all months, it represents more accurately the rest of the seasons.
@@
 However due to the atypical nature of the asymmetric component of the SAM during DJF (Section~\ref{definition-of-indices}) this should be taken only as preliminary evidence.
-The other significant trends is in the stratosphere during SON, where there is a positive trend in the variance explained by the S\nobreakdash-SAM of about 4\% per decade.
+The other significant trend is in the stratosphere during SON, where there is a positive trend in the variance explained by the S\nobreakdash-SAM of about 4\% per decade.
 This change could be the result of the forcing from ozone depletion.
@@
 Indeed, \citet{fogt2011} showed that there is a significant relationship between the SAM and the ENSO.
-The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is shown in Table~@ref\{tab:enso-cor-table\} for each SAM index and for each trimester.
-There is a significant correlation between SAM and ENSO.
-When divided in trimesters, this correlation is important only in DJF and SON.
+The correlation between the SAM and the ENSO as measured by the Oceanic Niño Index \citep[ONI,][]{bamston1997} is shown in Table~\ref{tab:enso-cor-table} for each SAM index and for each trimester.
+There is a significant correlation between SAM and ENSO. When divided in trimesters, this correlation is important only in DJF and SON.
 This relationship is captured mainly by the A\nobreakdash-SAM, as this index has signifcant partial correlations with ENSO while correlations with S\nobreakdash-SAM are all smaller and non-significant.
@@
 \begin{figure*}
-\includegraphics{regr-air-season-1} \caption{Regression of seasonal mean surface land air and sea temperature anomalies (Kelvin) with SAM, A\nobreakdash-SAM and S\nobreakdash-SAM for the 1979 -- 2018 period.
+\includegraphics{regr-air-season-1} \caption{Regression of seasonal mean 2-metre temperature anomalies (Kelvin) from ERA5 with SAM, A\nobreakdash-SAM and S\nobreakdash-SAM for the 1979 -- 2018 period.
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Note that the colour scale cuts-off at \(\pm0.6 \mathrm{K}\) so the areas of very high values in the Antarctic don't wash out detail in the mid-latitudes and tropics.}\label{fig:regr-air-season}
+Note that the colour scale cuts-off at \(\pm0.6 \mathrm{K}\) to highlight mid-latitudes and tropics features at the expense of the higher values in Polar regions.}\label{fig:regr-air-season}
 \end{figure*}
@@
 
-The same regressions using temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid are shown in Figure~@ref\{fig:A10\}.
+The same regressions using temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 \citep{smith2008, vose2012}, which blends land surface air temperature and sea surface (water) temperature analysis into a monthly global grid are shown in Figure~\ref{fig:A10}.
 Where data is available, the general patterns are similar, although with reduced magnitude, perhaps due to the fact that sea surface temperatures vary less.
@@
 \begin{figure}
-\includegraphics{global-pp-1} \caption{Regression of annual and seasonal mean precipitation anomalies (mm per day, shading) with a) SAM, b) A\nobreakdash-SAM and c) S\nobreakdash-SAM for the 1979 -- 2018 period.
+\includegraphics{global-pp-1} \caption{Regression of monthly precipitation anomalies (mm per day, shading) with a) SAM, b) A\nobreakdash-SAM and c) S\nobreakdash-SAM for the 1979 -- 2018 period.
 Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Note that the colour scale cuts-off at \(\pm0.25 \mathrm{K}\) so the areas of very high values in the Tropics don't wash out detail in the mid and high-latitudes.}\label{fig:global-pp}
+Note that the colour scale cuts-off at \(\pm0.25 \mathrm{K}\) to highlight mid- and high-latitude features at the expense of the very high values in the Tropics.}\label{fig:global-pp}
 \end{figure}
@@
 The precipitation signal associated with SAM shows generally decreased precipitation at around 45ºS, somewhat increase precipitation around 30ªS (Fig.~\ref{fig:global-pp}a) and increased precipitation over Antarctica, a pattern known from other studies \citep[e.g.][]{hendon2014}.
-This pattern is mostly unchanged between seasons albeit from varying in intensity.
+This pattern is mostly unchanged between seasons albeit varying in intensity (not shown).
 Panels b and c in Figure~\ref{fig:global-pp} shows that the A\nobreakdash-SAM signal is only in the tropics and mid-latitudes, while the S\nobreakdash-SAM signal is strong in the higher latitudes.
@@
 The seasonal-based regressions show statistically significant anomalies only in Spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \ref{fig:pp-regr-oceania}a.5).
-In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \ref{fig:pp-regr-oceania}c.5) while the positive precipitation anomalies related with positive A\nobreakdash-SAM affect all Eastern Australia (Figure \ref{fig:pp-regr-oceania}b.5).
+In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \ref{fig:pp-regr-oceania}c.5) while the positive precipitation anomalies related to positive A\nobreakdash-SAM affect all Eastern Australia (Figure \ref{fig:pp-regr-oceania}b.5).
 
@@
 The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
-In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, and a broad area of positive anomalies in the North-East to South-West direction.
-This structure seems to be associated with the S\nobreakdash-SAM, while the Northern positive values are associated with the A\nobreakdash-SAM.
-In Winter we see the same NE to SW aligned anomaly (although with much reduced amplitude) that is also present only in relation with the S\nobreakdash-SAM.
+In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, which are associated with the A\nobreakdash-SAM.
+In Winter regression coefficients are much weaker than during the rest of the year.
 None of these regression coefficients are statistically significant at the 95\% level.
@@
 
-All data used in this paper is freely available from their respective sources.
-ERA5 data can be obtained via the Copernicus Climate Data Store (\href{https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means}{\textless https://cds.climate.copernicus.eu/cdsapp\#!/dataset/reanalysis-era5-pressure-levels-monthly-means\textgreater{}}).
+The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at \url{http://www.cima.fcen.uba.ar/~elio.campitelli/asymsam/}.
+
+All data used in this paper is freely available from their respective sources:
+
+\begin{itemize}
+\item
+  ERA5 data can be obtained via the Copernicus Climate Data Store (\url{https://cds.climate.copernicus.eu/cdsapp/\#!/dataset/reanalysis-era5-pressure-levels-monthly-means/}).
+\item
   CMAP Precipitation data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at \url{https://psl.noaa.gov/data/gridded/data.cmap.html}.
+\item
   The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}.
-The SAM, A\nobreakdash-SAM and S\nobreakdash-SAM indices are available at \url{http://www.cima.fcen.uba.ar/~elio.campitelli/asymsam/}.
+\end{itemize}
 
@@
 A version-controlled repository of the code used to create this analysis, including the code used to download the data can be found at \url{https://github.com/eliocamp/asymsam}.
-
-\begin{acknowledgements}
-We are grateful to the reviewers for their constructive comments that improved the manuscript.
-\end{acknowledgements}
 

### 2021-05-28 analysis/paper/asymsam.Rmd
Cambia un en dash
Typo

@@
 
-(ref:method-cap) Spatial patterns of the first EOF of 700\ hPa geopotential height for 1979 -– 2018 period.
+(ref:method-cap) Spatial patterns of the first EOF of 700\ hPa geopotential height for 1979 -- 2018 period.
 (a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
@@
 Moreover, from Autumn to Spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM.
-In agreement, @marshall2016 found the same signal in association with SAM while that pattern is only evident in Autumm for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.
+In agreement, @marshall2016 found the same signal in association with SAM while that pattern is only evident in Autumn for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.
 

### 2021-05-28 analysis/paper/asymsam.Rmd
Spellcheck

@@
 
-<!-- The actual document text starts here: -->
-
-<!-- # Abstract -->
-
-<!-- The main mode of variability in the Southern Hemisphere extratropical circulation is the Southern Annular Mode (SAM); so called because of it's zonally symmetric ring-like shape.
-However, the SAM pattern actually contains noticeable deviations from zonal symmetry.
-We regress monthly geopotential height fields at each level onto the asymmetric and symmetric component of the SAM to create two new indices.
-In the troposphere, the Asymmetric SAM, is associated with a zonal wave 3 which is rotated a quarter wavelength with respect to the climatological zonal wave 3, is much stronger in the Pacific ocean, where it extends vertically to the stratosphere with an equivalent barotropic structure.
-The Symmetric SAM, on the other hand, is associated with a negative geopotential height anomalies over Antarctica surrounded by a zonally symmetric ring of positive geopotential height anomalies.
-The observed relationship between the El Niño Southern Oscillation and the SAM is explained fully by the Asymmetric SAM index.
-The positive trend of the SAM is present only in its asymmetric component.
-Despite this, the SAM is becoming more zonally asymmetric.
-The regional impacts of the SAM in temperature and precipitation are strongly affected by its asymmetric nature.
--->
-
 # Introduction
@@
 That means that zonal asymmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
-@fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
+@fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogeneous nature of the selected years for constructing the composites.
 To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700 hPa and 50 hPa levels (Figures \@ref(fig:A3) and \@ref(fig:A4)).
@@
 ```{r test-aao2}
-# test that my indices hace the correct sign. The Full index needs to be **positively** correlated
+# test that my indices have the correct sign. The Full index needs to be **positively** correlated
 # with the AAO index.
@@
 There is a significant correlation between SAM and ENSO. When divided in trimesters, this correlation is important only in DJF and SON. 
-This relationship is captured mainly by the A\nobreakdash-SAM, as this index has signifcant partial correlations with ENSO while correlations with S\nobreakdash-SAM are all smaller and non-significant.
+This relationship is captured mainly by the A\nobreakdash-SAM, as this index has significant partial correlations with ENSO while correlations with S\nobreakdash-SAM are all smaller and non-significant.
 In MAM, ENSO is not significantly correlated with SAM, but it is significantly correlated with A\nobreakdash-SAM in a level comparable with the ENSO-SAM correlation in SON.
@@
 Figure \@ref(fig:global-pp) shows regression of the SAM indices with precipitation for the Southern Hemisphere.
-The precipitation signal associated with SAM shows generally decreased precipitation at around 45ºS, somewhat increase precipitation around 30ªS (Fig. \@ref(fig:global-pp)a) and increased precipitation over Antarctica, a pattern known from other studies [e.g.
+The precipitation signal associated with SAM shows generally decreased precipitation at around 45\degree\ S, somewhat increase precipitation around 30\degree\S (Fig. \@ref(fig:global-pp)a) and increased precipitation over Antarctica, a pattern known from other studies [e.g.
 @hendon2014].

### 2021-05-28 analysis/paper/asymsam.tex
Render

@@
 That means that zonal asymmetries associated with positive SAM phase (SAM+) are almost opposite in sign and of the same magnitude to the ones associated with negative SAM phase (SAM-).
-\citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogenous nature of the selected years for constructing the composites.
+\citet{fogt2012}'s composites (their Figure 4) suggest that this might not be entirely valid, although much of that apparent non-linearity could be due to the heterogeneous nature of the selected years for constructing the composites.
 To test this assumption, we computed seasonal composites of zonal anomalies of geopotential height for SAM+ and SAM- (defined as months in which the SAM index is greater than 1 standard deviation and lower than minus 1 standard deviation, respectively) for the period from 1979 to 2018 at the 700~hPa and 50~hPa levels (Figures \ref{fig:A3} and \ref{fig:A4}).
@@
 There is a significant correlation between SAM and ENSO. When divided in trimesters, this correlation is important only in DJF and SON.
-This relationship is captured mainly by the A\nobreakdash-SAM, as this index has signifcant partial correlations with ENSO while correlations with S\nobreakdash-SAM are all smaller and non-significant.
+This relationship is captured mainly by the A\nobreakdash-SAM, as this index has significant partial correlations with ENSO while correlations with S\nobreakdash-SAM are all smaller and non-significant.
 In MAM, ENSO is not significantly correlated with SAM, but it is significantly correlated with A\nobreakdash-SAM in a level comparable with the ENSO-SAM correlation in SON.
@@
 Moreover, from Autumn to Spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM.
-In agreement, \citet{marshall2016} found the same signal in association with SAM while that pattern is only evident in Autumm for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.
+In agreement, \citet{marshall2016} found the same signal in association with SAM while that pattern is only evident in Autumn for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.
 
@@
 Figure~\ref{fig:global-pp} shows regression of the SAM indices with precipitation for the Southern Hemisphere.
-The precipitation signal associated with SAM shows generally decreased precipitation at around 45ºS, somewhat increase precipitation around 30ªS (Fig.~\ref{fig:global-pp}a) and increased precipitation over Antarctica, a pattern known from other studies \citep[e.g.][]{hendon2014}.
+The precipitation signal associated with SAM shows generally decreased precipitation at around 45\degree~S, somewhat increase precipitation around 30\degree\S (Fig.~\ref{fig:global-pp}a) and increased precipitation over Antarctica, a pattern known from other studies \citep[e.g.][]{hendon2014}.
 This pattern is mostly unchanged between seasons albeit varying in intensity (not shown).

### 2021-05-29 analysis/paper/asymsam.Rmd
Arregla caption de figura metodos

@@
 
-```{r method, fig.cap = "(ref:method-cap", fig.width=6, fig.height=2.7, fig.env="figure*"}
+```{r method, fig.cap = "(ref:method-cap)", fig.width=6, fig.height=2.7, fig.env="figure*"}
 sam_sep <- SAM[, eof[[1]]$right, by = .(lev)] %>% 

### 2021-05-29 analysis/paper/asymsam.tex
Render

@@
 \begin{figure*}
-\includegraphics{method-1} \caption{(ref:method-cap}\label{fig:method}
+\includegraphics{method-1} \caption{Spatial patterns of the first EOF of 700~hPa geopotential height for 1979 -- 2018 period.
+(a) Full field, (b) zonally asymmetric component and (c) zonally symmetric component.
+Arbitrary units; positive values in blue and negative values in red.}\label{fig:method}
 \end{figure*}

### 2021-07-08 analysis/paper/asymsam.Rmd
Mayúsculas

@@
 
-Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn [e.g. @fogt2020 and references therein].
+Positive trends in the SAM have been documented by various researchers using different indices, mostly in austral summer and autumn [e.g. @fogt2020 and references therein].
 It is thought that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables [@marshall2004; @gillett2005; @arblaster2006; @gillett2013].
@@
 The SAM index presents a statistically positive significant trend (Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "Annual")`) that extends throughout the troposphere up to about 50 hPa and reaches its maximum value at 100 hPa.
-The seasonal trends (rest of Figure \@ref(fig:trends) column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100 hPa maximum is much more defined.
-In Winter and Spring, we detect no statistically significant trend.
-This is consistent with the results of previous studies, which find large positive trends in Summer, smaller in Autumn and no trends in the other seasons [e.g.
+The seasonal trends (rest of Figure \@ref(fig:trends) column a) indicate that positive trends are present in autumn and particularly in summer, where the 100 hPa maximum is much more defined.
+In winter and spring, we detect no statistically significant trend.
+This is consistent with the results of previous studies, which find large positive trends in summer, smaller in autumn and no trends in the other seasons [e.g.
 @fogt2020 and references therein] using indices of the SAM based on surface or near-surface circulation.
@@
 column c in Figure \@ref(fig:trends)), but in some cases the trends become clearer.
-In Summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
-In Autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.
+In summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \@ref(fig:trends)`r get_tag(term == "full" & season == "DJF")` and `r get_tag(term == "sym" & season == "DJF")`).
+In autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.
 
@@
 Figure \@ref(fig:regr-air-season) shows regression coefficients of each index at 700 hPa with surface land air and sea temperature for each trimester.
-In Summer, positive values of the SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies in the mid latitudes.
+In summer, positive values of the SAM index (Figure \@ref(fig:regr-air-season)`r get_tag(season == "DJF" & term == "full")`) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies in the mid latitudes.
 The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E respectively.
@@
 
-In Autumn, Winter and Spring (rows `r combine_words(get_row(season %in% c("MAM", "JJA", "SON"), n = 3))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to Summer.
+In autumn, winter and spring (rows `r combine_words(get_row(season %in% c("MAM", "JJA", "SON"), n = 3))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to summer.
 There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
 Similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
-In Spring, the tropical signal of A\nobreakdash-SAM is similar to the one in Summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link. 
-Moreover, from Autumn to Spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM.
-In agreement, @marshall2016 found the same signal in association with SAM while that pattern is only evident in Autumn for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.
+In spring, the tropical signal of A\nobreakdash-SAM is similar to the one in summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link. 
+Moreover, from autumn to spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM.
+In agreement, @marshall2016 found the same signal in association with SAM while that pattern is only evident in autumn for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.
 
@@
 The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons -- although with varying intensity and small-scale details -- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
-It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in Summer and Spring).
+It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in summer and spring).
 Temperatures over Eastern Antarctica are most impacted by the S\nobreakdash-SAM, while on Western Antarctica they are more sensitive to the A\nobreakdash-SAM.
@@
 Figure \@ref(fig:regr-air-season) column b can be partially compared with Figure 11 from @fogt2012.
-Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.
+Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in summer and spring.
 
@@
 
-The seasonal-based regressions show statistically significant anomalies only in Spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
+The seasonal-based regressions show statistically significant anomalies only in spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
 In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related to positive A\nobreakdash-SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).
 
-In Summer, positive SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \@ref(fig:pp-regr-oceania)a.2).
+In summer, positive SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \@ref(fig:pp-regr-oceania)a.2).
 The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
-In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, which are associated with the A\nobreakdash-SAM.
-In Winter regression coefficients are much weaker than during the rest of the year. 
+In autumn, the regression with SAM shows positive anomalies in the North, similar to summer, which are associated with the A\nobreakdash-SAM.
+In winter regression coefficients are much weaker than during the rest of the year. 
 None of these regression coefficients are statistically significant at the 95% level.
-The Spring signal is broadly consistent with @hendon2007, but whereas they also detected a strong signal in Summer, Figure \@ref(fig:pp-regr-oceania)a.2 shows no statistically significant association (although the coefficients have the consistent sign).
+The spring signal is broadly consistent with @hendon2007, but whereas they also detected a strong signal in summer, Figure \@ref(fig:pp-regr-oceania)a.2 shows no statistically significant association (although the coefficients have the consistent sign).
 
@@
 
-Except Winter, seasonal-based regressions mirror this same pattern.
+Except winter, seasonal-based regressions mirror this same pattern.
 Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the SAM.
@@
 
-We observe positive trends of the SAM in Summer and Autumn, as was documented by previous studies [e.g. @fogt2020 and references therein] for low levels.
+We observe positive trends of the SAM in summer and autumn, as was documented by previous studies [e.g. @fogt2020 and references therein] for low levels.
 We show that these trends maximise at 100 hPa, and are explained by the zonally symmetric component.

### 2021-07-08 analysis/paper/asymsam.Rmd
Mayúsculas

@@
 @gong1999 defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 65\degree S, which is also the definition used by the station-based index in @marshall2003.
-@baldwin2009 proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
+@baldwin2009 proposed defining the Northern and Southern Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
 
@@
 The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at lower latitudes [@jones2019] (and vice versa for negative SAM).
-But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012].
+But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the south Atlantic [@fogt2012].
 The SAM-related signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry [@lim2016].
@@
 Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \@ref(fig:wave-amplitude) column `r get_col(term == "full")`) are dominated by the zonal mean (wavenumber 0) at both levels.
-However, zonal waves are important, particularly South of 50\degree S, with wavenumber 1 clearly dominating at 50 hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of waves of similar amplitude at 700 hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`).
+However, zonal waves are important, particularly south of 50\degree S, with wavenumber 1 clearly dominating at 50 hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 50)`) and a mix of waves of similar amplitude at 700 hPa (Figure \@ref(fig:wave-amplitude)`r get_tag(term == "full" & lev == 700)`).
 Figure \@ref(fig:wave-amplitude) column `r get_col(term == "asym")` shows that the A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 50)`), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \@ref(fig:wave-amplitude)`r get_tag(term== "asym" & lev == 700)`) with negligible amplitude of the zonal mean.
@@
 In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250 hPa.
-The anomalies are larger in the Western hemisphere, where they extend into the stratosphere.
-In the Eastern hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere.
+The anomalies are larger in the Western Hemisphere, where they extend into the stratosphere.
+In the Eastern Hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere.
 So, while the tropospheric A\nobreakdash-SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric A\nobreakdash-SAM.
@@
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Note that the colour scale cuts-off at $\pm0.6 \mathrm{K}$ to highlight mid-latitudes and tropics features at the expense of the higher values in Polar regions.
+Note that the colour scale cuts-off at $\pm0.6 \mathrm{K}$ to highlight mid-latitudes and tropics features at the expense of the higher values in polar regions.
 
@@
 Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics.
-Over Antarctica positive values of the SAM index are associated with negative temperature anomalies, particularly over the Eastern coast.
+Over Antarctica positive values of the SAM index are associated with negative temperature anomalies, particularly over the eastern coast.
 These anomalies are associated only with the S\nobreakdash-SAM.
@@
 @marshall2016].
-On the other hand, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.
+On the other hand, temperature anomalies in the Indian ocean, southern Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.
 
 In autumn, winter and spring (rows `r combine_words(get_row(season %in% c("MAM", "JJA", "SON"), n = 3))` in Figure \@ref(fig:regr-air-season)) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to summer.
-There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
+There are also negative anomalies in southern Australia, and positive anomalies over New Zealand and southern South America.
 Similar features were observed in station measurements by @jones2019, although using data from 1957 to 2016.
@@
 It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in summer and spring).
-Temperatures over Eastern Antarctica are most impacted by the S\nobreakdash-SAM, while on Western Antarctica they are more sensitive to the A\nobreakdash-SAM.
+Temperatures over East Antarctica are most impacted by the S\nobreakdash-SAM, while on West Antarctica they are more sensitive to the A\nobreakdash-SAM.
 Considering the positive trend in the S\nobreakdash-SAM index, this might be consistent with temperature trends, which show warming in the West and no trend in the East [@nicolas2014].
@@
 (ref:global-pp-cap) Regression of monthly precipitation anomalies (mm per day, shading) with a) SAM, b) A\nobreakdash-SAM and c) S\nobreakdash-SAM for the 1979 -- 2018 period.
-Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
+Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
 Note that the colour scale cuts-off at $\pm0.25 \mathrm{K}$ to highlight mid- and high-latitude features at the expense of the very high values in the Tropics.
@@
 ```{r pp-regr-oceania, dependson="def-plot_pp",  fig.cap = "Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700~hPa geopotential height anomalies (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) SAM, (column (b) A\\nobreakdash-SAM and (column c) S\\nobreakdash-SAM for the 1979 -- 2018 period.
-Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
+Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.", fig.width = 6, fig.height = 6.5, fig.env = "figure*"}
 plot_pp("oceania") 
@@
 In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the Southeastern region (Figure \@ref(fig:pp-regr-oceania)a.1), in agreement with @gillett2006.
-The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure \@ref(fig:pp-regr-oceania)c.1).
+The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the east coast (Figure \@ref(fig:pp-regr-oceania)c.1).
 Geopotential height anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by @hendon2007.
-A\nobreakdash-SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
-
-The seasonal-based regressions show statistically significant anomalies only in spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
-In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related to positive A\nobreakdash-SAM affect all Eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).
-
-In summer, positive SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \@ref(fig:pp-regr-oceania)a.2).
-The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
-In autumn, the regression with SAM shows positive anomalies in the North, similar to summer, which are associated with the A\nobreakdash-SAM.
+A\nobreakdash-SAM appears related to positive precipitation anomalies in the west coast of southeastern Australia (Figure \@ref(fig:pp-regr-oceania)b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
+
+The seasonal-based regressions show statistically significant anomalies only in spring, when positive SAM is associated with positive precipitation anomalies in eastern Australia (Figure \@ref(fig:pp-regr-oceania)a.5).
+In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the eastern coast (Figure \@ref(fig:pp-regr-oceania)c.5) while the positive precipitation anomalies related to positive A\nobreakdash-SAM affect all eastern Australia (Figure \@ref(fig:pp-regr-oceania)b.5).
+
+In summer, positive SAM index is associated with positive precipitation anomalies in western and eastern Australia, particularly in the north eastern region (Figure \@ref(fig:pp-regr-oceania)a.2).
+The eastern part being dominated by the relationship with S\nobreakdash-SAM and the western, by A\nobreakdash-SAM.
+In autumn, the regression with SAM shows positive anomalies in the north, similar to summer, which are associated with the A\nobreakdash-SAM.
 In winter regression coefficients are much weaker than during the rest of the year. 
@@
 
-In South America (Figure \@ref(fig:pp-regr-america)), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile, and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
+In South America (Figure \@ref(fig:pp-regr-america)), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and southern Chile, and non-significant positive anomalies in southern Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \@ref(fig:pp-regr-america)a.1).
 Figure \@ref(fig:pp-regr-america)b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with A\nobreakdash-SAM, that in southern Chile is related to S\nobreakdash-SAM.
@@
 
-The anomalous circulation at 700 hPa associated with S\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over Southern Chile.
+The anomalous circulation at 700 hPa associated with S\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)c.1) indicate anomalous easterly flow over southern Chile.
 This leads to reduced influx of moist air from the Pacific Ocean which is the main source of precipitable water in that region [e.g.
 @garreaud2007].
-On the other hand, the anomalous circulation associated with positive values of A\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
-This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region [@silvestri2009; @zamboni2010].
-This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events [@rosso2018].
+On the other hand, the anomalous circulation associated with positive values of A\nobreakdash-SAM (Figure \@ref(fig:pp-regr-america)b.1) in the Atlantic is anticyclonic in the south and cyclonic in the north.
+This promotes anomalous eouth-easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region [@silvestri2009; @zamboni2010].
+This same pattern was found to be associated with increased precipitation in southern Brazil during SACZ events [@rosso2018].
 There is a small area of significant positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by @gillett2006 that is explained by the A\nobreakdash-SAM.
@@
 The A\nobreakdash-SAM index explains a big portion of these deviations.
-In particular, the positive phase of A\nobreakdash-SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO\nobreakdash-SAM relationship.
+In particular, the positive phase of A\nobreakdash-SAM is associated with colder temperatures over southern Brazil, southern Africa and southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO\nobreakdash-SAM relationship.
 These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active [e.g. @cai2020a].
 
-In Australia the SAM is associated with positive precipitation anomalies in South East and this is explained by the S\nobreakdash-SAM.
-However, the A\nobreakdash-SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
-
-In South America, precipitation anomalies associated with the SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
+In Australia the SAM is associated with positive precipitation anomalies in the south east and this is explained by the S\nobreakdash-SAM.
+However, the A\nobreakdash-SAM is associated with a small area of positive precipitation anomalies in the eastern coast of western Australia, maybe due to advection of moist air from the Indian Ocean.
+
+In South America, precipitation anomalies associated with the SAM are negative both in southern Chile and Southeastern South America, and positive in southern Brazil.
 These features are cleanly separated between A\nobreakdash-SAM and S\nobreakdash-SAM.
-S\nobreakdash-SAM explains the negative anomalies in Southern Chile and A\nobreakdash-SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
+S\nobreakdash-SAM explains the negative anomalies in southern Chile and A\nobreakdash-SAM, the negative-positive dipole between Southeastern South America and southern Brazil.
 Individual seasons mostly follow this pattern.

### 2021-07-08 analysis/paper/asymsam.Rmd
Clarifica tema antártida - S-SAM - calentamiento global

@@
 Temperatures over East Antarctica are most impacted by the S\nobreakdash-SAM, while on West Antarctica they are more sensitive to the A\nobreakdash-SAM.
-Considering the positive trend in the S\nobreakdash-SAM index, this might be consistent with temperature trends, which show warming in the West and no trend in the East [@nicolas2014].
+Since the S\nobreakdash-SAM index is negatively correlated with temperatures over East Antarctica, it is possible that the positive trend in the S\nobreakdash-SAM index might help explain the lack of positive temperature trend in East Antarctica compared with West Antarctica in the context of global warming [@nicolas2014].
 

### 2021-07-08 analysis/paper/asymsam.tex
Render

@@
 \citet{gong1999} defined the SAM index as the zonal mean sea level pressure difference between 40\degree S and 65\degree S, which is also the definition used by the station-based index in \citet{marshall2003}.
-\citet{baldwin2009} proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
+\citet{baldwin2009} proposed defining the Northern and Southern Annular modes as the leading EOF of the zonally averaged geopotential height at each level.
 
@@
 
-Positive trends in the SAM have been documented by various researchers using different indices, mostly on austral Summer and Autumn \citep[e.g.][ and references therein]{fogt2020}.
+Positive trends in the SAM have been documented by various researchers using different indices, mostly in austral summer and autumn \citep[e.g.][ and references therein]{fogt2020}.
 It is thought that these trends are driven primarily by stratospheric ozone depletion and the increase in greenhouse gases, and understood in the context of zonal mean variables \citep{marshall2004, gillett2005, arblaster2006, gillett2013}.
@@
 The positive phase of the SAM is associated with colder-than-normal temperatures over Antarctica and warmer-than-normal temperatures at lower latitudes \citep{jones2019} (and vice versa for negative SAM).
-But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}.
+But there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the south Atlantic \citep{fogt2012}.
 The SAM-related signal on precipitation anomalies behaves similarly, although with even greater deviation from zonal symmetry \citep{lim2016}.
@@
 The SAM index presents a statistically positive significant trend (Figure \ref{fig:trends}a.1) that extends throughout the troposphere up to about 50~hPa and reaches its maximum value at 100~hPa.
-The seasonal trends (rest of Figure \ref{fig:trends} column a) indicate that positive trends are present in Autumn and particularly in Summer, where the 100~hPa maximum is much more defined.
-In Winter and Spring, we detect no statistically significant trend.
-This is consistent with the results of previous studies, which find large positive trends in Summer, smaller in Autumn and no trends in the other seasons \citep[e.g.][ and references therein]{fogt2020} using indices of the SAM based on surface or near-surface circulation.
+The seasonal trends (rest of Figure \ref{fig:trends} column a) indicate that positive trends are present in autumn and particularly in summer, where the 100~hPa maximum is much more defined.
+In winter and spring, we detect no statistically significant trend.
+This is consistent with the results of previous studies, which find large positive trends in summer, smaller in autumn and no trends in the other seasons \citep[e.g.][ and references therein]{fogt2020} using indices of the SAM based on surface or near-surface circulation.
 
 By separating the SAM signal in its asymmetric and symmetric parts, not only we can see that these trends are almost entirely due to the symmetric component (column b vs.~column c in Figure \ref{fig:trends}), but in some cases the trends become clearer.
-In Summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \ref{fig:trends}a.2 and c.2).
-In Autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.
+In summer, A\nobreakdash-SAM has a statistically non significant negative trend in the middle troposphere that obscures the trend in the SAM index; as a result, trends computed using only the Symmetric component are stronger (compare the shading region in Figure \ref{fig:trends}a.2 and c.2).
+In autumn, the S\nobreakdash-SAM index reveals a statistically significant positive trend in the stratosphere that is not significant using the SAM index.
 
@@
 Zonal wave amplitudes of the spatial pattern described by the SAM index (Figure \ref{fig:wave-amplitude} column a) are dominated by the zonal mean (wavenumber 0) at both levels.
-However, zonal waves are important, particularly South of 50\degree S, with wavenumber 1 clearly dominating at 50~hPa (Figure \ref{fig:wave-amplitude}a.1) and a mix of waves of similar amplitude at 700~hPa (Figure \ref{fig:wave-amplitude}a.2).
+However, zonal waves are important, particularly south of 50\degree S, with wavenumber 1 clearly dominating at 50~hPa (Figure \ref{fig:wave-amplitude}a.1) and a mix of waves of similar amplitude at 700~hPa (Figure \ref{fig:wave-amplitude}a.2).
 Figure \ref{fig:wave-amplitude} column b shows that the A\nobreakdash-SAM is overwhelmingly dominated by wave 1 in the stratosphere (Figure \ref{fig:wave-amplitude}b.1), while in the troposphere it is explained by a combination of zonal waves 3 to 1 in decreasing level of importance (Figure \ref{fig:wave-amplitude}b.2) with negligible amplitude of the zonal mean.
@@
 In the troposphere, the wave-3 structure is equivalent barotropic with maximum amplitude at roughly 250~hPa.
-The anomalies are larger in the Western hemisphere, where they extend into the stratosphere.
-In the Eastern hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere.
+The anomalies are larger in the Western Hemisphere, where they extend into the stratosphere.
+In the Eastern Hemisphere the wave-3 signal is smaller and confined to the troposphere while negative anomalies dominate in the stratosphere.
 So, while the tropospheric A\nobreakdash-SAM index is associated with stratospheric geopotential anomalies, these do not project strongly onto the stratospheric A\nobreakdash-SAM.
@@
 Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
-Note that the colour scale cuts-off at \(\pm0.6 \mathrm{K}\) to highlight mid-latitudes and tropics features at the expense of the higher values in Polar regions.}\label{fig:regr-air-season}
+Note that the colour scale cuts-off at \(\pm0.6 \mathrm{K}\) to highlight mid-latitudes and tropics features at the expense of the higher values in polar regions.}\label{fig:regr-air-season}
 \end{figure*}
@@
 Figure \ref{fig:regr-air-season} shows regression coefficients of each index at 700~hPa with surface land air and sea temperature for each trimester.
-In Summer, positive values of the SAM index (Figure \ref{fig:regr-air-season}a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies in the mid latitudes.
+In summer, positive values of the SAM index (Figure \ref{fig:regr-air-season}a.1) are associated with negative temperature anomalies near Antarctica which are surrounded by a ring of positive anomalies in the mid latitudes.
 The ring is not zonally symmetric, as there are four distinctive local maximums around 30\degree W, 120\degree W, 150\degree E and 90\degree E respectively.
@@
 Both the local maximums in the ring and the anomalies in the Pacific regions are present mostly on the A\nobreakdash-SAM regression map, while temperature patterns linked to positive S\nobreakdash-SAM show a more zonally consistent ring and less relation to the tropics.
-Over Antarctica positive values of the SAM index are associated with negative temperature anomalies, particularly over the Eastern coast.
+Over Antarctica positive values of the SAM index are associated with negative temperature anomalies, particularly over the eastern coast.
 These anomalies are associated only with the S\nobreakdash-SAM.
 Furthermore, the typical longitudinal variations of temperature anomalies along the Antarctic Peninsula are not evident in the regressions with the SAM, in agreement with previous works \citep[e.g.][]{marshall2016}.
-On the other hand, temperature anomalies in the Indian ocean, South Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.
+On the other hand, temperature anomalies in the Indian ocean, southern Africa and Australia are strongly related to A\nobreakdash-SAM and are not present in the regression pattern with the SAM.
 
-In Autumn, Winter and Spring (rows 2, 3, and 4 in Figure \ref{fig:regr-air-season}) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to Summer.
-There are also negative anomalies in Southern Australia, and positive anomalies over New Zealand and Southern South America.
+In autumn, winter and spring (rows 2, 3, and 4 in Figure \ref{fig:regr-air-season}) the positive ring is only present through its local maximums in the regression with the SAM, which reflects the more asymmetric nature of the SAM compared to summer.
+There are also negative anomalies in southern Australia, and positive anomalies over New Zealand and southern South America.
 Similar features were observed in station measurements by \citet{jones2019}, although using data from 1957 to 2016.
-In Spring, the tropical signal of A\nobreakdash-SAM is similar to the one in Summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link.
-Moreover, from Autumn to Spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM.
-In agreement, \citet{marshall2016} found the same signal in association with SAM while that pattern is only evident in Autumn for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.
+In spring, the tropical signal of A\nobreakdash-SAM is similar to the one in summer, again revealing the significance of the ENSO-A\nobreakdash-SAM link.
+Moreover, from autumn to spring, positive (negative) temperature anomalies are discernible at the northern (southern) portion of the Antarctic Peninsula in the regressions with the SAM, which are also evident in the regressions with A-SAM.
+In agreement, \citet{marshall2016} found the same signal in association with SAM while that pattern is only evident in autumn for the southern baroclinic annular mode, which is associated with variations in extratropical storm amplitude.
 
@@
 The pattern of negative anomalies in the pole surrounded by positive anomalies roughly seen in all seasons -- although with varying intensity and small-scale details -- translates to enhanced meridional temperature gradient maximised in the zero line, which is consistent with the intensification and poleward migration of the westerlies commonly linked to the SAM through thermal wind balance.
-It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in Summer and Spring).
-Temperatures over Eastern Antarctica are most impacted by the S\nobreakdash-SAM, while on Western Antarctica they are more sensitive to the A\nobreakdash-SAM.
-Considering the positive trend in the S\nobreakdash-SAM index, this might be consistent with temperature trends, which show warming in the West and no trend in the East \citep{nicolas2014}.
+It's then not surprising to see it more clearly in association with S\nobreakdash-SAM (at least in summer and spring).
+Temperatures over East Antarctica are most impacted by the S\nobreakdash-SAM, while on West Antarctica they are more sensitive to the A\nobreakdash-SAM.
+Since the S\nobreakdash-SAM index is negatively correlated with temperatures over East Antarctica, it is possible that the positive trend in the S\nobreakdash-SAM index might help explain the lack of positive temperature trend in East Antarctica compared with West Antarctica in the context of global warming \citep{nicolas2014}.
 
 Figure \ref{fig:regr-air-season} column b can be partially compared with Figure 11 from \citet{fogt2012}.
-Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in Summer and Spring.
+Although they used station data from 1958 to 2001, main features are reproduced here, such as the strong signal in New Zealand and Australia in summer and spring.
 
@@
 \includegraphics{global-pp-1} \caption{Regression of monthly precipitation anomalies (mm per day, shading) with a) SAM, b) A\nobreakdash-SAM and c) S\nobreakdash-SAM for the 1979 -- 2018 period.
-Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
+Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.
 Note that the colour scale cuts-off at \(\pm0.25 \mathrm{K}\) to highlight mid- and high-latitude features at the expense of the very high values in the Tropics.}\label{fig:global-pp}
@@
 \includegraphics{pp-regr-oceania-1} \caption{Regression of (row 1) annual and (rows 2 to 5) seasonal mean precipitation anomalies (mm per day, shading) and 700~hPa geopotential height anomalies (thin lines, positive values as solid lines and negative values as dashed lines) with (column a) SAM, (column (b) A\nobreakdash-SAM and (column c) S\nobreakdash-SAM for the 1979 -- 2018 period.
-Thin lines are the Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
+Black contours indicate areas with p-value smaller than 0.05 controlling for False Detection Rate.}\label{fig:pp-regr-oceania}
 \end{figure*}
@@
 In Australia, the annual regression shows that the SAM index is associated with positive precipitation anomalies in the Southeastern region (Figure \ref{fig:pp-regr-oceania}a.1), in agreement with \citet{gillett2006}.
-The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the East coast (Figure \ref{fig:pp-regr-oceania}c.1).
+The separation between A\nobreakdash-SAM and S\nobreakdash-SAM suggest that this positive anomaly is explained by the S\nobreakdash-SAM only in the east coast (Figure \ref{fig:pp-regr-oceania}c.1).
 Geopotential height anomalies associated with this index (black contours) are indicative of easterly flow from the Tasman Sea, which could explain the positive anomalies in precipitation as found by \citet{hendon2007}.
-A\nobreakdash-SAM appears related to positive precipitation anomalies in the West coast of Southeastern Australia (Figure \ref{fig:pp-regr-oceania}b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
+A\nobreakdash-SAM appears related to positive precipitation anomalies in the west coast of southeastern Australia (Figure \ref{fig:pp-regr-oceania}b.2), which could similarly be explained by the anomalous westerly circulation transporting moist air to the continent from the Indian Ocean.
 
-The seasonal-based regressions show statistically significant anomalies only in Spring, when positive SAM is associated with positive precipitation anomalies in Eastern Australia (Figure \ref{fig:pp-regr-oceania}a.5).
-In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the East Coast (Figure \ref{fig:pp-regr-oceania}c.5) while the positive precipitation anomalies related to positive A\nobreakdash-SAM affect all Eastern Australia (Figure \ref{fig:pp-regr-oceania}b.5).
+The seasonal-based regressions show statistically significant anomalies only in spring, when positive SAM is associated with positive precipitation anomalies in eastern Australia (Figure \ref{fig:pp-regr-oceania}a.5).
+In this trimester, S\nobreakdash-SAM seems to be associated with positive precipitation anomalies in a relatively reduced area of the eastern coast (Figure \ref{fig:pp-regr-oceania}c.5) while the positive precipitation anomalies related to positive A\nobreakdash-SAM affect all eastern Australia (Figure \ref{fig:pp-regr-oceania}b.5).
 
-In Summer, positive SAM index is associated with positive precipitation anomalies in Western and Eastern Australia, particularly in the North East (Figure \ref{fig:pp-regr-oceania}a.2).
-The Eastern part being dominated by the relationship with S\nobreakdash-SAM and the Western, by A\nobreakdash-SAM.
-In Autumn, the regression with SAM shows positive anomalies in the North, similar to Summer, which are associated with the A\nobreakdash-SAM.
-In Winter regression coefficients are much weaker than during the rest of the year.
+In summer, positive SAM index is associated with positive precipitation anomalies in western and eastern Australia, particularly in the north eastern region (Figure \ref{fig:pp-regr-oceania}a.2).
+The eastern part being dominated by the relationship with S\nobreakdash-SAM and the western, by A\nobreakdash-SAM.
+In autumn, the regression with SAM shows positive anomalies in the north, similar to summer, which are associated with the A\nobreakdash-SAM.
+In winter regression coefficients are much weaker than during the rest of the year.
 None of these regression coefficients are statistically significant at the 95\% level.
-The Spring signal is broadly consistent with \citet{hendon2007}, but whereas they also detected a strong signal in Summer, Figure \ref{fig:pp-regr-oceania}a.2 shows no statistically significant association (although the coefficients have the consistent sign).
+The spring signal is broadly consistent with \citet{hendon2007}, but whereas they also detected a strong signal in summer, Figure \ref{fig:pp-regr-oceania}a.2 shows no statistically significant association (although the coefficients have the consistent sign).
 
@@
 
-In South America (Figure \ref{fig:pp-regr-america}), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and Southern Chile, and non-significant positive anomalies in South Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \ref{fig:pp-regr-america}a.1).
+In South America (Figure \ref{fig:pp-regr-america}), regression using all seasons shows that positive SAM is associated with statistically significant negative precipitation anomalies in Southeastern South America (SESA) and southern Chile, and non-significant positive anomalies in southern Brazil, near the South Atlantic Convergence Zone (SACZ) (Figure \ref{fig:pp-regr-america}a.1).
 Figure \ref{fig:pp-regr-america}b.1 and c.1 show that while the signal over SESA and southern Brazil is associated with A\nobreakdash-SAM, that in southern Chile is related to S\nobreakdash-SAM.
 
-Except Winter, seasonal-based regressions mirror this same pattern.
+Except winter, seasonal-based regressions mirror this same pattern.
 Even if not statistically significant, they all show negative values in SESA and southern Chile along with positive values in southern Brazil in relation with the SAM.
@@
 
-The anomalous circulation at 700~hPa associated with S\nobreakdash-SAM (Figure \ref{fig:pp-regr-america}c.1) indicate anomalous easterly flow over Southern Chile.
+The anomalous circulation at 700~hPa associated with S\nobreakdash-SAM (Figure \ref{fig:pp-regr-america}c.1) indicate anomalous easterly flow over southern Chile.
 This leads to reduced influx of moist air from the Pacific Ocean which is the main source of precipitable water in that region \citep[e.g.][]{garreaud2007}.
-On the other hand, the anomalous circulation associated with positive values of A\nobreakdash-SAM (Figure \ref{fig:pp-regr-america}b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North.
-This promotes anomalous South-Easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region \citep{silvestri2009, zamboni2010}.
-This same pattern was found to be associated with increased precipitation in Southern Brazil during SACZ events \citep{rosso2018}.
+On the other hand, the anomalous circulation associated with positive values of A\nobreakdash-SAM (Figure \ref{fig:pp-regr-america}b.1) in the Atlantic is anticyclonic in the south and cyclonic in the north.
+This promotes anomalous eouth-easterly flow over SESA, which inhibits the flow of the South America Low-Level Jet to the region \citep{silvestri2009, zamboni2010}.
+This same pattern was found to be associated with increased precipitation in southern Brazil during SACZ events \citep{rosso2018}.
 There is a small area of significant positive precipitation anomalies with the SAM near central Argentina which is also present in the station-based analysis by \citet{gillett2006} that is explained by the A\nobreakdash-SAM.
@@
 
-We observe positive trends of the SAM in Summer and Autumn, as was documented by previous studies \citep[e.g.][ and references therein]{fogt2020} for low levels.
+We observe positive trends of the SAM in summer and autumn, as was documented by previous studies \citep[e.g.][ and references therein]{fogt2020} for low levels.
 We show that these trends maximise at 100~hPa, and are explained by the zonally symmetric component.
@@
 The A\nobreakdash-SAM index explains a big portion of these deviations.
-In particular, the positive phase of A\nobreakdash-SAM is associated with colder temperatures over Southern Brazil, South Africa and Southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO\nobreakdash-SAM relationship.
+In particular, the positive phase of A\nobreakdash-SAM is associated with colder temperatures over southern Brazil, southern Africa and southern Australia, as well as the negative anomalies in the equatorial Pacific consistent with the ENSO\nobreakdash-SAM relationship.
 These negative anomalies are particularly clear in the DJF and SON trimesters, which include the months in which the ENSO teleconnection is more active \citep[e.g.][]{cai2020a}.
 
-In Australia the SAM is associated with positive precipitation anomalies in South East and this is explained by the S\nobreakdash-SAM.
-However, the A\nobreakdash-SAM is associated with a small area of positive precipitation anomalies in the Eastern Coast of West Australia, maybe due to advection of moist air from the Indian Ocean.
+In Australia the SAM is associated with positive precipitation anomalies in the south east and this is explained by the S\nobreakdash-SAM.
+However, the A\nobreakdash-SAM is associated with a small area of positive precipitation anomalies in the eastern coast of western Australia, maybe due to advection of moist air from the Indian Ocean.
 
-In South America, precipitation anomalies associated with the SAM are negative both in Southern Chile and Southeastern South America, and positive in Southern Brazil.
+In South America, precipitation anomalies associated with the SAM are negative both in southern Chile and Southeastern South America, and positive in southern Brazil.
 These features are cleanly separated between A\nobreakdash-SAM and S\nobreakdash-SAM.
-S\nobreakdash-SAM explains the negative anomalies in Southern Chile and A\nobreakdash-SAM, the negative-positive dipole between Southeastern South America and Southern Brazil.
+S\nobreakdash-SAM explains the negative anomalies in southern Chile and A\nobreakdash-SAM, the negative-positive dipole between Southeastern South America and southern Brazil.
 Individual seasons mostly follow this pattern.

### 2021-07-13 analysis/paper/asymsam.Rmd
Render

--- 2021-07-08 analysis/paper/asymsam.Rmd
@@
   cache = TRUE,
-  cache.extra = 41,
+  cache.extra = 43,
   cache.path = "../cache/",
@@
 Temperatures over East Antarctica are most impacted by the S\nobreakdash-SAM, while on West Antarctica they are more sensitive to the A\nobreakdash-SAM.
-Considering the positive trend in the S\nobreakdash-SAM index, this might be consistent with temperature trends, which show warming in the West and no trend in the East [@nicolas2014].
+Since the S\nobreakdash-SAM index is negatively correlated with temperature over East Antarctica, it is possible that the positive trend in the S\nobreakdash-SAM index might help explain the lack of positive temperature trend in East Antarctica compared with West Antarctica in the context of global warming [@nicolas2014].
 

### 2021-07-13 analysis/paper/asymsam.tex
Render

@@
 Temperatures over East Antarctica are most impacted by the S\nobreakdash-SAM, while on West Antarctica they are more sensitive to the A\nobreakdash-SAM.
-Since the S\nobreakdash-SAM index is negatively correlated with temperatures over East Antarctica, it is possible that the positive trend in the S\nobreakdash-SAM index might help explain the lack of positive temperature trend in East Antarctica compared with West Antarctica in the context of global warming \citep{nicolas2014}.
+Since the S\nobreakdash-SAM index is negatively correlated with temperature over East Antarctica, it is possible that the positive trend in the S\nobreakdash-SAM index might help explain the lack of positive temperature trend in East Antarctica compared with West Antarctica in the context of global warming \citep{nicolas2014}.
 

### 2021-08-11 website/index.Rmd
Updates website

@@
 
-For more details, see @campitelli2021.
+For more details, see @campitelli2021 and [the companion blog post](https://eliocamp.github.io/scrapbook/posts/2021-08-09-asymsam/).
 
@@
 
-Campitelli, E., Diaz, L., & Vera, C. (2021). Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach. Climate Dynamics, (under review).
+Campitelli, E., Díaz, L.
+B., & Vera, C.
+(2021).
+Assessment of zonally symmetric and asymmetric components of the Southern Annular Mode using a novel approach.
+Climate Dynamics.
+https://doi.org/10.1007/s00382-021-05896-5
 
@@
   title = {Assessment of Zonally Symmetric and Asymmetric Components of the {{Southern Annular Mode}} Using a Novel Approach},
-  author = {Campitelli, Elio and Diaz, Leandro and Vera, Carolina},
-  year = {2021},
-  volume = {(under review)},
-  journal = {Climate Dynamics}
+  author = {Campitelli, Elio and Díaz, Leandro B. and Vera, Carolina},
+  date = {2021-08-09},
+  journaltitle = {Climate Dynamics},
+  shortjournal = {Clim Dyn},
+  issn = {1432-0894},
+  doi = {10.1007/s00382-021-05896-5},
+  url = {https://doi.org/10.1007/s00382-021-05896-5}
 }
+
 ```

### 2023-03-28 analysis/paper/asymsam.Rmd
Adds pure html output

@@
 bibliography: 
-  - AsymSAM
-  - packages
+  - AsymSAM.bib
+  - packages.bib
 bibstyle: spbasic
@@
 output: 
+  # bookdown::html_document2:
+  #     code_folding: "hide"
+
   bookdown::pdf_book: 
@@
       # pandoc_args: ["--lua-filter=abstract-to-meta.lua"]
+
 header-includes: 
@@
   comment = "#>",
-  dev = c("pdf", "png"),
+  dev = if (knitr::is_html_output()) c("png") else c("pdf", "png"),
   dpi = 300,
